### 2017-12-04 submission/manuscript.Rmd
upload template file for manuscript

---
title: ''
csl: peerj.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr"))

```

```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

```

# Insert Title Here

\begin{center}
\vspace{25mm}

Marc A Sze${^1}$ and Patrick D Schloss${^1}$${^\dagger}$

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\end{center}

Co-author e-mails:

* marcsze@med.umich.edu

\newpage
\linenumbers

## Abstract

**Background.**

**Methods.**

**Results.**

**Conclusions.**

\newpage

## Introduction

\newpage

## Materials & Methods

***Insert Sub section here***

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017.

\newpage

## Results

***Insert Sub title here***

\newpage

## Discussion

\newpage

## Conclusion

\newpage

## Acknowledgements

The authors would like to thank all the study participants who were a part of each of the individual studies utilized. We would also like to thank each of the study authors for making their data available for use. Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.

\newpage

## References

<div id="refs"></div>

\newpage

**Table 1: **

\footnotesize

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}

```

\normalsize
\newpage

**Table 2: **

\footnotesize

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}

```

\normalsize
\newpage

**Figure 1: .**

**Figure 2: .**

**Figure 3: .**

**Figure 4: .**

**Figure 5: .**

**Figure 6: .**

**Figure 7: .**

\newpage

**Figure S1: .**

**Figure S2: .**

**Figure S3: .**

**Figure S4: .**

**Figure S5: .**

**Figure S6: .**

**Figure S7: .**

\newpage

### 2017-12-04 submission/manuscript.tex
upload template file for manuscript

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\usepackage{amsmath}

\usepackage{mathtools}

\begin{document}

\section{Insert Title Here}\label{insert-title-here}

\begin{center}
\vspace{25mm}

Marc A Sze${^1}$ and Patrick D Schloss${^1}$${^\dagger}$

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\end{center}

Co-author e-mails:

\begin{itemize}
\tightlist
\item
  \href{mailto:marcsze@med.umich.edu}{\nolinkurl{marcsze@med.umich.edu}}
\end{itemize}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

\textbf{Background.}

\textbf{Methods.}

\textbf{Results.}

\textbf{Conclusions.}

\newpage

\subsection{Introduction}\label{introduction}

\newpage

\subsection{Materials \& Methods}\label{materials-methods}

\textbf{\emph{Insert Sub section here}}

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.

\newpage

\subsection{Results}\label{results}

\textbf{\emph{Insert Sub title here}}

\newpage

\subsection{Discussion}\label{discussion}

\newpage

\subsection{Conclusion}\label{conclusion}

\newpage

\subsection{Acknowledgements}\label{acknowledgements}

The authors would like to thank all the study participants who were a
part of each of the individual studies utilized. We would also like to
thank each of the study authors for making their data available for use.
Finally we would like to thank the members of the Schloss lab for
valuable feed back and proof reading during the formulation of this
manuscript.

\newpage

\subsection{References}\label{references}

\hypertarget{refs}{}

\newpage

\textbf{Table 1: }

\footnotesize

\normalsize
\newpage

\textbf{Table 2: }

\footnotesize

\normalsize
\newpage

\textbf{Figure 1: .}

\textbf{Figure 2: .}

\textbf{Figure 3: .}

\textbf{Figure 4: .}

\textbf{Figure 5: .}

\textbf{Figure 6: .}

\textbf{Figure 7: .}

\newpage

\textbf{Figure S1: .}

\textbf{Figure S2: .}

\textbf{Figure S3: .}

\textbf{Figure S4: .}

\textbf{Figure S5: .}

\textbf{Figure S6: .}

\textbf{Figure S7: .}

\newpage

\end{document}

### 2017-12-04 code/run_seq_error_table_creations.R
realized I don't need to separate by taq

@@ -167,12 +167,19 @@ get_summary_data <- function(dataList, depth){
 
 # Create vector with different subsample levels
 sub_sample_level <- c(50, 100, 500, 1000, 5000, 10000)
+#taqs_used <- c("acc", "k", "phu", "pl", "q5")
+
 
 # Read in needed count files
 count_table <- read_data("data/process/", "mock_error", ".count_table", "")
 
+#good_counts <- sapply(taqs_used, 
+#                      function(x) read_data("data/process/", x, ".count_table", "_data_only"), simplify = F)
+
 error_summary <- read_data("data/process/", "mock_error", ".summary", "")
 
+#good_error <- sapply(taqs_used, 
+#                      function(x) read_data("data/process/", x, ".summary", "_data_only"), simplify = F)
 
 # Read in master meta data file
 metadata <- read_csv("data/process/tables/meta_data.csv")
@@ -208,6 +215,18 @@ sapply(c(1:length(good_summary_data)),
 ###########################################################################################################################
 ############################### Run actual analysis for actual seq error ##################################################
 ###########################################################################################################################
+
+# Generate the random sampling 
+full_Recombine <- sapply(sub_sample_level, 
+                        function(x) get_random_sample(x, full_combined, unique(full_combined$full_name)), simplify = F)
+
+names(full_Recombine) <- sub_sample_level  
+
+
+
+
+
+
 
 
 

### 2017-12-04 code/run_seq_error_table_creations.R
add nucleotide substituion type rate

@@ -159,6 +159,52 @@ get_summary_data <- function(dataList, depth){
 }
 
 
+# Function to count the number of sequences type of error
+run_nucleotide_error_check <- function(dataTable){
+  
+  tempData <- dataTable %>% 
+    mutate(at = ifelse(AT == 0, invisible(0), invisible(1)), 
+           ag = ifelse(AG == 0, invisible(0), invisible(1)), 
+           ac = ifelse(AC == 0, invisible(0), invisible(1)), 
+           ta = ifelse(TA == 0, invisible(0), invisible(1)), 
+           tg = ifelse(TG == 0, invisible(0), invisible(1)), 
+           tc = ifelse(TC == 0, invisible(0), invisible(1)), 
+           ga = ifelse(GA == 0, invisible(0), invisible(1)), 
+           gt = ifelse(GT == 0, invisible(0), invisible(1)), 
+           gc = ifelse(GC == 0, invisible(0), invisible(1)), 
+           ca = ifelse(CA == 0, invisible(0), invisible(1)), 
+           ct = ifelse(CT == 0, invisible(0), invisible(1)), 
+           cg = ifelse(CG == 0, invisible(0), invisible(1)))
+  
+  return(tempData)
+}
+
+
+# Function to create individual nucleotide summary data
+get_nucleotide_summary_data <- function(dataList, depth){
+  
+  dataList <- dataList[[as.character(depth)]]
+  
+  tempData <- dataList %>% 
+    bind_rows() %>% 
+    group_by(taq, cycles, sample_name.y) %>% 
+    summarise(at_rate = sum(AT)/sum(total.x), ag_rate = sum(AG)/sum(total.x), 
+              ac_rate = sum(AC)/sum(total.x), ta_rate = sum(TA)/sum(total.x), 
+              tg_rate = sum(TG)/sum(total.x), tc_rate = sum(TC)/sum(total.x), 
+              ga_rate = sum(GA)/sum(total.x), gt_rate = sum(GT)/sum(total.x), 
+              gc_rate = sum(GC)/sum(total.x), ca_rate = sum(CA)/sum(total.x), 
+              ct_rate = sum(CT)/sum(total.x), cg_rate = sum(CG)/sum(total.x))
+      
+      
+      
+      #mean_error = mean(error, na.rm = T), sd_error = sd(error, na.rm = T), 
+      #        chimera_prevalence = sum(chimera)/depth, 
[…]

### 2017-12-04 code/run_seq_error_table_creations.R
complete nucleotide error workflow

@@ -193,14 +193,16 @@ get_nucleotide_summary_data <- function(dataList, depth){
               tg_rate = sum(TG)/sum(total.x), tc_rate = sum(TC)/sum(total.x), 
               ga_rate = sum(GA)/sum(total.x), gt_rate = sum(GT)/sum(total.x), 
               gc_rate = sum(GC)/sum(total.x), ca_rate = sum(CA)/sum(total.x), 
-              ct_rate = sum(CT)/sum(total.x), cg_rate = sum(CG)/sum(total.x))
+              ct_rate = sum(CT)/sum(total.x), cg_rate = sum(CG)/sum(total.x), 
+              at_seq_prev = sum(at)/depth, ag_seq_prev = sum(ag)/depth, 
+              ac_seq_prev = sum(ac)/depth, ta_seq_prev = sum(ta)/depth, 
+              tg_seq_prev = sum(tg)/depth, tc_seq_prev = sum(tc)/depth, 
+              ga_seq_prev = sum(ga)/depth, gt_seq_prev = sum(gt)/depth, 
+              gc_seq_prev = sum(gc)/depth, ca_seq_prev = sum(ca)/depth, 
+              ct_seq_prev = sum(ct)/depth, cg_seq_prev = sum(cg)/depth) %>% 
+    rename(sample_name = sample_name.y)
       
-      
-      
-      #mean_error = mean(error, na.rm = T), sd_error = sd(error, na.rm = T), 
-      #        chimera_prevalence = sum(chimera)/depth, 
-       #       seq_error_prevalence = sum(seq_error)/depth)
-  
+
   return(tempData)
 }
 
@@ -275,9 +277,14 @@ full_Recombine <- lapply(full_Recombine,
 # Generate Nucleotide Summary Data
 nucleotide_summary_data <- sapply(sub_sample_level, 
                             function(x) get_nucleotide_summary_data(full_Recombine, x), simplify = F)
+names(nucleotide_summary_data) <- sub_sample_level
 
 
-
+# Write out summarized data tables for graphing
+sapply(c(1:length(nucleotide_summary_data)), 
+       function(x) write_csv(nucleotide_summary_data[[x]], 
+                             paste("data/process/tables/nucleotide_error_", sub_sample_level[x], 
+                                   "_summary.csv", sep = "")))
 
 
 

### 2017-12-05 code/make_nucleotide_graphs.R
first commit

@@ -1,0 +1,97 @@
+### Create graphs of the number of sequences with a chimera based on cycle number and sub sampling
+### Mock Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "viridis", "gridExtra"))
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = ""))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+
+# Function to gatger the data
+run_data_gather <- function(i, dataList){
+  
+  tempData <- dataList[[i]] %>% 
+    gather("sub_type", "rate", contains("_rate")) %>% 
+    mutate(sub_type = str_replace(sub_type, "_rate", "")) %>% 
+    gather("prev_subtype", "seq_prevalence", contains("_seq_prev")) %>% 
+    mutate(prev_subtype = str_replace(prev_subtype, "_seq_prev", ""))
+  
+  return(tempData)
+  
+}
+
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+
+# Vector of sub samples used
+sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+
+# Read in the count data
+nucleotide_data <- sapply(sub_sample_level, 
+                     function(x) read_data("data/process/tables/nucleotide_error_", "", "_summary.csv", x), 
+                     simplify = F)
+
+# Transform data for better graphing
+nucleotide_data <- sapply(sub_sample_level, 
+               function(x) run_data_gather(x, nucleotide_data), simplify = F)
+
+
+
[…]

### 2017-12-05 code/make_nucleotide_graphs.R
complete sub rate work flow

@@ -60,7 +60,7 @@ nucleotide_data <- sapply(sub_sample_level,
 
 
 # Generate graph of Mock DNA samples 
-nucleotide_data[["1000"]] %>% 
+thousand <- nucleotide_data[["1000"]] %>% 
   mutate(sub_type = factor(sub_type, 
                        levels = c("ac", "ag", "at", "ca", "cg", "ct", "ga", "gc", "gt", "ta", "tc", "tg"), 
                        labels = c("A => C", "A => G", "A => T", "C => A", "C => G", "C => T", "G => A", "G => C", 
@@ -75,9 +75,8 @@ nucleotide_data[["1000"]] %>%
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "", y = expression(Log["10"]~Substitution~Rate)) + 
   coord_cartesian(ylim = c(0, -6)) + 
-  ggtitle("A") +  
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.7, y = 0.102, size = 2.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
+  ggtitle("Sub-Sampled to 1000 Sequences") +  
+  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 10), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10), 
@@ -88,10 +87,61 @@ nucleotide_data[["1000"]] %>%
   
   
   
- 
+five_thousand <- nucleotide_data[["5000"]] %>% 
+  mutate(sub_type = factor(sub_type, 
+                           levels = c("ac", "ag", "at", "ca", "cg", "ct", "ga", "gc", "gt", "ta", "tc", "tg"), 
+                           labels = c("A => C", "A => G", "A => T", "C => A", "C => G", "C => T", "G => A", "G => C", 
+                                      "G => T", "T => A", "T => C", "T => G"))) %>% 
+  ggplot(aes(sub_type, log10(rate + 0.000001), color = taq, group = taq)) + 
+  geom_point(size = 2, alpha = 0.2, position = position_dodge(width = 0.7), show.legend = F, color = "gray") + 
+  theme_bw() + 
+  stat_summary(aes(color = taq), fun.y = median, fun.ymin = median, fun.ymax = median, 
[…]

### 2017-12-05 code/make_fecal_numOTU_graphs.R
update figure titles

@@ -174,4 +174,4 @@ twenty_thousand_graph <- numOTU_data[["20000"]] %>%
 combined_graph <- grid.arrange(thousand_graph, five_thousand_graph, ten_thousand_graph, fifteen_thousand_graph, 
                                layout_matrix = rbind(c(1, 2), c(3, 4)))
 
-ggsave("results/figures/fecal_numOTU_graph.pdf", combined_graph, width = 11, height = 8, dpi = 300)
+ggsave("results/figures/Figure1.pdf", combined_graph, width = 11, height = 8, dpi = 300)

### 2017-12-05 code/make_mock_chimera_count_graphs.R

@@ -137,5 +137,5 @@ ten_thousand <- error_data[["10000"]] %>%
 
 combined_graph <- grid.arrange(thousand, five_thousand, ten_thousand, ncol = 3)
 
-ggsave("results/figures/mock_chimera_graph.pdf", combined_graph, width = 11, height = 7, dpi = 300)
+ggsave("results/figures/Figure4.pdf", combined_graph, width = 11, height = 7, dpi = 300)
 

### 2017-12-05 code/make_mock_error_graphs.R

@@ -133,4 +133,4 @@ ten_thousand <-  error_data[["10000"]] %>%
 
 combined_graph <- grid.arrange(thousand, five_thousand, ten_thousand, ncol = 3)
 
-ggsave("results/figures/mock_error_graph.pdf", combined_graph, width = 11, height = 7, dpi = 300)
+ggsave("results/figures/Figure3.pdf", combined_graph, width = 11, height = 7, dpi = 300)

### 2017-12-05 code/make_mock_seq_error_count_graphs.R

@@ -136,4 +136,4 @@ ten_thousand <- error_data[["10000"]] %>%
 
 combined_graph <- grid.arrange(thousand, five_thousand, ten_thousand, ncol = 3)
 
-ggsave("results/figures/mock_seq_count_error_graph.pdf", combined_graph, width = 11, height = 7, dpi = 300)
+ggsave("results/figures/FigureS1.pdf", combined_graph, width = 11, height = 7, dpi = 300)

### 2017-12-05 code/make_nucleotide_graphs.R

@@ -139,9 +139,9 @@ ten_thousand <- nucleotide_data[["10000"]] %>%
         legend.background = element_rect(color = "black"))
 
 
-ggsave("results/figures/mock_1000_sub_rate_graph.pdf", thousand, width = 8, height = 7, dpi = 300)
-ggsave("results/figures/mock_5000_sub_rate_graph.pdf", five_thousand, width = 8, height = 7, dpi = 300)
-ggsave("results/figures/mock_10000_sub_rate_graph.pdf", ten_thousand, width = 8, height = 7, dpi = 300)
+ggsave("results/figures/FigureS2.pdf", thousand, width = 8, height = 7, dpi = 300)
+ggsave("results/figures/FigureS3.pdf", five_thousand, width = 8, height = 7, dpi = 300)
+ggsave("results/figures/FigureS4.pdf", ten_thousand, width = 8, height = 7, dpi = 300)
 
 
 

### 2017-12-05 code/run_error_analysis.R
fix incorrect saving file name

@@ -197,7 +197,7 @@ write_csv(combined_kruskal_table_seq_error_count, "data/process/tables/mock_erro
 write_csv(combined_dunn_seq_error_count, "data/process/tables/mock_error_count_overall_dunn_results.csv")
 
 write_csv(combined_kruskal_table_chimera, "data/process/tables/mock_chimera_overall_kruskal_results.csv")
-write_csv(combined_dunn_chimera, "data/process/tables/mock_chimera_overall_tukey_results.csv")
+write_csv(combined_dunn_chimera, "data/process/tables/mock_chimera_overall_dunn_results.csv")
 
 
 

### 2017-12-05 submission/manuscript.Rmd
update title
add comparison tables to be used

@@
 
+# Read in fecal sample number of OTU results
+fs_anova_summary <- read_csv("../data/process/tables/fecal_overall_anova_results.csv")
+fs_tukey_summary <- read_csv("../data/process/tables/fecal_overall_tukey_results.csv")
+
+# Read in error rate results
+krsukal_error_rate <- read_csv("../data/process/tables/mock_error_overall_kruskal_results.csv")
+dunn_error_rate <- read_csv("../data/process/tables/mock_error_overall_dunn_results.csv")
+
 ```
 
-# Insert Title Here
+# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Sequencing
 

### 2017-12-05 submission/manuscript.Rmd
complete addition of tables to be used in mansucript

@@
 
+# Read in mock sample number of OTU results
+mock_anova_summary <- read_csv("../data/process/tables/mock_overall_anova_results.csv")
+mock_tukey_summary <- read_csv("../data/process/tables/mock_overall_tukey_results.csv")
+
 # Read in error rate results
@@
 dunn_error_rate <- read_csv("../data/process/tables/mock_error_overall_dunn_results.csv")
+
+# Read in sequence error count results
+kruskal_seq_error_count <- read_csv("../data/process/tables/mock_error_count_overall_kruskal_results.csv")
+dunn_seq_error_count <- read_csv("../data/process/tables/mock_error_count_overall_dunn_results.csv")
+
+# Read in chimera prevalence results
+kruskal_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv")
+dunn_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")
 

### 2017-12-05 submission/manuscript.Rmd
add part of methods

@@
 
-***Insert Sub section here***
+***Human and Mock Samples:***
+A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN) and the processing and storage of these samples have been published previously [@erin_seekatz_2016].
+Clinical data and other types of meta data were not utilized or accessed for this study.
+All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance  (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+
+***PCR Protocol:***
+The five different high fidelity Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA).
 

### 2017-12-05 submission/manuscript.tex
update title
add part of methods

@@
 
-\section{Insert Title Here}\label{insert-title-here}
+\section{What the Taq? The Influence of Different Hi-Fidelity Taq
+Polymerase on 16S rRNA
+Sequencing}\label{what-the-taq-the-influence-of-different-hi-fidelity-taq-polymerase-on-16s-rrna-sequencing}
 
@@
 
-\textbf{\emph{Insert Sub section here}}
+\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
+obtained from 4 individuals who were part of the Enterics Research
+Investigational Network (ERIN) and the processing and storage of these
+samples have been published previously (Seekatz et al., 2016). Clinical
+data and other types of meta data were not utilized or accessed for this
+study. All samples were extracted using the MOBIO\textsuperscript{TM}
+PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The
+ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
+CA, USA) was used in this study and is made up of \emph{Pseudomonas
+aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
+\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
+\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
+\emph{Bacillus subtilis} at equal genomic DNA abundance
+(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
+
+\textbf{\emph{PCR Protocol:}} The five different high fidelity Taq DNA
+polymerase that were tested were AccuPrime\textsuperscript{TM}
+(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA).
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-erin_seekatz_2016}{}
+Seekatz AM., Rao K., Santhosh K., Young VB. 2016. Dynamics of the fecal
+microbiome in patients with recurrent and nonrecurrent clostridium
+difficile infection. \emph{Genome Medicine} 8. DOI:
+\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
 

### 2017-12-06 submission/manuscript.Rmd
add PCR protocol to methods

@@
 ```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+#Read in the meta data file
+metadata <- read_csv("../data/process/tables/meta_data.csv")
 
@@
 ***PCR Protocol:***
-The five different high fidelity Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA).
+The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The PCR cycle conditions were the same for every primer [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
+If the HiFi Taq had a specific activation time that was different then 2 minutes that was used instead.
+The 30 cycle default was used but the cycle conditions started at 15 and increased by 5 up to 35 cycles and was used for both fecal and mock samples.
+The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100).
+Although, the mock communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10 replicates were used for 30 cycles for all Taq (total samples = 130).
+For all the mock community samples there was not enough PCR product at 15 cycles for adequate sequencing.
 

### 2017-12-06 submission/manuscript.tex
add PCR protocol to methods

@@
 
-\textbf{\emph{PCR Protocol:}} The five different high fidelity Taq DNA
-polymerase that were tested were AccuPrime\textsuperscript{TM}
-(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA).
+\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
+Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
+(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
+(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
+England Biolabs, MA, USA). The PCR cycle conditions were the same for
+every primer (Kozich et al., 2013)
+(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
+If the HiFi Taq had a specific activation time that was different then 2
+minutes that was used instead. The 30 cycle default was used but the
+cycle conditions started at 15 and increased by 5 up to 35 cycles and
+was used for both fecal and mock samples. The fecal PCR consisted of all
+4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples =
+100). Although, the mock communities also had 4 replicates used for 15,
+20, 25, and 35 cycles, 10 replicates were used for 30 cycles for all Taq
+(total samples = 130). For all the mock community samples there was not
+enough PCR product at 15 cycles for adequate sequencing.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-protocol_Kozich_2013}{}
+Kozich JJ., Westcott SL., Baxter NT., Highlander SK., Schloss PD. 2013.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. \emph{Applied and Environmental Microbiology} 79:5112--5120.
+DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
 \hypertarget{ref-erin_seekatz_2016}{}

### 2017-12-06 submission/manuscript.Rmd
add sequence processing methods section

@@
 
+***Sequence Processing:***
+The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009].
+The protocol followed was similar to that used by Kozich, *et al.* [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for OTU generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+
 ***Reproducible Methods:***

### 2017-12-06 submission/manuscript.tex
add sequence processing methods section

@@
 
+\textbf{\emph{Sequence Processing:}} The mothur software program was
+utilized for all sequence processing steps (Schloss et al., 2009). The
+protocol followed was similar to that used by Kozich, \emph{et al.}
+(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
+major differences from the stated protocol were the use VSEARCH instead
+of UCHIME for chimera detection and the use of the OptiClust algorithmn
+instead of average neighbor for OTU generation (Edgar et al., 2011;
+Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error was
+determined using the seq.error command on mock samples after chimera
+removal and classification to the RDP to remove non-bacterial sequences
+(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
@@
 \hypertarget{refs}{}
+\hypertarget{ref-rdp_Cole_2013}{}
+Cole JR., Wang Q., Fish JA., Chai B., McGarrell DM., Sun Y., Brown CT.,
+Porras-Alfaro A., Kuske CR., Tiedje JM. 2013. Ribosomal database
+project: Data and tools for high throughput rRNA analysis. \emph{Nucleic
+Acids Research} 42:D633--D642. DOI:
+\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+
+\hypertarget{ref-uchime_Edgar_2011}{}
+Edgar RC., Haas BJ., Clemente JC., Quince C., Knight R. 2011. UCHIME
+improves sensitivity and speed of chimera detection.
+\emph{Bioinformatics} 27:2194--2200. DOI:
+\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+
 \hypertarget{ref-protocol_Kozich_2013}{}
@@
 
+\hypertarget{ref-vsearch_Rognes_2016}{}
+Rognes T., Flouri T., Nichols B., Quince C., Mahé F. 2016. VSEARCH: A
+versatile open source tool for metagenomics. \emph{PeerJ} 4:e2584. DOI:
+\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-mothur_schloss_2009}{}
+Schloss PD., Westcott SL., Ryabin T., Hall JR., Hartmann M., Hollister
+EB., Lesniewski RA., Oakley BB., Parks DH., Robinson CJ., Sahl JW.,
+Stres B., Thallinger GG., Horn DJV., Weber CF. 2009. Introducing mothur:
+Open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. \emph{Applied and
+Environmental Microbiology} 75:7537--7541. DOI:
+\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
 \hypertarget{ref-erin_seekatz_2016}{}
@@
 \href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+
+\hypertarget{ref-opticlust_Westcott_2017}{}
+Westcott SL., Schloss PD. 2017. OptiClust, an improved method for
+assigning amplicon-based sequence data to operational taxonomic units.
+\emph{mSphere} 2:e00073--17. DOI:
+\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
 

### 2017-12-06 submission/manuscript.Rmd
add statistical analysis section

@@
 The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009].
-The protocol followed was similar to that used by Kozich, *et al.* [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
 Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for OTU generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
 Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+
+***Statistical Analysis:***
+All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
+Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
+The total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
+For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between different people.
+For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
+Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
 

### 2017-12-06 submission/manuscript.tex
add statistical analysis section

@@
 utilized for all sequence processing steps (Schloss et al., 2009). The
-protocol followed was similar to that used by Kozich, \emph{et al.}
+protocol followed was similar to what has been previously publsihed
 (Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
@@
 (Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+
+\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
+(v 3.4.2) software package (R Core Team, 2017). Data tranformation and
+graphing was completed using the tidyverse package (v 1.1.1) and colors
+selected using the viridis package (v 0.4.0) (Garnier, 2017; Wickham,
+2017). The total number of OTUs were analyzed using an ANOVA with a
+tukey post-hoc test. For the fecal samples the data was normalized to
+each individual by cycle number to account for the biological variation
+between different people. For both error and chimera analysis, samples
+were tested using Kruskal-Wallis with a Dunns post-hoc test. Where
+applicable correction for multiple comparison utilized the
+Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-benjamini_controlling_1995}{}
+Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
+practical and powerful approach to multiple testing. \emph{Journal of
+the Royal Statistical Society. Series B (Methodological)} 57:289--300.
+
 \hypertarget{ref-rdp_Cole_2013}{}
@@
 
+\hypertarget{ref-viridis_citation_2017}{}
+Garnier S. 2017. \emph{Viridis: Default color maps from 'matplotlib'}.
+
 \hypertarget{ref-protocol_Kozich_2013}{}
@@
 DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-r_citation_2017}{}
+R Core Team. 2017. \emph{R: A language and environment for statistical
+computing}. Vienna, Austria: R Foundation for Statistical Computing.
 
@@
 \href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
+\hypertarget{ref-tidyverse_2017}{}
+Wickham H. 2017. \emph{Tidyverse: Easily install and load 'tidyverse'
+packages}.
 

### 2017-12-06 submission/manuscript.Rmd
add general analysis workflow to methods

@@
 
+***Analysis Workflow:***
+The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
+From these observations we wanted to next analyze potential reasons as to why some of these differences may have occured.
+First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each Taq.
+After assessing these errors, the total number of chimeras was determined after sequence processing.
+The fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.
+
 ***Reproducible Methods:***

### 2017-12-06 submission/manuscript.tex
add general analysis workflow to methods

@@
 
+\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
+sub-sampling was analyzed for both the fecal and mock community samples.
+From these observations we wanted to next analyze potential reasons as
+to why some of these differences may have occured. First, analysis of
+general sequence error rate, number of sequences with an error, and base
+subsitution were assessed in the mock community for each Taq. After
+assessing these errors, the total number of chimeras was determined
+after sequence processing. The fecal samples were analyzed at 4
+different sub-sampling levels, 1000, 5000, 10000, and 15000 while the
+mock community samples were analysed at 3 levels, 1000, 5000, 10000.
+
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found

### 2017-12-06 submission/manuscript.Rmd
add reproducible research section and update acknowledgements

@@
 The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017.
+The raw sequences can be found in the SRA at the following accesssion number **need to upload still**.
 
@@
 
-The authors would like to thank all the study participants who were a part of each of the individual studies utilized.
-We would also like to thank each of the study authors for making their data available for use.
-Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.
+The authors would like to thank all the study participants ERIN whose samples were utilized.
+We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
+Salary support for Marc Sze came from the Canadian Institue of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.
 

### 2017-12-06 submission/manuscript.tex
add reproducible research section and update acknowledgements

@@
 here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
+The raw sequences can be found in the SRA at the following accesssion
+number \textbf{need to upload still}.
 
@@
 
-The authors would like to thank all the study participants who were a
-part of each of the individual studies utilized. We would also like to
-thank each of the study authors for making their data available for use.
-Finally we would like to thank the members of the Schloss lab for
-valuable feed back and proof reading during the formulation of this
-manuscript.
+The authors would like to thank all the study participants ERIN whose
+samples were utilized. We would also like to thank Judy Opp and April
+Cockburn for their effort in sequencing the samples as part of the
+Microbiome Core Facility at the University of Michigan. Salary support
+for Marc Sze came from the Canadian Institue of Health Research and the
+Michigan Institute for Clinical and Health Research Postdoctoral
+Translational Scholar Program.
 

### 2017-12-06 submission/manuscript.Rmd
update results section headers
add acronym definition

@@
 The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for OTU generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
 Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
@@
 
-***Insert Sub title here***
+***The Number of OTUs is Dependent on HiFi Taq Used:***
+
+***Per Base Error Rate is Dependent on both Taq and Cycle Number Used:***
+
+***Chimeric Sequences are HiFi Taq Dependent:***
 

### 2017-12-06 submission/manuscript.Rmd
add result template first section

@@
 ***The Number of OTUs is Dependent on HiFi Taq Used:***
+After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi Taq used and this difference increased as the depth of sub-sampling increased [Figure 1].
+Lower cycle numbers (15-25) resulted in less differences between Taq while cycle numbers of 30 and 35 had larger clearer defined differences [Figure 1].
+Generally, the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum [Figure 1].
+This Taq dependent difference in the number of OTUs was also observed in the mock community samples with the same Taq polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2].
 

### 2017-12-08 submission/manuscript.Rmd
add results template write up

@@
 ***Per Base Error Rate is Dependent on both Taq and Cycle Number Used:***
+The median average per base error was highest for the Kappa HiFI Taq [FIgure 3A].
+Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B & C].
+Generally, there were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number [Figure 3B & C].
+Platinum HiFi Taq consistently had the highes median average per base error rate while Phusion and Accuprime had the lowest [Figure 3B & C].
+These differences though were realtively small between the different HiFi Taq [Figure 3].
+The total seqeunces with at least one error matched the median average per base error except for the Accuprime HiFi Taq [Figure S1].
+Although it had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number of sub-sampling depth [Figure S1].
+Investigation of whether there were Taq dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub sampling depth [Figure S2-S4].
+Further, the variation in subsitution error seems to reduce as the sequencing depth increases [Figure S2-S4].
 
 ***Chimeric Sequences are HiFi Taq Dependent:***
+After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
+From this we observed that Platinum HiFi conssitently had the highest percentage of chimeric reads regardless of amplification cycle number and sub-sampling depth [Figure 4].
+These results mirror the observations of increased number of OTUs and the increased median average per base error [Figure 2-4].
 

### 2017-12-08 code/make_correlation_graphs.R
first commit

@@ -1,0 +1,50 @@
+### Correlations between error and chimera with number of OTUs
+### Mock Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "dunn.test"))
+
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = ""))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+
+
+
+
+
+
+
+
+
+
+
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+
+
+
+

### 2017-12-08 code/make_correlation_graphs.R
things o do

@@ -5,7 +5,7 @@
 # Load in needed functions and libraries
 source('code/functions.R')
 
-loadLibs(c("tidyverse", "stringr", "dunn.test"))
+loadLibs(c("tidyverse", "stringr", "viridis"))
 
 
 ###########################################################################################################################
@@ -29,11 +29,24 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 }
 
 
+# FUnction to create a new common column
+
+# Function to join data together
 
 
+# Read in subsample.shared files
+sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+
+# Read in the count data
+error_data <- sapply(sub_sample_level, 
+                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
+                     simplify = F)
 
 
-
+# Read in the count data
+numOTU_data <- sapply(sub_sample_level, 
+                      function(x) read_data("data/process/tables/", "mock_sub_sample_", "_count_table.csv", x), 
+                      simplify = F)
 
 
 

### 2017-12-08 code/make_correlation_graphs.R
add combination of data for correlations

@@ -30,8 +30,24 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 
 
 # FUnction to create a new common column
+make_common_column <- function(i, dataList){
+  
+  tempData <- dataList[[i]] %>% 
+    mutate(common_column = paste(cycles, "_", taq, "_", sample_name, sep = ""))
+  
+  return(tempData)
+  
+}
 
 # Function to join data together
+combine_data <- function(i, dataList, to_match_list){
+  
+  tempData <- dataList[[i]] %>% 
+    left_join(to_match_list[[i]], by = "common_column")
+  
+  return(tempData)
+    
+}
 
 
 # Read in subsample.shared files
@@ -49,8 +65,16 @@ numOTU_data <- sapply(sub_sample_level,
                       simplify = F)
 
 
+# Create new common column
+up_error_data <- sapply(sub_sample_level, 
+                        function(x) make_common_column(x, error_data), simplify = F)
 
+up_numOTU_data <- sapply(sub_sample_level, 
+                         function(x) make_common_column(x, numOTU_data), simplify = F)
 
+# Combine the data together
+combined_list <- sapply(sub_sample_level, 
+                        function(x) combine_data(x, up_error_data, up_numOTU_data), simplify = F)
 
 
 ###########################################################################################################################

### 2017-12-08 code/make_correlation_graphs.R
rough graphing

@@ -77,6 +77,29 @@ combined_list <- sapply(sub_sample_level,
                         function(x) combine_data(x, up_error_data, up_numOTU_data), simplify = F)
 
 
+combined_list[["1000"]] %>% 
+  mutate(taq = factor(taq.x, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
+  ggplot(aes(cycles, log2(numOTUs), color = taq, group = taq)) + 
+  geom_smooth(size = 1, method = "lm", formula = y ~ poly(x, 2), se = FALSE) + 
+  geom_point(size = 2, alpha = 0.7) + theme_bw() + 
+  scale_color_manual(name = "Taq Used", 
+                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
+  labs(x = "Amplification Cycles", y = expression(Log["2"]~Number~of~OTUs)) + 
+  ggtitle("A") + coord_cartesian(ylim = c(0, 8)) + 
+  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.5, y = 8.2, size = 2.5) + 
+  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        axis.text.y = element_text(size = 10), 
+        legend.position = c(0.20, 0.18), 
+        legend.title = element_blank(), 
+        legend.key = element_blank(), 
+        legend.background = element_rect(color = "black"))
+
+
+
 ###########################################################################################################################
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################

### 2017-12-08 code/make_correlation_graphs.R
add R2 formula code --rough

@@ -50,6 +50,7 @@ combine_data <- function(i, dataList, to_match_list){
 }
 
 
+
 # Read in subsample.shared files
 sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
 
@@ -77,16 +78,32 @@ combined_list <- sapply(sub_sample_level,
                         function(x) combine_data(x, up_error_data, up_numOTU_data), simplify = F)
 
 
+lm_eqn = function(df){
+  m = lm((chimera_prevalence*1000) ~ numOTUs, df);
+  eq <- substitute(~~R^2~"="~r2, 
+                   list(r2 = format(summary(m)$r.squared, digits = 3)))
+  as.character(as.expression(eq));                 
+}
+
+
+
+eqns <- by(combined_list[["1000"]], combined_list[["1000"]]$taq.x, lm_eqn)
+df2 <- data.frame(eq = unclass(eqns), taq = as.numeric(names(eqns)))
+
 combined_list[["1000"]] %>% 
   mutate(taq = factor(taq.x, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
-  ggplot(aes(cycles, log2(numOTUs), color = taq, group = taq)) + 
-  geom_smooth(size = 1, method = "lm", formula = y ~ poly(x, 2), se = FALSE) + 
+  ggplot(aes(chimera_prevalence*100, numOTUs, color = cycles.x, group = taq)) + 
+  geom_smooth(size = 1, method = "lm", se = FALSE, color = "black") + 
   geom_point(size = 2, alpha = 0.7) + theme_bw() + 
-  scale_color_manual(name = "Taq Used", 
-                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = expression(Log["2"]~Number~of~OTUs)) + 
+  facet_grid(taq ~.) + 
+  scale_color_manual(name = "Cycle Number", 
+                     values = c("#0000FF", "#00C957", "#CD8500", "#FF1493")) + 
+  labs(x = "Percent Chimera Prevalence", y = "Number of OTUs") + 
+  geom_text(x = 25, y = 300, label = lm_eqn(), parse = TRUE)
+  
+  
[…]

### 2017-12-08 code/make_correlation_graphs.R
update workflow for correlation graph

@@ -49,6 +49,14 @@ combine_data <- function(i, dataList, to_match_list){
     
 }
 
+# Function to make label for R2 (found on stack overflow)
+lm_eqn = function(df){
+  m = lm((chimera_prevalence*1000) ~ numOTUs, df);
+  eq <- substitute(~~R^2~"="~r2, 
+                   list(r2 = format(summary(m)$r.squared, digits = 3)))
+  as.character(as.expression(eq));                 
+}
+
 
 
 # Read in subsample.shared files
@@ -78,17 +86,13 @@ combined_list <- sapply(sub_sample_level,
                         function(x) combine_data(x, up_error_data, up_numOTU_data), simplify = F)
 
 
-lm_eqn = function(df){
-  m = lm((chimera_prevalence*1000) ~ numOTUs, df);
-  eq <- substitute(~~R^2~"="~r2, 
-                   list(r2 = format(summary(m)$r.squared, digits = 3)))
-  as.character(as.expression(eq));                 
-}
 
 
-
-eqns <- by(combined_list[["1000"]], combined_list[["1000"]]$taq.x, lm_eqn)
-df2 <- data.frame(eq = unclass(eqns), taq = as.numeric(names(eqns)))
+eqns1000 <- by(combined_list[["1000"]], combined_list[["1000"]]$taq.x, lm_eqn)
+df1000 <- data.frame(eq = unclass(eqns1000), taq = rownames(eqns1000)) %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
 
 combined_list[["1000"]] %>% 
   mutate(taq = factor(taq.x, 
@@ -101,21 +105,50 @@ combined_list[["1000"]] %>%
   scale_color_manual(name = "Cycle Number", 
                      values = c("#0000FF", "#00C957", "#CD8500", "#FF1493")) + 
   labs(x = "Percent Chimera Prevalence", y = "Number of OTUs") + 
-  geom_text(x = 25, y = 300, label = lm_eqn(), parse = TRUE)
-  
-  
-  ggtitle("A") + coord_cartesian(ylim = c(0, 8)) + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.5, y = 8.2, size = 2.5) + 
[…]

### 2017-12-08 code/make_correlation_graphs.R
incoporate correlation between OTUs and chimera

@@ -5,7 +5,7 @@
 # Load in needed functions and libraries
 source('code/functions.R')
 
-loadLibs(c("tidyverse", "stringr", "viridis"))
+loadLibs(c("tidyverse", "stringr", "viridis", "gridExtra"))
 
 
 ###########################################################################################################################
@@ -94,7 +94,7 @@ df1000 <- data.frame(eq = unclass(eqns1000), taq = rownames(eqns1000)) %>%
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
 
-combined_list[["1000"]] %>% 
+thousand <- combined_list[["1000"]] %>% 
   mutate(taq = factor(taq.x, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
@@ -105,10 +105,10 @@ combined_list[["1000"]] %>%
   scale_color_manual(name = "Cycle Number", 
                      values = c("#0000FF", "#00C957", "#CD8500", "#FF1493")) + 
   labs(x = "Percent Chimera Prevalence", y = "Number of OTUs") + 
-  geom_text(data = df1000, aes(x = 0.5, y = 65, label = eq, family = "arial"), 
+  geom_text(data = df1000, aes(x = 1.5, y = 75, label = eq), 
             color = 'black',  parse = TRUE) + 
   ggtitle("A") + coord_cartesian(ylim = c(0, 80)) + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 24, y = 8.2, size = 2.5) + 
+  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 22, y = 8.2, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -126,7 +126,7 @@ df5000 <- data.frame(eq = unclass(eqns5000), taq = rownames(eqns5000)) %>%
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
 
-combined_list[["5000"]] %>% 
+five_thousand <- combined_list[["5000"]] %>% 
[…]

### 2017-12-08 submission/manuscript.Rmd
add completed rough Results workflow -- need stats still

@@
 
-***Chimeric Sequences are HiFi Taq Dependent:***
+***Chimeric Sequences Corelate with OTUs and are HiFi Taq Dependent:***
 After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
 From this we observed that Platinum HiFi conssitently had the highest percentage of chimeric reads regardless of amplification cycle number and sub-sampling depth [Figure 4].
-These results mirror the observations of increased number of OTUs and the increased median average per base error [Figure 2-4].
+For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs and this correlation was strongest for Platinum and Phusion HiFi Taq [Figure 5].
+In general the correlations between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
 

### 2017-12-08 submission/manuscript.tex
add completed rough Results workflow -- need stats still

@@
 of UCHIME for chimera detection and the use of the OptiClust algorithmn
-instead of average neighbor for OTU generation (Edgar et al., 2011;
-Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error was
-determined using the seq.error command on mock samples after chimera
-removal and classification to the RDP to remove non-bacterial sequences
-(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+instead of average neighbor for Operational Taxonomic Unit (OTU)
+generation (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
+Schloss, 2017). Sequence error was determined using the seq.error
+command on mock samples after chimera removal and classification to the
+RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
+al., 2013; Rognes et al., 2016).
 
@@
 
-\textbf{\emph{Insert Sub title here}}
+\textbf{\emph{The Number of OTUs is Dependent on HiFi Taq Used:}} After
+normalization by individual, for each cycle number, we observed that for
+fecal samples the number of OTUs identified was dependent upon the HiFi
+Taq used and this difference increased as the depth of sub-sampling
+increased {[}Figure 1{]}. Lower cycle numbers (15-25) resulted in less
+differences between Taq while cycle numbers of 30 and 35 had larger
+clearer defined differences {[}Figure 1{]}. Generally, the lowest number
+of OTUs identified was from Accuprime\textsuperscript{TM} while the
+highest was from Platinum {[}Figure 1{]}. This Taq dependent difference
+in the number of OTUs was also observed in the mock community samples
+with the same Taq polymerases being high (Platinum) and low (Accuprime)
+respectively {[}Figure 2{]}.
+
+\textbf{\emph{Per Base Error Rate is Dependent on both Taq and Cycle
+Number Used:}} The median average per base error was highest for the
+Kappa HiFI Taq {[}FIgure 3A{]}. Sub-sampling depth seems to have little
+effect on this rate with both 5000 and 10000 sub-sampled sequences
+showing similar results {[}Figure 3B \& C{]}. Generally, there were
+small differences between the various HiFi Taq at lower cycle number but
+larger differences at higher cycle number {[}Figure 3B \& C{]}. Platinum
+HiFi Taq consistently had the highes median average per base error rate
+while Phusion and Accuprime had the lowest {[}Figure 3B \& C{]}. These
+differences though were realtively small between the different HiFi Taq
+{[}Figure 3{]}. The total seqeunces with at least one error matched the
+median average per base error except for the Accuprime HiFi Taq
+{[}Figure S1{]}. Although it had the lowest per base error rate it had
+either the largest or second largest number of sequences with at least
+one error regardless of cycle number of sub-sampling depth {[}Figure
+S1{]}. Investigation of whether there were Taq dependent effects on base
+subsitution found that there was no clear bias and this was indpendent
+of sub sampling depth {[}Figure S2-S4{]}. Further, the variation in
+subsitution error seems to reduce as the sequencing depth increases
+{[}Figure S2-S4{]}.
+
+\textbf{\emph{Chimeric Sequences Corelate with OTUs and are HiFi Taq
+Dependent:}} After chimera removal using VSEARCH and removal of
+sequences that did not classify as bacteria we assessed the percentage
+of sequences that were still chimeric within our mock community. From
+this we observed that Platinum HiFi conssitently had the highest
+percentage of chimeric reads regardless of amplification cycle number
+and sub-sampling depth {[}Figure 4{]}. For all Taqs a positive
+correlation was observed between chimeric sequences and number of OTUs
+and this correlation was strongest for Platinum and Phusion HiFi Taq
+{[}Figure 5{]}. In general the correlations between the number of OTUs
+and chimeric sequences became stronger as sub-sampling depth increased
+{[}Figure 5{]}.
 

### 2017-12-11 submission/supplemental.Rmd
first commit

---
title: ''
csl: peerj.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: null
---

# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Sequencing

\vspace{10mm}

\begin{center}
Marc A Sze and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}

\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))

```

**Table S1: ANOVA Results of HiFi Taq Differences in Fecal Samples**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

fs_anova <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv")) %>% 
  mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))

colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycles", "Sub-Sample Depth")

kable(fs_anova, format = "markdown", align = 'c', padding = 2)
```

\newpage

**Table S2:**

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S3: **

```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S4: **

```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S5: **

```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S6: **

```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S7: **

```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}

```

### 2017-12-11 submission/manuscript.Rmd
add pvalues to fecal section

@@
 # Read in fecal sample number of OTU results
-fs_anova_summary <- read_csv("../data/process/tables/fecal_overall_anova_results.csv")
+fs_anova_summary <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv"))
 fs_tukey_summary <- read_csv("../data/process/tables/fecal_overall_tukey_results.csv")
@@
 After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi Taq used and this difference increased as the depth of sub-sampling increased [Figure 1].
-Lower cycle numbers (15-25) resulted in less differences between Taq while cycle numbers of 30 and 35 had larger clearer defined differences [Figure 1].
+Lower cycle numbers (15-20) resulted in less differences between Taq while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
+Only 35 cycles had HiFi Taq differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi Taq differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05).
 Generally, the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum [Figure 1].

### 2017-12-11 submission/manuscript.tex
add pvalues to fecal section

@@
 Taq used and this difference increased as the depth of sub-sampling
-increased {[}Figure 1{]}. Lower cycle numbers (15-25) resulted in less
-differences between Taq while cycle numbers of 30 and 35 had larger
-clearer defined differences {[}Figure 1{]}. Generally, the lowest number
-of OTUs identified was from Accuprime\textsuperscript{TM} while the
-highest was from Platinum {[}Figure 1{]}. This Taq dependent difference
-in the number of OTUs was also observed in the mock community samples
-with the same Taq polymerases being high (Platinum) and low (Accuprime)
-respectively {[}Figure 2{]}.
+increased {[}Figure 1{]}. Lower cycle numbers (15-20) resulted in less
+differences between Taq while cycle numbers of 25, 30, and 35 had larger
+clearer defined differences {[}Figure 1{]}. Only 35 cycles had HiFi Taq
+differences that were significantly different at all sub-sampling levels
+(P-value \textless{} 0.0001). At sub-sampling depth of 5000 or higher 25
+and 30 cycles had HiFi Taq differences (P-value \textless{} 0.05).
+Generally, the lowest number of OTUs identified was from
+Accuprime\textsuperscript{TM} while the highest was from Platinum
+{[}Figure 1{]}. This Taq dependent difference in the number of OTUs was
+also observed in the mock community samples with the same Taq
+polymerases being high (Platinum) and low (Accuprime) respectively
+{[}Figure 2{]}.
 

### 2017-12-11 submission/manuscript.Rmd
add supplemental tables and more pvslues to first section results

--- 2017-12-08 submission/manuscript.Rmd
@@
 # Read in fecal sample number of OTU results
-fs_anova_summary <- read_csv("../data/process/tables/fecal_overall_anova_results.csv")
-fs_tukey_summary <- read_csv("../data/process/tables/fecal_overall_tukey_results.csv")
+fs_anova_summary <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv"))
+fs_tukey_summary <- as.data.frame(read_csv("../data/process/tables/fecal_overall_tukey_results.csv")) %>% 
+  rename(padj = `p adj`)
 
@@
 After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi Taq used and this difference increased as the depth of sub-sampling increased [Figure 1].
-Lower cycle numbers (15-25) resulted in less differences between Taq while cycle numbers of 30 and 35 had larger clearer defined differences [Figure 1].
-Generally, the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum [Figure 1].
+Lower cycle numbers (15-20) resulted in less differences between Taq while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
+Only 35 cycles had HiFi Taq differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi Taq differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
+Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other Taq across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Across sub-sampling depth and cycle number the lowest number of OTUs identified was generally from Accuprime^TM^ while the highest was from Platinum [Figure 1].
 This Taq dependent difference in the number of OTUs was also observed in the mock community samples with the same Taq polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2].

### 2017-12-11 submission/manuscript.tex
add supplemental tables and more pvslues to first section results

@@
 differences that were significantly different at all sub-sampling levels
-(P-value \textless{} 0.0001). At sub-sampling depth of 5000 or higher 25
-and 30 cycles had HiFi Taq differences (P-value \textless{} 0.05).
-Generally, the lowest number of OTUs identified was from
+(P-value \textless{} 0.0001) {[}Table S1{]}. At sub-sampling depth of
+5000 or higher 25 and 30 cycles had HiFi Taq differences (P-value
+\textless{} 0.05) {[}Table S1{]}. Using a Tukey post-hoc test only at 35
+cycles were significant differences found to be mainly driven by
+Platinum being different than all other Taq across sub-sampling depth
+(P-value \textless{} 0.05) {[}Table S2{]}. Across sub-sampling depth and
+cycle number the lowest number of OTUs identified was generally from
 Accuprime\textsuperscript{TM} while the highest was from Platinum

### 2017-12-11 submission/supplemental.Rmd
add table s2

@@
 
-colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycles", "Sub-Sample Depth")
+colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sub-Sample Depth")
 
@@
 
-**Table S2:**
+**Table S2: Tukey Post-Hoc Results of Significant HiFi Taq Differences in Fecal Samples**
 
 ```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+
+fs_tukey <- as.data.frame(read_csv("../data/process/tables/fecal_overall_tukey_results.csv")) %>% 
+  mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+  rename(padj = `p adj`) %>% 
+  mutate(padj = as.numeric(padj)) %>% 
+  filter(padj < 0.05) %>% 
+  mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+
+colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+
+kable(fs_tukey, format = "markdown", align = 'c', padding = 2)
 

### 2017-12-11 submission/supplemental.Rmd
add table s3

@@
 
-**Table S1: ANOVA Results of HiFi Taq Differences in Fecal Samples**
+**Table S1: ANOVA Results of Number of OTUs Differences between HiFi Taq in Fecal Samples**
 
@@
 
-**Table S2: Tukey Post-Hoc Results of Significant HiFi Taq Differences in Fecal Samples**
+**Table S2: Tukey Post-Hoc Results of Number of OTUs Differences between HiFi Taq in Fecal Samples**
 
@@
 
-**Table S3: **
+**Table S3: ANOVA Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
 
@@
 
+mock_anova <- as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv")) %>% 
+  mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  filter(sub_sample_level > 500)
+
+colnames(mock_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+
+kable(mock_anova, format = "markdown", align = 'c', padding = 2)
 ```

### 2017-12-11 submission/manuscript.Rmd
complete first results section -- better version

@@
 # Read in mock sample number of OTU results
-mock_anova_summary <- read_csv("../data/process/tables/mock_overall_anova_results.csv")
+mock_anova_summary <- as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv"))
 mock_tukey_summary <- read_csv("../data/process/tables/mock_overall_tukey_results.csv")
@@
 Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other Taq across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
-Across sub-sampling depth and cycle number the lowest number of OTUs identified was generally from Accuprime^TM^ while the highest was from Platinum [Figure 1].
-This Taq dependent difference in the number of OTUs was also observed in the mock community samples with the same Taq polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2].
+
+This Taq dependent difference in the number of OTUs was also observed in the mock community samples with the same Taq polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2 & Table S3].
+Conversely, differences between HiFi Taq were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+Across sub-sampling depth and cycle number the lowest number of OTUs identified was generally from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
 

### 2017-12-11 submission/manuscript.tex
complete first results section -- better version

@@
 Platinum being different than all other Taq across sub-sampling depth
-(P-value \textless{} 0.05) {[}Table S2{]}. Across sub-sampling depth and
-cycle number the lowest number of OTUs identified was generally from
-Accuprime\textsuperscript{TM} while the highest was from Platinum
-{[}Figure 1{]}. This Taq dependent difference in the number of OTUs was
-also observed in the mock community samples with the same Taq
-polymerases being high (Platinum) and low (Accuprime) respectively
-{[}Figure 2{]}.
+(P-value \textless{} 0.05) {[}Table S2{]}.
+
+This Taq dependent difference in the number of OTUs was also observed in
+the mock community samples with the same Taq polymerases being high
+(Platinum) and low (Accuprime) respectively {[}Figure 2 \& Table S3{]}.
+Conversely, differences between HiFi Taq were observed as early as 20
+cycles and a sub-sampling depth of 1000 sequences (P-value = 0.002)
+{[}Table S3{]}. Across sub-sampling depth and cycle number the lowest
+number of OTUs identified was generally from
+Accuprime\textsuperscript{TM} while the highest was from Platinum for
+both fecal and mock samples {[}Figure 1 \& 2{]}.
 

### 2017-12-11 submission/supplemental.Rmd
add table s4

@@
 
-**Table S4: **
+**Table S4: Tukey Post-Hoc Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
 
 ```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+
+mock_tukey <- as.data.frame(read_csv("../data/process/tables/mock_overall_tukey_results.csv")) %>% 
+  mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+  rename(padj = `p adj`) %>% 
+  mutate(padj = as.numeric(padj)) %>% 
+  filter(padj < 0.05, sub_sample_level > 500) %>% 
+  arrange(padj) %>% 
+  mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+
+colnames(mock_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+
+kable(mock_tukey, format = "markdown", align = 'c', padding = 2)
 

### 2017-12-11 submission/manuscript.Rmd
add table s4 to results

@@
 Conversely, differences between HiFi Taq were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-Across sub-sampling depth and cycle number the lowest number of OTUs identified was generally from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
+Using a Tukey post-hoc test differences between Platinum and the other HiFi Taqs was the major driver of the differences seen at different cycle numbers and sub-sampling depths [Table S4].
+Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
 
 ***Per Base Error Rate is Dependent on both Taq and Cycle Number Used:***
-The median average per base error was highest for the Kappa HiFI Taq [FIgure 3A].
+Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [].
+Generally, the highest values were for the Kappa HiFI Taq [Figure 3A].
 Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B & C].

### 2017-12-11 submission/manuscript.tex
add table s4 to results

@@
 cycles and a sub-sampling depth of 1000 sequences (P-value = 0.002)
-{[}Table S3{]}. Across sub-sampling depth and cycle number the lowest
-number of OTUs identified was generally from
+{[}Table S3{]}. Using a Tukey post-hoc test differences between Platinum
+and the other HiFi Taqs was the major driver of the differences seen at
+different cycle numbers and sub-sampling depths {[}Table S4{]}. Both
+fecal and mock samples consistently showed that across sub-sampling
+depth and cycle number the lowest number of OTUs identified was from
 Accuprime\textsuperscript{TM} while the highest was from Platinum for
@@
 \textbf{\emph{Per Base Error Rate is Dependent on both Taq and Cycle
-Number Used:}} The median average per base error was highest for the
-Kappa HiFI Taq {[}FIgure 3A{]}. Sub-sampling depth seems to have little
-effect on this rate with both 5000 and 10000 sub-sampled sequences
-showing similar results {[}Figure 3B \& C{]}. Generally, there were
-small differences between the various HiFi Taq at lower cycle number but
-larger differences at higher cycle number {[}Figure 3B \& C{]}. Platinum
-HiFi Taq consistently had the highes median average per base error rate
-while Phusion and Accuprime had the lowest {[}Figure 3B \& C{]}. These
-differences though were realtively small between the different HiFi Taq
-{[}Figure 3{]}. The total seqeunces with at least one error matched the
-median average per base error except for the Accuprime HiFi Taq
-{[}Figure S1{]}. Although it had the lowest per base error rate it had
-either the largest or second largest number of sequences with at least
-one error regardless of cycle number of sub-sampling depth {[}Figure
-S1{]}. Investigation of whether there were Taq dependent effects on base
-subsitution found that there was no clear bias and this was indpendent
-of sub sampling depth {[}Figure S2-S4{]}. Further, the variation in
-subsitution error seems to reduce as the sequencing depth increases
-{[}Figure S2-S4{]}.
+Number Used:}} Differences by HiFi Taq in the median average per base
+error varied without a clear pattern across sub-sampling depth {[}{]}.
+Generally, the highest values were for the Kappa HiFI Taq {[}Figure
+3A{]}. Sub-sampling depth seems to have little effect on this rate with
+both 5000 and 10000 sub-sampled sequences showing similar results
+{[}Figure 3B \& C{]}. Generally, there were small differences between
+the various HiFi Taq at lower cycle number but larger differences at
+higher cycle number {[}Figure 3B \& C{]}. Platinum HiFi Taq consistently
+had the highes median average per base error rate while Phusion and
+Accuprime had the lowest {[}Figure 3B \& C{]}. These differences though
+were realtively small between the different HiFi Taq {[}Figure 3{]}. The
+total seqeunces with at least one error matched the median average per
+base error except for the Accuprime HiFi Taq {[}Figure S1{]}. Although
+it had the lowest per base error rate it had either the largest or
+second largest number of sequences with at least one error regardless of
+cycle number of sub-sampling depth {[}Figure S1{]}. Investigation of
+whether there were Taq dependent effects on base subsitution found that
+there was no clear bias and this was indpendent of sub sampling depth
+{[}Figure S2-S4{]}. Further, the variation in subsitution error seems to
+reduce as the sequencing depth increases {[}Figure S2-S4{]}.
 

### 2017-12-11 submission/supplemental.Rmd
add table s5

@@
 
-**Table S5: **
+**Table S5: Kruskal-Wallis Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
 
@@
 
+mock_error_rate_kruskal <- as.data.frame(read_csv("../data/process/tables/mock_error_overall_kruskal_results.csv")) %>% 
+  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  filter(sub_sample_level > 500)
+
+colnames(mock_error_rate_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+
+kable(mock_error_rate_kruskal, format = "markdown", align = 'c', padding = 2)
 ```

### 2017-12-11 submission/supplemental.Rmd
add table s6

@@
 
+mock_error_rate_dunn <- as.data.frame(read_csv("../data/process/tables/mock_error_overall_dunn_results.csv")) %>% 
+  mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  arrange(pvalue) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  select(-bh)
+
+colnames(mock_error_rate_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+
+kable(mock_error_rate_dunn, format = "markdown", align = 'c', padding = 2)
 ```

### 2017-12-11 submission/supplemental.Rmd
fix titles table s5 and s6
add table s7 and s8

@@
 
-**Table S5: Kruskal-Wallis Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
+**Table S5: Kruskal-Wallis Results of Per Base Error Rate Differences between HiFi Taq in Mock Samples**
 
@@
 
-**Table S6: **
+**Table S6: Dunn's Post-Hoc Results of Per Base Error Rate between HiFi Taq in Mock Samples**
 
@@
 
-**Table S7: **
+**Table S7: Kruskal-Wallis Results of Sequences with Error Differences between HiFi Taq in Mock Samples**
 
@@
 
+mock_error_count_kruskal <- as.data.frame(
+  read_csv("../data/process/tables/mock_error_count_overall_kruskal_results.csv")) %>% 
+  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  filter(sub_sample_level > 500)
+
+colnames(mock_error_count_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+
+kable(mock_error_count_kruskal, format = "markdown", align = 'c', padding = 2)
+
 ```
+
+\newpage
+
+**Table S8: Dunn's Post-Hoc Results of Sequences with Error Differences between HiFi Taq in Mock Samples**
+
+```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}
+
+mock_error_count_dunn <- as.data.frame(read_csv("../data/process/tables/mock_error_count_overall_dunn_results.csv")) %>% 
+  mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  arrange(pvalue) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  select(-bh)
+
+colnames(mock_error_count_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+
+kable(mock_error_count_dunn, format = "markdown", align = 'c', padding = 2)
+
+```

### 2017-12-11 submission/manuscript.Rmd
complete error result section -- more finished

@@
 
-***Per Base Error Rate is Dependent on both Taq and Cycle Number Used:***
-Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [].
+***Sequence Error is Dependent on both Taq and Cycle Number Used:***
+Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
 Generally, the highest values were for the Kappa HiFI Taq [Figure 3A].
-Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B & C].
-Generally, there were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number [Figure 3B & C].
-Platinum HiFi Taq consistently had the highes median average per base error rate while Phusion and Accuprime had the lowest [Figure 3B & C].
-These differences though were realtively small between the different HiFi Taq [Figure 3].
-The total seqeunces with at least one error matched the median average per base error except for the Accuprime HiFi Taq [Figure S1].
-Although it had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number of sub-sampling depth [Figure S1].
-Investigation of whether there were Taq dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub sampling depth [Figure S2-S4].
-Further, the variation in subsitution error seems to reduce as the sequencing depth increases [Figure S2-S4].
+Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B-C].
+There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi Taq [Figure 3B-C and Table S6].
+
+The total seqeunces with at least one error had multiple differences at different cycle numbers and sub-sampling depth driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi Taq tested [Figure S1 & Table S7 & S8].
+Although Accuprime^TM^ had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number or sub-sampling depth [Figure S1].
+Investigation of whether there were Taq dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub-sampling depth [Figure S2-S4].
+Further, the variation in subsitution error seems to reduce as the sub-sampling depth increases [Figure S2-S4].
 

### 2017-12-11 submission/manuscript.tex
complete error result section -- more finished

@@
 
-\textbf{\emph{Per Base Error Rate is Dependent on both Taq and Cycle
-Number Used:}} Differences by HiFi Taq in the median average per base
-error varied without a clear pattern across sub-sampling depth {[}{]}.
+\textbf{\emph{Sequence Error is Dependent on both Taq and Cycle Number
+Used:}} Differences by HiFi Taq in the median average per base error
+varied without a clear pattern across sub-sampling depth {[}Table S5{]}.
 Generally, the highest values were for the Kappa HiFI Taq {[}Figure
@@
 both 5000 and 10000 sub-sampled sequences showing similar results
-{[}Figure 3B \& C{]}. Generally, there were small differences between
-the various HiFi Taq at lower cycle number but larger differences at
-higher cycle number {[}Figure 3B \& C{]}. Platinum HiFi Taq consistently
-had the highes median average per base error rate while Phusion and
-Accuprime had the lowest {[}Figure 3B \& C{]}. These differences though
-were realtively small between the different HiFi Taq {[}Figure 3{]}. The
-total seqeunces with at least one error matched the median average per
-base error except for the Accuprime HiFi Taq {[}Figure S1{]}. Although
-it had the lowest per base error rate it had either the largest or
-second largest number of sequences with at least one error regardless of
-cycle number of sub-sampling depth {[}Figure S1{]}. Investigation of
-whether there were Taq dependent effects on base subsitution found that
-there was no clear bias and this was indpendent of sub sampling depth
-{[}Figure S2-S4{]}. Further, the variation in subsitution error seems to
-reduce as the sequencing depth increases {[}Figure S2-S4{]}.
+{[}Figure 3B-C{]}. There were small differences between the various HiFi
+Taq at lower cycle number but larger differences at higher cycle number
+with Platinum having large differences between all other HiFi Taq
+{[}Figure 3B-C and Table S6{]}.
+
+The total seqeunces with at least one error had multiple differences at
+different cycle numbers and sub-sampling depth driven by large
+differences in Accuprime\textsuperscript{TM} and Platinum versus the
+other HiFi Taq tested {[}Figure S1 \& Table S7 \& S8{]}. Although
+Accuprime\textsuperscript{TM} had the lowest per base error rate it had
+either the largest or second largest number of sequences with at least
+one error regardless of cycle number or sub-sampling depth {[}Figure
+S1{]}. Investigation of whether there were Taq dependent effects on base
+subsitution found that there was no clear bias and this was indpendent
+of sub-sampling depth {[}Figure S2-S4{]}. Further, the variation in
+subsitution error seems to reduce as the sub-sampling depth increases
+{[}Figure S2-S4{]}.
 

### 2017-12-11 submission/supplemental.Rmd
add table s9 and 10

@@
 ```
+
+\newpage
+
+**Table S9: Kruskal-Wallis Results of Chimera Differences between HiFi Taq in Mock Samples**
+
+```{r supplemental_table9, echo=FALSE, message=FALSE, warning=FALSE}
+
+mock_chimera_kruskal <- as.data.frame(
+  read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv")) %>% 
+  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  filter(sub_sample_level > 500)
+
+colnames(mock_chimera_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+
+kable(mock_chimera_kruskal, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage
+
+**Table S10: Dunn's Post-Hoc Results of Chimera Differences between HiFi Taq in Mock Samples**
+
+```{r supplemental_table10, echo=FALSE, message=FALSE, warning=FALSE}
+
+mock_chimera_dunn <- as.data.frame(read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")) %>% 
+  mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  arrange(pvalue) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  select(-bh)
+
+colnames(mock_chimera_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+
+kable(mock_chimera_dunn, format = "markdown", align = 'c', padding = 2)
+
+```

### 2017-12-11 submission/manuscript.Rmd
refine chimera section -- closer to final edit

@@
 # Read in chimera prevalence results
-kruskal_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv")
+kruskal_chimera_prev <- as.data.frame(read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv"))
 dunn_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")
@@
 After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
-From this we observed that Platinum HiFi conssitently had the highest percentage of chimeric reads regardless of amplification cycle number and sub-sampling depth [Figure 4].
-For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs and this correlation was strongest for Platinum and Phusion HiFi Taq [Figure 5].
-In general the correlations between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
+At all levels of sub-sampling and cycle number there were significant differences between the HiFi Taq used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi Taq across cycle number and sub-sampling depth [Table S10].
+Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi Taq [Figure 4].
+
+For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs with this correlation being strongest for Platinum and Phusion HiFi Taq [Figure 5].
+In general, the R^2^ value between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
+Interestingly, Accuprime^TM^ not only had one of the lowest chimera prevalence but also consistently had the lowest correlations between the number of OTUs and chimeric sequences [Figure 5].
 

### 2017-12-11 submission/manuscript.tex
refine chimera section -- closer to final edit

@@
 sequences that did not classify as bacteria we assessed the percentage
-of sequences that were still chimeric within our mock community. From
-this we observed that Platinum HiFi conssitently had the highest
-percentage of chimeric reads regardless of amplification cycle number
-and sub-sampling depth {[}Figure 4{]}. For all Taqs a positive
-correlation was observed between chimeric sequences and number of OTUs
-and this correlation was strongest for Platinum and Phusion HiFi Taq
-{[}Figure 5{]}. In general the correlations between the number of OTUs
-and chimeric sequences became stronger as sub-sampling depth increased
-{[}Figure 5{]}.
+of sequences that were still chimeric within our mock community. At all
+levels of sub-sampling and cycle number there were significant
+differences between the HiFi Taq used (P-value \textless{} 0.05)
+{[}Table S9{]}. Using a Dunn's post-hoc test the vast majority of these
+differences were driven by Platinum being different then all other HiFi
+Taq across cycle number and sub-sampling depth {[}Table S10{]}.
+Generally, across sub-sampling depth and cycle number
+Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
+the HiFi Taq {[}Figure 4{]}.
+
+For all Taqs a positive correlation was observed between chimeric
+sequences and number of OTUs with this correlation being strongest for
+Platinum and Phusion HiFi Taq {[}Figure 5{]}. In general, the
+R\textsuperscript{2} value between the number of OTUs and chimeric
+sequences became stronger as sub-sampling depth increased {[}Figure
+5{]}. Interestingly, Accuprime\textsuperscript{TM} not only had one of
+the lowest chimera prevalence but also consistently had the lowest
+correlations between the number of OTUs and chimeric sequences {[}Figure
+5{]}.
 

### 2017-12-12 submission/manuscript.Rmd
add results -- final section better edit

@@
 
-For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs with this correlation being strongest for Platinum and Phusion HiFi Taq [Figure 5].
+For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Platinum and Phusion HiFi Taq [Figure 5].
 In general, the R^2^ value between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
-Interestingly, Accuprime^TM^ not only had one of the lowest chimera prevalence but also consistently had the lowest correlations between the number of OTUs and chimeric sequences [Figure 5].
+Taken together this data suggests that a strong correlation exists between the number of OTUs and chimera sequence prevalence.
+Interestingly, Accuprime^TM^ not only had one of the lowest prevalence of chimeric sequences but also, consistently, had the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 5].
 

### 2017-12-12 submission/manuscript.tex
add results -- final section better edit

@@
 For all Taqs a positive correlation was observed between chimeric
-sequences and number of OTUs with this correlation being strongest for
+sequences and number of OTUs, with this correlation being strongest for
 Platinum and Phusion HiFi Taq {[}Figure 5{]}. In general, the
@@
 sequences became stronger as sub-sampling depth increased {[}Figure
-5{]}. Interestingly, Accuprime\textsuperscript{TM} not only had one of
-the lowest chimera prevalence but also consistently had the lowest
-correlations between the number of OTUs and chimeric sequences {[}Figure
-5{]}.
+5{]}. Taken together this data suggests that a strong correlation exists
+between the number of OTUs and chimera sequence prevalence.
+Interestingly, Accuprime\textsuperscript{TM} not only had one of the
+lowest prevalence of chimeric sequences but also, consistently, had the
+lowest correlations between the number of OTUs and chimeric sequences
+across sub-sampling depth {[}Figure 5{]}.
 

### 2017-12-12 submission/manuscript.Rmd
update methods with Gohl reference

@@
 The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-The PCR cycle conditions were the same for every primer [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
+The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
 If the HiFi Taq had a specific activation time that was different then 2 minutes that was used instead.
+For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016].
+For Phusion the company defined conditions were used but the same extension time as that used for Accuprime and Platinum was used.
+
 The 30 cycle default was used but the cycle conditions started at 15 and increased by 5 up to 35 cycles and was used for both fecal and mock samples.

### 2017-12-12 submission/manuscript.tex
update methods with Gohl reference

@@
 (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
-England Biolabs, MA, USA). The PCR cycle conditions were the same for
-every primer (Kozich et al., 2013)
+England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
+Accuprime followed a previously published protocol (Kozich et al., 2013)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
 If the HiFi Taq had a specific activation time that was different then 2
-minutes that was used instead. The 30 cycle default was used but the
-cycle conditions started at 15 and increased by 5 up to 35 cycles and
-was used for both fecal and mock samples. The fecal PCR consisted of all
-4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples =
-100). Although, the mock communities also had 4 replicates used for 15,
-20, 25, and 35 cycles, 10 replicates were used for 30 cycles for all Taq
-(total samples = 130). For all the mock community samples there was not
-enough PCR product at 15 cycles for adequate sequencing.
+minutes that was used instead. For Kappa and Q5 the protocol previously
+published by Gohl and colleagues was used (Gohl et al., 2016). For
+Phusion the company defined conditions were used but the same extension
+time as that used for Accuprime and Platinum was used.
+
+The 30 cycle default was used but the cycle conditions started at 15 and
+increased by 5 up to 35 cycles and was used for both fecal and mock
+samples. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and
+35 cycles for each Taq (total samples = 100). Although, the mock
+communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10
+replicates were used for 30 cycles for all Taq (total samples = 130).
+For all the mock community samples there was not enough PCR product at
+15 cycles for adequate sequencing.
 
@@
 
+\hypertarget{ref-taq_Gohl_2016}{}
+Gohl DM., Vangay P., Garbe J., MacLean A., Hauge A., Becker A., Gould
+TJ., Clayton JB., Johnson TJ., Hunter R., Knights D., Beckman KB. 2016.
+Systematic improvement of amplicon marker gene methods for increased
+accuracy in microbiome studies. \emph{Nature Biotechnology} 34:942--949.
+DOI: \href{https://doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
+
 \hypertarget{ref-protocol_Kozich_2013}{}

### 2017-12-13 submission/manuscript.tex
change title

@@
 \section{What the Taq? The Influence of Different Hi-Fidelity Taq
-Polymerase on 16S rRNA
-Sequencing}\label{what-the-taq-the-influence-of-different-hi-fidelity-taq-polymerase-on-16s-rrna-sequencing}
+Polymerase on 16S rRNA Gene
+Sequencing}\label{what-the-taq-the-influence-of-different-hi-fidelity-taq-polymerase-on-16s-rrna-gene-sequencing}
 

### 2017-12-13 code/run_nmds_fecal_data.R
rough layout of b-diversity -- just in case

@@ -1,0 +1,111 @@
+### Create graphs of the different numOTUs based on cycle number and sub sampling
+### Fecal Sample Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "vegan"))
+
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = "")) %>% 
+    filter(sample_name %in% c("DA10001", "DA10016", "DA10029", "DA10040"))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+# Vector of sub samples used
+sub_sample_level <- c("1000", "5000", "10000", "15000", "20000")
+
+
+# Read in distance matrix
+braycurtis_dist <- read.dist('data/process/all_amp.braycurtis.0.03.lt.1000.dist')
+
+
+# Read in meta data
+metadata <- read_data("data/process/tables/", "meta_data", ".csv", "") %>% 
+  filter(full_name %in% rownames(braycurtis_dist))
+
+sample_to_keep <- metadata$full_name
+
+braycurtis_dist <- braycurtis_dist[sample_to_keep, sample_to_keep]
+
+# create select Taq distance table
+phu <- filter(metadata, taq == "PHU")[, "full_name"]
[…]

### 2017-12-13 submission/manuscript.Rmd
change title
fix polymerase statement and add first paragraph discussion

@@
 
-# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Sequencing
+# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Gene Sequencing
 
@@
 ***The Number of OTUs is Dependent on HiFi Taq Used:***
-After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi Taq used and this difference increased as the depth of sub-sampling increased [Figure 1].
-Lower cycle numbers (15-20) resulted in less differences between Taq while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
+After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi DNA polymerase used and this difference increased as the depth of sub-sampling increased [Figure 1].
+Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
 Only 35 cycles had HiFi Taq differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
-At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi Taq differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
+At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA polymerase differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
 Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other Taq across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
 
-This Taq dependent difference in the number of OTUs was also observed in the mock community samples with the same Taq polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2 & Table S3].
-Conversely, differences between HiFi Taq were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-Using a Tukey post-hoc test differences between Platinum and the other HiFi Taqs was the major driver of the differences seen at different cycle numbers and sub-sampling depths [Table S4].
+This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2 & Table S3].
+Conversely, differences between HiFi DNA polymerase were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+Using a Tukey post-hoc test differences between Platinum and the other HiFi DNA polymerases was the major driver of the differences seen at different cycle numbers and sub-sampling depths [Table S4].
 Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
@@
 Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest values were for the Kappa HiFI Taq [Figure 3A].
+Generally, the highest values were for the Kappa HiFI DNA polymerase [Figure 3A].
 Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B-C].
-There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi Taq [Figure 3B-C and Table S6].
+There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 3B-C and Table S6].
 
@@
 Although Accuprime^TM^ had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number or sub-sampling depth [Figure S1].
-Investigation of whether there were Taq dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub-sampling depth [Figure S2-S4].
+Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub-sampling depth [Figure S2-S4].
 Further, the variation in subsitution error seems to reduce as the sub-sampling depth increases [Figure S2-S4].
@@
 After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
-At all levels of sub-sampling and cycle number there were significant differences between the HiFi Taq used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi Taq across cycle number and sub-sampling depth [Table S10].
-Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi Taq [Figure 4].
+At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi DNA polymerase across cycle number and sub-sampling depth [Table S10].
+Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase [Figure 4].
 
-For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Platinum and Phusion HiFi Taq [Figure 5].
+For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Platinum and Phusion HiFi DNA Polymerase [Figure 5].
 In general, the R^2^ value between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
@@
 ## Discussion
+
+Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA Polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
+This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results as well as the findings.
+Although the attention has mostly been on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4].
+
+```{r}
+#Results have an impact on downstream diversity measurements due to effects on total OTUs
+
+#Need to cover that different types of polymerase may affect outcome
+#- Accuprime HiFi from recombinant A-Family Taq DNA polymerase, Pyrococcus spp GB-D polymerase (proof-reading), platinum Taq Antibody (inhibits room temp activity)
+
+#- Kappa HiFi from B-family DNA polymerase (proof-reading baked in), propietary antibody (inhibit room temp activity)
+
+#- Q5 HiFi from B-family thermostable DNA polymerase with 3´→ 5´ exonuclease activity, fused to a processivity-enhancing Sso7d domain to support robust DNA amplification
+
+#-Phusion HiFi B-family DNA polymerase a novel Pyrococcus-like enzyme fused with a processivity-enhancing domain, generate blunt-ended products.
+
+#- Platinum HiFi recombinant A-Family Taq DNA polymerase, Pyrococcus species GB-D polymeras, Platinum Taq Antibody
+
+```
 

### 2017-12-13 submission/manuscript.tex
fix polymerase statement and add first paragraph discussion

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
 \usepackage{graphicx,grffile}
@@
 fecal samples the number of OTUs identified was dependent upon the HiFi
-Taq used and this difference increased as the depth of sub-sampling
-increased {[}Figure 1{]}. Lower cycle numbers (15-20) resulted in less
-differences between Taq while cycle numbers of 25, 30, and 35 had larger
-clearer defined differences {[}Figure 1{]}. Only 35 cycles had HiFi Taq
-differences that were significantly different at all sub-sampling levels
-(P-value \textless{} 0.0001) {[}Table S1{]}. At sub-sampling depth of
-5000 or higher 25 and 30 cycles had HiFi Taq differences (P-value
-\textless{} 0.05) {[}Table S1{]}. Using a Tukey post-hoc test only at 35
-cycles were significant differences found to be mainly driven by
-Platinum being different than all other Taq across sub-sampling depth
-(P-value \textless{} 0.05) {[}Table S2{]}.
+DNA polymerase used and this difference increased as the depth of
+sub-sampling increased {[}Figure 1{]}. Lower cycle numbers (15-20)
+resulted in less differences between HiFI DNA polymerase while cycle
+numbers of 25, 30, and 35 had larger clearer defined differences
+{[}Figure 1{]}. Only 35 cycles had HiFi Taq differences that were
+significantly different at all sub-sampling levels (P-value \textless{}
+0.0001) {[}Table S1{]}. At sub-sampling depth of 5000 or higher 25 and
+30 cycles had HiFi DNA polymerase differences (P-value \textless{} 0.05)
+{[}Table S1{]}. Using a Tukey post-hoc test only at 35 cycles were
+significant differences found to be mainly driven by Platinum being
+different than all other Taq across sub-sampling depth (P-value
+\textless{} 0.05) {[}Table S2{]}.
 
-This Taq dependent difference in the number of OTUs was also observed in
-the mock community samples with the same Taq polymerases being high
-(Platinum) and low (Accuprime) respectively {[}Figure 2 \& Table S3{]}.
-Conversely, differences between HiFi Taq were observed as early as 20
-cycles and a sub-sampling depth of 1000 sequences (P-value = 0.002)
-{[}Table S3{]}. Using a Tukey post-hoc test differences between Platinum
-and the other HiFi Taqs was the major driver of the differences seen at
-different cycle numbers and sub-sampling depths {[}Table S4{]}. Both
-fecal and mock samples consistently showed that across sub-sampling
-depth and cycle number the lowest number of OTUs identified was from
+This HiFi DNA polymerase dependent difference in the number of OTUs was
+also observed in the mock community samples with the same DNA
+polymerases being high (Platinum) and low (Accuprime) respectively
+{[}Figure 2 \& Table S3{]}. Conversely, differences between HiFi DNA
+polymerase were observed as early as 20 cycles and a sub-sampling depth
+of 1000 sequences (P-value = 0.002) {[}Table S3{]}. Using a Tukey
+post-hoc test differences between Platinum and the other HiFi DNA
+polymerases was the major driver of the differences seen at different
+cycle numbers and sub-sampling depths {[}Table S4{]}. Both fecal and
+mock samples consistently showed that across sub-sampling depth and
+cycle number the lowest number of OTUs identified was from
 Accuprime\textsuperscript{TM} while the highest was from Platinum for
@@
 varied without a clear pattern across sub-sampling depth {[}Table S5{]}.
-Generally, the highest values were for the Kappa HiFI Taq {[}Figure
-3A{]}. Sub-sampling depth seems to have little effect on this rate with
-both 5000 and 10000 sub-sampled sequences showing similar results
-{[}Figure 3B-C{]}. There were small differences between the various HiFi
-Taq at lower cycle number but larger differences at higher cycle number
-with Platinum having large differences between all other HiFi Taq
-{[}Figure 3B-C and Table S6{]}.
+Generally, the highest values were for the Kappa HiFI DNA polymerase
+{[}Figure 3A{]}. Sub-sampling depth seems to have little effect on this
+rate with both 5000 and 10000 sub-sampled sequences showing similar
+results {[}Figure 3B-C{]}. There were small differences between the
+various HiFi Taq at lower cycle number but larger differences at higher
+cycle number with Platinum having large differences between all other
+HiFi DNA polymerase {[}Figure 3B-C and Table S6{]}.
 
@@
 one error regardless of cycle number or sub-sampling depth {[}Figure
-S1{]}. Investigation of whether there were Taq dependent effects on base
-subsitution found that there was no clear bias and this was indpendent
-of sub-sampling depth {[}Figure S2-S4{]}. Further, the variation in
-subsitution error seems to reduce as the sub-sampling depth increases
-{[}Figure S2-S4{]}.
+S1{]}. Investigation of whether there were HiFi DNA polymerase dependent
+effects on base subsitution found that there was no clear bias and this
+was indpendent of sub-sampling depth {[}Figure S2-S4{]}. Further, the
+variation in subsitution error seems to reduce as the sub-sampling depth
+increases {[}Figure S2-S4{]}.
 
@@
 levels of sub-sampling and cycle number there were significant
-differences between the HiFi Taq used (P-value \textless{} 0.05)
-{[}Table S9{]}. Using a Dunn's post-hoc test the vast majority of these
-differences were driven by Platinum being different then all other HiFi
-Taq across cycle number and sub-sampling depth {[}Table S10{]}.
-Generally, across sub-sampling depth and cycle number
+differences between the HiFi DNA polymerase used (P-value \textless{}
+0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast majority of
+these differences were driven by Platinum being different then all other
+HiFi DNA polymerase across cycle number and sub-sampling depth {[}Table
+S10{]}. Generally, across sub-sampling depth and cycle number
 Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
-the HiFi Taq {[}Figure 4{]}.
+the HiFi DNA Polymerase {[}Figure 4{]}.
 
@@
 sequences and number of OTUs, with this correlation being strongest for
-Platinum and Phusion HiFi Taq {[}Figure 5{]}. In general, the
+Platinum and Phusion HiFi DNA Polymerase {[}Figure 5{]}. In general, the
 R\textsuperscript{2} value between the number of OTUs and chimeric
@@
 \subsection{Discussion}\label{discussion}
+
+Our observations build upon previous studies (Gohl et al., 2016) by
+showing that even different HiFi DNA Polymerase have significant
+differences in the number of OTUs and that the changes to total OTUs
+correlate with chimeras not removed after sequence processing {[}Figure
+1-2 \& 5{]}. This is important since many diversity metrics rely on the
+total number of OTUs as part of their calculations and changes to the
+total number of OTUs could drastically change the results as well as the
+findings. Although the attention has mostly been on standardizing and
+improving collection and extraction methods (Salter et al., 2014) our
+observations show that independent of this consideration HiFi DNA
+polymerase can have a noticeable affect on the OTUs generated that can
+be found across sub-sampling depth and PCR cycle number {[}Figure
+2-4{]}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#Results have an impact on downstream diversity measurements due to effects on total OTUs}
+
+\CommentTok{#Need to cover that different types of polymerase may affect outcome}
+\CommentTok{#- Accuprime HiFi from recombinant A-Family Taq DNA polymerase, Pyrococcus spp GB-D polymerase (proof-reading), platinum Taq Antibody (inhibits room temp activity)}
+
+\CommentTok{#- Kappa HiFi from B-family DNA polymerase (proof-reading baked in), propietary antibody (inhibit room temp activity)}
+
+\CommentTok{#- Q5 HiFi from B-family thermostable DNA polymerase with 3´→ 5´ exonuclease activity, fused to a processivity-enhancing Sso7d domain to support robust DNA amplification}
+
+\CommentTok{#-Phusion HiFi B-family DNA polymerase a novel Pyrococcus-like enzyme fused with a processivity-enhancing domain, generate blunt-ended products.}
+
+\CommentTok{#- Platinum HiFi recombinant A-Family Taq DNA polymerase, Pyrococcus species GB-D polymeras, Platinum Taq Antibody}
+\end{Highlighting}
+\end{Shaded}
 
@@
 \href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-contamination_Salter2014}{}
+Salter SJ., Cox MJ., Turek EM., Calus ST., Cookson WO., Moffatt MF.,
+Turner P., Parkhill J., Loman NJ., Walker AW. 2014. Reagent and
+laboratory contamination can critically impact sequence-based microbiome
+analyses. \emph{BMC Biology} 12. DOI:
+\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
 

### 2017-12-13 code/run_nmds_fecal_data.R
first commit

@@ -54,12 +54,13 @@ sub_sample_level <- c("1000", "5000", "10000", "15000", "20000")
 
 
 # Read in distance matrix
-braycurtis_dist <- read.dist('data/process/all_amp.braycurtis.0.03.lt.1000.dist')
+braycurtis_dist <- sapply(sub_sample_level, 
+                          function(x) read.dist(paste("data/process/all_amp.braycurtis.0.03.lt.", 
+                                                      x, ".dist", sep = "")), simplify = F)
 
 
 # Read in meta data
-metadata <- read_data("data/process/tables/", "meta_data", ".csv", "") %>% 
-  filter(full_name %in% rownames(braycurtis_dist))
+metadata <- read_data("data/process/tables/", "meta_data", ".csv", "")
 
 sample_to_keep <- metadata$full_name
 

### 2017-12-13 code/run_nmds_fecal_data.R
complete initial aggregation and munging

@@ -29,7 +29,27 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 }
 
 
+# Create unique metadata files for each sub-sampling
+paredown_meta <- function(depth, metafile, distfile){
+  
+  tempvalues <- rownames(distfile[[depth]])
+  
+  metafile <- metafile %>% filter(full_name %in% tempvalues)
+  
+  return(metafile)
+  
+}
 
+# Pare down distance matrix to match metadata file for each sub-sampling
+paredown_dist <- function(depth, metaList, distList){
+  
+  tempMeta_values <- as.data.frame(metaList[[depth]])[, "full_name"]
+  
+  tempData <- distList[[depth]][tempMeta_values, tempMeta_values]
+  
+  
+  return(tempData)
+}
 
 
 
@@ -61,6 +81,16 @@ braycurtis_dist <- sapply(sub_sample_level,
 
 # Read in meta data
 metadata <- read_data("data/process/tables/", "meta_data", ".csv", "")
+
+# Generate pared down meta matching each sub-sampling
+meta_list <- sapply(sub_sample_level, 
+                    function(x) paredown_meta(x, metadata, braycurtis_dist), simplify = F)
+
+# Generate pared down dist list for each sub-sampling that matches meta-data
+red_bray_dist <- sapply(sub_sample_level, 
+                    function(x) paredown_dist(x, meta_list, braycurtis_dist), simplify = F)
+
+
 
 sample_to_keep <- metadata$full_name
 

### 2017-12-13 code/run_nmds_fecal_data.R
complete data munging

@@ -52,7 +52,27 @@ paredown_dist <- function(depth, metaList, distList){
 }
 
 
-
+# Function to seperate each DNA polymerase within each sub sampling
+separate_polymerase <- function(depth, metaList, distList, 
+                                taq_used = c("ACC", "K", "PHU", "PL", "Q5"), meta_readout = F){
+  
+  tempMeta <- sapply(taq_used, 
+                     function(x) filter(metaList[[depth]], taq == x), simplify = F)
+  
+  namesList <- sapply(taq_used, function(x) as.data.frame(tempMeta[[x]])[, "full_name"])
+  
+  tempDist <- sapply(taq_used, 
+                     function(x) distList[[depth]][namesList[[x]], namesList[[x]]], simplify = F)
+  
+  if(meta_readout == F){
+    
+    return(tempDist)
+  } else{
+    
+    return(tempMeta)
+  }
+  
+}
 
 
 
@@ -89,6 +109,25 @@ meta_list <- sapply(sub_sample_level,
 # Generate pared down dist list for each sub-sampling that matches meta-data
 red_bray_dist <- sapply(sub_sample_level, 
                     function(x) paredown_dist(x, meta_list, braycurtis_dist), simplify = F)
+
+
+# Separate each polymerase within each sub-sampling
+sep_red_bray_dist <- sapply(sub_sample_level, 
+                        function(x) separate_polymerase(x, meta_list, braycurtis_dist), simplify = F)
+
+sep_meta_list <- sapply(sub_sample_level, 
+                            function(x) separate_polymerase(x, meta_list, 
+                                                            braycurtis_dist, meta_readout = T), simplify = F)
+
+
+
+
+
+
+
+
+
+
 
 
 

### 2017-12-13 submission/manuscript.Rmd
complete first paragraph of discussion

@@
 Although the attention has mostly been on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4].
+These differences were observed in high biomass samples, where biases introduced by such components like kit contamination have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 #Results have an impact on downstream diversity measurements due to effects on total OTUs

### 2017-12-13 submission/manuscript.tex
complete first paragraph of discussion

@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
 \usepackage{graphicx,grffile}
@@
 be found across sub-sampling depth and PCR cycle number {[}Figure
-2-4{]}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#Results have an impact on downstream diversity measurements due to effects on total OTUs}
-
-\CommentTok{#Need to cover that different types of polymerase may affect outcome}
-\CommentTok{#- Accuprime HiFi from recombinant A-Family Taq DNA polymerase, Pyrococcus spp GB-D polymerase (proof-reading), platinum Taq Antibody (inhibits room temp activity)}
-
-\CommentTok{#- Kappa HiFi from B-family DNA polymerase (proof-reading baked in), propietary antibody (inhibit room temp activity)}
-
-\CommentTok{#- Q5 HiFi from B-family thermostable DNA polymerase with 3´→ 5´ exonuclease activity, fused to a processivity-enhancing Sso7d domain to support robust DNA amplification}
-
-\CommentTok{#-Phusion HiFi B-family DNA polymerase a novel Pyrococcus-like enzyme fused with a processivity-enhancing domain, generate blunt-ended products.}
-
-\CommentTok{#- Platinum HiFi recombinant A-Family Taq DNA polymerase, Pyrococcus species GB-D polymeras, Platinum Taq Antibody}
-\end{Highlighting}
-\end{Shaded}
+2-4{]}. These differences were observed in high biomass samples, where
+biases introduced by such components like kit contamination have less of
+an effect, suggesting that these differences may be exacerbated in low
+biomass samples.
 

### 2017-12-13 code/run_nmds_fecal_data.R
add distance value picking

@@ -75,13 +75,78 @@ separate_polymerase <- function(depth, metaList, distList,
 }
 
 
+# Function that pulls distances from the same individual but 5 cycles later from table 
+get_next_distance <- function(depth, metaList, distList, 
+                              taq_used = c("ACC", "K", "PHU", "PL", "Q5")){
+  
+  tempMeta <- metaList[[depth]]
+  tempDist <- distList[[depth]]
+  
+  final_data <- sapply(taq_used, 
+         function(x) pull_dist_data(x, tempMeta, tempDist), simplify = F)
+  
+  final_data <- try(final_data %>% bind_rows())
+  
+  return(as.data.frame(final_data))
+  
+}
 
 
+# Function that actuall gets the distances of the between current and previous cycle
+pull_dist_data <- function(taq_of_int, metaList, distList){
+  
+  tempMeta <- metaList[[taq_of_int]]
+  tempDist <- distList[[taq_of_int]]
+  
+  unique_names <- unique(tempMeta$sample_name)
+  
+  tempData <- sapply(unique_names, 
+         function(x) run_each_sample(x, taq_of_int, tempMeta, tempDist), simplify = F)
+  
+  tempData <- try(tempData %>% bind_rows())
+    
+  return(tempData)
+  
+}
 
 
-
-
-
+# Function that controls the flow within a single sample 
+run_each_sample <- function(u_names, taq_used, meta_file, distList){
+  
+  temp_names <- as.data.frame(filter(meta_file, sample_name == u_names))[, "full_name"]
+  
+  row_values <- temp_names[1:(length(temp_names)-1)]
+  
+  col_values <- temp_names[2:length(temp_names)]
+  
+  tempData <- t(sapply(c(1:length(row_values)), 
+                     function(x) 
+                       c(initial = row_values[x], 
+                         final = col_values[x],
+                         sample_name = u_names, 
+                         taq = taq_used, 
+                         distance = try(distList[row_values[x], col_values[x]])), simplify = T))
+  
+  if("distance" %in% colnames(tempData)){
+    
[…]

### 2017-12-13 code/run_nmds_fecal_data.R
YESgit add code/! complete selective pulling of distances by sample

@@ -149,6 +149,26 @@ run_each_sample <- function(u_names, taq_used, meta_file, distList){
 }
 
 
+# Function to expand meta data tables into a nice table
+make_nice_table <- function(depth, dataList){
+  
+  tempData <- dataList[[depth]] %>% 
+    select(-taq, -sample_name) %>% 
+    separate(initial, c("lower_cycle", "taq", "kit", "sample_type", "sample_name")) %>% 
+    select(-taq, -kit, -sample_type, -sample_name) %>% 
+    separate(final, c("upper_cycle", "taq", "kit", "sample_type", "sample_name")) %>% 
+    select(-kit, -sample_type) %>% 
+    mutate(
+      cycle_compare = paste(
+        str_replace(lower_cycle, "x", "") , "to", str_replace(upper_cycle, "x", ""), sep = "")) %>% 
+    select(cycle_compare, taq, sample_name, distance)
+    
+    
+  return(tempData)
+}
+
+
+
 
 ###########################################################################################################################
 ############################### Run actual analysis programs  #############################################################
@@ -188,7 +208,8 @@ sep_meta_list <- sapply(sub_sample_level,
 next_distance_values <- sapply(sub_sample_level, 
                function(x) get_next_distance(x, sep_meta_list, sep_red_bray_dist), simplify = F)
 
-
+finalized_tables <- sapply(sub_sample_level, 
+               function(x) make_nice_table(x, next_distance_values), simplify = F)
 
 
 

### 2017-12-13 code/run_nmds_fecal_data.R
completed krukal test function

@@ -160,7 +160,8 @@ make_nice_table <- function(depth, dataList){
     select(-kit, -sample_type) %>% 
     mutate(
       cycle_compare = paste(
-        str_replace(lower_cycle, "x", "") , "to", str_replace(upper_cycle, "x", ""), sep = "")) %>% 
+        str_replace(lower_cycle, "x", "") , "to", str_replace(upper_cycle, "x", ""), sep = ""), 
+      distance = as.numeric(distance)) %>% 
     select(cycle_compare, taq, sample_name, distance)
     
     
@@ -168,6 +169,27 @@ make_nice_table <- function(depth, dataList){
 }
 
 
+# Function to run the krukal wallis test and store the P-values
+run_kruskal <- function(depth, dataList){
+  
+  taq_used <- c("ACC", "K", "PHU", "PL", "Q5")
+  
+  tempTest <- sapply(taq_used, 
+                     function(x) try(kruskal.test(distance ~ factor(cycle_compare), 
+                           data = filter(dataList[[depth]], taq == x))$p.value), simplify = F)
+
+  tempTest <- as.data.frame(t(tempTest %>% 
+    bind_rows())) %>% 
+    mutate(taq = rownames(.), 
+           pvalue = V1, 
+           pvalue = ifelse(grepl("Error", pvalue) == T, invisible(NA), 
+                           invisible(as.numeric(as.character(pvalue)))), 
+           bh = p.adjust(pvalue, method = "BH")) %>% 
+    select(taq, pvalue, bh)
+    
+  
+  return(tempTest)
+}
 
 
 ###########################################################################################################################
@@ -214,8 +236,11 @@ finalized_tables <- sapply(sub_sample_level,
 
 
 
+kruskal_test_data <- sapply(sub_sample_level, 
+               function(x) run_kruskal(x, finalized_tables), simplify = F)
 
 
+### Need to do PERMANOVAs by taq and subsampling ---probably don't need visualizations
 
 
 #sample_to_keep <- metadata$full_name

### 2017-12-13 code/run_nmds_fecal_data.R
add permanova results

@@ -184,11 +184,37 @@ run_kruskal <- function(depth, dataList){
            pvalue = V1, 
            pvalue = ifelse(grepl("Error", pvalue) == T, invisible(NA), 
                            invisible(as.numeric(as.character(pvalue)))), 
-           bh = p.adjust(pvalue, method = "BH")) %>% 
-    select(taq, pvalue, bh)
+           bh = p.adjust(pvalue, method = "BH"), 
+           sub_sample_level = depth) %>% 
+    select(taq, sub_sample_level, pvalue, bh)
     
   
   return(tempTest)
+}
+
+# Create function to run PERMANOVA analysis by subsampling by Taq
+get_permanova <- function(depth, metaList, dataList){
+  
+  taq_used <- c("ACC", "K", "PHU", "PL", "Q5")
+  
+  tempMeta <- metaList[[depth]]
+  tempData <- dataList[[depth]]
+  
+  set.seed(12345)
+  tempTest <- sapply(taq_used, 
+                     function(x) 
+                       try(adonis(as.dist(tempData[[x]]) ~ factor(tempMeta[[x]]$cycles))$aov.tab$`Pr(>F)`[1]), 
+                     simplify = F) %>% bind_rows() %>% t() %>% as.data.frame() %>% 
+    mutate(taq = rownames(.), 
+           pvalue = V1, 
+           pvalue = ifelse(grepl("Error", pvalue) == T, invisible(NA), invisible(as.character(pvalue))), 
+           pvalue = as.numeric(pvalue), 
+           bh = p.adjust(pvalue, method = "BH"), 
+           sub_sample_level = depth) %>% 
+    select(taq, sub_sample_level, pvalue, bh)
+    
+  return(tempTest)
+  
 }
 
 
@@ -233,61 +259,20 @@ next_distance_values <- sapply(sub_sample_level,
 finalized_tables <- sapply(sub_sample_level, 
                function(x) make_nice_table(x, next_distance_values), simplify = F)
 
-
-
-
+# Get kruskal-wallis results on similarity to last cycle group
 kruskal_test_data <- sapply(sub_sample_level, 
-               function(x) run_kruskal(x, finalized_tables), simplify = F)
[…]

### 2017-12-14 submission/manuscript.Rmd
add bray-curtis diffs to manuscript

@@
 dunn_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")
+
+# Read in results from PERMANOVA and bray analysis
+bray_sim_to_prev_cycle_kruskal_results <- as.data.frame(read_csv("../data/process/tables/bray_sim_to_prev_cycle_kruskal_results.csv"))
+bray_permanova_by_taq_results <- as.data.frame(read_csv("../data/process/tables/bray_permanova_by_taq_results.csv"))
 
@@
 For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between different people.
+Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
+The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R.
 For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
@@
 The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
+Cycle dependent affects on Bray-Curits distances were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
+20 versus 25, 25 versus 30, etc.).
 From these observations we wanted to next analyze potential reasons as to why some of these differences may have occured.
@@
 Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
+
+***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
+Using PERMANOVA to test whether there were any differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
+Interestingly, Phusion was one of only two DNA polymerase that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
+Next, we assessed whether there were any major differences between 5 cycle increments within each sample.
+We found that there was no detectable difference in distance when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
+It should be noted that at higher sub-sampling depths these differences in Bray-Curits distance disappear for both differences in cycle number and within 5 cycle increments within an individual.
 

### 2017-12-14 submission/manuscript.tex
add bray-curtis diffs to manuscript

@@
 each individual by cycle number to account for the biological variation
-between different people. For both error and chimera analysis, samples
-were tested using Kruskal-Wallis with a Dunns post-hoc test. Where
-applicable correction for multiple comparison utilized the
-Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
+between different people. Bray-Curtis distance matrices were generated
+using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
+The distance matrix data was analyzed using PERMANOVA with the vegan
+package (v 2.4.4) (Oksanen et al., 2017) and kruskal-wallis tests within
+R. For both error and chimera analysis, samples were tested using
+Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
+for multiple comparison utilized the Benjamini-Hochberg method
+(Benjamini \& Hochberg, 1995).
 
@@
 sub-sampling was analyzed for both the fecal and mock community samples.
-From these observations we wanted to next analyze potential reasons as
-to why some of these differences may have occured. First, analysis of
-general sequence error rate, number of sequences with an error, and base
-subsitution were assessed in the mock community for each Taq. After
-assessing these errors, the total number of chimeras was determined
-after sequence processing. The fecal samples were analyzed at 4
-different sub-sampling levels, 1000, 5000, 10000, and 15000 while the
+Cycle dependent affects on Bray-Curits distances were next assessed for
+the fecal samples looking at both overall cycle differences and within
+individual differences for the previous cycle (e.g.~20 versus 25, 25
+versus 30, etc.). From these observations we wanted to next analyze
+potential reasons as to why some of these differences may have occured.
+First, analysis of general sequence error rate, number of sequences with
+an error, and base subsitution were assessed in the mock community for
+each Taq. After assessing these errors, the total number of chimeras was
+determined after sequence processing. The fecal samples were analyzed at
+4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the
 mock community samples were analysed at 3 levels, 1000, 5000, 10000.
@@
 both fecal and mock samples {[}Figure 1 \& 2{]}.
+
+\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
+Number:}} Using PERMANOVA to test whether there were any differences
+within HiFi DNA polymerase groups based on cycle number, only Phusion
+had cycle dependent differences at 1000 and 5000 sub-sampling depth
+(P-value = 0.03 and 0.01, respectively). Interestingly, Phusion was one
+of only two DNA polymerase that managed to have samples for the 1000
+sub-sampling depth at 15 cycles. Next, we assessed whether there were
+any major differences between 5 cycle increments within each sample. We
+found that there was no detectable difference in distance when comparing
+to the previous 5 cycle increment (P-value \textgreater{} 0.05).
+However, Phusion at 1000 sub-sampling depth had a P-value = 0.02 before
+multiple comparison correction. It should be noted that at higher
+sub-sampling depths these differences in Bray-Curits distance disappear
+for both differences in cycle number and within 5 cycle increments
+within an individual.
 
@@
 
+\hypertarget{ref-vegan_citation}{}
+Oksanen J., Blanchet FG., Friendly M., Kindt R., Legendre P., McGlinn
+D., Minchin PR., O'Hara RB., Simpson GL., Solymos P., Stevens MHH.,
+Szoecs E., Wagner H. 2017. \emph{Vegan: Community ecology package}.
+
 \hypertarget{ref-r_citation_2017}{}

### 2017-12-14 submission/manuscript.Rmd
add paragraph on bray-curtis to discussion

@@
 For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between different people.
-Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
+Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
 The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R.
@@
 The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curits distances were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
+Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
 20 versus 25, 25 versus 30, etc.).
@@
 Next, we assessed whether there were any major differences between 5 cycle increments within each sample.
-We found that there was no detectable difference in distance when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+We found that there was no detectable difference in Bray-Curits index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
 However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
-It should be noted that at higher sub-sampling depths these differences in Bray-Curits distance disappear for both differences in cycle number and within 5 cycle increments within an individual.
+It should be noted that at higher sub-sampling depths these differences in Bray-Curits indicies disappear for both differences in cycle number and within 5 cycle increments within an individual.
 
@@
 These differences were observed in high biomass samples, where biases introduced by such components like kit contamination have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
+
+Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30, that are commonly used.
+What is most interesting is that there was no difference within individuals between corresponding 5 cycle increments (e.g.
+15 to 20, 20 to 25, etc.).
+This finding, in conjunction with the PERMANOVA results, would suggest that differences based on Bray-Curtis incrementaly build up so that within 5 cycle increments the differences are not that large but when comparing larger cycle differences, as with the PERMANOVA, these differences start to become detectable.
+Although our data suggests that it is possible to overcome these types of cycle dependent differences with increased sub-sampling depth, for some DNA polymerase increased sub-sampling depth increases chimera prevalence and per base error rate.
+Conversely, it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant and the Bray-Curtis index is able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
 

### 2017-12-14 code/run_seq_error_table_creations.R
run full data without sub-sampling

@@ -103,30 +103,32 @@ run_recombine_data <- function(i, RsamplingList, fullDataList){
 
 
 # Function to randomly sample each sample to a specified depth
-get_random_sample <- function(depth, dataTable, sample_vector){
+get_samples <- function(depth, dataTable, sample_vector){
 
+  
   tempList <- sapply(sample_vector, 
                      function(x) filter(dataTable, full_name == x), simplify = F)
   
   tempSamplingList <- sapply(sample_vector, 
                              function(x) get_rand_vectors(x, tempList))
   
-  above_depth_List <- sapply(sample_vector, 
-                             function(x) get_depth_check(x, depth, tempSamplingList), simplify = F)
+  #above_depth_List <- sapply(sample_vector, 
+   #                          function(x) get_depth_check(x, depth, tempSamplingList), simplify = F)
   
   # Remove null elements
-  above_depth_List <- above_depth_List[!sapply(above_depth_List, is.null)] 
+  #above_depth_List <- above_depth_List[!sapply(above_depth_List, is.null)] 
   
   # Runs the random sampling across samples
-  tempRSamplingList <- lapply(above_depth_List, 
-                              function(x) run_random_sampling(depth, x))
+  #tempRSamplingList <- lapply(above_depth_List, 
+  #                            function(x) run_random_sampling(depth, x))
   
   # Replace the current sample vectors with only those included in the random sampling
-  sample_vector <- names(tempRSamplingList)
+  #sample_vector <- names(tempRSamplingList)
+  sample_vector <- names(tempSamplingList)
   
   # Recombine with full data set
   tempRecombinedList <- sapply(sample_vector, 
-                               function(x) run_recombine_data(x, tempRSamplingList, tempList), simplify = F)
+                               function(x) run_recombine_data(x, tempSamplingList, tempList), simplify = F)
   
   return(tempRecombinedList)
 }
@@ -146,14 +148,14 @@ run_checks <- function(dataTable){
 # Function to create summary data
 get_summary_data <- function(dataList, depth){
   
-  dataList <- dataList[[as.character(depth)]]
+  #dataList <- dataList[[as.character(depth)]]
   
   tempData <- dataList %>% 
[…]

### 2017-12-14 code/run_seq_error_table_creations.R
add total sequence component

@@ -74,9 +74,10 @@ get_rand_vectors <- function(i, dataList){
 # Function to check if total sequence is equal to or greater than depth and toss those that arent
 get_depth_check <- function(i, depth_check, dataList){
   
-  tempData <- dataList[[i]]
+  tempData <- dataList
+  #tempData <- dataList[[i]]
   
-  ifelse(length(tempData) >= depth_check, 
+  ifelse(length(rownames(tempData)) >= depth_check, 
          return(tempData), return(NULL))
   
 }
@@ -113,12 +114,14 @@ get_samples <- function(depth, dataTable, sample_vector){
                              function(x) get_rand_vectors(x, tempList))
   
   #above_depth_List <- sapply(sample_vector, 
-   #                          function(x) get_depth_check(x, depth, tempSamplingList), simplify = F)
+  #                           function(x) get_depth_check(x, depth, tempSamplingList), simplify = F)
   
   # Remove null elements
   #above_depth_List <- above_depth_List[!sapply(above_depth_List, is.null)] 
   
   # Runs the random sampling across samples
+  
+  
   #tempRSamplingList <- lapply(above_depth_List, 
   #                            function(x) run_random_sampling(depth, x))
   
@@ -254,8 +257,14 @@ testRecombine <- lapply(testRecombine, function(y) run_checks(y))
 
 # Generate Summary Data
 good_summary_data <- get_summary_data(testRecombine, "full_data") %>% 
+  mutate(sample_name = ifelse(grepl("Zmock_A_111716", full_name) == T, invisible("A"), 
+                              ifelse(grepl("Zmock_B_", full_name) == T, invisible("B"), 
+                                     ifelse(grepl("Zmock_C_", full_name) == T, invisible("C"), 
+                                            ifelse(grepl("Zmock_D_", full_name) == T, 
+                                                   invisible("D"), invisible(sample_name)))))) %>% 
   group_by(taq, cycles, sample_name) %>% 
-  summarise(mean_error = mean(error, na.rm = T), sd_error = sd(error, na.rm = T),  
+  summarise(mean_error = mean(error, na.rm = T), sd_error = sd(error, na.rm = T), 
+            total_seqs = length(query), 
             chimera_prevalence = sum(chimera)/length(query), 
             seq_error_prevalence = sum(seq_error)/length(query)) %>% 
[…]

### 2017-12-14 code/make_mock_chimera_count_graphs.R
update to use all data with at least 1000 sequences present

@@ -32,16 +32,19 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 
 
 # Vector of sub samples used
-sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+#sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
 
 # Read in the count data
-error_data <- sapply(sub_sample_level, 
-                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
-                     simplify = F)
+error_data <- read_csv("data/process/tables/full_error_summary.csv")
+
+#error_data <- sapply(sub_sample_level, 
+#                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
+#                     simplify = F)
 
 
 # Generate graph of Mock DNA samples 
-thousand <- error_data[["1000"]] %>% 
+full_graph <- error_data %>% 
+  filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
@@ -60,82 +63,115 @@ thousand <- error_data[["1000"]] %>%
   labs(x = "Amplification Cycles", y = "Chimeric Sequences") + 
   coord_cartesian(ylim = c(0, 0.10)) + 
   scale_y_continuous(labels = scales::percent) + 
-  ggtitle("A") + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.7, y = 0.102, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10), 
-        legend.position = c(0.25, 0.83), 
-        legend.title = element_blank(), 
-        legend.key = element_blank(), 
-        legend.background = element_rect(color = "black"))
-
-
-five_thousand <- error_data[["5000"]] %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
[…]

### 2017-12-14 code/make_mock_error_graphs.R
use data with at least 1000 sequences

@@ -32,19 +32,12 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################
 
-
-# Vector of sub samples used
-sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
-
 # Read in the count data
-error_data <- sapply(sub_sample_level, 
-                      function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
-                      simplify = F)
-
-
+error_data <- read_csv("data/process/tables/full_error_summary.csv")
 
 # Generate graph of Mock DNA samples 
-thousand <- error_data[["1000"]] %>% 
+full_graph <- error_data %>% 
+  filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
@@ -53,84 +46,124 @@ thousand <- error_data[["1000"]] %>%
             group_iqr75 = quantile(mean_error)["75%"]) %>% 
   ggplot(aes(cycles, group_median, color = taq, group = taq)) + 
   geom_line() + 
-  geom_errorbar(aes(ymin=group_iqr25, ymax=group_iqr75), width = 0.1, size = 0.5, alpha = 0.4) + 
+  geom_errorbar(aes(ymin=group_iqr25, 
+                    ymax=group_iqr75), 
+                width = 0.1, size = 0.5, alpha = 0.4) + 
   geom_point(size = 2, alpha = 0.7) + 
   theme_bw() + 
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Amplification Cycles", y = "Median Average Error Rate Per Base") + 
-  coord_cartesian(ylim = c(0, 0.03)) + 
-  scale_y_continuous(labels = scales::percent) + 
-  ggtitle("A") + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.5, y = 0.031, size = 2.5) + 
[…]

### 2017-12-14 code/make_mock_error_graphs.R
completed code

@@ -166,4 +166,4 @@ full_graph <- error_data %>%
 # 
 # combined_graph <- grid.arrange(thousand, five_thousand, ten_thousand, ncol = 3)
 
-ggsave("results/figures/Figure3.pdf", full_graph, width = 11, height = 7, dpi = 300)
+ggsave("results/figures/Figure3.pdf", full_graph, width = 7, height = 7, dpi = 300)

### 2017-12-14 code/make_mock_seq_error_count_graphs.R
use all seqs in samples with at least 1000

@@ -30,18 +30,17 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################
 
-
-# Vector of sub samples used
-sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
-
 # Read in the count data
-error_data <- sapply(sub_sample_level, 
-                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
-                     simplify = F)
+error_data <- read_csv("data/process/tables/full_error_summary.csv")
+
+#error_data <- sapply(sub_sample_level, 
+#                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
+#                     simplify = F)
 
 
 # Generate graph of Mock DNA samples 
-thousand <- error_data[["1000"]] %>% 
+full_graph <- error_data %>% 
+  filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
@@ -58,82 +57,123 @@ thousand <- error_data[["1000"]] %>%
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Amplification Cycles", y = "Percent of Sequences with Error") + 
-  coord_cartesian(ylim = c(0, 0.15)) + 
+  coord_cartesian(ylim = c(0, 0.10)) + 
   scale_y_continuous(labels = scales::percent) + 
-  ggtitle("A") + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 1.7, y = 0.155, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10), 
-        legend.position = c(0.75, 0.85), 
[…]

### 2017-12-14 code/run_error_analysis.R
update to test all sequences in samples with at least 1000 sequences

@@ -34,7 +34,8 @@ run_comparison <- function(i, col_of_int, taq_name, dataList){
   # Add a vector with all possible cycles analyzed
   amp_cycles <- c("15x", "20x", "25x", "30x", "35x")
   # Pull out data table of interest
-  tempData <- dataList[[i]]
+  tempData <- dataList
+  #tempData <- dataList[[i]]
   # Runs the ANOVA across all the different amp cycles
   tempTests <- sapply(amp_cycles, 
                       function(x) run_kruskal(x, i, col_of_int, taq_name, tempData), simplify = F)
@@ -86,8 +87,9 @@ run_dunn <- function(i, col_of_int, taq_name, dataList, rawData){
   # rawData is is the combined count and meta data list
   
   # Remove all samples that have a BH corrected value of over 0.05
-  tempData <- dataList[[i]] %>% 
-    filter(bh < 0.05)
+  tempData <- dataList %>% filter(bh < 0.05)
+  # tempData <- dataList[[i]] %>% 
+  #   filter(bh < 0.05)
   # Pull out the specific cycle numbers that this occured for
   tempVector <- as.data.frame(tempData)[, "cycle"]
   # Check if the vector is empty or not
@@ -113,7 +115,8 @@ get_dunn_test <- function(subsample, cycle_num, comparator_var, taq_var, dataTab
   # dataTable is the combined count and meta data list
   
   # filter the data based on subsample level and cycle number of interest
-  tempData <- as.data.frame(dataTable[[subsample]] %>% filter(cycles == cycle_num))
+  tempData <- as.data.frame(dataTable %>% filter(cycles == cycle_num))
+  #tempData <- as.data.frame(dataTable[[subsample]] %>% filter(cycles == cycle_num))
   
   tempValues <- tempData[, comparator_var]
   tempLabels <- as.factor(tempData[, taq_var])
@@ -139,65 +142,75 @@ get_dunn_test <- function(subsample, cycle_num, comparator_var, taq_var, dataTab
 ###########################################################################################################################
 
 # Read in subsample.shared files
-sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+# sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+# 
+# # Read in the count data
+error_data <- read_csv("data/process/tables/full_error_summary.csv") %>% 
+  filter(total_seqs >= 1000)
 
-# Read in the count data
-error_data <- sapply(sub_sample_level, 
[…]

### 2017-12-14 code/make_nucleotide_graphs.R
use all seqs in samples with more than 1000 seqs

@@ -44,104 +44,137 @@ run_data_gather <- function(i, dataList){
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################
 
-
-# Vector of sub samples used
-sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
-
 # Read in the count data
-nucleotide_data <- sapply(sub_sample_level, 
-                     function(x) read_data("data/process/tables/nucleotide_error_", "", "_summary.csv", x), 
-                     simplify = F)
-
-# Transform data for better graphing
-nucleotide_data <- sapply(sub_sample_level, 
-               function(x) run_data_gather(x, nucleotide_data), simplify = F)
-
+nucleotide_data <- read_csv("data/process/tables/full_nucleotide_error_summary.csv") %>% 
+  filter(total_seqs >= 1000) %>% 
+  gather("sub_type", "rate", contains("_rate")) %>% 
+  mutate(sub_type = str_replace(sub_type, "_rate", "")) %>% 
+  gather("prev_subtype", "seq_prevalence", contains("_seq_prev")) %>% 
+  mutate(prev_subtype = str_replace(prev_subtype, "_seq_prev", ""))
 
 
-# Generate graph of Mock DNA samples 
-thousand <- nucleotide_data[["1000"]] %>% 
-  mutate(sub_type = factor(sub_type, 
-                       levels = c("ac", "ag", "at", "ca", "cg", "ct", "ga", "gc", "gt", "ta", "tc", "tg"), 
-                       labels = c("A => C", "A => G", "A => T", "C => A", "C => G", "C => T", "G => A", "G => C", 
-                                  "G => T", "T => A", "T => C", "T => G"))) %>% 
-  ggplot(aes(sub_type, log10(rate + 0.000001), color = taq, group = taq)) + 
-  geom_point(size = 2, alpha = 0.2, position = position_dodge(width = 0.7), show.legend = F, color = "gray") + 
-  theme_bw() + 
-  stat_summary(aes(color = taq), fun.y = median, fun.ymin = median, fun.ymax = median, 
-               position = position_dodge(width = 0.7), geom = "crossbar", size = 0.5, width = 0.7) + 
-  facet_grid(cycles~.) + 
-  scale_color_manual(name = "Taq Used", 
[…]

### 2017-12-15 submission/manuscript.tex

@@
 each individual by cycle number to account for the biological variation
-between different people. Bray-Curtis distance matrices were generated
-using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
-The distance matrix data was analyzed using PERMANOVA with the vegan
-package (v 2.4.4) (Oksanen et al., 2017) and kruskal-wallis tests within
-R. For both error and chimera analysis, samples were tested using
+between different people. Bray-Curtis matrices were generated using
+mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The
+distance matrix data was analyzed using PERMANOVA with the vegan package
+(v 2.4.4) (Oksanen et al., 2017) and kruskal-wallis tests within R. For
+both error and chimera analysis, samples were tested using
 Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
@@
 sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curits distances were next assessed for
+Cycle dependent affects on Bray-Curits indicies were next assessed for
 the fecal samples looking at both overall cycle differences and within
@@
 any major differences between 5 cycle increments within each sample. We
-found that there was no detectable difference in distance when comparing
-to the previous 5 cycle increment (P-value \textgreater{} 0.05).
-However, Phusion at 1000 sub-sampling depth had a P-value = 0.02 before
-multiple comparison correction. It should be noted that at higher
-sub-sampling depths these differences in Bray-Curits distance disappear
+found that there was no detectable difference in Bray-Curits index when
+comparing to the previous 5 cycle increment (P-value \textgreater{}
+0.05). However, Phusion at 1000 sub-sampling depth had a P-value = 0.02
+before multiple comparison correction. It should be noted that at higher
+sub-sampling depths these differences in Bray-Curits indicies disappear
 for both differences in cycle number and within 5 cycle increments
@@
 
+Although we did not observe strong differences, based on cycle number,
+using the Bray-Curtis index the data suggests that there may be
+differences between 15 cycles and higher cycle numbers, such as 30, that
+are commonly used. What is most interesting is that there was no
+difference within individuals between corresponding 5 cycle increments
+(e.g.~15 to 20, 20 to 25, etc.). This finding, in conjunction with the
+PERMANOVA results, would suggest that differences based on Bray-Curtis
+incrementaly build up so that within 5 cycle increments the differences
+are not that large but when comparing larger cycle differences, as with
+the PERMANOVA, these differences start to become detectable. Although
+our data suggests that it is possible to overcome these types of cycle
+dependent differences with increased sub-sampling depth, for some DNA
+polymerase increased sub-sampling depth increases chimera prevalence and
+per base error rate. Conversely, it is possible that many of the
+increased number of OTUs, generated as cycle number increases, are not
+highly abundant and the Bray-Curtis index is able to successfully
+downweight these respective OTUs (Minchin, 1987).
+
 \newpage
@@
 DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-bc_index_Minchin1987}{}
+Minchin PR. 1987. An evaluation of the relative robustness of techniques
+for ecological ordination. \emph{Vegetatio} 69:89--107. DOI:
+\href{https://doi.org/10.1007/bf00038690}{10.1007/bf00038690}.
 

### 2017-12-15 code/run_pre_precluster_error_analysis.R
first commit

@@ -1,0 +1,214 @@
+### Create analysis of the mock error before pre.cluster command
+### Mock Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "dunn.test"))
+
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = ""))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+
+# Create ANOVA function to test differences between groups at each amplification cycle
+run_comparison <- function(i, col_of_int, taq_name, dataList){
+  # i is the sub sampled level of interest
+  # dataList is the combined meta data with OTU count data table list
+  
+  # Add a vector with all possible cycles analyzed
+  amp_cycles <- c("15x", "20x", "25x", "30x", "35x")
+  # Pull out data table of interest
+  tempData <- dataList
+  #tempData <- dataList[[i]]
+  # Runs the ANOVA across all the different amp cycles
+  tempTests <- sapply(amp_cycles, 
+                      function(x) run_kruskal(x, i, col_of_int, taq_name, tempData), simplify = F)
+  # removes all the cycles that do not have any values
+  tempTests <- tempTests[!sapply(tempTests, is.null)]
+  # converts the list to a data frame, run the BH correction, and reorganizes the table
+  tempTests <- tempTests %>% bind_rows() %>% 
+    mutate(bh = p.adjust(pvalue, method = "BH")) %>% 
+    select(statistic, parameter, pvalue, bh, cycle, sub_sample_level)
+  # Return the results to the global work environment
+  return(tempTests)
+  
+}
+
+
+# Function that runs the actual ANOVA with a select data set
+run_kruskal <- function(ac, subsample, comparator_var, taq_var, dataTable){
[…]

### 2017-12-15 code/run_seq_error_table_creations.R
add error analysis tables from precluster and chimera

@@ -8,9 +8,6 @@ source('code/functions.R')
 loadLibs(c("tidyverse", "stringr"))
 
 
-### Recognized an error in the error calculation and will need to go back and run through mothur to get correct errors ###
-
-
 ###########################################################################################################################
 ############################### List of functions to make analysis work ###################################################
 ###########################################################################################################################
@@ -221,42 +218,50 @@ get_nucleotide_summary_data <- function(dataList, depth){
 # Create vector with different subsample levels
 #sub_sample_level <- c(50, 100, 500, 1000, 5000, 10000)
 #taqs_used <- c("acc", "k", "phu", "pl", "q5")
-
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error")
 
 # Read in needed count files
-count_table <- read_data("data/process/", "mock_error", ".count_table", "")
+#count_table <- read_data("data/process/", "mock_error", ".count_table", "")
 
-#good_counts <- sapply(taqs_used, 
-#                      function(x) read_data("data/process/", x, ".count_table", "_data_only"), simplify = F)
+good_counts <- sapply(error_files, 
+                      function(x) read_data("data/process/", x, ".count_table", ""), simplify = F)
 
-error_summary <- read_data("data/process/", "mock_error", ".summary", "")
+#error_summary <- read_data("data/process/", "mock_error", ".summary", "")
 
-#good_error <- sapply(taqs_used, 
-#                      function(x) read_data("data/process/", x, ".summary", "_data_only"), simplify = F)
+good_error <- sapply(error_files, 
+                      function(x) read_data("data/process/", x, ".summary", ""), simplify = F)
 
 # Read in master meta data file
 metadata <- read_csv("data/process/tables/meta_data.csv")
 
 # Combine and make full data tables
-full_combined <- combine_data(count_table, error_summary, metadata) %>% 
-  rename(full_name = sample_name)
+full_combined <- sapply(error_files, 
+                        function(x) combine_data(good_counts[[x]], good_error[[x]], metadata) %>% 
+                          rename(full_name = sample_name), simplify = F)
+
+#full_combined <- combine_data(count_table, error_summary, metadata) %>% 
+#  rename(full_name = sample_name)
 
 # Combine and make slim data table
-slim_combined <- combine_data(count_table, error_summary, metadata, simple = T)
-  
[…]

### 2017-12-15 code/run_seq_error_table_creations.R
add completed nucleotide error from all analysis

@@ -291,7 +291,9 @@ sapply(c(1:length(good_summary_data)),
 ###########################################################################################################################
 
 # Generate the random sampling 
-full_Recombine <- get_samples("full_data", full_combined, unique(full_combined$full_name))
+full_Recombine <- sapply(error_files, 
+                         function(x) get_samples("full_data", full_combined[[x]], 
+                                                 unique(full_combined[[x]]$full_name)), simplify = F)
 
 #full_Recombine <- sapply(sub_sample_level, 
 #                        function(x) get_random_sample(x, full_combined, unique(full_combined$full_name)), simplify = F)
@@ -299,10 +301,11 @@ full_Recombine <- get_samples("full_data", full_combined, unique(full_combined$f
 #names(full_Recombine) <- sub_sample_level  
 
 # Add nucleotide changes checked column
-full_Recombine <- lapply(full_Recombine, function(y) run_nucleotide_error_check(y))
+full_Recombine <- lapply(full_Recombine, function(x) 
+  lapply(x, function(y) run_nucleotide_error_check(y)))
 
 # Generate Nucleotide Summary Data
-nucleotide_summary_data <- get_nucleotide_summary_data(full_Recombine, "full_data") %>% 
+nucleotide_summary_data <- sapply(error_files, function(x) get_nucleotide_summary_data(full_Recombine[[x]], "full_data") %>% 
   mutate(sample_name.y = ifelse(grepl("Zmock_A_111716", full_name) == T, invisible("A"), 
                               ifelse(grepl("Zmock_B_", full_name) == T, invisible("B"), 
                                      ifelse(grepl("Zmock_C_", full_name) == T, invisible("C"), 
@@ -323,7 +326,7 @@ nucleotide_summary_data <- get_nucleotide_summary_data(full_Recombine, "full_dat
             gc_seq_prev = sum(gc)/length(query), ca_seq_prev = sum(ca)/length(query), 
             ct_seq_prev = sum(ct)/length(query), cg_seq_prev = sum(cg)/length(query)) %>% 
   rename(sample_name = sample_name.y) %>% 
-  filter(!is.na(sample_name))
+  filter(!is.na(sample_name)), simplify = F)
 
 
 #nucleotide_summary_data <- sapply(sub_sample_level, 
@@ -332,12 +335,12 @@ nucleotide_summary_data <- get_nucleotide_summary_data(full_Recombine, "full_dat
 
 
 # Write out summarized data tables for graphing
[…]

### 2017-12-15 code/run_error_analysis.R
complete workflow to add all error to workflow

@@ -143,10 +143,11 @@ get_dunn_test <- function(subsample, cycle_num, comparator_var, taq_var, dataTab
 
 # Read in subsample.shared files
 # sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
-# 
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error") 
+
 # # Read in the count data
-error_data <- read_csv("data/process/tables/full_error_summary.csv") %>% 
-  filter(total_seqs >= 1000)
+error_data <- sapply(error_files, function(x) read_csv(paste("data/process/tables/", x, "_summary.csv", sep = "")) %>% 
+  filter(total_seqs >= 1000), simplify = F)
 
 # error_data <- sapply(sub_sample_level, 
 #                      function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
@@ -154,7 +155,8 @@ error_data <- read_csv("data/process/tables/full_error_summary.csv") %>%
 
 
 # Run the kruskal comparisons between amp cycle across subsamplings error per base
-krsukal_tests_error <- run_comparison("full_data", "mean_error", "taq", error_data)
+krsukal_tests_error <- sapply(error_files, 
+                              function(x) run_comparison("full_data", "mean_error", "taq", error_data[[x]]), simplify = F)
 
 #  kruskal_tests_error <- sapply(sub_sample_level, 
 #                        function(x) run_comparison(x, "mean_error", "taq", error_data), simplify = F)
@@ -162,7 +164,9 @@ krsukal_tests_error <- run_comparison("full_data", "mean_error", "taq", error_da
 # combined_kruskal_table_error <- kruskal_tests_error %>% bind_rows()
 
 # Run the kruskal comparisons between amp cycle across subsamplings number of seqs with error
-kruskal_tests_seq_error_count <- run_comparison("full_data", "seq_error_prevalence", "taq", error_data)
+kruskal_tests_seq_error_count <- sapply(error_files, 
+                                        function(x) run_comparison("full_data", "seq_error_prevalence", "taq", 
+                                                                   error_data[[x]]), simplify = F)
 
 # kruskal_tests_seq_error_count <- sapply(sub_sample_level, 
 #                             function(x) run_comparison(x, "seq_error_prevalence", "taq", error_data), simplify = F)
[…]

### 2017-12-15 code/make_mock_error_graphs.R
incorporate full error analysis

@@ -32,11 +32,15 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################
 
-# Read in the count data
-error_data <- read_csv("data/process/tables/full_error_summary.csv")
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error") 
+
+# # Read in the count data
+error_data <- sapply(error_files, function(x) read_csv(paste("data/process/tables/", x, "_summary.csv", sep = "")) %>% 
+                       filter(total_seqs >= 1000), simplify = F)
+
 
 # Generate graph of Mock DNA samples 
-full_graph <- error_data %>% 
+before_precluster <- error_data[["mock_precluster_error"]] %>% 
   filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
@@ -45,125 +49,89 @@ full_graph <- error_data %>%
   summarise(group_median = median(mean_error), group_iqr25 = quantile(mean_error)["25%"], 
             group_iqr75 = quantile(mean_error)["75%"]) %>% 
   ggplot(aes(cycles, group_median, color = taq, group = taq)) + 
-  geom_line() + 
+  geom_line(show.legend = T) + 
   geom_errorbar(aes(ymin=group_iqr25, 
-                    ymax=group_iqr75), 
+                    ymax=group_iqr75), show.legend = T,  
                 width = 0.1, size = 0.5, alpha = 0.4) + 
-  geom_point(size = 2, alpha = 0.7) + 
+  geom_point(size = 2, alpha = 0.7, show.legend = T) + 
   theme_bw() + 
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Amplification Cycles", y = "Median Average Error Rate Per Base") + 
+  ggtitle("A") + 
+  annotate("text", label = paste("Before Pre-Cluster Step"), x = 2.5, y = 0.031, size = 2.5) + 
[…]

### 2017-12-15 code/make_mock_seq_error_count_graphs.R
update with all error

@@ -30,16 +30,14 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ############################### Run actual analysis programs  #############################################################
 ###########################################################################################################################
 
-# Read in the count data
-error_data <- read_csv("data/process/tables/full_error_summary.csv")
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error") 
 
-#error_data <- sapply(sub_sample_level, 
-#                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
-#                     simplify = F)
-
+# # Read in the count data
+error_data <- sapply(error_files, function(x) read_csv(paste("data/process/tables/", x, "_summary.csv", sep = "")) %>% 
+                       filter(total_seqs >= 1000), simplify = F)
 
 # Generate graph of Mock DNA samples 
-full_graph <- error_data %>% 
+before_precluster <- error_data[["mock_precluster_error"]] %>% 
   filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
@@ -57,123 +55,84 @@ full_graph <- error_data %>%
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Amplification Cycles", y = "Percent of Sequences with Error") + 
-  coord_cartesian(ylim = c(0, 0.10)) + 
+  coord_cartesian(ylim = c(0, 0.40)) + 
   scale_y_continuous(labels = scales::percent) + 
+  ggtitle("A") + 
+  annotate("text", label = paste("Before Pre-Cluster Step"), x = 2.5, y = 0.41, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10), 
-        legend.position = c(0.10, 0.83), 
+        legend.position = c(0.20, 0.83), 
         legend.title = element_blank(), 
[…]

### 2017-12-15 code/make_mock_error_graphs.R
change sizing

@@ -64,7 +64,7 @@ before_precluster <- error_data[["mock_precluster_error"]] %>%
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10), 
-        legend.position = c(0.40, 0.80), 
+        legend.position = c(0.20, 0.80), 
         legend.title = element_blank(), 
         legend.key = element_blank(), 
         legend.background = element_rect(color = "black"))
@@ -134,4 +134,4 @@ full_pipeline <- error_data[["mock_error"]] %>%
  
  combined_graph <- grid.arrange(before_precluster, before_chimera_vsearch, full_pipeline, ncol = 3)
 
-ggsave("results/figures/Figure3.pdf", combined_graph, width = 7, height = 7, dpi = 300)
+ggsave("results/figures/Figure3.pdf", combined_graph, width = 11, height = 7, dpi = 300)

### 2017-12-15 code/make_mock_chimera_count_graphs.R
add full analysis for chimera

@@ -31,19 +31,15 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ###########################################################################################################################
 
 
-# Vector of sub samples used
-#sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error") 
 
-# Read in the count data
-error_data <- read_csv("data/process/tables/full_error_summary.csv")
-
-#error_data <- sapply(sub_sample_level, 
-#                     function(x) read_data("data/process/tables/error_", "", "_summary.csv", x), 
-#                     simplify = F)
+# # Read in the count data
+error_data <- sapply(error_files, function(x) read_csv(paste("data/process/tables/", x, "_summary.csv", sep = "")) %>% 
+                       filter(total_seqs >= 1000), simplify = F)
 
 
 # Generate graph of Mock DNA samples 
-full_graph <- error_data %>% 
+before_precluster <- error_data[["mock_precluster_error"]]  %>% 
   filter(total_seqs >= 1000) %>% 
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
@@ -52,17 +48,83 @@ full_graph <- error_data %>%
   summarise(group_chimera = median(chimera_prevalence), group_iqr25 = quantile(chimera_prevalence)["25%"], 
             group_iqr75 = quantile(chimera_prevalence)["75%"]) %>% 
   ggplot(aes(cycles, group_chimera, color = taq, group = taq)) + 
-  geom_line() + 
+  geom_line(show.legend = T) + 
   geom_errorbar(aes(ymin=group_iqr25, 
-                    ymax=group_iqr75), 
+                    ymax=group_iqr75), show.legend = T,  
                 width = 0.1, size = 0.5, alpha = 0.4) + 
-  geom_point(size = 2, alpha = 0.7) + 
+  geom_point(size = 2, alpha = 0.7, show.legend = T) + 
   theme_bw() + 
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
[…]

### 2017-12-15 code/run_nmds_fecal_data.R
add 5 cycle distance comparison to write out to work flow

@@ -274,5 +274,7 @@ permanova_results <- sapply(sub_sample_level,
 write_csv(kruskal_test_data, "data/process/tables/bray_sim_to_prev_cycle_kruskal_results.csv")
 write_csv(permanova_results, "data/process/tables/bray_permanova_by_taq_results.csv")
   
-
+sapply(c(1:length(finalized_tables)), 
+       function(x) write_csv(finalized_tables[[x]], 
+                             paste("data/process/tables/bray_5_cycle_dist_", sub_sample_level[x], "_data.csv", sep = "")))
   

### 2017-12-15 code/run_nmds_fecal_data.R
add spacing

@@ -277,4 +277,6 @@ write_csv(permanova_results, "data/process/tables/bray_permanova_by_taq_results.
 sapply(c(1:length(finalized_tables)), 
        function(x) write_csv(finalized_tables[[x]], 
                              paste("data/process/tables/bray_5_cycle_dist_", sub_sample_level[x], "_data.csv", sep = "")))
-  
+
+
+

### 2017-12-15 code/make_bray_distance_graphs.R
first commit

@@ -1,0 +1,50 @@
+### Create graphs of the distances compared in 5 cycle increments (Bray)
+### Mock Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "viridis", "gridExtra"))
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = ""))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+# Vector of sub samples used
+sub_sample_level <- c("1000", "5000", "10000", "15000", "20000")
+
+dist_data <- sapply(sub_sample_level, 
+                    function(x) read_csv(paste("data/process/tables/bray_5_cycle_dist_", 
+                                               x, "_data.csv", sep = "")), simplify = F)
+
+
+
+
+
+
+
+
+
+
+
+

### 2018-01-12 code/make_correlation_graphs.R
update to make pretty

@@ -67,15 +67,17 @@ error_data <- sapply(error_files, function(x) read_csv(paste("data/process/table
                        filter(total_seqs >= 1000), simplify = F)
 
 # Read in the count data
-#numOTU_data <- sapply(sub_sample_level, 
- #                     function(x) read_data("data/process/tables/", "mock_sub_sample_", "_count_table.csv", x), 
-  #                    simplify = F)
-
-
-
-
 test <- read_tsv("data/process/all_amp.shared") %>% 
   select(-label, -numOtus) %>% 
+  filter(str_detect(Group, "FS") == FALSE)
+
+test2 <- select(test, contains("Otu"))
+test <- test2[, colSums(test2) > 0] %>% 
+  mutate(Group = test$Group)
+
+rm(test2)
+
+test <- test %>%   
   mutate_at(vars(contains("Otu")), function(x) ifelse(x > 0, invisible(1), invisible(0))) %>% 
   mutate(total_otus = select(., -Group) %>% rowSums(.)) %>% 
   select(Group, total_otus)
@@ -118,10 +120,10 @@ mock_pc <- combined_list[["mock_precluster_error"]] %>%
   scale_color_manual(name = "Cycle Number", 
                      values = c("#0000FF", "#00C957", "#CD8500", "#FF1493")) + 
   labs(x = "Chimera Sequence Prevalence (%)", y = "Number of OTUs") + 
-  geom_text(data = df_m_pc_error, aes(x = 1.5, y = 350, label = eq), 
+  geom_text(data = df_m_pc_error, aes(x = 32, y = 50, label = eq), 
             color = 'black',  parse = TRUE) + 
-  ggtitle("A") + coord_cartesian(ylim = c(0, 400)) + 
-  annotate("text", label = paste("Before Precluster Step"), x = 22, y = 100, size = 2.5) + 
+  ggtitle("A") + coord_cartesian(ylim = c(0, 500)) + scale_x_continuous(limits = c(0, 40)) + 
+  annotate("text", label = paste("Before Precluster Step"), x = 32, y = 110, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -152,10 +154,10 @@ mock_ch <- combined_list[["mock_chimera_error"]] %>%
   scale_color_manual(name = "Cycle Number", 
[…]

### 2018-01-12 code/run_chimera_removed_kept.R
not sure what rate of chimeras will tell me -archive for now

@@ -40,8 +40,9 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 
 # Read in subsample.shared files
 sub_sample_level <- c("1000", "5000", "10000")
+error_files <- c("mock_precluster_error", "mock_chimera_error", "mock_error") 
 
-mock_count_tables <- sapply(sub_sample_level, 
+mock_count_tables <- sapply(error_files, 
                           function(x) read_data("data/process/tables/", "mock_sub_sample_", "_count_table.csv", x), 
                           simplify = F)
 
@@ -64,4 +65,5 @@ rearranged_combined_table <- combined_table %>% gather(sample_name, seq_count, 3
   mutate(chimera_positive = ifelse(chimera == "N", invisible(0), invisible(1))) %>% 
   group_by(sample_name, chimera) %>% 
   summarise(total_seqs = sum(seq_count))
+
 

### 2018-01-12 code/make_nucleotide_graphs.R
remove sub-sample title

@@ -68,7 +68,6 @@ full_graph <- nucleotide_data %>%
                       values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
    labs(x = "", y = expression(Log["10"]~Substitution~Rate)) + 
    coord_cartesian(ylim = c(0, -6)) + 
-   ggtitle("Sub-Sampled to 1000 Sequences") +  
    theme(plot.title = element_text(face="bold", hjust = -0.07, size = 10), 
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 

### 2018-01-12 submission/manuscript.Rmd
add figure legends and update discussion and results

@@
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
+Overall, there was very little difference between each respective 5-cycle increment (e.g.
+15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3].
+Two exceptions to the low differences between 5-cycle increments can be found.
+The first that there seems to be large differences for fecal samples between 20x vs.
+25x that is robust against sub-sampling depth [Figure 3A-B].
+Second, for the mocks, where data is available, there appears to be a similar large difference detected between 20x and 25x [Figure 3D].
+Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community [Figure 3].
+
 Using PERMANOVA to test whether there were any differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
@@
 Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest values were for the Kappa HiFI DNA polymerase [Figure 3A].
-Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 3B-C].
-There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 3B-C and Table S6].
+Generally, the highest values were for the Kappa HiFI DNA polymerase [Figure 4A].
+Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 4B-C].
+There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].
 
@@
 Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi DNA polymerase across cycle number and sub-sampling depth [Table S10].
-Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase [Figure 4].
+Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had been used [Figure 5].
 
-For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Platinum and Phusion HiFi DNA Polymerase [Figure 5].
-In general, the R^2^ value between the number of OTUs and chimeric sequences became stronger as sub-sampling depth increased [Figure 5].
-Taken together this data suggests that a strong correlation exists between the number of OTUs and chimera sequence prevalence.
-Interestingly, Accuprime^TM^ not only had one of the lowest prevalence of chimeric sequences but also, consistently, had the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 5].
+For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
+In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6].
+Taken together this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.
+Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].
 
@@
 
-Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30, that are commonly used.
+Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30x, that are commonly used.
 What is most interesting is that there was no difference within individuals between corresponding 5 cycle increments (e.g.
 15 to 20, 20 to 25, etc.).
-This finding, in conjunction with the PERMANOVA results, would suggest that differences based on Bray-Curtis incrementaly build up so that within 5 cycle increments the differences are not that large but when comparing larger cycle differences, as with the PERMANOVA, these differences start to become detectable.
-Although our data suggests that it is possible to overcome these types of cycle dependent differences with increased sub-sampling depth, for some DNA polymerase increased sub-sampling depth increases chimera prevalence and per base error rate.
-Conversely, it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant and the Bray-Curtis index is able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
+Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different [Figure 3].
+This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once at least 25 cylces are reached.
+Increasing the sub-sampling depth, for some DNA polymerase may reduce some of these observed community differences at lower cycle numbers.
+
+Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5].
+The chimera prevalence was very strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations.
+However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase.
+Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant that the Bray-Curtis index is able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
+So, choice of downstream diversity metric could be an important consideration to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase such as Platinum.
+
+Our observations suggest that there are clear HiFi dependent differences in both per base error rate and chimeras that cannot be removed using bioinformatic approaches [Figure 4 & 5].
+Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A polymerase [@polymerase_Ishino_2014].
+In fact, from the material safety data sheet (MSDS), it is not clear what the difference between the two different mixes really is.
+Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pycrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate since all samples were also sequenced at the same time as well as amplified using the same machine.
 
@@
 
+Our findings show that measures that rely on number of OTUs will be specific for a particular study and may not be easily generalized to other studies that may be studying a similar area.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed as well as potentially influence diversity based metrics that do not down weight rare observations.
+
 \newpage
@@
 
-**Table 1: **
+**Figure 1: Normalized Fecal Number of OTUs.** A) Sub-sampled to 1000 reads.
+B) Sub-sampled to 5000 reads.
+C) Sub-sampled to 10000 reads.
+D) Sub-sampled to 15000 reads.
+The dotted line represents no change from the mean number of OTUs within that specific individual.
 
-\footnotesize
+**Figure 2:  Mock Sample Variability in Number of OTUs based on HiFi DNA Polymerase.** A) Sub-sampled to 1000 reads.
+B) Sub-sampled to 5000 reads.
+C) Sub-sampled to 10000 reads.
+The dotted line represents the number of OTUs genereated when the mock reference sequences are run through the pipeline.
 
-```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
+**Figure 3: Five Cycle Interval Community Differences.** A) Fecal samples sub-sampled to 1000 reads.
+B) Fecal samples sub-sampled to 5000 reads.
+C) Fecal samples sub-sampled to 10000 reads.
+D) Mock samples sub-sampled to 1000 reads.
+E) Mock samples sub-sampled to 5000 reads.
+F) Mock samples sub-sampled to 10000 reads.
+The solid black lines represent the median Bray-Curtis index difference within sample for each 5 cycle interval.
 
-```
+**Figure 4: HiFi DNA Polymerase Per Base Error Rate in Mock Samples.** A) Error rate before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
+B) Error rate before the removal of chimeras with VSEARCH.
+C) Full pipeline.
+The error bars represent the 75% interquartile range of the median.
 
-\normalsize
-\newpage
+**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Chimera sequence percentage before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
+B) Chimera sequence percentage before the removal of chimeras with VSEARCH.
+C) Full pipeline.
+The error bars represent the 75% interquartile range of the median.
 
-**Table 2: **
-
-\footnotesize
-
-```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-
-\normalsize
-\newpage
-
-**Figure 1: .**
-
-**Figure 2: .**
-
-**Figure 3: .**
-
-**Figure 4: .**
-
-**Figure 5: .**
-
-**Figure 6: .**
-
-**Figure 7: .**
+**Figure 6: The Correlation between Number of OTUs and Chimeras.** A) Correlation before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
+B) Correlation before the removal of chimeras with VSEARCH.
+C) Correlation with full pipeline.
 
@@
 
-**Figure S1: .**
+**Figure S1: HiFi DNA Polymerase Sequence Error Prevalence in Mock Samples.** A) Sequence error prevalence before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
+B) Sequence error prevalence before the removal of chimeras with VSEARCH.
+C) Full pipeline.
+The error bars represent the 75% interquartile range of the median.
 
-**Figure S2: .**
-
-**Figure S3: .**
-
-**Figure S4: .**
-
-**Figure S5: .**
-
-**Figure S6: .**
-
-**Figure S7: .**
-
-\newpage
+**Figure S2: HiFi DNA Polymerase Nucleotide Subsitutions in Mock Samples.**

### 2018-01-12 submission/manuscript.tex
add figure legends and update discussion and results

@@
 \textbf{\emph{Statistical Analysis:}} All analysis was done with the R
-(v 3.4.2) software package (R Core Team, 2017). Data tranformation and
+(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
 graphing was completed using the tidyverse package (v 1.1.1) and colors
@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} Using PERMANOVA to test whether there were any differences
-within HiFi DNA polymerase groups based on cycle number, only Phusion
-had cycle dependent differences at 1000 and 5000 sub-sampling depth
-(P-value = 0.03 and 0.01, respectively). Interestingly, Phusion was one
-of only two DNA polymerase that managed to have samples for the 1000
-sub-sampling depth at 15 cycles. Next, we assessed whether there were
-any major differences between 5 cycle increments within each sample. We
-found that there was no detectable difference in Bray-Curits index when
-comparing to the previous 5 cycle increment (P-value \textgreater{}
-0.05). However, Phusion at 1000 sub-sampling depth had a P-value = 0.02
-before multiple comparison correction. It should be noted that at higher
+Number:}} Overall, there was very little difference between each
+respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
+samples and this was consistent across the different sub-samplings used
+{[}Figure 3{]}. Two exceptions to the low differences between 5-cycle
+increments can be found. The first that there seems to be large
+differences for fecal samples between 20x vs.~25x that is robust against
+sub-sampling depth {[}Figure 3A-B{]}. Second, for the mocks, where data
+is available, there appears to be a similar large difference detected
+between 20x and 25x {[}Figure 3D{]}. Regardless, by the time PCR cycles
+reach 25x there does not seem to be large differences in the community
+{[}Figure 3{]}.
+
+Using PERMANOVA to test whether there were any differences within HiFi
+DNA polymerase groups based on cycle number, only Phusion had cycle
+dependent differences at 1000 and 5000 sub-sampling depth (P-value =
+0.03 and 0.01, respectively). Interestingly, Phusion was one of only two
+DNA polymerase that managed to have samples for the 1000 sub-sampling
+depth at 15 cycles. Next, we assessed whether there were any major
+differences between 5 cycle increments within each sample. We found that
+there was no detectable difference in Bray-Curits index when comparing
+to the previous 5 cycle increment (P-value \textgreater{} 0.05).
+However, Phusion at 1000 sub-sampling depth had a P-value = 0.02 before
+multiple comparison correction. It should be noted that at higher
 sub-sampling depths these differences in Bray-Curits indicies disappear
@@
 Generally, the highest values were for the Kappa HiFI DNA polymerase
-{[}Figure 3A{]}. Sub-sampling depth seems to have little effect on this
+{[}Figure 4A{]}. Sub-sampling depth seems to have little effect on this
 rate with both 5000 and 10000 sub-sampled sequences showing similar
-results {[}Figure 3B-C{]}. There were small differences between the
+results {[}Figure 4B-C{]}. There were small differences between the
 various HiFi Taq at lower cycle number but larger differences at higher
 cycle number with Platinum having large differences between all other
-HiFi DNA polymerase {[}Figure 3B-C and Table S6{]}.
+HiFi DNA polymerase {[}Figure 4B-C and Table S6{]}.
 
@@
 Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
-the HiFi DNA Polymerase {[}Figure 4{]}.
+the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had
+been used {[}Figure 5{]}.
 
@@
 sequences and number of OTUs, with this correlation being strongest for
-Platinum and Phusion HiFi DNA Polymerase {[}Figure 5{]}. In general, the
-R\textsuperscript{2} value between the number of OTUs and chimeric
-sequences became stronger as sub-sampling depth increased {[}Figure
-5{]}. Taken together this data suggests that a strong correlation exists
-between the number of OTUs and chimera sequence prevalence.
-Interestingly, Accuprime\textsuperscript{TM} not only had one of the
-lowest prevalence of chimeric sequences but also, consistently, had the
+Accuprime, Platinum and Phusion HiFi DNA Polymerase {[}Figure 6{]}. In
+general, the R\textsuperscript{2} value between the number of OTUs and
+chimeric sequences did not change from the use of pre.cluster and
+VSEARCH {[}Figure 6{]}. Taken together this data suggests that a strong
+correlation exists between the number of OTUs and the prevalence of
+chimeric sequences. Kappa had the highest per base error rate and the
 lowest correlations between the number of OTUs and chimeric sequences
-across sub-sampling depth {[}Figure 5{]}.
+across sub-sampling depth {[}Figure 4 \& 6{]}.
 
@@
 using the Bray-Curtis index the data suggests that there may be
-differences between 15 cycles and higher cycle numbers, such as 30, that
-are commonly used. What is most interesting is that there was no
+differences between 15 cycles and higher cycle numbers, such as 30x,
+that are commonly used. What is most interesting is that there was no
 difference within individuals between corresponding 5 cycle increments
-(e.g.~15 to 20, 20 to 25, etc.). This finding, in conjunction with the
-PERMANOVA results, would suggest that differences based on Bray-Curtis
-incrementaly build up so that within 5 cycle increments the differences
-are not that large but when comparing larger cycle differences, as with
-the PERMANOVA, these differences start to become detectable. Although
-our data suggests that it is possible to overcome these types of cycle
-dependent differences with increased sub-sampling depth, for some DNA
-polymerase increased sub-sampling depth increases chimera prevalence and
-per base error rate. Conversely, it is possible that many of the
-increased number of OTUs, generated as cycle number increases, are not
-highly abundant and the Bray-Curtis index is able to successfully
-downweight these respective OTUs (Minchin, 1987).
+(e.g.~15 to 20, 20 to 25, etc.). Conversely, this may be due to low
+power and on observation there does seem to be a trend that 20x and 25x
+communities are very different {[}Figure 3{]}. This finding, in
+conjunction with the PERMANOVA results, suggest that cycle number can
+change bacterial community calculations but that these differences are
+minimal once at least 25 cylces are reached. Increasing the sub-sampling
+depth, for some DNA polymerase may reduce some of these observed
+community differences at lower cycle numbers.
+
+Increasing the cycle number also exacerbated chimera prevalence
+differences between the different HiFi DNA polymerases {[}Figure 5{]}.
+The chimera prevalence was very strongly correlated with the number of
+OTUs and this value is relied upon heavily for many different downstream
+community metric calculations. However, Bray-Curtis analysis with
+PERMANOVA showed few differences based on DNA polymerase. Since it is
+possible that many of the increased number of OTUs, generated as cycle
+number increases, are not highly abundant that the Bray-Curtis index is
+able to successfully downweight these respective OTUs (Minchin, 1987).
+So, choice of downstream diversity metric could be an important
+consideration to mitigate these observed changes due to high chimera
+prevalence in HiFi DNA polymerase such as Platinum.
+
+Our observations suggest that there are clear HiFi dependent differences
+in both per base error rate and chimeras that cannot be removed using
+bioinformatic approaches {[}Figure 4 \& 5{]}. Although it may be a
+natural assumption that the variation may be due to the DNA polymerase
+family, the highest chimera rate, from Platinum, was a family A
+polymerase while the lowest, from Accuprime, was also an A polymerase
+(Ishino \& Ishino, 2014). In fact, from the material safety data sheet
+(MSDS), it is not clear what the difference between the two different
+mixes really is. Both Accuprime and Platinum contain a recombinant
+\emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D polymerase and
+a platinum \emph{Taq} antibody. It is possible that differences in how
+the recombinant \emph{Taq} was generated could be the main reason for
+the differences in chimera rate since all samples were also sequenced at
+the same time as well as amplified using the same machine.
 
@@
 \subsection{Conclusion}\label{conclusion}
+
+Our findings show that measures that rely on number of OTUs will be
+specific for a particular study and may not be easily generalized to
+other studies that may be studying a similar area. Care should be taken
+when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since
+intrinsic differences can change the number of OTUs observed as well as
+potentially influence diversity based metrics that do not down weight
+rare observations.
 
@@
 DOI: \href{https://doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
+
+\hypertarget{ref-polymerase_Ishino_2014}{}
+Ishino S., Ishino Y. 2014. DNA polymerases as useful reagents for
+biotechnology â the history of developmental research in the field.
+\emph{Frontiers in Microbiology} 5. DOI:
+\href{https://doi.org/10.3389/fmicb.2014.00465}{10.3389/fmicb.2014.00465}.
 
@@
 
-\textbf{Table 1: }
+\textbf{Figure 1: Normalized Fecal Number of OTUs.} A) Sub-sampled to
+1000 reads. B) Sub-sampled to 5000 reads. C) Sub-sampled to 10000 reads.
+D) Sub-sampled to 15000 reads. The dotted line represents no change from
+the mean number of OTUs within that specific individual.
 
-\footnotesize
+\textbf{Figure 2: Mock Sample Variability in Number of OTUs based on
+HiFi DNA Polymerase.} A) Sub-sampled to 1000 reads. B) Sub-sampled to
+5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents
+the number of OTUs genereated when the mock reference sequences are run
+through the pipeline.
 
-\normalsize
-\newpage
+\textbf{Figure 3: Five Cycle Interval Community Differences.} A) Fecal
+samples sub-sampled to 1000 reads. B) Fecal samples sub-sampled to 5000
+reads. C) Fecal samples sub-sampled to 10000 reads. D) Mock samples
+sub-sampled to 1000 reads. E) Mock samples sub-sampled to 5000 reads. F)
+Mock samples sub-sampled to 10000 reads. The solid black lines represent
+the median Bray-Curtis index difference within sample for each 5 cycle
+interval.
 
-\textbf{Table 2: }
+\textbf{Figure 4: HiFi DNA Polymerase Per Base Error Rate in Mock
+Samples.} A) Error rate before the merger of sequences with pre.cluster
+and the removal of chimeras with VSEARCH. B) Error rate before the
+removal of chimeras with VSEARCH. C) Full pipeline. The error bars
+represent the 75\% interquartile range of the median.
 
-\footnotesize
+\textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
+Samples.} A) Chimera sequence percentage before the merger of sequences
+with pre.cluster and the removal of chimeras with VSEARCH. B) Chimera
+sequence percentage before the removal of chimeras with VSEARCH. C) Full
+pipeline. The error bars represent the 75\% interquartile range of the
+median.
 
-\normalsize
-\newpage
-
-\textbf{Figure 1: .}
-
-\textbf{Figure 2: .}
-
-\textbf{Figure 3: .}
-
-\textbf{Figure 4: .}
-
-\textbf{Figure 5: .}
-
-\textbf{Figure 6: .}
-
-\textbf{Figure 7: .}
+\textbf{Figure 6: The Correlation between Number of OTUs and Chimeras.}
+A) Correlation before the merger of sequences with pre.cluster and the
+removal of chimeras with VSEARCH. B) Correlation before the removal of
+chimeras with VSEARCH. C) Correlation with full pipeline.
 
@@
 
-\textbf{Figure S1: .}
+\textbf{Figure S1: HiFi DNA Polymerase Sequence Error Prevalence in Mock
+Samples.} A) Sequence error prevalence before the merger of sequences
+with pre.cluster and the removal of chimeras with VSEARCH. B) Sequence
+error prevalence before the removal of chimeras with VSEARCH. C) Full
+pipeline. The error bars represent the 75\% interquartile range of the
+median.
 
-\textbf{Figure S2: .}
-
-\textbf{Figure S3: .}
-
-\textbf{Figure S4: .}
-
-\textbf{Figure S5: .}
-
-\textbf{Figure S6: .}
-
-\textbf{Figure S7: .}
-
-\newpage
+\textbf{Figure S2: HiFi DNA Polymerase Nucleotide Subsitutions in Mock
+Samples.}
 

### 2018-01-12 submission/manuscript.Rmd
add abstract

@@
 
-**Background.**
+**Background.** Increasing research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys.
+These components include, but are not necessarily limited to, preservation media, extraction kit, bead beating time, and primers.
+Both cycle number and Hifidelity (HiFi) DNA polymerase are sometimes overlooked during this conversation of bias considerations.
+Here we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.
 
-**Methods.**
+**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+Mock communities (technical replicates n = 4) consisting of the whole genome of 8 different bacteria were also amplified using the same approach.
+The number of OTUs (Operational Taxonomic Units) was first examined for both fecal samples and mock communites.
+Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence, and the chimera prevalence correlation with number of OTUs was assessed for either cycle number or HiFi DNA polymerase dependent differences.
 
-**Results.**
-
-**Conclusions.**
+**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH.
+Further, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample that was also not changed by chimera removal with VSEARCH.
+**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras.
+So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUS that are observed.
 
@@
 
-***Chimeric Sequences Corelate with OTUs and are HiFi Taq Dependent:***
+***Chimeric Sequences Correlate with OTUs and are HiFi Taq Dependent:***
 After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.

### 2018-01-12 submission/manuscript.tex
add abstract

@@
 
-\textbf{Background.}
+\textbf{Background.} Increasing research has found that various
+methodological steps can have an impact on the observed microbial
+community when using 16S rRNA gene surveys. These components include,
+but are not necessarily limited to, preservation media, extraction kit,
+bead beating time, and primers. Both cycle number and Hifidelity (HiFi)
+DNA polymerase are sometimes overlooked during this conversation of bias
+considerations. Here we critically examine both cycle number and HiFi
+DNA polymerase for biases that may influence downstream diversity
+measures of 16S rRNA gene surveys.
 
-\textbf{Methods.}
+\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
+single PowerMag DNA extraction kit with a 10 minute bead beating step
+and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa,
+Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
+(technical replicates n = 4) consisting of the whole genome of 8
+different bacteria were also amplified using the same approach. The
+number of OTUs (Operational Taxonomic Units) was first examined for both
+fecal samples and mock communites. Next, Bray-Curtis index, the error
+rate, sequence error prevalence, and chimera prevalence, and the chimera
+prevalence correlation with number of OTUs was assessed for either cycle
+number or HiFi DNA polymerase dependent differences.
 
-\textbf{Results.}
-
-\textbf{Conclusions.}
+\textbf{Results.} At 35 cycles there were significant differences
+between HiFi DNA polymerase for fecal samples (P-value \textless{}
+0.0001). These HiFi dependent differences in the number of OTUs could be
+identified as early as 20 cycles in the mock communities (P-value =
+0.002). Chimera prevalence varied by HiFi DNA polymerase and these
+differences were still observed after chimera removal using VSEARCH.
+Further, the chimera prevalence had a strong positive correlation with
+the number of OTUs observed in a sample that was also not changed by
+chimera removal with VSEARCH.\\
+\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
+number of OTUs, common diversity metrics that incorporate this value
+could give artificially inflated numbers due to higher undetected
+chimeras. So, when designing 16S rRNA gene survey studies it is
+important to consider both the cycle number and the type of HiFi DNA
+polymerase that will be used since it can increase or decrease the
+number of OTUS that are observed.
 
@@
 
-\textbf{\emph{Chimeric Sequences Corelate with OTUs and are HiFi Taq
+\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi Taq
 Dependent:}} After chimera removal using VSEARCH and removal of

### 2018-01-15 submission/manuscript.Rmd
add first 3 paragraphs of the intro

@@
 ## Introduction
+
+Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [].
+In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [].
+Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016].
+Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequece library preparation.
+Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.
+
+The most commonly studied biases are by far the skews introduced to the data via preservation media used, storage conditions, and DNA extraction kits chosen.
+The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g.
+Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/).
+Within the literature DNA extraction kits and storage condition chosen have consistently been shown to add bias to downstream analysis [].
+However, the current literature on preservation media has been more mixed, with some studies showing biases while others do not [].
+Although these are important sources of bias they are not the only sources that should be critically examined.
+The type of DNA polymerase chosen could have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
+
+A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
+Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
+Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
+There is some reason to think that they may since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
+
+  - Only a few have focused on DNA polymerase (highlight Knights paper)
+    - HiFi DNA polymerase is generally seen as better 
+  - Are all HiFi DNA polymerase created equal
+    - add that there are many families of DNA polymerase
+  - Brief overview of what was done
 

### 2018-01-15 submission/manuscript.Rmd
complete first pass of full intro addition

@@
 There is some reason to think that they may since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
+In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 
-  - Only a few have focused on DNA polymerase (highlight Knights paper)
-    - HiFi DNA polymerase is generally seen as better 
-  - Are all HiFi DNA polymerase created equal
-    - add that there are many families of DNA polymerase
-  - Brief overview of what was done
+We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase.
+First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples.
+Next, since there were differences based of HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected.
+After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence.
+Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.
 

### 2018-01-15 submission/manuscript.tex
complete first pass of full intro addition

@@
 \subsection{Introduction}\label{introduction}
+
+Over recent years there has been an increasing focus on standardizing
+methodological approaches in microbiota research {[}{]}. In particular,
+a lot of the attention of this standardization push has been invested in
+investigating ways that 16S rRNA gene surveys can be made more
+replicatable {[}{]}. Although 16S rRNA gene sequencing has been much
+maligned for introduced bias, many of these same considerations also
+affect metagenomic sequencing (Nayfach \& Pollard, 2016). Between the
+two approaches similar bias considerations include, but are not limited
+to, preservation media, storage conditions, DNA extraction kit, PCR, and
+sequece library preparation. Thus what is identified as a source of bias
+in 16S rRNA gene sequencing for these overlapping considerations will
+also influence metagenomic sequencing results.
+
+The most commonly studied biases are by far the skews introduced to the
+data via preservation media used, storage conditions, and DNA extraction
+kits chosen. The study of these specific biases has become so large,
+aggregating them all together has become a difficult task and some
+researchers provide resources to actively track new findings
+(e.g.~Microbiome Digest -
+\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
+Within the literature DNA extraction kits and storage condition chosen
+have consistently been shown to add bias to downstream analysis {[}{]}.
+However, the current literature on preservation media has been more
+mixed, with some studies showing biases while others do not {[}{]}.
+Although these are important sources of bias they are not the only
+sources that should be critically examined. The type of DNA polymerase
+chosen could have a wide ranging affect on downstream results due to
+error rates and chimeras that may not be easily resolved using
+bioinformatic approaches.
+
+A recent study in \emph{Nature Biotechnology} showed that there were
+clear differences between normal and high fidelity (HiFi) DNA polymerase
+and that you could reduce error and chimera generation by optimizing the
+PCR protocol (Gohl et al., 2016). Another important component of this
+study was that differences, based on DNA polymerase, in the number of
+OTUs generated were not easily removed using the authors chosen
+bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
+surprising that normal DNA polymerase performed worse than HiFi DNA
+polymerase, it is natural to extend this line of inquiry and ask whether
+different HiFi DNA polymerase contribute different biases to downstream
+sequencing results. There is some reason to think that they may since
+many of these HiFi DNA polymerase come from different families and may
+intrinsically have different error rates that cannot be completely
+removed with modifications (Ishino \& Ishino, 2014). In this study we
+critically examine if any of five different HiFi DNA polymerase
+introduce noticeable biases into 16S rRNA gene surveys, if this is a
+cycle dependent phenomenon, and whether they can be removed using a
+standard bioinformatic pipeline.
+
+We amplified the 16S rRNA gene in both fecal and mock community samples
+using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
+polymerase. First, we tested if we could identify differences in the
+number of OTUs between the differen HiFi DNA polymerase for both the
+fecal and mock samples. Next, since there were differences based of HiFi
+DNA polymerase, we examined if a community measure such as the
+Bray-Curtis index would be affected. After assessing these community
+measures we then examined if the different HiFi DNA polymerase had
+different per base error rates as well as chimera prevalence. Finally,
+since differences in chimera prevalence, based on HiFi DNA polymerase,
+could not be completely removed using bioinformatic approaches we tested
+whether the chimera prevalence correlated well with total number of
+observed OTUs.
 
@@
 
+\hypertarget{ref-metagenomcis_bias_Nayfach_2016}{}
+Nayfach S., Pollard KS. 2016. Toward accurate and quantitative
+comparative metagenomics. \emph{Cell} 166:1103--1116. DOI:
+\href{https://doi.org/10.1016/j.cell.2016.08.007}{10.1016/j.cell.2016.08.007}.
+
 \hypertarget{ref-vegan_citation}{}

### 2018-01-15 submission/manuscript.Rmd
add completed intro with citations

@@
 
-Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [].
-In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [].
-Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016].
+Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
+In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
+Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
 Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequece library preparation.
@@
 Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/).
-Within the literature DNA extraction kits and storage condition chosen have consistently been shown to add bias to downstream analysis [].
-However, the current literature on preservation media has been more mixed, with some studies showing biases while others do not [].
+Within the literature DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017].
+However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015].
 Although these are important sources of bias they are not the only sources that should be critically examined.

### 2018-01-15 submission/manuscript.tex
add completed intro with citations

@@
 Over recent years there has been an increasing focus on standardizing
-methodological approaches in microbiota research {[}{]}. In particular,
-a lot of the attention of this standardization push has been invested in
-investigating ways that 16S rRNA gene surveys can be made more
-replicatable {[}{]}. Although 16S rRNA gene sequencing has been much
-maligned for introduced bias, many of these same considerations also
-affect metagenomic sequencing (Nayfach \& Pollard, 2016). Between the
+methodological approaches in microbiota research (Kim et al., 2017;
+Hugerth \& Andersson, 2017). In particular, a lot of the attention of
+this standardization push has been invested in investigating ways that
+16S rRNA gene surveys can be made more replicatable (Lauber et al.,
+2010; Salter et al., 2014; Song et al., 2016; Gohl et al., 2016).
+Although 16S rRNA gene sequencing has been much maligned for introduced
+bias, many of these same considerations also affect metagenomic
+sequencing (Nayfach \& Pollard, 2016; Costea et al., 2017). Between the
 two approaches similar bias considerations include, but are not limited
@@
 \url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
-Within the literature DNA extraction kits and storage condition chosen
-have consistently been shown to add bias to downstream analysis {[}{]}.
-However, the current literature on preservation media has been more
-mixed, with some studies showing biases while others do not {[}{]}.
+Within the literature DNA extraction kits have consistently been shown
+to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
+2017). However, the current literature on preservation media and storage
+conditions has been more mixed, with some studies showing biases while
+others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
+al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
 Although these are important sources of bias they are not the only
@@
 \hypertarget{refs}{}
+\hypertarget{ref-storage_Bassis_2017}{}
+Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
+Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
+samples and storage conditions on bacterial community profiles.
+\emph{BMC Microbiology} 17. DOI:
+\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
+
 \hypertarget{ref-benjamini_controlling_1995}{}
@@
 \href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+
+\hypertarget{ref-metagenomcis_bias_Costea_2017}{}
+Costea PI., Zeller G., Sunagawa S., Pelletier E., Alberti A., Levenez
+F., Tramontano M., Driessen M., Hercog R., Jung F-E., Kultima JR.,
+Hayward MR., Coelho LP., Allen-Vercoe E., Bertrand L., Blaut M., Brown
+JRM., Carton T., Cools-Portier S., Daigneault M., Derrien M., Druesne
+A., Vos WM de., Finlay BB., Flint HJ., Guarner F., Hattori M., Heilig
+H., Luna RA., Hylckama Vlieg J van., Junick J., Klymiuk I., Langella P.,
+Chatelier EL., Mai V., Manichanh C., Martin JC., Mery C., Morita H.,
+O'Toole PW., Orvain C., Patil KR., Penders J., Persson S., Pons N.,
+Popova M., Salonen A., Saulnier D., Scott KP., Singh B., Slezak K.,
+Veiga P., Versalovic J., Zhao L., Zoetendal EG., Ehrlich SD., Dore J.,
+Bork P. 2017. Towards standards for human fecal sample processing in
+metagenomic studies. \emph{Nature Biotechnology}. DOI:
+\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.
+
+\hypertarget{ref-preservation_Dominianni_2014}{}
+Dominianni C., Wu J., Hayes RB., Ahn J. 2014. Comparison of methods for
+fecal microbiome biospecimen collection. \emph{BMC Microbiology} 14:103.
+DOI:
+\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.
 
@@
 
+\hypertarget{ref-review_Hugerth_2017}{}
+Hugerth LW., Andersson AF. 2017. Analysing microbial community
+composition through amplicon sequencing: From sampling to hypothesis
+testing. \emph{Frontiers in Microbiology} 8. DOI:
+\href{https://doi.org/10.3389/fmicb.2017.01561}{10.3389/fmicb.2017.01561}.
+
 \hypertarget{ref-polymerase_Ishino_2014}{}
@@
 \href{https://doi.org/10.3389/fmicb.2014.00465}{10.3389/fmicb.2014.00465}.
+
+\hypertarget{ref-review_Kim_2017}{}
+Kim D., Hofstaedter CE., Zhao C., Mattei L., Tanes C., Clarke E., Lauder
+A., Sherrill-Mix S., Chehoud C., Kelsen J., Conrad M., Collman RG.,
+Baldassano R., Bushman FD., Bittinger K. 2017. Optimizing methods and
+dodging pitfalls in microbiome research. \emph{Microbiome} 5. DOI:
+\href{https://doi.org/10.1186/s40168-017-0267-5}{10.1186/s40168-017-0267-5}.
 
@@
 DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-storage_Lauber_2010}{}
+Lauber CL., Zhou N., Gordon JI., Knight R., Fierer N. 2010. Effect of
+storage conditions on the assessment of bacterial community structure in
+soil and human-associated samples. \emph{FEMS Microbiology Letters}
+307:80--86. DOI:
+\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.
+
+\hypertarget{ref-preservation_Luo_2016}{}
+Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
+E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
+2016. Effects of specimen collection methodologies and storage
+conditions on the short-term stability of oral microbiome taxonomy.
+\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
+\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
 
@@
 \href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+
+\hypertarget{ref-preservation_Sinha_2015}{}
+Sinha R., Chen J., Amir A., Vogtmann E., Shi J., Inman KS., Flores R.,
+Sampson J., Knight R., Chia N. 2015. Collecting fecal samples for
+microbiome analyses in epidemiology studies. \emph{Cancer Epidemiology
+Biomarkers \& Prevention} 25:407--416. DOI:
+\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.
+
+\hypertarget{ref-preservation_Song_2016}{}
+Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
+R. 2016. Preservation methods differ in fecal microbiome stability,
+affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
+DOI:
+\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
 

### 2018-01-17 submission/manuscript.Rmd
grammar corrections before a lab reading

@@
 Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH.
-Further, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample that was also not changed by chimera removal with VSEARCH.
+Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and was also not changed by chimera removal with VSEARCH.
+
 **Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras.
@@
 Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
-Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequece library preparation.
+Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
 Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.
@@
 Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
-There is some reason to think that they may since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
+There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
 In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
@@
 First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples.
-Next, since there were differences based of HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected.
+Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected.
 After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence.
@@
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-If the HiFi Taq had a specific activation time that was different then 2 minutes that was used instead.
+If the HiFi Taq had a specific activation time that was different then 2 minutes, the specified activation was used instead.
 For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016].
@@
 The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the stated protocol were the use VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
 Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
@@
 Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
-The total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between different people.
+Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
+For the fecal samples the data was normalized to each individual by cycle number, to account for the biological variation between different people.
 Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
@@
 20 versus 25, 25 versus 30, etc.).
-From these observations we wanted to next analyze potential reasons as to why some of these differences may have occured.
-First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each Taq.
+Based on these observations we wanted to next analyze potential reasons as to why some of these differences may have occured.
+First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase.
 After assessing these errors, the total number of chimeras was determined after sequence processing.
-The fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.
+For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.
 
@@
 
-***The Number of OTUs is Dependent on HiFi Taq Used:***
+***The Number of OTUs is Dependent on HiFi DNA Polymerase Used:***
 After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi DNA polymerase used and this difference increased as the depth of sub-sampling increased [Figure 1].
 Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
-Only 35 cycles had HiFi Taq differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+Only 35 cycles had HiFi DNA polymerase differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
 At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA polymerase differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other Taq across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other DNA polymerases across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
 
@@
 Two exceptions to the low differences between 5-cycle increments can be found.
-The first that there seems to be large differences for fecal samples between 20x vs.
+The first, that there seems to be large differences for fecal samples between 20x vs.
 25x that is robust against sub-sampling depth [Figure 3A-B].
 Second, for the mocks, where data is available, there appears to be a similar large difference detected between 20x and 25x [Figure 3D].
-Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community [Figure 3].
+Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community between the same sample [Figure 3].
 
@@
 
-***Sequence Error is Dependent on both Taq and Cycle Number Used:***
-Differences by HiFi Taq in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest values were for the Kappa HiFI DNA polymerase [Figure 4A].
-Sub-sampling depth seems to have little effect on this rate with both 5000 and 10000 sub-sampled sequences showing similar results [Figure 4B-C].
-There were small differences between the various HiFi Taq at lower cycle number but larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].
+***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number Used:***
+Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
+Generally, the highest values were for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera removal by VSEARCH had little effect on this rate [Figure 4].
+Generally, there were small differences between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].
 
-The total seqeunces with at least one error had multiple differences at different cycle numbers and sub-sampling depth driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi Taq tested [Figure S1 & Table S7 & S8].
-Although Accuprime^TM^ had the lowest per base error rate it had either the largest or second largest number of sequences with at least one error regardless of cycle number or sub-sampling depth [Figure S1].
-Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there was no clear bias and this was indpendent of sub-sampling depth [Figure S2-S4].
-Further, the variation in subsitution error seems to reduce as the sub-sampling depth increases [Figure S2-S4].
+The total seqeunces with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1].
+Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8].
+Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1].
+However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1].
+Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there were generally no biases in the types of subsitution made [Figure S2].
 
-***Chimeric Sequences Correlate with OTUs and are HiFi Taq Dependent:***
+***Chimeric Sequences Correlate with OTUs and are HiFi DNA Polymerase Dependent:***
 After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
@@
 
-For all Taqs a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
+For all DNA polymerase, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
 In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6].
 Taken together this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.
-Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].
+Additionally, Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].
 
@@
 
-Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA Polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
-This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results as well as the findings.
-Although the attention has mostly been on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4].
-These differences were observed in high biomass samples, where biases introduced by such components like kit contamination have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
+Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
+This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results, as well as the findings.
+Although the attention has mostly been focused on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4].
+These differences were observed in high biomass samples, where biases introduced by such components like kit contamination may have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
 
 Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30x, that are commonly used.
-What is most interesting is that there was no difference within individuals between corresponding 5 cycle increments (e.g.
+Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g.
 15 to 20, 20 to 25, etc.).
-Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different [Figure 3].
-This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once at least 25 cylces are reached.
-Increasing the sub-sampling depth, for some DNA polymerase may reduce some of these observed community differences at lower cycle numbers.
+Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different from each other [Figure 3].
+This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cylces are reached.
+Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.
 
 Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5].
-The chimera prevalence was very strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations.
+The chimera prevalence was strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations.
 However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase.
-Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant that the Bray-Curtis index is able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
-So, choice of downstream diversity metric could be an important consideration to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase such as Platinum.
+Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
+So, choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase.
 

### 2018-01-17 submission/manuscript.tex
grammar corrections before a lab reading

@@
 differences were still observed after chimera removal using VSEARCH.
-Further, the chimera prevalence had a strong positive correlation with
-the number of OTUs observed in a sample that was also not changed by
-chimera removal with VSEARCH.\\
+Additionally, the chimera prevalence had a strong positive correlation
+with the number of OTUs observed in a sample and was also not changed by
+chimera removal with VSEARCH.
+
 \textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
@@
 to, preservation media, storage conditions, DNA extraction kit, PCR, and
-sequece library preparation. Thus what is identified as a source of bias
-in 16S rRNA gene sequencing for these overlapping considerations will
-also influence metagenomic sequencing results.
+sequence library preparation. Thus what is identified as a source of
+bias in 16S rRNA gene sequencing for these overlapping considerations
+will also influence metagenomic sequencing results.
 
@@
 different HiFi DNA polymerase contribute different biases to downstream
-sequencing results. There is some reason to think that they may since
-many of these HiFi DNA polymerase come from different families and may
-intrinsically have different error rates that cannot be completely
-removed with modifications (Ishino \& Ishino, 2014). In this study we
-critically examine if any of five different HiFi DNA polymerase
+sequencing results. There is some reason to think that this may be the
+case since many of these HiFi DNA polymerase come from different
+families and may intrinsically have different error rates that cannot be
+completely removed with modifications (Ishino \& Ishino, 2014). In this
+study we critically examine if any of five different HiFi DNA polymerase
 introduce noticeable biases into 16S rRNA gene surveys, if this is a
@@
 number of OTUs between the differen HiFi DNA polymerase for both the
-fecal and mock samples. Next, since there were differences based of HiFi
+fecal and mock samples. Next, since there were differences based on HiFi
 DNA polymerase, we examined if a community measure such as the
@@
 If the HiFi Taq had a specific activation time that was different then 2
-minutes that was used instead. For Kappa and Q5 the protocol previously
-published by Gohl and colleagues was used (Gohl et al., 2016). For
-Phusion the company defined conditions were used but the same extension
-time as that used for Accuprime and Platinum was used.
+minutes, the specified activation was used instead. For Kappa and Q5 the
+protocol previously published by Gohl and colleagues was used (Gohl et
+al., 2016). For Phusion the company defined conditions were used but the
+same extension time as that used for Accuprime and Platinum was used.
 
@@
 (Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
-major differences from the stated protocol were the use VSEARCH instead
-of UCHIME for chimera detection and the use of the OptiClust algorithmn
-instead of average neighbor for Operational Taxonomic Unit (OTU)
-generation (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
-Schloss, 2017). Sequence error was determined using the seq.error
-command on mock samples after chimera removal and classification to the
-RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
-al., 2013; Rognes et al., 2016).
+major differences from the stated protocol were the use of VSEARCH
+instead of UCHIME for chimera detection and the use of the OptiClust
+algorithmn instead of average neighbor for Operational Taxonomic Unit
+(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
+2016; Westcott \& Schloss, 2017). Sequence error was determined using
+the seq.error command on mock samples after chimera removal and
+classification to the RDP to remove non-bacterial sequences (Schloss et
+al., 2009; Cole et al., 2013; Rognes et al., 2016).
 
@@
 (v 3.4.3) software package (R Core Team, 2017). Data tranformation and
-graphing was completed using the tidyverse package (v 1.1.1) and colors
-selected using the viridis package (v 0.4.0) (Garnier, 2017; Wickham,
-2017). The total number of OTUs were analyzed using an ANOVA with a
-tukey post-hoc test. For the fecal samples the data was normalized to
-each individual by cycle number to account for the biological variation
-between different people. Bray-Curtis matrices were generated using
-mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The
-distance matrix data was analyzed using PERMANOVA with the vegan package
-(v 2.4.4) (Oksanen et al., 2017) and kruskal-wallis tests within R. For
-both error and chimera analysis, samples were tested using
+graphing was completed using the tidyverse package (v 1.2.1) and colors
+selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
+2017). Differences in the total number of OTUs were analyzed using an
+ANOVA with a tukey post-hoc test. For the fecal samples the data was
+normalized to each individual by cycle number, to account for the
+biological variation between different people. Bray-Curtis matrices were
+generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
+15000. The distance matrix data was analyzed using PERMANOVA with the
+vegan package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests
+within R. For both error and chimera analysis, samples were tested using
 Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
@@
 individual differences for the previous cycle (e.g.~20 versus 25, 25
-versus 30, etc.). From these observations we wanted to next analyze
+versus 30, etc.). Based on these observations we wanted to next analyze
 potential reasons as to why some of these differences may have occured.
@@
 an error, and base subsitution were assessed in the mock community for
-each Taq. After assessing these errors, the total number of chimeras was
-determined after sequence processing. The fecal samples were analyzed at
-4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the
-mock community samples were analysed at 3 levels, 1000, 5000, 10000.
+each DNA polymerase. After assessing these errors, the total number of
+chimeras was determined after sequence processing. For the community
+based measures, the fecal samples were analyzed at 4 different
+sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock
+community samples were analysed at 3 levels, 1000, 5000, 10000.
 
@@
 
-\textbf{\emph{The Number of OTUs is Dependent on HiFi Taq Used:}} After
-normalization by individual, for each cycle number, we observed that for
-fecal samples the number of OTUs identified was dependent upon the HiFi
-DNA polymerase used and this difference increased as the depth of
-sub-sampling increased {[}Figure 1{]}. Lower cycle numbers (15-20)
-resulted in less differences between HiFI DNA polymerase while cycle
-numbers of 25, 30, and 35 had larger clearer defined differences
-{[}Figure 1{]}. Only 35 cycles had HiFi Taq differences that were
-significantly different at all sub-sampling levels (P-value \textless{}
-0.0001) {[}Table S1{]}. At sub-sampling depth of 5000 or higher 25 and
-30 cycles had HiFi DNA polymerase differences (P-value \textless{} 0.05)
-{[}Table S1{]}. Using a Tukey post-hoc test only at 35 cycles were
-significant differences found to be mainly driven by Platinum being
-different than all other Taq across sub-sampling depth (P-value
-\textless{} 0.05) {[}Table S2{]}.
+\textbf{\emph{The Number of OTUs is Dependent on HiFi DNA Polymerase
+Used:}} After normalization by individual, for each cycle number, we
+observed that for fecal samples the number of OTUs identified was
+dependent upon the HiFi DNA polymerase used and this difference
+increased as the depth of sub-sampling increased {[}Figure 1{]}. Lower
+cycle numbers (15-20) resulted in less differences between HiFI DNA
+polymerase while cycle numbers of 25, 30, and 35 had larger clearer
+defined differences {[}Figure 1{]}. Only 35 cycles had HiFi DNA
+polymerase differences that were significantly different at all
+sub-sampling levels (P-value \textless{} 0.0001) {[}Table S1{]}. At
+sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA
+polymerase differences (P-value \textless{} 0.05) {[}Table S1{]}. Using
+a Tukey post-hoc test only at 35 cycles were significant differences
+found to be mainly driven by Platinum being different than all other DNA
+polymerases across sub-sampling depth (P-value \textless{} 0.05)
+{[}Table S2{]}.
 
@@
 {[}Figure 3{]}. Two exceptions to the low differences between 5-cycle
-increments can be found. The first that there seems to be large
+increments can be found. The first, that there seems to be large
 differences for fecal samples between 20x vs.~25x that is robust against
@@
 reach 25x there does not seem to be large differences in the community
-{[}Figure 3{]}.
+between the same sample {[}Figure 3{]}.
 
@@
 
-\textbf{\emph{Sequence Error is Dependent on both Taq and Cycle Number
-Used:}} Differences by HiFi Taq in the median average per base error
-varied without a clear pattern across sub-sampling depth {[}Table S5{]}.
-Generally, the highest values were for the Kappa HiFI DNA polymerase
-{[}Figure 4A{]}. Sub-sampling depth seems to have little effect on this
-rate with both 5000 and 10000 sub-sampled sequences showing similar
-results {[}Figure 4B-C{]}. There were small differences between the
-various HiFi Taq at lower cycle number but larger differences at higher
-cycle number with Platinum having large differences between all other
-HiFi DNA polymerase {[}Figure 4B-C and Table S6{]}.
+\textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
+and Cycle Number Used:}} Differences by HiFi DNA polymerase in the
+median average per base error varied without a clear pattern across
+sub-sampling depth {[}Table S5{]}. Generally, the highest values were
+for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera
+removal by VSEARCH had little effect on this rate {[}Figure 4{]}.
+Generally, there were small differences between the various HiFi DNA
+polymerase at lower cycle numbers and larger differences at higher cycle
+number with Platinum having large differences between all other HiFi DNA
+polymerase {[}Figure 4B-C and Table S6{]}.
 
 The total seqeunces with at least one error had multiple differences at
-different cycle numbers and sub-sampling depth driven by large
-differences in Accuprime\textsuperscript{TM} and Platinum versus the
-other HiFi Taq tested {[}Figure S1 \& Table S7 \& S8{]}. Although
+different cycle numbers and was mostly alleviated by the use of the
+pre.cluster step {[}Figure S1{]}. Major differences before this
+pre.cluster step were driven by large differences in
+Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
+polymerase tested {[}Figure S1 \& Table S7 \& S8{]}. Although
 Accuprime\textsuperscript{TM} had the lowest per base error rate it had
-either the largest or second largest number of sequences with at least
-one error regardless of cycle number or sub-sampling depth {[}Figure
-S1{]}. Investigation of whether there were HiFi DNA polymerase dependent
-effects on base subsitution found that there was no clear bias and this
-was indpendent of sub-sampling depth {[}Figure S2-S4{]}. Further, the
-variation in subsitution error seems to reduce as the sub-sampling depth
-increases {[}Figure S2-S4{]}.
+the largest number of sequences with at least one error, regardless of
+cycle number or sub-sampling depth {[}Figure S1{]}. However, this
+increased number of sequences with an error can be drastically lowered
+with existing bioinformatic approaches {[}Figure S1{]}. Investigation of
+whether there were HiFi DNA polymerase dependent effects on base
+subsitution found that there were generally no biases in the types of
+subsitution made {[}Figure S2{]}.
 
-\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi Taq
-Dependent:}} After chimera removal using VSEARCH and removal of
-sequences that did not classify as bacteria we assessed the percentage
-of sequences that were still chimeric within our mock community. At all
-levels of sub-sampling and cycle number there were significant
-differences between the HiFi DNA polymerase used (P-value \textless{}
-0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast majority of
-these differences were driven by Platinum being different then all other
-HiFi DNA polymerase across cycle number and sub-sampling depth {[}Table
-S10{]}. Generally, across sub-sampling depth and cycle number
-Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
-the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had
-been used {[}Figure 5{]}.
+\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi DNA
+Polymerase Dependent:}} After chimera removal using VSEARCH and removal
+of sequences that did not classify as bacteria we assessed the
+percentage of sequences that were still chimeric within our mock
+community. At all levels of sub-sampling and cycle number there were
+significant differences between the HiFi DNA polymerase used (P-value
+\textless{} 0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast
+majority of these differences were driven by Platinum being different
+then all other HiFi DNA polymerase across cycle number and sub-sampling
+depth {[}Table S10{]}. Generally, across sub-sampling depth and cycle
+number Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+of all the HiFi DNA Polymerase regardless of whether pre.cluster or
+VSEARCH had been used {[}Figure 5{]}.
 
-For all Taqs a positive correlation was observed between chimeric
-sequences and number of OTUs, with this correlation being strongest for
-Accuprime, Platinum and Phusion HiFi DNA Polymerase {[}Figure 6{]}. In
-general, the R\textsuperscript{2} value between the number of OTUs and
-chimeric sequences did not change from the use of pre.cluster and
-VSEARCH {[}Figure 6{]}. Taken together this data suggests that a strong
-correlation exists between the number of OTUs and the prevalence of
-chimeric sequences. Kappa had the highest per base error rate and the
-lowest correlations between the number of OTUs and chimeric sequences
-across sub-sampling depth {[}Figure 4 \& 6{]}.
+For all DNA polymerase, a positive correlation was observed between
+chimeric sequences and number of OTUs, with this correlation being
+strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
+{[}Figure 6{]}. In general, the R\textsuperscript{2} value between the
+number of OTUs and chimeric sequences did not change from the use of
+pre.cluster and VSEARCH {[}Figure 6{]}. Taken together this data
+suggests that a strong correlation exists between the number of OTUs and
+the prevalence of chimeric sequences. Additionally, Kappa had the
+highest per base error rate and the lowest correlations between the
+number of OTUs and chimeric sequences across sub-sampling depth
+{[}Figure 4 \& 6{]}.
 
@@
 Our observations build upon previous studies (Gohl et al., 2016) by
-showing that even different HiFi DNA Polymerase have significant
+showing that even different HiFi DNA polymerase have significant
 differences in the number of OTUs and that the changes to total OTUs
@@
 total number of OTUs as part of their calculations and changes to the
-total number of OTUs could drastically change the results as well as the
-findings. Although the attention has mostly been on standardizing and
-improving collection and extraction methods (Salter et al., 2014) our
-observations show that independent of this consideration HiFi DNA
-polymerase can have a noticeable affect on the OTUs generated that can
-be found across sub-sampling depth and PCR cycle number {[}Figure
-2-4{]}. These differences were observed in high biomass samples, where
-biases introduced by such components like kit contamination have less of
-an effect, suggesting that these differences may be exacerbated in low
-biomass samples.
+total number of OTUs could drastically change the results, as well as
+the findings. Although the attention has mostly been focused on
+standardizing and improving collection and extraction methods (Salter et
+al., 2014) our observations show that independent of this consideration
+HiFi DNA polymerase can have a noticeable affect on the OTUs generated
+that can be found across sub-sampling depth and PCR cycle number
+{[}Figure 2-4{]}. These differences were observed in high biomass
+samples, where biases introduced by such components like kit
+contamination may have less of an effect, suggesting that these
+differences may be exacerbated in low biomass samples.
 
@@
 differences between 15 cycles and higher cycle numbers, such as 30x,
-that are commonly used. What is most interesting is that there was no
-difference within individuals between corresponding 5 cycle increments
-(e.g.~15 to 20, 20 to 25, etc.). Conversely, this may be due to low
-power and on observation there does seem to be a trend that 20x and 25x
-communities are very different {[}Figure 3{]}. This finding, in
+that are commonly used. Additionally, there was no difference within
+individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
+to 25, etc.). Conversely, this may be due to low power and on
+observation there does seem to be a trend that 20x and 25x communities
+are very different from each other {[}Figure 3{]}. This finding, in
 conjunction with the PERMANOVA results, suggest that cycle number can
 change bacterial community calculations but that these differences are
-minimal once at least 25 cylces are reached. Increasing the sub-sampling
-depth, for some DNA polymerase may reduce some of these observed
-community differences at lower cycle numbers.
+minimal once 25 cylces are reached. Increasing the sub-sampling depth,
+for some DNA polymerase, may reduce some of these observed community
+differences at lower cycle numbers.
 
@@
 differences between the different HiFi DNA polymerases {[}Figure 5{]}.
-The chimera prevalence was very strongly correlated with the number of
-OTUs and this value is relied upon heavily for many different downstream
+The chimera prevalence was strongly correlated with the number of OTUs
+and this value is relied upon heavily for many different downstream
 community metric calculations. However, Bray-Curtis analysis with
@@
 possible that many of the increased number of OTUs, generated as cycle
-number increases, are not highly abundant that the Bray-Curtis index is
-able to successfully downweight these respective OTUs (Minchin, 1987).
-So, choice of downstream diversity metric could be an important
-consideration to mitigate these observed changes due to high chimera
-prevalence in HiFi DNA polymerase such as Platinum.
+number increases, are not highly abundant allowing the Bray-Curtis index
+to be able to successfully downweight these respective OTUs (Minchin,
+1987). So, choice of downstream diversity metric could be an important
+consideration in helping to mitigate these observed changes due to high
+chimera prevalence in HiFi DNA polymerase.
 

### 2018-01-29 submission/manuscript.Rmd
incorporate nick suggestions - abstract

@@
 
-# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Gene Sequencing
+# What the Taq? The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene Sequencing
 
@@
 These components include, but are not necessarily limited to, preservation media, extraction kit, bead beating time, and primers.
-Both cycle number and Hifidelity (HiFi) DNA polymerase are sometimes overlooked during this conversation of bias considerations.
+Both cycle number and high fidelity (HiFi) DNA polymerase are sometimes overlooked when sources of bias are being considered.
 Here we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.
@@
 **Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Mock communities (technical replicates n = 4) consisting of the whole genome of 8 different bacteria were also amplified using the same approach.
-The number of OTUs (Operational Taxonomic Units) was first examined for both fecal samples and mock communites.
-Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence, and the chimera prevalence correlation with number of OTUs was assessed for either cycle number or HiFi DNA polymerase dependent differences.
+Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach.
+First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communites.
+Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase.
+Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.
 
@@
 **Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras.
-So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUS that are observed.
+So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.
 

### 2018-01-29 submission/manuscript.tex
incorporate nick suggestions - abstract

@@
 
-\section{What the Taq? The Influence of Different Hi-Fidelity Taq
-Polymerase on 16S rRNA Gene
-Sequencing}\label{what-the-taq-the-influence-of-different-hi-fidelity-taq-polymerase-on-16s-rrna-gene-sequencing}
+\section{What the Taq? The Influence of High Fidelity DNA Polymerase on
+16S rRNA Gene
+Sequencing}\label{what-the-taq-the-influence-of-high-fidelity-dna-polymerase-on-16s-rrna-gene-sequencing}
 
@@
 but are not necessarily limited to, preservation media, extraction kit,
-bead beating time, and primers. Both cycle number and Hifidelity (HiFi)
-DNA polymerase are sometimes overlooked during this conversation of bias
-considerations. Here we critically examine both cycle number and HiFi
+bead beating time, and primers. Both cycle number and high fidelity
+(HiFi) DNA polymerase are sometimes overlooked when sources of bias are
+being considered. Here we critically examine both cycle number and HiFi
 DNA polymerase for biases that may influence downstream diversity
@@
 Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
-(technical replicates n = 4) consisting of the whole genome of 8
-different bacteria were also amplified using the same approach. The
-number of OTUs (Operational Taxonomic Units) was first examined for both
-fecal samples and mock communites. Next, Bray-Curtis index, the error
-rate, sequence error prevalence, and chimera prevalence, and the chimera
-prevalence correlation with number of OTUs was assessed for either cycle
-number or HiFi DNA polymerase dependent differences.
+(technical replicates n = 4) consisting of previously isolated whole
+genomes of 8 different bacteria were also amplified using the same PCR
+amplification approach. First, the number of OTUs (Operational Taxonomic
+Units) was examined for both fecal samples and mock communites. Next,
+Bray-Curtis index, the error rate, sequence error prevalence, and
+chimera prevalence were assessed based on cycle number and HiFi DNA
+polymerase. Finally, the chimera prevalence correlation with number of
+OTUs was assessed for both cycle number and HiFi DNA polymerase
+dependent differences.
 
@@
 polymerase that will be used since it can increase or decrease the
-number of OTUS that are observed.
+number of OTUs that are observed.
 

### 2018-01-29 submission/manuscript.Rmd
nick intro suggestions

@@
 Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
-In particular, a lot of the attention of this standardization push has been invested in investigating ways that 16S rRNA gene surveys can be made more replicatable [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
+In particular, investigating ways that 16S rRNA gene surveys can be made more replicatable has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
 Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
 Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
-Thus what is identified as a source of bias in 16S rRNA gene sequencing for these overlapping considerations will also influence metagenomic sequencing results.
+Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
 
-The most commonly studied biases are by far the skews introduced to the data via preservation media used, storage conditions, and DNA extraction kits chosen.
+Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases.
 The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g.
 Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/).
-Within the literature DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017].
+Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017].
 However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015].
@@
 Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
-There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-In this study we critically examine if any of five different HiFi DNA polymerase introduce noticeable biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
+*Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
+In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 
 We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase.
-First, we tested if we could identify differences in the number of OTUs between the differen HiFi DNA polymerase for both the fecal and mock samples.
+First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples.
 Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected.

### 2018-01-29 submission/manuscript.tex
nick intro suggestions

@@
 methodological approaches in microbiota research (Kim et al., 2017;
-Hugerth \& Andersson, 2017). In particular, a lot of the attention of
-this standardization push has been invested in investigating ways that
-16S rRNA gene surveys can be made more replicatable (Lauber et al.,
-2010; Salter et al., 2014; Song et al., 2016; Gohl et al., 2016).
-Although 16S rRNA gene sequencing has been much maligned for introduced
-bias, many of these same considerations also affect metagenomic
-sequencing (Nayfach \& Pollard, 2016; Costea et al., 2017). Between the
-two approaches similar bias considerations include, but are not limited
-to, preservation media, storage conditions, DNA extraction kit, PCR, and
-sequence library preparation. Thus what is identified as a source of
-bias in 16S rRNA gene sequencing for these overlapping considerations
-will also influence metagenomic sequencing results.
+Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
+rRNA gene surveys can be made more replicatable has been a predominant
+driver of this standardization push (Lauber et al., 2010; Salter et al.,
+2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
+sequencing has been much maligned for introduced bias, many of these
+same considerations also affect metagenomic sequencing (Nayfach \&
+Pollard, 2016; Costea et al., 2017). Between the two approaches similar
+bias considerations include, but are not limited to, preservation media,
+storage conditions, DNA extraction kit, PCR, and sequence library
+preparation. Thus, for these overlapping considerations biases
+identified for 16S rRNA gene sequencing will also likely influence
+metagenomic sequencing results.
 
-The most commonly studied biases are by far the skews introduced to the
-data via preservation media used, storage conditions, and DNA extraction
-kits chosen. The study of these specific biases has become so large,
-aggregating them all together has become a difficult task and some
-researchers provide resources to actively track new findings
-(e.g.~Microbiome Digest -
+Currently, preservation media used, storage conditions, and DNA
+extraction kits chosen are the most commonly studied biases. The study
+of these specific biases has become so large, aggregating them all
+together has become a difficult task and some researchers provide
+resources to actively track new findings (e.g.~Microbiome Digest -
 \url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
-Within the literature DNA extraction kits have consistently been shown
+Within the literature, DNA extraction kits have consistently been shown
 to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
@@
 case since many of these HiFi DNA polymerase come from different
-families and may intrinsically have different error rates that cannot be
-completely removed with modifications (Ishino \& Ishino, 2014). In this
-study we critically examine if any of five different HiFi DNA polymerase
-introduce noticeable biases into 16S rRNA gene surveys, if this is a
+families (e.g. \emph{Taq} belongs to the family A polymerases) and may
+intrinsically have different error rates that cannot be completely
+removed with modifications (Ishino \& Ishino, 2014). In this study we
+critically examine if any of five different HiFi DNA polymerases
+introduce significant biases into 16S rRNA gene surveys, if this is a
 cycle dependent phenomenon, and whether they can be removed using a
@@
 polymerase. First, we tested if we could identify differences in the
-number of OTUs between the differen HiFi DNA polymerase for both the
+number of OTUs between the different HiFi DNA polymerase for both the
 fecal and mock samples. Next, since there were differences based on HiFi

### 2018-01-29 submission/manuscript.Rmd
add nick edits -- methods

@@
 ***Human and Mock Samples:***
-A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN) and the processing and storage of these samples have been published previously [@erin_seekatz_2016].
+A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
+The processing and storage of these samples have been published previously [@erin_seekatz_2016].
 Clinical data and other types of meta data were not utilized or accessed for this study.
@@
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-If the HiFi Taq had a specific activation time that was different then 2 minutes, the specified activation was used instead.
-For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016].
-For Phusion the company defined conditions were used but the same extension time as that used for Accuprime and Platinum was used.
+The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified.
+For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016].
+For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.
 
-The 30 cycle default was used but the cycle conditions started at 15 and increased by 5 up to 35 cycles and was used for both fecal and mock samples.
+Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
 The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100).
-Although, the mock communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10 replicates were used for 30 cycles for all Taq (total samples = 130).
-For all the mock community samples there was not enough PCR product at 15 cycles for adequate sequencing.
+Although, the mock communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130).
+No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
 
@@
 The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009].
-The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Generally, the protocol followed what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
 Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
@@
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the fecal samples the data was normalized to each individual by cycle number, to account for the biological variation between different people.
+For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
 Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
@@
 20 versus 25, 25 versus 30, etc.).
-Based on these observations we wanted to next analyze potential reasons as to why some of these differences may have occured.
+Based on these observations we wanted to next analyze potential reasons for these differences.
 First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase.
 After assessing these errors, the total number of chimeras was determined after sequence processing.
-For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.
+For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
 

### 2018-01-29 submission/manuscript.tex
add nick edits -- methods

@@
 obtained from 4 individuals who were part of the Enterics Research
-Investigational Network (ERIN) and the processing and storage of these
+Investigational Network (ERIN). The processing and storage of these
 samples have been published previously (Seekatz et al., 2016). Clinical
@@
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
-If the HiFi Taq had a specific activation time that was different then 2
-minutes, the specified activation was used instead. For Kappa and Q5 the
-protocol previously published by Gohl and colleagues was used (Gohl et
-al., 2016). For Phusion the company defined conditions were used but the
-same extension time as that used for Accuprime and Platinum was used.
+The HiFi DNA polymerase activation time was 2 minutes, unless a
+different activation was specified. For Kappa and Q5, the protocol
+previously published by Gohl and colleagues was used (Gohl et al.,
+2016). For Phusion, the company defined conditions were used except for
+extension time, where the Accuprime and Platinum settings were used.
 
-The 30 cycle default was used but the cycle conditions started at 15 and
-increased by 5 up to 35 cycles and was used for both fecal and mock
-samples. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and
-35 cycles for each Taq (total samples = 100). Although, the mock
-communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10
-replicates were used for 30 cycles for all Taq (total samples = 130).
-For all the mock community samples there was not enough PCR product at
-15 cycles for adequate sequencing.
+Both fecal and mock samples cycle conditions started at 15 and increased
+by 5 up to 35 cycles with amplicons used at each 5-step increase for
+sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
+and 35 cycles for each Taq (total samples = 100). Although, the mock
+communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10
+replicates for 30 cycles for all HiFi DNA polymerase (total samples =
+130). No mock community sample had enough PCR product at 15 cycles for
+adequate 16S rRNA gene sequencing.
 
 \textbf{\emph{Sequence Processing:}} The mothur software program was
-utilized for all sequence processing steps (Schloss et al., 2009). The
-protocol followed was similar to what has been previously publsihed
+utilized for all sequence processing steps (Schloss et al., 2009).
+Generally, the protocol followed what has been previously publsihed
 (Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
@@
 ANOVA with a tukey post-hoc test. For the fecal samples the data was
-normalized to each individual by cycle number, to account for the
-biological variation between different people. Bray-Curtis matrices were
-generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
-15000. The distance matrix data was analyzed using PERMANOVA with the
-vegan package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests
-within R. For both error and chimera analysis, samples were tested using
+normalized to each individual by cycle number to account for the
+biological variation between people. Bray-Curtis matrices were generated
+using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
+The distance matrix data was analyzed using PERMANOVA with the vegan
+package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests within
+R. For both error and chimera analysis, samples were tested using
 Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
@@
 versus 30, etc.). Based on these observations we wanted to next analyze
-potential reasons as to why some of these differences may have occured.
-First, analysis of general sequence error rate, number of sequences with
-an error, and base subsitution were assessed in the mock community for
-each DNA polymerase. After assessing these errors, the total number of
-chimeras was determined after sequence processing. For the community
-based measures, the fecal samples were analyzed at 4 different
-sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock
-community samples were analysed at 3 levels, 1000, 5000, 10000.
+potential reasons for these differences. First, analysis of general
+sequence error rate, number of sequences with an error, and base
+subsitution were assessed in the mock community for each DNA polymerase.
+After assessing these errors, the total number of chimeras was
+determined after sequence processing. For the community based measures,
+the fecal samples were analyzed at 4 different sub-sampling levels
+(1000, 5000, 10000, and 15000) while the mock community samples were
+analysed at 3 levels (1000, 5000, 10000).
 

### 2018-01-29 submission/manuscript.Rmd
nick suggestions results

@@
 bray_permanova_by_taq_results <- as.data.frame(read_csv("../data/process/tables/bray_permanova_by_taq_results.csv"))
+
+# Vector of sub samples used
+sub_sample_level <- c("1000", "5000", "10000")
+
+dist_data <- sapply(sub_sample_level, 
+                    function(x) as.data.frame(read_csv(paste("../data/process/tables/bray_5_cycle_dist_",                                                x, "_data.csv", sep = ""))), simplify = F)
+
+mock_dist_data <- sapply(sub_sample_level, 
+                    function(x) as.data.frame(read_csv(paste("../data/process/tables/mock_bray_5_cycle_dist_", 
+                                               x, "_data.csv", sep = ""))), simplify = F)
 
@@
 
-***The Number of OTUs is Dependent on HiFi DNA Polymerase Used:***
-After normalization by individual, for each cycle number, we observed that for fecal samples the number of OTUs identified was dependent upon the HiFi DNA polymerase used and this difference increased as the depth of sub-sampling increased [Figure 1].
-Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 had larger clearer defined differences [Figure 1].
-Only 35 cycles had HiFi DNA polymerase differences that were significantly different at all sub-sampling levels (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
-At sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA polymerase differences (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-Using a Tukey post-hoc test only at 35 cycles were significant differences found to be mainly driven by Platinum being different than all other DNA polymerases across sub-sampling depth (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+***Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
+In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data.
+After normalization, we identified that there was a HiFi DNA polymerase dependent difference that became more apparent as the depth of sub-sampling increased [Figure 1].
+Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 having distinct differences [Figure 1].
+All subsampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
+Platinum HiFi DNA polymerase was the main driver of the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
 
-This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases being high (Platinum) and low (Accuprime) respectively [Figure 2 & Table S3].
+This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases having high (Platinum) and low (Accuprime) number of OTUs [Figure 2 & Table S3].
 Conversely, differences between HiFi DNA polymerase were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
@@
 15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3].
-Two exceptions to the low differences between 5-cycle increments can be found.
-The first, that there seems to be large differences for fecal samples between 20x vs.
-25x that is robust against sub-sampling depth [Figure 3A-B].
-Second, for the mocks, where data is available, there appears to be a similar large difference detected between 20x and 25x [Figure 3D].
-Regardless, by the time PCR cycles reach 25x there does not seem to be large differences in the community between the same sample [Figure 3].
+There were only two differences between 5-cycle increments that were identified.
+First, there were differences for the same fecal sample between 20x vs.
+25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (subsampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), subsampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), subsampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A-B].
+Second, for the mocks, where data is available, there were larger difference between 20x and 25x (subsampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3D].
+However, these differences between the next 5 cycle increment do not persist once 25 cylces are reached [Figure 3].
 
-Using PERMANOVA to test whether there were any differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Interestingly, Phusion was one of only two DNA polymerase that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
+Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
+Phusion was one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
 Next, we assessed whether there were any major differences between 5 cycle increments within each sample.
@@
 
-***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number Used:***
+***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number:***
 Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest values were for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera removal by VSEARCH had little effect on this rate [Figure 4].
-Generally, there were small differences between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having large differences between all other HiFi DNA polymerase [Figure 4B-C and Table S6].
+Generally, the highest per base median average error rates were for the Kappa HiFI DNA polymerase [Figure 4].
+This error rate was minimally affected by both the pre-cluster step and chimera removal by VSEARCH [Figure 4].
+There were small differences in the per base error rate between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having the largest differences of all the HiFi DNA polymerase [Figure 4B-C and Table S6].
 
@@
 At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Using a Dunn's post-hoc test the vast majority of these differences were driven by Platinum being different then all other HiFi DNA polymerase across cycle number and sub-sampling depth [Table S10].
-Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had been used [Figure 5].
+Differences between Platinum and all other HiFi DNA polymerases accounted for the vast majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
+Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether pre.cluster or VSEARCH had been used [Figure 5].
 
@@
 In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6].
-Taken together this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.
-Additionally, Kappa had the highest per base error rate and the lowest correlations between the number of OTUs and chimeric sequences across sub-sampling depth [Figure 4 & 6].
+Taken together, this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.
 

### 2018-01-29 submission/manuscript.tex
nick suggestions results

@@
 
-\textbf{\emph{The Number of OTUs is Dependent on HiFi DNA Polymerase
-Used:}} After normalization by individual, for each cycle number, we
-observed that for fecal samples the number of OTUs identified was
-dependent upon the HiFi DNA polymerase used and this difference
-increased as the depth of sub-sampling increased {[}Figure 1{]}. Lower
-cycle numbers (15-20) resulted in less differences between HiFI DNA
-polymerase while cycle numbers of 25, 30, and 35 had larger clearer
-defined differences {[}Figure 1{]}. Only 35 cycles had HiFi DNA
-polymerase differences that were significantly different at all
-sub-sampling levels (P-value \textless{} 0.0001) {[}Table S1{]}. At
-sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA
-polymerase differences (P-value \textless{} 0.05) {[}Table S1{]}. Using
-a Tukey post-hoc test only at 35 cycles were significant differences
-found to be mainly driven by Platinum being different than all other DNA
-polymerases across sub-sampling depth (P-value \textless{} 0.05)
-{[}Table S2{]}.
+\textbf{\emph{Number of OTUs is Dependent on the HiFi DNA Polymerase
+Used:}} In order to compare the number of OTUs across individuals a
+Z-score normalization by individual, by cycle number, was employed on
+the number of OTUs data. After normalization, we identified that there
+was a HiFi DNA polymerase dependent difference that became more apparent
+as the depth of sub-sampling increased {[}Figure 1{]}. Lower cycle
+numbers (15-20) resulted in less differences between HiFI DNA polymerase
+while cycle numbers of 25, 30, and 35 having distinct differences
+{[}Figure 1{]}. All subsampling levels were significantly different only
+at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There were
+differences in HiFi DNA polymerase at 25 and 30 cycles but the
+sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
+{[}Table S1{]}. Platinum HiFi DNA polymerase was the main driver of the
+differences observed across all sub-sampling depths at 35 cycles, based
+on a Tukey post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}.
 
@@
 also observed in the mock community samples with the same DNA
-polymerases being high (Platinum) and low (Accuprime) respectively
+polymerases having high (Platinum) and low (Accuprime) number of OTUs
 {[}Figure 2 \& Table S3{]}. Conversely, differences between HiFi DNA
@@
 samples and this was consistent across the different sub-samplings used
-{[}Figure 3{]}. Two exceptions to the low differences between 5-cycle
-increments can be found. The first, that there seems to be large
-differences for fecal samples between 20x vs.~25x that is robust against
-sub-sampling depth {[}Figure 3A-B{]}. Second, for the mocks, where data
-is available, there appears to be a similar large difference detected
-between 20x and 25x {[}Figure 3D{]}. Regardless, by the time PCR cycles
-reach 25x there does not seem to be large differences in the community
-between the same sample {[}Figure 3{]}.
+{[}Figure 3{]}. There were only two differences between 5-cycle
+increments that were identified. First, there were differences for the
+same fecal sample between 20x vs.~25x that was independent of HiFi DNA
+polymerase but dependent on sub-sampling depth (subsampled to 1000 =
+0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), subsampled to 5000 =
+0.43 (0.33 - 0.63), subsampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
+3A-B{]}. Second, for the mocks, where data is available, there were
+larger difference between 20x and 25x (subsampled to 1000 = 0.88 (0.42 -
+0.91)) {[}Figure 3D{]}. However, these differences between the next 5
+cycle increment do not persist once 25 cylces are reached {[}Figure
+3{]}.
 
-Using PERMANOVA to test whether there were any differences within HiFi
-DNA polymerase groups based on cycle number, only Phusion had cycle
-dependent differences at 1000 and 5000 sub-sampling depth (P-value =
-0.03 and 0.01, respectively). Interestingly, Phusion was one of only two
-DNA polymerase that managed to have samples for the 1000 sub-sampling
-depth at 15 cycles. Next, we assessed whether there were any major
-differences between 5 cycle increments within each sample. We found that
-there was no detectable difference in Bray-Curits index when comparing
-to the previous 5 cycle increment (P-value \textgreater{} 0.05).
-However, Phusion at 1000 sub-sampling depth had a P-value = 0.02 before
-multiple comparison correction. It should be noted that at higher
-sub-sampling depths these differences in Bray-Curits indicies disappear
-for both differences in cycle number and within 5 cycle increments
-within an individual.
+Using PERMANOVA to test for differences within HiFi DNA polymerase
+groups based on cycle number, only Phusion had cycle dependent
+differences at 1000 and 5000 sub-sampling depth (P-value = 0.03 and
+0.01, respectively). Phusion was one of only two HiFi DNA polymerases
+that managed to have samples for the 1000 sub-sampling depth at 15
+cycles. Next, we assessed whether there were any major differences
+between 5 cycle increments within each sample. We found that there was
+no detectable difference in Bray-Curits index when comparing to the
+previous 5 cycle increment (P-value \textgreater{} 0.05). However,
+Phusion at 1000 sub-sampling depth had a P-value = 0.02 before multiple
+comparison correction. It should be noted that at higher sub-sampling
+depths these differences in Bray-Curits indicies disappear for both
+differences in cycle number and within 5 cycle increments within an
+individual.
 
 \textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
-and Cycle Number Used:}} Differences by HiFi DNA polymerase in the
-median average per base error varied without a clear pattern across
-sub-sampling depth {[}Table S5{]}. Generally, the highest values were
-for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera
-removal by VSEARCH had little effect on this rate {[}Figure 4{]}.
-Generally, there were small differences between the various HiFi DNA
-polymerase at lower cycle numbers and larger differences at higher cycle
-number with Platinum having large differences between all other HiFi DNA
-polymerase {[}Figure 4B-C and Table S6{]}.
+and Cycle Number:}} Differences by HiFi DNA polymerase in the median
+average per base error varied without a clear pattern across
+sub-sampling depth {[}Table S5{]}. Generally, the highest per base
+median average error rates were for the Kappa HiFI DNA polymerase
+{[}Figure 4{]}. This error rate was minimally affected by both the
+pre-cluster step and chimera removal by VSEARCH {[}Figure 4{]}. There
+were small differences in the per base error rate between the various
+HiFi DNA polymerase at lower cycle numbers and larger differences at
+higher cycle number with Platinum having the largest differences of all
+the HiFi DNA polymerase {[}Figure 4B-C and Table S6{]}.
 
@@
 significant differences between the HiFi DNA polymerase used (P-value
-\textless{} 0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast
-majority of these differences were driven by Platinum being different
-then all other HiFi DNA polymerase across cycle number and sub-sampling
-depth {[}Table S10{]}. Generally, across sub-sampling depth and cycle
-number Accuprime\textsuperscript{TM} had the lowest chimera prevalence
-of all the HiFi DNA Polymerase regardless of whether pre.cluster or
-VSEARCH had been used {[}Figure 5{]}.
+\textless{} 0.05) {[}Table S9{]}. Differences between Platinum and all
+other HiFi DNA polymerases accounted for the vast majority of these
+differences independent of cycle number and sub-sampling depth when
+using a Dunn's post-hoc test {[}Table S10{]}. Generally, across
+sub-sampling depth and cycle number Accuprime\textsuperscript{TM} had
+the lowest chimera prevalence of all the HiFi DNA Polymerases regardless
+of whether pre.cluster or VSEARCH had been used {[}Figure 5{]}.
 
@@
 number of OTUs and chimeric sequences did not change from the use of
-pre.cluster and VSEARCH {[}Figure 6{]}. Taken together this data
+pre.cluster and VSEARCH {[}Figure 6{]}. Taken together, this data
 suggests that a strong correlation exists between the number of OTUs and
-the prevalence of chimeric sequences. Additionally, Kappa had the
-highest per base error rate and the lowest correlations between the
-number of OTUs and chimeric sequences across sub-sampling depth
-{[}Figure 4 \& 6{]}.
+the prevalence of chimeric sequences.
 

### 2018-01-29 submission/manuscript.Rmd
nick discussion additions

@@
 
-Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA polymerase have significant differences in the number of OTUs and that the changes to total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
+Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
 This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results, as well as the findings.
-Although the attention has mostly been focused on standardizing and improving collection and extraction methods [@contamination_Salter2014] our observations show that independent of this consideration HiFi DNA polymerase can have a noticeable affect on the OTUs generated that can be found across sub-sampling depth and PCR cycle number [Figure 2-4].
+Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4].
 These differences were observed in high biomass samples, where biases introduced by such components like kit contamination may have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
 
-Although we did not observe strong differences, based on cycle number, using the Bray-Curtis index the data suggests that there may be differences between 15 cycles and higher cycle numbers, such as 30x, that are commonly used.
+Although we did not observe strong differences in the Bray-Curtis index, based on cycle number, the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used.
 Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g.
 15 to 20, 20 to 25, etc.).
-Conversely, this may be due to low power and on observation there does seem to be a trend that 20x and 25x communities are very different from each other [Figure 3].
+Conversely, we may not have enough samples to adequately judge the true magnitude difference between 20x and 25x communities and higher cycle numbers analyzed, even though there is a clear trend to suggest that 20x is very different then 25x [Figure 3].
 This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cylces are reached.
@@
 Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5].
-The chimera prevalence was strongly correlated with the number of OTUs and this value is relied upon heavily for many different downstream community metric calculations.
+The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations.
 However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase.
-Since it is possible that many of the increased number of OTUs, generated as cycle number increases, are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
-So, choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in HiFi DNA polymerase.
+It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g.
+Platinum).
 
-Our observations suggest that there are clear HiFi dependent differences in both per base error rate and chimeras that cannot be removed using bioinformatic approaches [Figure 4 & 5].
+Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimeras that cannot be removed using bioinformatic approaches [Figure 4 & 5].
 Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A polymerase [@polymerase_Ishino_2014].
-In fact, from the material safety data sheet (MSDS), it is not clear what the difference between the two different mixes really is.
+In fact, from the material safety data sheet (MSDS), the differences between the two mixes is not immediately apparent.
 Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pycrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate since all samples were also sequenced at the same time as well as amplified using the same machine.
+It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.
 

### 2018-01-29 submission/manuscript.tex
nick discussion additions

@@
 Our observations build upon previous studies (Gohl et al., 2016) by
-showing that even different HiFi DNA polymerase have significant
-differences in the number of OTUs and that the changes to total OTUs
+showing that even different HiFi DNA polymerases have significant
+differences in the number of OTUs and that changes in total OTUs
 correlate with chimeras not removed after sequence processing {[}Figure
@@
 total number of OTUs could drastically change the results, as well as
-the findings. Although the attention has mostly been focused on
-standardizing and improving collection and extraction methods (Salter et
-al., 2014) our observations show that independent of this consideration
-HiFi DNA polymerase can have a noticeable affect on the OTUs generated
-that can be found across sub-sampling depth and PCR cycle number
-{[}Figure 2-4{]}. These differences were observed in high biomass
-samples, where biases introduced by such components like kit
-contamination may have less of an effect, suggesting that these
-differences may be exacerbated in low biomass samples.
+the findings. Our observations show that HiFi DNA polymerase can have a
+noticeable affect on the OTUs generated and these differences are
+consistent across sub-sampling depth and PCR cycle number {[}Figure
+2-4{]}. These differences were observed in high biomass samples, where
+biases introduced by such components like kit contamination may have
+less of an effect, suggesting that these differences may be exacerbated
+in low biomass samples.
 
-Although we did not observe strong differences, based on cycle number,
-using the Bray-Curtis index the data suggests that there may be
-differences between 15 cycles and higher cycle numbers, such as 30x,
-that are commonly used. Additionally, there was no difference within
+Although we did not observe strong differences in the Bray-Curtis index,
+based on cycle number, the data suggests that there may be differences
+between 15 and 20 cycles versus higher cycle numbers, such as 30x, that
+are commonly used. Additionally, there was no difference within
 individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
-to 25, etc.). Conversely, this may be due to low power and on
-observation there does seem to be a trend that 20x and 25x communities
-are very different from each other {[}Figure 3{]}. This finding, in
-conjunction with the PERMANOVA results, suggest that cycle number can
-change bacterial community calculations but that these differences are
-minimal once 25 cylces are reached. Increasing the sub-sampling depth,
-for some DNA polymerase, may reduce some of these observed community
-differences at lower cycle numbers.
+to 25, etc.). Conversely, we may not have enough samples to adequately
+judge the true magnitude difference between 20x and 25x communities and
+higher cycle numbers analyzed, even though there is a clear trend to
+suggest that 20x is very different then 25x {[}Figure 3{]}. This
+finding, in conjunction with the PERMANOVA results, suggest that cycle
+number can change bacterial community calculations but that these
+differences are minimal once 25 cylces are reached. Increasing the
+sub-sampling depth, for some DNA polymerase, may reduce some of these
+observed community differences at lower cycle numbers.
 
@@
 The chimera prevalence was strongly correlated with the number of OTUs
-and this value is relied upon heavily for many different downstream
-community metric calculations. However, Bray-Curtis analysis with
-PERMANOVA showed few differences based on DNA polymerase. Since it is
-possible that many of the increased number of OTUs, generated as cycle
-number increases, are not highly abundant allowing the Bray-Curtis index
-to be able to successfully downweight these respective OTUs (Minchin,
-1987). So, choice of downstream diversity metric could be an important
+which is relied upon heavily for many community metric calculations.
+However, Bray-Curtis analysis with PERMANOVA showed few differences
+based on DNA polymerase. It is possible that many of the increased
+number of OTUs are not highly abundant allowing the Bray-Curtis index to
+be able to successfully downweight these respective OTUs (Minchin,
+1987). The choice of downstream diversity metric could be an important
 consideration in helping to mitigate these observed changes due to high
-chimera prevalence in HiFi DNA polymerase.
+chimera prevalence in specific HiFi DNA polymerases (e.g.~Platinum).
 
-Our observations suggest that there are clear HiFi dependent differences
-in both per base error rate and chimeras that cannot be removed using
-bioinformatic approaches {[}Figure 4 \& 5{]}. Although it may be a
-natural assumption that the variation may be due to the DNA polymerase
-family, the highest chimera rate, from Platinum, was a family A
-polymerase while the lowest, from Accuprime, was also an A polymerase
-(Ishino \& Ishino, 2014). In fact, from the material safety data sheet
-(MSDS), it is not clear what the difference between the two different
-mixes really is. Both Accuprime and Platinum contain a recombinant
+Our observations suggest that there are clear HiFi DNA polymerase
+dependent differences in both per base error rate and chimeras that
+cannot be removed using bioinformatic approaches {[}Figure 4 \& 5{]}.
+Although it may be a natural assumption that the variation may be due to
+the DNA polymerase family, the highest chimera rate, from Platinum, was
+a family A polymerase while the lowest, from Accuprime, was also an A
+polymerase (Ishino \& Ishino, 2014). In fact, from the material safety
+data sheet (MSDS), the differences between the two mixes is not
+immediately apparent. Both Accuprime and Platinum contain a recombinant
 \emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D polymerase and
@@
 the recombinant \emph{Taq} was generated could be the main reason for
-the differences in chimera rate since all samples were also sequenced at
-the same time as well as amplified using the same machine.
+the differences in chimera rate.
 

### 2018-01-29 submission/manuscript.Rmd
add nick conclusion suggestions

@@
 
-Our findings show that measures that rely on number of OTUs will be specific for a particular study and may not be easily generalized to other studies that may be studying a similar area.
+Our findings show that measures that rely on number of OTUs will be specific for a particular study and may not be easily generalized to other studies investigating a similar area.
 Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed as well as potentially influence diversity based metrics that do not down weight rare observations.

### 2018-01-29 submission/manuscript.tex
add nick conclusion suggestions

@@
 specific for a particular study and may not be easily generalized to
-other studies that may be studying a similar area. Care should be taken
-when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since
-intrinsic differences can change the number of OTUs observed as well as
+other studies investigating a similar area. Care should be taken when
+choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic
+differences can change the number of OTUs observed as well as
 potentially influence diversity based metrics that do not down weight

### 2018-01-30 code/make_fecal_numOTU_graphs.R
change overall figure --simplify the message

@@ -37,141 +37,35 @@ sub_sample_level <- c("1000", "5000", "10000", "15000", "20000")
 
 # Read in the count data
 numOTU_data <- sapply(sub_sample_level, 
-                      function(x) read_data("data/process/tables/", "fecal_zscore_sub_sample_", "_count_table.csv", x), 
-                      simplify = F)
-
-
-thousand_graph <- numOTU_data[["1000"]] %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")), 
-         cycle_num = factor(cycle_num, 
-                         levels = c(15, 20, 25, 30, 35), 
-                         labels = c("15x", "20x", "25x", "30x", "35x"))) %>% 
-  ggplot(aes(cycle_num, scaled_numOTU, color = taq, group = taq)) + 
-  geom_point(size = 2, alpha = 0.7, show.legend = T) + theme_bw() + 
-  geom_vline(xintercept = c(1.5, 2.5, 3.5), color = "gray") + 
-  geom_hline(yintercept = 0, linetype = "dashed") + 
-  facet_grid(. ~ sample_name) + 
-  scale_color_manual(name = "Taq Used", 
-                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = "Z-Score Normalized Number of OTUs") + 
-  ggtitle("A") + coord_cartesian(ylim = c(-2.0, 2.0)) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10), 
-        legend.position = "bottom",  
-        legend.title = element_blank(), 
-        legend.key = element_blank(), 
-        legend.background = element_rect(color = "black"))
-
-
-five_thousand_graph <- numOTU_data[["5000"]] %>% 
+                      function(x) read_data("data/process/tables/", "fecal_zscore_sub_sample_", "_count_table.csv", x) %>% 
+                        mutate(depth_level = as.numeric(x)), 
+                      simplify = F) %>% bind_rows() %>% 
[…]

### 2018-01-30 code/make_fecal_numOTU_graphs.R
complete code update

@@ -49,23 +49,27 @@ numOTU_data <- sapply(sub_sample_level,
   
 
 summary_data <- numOTU_data %>% group_by(taq) %>% 
-  summarise(taq_median = median(scaled_numOTU, na.rm = T))
+  summarise(taq_mean = mean(scaled_numOTU, na.rm = T))
 
 fecal_samples <- numOTU_data %>% 
   ggplot(aes(depth_level, scaled_numOTU, color = cycle_num, group = taq)) + 
   geom_point(size = 2, alpha = 0.9, show.legend = T) + theme_bw() + 
   geom_hline(yintercept = 0, linetype = "dashed", size = 1) + 
-  geom_hline(aes(yintercept = taq_median), summary_data, color = "red", size = 1) + 
+  geom_hline(aes(yintercept = taq_mean), summary_data, color = "red", size = 1) + 
   facet_grid(. ~ taq) + 
   labs(x = "Sub-sampling Depth", y = "Z-Score Normalized Number of OTUs") + 
   scale_color_manual(name = "Cycle Number", 
                      values = c("#006400", "#00FF7F", "#63B8FF", "#104E8B")) + 
+  scale_x_continuous(breaks = c(1000, 5000, 10000, 15000, 20000), 
+                    labels = c("1000", "5000", "10000", "15000", "20000")) + 
   theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10), 
+        axis.text.y = element_text(size = 10),
+        axis.text.x = element_text(size = 8, angle = 70, hjust = 1), 
         legend.key = element_blank(), 
+        legend.position = "bottom", 
         legend.background = element_rect(color = "black"))
   
 
-ggsave("results/figures/Figure1.pdf", fecal_samples, width = 10, height = 5, dpi = 300)
+ggsave("results/figures/Figure1.pdf", fecal_samples, width = 7, height = 5, dpi = 300)

### 2018-01-30 submission/manuscript.Rmd
update with new fig1

@@
 In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data.
-After normalization, we identified that there was a HiFi DNA polymerase dependent difference that became more apparent as the depth of sub-sampling increased [Figure 1].
-Lower cycle numbers (15-20) resulted in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 having distinct differences [Figure 1].
+After normalization, we identified that there was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1].
+Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1].
 All subsampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
@@
 
-**Figure 1: Normalized Fecal Number of OTUs.** A) Sub-sampled to 1000 reads.
-B) Sub-sampled to 5000 reads.
-C) Sub-sampled to 10000 reads.
-D) Sub-sampled to 15000 reads.
-The dotted line represents no change from the mean number of OTUs within that specific individual.
+**Figure 1: Normalized Fecal Number of OTUs.** The x-axis represents the different sub-sampling depths used and the y-axis is the normalized within individual number of OTUs.
+The red line represents the overall mean Z-score normalized number of OTUs for each respective HiFi DNA polymerase.
+The dashed black line represents the overall Z-score normalized mean number of OTUs.
 

### 2018-01-30 submission/manuscript.tex
update with new fig1

@@
 the number of OTUs data. After normalization, we identified that there
-was a HiFi DNA polymerase dependent difference that became more apparent
-as the depth of sub-sampling increased {[}Figure 1{]}. Lower cycle
-numbers (15-20) resulted in less differences between HiFI DNA polymerase
-while cycle numbers of 25, 30, and 35 having distinct differences
+was a HiFi DNA polymerase dependent difference that was consistent
+across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
+to result in less differences between HiFI DNA polymerase while cycle
+numbers of 25, 30, and 35 tended to have more distinct differences
 {[}Figure 1{]}. All subsampling levels were significantly different only
@@
 
-\textbf{Figure 1: Normalized Fecal Number of OTUs.} A) Sub-sampled to
-1000 reads. B) Sub-sampled to 5000 reads. C) Sub-sampled to 10000 reads.
-D) Sub-sampled to 15000 reads. The dotted line represents no change from
-the mean number of OTUs within that specific individual.
+\textbf{Figure 1: Normalized Fecal Number of OTUs.} The x-axis
+represents the different sub-sampling depths used and the y-axis is the
+normalized within individual number of OTUs. The red line represents the
+overall mean Z-score normalized number of OTUs for each respective HiFi
+DNA polymerase. The dashed black line represents the overall Z-score
+normalized mean number of OTUs.
 

### 2018-01-30 code/make_numOTU_graphs.R
update with nick suggestions

@@ -37,27 +37,31 @@ sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
 
 # Read in the count data
 numOTU_data <- sapply(sub_sample_level, 
-                          function(x) read_data("data/process/tables/", "mock_sub_sample_", "_count_table.csv", x), 
-                          simplify = F)
+                          function(x) 
+                            read_data("data/process/tables/", 
+                                      "mock_sub_sample_", "_count_table.csv", x) %>% 
+                        mutate(depth_level = as.numeric(x)), 
+                          simplify = F) %>% bind_rows() %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
+  
 
 ref_otu_num <- read_tsv("data/process/zymo_mock_only.shared") %>% 
   select(numOtus)
 
 # Generate graph of Mock DNA samples 
-thousand_graph <- numOTU_data[["1000"]] %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
+thousand_graph <- numOTU_data %>% filter(depth_level == 1000) %>% 
   ggplot(aes(cycles, log2(numOTUs), color = taq, group = taq)) + 
-  geom_smooth(size = 1, method = "lm", formula = y ~ poly(x, 2), se = FALSE) + 
-  geom_point(size = 2, alpha = 0.7) + 
-  geom_hline(yintercept = log2(ref_otu_num$numOtus), linetype = "dashed") + 
+  geom_smooth(size = 1, method = "lm", formula = y ~ poly(x, 2), se = FALSE, show.legend = F) + 
+  geom_point(size = 2, alpha = 0.7, show.legend = F) + 
+  geom_hline(yintercept = log2(ref_otu_num$numOtus), linetype = "dashed", size = 1) + 
   theme_bw() + 
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
[…]

### 2018-01-30 code/make_bray_distance_graphs.R
simplify message of graph

@@ -38,203 +38,82 @@ sub_sample_level <- c("1000", "5000", "10000")
 
 dist_data <- sapply(sub_sample_level, 
                     function(x) read_csv(paste("data/process/tables/bray_5_cycle_dist_", 
-                                               x, "_data.csv", sep = "")), simplify = F)
+                                               x, "_data.csv", sep = "")) %>% 
+                      mutate(depth_level = as.numeric(x)), simplify = F) %>% bind_rows() %>% 
+  filter(taq != "K") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Phusion", "Platinum", "Q5")), 
+         cycle_compare = factor(cycle_compare, 
+                                levels = c("15to20", "20to25", "25to30", "30to35"), 
+                                labels = c("15x vs 20x", "20x vs 25x", 
+                                           "25x vs 30x", "30x vs 35x")))
+  
 
 mock_dist_data <- sapply(sub_sample_level, 
                     function(x) read_csv(paste("data/process/tables/mock_bray_5_cycle_dist_", 
-                                               x, "_data.csv", sep = "")), simplify = F)
+                                               x, "_data.csv", sep = "")) %>% 
+                      mutate(depth_level = as.numeric(x)), simplify = F) %>% bind_rows() %>% 
+  filter(taq != "K") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Phusion", "Platinum", "Q5")), 
+         cycle_compare = factor(cycle_compare, 
+                                levels = c("15to20", "20to25", "25to30", "30to35"), 
+                                labels = c("15x vs 20x", "20x vs 25x", 
+                                           "25x vs 30x", "30x vs 35x")))
 
 ann_text <- data.frame(cycle_compare = 2.2, distance = 0.95, lab = "Sub-Sampled", 
                        taq = factor("ACC", levels = c("ACC", "PHU", "PL", "Q5"), 
[…]

### 2018-01-30 submission/manuscript.Rmd
update figure3 in manuscirpt

@@
 First, there were differences for the same fecal sample between 20x vs.
-25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (subsampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), subsampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), subsampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A-B].
-Second, for the mocks, where data is available, there were larger difference between 20x and 25x (subsampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3D].
+25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (subsampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), subsampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), subsampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Second, for the mocks, where data is available, there were larger difference between 20x and 25x (subsampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
 However, these differences between the next 5 cycle increment do not persist once 25 cylces are reached [Figure 3].
@@
 
-**Figure 3: Five Cycle Interval Community Differences.** A) Fecal samples sub-sampled to 1000 reads.
-B) Fecal samples sub-sampled to 5000 reads.
-C) Fecal samples sub-sampled to 10000 reads.
-D) Mock samples sub-sampled to 1000 reads.
-E) Mock samples sub-sampled to 5000 reads.
-F) Mock samples sub-sampled to 10000 reads.
-The solid black lines represent the median Bray-Curtis index difference within sample for each 5 cycle interval.
+**Figure 3: Community Differences by Five-Cycle Intervals and Sub-sampling Depth.** A) Fecal samples within person difference based on the next 5-cycle PCR interval.
+B) Mock samples within replicate difference based on the next 5-cycle PCR interval.
 

### 2018-01-30 submission/manuscript.tex
update figure3 in manuscirpt

@@
 0.43 (0.33 - 0.63), subsampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
-3A-B{]}. Second, for the mocks, where data is available, there were
-larger difference between 20x and 25x (subsampled to 1000 = 0.88 (0.42 -
-0.91)) {[}Figure 3D{]}. However, these differences between the next 5
-cycle increment do not persist once 25 cylces are reached {[}Figure
-3{]}.
+3A{]}. Second, for the mocks, where data is available, there were larger
+difference between 20x and 25x (subsampled to 1000 = 0.88 (0.42 - 0.91))
+{[}Figure 3B{]}. However, these differences between the next 5 cycle
+increment do not persist once 25 cylces are reached {[}Figure 3{]}.
 
@@
 
-\textbf{Figure 3: Five Cycle Interval Community Differences.} A) Fecal
-samples sub-sampled to 1000 reads. B) Fecal samples sub-sampled to 5000
-reads. C) Fecal samples sub-sampled to 10000 reads. D) Mock samples
-sub-sampled to 1000 reads. E) Mock samples sub-sampled to 5000 reads. F)
-Mock samples sub-sampled to 10000 reads. The solid black lines represent
-the median Bray-Curtis index difference within sample for each 5 cycle
-interval.
+\textbf{Figure 3: Community Differences by Five-Cycle Intervals and
+Sub-sampling Depth.} A) Fecal samples within person difference based on
+the next 5-cycle PCR interval. B) Mock samples within replicate
+difference based on the next 5-cycle PCR interval.
 

### 2018-02-01 submission/manuscript.Rmd
abstract changes

@@
 
-**Background.** Increasing research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys.
-These components include, but are not necessarily limited to, preservation media, extraction kit, bead beating time, and primers.
+**Background.** An increasing body of research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys.
+These components include, but are not limited to, preservation media, extraction kit, bead beating time, and primers.
 Both cycle number and high fidelity (HiFi) DNA polymerase are sometimes overlooked when sources of bias are being considered.
@@
 
-**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
 Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach.
@@
 Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH.
-Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and was also not changed by chimera removal with VSEARCH.
+Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and this association was not changed by chimera removal with VSEARCH.
 
 **Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras.
-So, when designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.
+When designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.
 

### 2018-02-01 submission/manuscript.tex
abstract changes

@@
 
-\textbf{Background.} Increasing research has found that various
-methodological steps can have an impact on the observed microbial
-community when using 16S rRNA gene surveys. These components include,
-but are not necessarily limited to, preservation media, extraction kit,
+\textbf{Background.} An increasing body of research has found that
+various methodological steps can have an impact on the observed
+microbial community when using 16S rRNA gene surveys. These components
+include, but are not limited to, preservation media, extraction kit,
 bead beating time, and primers. Both cycle number and high fidelity
@@
 single PowerMag DNA extraction kit with a 10 minute bead beating step
-and amplified at 15x, 20x, 25x, 30x, and 35x using Accuprime, Kappa,
+and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
 Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
@@
 Additionally, the chimera prevalence had a strong positive correlation
-with the number of OTUs observed in a sample and was also not changed by
-chimera removal with VSEARCH.
+with the number of OTUs observed in a sample and this association was
+not changed by chimera removal with VSEARCH.
 
@@
 could give artificially inflated numbers due to higher undetected
-chimeras. So, when designing 16S rRNA gene survey studies it is
-important to consider both the cycle number and the type of HiFi DNA
-polymerase that will be used since it can increase or decrease the
-number of OTUs that are observed.
+chimeras. When designing 16S rRNA gene survey studies it is important to
+consider both the cycle number and the type of HiFi DNA polymerase that
+will be used since it can increase or decrease the number of OTUs that
+are observed.
 

### 2018-02-01 submission/manuscript.Rmd
run spell-check

@@
 Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach.
-First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communites.
+First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communities.
 Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase.
@@
 
-Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
-In particular, investigating ways that 16S rRNA gene surveys can be made more replicatable has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
+Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
+In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
 Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
@@
 ***PCR Protocol:***
-The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The five different high fidelity (HiFi) DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
@@
 The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009].
-Generally, the protocol followed what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Generally, the protocol followed what has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
 Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
@@
 All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
+Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
@@
 The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
+Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
 20 versus 25, 25 versus 30, etc.).
 Based on these observations we wanted to next analyze potential reasons for these differences.
-First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase.
+First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase.
 After assessing these errors, the total number of chimeras was determined after sequence processing.
@@
 ***Reproducible Methods:***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017.
-The raw sequences can be found in the SRA at the following accesssion number **need to upload still**.
+The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number **need to upload still**.
 
@@
 Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1].
-All subsampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+All sub-sampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
 There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
@@
 First, there were differences for the same fecal sample between 20x vs.
-25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (subsampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), subsampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), subsampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Second, for the mocks, where data is available, there were larger difference between 20x and 25x (subsampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
-However, these differences between the next 5 cycle increment do not persist once 25 cylces are reached [Figure 3].
+25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Second, for the mocks, where data is available, there were larger difference between 20x and 25x (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
+However, these differences between the next 5 cycle increment do not persist once 25 cycles are reached [Figure 3].
 
@@
 Next, we assessed whether there were any major differences between 5 cycle increments within each sample.
-We found that there was no detectable difference in Bray-Curits index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+We found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
 However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
-It should be noted that at higher sub-sampling depths these differences in Bray-Curits indicies disappear for both differences in cycle number and within 5 cycle increments within an individual.
+It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5 cycle increments within an individual.
 
@@
 
-The total seqeunces with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1].
+The total sequences with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1].
 Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8].
@@
 However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1].
-Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there were generally no biases in the types of subsitution made [Figure S2].
+Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were generally no biases in the types of substitution made [Figure S2].
 
@@
 Conversely, we may not have enough samples to adequately judge the true magnitude difference between 20x and 25x communities and higher cycle numbers analyzed, even though there is a clear trend to suggest that 20x is very different then 25x [Figure 3].
-This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cylces are reached.
+This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
 Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.
@@
 However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase.
-It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987].
+It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987].
 The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g.
@@
 We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
-Salary support for Marc Sze came from the Canadian Institue of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.
+Salary support for Marc Sze came from the Canadian Institute of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.
 
@@
 C) Sub-sampled to 10000 reads.
-The dotted line represents the number of OTUs genereated when the mock reference sequences are run through the pipeline.
+The dotted line represents the number of OTUs generated when the mock reference sequences are run through the pipeline.
 

### 2018-02-01 submission/manuscript.tex
run spell-check

@@
 amplification approach. First, the number of OTUs (Operational Taxonomic
-Units) was examined for both fecal samples and mock communites. Next,
+Units) was examined for both fecal samples and mock communities. Next,
 Bray-Curtis index, the error rate, sequence error prevalence, and
@@
 
-Over recent years there has been an increasing focus on standardizing
+Recently there has been an increasing focus on standardizing
 methodological approaches in microbiota research (Kim et al., 2017;
 Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
-rRNA gene surveys can be made more replicatable has been a predominant
+rRNA gene surveys can be made more reproducible has been a predominant
 driver of this standardization push (Lauber et al., 2010; Salter et al.,
@@
 \textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
-Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
+DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
 (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
@@
 utilized for all sequence processing steps (Schloss et al., 2009).
-Generally, the protocol followed what has been previously publsihed
+Generally, the protocol followed what has been previously published
 (Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
@@
 instead of UCHIME for chimera detection and the use of the OptiClust
-algorithmn instead of average neighbor for Operational Taxonomic Unit
+algorithm instead of average neighbor for Operational Taxonomic Unit
 (OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
@@
 \textbf{\emph{Statistical Analysis:}} All analysis was done with the R
-(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
+(v 3.4.3) software package (R Core Team, 2017). Data transformation and
 graphing was completed using the tidyverse package (v 1.2.1) and colors
@@
 sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curits indicies were next assessed for
+Cycle dependent affects on Bray-Curtis indices were next assessed for
 the fecal samples looking at both overall cycle differences and within
@@
 sequence error rate, number of sequences with an error, and base
-subsitution were assessed in the mock community for each DNA polymerase.
-After assessing these errors, the total number of chimeras was
-determined after sequence processing. For the community based measures,
-the fecal samples were analyzed at 4 different sub-sampling levels
-(1000, 5000, 10000, and 15000) while the mock community samples were
-analysed at 3 levels (1000, 5000, 10000).
+substitution were assessed in the mock community for each DNA
+polymerase. After assessing these errors, the total number of chimeras
+was determined after sequence processing. For the community based
+measures, the fecal samples were analyzed at 4 different sub-sampling
+levels (1000, 5000, 10000, and 15000) while the mock community samples
+were analysed at 3 levels (1000, 5000, 10000).
 
@@
 here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
-The raw sequences can be found in the SRA at the following accesssion
+The raw sequences can be found in the SRA at the following accession
 number \textbf{need to upload still}.
@@
 numbers of 25, 30, and 35 tended to have more distinct differences
-{[}Figure 1{]}. All subsampling levels were significantly different only
-at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There were
-differences in HiFi DNA polymerase at 25 and 30 cycles but the
+{[}Figure 1{]}. All sub-sampling levels were significantly different
+only at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There
+were differences in HiFi DNA polymerase at 25 and 30 cycles but the
 sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
@@
 same fecal sample between 20x vs.~25x that was independent of HiFi DNA
-polymerase but dependent on sub-sampling depth (subsampled to 1000 =
-0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), subsampled to 5000 =
-0.43 (0.33 - 0.63), subsampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
+polymerase but dependent on sub-sampling depth (sub-sampled to 1000 =
+0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), sub-sampled to 5000 =
+0.43 (0.33 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
 3A{]}. Second, for the mocks, where data is available, there were larger
-difference between 20x and 25x (subsampled to 1000 = 0.88 (0.42 - 0.91))
-{[}Figure 3B{]}. However, these differences between the next 5 cycle
-increment do not persist once 25 cylces are reached {[}Figure 3{]}.
+difference between 20x and 25x (sub-sampled to 1000 = 0.88 (0.42 -
+0.91)) {[}Figure 3B{]}. However, these differences between the next 5
+cycle increment do not persist once 25 cycles are reached {[}Figure
+3{]}.
 
@@
 between 5 cycle increments within each sample. We found that there was
-no detectable difference in Bray-Curits index when comparing to the
+no detectable difference in Bray-Curtis index when comparing to the
 previous 5 cycle increment (P-value \textgreater{} 0.05). However,
@@
 comparison correction. It should be noted that at higher sub-sampling
-depths these differences in Bray-Curits indicies disappear for both
+depths these differences in Bray-Curtis indices disappear for both
 differences in cycle number and within 5 cycle increments within an
@@
 
-The total seqeunces with at least one error had multiple differences at
+The total sequences with at least one error had multiple differences at
 different cycle numbers and was mostly alleviated by the use of the
@@
 whether there were HiFi DNA polymerase dependent effects on base
-subsitution found that there were generally no biases in the types of
-subsitution made {[}Figure S2{]}.
+substitution found that there were generally no biases in the types of
+substitution made {[}Figure S2{]}.
 
@@
 number can change bacterial community calculations but that these
-differences are minimal once 25 cylces are reached. Increasing the
+differences are minimal once 25 cycles are reached. Increasing the
 sub-sampling depth, for some DNA polymerase, may reduce some of these
@@
 number of OTUs are not highly abundant allowing the Bray-Curtis index to
-be able to successfully downweight these respective OTUs (Minchin,
+be able to successfully down-weight these respective OTUs (Minchin,
 1987). The choice of downstream diversity metric could be an important
@@
 Microbiome Core Facility at the University of Michigan. Salary support
-for Marc Sze came from the Canadian Institue of Health Research and the
+for Marc Sze came from the Canadian Institute of Health Research and the
 Michigan Institute for Clinical and Health Research Postdoctoral
@@
 5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents
-the number of OTUs genereated when the mock reference sequences are run
+the number of OTUs generated when the mock reference sequences are run
 through the pipeline.

### 2018-02-01 submission/manuscript.Rmd
intro modifications

@@
 In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
-Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
-Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
+Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
+Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
 Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
@@
 A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
-Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
+Another important component of this study was that the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline regardless of DNA polymerase used [@taq_Gohl_2016].
 Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
@@
 
-We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase.
+We amplified the V4 region of the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase.
 First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples.

### 2018-02-01 submission/manuscript.tex
intro modifications

@@
 2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
-sequencing has been much maligned for introduced bias, many of these
+sequencing has been much maligned for introduced biases, many of these
 same considerations also affect metagenomic sequencing (Nayfach \&
 Pollard, 2016; Costea et al., 2017). Between the two approaches similar
-bias considerations include, but are not limited to, preservation media,
-storage conditions, DNA extraction kit, PCR, and sequence library
-preparation. Thus, for these overlapping considerations biases
-identified for 16S rRNA gene sequencing will also likely influence
-metagenomic sequencing results.
+bias considerations include preservation media, storage conditions, DNA
+extraction kit, PCR, and sequence library preparation. Thus, for these
+overlapping considerations biases identified for 16S rRNA gene
+sequencing will also likely influence metagenomic sequencing results.
 
@@
 PCR protocol (Gohl et al., 2016). Another important component of this
-study was that differences, based on DNA polymerase, in the number of
-OTUs generated were not easily removed using the authors chosen
-bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
+study was that the number of OTUs generated were not easily removed
+using the authors chosen bioinformatic pipeline regardless of DNA
+polymerase used (Gohl et al., 2016). Although it is probably not
 surprising that normal DNA polymerase performed worse than HiFi DNA
@@
 
-We amplified the 16S rRNA gene in both fecal and mock community samples
-using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
-polymerase. First, we tested if we could identify differences in the
-number of OTUs between the different HiFi DNA polymerase for both the
-fecal and mock samples. Next, since there were differences based on HiFi
-DNA polymerase, we examined if a community measure such as the
-Bray-Curtis index would be affected. After assessing these community
-measures we then examined if the different HiFi DNA polymerase had
-different per base error rates as well as chimera prevalence. Finally,
-since differences in chimera prevalence, based on HiFi DNA polymerase,
-could not be completely removed using bioinformatic approaches we tested
-whether the chimera prevalence correlated well with total number of
-observed OTUs.
+We amplified the V4 region of the 16S rRNA gene in both fecal and mock
+community samples using either Accuprime, Kappa, Q5, Phusion, or
+Platinum HiFi DNA polymerase. First, we tested if we could identify
+differences in the number of OTUs between the different HiFi DNA
+polymerase for both the fecal and mock samples. Next, since there were
+differences based on HiFi DNA polymerase, we examined if a community
+measure such as the Bray-Curtis index would be affected. After assessing
+these community measures we then examined if the different HiFi DNA
+polymerase had different per base error rates as well as chimera
+prevalence. Finally, since differences in chimera prevalence, based on
+HiFi DNA polymerase, could not be completely removed using bioinformatic
+approaches we tested whether the chimera prevalence correlated well with
+total number of observed OTUs.
 

### 2018-02-01 submission/manuscript.tex
remove what the taq component

@@
 
-\section{What the Taq? The Influence of High Fidelity DNA Polymerase on
-16S rRNA Gene
-Sequencing}\label{what-the-taq-the-influence-of-high-fidelity-dna-polymerase-on-16s-rrna-gene-sequencing}
+\section{The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene
+Sequencing}\label{the-influence-of-high-fidelity-dna-polymerase-on-16s-rrna-gene-sequencing}
 

### 2018-02-01 submission/manuscript.Rmd
remove what the taq component
methods edits

@@
 
-# What the Taq? The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene Sequencing
+# The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene Sequencing
 
@@
 Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+Sequence error was determined using the 'seq.error' command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
 
@@
 For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
-Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
-The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R.
+Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth.
+The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
 For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
@@
 After assessing these errors, the total number of chimeras was determined after sequence processing.
-For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
+For the community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
 

### 2018-02-01 submission/manuscript.tex
methods edits

@@
 2016; Westcott \& Schloss, 2017). Sequence error was determined using
-the seq.error command on mock samples after chimera removal and
+the `seq.error' command on mock samples after chimera removal and
 classification to the RDP to remove non-bacterial sequences (Schloss et
@@
 biological variation between people. Bray-Curtis matrices were generated
-using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
-The distance matrix data was analyzed using PERMANOVA with the vegan
-package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests within
-R. For both error and chimera analysis, samples were tested using
-Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
-for multiple comparison utilized the Benjamini-Hochberg method
-(Benjamini \& Hochberg, 1995).
+using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000
+sequence depth. The distance matrix data was analyzed using PERMANOVA
+with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
+Kruskal-Wallis tests within R. For both error and chimera analysis,
+samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
+Where applicable correction for multiple comparison utilized the
+Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
 
@@
 measures, the fecal samples were analyzed at 4 different sub-sampling
-levels (1000, 5000, 10000, and 15000) while the mock community samples
-were analysed at 3 levels (1000, 5000, 10000).
+sequence depths (1000, 5000, 10000, and 15000) while the mock community
+samples were analysed at 3 levels (1000, 5000, 10000).
 

### 2018-02-01 submission/manuscript.Rmd
result section edits

@@
 
-***Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
+***The Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
 In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data.
@@
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
-Overall, there was very little difference between each respective 5-cycle increment (e.g.
-15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3].
+Overall, there was little difference between each respective 5-cycle increment at higher cycle numbers, for both fecal and mock samples, and this was consistent across the different sub-samplings [Figure 3].
 There were only two differences between 5-cycle increments that were identified.
-First, there were differences for the same fecal sample between 20x vs.
-25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Second, for the mocks, where data is available, there were larger difference between 20x and 25x (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
-However, these differences between the next 5 cycle increment do not persist once 25 cycles are reached [Figure 3].
+First, there were differences for the same fecal sample between 20 versus 25 cycles that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Second, for the mocks, where data is available, there were larger difference between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
+However, these differences between the next 5-cycle increment do not persist once 25 cycles are reached [Figure 3].
 
@@
 Phusion was one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
-Next, we assessed whether there were any major differences between 5 cycle increments within each sample.
-We found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+Next, we assessed whether there were any major differences between 5-cycle increments within each sample.
+We found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
 However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
-It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5 cycle increments within an individual.
+It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5-cycle increments within an individual.
 
@@
 Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest per base median average error rates were for the Kappa HiFI DNA polymerase [Figure 4].
-This error rate was minimally affected by both the pre-cluster step and chimera removal by VSEARCH [Figure 4].
+Generally, the highest per base median average error rates were for the Kappa HiFi DNA polymerase [Figure 4].
+This error rate was minimally affected by both the 'pre.cluster' step and chimera removal by VSEARCH [Figure 4].
 There were small differences in the per base error rate between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having the largest differences of all the HiFi DNA polymerase [Figure 4B-C and Table S6].
 
-The total sequences with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1].
-Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8].
+The total sequences with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the 'pre.cluster' step [Figure S1].
+Major differences before this 'pre.cluster' step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8].
 Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1].
@@
 
-***Chimeric Sequences Correlate with OTUs and are HiFi DNA Polymerase Dependent:***
-After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria we assessed the percentage of sequences that were still chimeric within our mock community.
+***Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with Number of OTUs:***
+After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria, we assessed the percentage of sequences that were still chimeric within our mock community.
 At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
 Differences between Platinum and all other HiFi DNA polymerases accounted for the vast majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
-Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether pre.cluster or VSEARCH had been used [Figure 5].
+Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH had been used [Figure 5].
 
 For all DNA polymerase, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
-In general, the R^2^ value between the number of OTUs and chimeric sequences did not change from the use of pre.cluster and VSEARCH [Figure 6].
+The R^2^ value between the number of OTUs and chimeric sequences did not change from the use of 'pre.cluster' and the removal of chimeras with VSEARCH [Figure 6].
 Taken together, this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.

### 2018-02-01 submission/manuscript.tex
result section edits

@@
 
-\textbf{\emph{Number of OTUs is Dependent on the HiFi DNA Polymerase
+\textbf{\emph{The Number of OTUs is Dependent on the HiFi DNA Polymerase
 Used:}} In order to compare the number of OTUs across individuals a
@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} Overall, there was very little difference between each
-respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
-samples and this was consistent across the different sub-samplings used
+Number:}} Overall, there was little difference between each respective
+5-cycle increment at higher cycle numbers, for both fecal and mock
+samples, and this was consistent across the different sub-samplings
 {[}Figure 3{]}. There were only two differences between 5-cycle
 increments that were identified. First, there were differences for the
-same fecal sample between 20x vs.~25x that was independent of HiFi DNA
-polymerase but dependent on sub-sampling depth (sub-sampled to 1000 =
-0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), sub-sampled to 5000 =
-0.43 (0.33 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
-3A{]}. Second, for the mocks, where data is available, there were larger
-difference between 20x and 25x (sub-sampled to 1000 = 0.88 (0.42 -
-0.91)) {[}Figure 3B{]}. However, these differences between the next 5
-cycle increment do not persist once 25 cycles are reached {[}Figure
-3{]}.
+same fecal sample between 20 versus 25 cycles that was independent of
+HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to
+1000 = 0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), sub-sampled to
+5000 = 0.43 (0.33 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43))
+{[}Figure 3A{]}. Second, for the mocks, where data is available, there
+were larger difference between 20 and 25 cycles (sub-sampled to 1000 =
+0.88 (0.42 - 0.91)) {[}Figure 3B{]}. However, these differences between
+the next 5-cycle increment do not persist once 25 cycles are reached
+{[}Figure 3{]}.
 
@@
 cycles. Next, we assessed whether there were any major differences
-between 5 cycle increments within each sample. We found that there was
+between 5-cycle increments within each sample. We found that there was
 no detectable difference in Bray-Curtis index when comparing to the
-previous 5 cycle increment (P-value \textgreater{} 0.05). However,
+previous 5-cycle increment (P-value \textgreater{} 0.05). However,
 Phusion at 1000 sub-sampling depth had a P-value = 0.02 before multiple
@@
 depths these differences in Bray-Curtis indices disappear for both
-differences in cycle number and within 5 cycle increments within an
+differences in cycle number and within 5-cycle increments within an
 individual.
@@
 sub-sampling depth {[}Table S5{]}. Generally, the highest per base
-median average error rates were for the Kappa HiFI DNA polymerase
+median average error rates were for the Kappa HiFi DNA polymerase
 {[}Figure 4{]}. This error rate was minimally affected by both the
-pre-cluster step and chimera removal by VSEARCH {[}Figure 4{]}. There
+`pre.cluster' step and chimera removal by VSEARCH {[}Figure 4{]}. There
 were small differences in the per base error rate between the various
@@
 different cycle numbers and was mostly alleviated by the use of the
-pre.cluster step {[}Figure S1{]}. Major differences before this
-pre.cluster step were driven by large differences in
+`pre.cluster' step {[}Figure S1{]}. Major differences before this
+`pre.cluster' step were driven by large differences in
 Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
@@
 
-\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi DNA
-Polymerase Dependent:}} After chimera removal using VSEARCH and removal
-of sequences that did not classify as bacteria we assessed the
+\textbf{\emph{Chimeric Sequences are HiFi DNA Polymerase Dependent and
+Correlate with Number of OTUs:}} After chimera removal using VSEARCH and
+removal of sequences that did not classify as bacteria, we assessed the
 percentage of sequences that were still chimeric within our mock
@@
 the lowest chimera prevalence of all the HiFi DNA Polymerases regardless
-of whether pre.cluster or VSEARCH had been used {[}Figure 5{]}.
+of whether `pre.cluster' or chimera removal with VSEARCH had been used
+{[}Figure 5{]}.
 
@@
 strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
-{[}Figure 6{]}. In general, the R\textsuperscript{2} value between the
-number of OTUs and chimeric sequences did not change from the use of
-pre.cluster and VSEARCH {[}Figure 6{]}. Taken together, this data
-suggests that a strong correlation exists between the number of OTUs and
-the prevalence of chimeric sequences.
+{[}Figure 6{]}. The R\textsuperscript{2} value between the number of
+OTUs and chimeric sequences did not change from the use of `pre.cluster'
+and the removal of chimeras with VSEARCH {[}Figure 6{]}. Taken together,
+this data suggests that a strong correlation exists between the number
+of OTUs and the prevalence of chimeric sequences.
 

### 2018-02-01 submission/manuscript.Rmd
discussion edits

@@
 
-Our observations build upon previous studies [@taq_Gohl_2016] by showing that even different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
-This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results, as well as the findings.
+Our observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have both significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
+This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results and conclusions.
 Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4].
@@
 
-Although we did not observe strong differences in the Bray-Curtis index, based on cycle number, the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used.
-Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g.
+Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used.
+There was few differences within individuals between corresponding 5-cycle increments (e.g.
 15 to 20, 20 to 25, etc.).
-Conversely, we may not have enough samples to adequately judge the true magnitude difference between 20x and 25x communities and higher cycle numbers analyzed, even though there is a clear trend to suggest that 20x is very different then 25x [Figure 3].
-This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
-Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.
+However, there is a clear trend that suggests that 20 cycles is very different then 25 cycles [Figure 3].
+We may just not have enough samples to adequately assess the true magnitude difference between 20 and 25 cycle communities and higher cycle numbers analyzed.
+These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
+For some DNA polymerase, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
 
@@
 Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimeras that cannot be removed using bioinformatic approaches [Figure 4 & 5].
-Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A polymerase [@polymerase_Ishino_2014].
+Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014].
 In fact, from the material safety data sheet (MSDS), the differences between the two mixes is not immediately apparent.

### 2018-02-01 submission/manuscript.tex
discussion edits

@@
 Our observations build upon previous studies (Gohl et al., 2016) by
-showing that even different HiFi DNA polymerases have significant
+showing that different HiFi DNA polymerases have both significant
 differences in the number of OTUs and that changes in total OTUs
@@
 total number of OTUs as part of their calculations and changes to the
-total number of OTUs could drastically change the results, as well as
-the findings. Our observations show that HiFi DNA polymerase can have a
+total number of OTUs could drastically change the results and
+conclusions. Our observations show that HiFi DNA polymerase can have a
 noticeable affect on the OTUs generated and these differences are
@@
 
-Although we did not observe strong differences in the Bray-Curtis index,
-based on cycle number, the data suggests that there may be differences
-between 15 and 20 cycles versus higher cycle numbers, such as 30x, that
-are commonly used. Additionally, there was no difference within
-individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
-to 25, etc.). Conversely, we may not have enough samples to adequately
-judge the true magnitude difference between 20x and 25x communities and
-higher cycle numbers analyzed, even though there is a clear trend to
-suggest that 20x is very different then 25x {[}Figure 3{]}. This
-finding, in conjunction with the PERMANOVA results, suggest that cycle
-number can change bacterial community calculations but that these
-differences are minimal once 25 cycles are reached. Increasing the
-sub-sampling depth, for some DNA polymerase, may reduce some of these
-observed community differences at lower cycle numbers.
+Although we did not observe strong differences in the Bray-Curtis index
+the data suggests that there may be differences between 15 and 20 cycles
+versus higher cycle numbers, such as 30x, that are commonly used. There
+was few differences within individuals between corresponding 5-cycle
+increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a clear
+trend that suggests that 20 cycles is very different then 25 cycles
+{[}Figure 3{]}. We may just not have enough samples to adequately assess
+the true magnitude difference between 20 and 25 cycle communities and
+higher cycle numbers analyzed. These findings, in conjunction with the
+PERMANOVA results, suggest that cycle number can change bacterial
+community calculations but that these differences are minimal once 25
+cycles are reached. For some DNA polymerase, increasing the sub-sampling
+depth may reduce some of these observed community differences at lower
+cycle numbers.
 
@@
 a family A polymerase while the lowest, from Accuprime, was also an A
-polymerase (Ishino \& Ishino, 2014). In fact, from the material safety
-data sheet (MSDS), the differences between the two mixes is not
+family polymerase (Ishino \& Ishino, 2014). In fact, from the material
+safety data sheet (MSDS), the differences between the two mixes is not
 immediately apparent. Both Accuprime and Platinum contain a recombinant

### 2018-02-01 submission/manuscript.Rmd
conclusion and acknowledgement edits

@@
 
-Our findings show that measures that rely on number of OTUs will be specific for a particular study and may not be easily generalized to other studies investigating a similar area.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed as well as potentially influence diversity based metrics that do not down weight rare observations.
+Our findings show that measures that rely on the number of OTUs will be specific for a particular study due to both HiFi DNA polymerase and cycle number chosen.
+Due in part to these consderations findings may not be easily generalized to other studies investigating a similar area.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed as well as potentially influence diversity based metrics that do not down-weight rare observations.
 
@@
 
-The authors would like to thank all the study participants ERIN whose samples were utilized.
+The authors would like to thank all the study participants in ERIN whose samples were utilized.
 We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.

### 2018-02-01 submission/manuscript.tex
conclusion and acknowledgement edits

@@
 
-Our findings show that measures that rely on number of OTUs will be
-specific for a particular study and may not be easily generalized to
-other studies investigating a similar area. Care should be taken when
-choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic
-differences can change the number of OTUs observed as well as
-potentially influence diversity based metrics that do not down weight
-rare observations.
+Our findings show that measures that rely on the number of OTUs will be
+specific for a particular study due to both HiFi DNA polymerase and
+cycle number chosen. Due in part to these consderations findings may not
+be easily generalized to other studies investigating a similar area.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA
+gene surveys since intrinsic differences can change the number of OTUs
+observed as well as potentially influence diversity based metrics that
+do not down-weight rare observations.
 
@@
 
-The authors would like to thank all the study participants ERIN whose
+The authors would like to thank all the study participants in ERIN whose
 samples were utilized. We would also like to thank Judy Opp and April

### 2018-02-12 submission/manuscript.Rmd
update with bringing the lead to the fore
update abstract, intro, and methods

@@
 
-**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+**Methods.** DNA from fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
 Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach.
@@
 Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
-Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
+Thus, for these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
+This current study specifically addresses the PCR contribution of bias to this overall workflow.
 
@@
 Although these are important sources of bias they are not the only sources that should be critically examined.
-The type of DNA polymerase chosen could have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
+The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
 
 A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
-Another important component of this study was that the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline regardless of DNA polymerase used [@taq_Gohl_2016].
+Another important component of this study was that the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline, regardless of DNA polymerase used [@taq_Gohl_2016].
 Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
@@
 The processing and storage of these samples have been published previously [@erin_seekatz_2016].
-Clinical data and other types of meta data were not utilized or accessed for this study.
+Other than confirmation that none of these individuals had a *Clostridium difficle* infection clinical data and other types of meta data were not utilized or accessed for this study.
 All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
@@
 ***PCR Protocol:***
-The five different high fidelity (HiFi) DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The five different high fidelity (HiFi) DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
@@
 Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
-The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100).
-Although, the mock communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130).
+The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total samples = 100).
+The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130).
 No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
@@
 For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
-Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth.
+Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth.
 The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
@@
 20 versus 25, 25 versus 30, etc.).
-Based on these observations we wanted to next analyze potential reasons for these differences.
+For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
+Next, based on these observations we wanted to analyze potential reasons for these differences.
 First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase.
 After assessing these errors, the total number of chimeras was determined after sequence processing.
-For the community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
 

### 2018-02-12 submission/manuscript.tex
update with bringing the lead to the fore
update abstract, intro, and methods

@@
 
-\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
+\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
 single PowerMag DNA extraction kit with a 10 minute bead beating step
@@
 extraction kit, PCR, and sequence library preparation. Thus, for these
-overlapping considerations biases identified for 16S rRNA gene
+overlapping considerations, biases identified for 16S rRNA gene
 sequencing will also likely influence metagenomic sequencing results.
+This current study specifically addresses the PCR contribution of bias
+to this overall workflow.
 
@@
 sources that should be critically examined. The type of DNA polymerase
-chosen could have a wide ranging affect on downstream results due to
-error rates and chimeras that may not be easily resolved using
+chosen could also have a wide ranging affect on downstream results due
+to error rates and chimeras that may not be easily resolved using
 bioinformatic approaches.
@@
 PCR protocol (Gohl et al., 2016). Another important component of this
-study was that the number of OTUs generated were not easily removed
-using the authors chosen bioinformatic pipeline regardless of DNA
-polymerase used (Gohl et al., 2016). Although it is probably not
-surprising that normal DNA polymerase performed worse than HiFi DNA
-polymerase, it is natural to extend this line of inquiry and ask whether
-different HiFi DNA polymerase contribute different biases to downstream
-sequencing results. There is some reason to think that this may be the
-case since many of these HiFi DNA polymerase come from different
-families (e.g. \emph{Taq} belongs to the family A polymerases) and may
-intrinsically have different error rates that cannot be completely
-removed with modifications (Ishino \& Ishino, 2014). In this study we
-critically examine if any of five different HiFi DNA polymerases
-introduce significant biases into 16S rRNA gene surveys, if this is a
-cycle dependent phenomenon, and whether they can be removed using a
-standard bioinformatic pipeline.
+study was that the number of Operational Taxonomic Units (OTUs)
+generated were not easily removed using the authors chosen bioinformatic
+pipeline, regardless of DNA polymerase used (Gohl et al., 2016).
+Although it is probably not surprising that normal DNA polymerase
+performed worse than HiFi DNA polymerase, it is natural to extend this
+line of inquiry and ask whether different HiFi DNA polymerase contribute
+different biases to downstream sequencing results. There is some reason
+to think that this may be the case since many of these HiFi DNA
+polymerase come from different families (e.g. \emph{Taq} belongs to the
+family A polymerases) and may intrinsically have different error rates
+that cannot be completely removed with modifications (Ishino \& Ishino,
+2014). In this study we critically examine if any of five different HiFi
+DNA polymerases introduce significant biases into 16S rRNA gene surveys,
+if this is a cycle dependent phenomenon, and whether they can be removed
+using a standard bioinformatic pipeline.
 
@@
 Investigational Network (ERIN). The processing and storage of these
-samples have been published previously (Seekatz et al., 2016). Clinical
-data and other types of meta data were not utilized or accessed for this
-study. All samples were extracted using the MOBIO\textsuperscript{TM}
-PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The
-ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
-CA, USA) was used in this study and is made up of \emph{Pseudomonas
-aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
-\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
-\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
-\emph{Bacillus subtilis} at equal genomic DNA abundance
+samples have been published previously (Seekatz et al., 2016). Other
+than confirmation that none of these individuals had a \emph{Clostridium
+difficle} infection clinical data and other types of meta data were not
+utilized or accessed for this study. All samples were extracted using
+the MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
+(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
+Community DNA Standard (Zymo, CA, USA) was used in this study and is
+made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
+\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
+\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
+\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
+genomic DNA abundance
 (\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
@@
 \textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
-DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
+DNA polymerases that were tested included AccuPrime\textsuperscript{TM}
 (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
@@
 sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
-and 35 cycles for each Taq (total samples = 100). Although, the mock
-communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10
+and 35 cycles for each HiFi DNA polymerase (total samples = 100). The
+mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10
 replicates for 30 cycles for all HiFi DNA polymerase (total samples =
@@
 normalized to each individual by cycle number to account for the
-biological variation between people. Bray-Curtis matrices were generated
-using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000
-sequence depth. The distance matrix data was analyzed using PERMANOVA
-with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
+biological variation between people. Bray-Curtis distance matrices were
+generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
+15000 sequence depth. The distance matrix data was analyzed using
+PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
 Kruskal-Wallis tests within R. For both error and chimera analysis,
@@
 individual differences for the previous cycle (e.g.~20 versus 25, 25
-versus 30, etc.). Based on these observations we wanted to next analyze
-potential reasons for these differences. First, analysis of general
-sequence error rate, number of sequences with an error, and base
+versus 30, etc.). For these community based measures, the fecal samples
+were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
+10000, and 15000) while the mock community samples were analysed at 3
+levels (1000, 5000, 10000). Next, based on these observations we wanted
+to analyze potential reasons for these differences. First, analysis of
+general sequence error rate, number of sequences with an error, and base
 substitution were assessed in the mock community for each DNA
 polymerase. After assessing these errors, the total number of chimeras
-was determined after sequence processing. For the community based
-measures, the fecal samples were analyzed at 4 different sub-sampling
-sequence depths (1000, 5000, 10000, and 15000) while the mock community
-samples were analysed at 3 levels (1000, 5000, 10000).
+was determined after sequence processing.
 

### 2018-02-12 submission/manuscript.Rmd
update discussion and conclusion

@@
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
-Overall, there was little difference between each respective 5-cycle increment at higher cycle numbers, for both fecal and mock samples, and this was consistent across the different sub-samplings [Figure 3].
-There were only two differences between 5-cycle increments that were identified.
+For both fecal and mock samples, there was little difference between each respective 5-cycle increment at higher cycle numbers and this was consistent across the different sub-samplings [Figure 3].
+Overall, there were only two differences between 5-cycle increments that were identified.
 First, there were differences for the same fecal sample between 20 versus 25 cycles that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
@@
 Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Phusion was one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
+Phusion was also one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
 Next, we assessed whether there were any major differences between 5-cycle increments within each sample.
@@
 Differences between Platinum and all other HiFi DNA polymerases accounted for the vast majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
-Generally, across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH had been used [Figure 5].
+Across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH had been used [Figure 5].
 
-For all DNA polymerase, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
+For all HiFi DNA polymerases, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
 The R^2^ value between the number of OTUs and chimeric sequences did not change from the use of 'pre.cluster' and the removal of chimeras with VSEARCH [Figure 6].
@@
 Our observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have both significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
-This is important since many diversity metrics rely on the total number of OTUs as part of their calculations and changes to the total number of OTUs could drastically change the results and conclusions.
+HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this metric as part of their calculations.
 Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4].
@@
 These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
-For some DNA polymerase, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
+For some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
 
@@
 The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations.
-However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase.
+However, Bray-Curtis analysis with PERMANOVA showed few differences based on HiFi DNA polymerase.
 It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987].
@@
 
-Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimeras that cannot be removed using bioinformatic approaches [Figure 4 & 5].
+Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimera prevalence that cannot be removed using bioinformatic approaches [Figure 4 & 5].
 Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014].
@@
 
-Our findings show that measures that rely on the number of OTUs will be specific for a particular study due to both HiFi DNA polymerase and cycle number chosen.
-Due in part to these consderations findings may not be easily generalized to other studies investigating a similar area.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed as well as potentially influence diversity based metrics that do not down-weight rare observations.
+Our findings show that measures that rely on the number of OTUs will be dependent on both HiFi DNA polymerase and cycle number chosen.
+Due in part to these consderations, studies investigating similar areas may not be easily generalized with each other.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
 

### 2018-02-12 submission/manuscript.tex
update discussion and conclusion

@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} Overall, there was little difference between each respective
-5-cycle increment at higher cycle numbers, for both fecal and mock
-samples, and this was consistent across the different sub-samplings
-{[}Figure 3{]}. There were only two differences between 5-cycle
-increments that were identified. First, there were differences for the
-same fecal sample between 20 versus 25 cycles that was independent of
-HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to
-1000 = 0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), sub-sampled to
-5000 = 0.43 (0.33 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43))
-{[}Figure 3A{]}. Second, for the mocks, where data is available, there
-were larger difference between 20 and 25 cycles (sub-sampled to 1000 =
-0.88 (0.42 - 0.91)) {[}Figure 3B{]}. However, these differences between
-the next 5-cycle increment do not persist once 25 cycles are reached
-{[}Figure 3{]}.
+Number:}} For both fecal and mock samples, there was little difference
+between each respective 5-cycle increment at higher cycle numbers and
+this was consistent across the different sub-samplings {[}Figure 3{]}.
+Overall, there were only two differences between 5-cycle increments that
+were identified. First, there were differences for the same fecal sample
+between 20 versus 25 cycles that was independent of HiFi DNA polymerase
+but dependent on sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 -
+0.79) (median (25\% - 75\% quantile)), sub-sampled to 5000 = 0.43 (0.33
+- 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}.
+Second, for the mocks, where data is available, there were larger
+difference between 20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 -
+0.91)) {[}Figure 3B{]}. However, these differences between the next
+5-cycle increment do not persist once 25 cycles are reached {[}Figure
+3{]}.
 
@@
 differences at 1000 and 5000 sub-sampling depth (P-value = 0.03 and
-0.01, respectively). Phusion was one of only two HiFi DNA polymerases
-that managed to have samples for the 1000 sub-sampling depth at 15
-cycles. Next, we assessed whether there were any major differences
+0.01, respectively). Phusion was also one of only two HiFi DNA
+polymerases that managed to have samples for the 1000 sub-sampling depth
+at 15 cycles. Next, we assessed whether there were any major differences
 between 5-cycle increments within each sample. We found that there was
@@
 differences independent of cycle number and sub-sampling depth when
-using a Dunn's post-hoc test {[}Table S10{]}. Generally, across
-sub-sampling depth and cycle number Accuprime\textsuperscript{TM} had
-the lowest chimera prevalence of all the HiFi DNA Polymerases regardless
-of whether `pre.cluster' or chimera removal with VSEARCH had been used
-{[}Figure 5{]}.
+using a Dunn's post-hoc test {[}Table S10{]}. Across sub-sampling depth
+and cycle number Accuprime\textsuperscript{TM} had the lowest chimera
+prevalence of all the HiFi DNA Polymerases regardless of whether
+`pre.cluster' or chimera removal with VSEARCH had been used {[}Figure
+5{]}.
 
-For all DNA polymerase, a positive correlation was observed between
-chimeric sequences and number of OTUs, with this correlation being
-strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
+For all HiFi DNA polymerases, a positive correlation was observed
+between chimeric sequences and number of OTUs, with this correlation
+being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
 {[}Figure 6{]}. The R\textsuperscript{2} value between the number of
@@
 correlate with chimeras not removed after sequence processing {[}Figure
-1-2 \& 5{]}. This is important since many diversity metrics rely on the
-total number of OTUs as part of their calculations and changes to the
-total number of OTUs could drastically change the results and
-conclusions. Our observations show that HiFi DNA polymerase can have a
-noticeable affect on the OTUs generated and these differences are
-consistent across sub-sampling depth and PCR cycle number {[}Figure
-2-4{]}. These differences were observed in high biomass samples, where
-biases introduced by such components like kit contamination may have
-less of an effect, suggesting that these differences may be exacerbated
-in low biomass samples.
+1-2 \& 5{]}. HiFi DNA polymerase dependent differences in total number
+of OTUs are important to consider since many diversity metrics rely on
+this metric as part of their calculations. Our observations show that
+HiFi DNA polymerase can have a noticeable affect on the OTUs generated
+and these differences are consistent across sub-sampling depth and PCR
+cycle number {[}Figure 2-4{]}. These differences were observed in high
+biomass samples, where biases introduced by such components like kit
+contamination may have less of an effect, suggesting that these
+differences may be exacerbated in low biomass samples.
 
@@
 community calculations but that these differences are minimal once 25
-cycles are reached. For some DNA polymerase, increasing the sub-sampling
-depth may reduce some of these observed community differences at lower
-cycle numbers.
+cycles are reached. For some HiFI DNA polymerases, increasing the
+sub-sampling depth may reduce some of these observed community
+differences at lower cycle numbers.
 
@@
 However, Bray-Curtis analysis with PERMANOVA showed few differences
-based on DNA polymerase. It is possible that many of the increased
+based on HiFi DNA polymerase. It is possible that many of the increased
 number of OTUs are not highly abundant allowing the Bray-Curtis index to
@@
 Our observations suggest that there are clear HiFi DNA polymerase
-dependent differences in both per base error rate and chimeras that
-cannot be removed using bioinformatic approaches {[}Figure 4 \& 5{]}.
-Although it may be a natural assumption that the variation may be due to
-the DNA polymerase family, the highest chimera rate, from Platinum, was
-a family A polymerase while the lowest, from Accuprime, was also an A
-family polymerase (Ishino \& Ishino, 2014). In fact, from the material
-safety data sheet (MSDS), the differences between the two mixes is not
-immediately apparent. Both Accuprime and Platinum contain a recombinant
-\emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D polymerase and
-a platinum \emph{Taq} antibody. It is possible that differences in how
-the recombinant \emph{Taq} was generated could be the main reason for
-the differences in chimera rate.
+dependent differences in both per base error rate and chimera prevalence
+that cannot be removed using bioinformatic approaches {[}Figure 4 \&
+5{]}. Although it may be a natural assumption that the variation may be
+due to the DNA polymerase family, the highest chimera rate, from
+Platinum, was a family A polymerase while the lowest, from Accuprime,
+was also an A family polymerase (Ishino \& Ishino, 2014). In fact, from
+the material safety data sheet (MSDS), the differences between the two
+mixes is not immediately apparent. Both Accuprime and Platinum contain a
+recombinant \emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D
+polymerase and a platinum \emph{Taq} antibody. It is possible that
+differences in how the recombinant \emph{Taq} was generated could be the
+main reason for the differences in chimera rate.
 
@@
 Our findings show that measures that rely on the number of OTUs will be
-specific for a particular study due to both HiFi DNA polymerase and
-cycle number chosen. Due in part to these consderations findings may not
-be easily generalized to other studies investigating a similar area.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA
-gene surveys since intrinsic differences can change the number of OTUs
-observed as well as potentially influence diversity based metrics that
-do not down-weight rare observations.
+dependent on both HiFi DNA polymerase and cycle number chosen. Due in
+part to these consderations, studies investigating similar areas may not
+be easily generalized with each other. Care should be taken when
+choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic
+differences can change the number of OTUs observed and influence
+diversity based metrics that do not down-weight rare observations.
 

### 2018-02-19 submission/manuscript.Rmd
add sra accession number
update abstract

@@
 
-**Background.** An increasing body of research has found that various methodological steps can have an impact on the observed microbial community when using 16S rRNA gene surveys.
-These components include, but are not limited to, preservation media, extraction kit, bead beating time, and primers.
-Both cycle number and high fidelity (HiFi) DNA polymerase are sometimes overlooked when sources of bias are being considered.
-Here we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.
+**Background.** An increasing body of research has found that various methodological steps along the 16S rRNA gene survey pipeline can change the observed microbial community.
+The most common studied methodological components include preservation media, extraction kit, bead beating time, and primers.
+Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias.
+Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.
 
-**Methods.** DNA from fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach.
-First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communities.
-Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase.
-Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.
+**Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR approach.
+First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities.
+Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated.
+Finally, the chimera prevalence correlation with number of OTUs was assessed.
 
@@
 Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH.
-Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs observed in a sample and this association was not changed by chimera removal with VSEARCH.
+Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs and this association was not changed by chimera removal with VSEARCH.
 
-**Conclusions.** Due to the impact of HiFi DNA polymerase on the number of OTUs, common diversity metrics that incorporate this value could give artificially inflated numbers due to higher undetected chimeras.
-When designing 16S rRNA gene survey studies it is important to consider both the cycle number and the type of HiFi DNA polymerase that will be used since it can increase or decrease the number of OTUs that are observed.
+**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have artificially inflated values that are not comparable across studies.
 
@@
 ***Reproducible Methods:***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number **need to upload still**.
+The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number SRP132931.
 

### 2018-02-19 submission/manuscript.tex
add sra accession number
update abstract

@@
 \textbf{Background.} An increasing body of research has found that
-various methodological steps can have an impact on the observed
-microbial community when using 16S rRNA gene surveys. These components
-include, but are not limited to, preservation media, extraction kit,
-bead beating time, and primers. Both cycle number and high fidelity
-(HiFi) DNA polymerase are sometimes overlooked when sources of bias are
-being considered. Here we critically examine both cycle number and HiFi
+various methodological steps along the 16S rRNA gene survey pipeline can
+change the observed microbial community. The most common studied
+methodological components include preservation media, extraction kit,
+bead beating time, and primers. Although cycle number and high fidelity
+(HiFi) DNA polymerase are studied less often, they are still important
+sources of bias. Here, we critically examine both cycle number and HiFi
 DNA polymerase for biases that may influence downstream diversity
@@
 \textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
-single PowerMag DNA extraction kit with a 10 minute bead beating step
-and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
+PowerMag DNA extraction kit with a 10 minute bead beating step and
+amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
 Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
@@
 genomes of 8 different bacteria were also amplified using the same PCR
-amplification approach. First, the number of OTUs (Operational Taxonomic
-Units) was examined for both fecal samples and mock communities. Next,
-Bray-Curtis index, the error rate, sequence error prevalence, and
-chimera prevalence were assessed based on cycle number and HiFi DNA
-polymerase. Finally, the chimera prevalence correlation with number of
-OTUs was assessed for both cycle number and HiFi DNA polymerase
-dependent differences.
+approach. First, the number of OTUs (Operational Taxonomic Units) were
+examined for both fecal samples and mock communities. Next, Bray-Curtis
+index, the error rate, sequence error prevalence, and chimera prevalence
+were investigated. Finally, the chimera prevalence correlation with
+number of OTUs was assessed.
 
@@
 Additionally, the chimera prevalence had a strong positive correlation
-with the number of OTUs observed in a sample and this association was
-not changed by chimera removal with VSEARCH.
+with the number of OTUs and this association was not changed by chimera
+removal with VSEARCH.
 
-\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
-number of OTUs, common diversity metrics that incorporate this value
-could give artificially inflated numbers due to higher undetected
-chimeras. When designing 16S rRNA gene survey studies it is important to
-consider both the cycle number and the type of HiFi DNA polymerase that
-will be used since it can increase or decrease the number of OTUs that
-are observed.
+\textbf{Conclusions.} Due to HiFi DNA polymerase dependent differences
+in the number of OTUs and chimera prevalence, common diversity metrics
+could have artificially inflated values that are not comparable across
+studies.
 
@@
 The raw sequences can be found in the SRA at the following accession
-number \textbf{need to upload still}.
+number SRP132931.
 

### 2018-02-20 submission/manuscript.Rmd
update abstract and intro

@@
 
-**Background.** An increasing body of research has found that various methodological steps along the 16S rRNA gene survey pipeline can change the observed microbial community.
-The most common studied methodological components include preservation media, extraction kit, bead beating time, and primers.
+**Background.** Research has identified various methodological steps along the 16S rRNA gene survey pipeline that can change microbial community results.
 Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias.
-Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream diversity measures of 16S rRNA gene surveys.
+Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream bacterial community results.
 
 **Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR approach.
+Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach.
 First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities.
@@
 
-**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have artificially inflated values that are not comparable across studies.
+**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have values that are not comparable across studies.
 
@@
 
-Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
-In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
-Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
+Recently, there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
+Investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant focal point of this push towards standardization [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
+Due to this focus, 16S rRNA gene sequencing has been much maligned for introduced biases.
+Yet, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
 Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
-Thus, for these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
-This current study specifically addresses the PCR contribution of bias to this overall workflow.
+For these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
 
-Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases.
-The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g.
+The most commonly studied biases are preservation media, storage conditions, and DNA extraction kits.
+Although DNA extraction kits have consistently been shown to add bias to downstream analysis the current literature on preservation media and storage conditions has been mixed, with some studies showing biases while others do not [@contamination_Salter2014; @metagenomcis_bias_Costea_2017; @preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015].
+The study of these specific biases has become so large, aggregating them all together has become a difficult task with some researchers providing resources to actively track new findings (e.g.
 Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/).
-Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017].
-However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015].
-Although these are important sources of bias they are not the only sources that should be critically examined.
+It is evident that these three sources of bias have been extensively covered within the literature but they are not the only sources that should be critically examined.
 The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
 
-A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
-Another important component of this study was that the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline, regardless of DNA polymerase used [@taq_Gohl_2016].
-Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
+A recent study in *Nature Biotechnology* found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
+This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016].
+It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
 There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
@@
 In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
-
-We amplified the V4 region of the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase.
-First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples.
-Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected.
-After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence.
-Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.
 

### 2018-02-20 submission/manuscript.tex
update abstract and intro

@@
 
-\textbf{Background.} An increasing body of research has found that
-various methodological steps along the 16S rRNA gene survey pipeline can
-change the observed microbial community. The most common studied
-methodological components include preservation media, extraction kit,
-bead beating time, and primers. Although cycle number and high fidelity
-(HiFi) DNA polymerase are studied less often, they are still important
-sources of bias. Here, we critically examine both cycle number and HiFi
-DNA polymerase for biases that may influence downstream diversity
-measures of 16S rRNA gene surveys.
+\textbf{Background.} Research has identified various methodological
+steps along the 16S rRNA gene survey pipeline that can change microbial
+community results. Although cycle number and high fidelity (HiFi) DNA
+polymerase are studied less often, they are still important sources of
+bias. Here, we critically examine both cycle number and HiFi DNA
+polymerase for biases that may influence downstream bacterial community
+results.
 
@@
 (technical replicates n = 4) consisting of previously isolated whole
-genomes of 8 different bacteria were also amplified using the same PCR
+genomes of 8 different bacteria were also amplified using the same
 approach. First, the number of OTUs (Operational Taxonomic Units) were
@@
 in the number of OTUs and chimera prevalence, common diversity metrics
-could have artificially inflated values that are not comparable across
-studies.
+could have values that are not comparable across studies.
 
@@
 
-Recently there has been an increasing focus on standardizing
+Recently, there has been an increasing focus on standardizing
 methodological approaches in microbiota research (Kim et al., 2017;
-Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
-rRNA gene surveys can be made more reproducible has been a predominant
-driver of this standardization push (Lauber et al., 2010; Salter et al.,
-2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
-sequencing has been much maligned for introduced biases, many of these
-same considerations also affect metagenomic sequencing (Nayfach \&
-Pollard, 2016; Costea et al., 2017). Between the two approaches similar
-bias considerations include preservation media, storage conditions, DNA
-extraction kit, PCR, and sequence library preparation. Thus, for these
-overlapping considerations, biases identified for 16S rRNA gene
-sequencing will also likely influence metagenomic sequencing results.
-This current study specifically addresses the PCR contribution of bias
-to this overall workflow.
+Hugerth \& Andersson, 2017). Investigating ways that 16S rRNA gene
+surveys can be made more reproducible has been a predominant focal point
+of this push towards standardization (Lauber et al., 2010; Salter et
+al., 2014; Song et al., 2016; Gohl et al., 2016). Due to this focus, 16S
+rRNA gene sequencing has been much maligned for introduced biases. Yet,
+many of these same considerations also affect metagenomic sequencing
+(Nayfach \& Pollard, 2016; Costea et al., 2017). Between the two
+approaches similar bias considerations include preservation media,
+storage conditions, DNA extraction kit, PCR, and sequence library
+preparation. For these overlapping considerations, biases identified for
+16S rRNA gene sequencing will also likely influence metagenomic
+sequencing results.
 
-Currently, preservation media used, storage conditions, and DNA
-extraction kits chosen are the most commonly studied biases. The study
-of these specific biases has become so large, aggregating them all
-together has become a difficult task and some researchers provide
-resources to actively track new findings (e.g.~Microbiome Digest -
+The most commonly studied biases are preservation media, storage
+conditions, and DNA extraction kits. Although DNA extraction kits have
+consistently been shown to add bias to downstream analysis the current
+literature on preservation media and storage conditions has been mixed,
+with some studies showing biases while others do not (Lauber et al.,
+2010; Dominianni et al., 2014; Salter et al., 2014; Sinha et al., 2015;
+Song et al., 2016; Luo et al., 2016; Bassis et al., 2017; Costea et al.,
+2017). The study of these specific biases has become so large,
+aggregating them all together has become a difficult task with some
+researchers providing resources to actively track new findings
+(e.g.~Microbiome Digest -
 \url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
-Within the literature, DNA extraction kits have consistently been shown
-to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
-2017). However, the current literature on preservation media and storage
-conditions has been more mixed, with some studies showing biases while
-others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
-al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
-Although these are important sources of bias they are not the only
-sources that should be critically examined. The type of DNA polymerase
-chosen could also have a wide ranging affect on downstream results due
-to error rates and chimeras that may not be easily resolved using
-bioinformatic approaches.
+It is evident that these three sources of bias have been extensively
+covered within the literature but they are not the only sources that
+should be critically examined. The type of DNA polymerase chosen could
+also have a wide ranging affect on downstream results due to error rates
+and chimeras that may not be easily resolved using bioinformatic
+approaches.
 
-A recent study in \emph{Nature Biotechnology} showed that there were
-clear differences between normal and high fidelity (HiFi) DNA polymerase
-and that you could reduce error and chimera generation by optimizing the
-PCR protocol (Gohl et al., 2016). Another important component of this
-study was that the number of Operational Taxonomic Units (OTUs)
-generated were not easily removed using the authors chosen bioinformatic
-pipeline, regardless of DNA polymerase used (Gohl et al., 2016).
-Although it is probably not surprising that normal DNA polymerase
-performed worse than HiFi DNA polymerase, it is natural to extend this
-line of inquiry and ask whether different HiFi DNA polymerase contribute
-different biases to downstream sequencing results. There is some reason
-to think that this may be the case since many of these HiFi DNA
-polymerase come from different families (e.g. \emph{Taq} belongs to the
-family A polymerases) and may intrinsically have different error rates
-that cannot be completely removed with modifications (Ishino \& Ishino,
-2014). In this study we critically examine if any of five different HiFi
-DNA polymerases introduce significant biases into 16S rRNA gene surveys,
-if this is a cycle dependent phenomenon, and whether they can be removed
+A recent study in \emph{Nature Biotechnology} found clear differences
+between normal and high fidelity (HiFi) DNA polymerase and that you
+could reduce error and chimera generation by optimizing the PCR protocol
+(Gohl et al., 2016). This study also found that, regardless of DNA
+polymerase, the number of Operational Taxonomic Units (OTUs) generated
+were not easily removed using the authors chosen bioinformatic pipeline
+(Gohl et al., 2016). It is natural to extend this line of inquiry and
+ask whether different HiFi DNA polymerase contribute different biases to
+downstream sequencing results. There is some reason to think that this
+may be the case since many of these HiFi DNA polymerase come from
+different families (e.g. \emph{Taq} belongs to the family A polymerases)
+and may intrinsically have different error rates that cannot be
+completely removed with modifications (Ishino \& Ishino, 2014). In this
+study we critically examine if any of five different HiFi DNA
+polymerases introduce significant biases into 16S rRNA gene surveys, if
+this is a cycle dependent phenomenon, and whether they can be removed
 using a standard bioinformatic pipeline.
-
-We amplified the V4 region of the 16S rRNA gene in both fecal and mock
-community samples using either Accuprime, Kappa, Q5, Phusion, or
-Platinum HiFi DNA polymerase. First, we tested if we could identify
-differences in the number of OTUs between the different HiFi DNA
-polymerase for both the fecal and mock samples. Next, since there were
-differences based on HiFi DNA polymerase, we examined if a community
-measure such as the Bray-Curtis index would be affected. After assessing
-these community measures we then examined if the different HiFi DNA
-polymerase had different per base error rates as well as chimera
-prevalence. Finally, since differences in chimera prevalence, based on
-HiFi DNA polymerase, could not be completely removed using bioinformatic
-approaches we tested whether the chimera prevalence correlated well with
-total number of observed OTUs.
 

### 2018-02-20 submission/manuscript.Rmd
update methods

@@
 A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
-The processing and storage of these samples have been published previously [@erin_seekatz_2016].
-Other than confirmation that none of these individuals had a *Clostridium difficle* infection clinical data and other types of meta data were not utilized or accessed for this study.
+The processing and storage of these samples have been previously published [@erin_seekatz_2016].
+Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study.
 All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
@@
 ***PCR Protocol:***
-The five different high fidelity (HiFi) DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The five different HiFi DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
@@
 Generally, the protocol followed what has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
 Sequence error was determined using the 'seq.error' command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+
+***Analysis Workflow:***
+The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
+Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
+20 versus 25, 25 versus 30, etc.).
+For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
+Next, based on these observations we wanted to analyze potential reasons for these differences.
+First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each HiFi DNA polymerase.
+After assessing these errors, the total number of chimeras was determined after sequence processing and correlations to the number of OTUs also assessed.
 
@@
 Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
-
-***Analysis Workflow:***
-The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
-20 versus 25, 25 versus 30, etc.).
-For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
-Next, based on these observations we wanted to analyze potential reasons for these differences.
-First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase.
-After assessing these errors, the total number of chimeras was determined after sequence processing.
 

### 2018-02-20 submission/manuscript.tex
update methods

@@
 Investigational Network (ERIN). The processing and storage of these
-samples have been published previously (Seekatz et al., 2016). Other
+samples have been previously published (Seekatz et al., 2016). Other
 than confirmation that none of these individuals had a \emph{Clostridium
-difficle} infection clinical data and other types of meta data were not
+difficle} infection, clinical data and other types of meta data were not
 utilized or accessed for this study. All samples were extracted using
@@
 
-\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
-DNA polymerases that were tested included AccuPrime\textsuperscript{TM}
-(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
-(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
-England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
-Accuprime followed a previously published protocol (Kozich et al., 2013)
+\textbf{\emph{PCR Protocol:}} The five different HiFi DNA polymerases
+that were tested included AccuPrime\textsuperscript{TM} (ThermoFisher,
+MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA),
+Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The PCR cycle conditions for Platinum and Accuprime followed a
+previously published protocol (Kozich et al., 2013)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
@@
 instead of UCHIME for chimera detection and the use of the OptiClust
-algorithm instead of average neighbor for Operational Taxonomic Unit
-(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
-2016; Westcott \& Schloss, 2017). Sequence error was determined using
-the `seq.error' command on mock samples after chimera removal and
-classification to the RDP to remove non-bacterial sequences (Schloss et
-al., 2009; Cole et al., 2013; Rognes et al., 2016).
+algorithm instead of average neighbor for OTU generation at 97\%
+similarity (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
+Schloss, 2017). Sequence error was determined using the `seq.error'
+command on mock samples after chimera removal and classification to the
+RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
+al., 2013; Rognes et al., 2016).
+
+\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
+sub-sampling was analyzed for both the fecal and mock community samples.
+Cycle dependent affects on Bray-Curtis indices were next assessed for
+the fecal samples looking at both overall cycle differences and within
+individual differences for the previous cycle (e.g.~20 versus 25, 25
+versus 30, etc.). For these community based measures, the fecal samples
+were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
+10000, and 15000) while the mock community samples were analysed at 3
+levels (1000, 5000, 10000). Next, based on these observations we wanted
+to analyze potential reasons for these differences. First, analysis of
+general sequence error rate, number of sequences with an error, and base
+substitution were assessed in the mock community for each HiFi DNA
+polymerase. After assessing these errors, the total number of chimeras
+was determined after sequence processing and correlations to the number
+of OTUs also assessed.
 
@@
 Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
-
-\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
-sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curtis indices were next assessed for
-the fecal samples looking at both overall cycle differences and within
-individual differences for the previous cycle (e.g.~20 versus 25, 25
-versus 30, etc.). For these community based measures, the fecal samples
-were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
-10000, and 15000) while the mock community samples were analysed at 3
-levels (1000, 5000, 10000). Next, based on these observations we wanted
-to analyze potential reasons for these differences. First, analysis of
-general sequence error rate, number of sequences with an error, and base
-substitution were assessed in the mock community for each DNA
-polymerase. After assessing these errors, the total number of chimeras
-was determined after sequence processing.
 

### 2018-02-20 submission/manuscript.Rmd
update results section

@@
 
-***The Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
-In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data.
-After normalization, we identified that there was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1].
+***The Number of OTUs is Dependent on HiFi DNA Polymerase:***
+There was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1].
 Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1].
@@
 There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-Platinum HiFi DNA polymerase was the main driver of the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Platinum HiFi DNA polymerase was the main reason for the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
 
 This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases having high (Platinum) and low (Accuprime) number of OTUs [Figure 2 & Table S3].
-Conversely, differences between HiFi DNA polymerase were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+Differences between HiFi DNA polymerase were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
 Using a Tukey post-hoc test differences between Platinum and the other HiFi DNA polymerases was the major driver of the differences seen at different cycle numbers and sub-sampling depths [Table S4].
-Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum for both fecal and mock samples [Figure 1 & 2].
+Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum [Figure 1 & 2].
+Based on these observations it is clear that HiFi DNA polymerase affects the total number of OTUs observed within a sample.
 
@@
 For both fecal and mock samples, there was little difference between each respective 5-cycle increment at higher cycle numbers and this was consistent across the different sub-samplings [Figure 3].
-Overall, there were only two differences between 5-cycle increments that were identified.
+There were only two differences between 5-cycle increments that were identified.
 First, there were differences for the same fecal sample between 20 versus 25 cycles that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Second, for the mocks, where data is available, there were larger difference between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
+Second, where data is available for the mock communities, there were larger difference between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
 However, these differences between the next 5-cycle increment do not persist once 25 cycles are reached [Figure 3].
@@
 Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Phusion was also one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles.
+Phusion was also one of only two HiFi DNA polymerases that that had samples for the 1000 sub-sampling depth at 15 cycles.
 Next, we assessed whether there were any major differences between 5-cycle increments within each sample.
@@
 It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5-cycle increments within an individual.
+Overall, HiFi DNA polymerase differences in Bray-Curtis index are dependent on both sub-sampling depth and cycle number.
 
@@
 Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-Generally, the highest per base median average error rates were for the Kappa HiFi DNA polymerase [Figure 4].
+The highest per base median average error rates were for the Kappa HiFi DNA polymerase [Figure 4].
 This error rate was minimally affected by both the 'pre.cluster' step and chimera removal by VSEARCH [Figure 4].
@@
 Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1].
-However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1].
-Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were generally no biases in the types of substitution made [Figure S2].
+However, this increased number of sequences with an error was lowered with existing bioinformatic approaches [Figure S1].
+Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were no biases in the types of substitution made [Figure S2].
+Although sequence error is dependent on HiFi DNA polymerase it can be corrected using exsiting bioinformatic approaches.
 
 ***Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with Number of OTUs:***
-After chimera removal using VSEARCH and removal of sequences that did not classify as bacteria, we assessed the percentage of sequences that were still chimeric within our mock community.
-At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Differences between Platinum and all other HiFi DNA polymerases accounted for the vast majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
-Across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH had been used [Figure 5].
+At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used and chimera prevalence (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Differences between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
+Across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH was used [Figure 5].
 

### 2018-02-20 submission/manuscript.tex
update results section

@@
 
-\textbf{\emph{The Number of OTUs is Dependent on the HiFi DNA Polymerase
-Used:}} In order to compare the number of OTUs across individuals a
-Z-score normalization by individual, by cycle number, was employed on
-the number of OTUs data. After normalization, we identified that there
-was a HiFi DNA polymerase dependent difference that was consistent
+\textbf{\emph{The Number of OTUs is Dependent on HiFi DNA Polymerase:}}
+There was a HiFi DNA polymerase dependent difference that was consistent
 across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
@@
 sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
-{[}Table S1{]}. Platinum HiFi DNA polymerase was the main driver of the
+{[}Table S1{]}. Platinum HiFi DNA polymerase was the main reason for the
 differences observed across all sub-sampling depths at 35 cycles, based
@@
 polymerases having high (Platinum) and low (Accuprime) number of OTUs
-{[}Figure 2 \& Table S3{]}. Conversely, differences between HiFi DNA
-polymerase were observed as early as 20 cycles and a sub-sampling depth
-of 1000 sequences (P-value = 0.002) {[}Table S3{]}. Using a Tukey
-post-hoc test differences between Platinum and the other HiFi DNA
-polymerases was the major driver of the differences seen at different
-cycle numbers and sub-sampling depths {[}Table S4{]}. Both fecal and
-mock samples consistently showed that across sub-sampling depth and
-cycle number the lowest number of OTUs identified was from
-Accuprime\textsuperscript{TM} while the highest was from Platinum for
-both fecal and mock samples {[}Figure 1 \& 2{]}.
+{[}Figure 2 \& Table S3{]}. Differences between HiFi DNA polymerase were
+observed as early as 20 cycles and a sub-sampling depth of 1000
+sequences (P-value = 0.002) {[}Table S3{]}. Using a Tukey post-hoc test
+differences between Platinum and the other HiFi DNA polymerases was the
+major driver of the differences seen at different cycle numbers and
+sub-sampling depths {[}Table S4{]}. Both fecal and mock samples
+consistently showed that across sub-sampling depth and cycle number the
+lowest number of OTUs identified was from Accuprime\textsuperscript{TM}
+while the highest was from Platinum {[}Figure 1 \& 2{]}. Based on these
+observations it is clear that HiFi DNA polymerase affects the total
+number of OTUs observed within a sample.
 
@@
 this was consistent across the different sub-samplings {[}Figure 3{]}.
-Overall, there were only two differences between 5-cycle increments that
-were identified. First, there were differences for the same fecal sample
+There were only two differences between 5-cycle increments that were
+identified. First, there were differences for the same fecal sample
 between 20 versus 25 cycles that was independent of HiFi DNA polymerase
@@
 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}.
-Second, for the mocks, where data is available, there were larger
-difference between 20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 -
-0.91)) {[}Figure 3B{]}. However, these differences between the next
-5-cycle increment do not persist once 25 cycles are reached {[}Figure
-3{]}.
+Second, where data is available for the mock communities, there were
+larger difference between 20 and 25 cycles (sub-sampled to 1000 = 0.88
+(0.42 - 0.91)) {[}Figure 3B{]}. However, these differences between the
+next 5-cycle increment do not persist once 25 cycles are reached
+{[}Figure 3{]}.
 
@@
 0.01, respectively). Phusion was also one of only two HiFi DNA
-polymerases that managed to have samples for the 1000 sub-sampling depth
-at 15 cycles. Next, we assessed whether there were any major differences
+polymerases that that had samples for the 1000 sub-sampling depth at 15
+cycles. Next, we assessed whether there were any major differences
 between 5-cycle increments within each sample. We found that there was
@@
 differences in cycle number and within 5-cycle increments within an
-individual.
+individual. Overall, HiFi DNA polymerase differences in Bray-Curtis
+index are dependent on both sub-sampling depth and cycle number.
 
@@
 average per base error varied without a clear pattern across
-sub-sampling depth {[}Table S5{]}. Generally, the highest per base
-median average error rates were for the Kappa HiFi DNA polymerase
-{[}Figure 4{]}. This error rate was minimally affected by both the
-`pre.cluster' step and chimera removal by VSEARCH {[}Figure 4{]}. There
-were small differences in the per base error rate between the various
-HiFi DNA polymerase at lower cycle numbers and larger differences at
-higher cycle number with Platinum having the largest differences of all
-the HiFi DNA polymerase {[}Figure 4B-C and Table S6{]}.
+sub-sampling depth {[}Table S5{]}. The highest per base median average
+error rates were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. This
+error rate was minimally affected by both the `pre.cluster' step and
+chimera removal by VSEARCH {[}Figure 4{]}. There were small differences
+in the per base error rate between the various HiFi DNA polymerase at
+lower cycle numbers and larger differences at higher cycle number with
+Platinum having the largest differences of all the HiFi DNA polymerase
+{[}Figure 4B-C and Table S6{]}.
 
@@
 cycle number or sub-sampling depth {[}Figure S1{]}. However, this
-increased number of sequences with an error can be drastically lowered
-with existing bioinformatic approaches {[}Figure S1{]}. Investigation of
-whether there were HiFi DNA polymerase dependent effects on base
-substitution found that there were generally no biases in the types of
-substitution made {[}Figure S2{]}.
+increased number of sequences with an error was lowered with existing
+bioinformatic approaches {[}Figure S1{]}. Investigation of whether there
+were HiFi DNA polymerase dependent effects on base substitution found
+that there were no biases in the types of substitution made {[}Figure
+S2{]}. Although sequence error is dependent on HiFi DNA polymerase it
+can be corrected using exsiting bioinformatic approaches.
 
 \textbf{\emph{Chimeric Sequences are HiFi DNA Polymerase Dependent and
-Correlate with Number of OTUs:}} After chimera removal using VSEARCH and
-removal of sequences that did not classify as bacteria, we assessed the
-percentage of sequences that were still chimeric within our mock
-community. At all levels of sub-sampling and cycle number there were
-significant differences between the HiFi DNA polymerase used (P-value
-\textless{} 0.05) {[}Table S9{]}. Differences between Platinum and all
-other HiFi DNA polymerases accounted for the vast majority of these
-differences independent of cycle number and sub-sampling depth when
-using a Dunn's post-hoc test {[}Table S10{]}. Across sub-sampling depth
-and cycle number Accuprime\textsuperscript{TM} had the lowest chimera
-prevalence of all the HiFi DNA Polymerases regardless of whether
-`pre.cluster' or chimera removal with VSEARCH had been used {[}Figure
-5{]}.
+Correlate with Number of OTUs:}} At all levels of sub-sampling and cycle
+number there were significant differences between the HiFi DNA
+polymerase used and chimera prevalence (P-value \textless{} 0.05)
+{[}Table S9{]}. Differences between Platinum and all other HiFi DNA
+polymerases accounted for the majority of these differences independent
+of cycle number and sub-sampling depth when using a Dunn's post-hoc test
+{[}Table S10{]}. Across sub-sampling depth and cycle number
+Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
+the HiFi DNA Polymerases regardless of whether `pre.cluster' or chimera
+removal with VSEARCH was used {[}Figure 5{]}.
 

### 2018-02-20 submission/manuscript.Rmd
update discussion

@@
 
-Our observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have both significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
-HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this metric as part of their calculations.
+These observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
+HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this measure as part of their calculations.
 Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4].
-These differences were observed in high biomass samples, where biases introduced by such components like kit contamination may have less of an effect, suggesting that these differences may be exacerbated in low biomass samples.
+Importantly, high biomass samples may underestimate the biases present within low biomass samples.
+An indication that this may be the case is the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples.
 
-Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used.
+Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30 cycles, that are commonly used.
 There was few differences within individuals between corresponding 5-cycle increments (e.g.
 15 to 20, 20 to 25, etc.).
-However, there is a clear trend that suggests that 20 cycles is very different then 25 cycles [Figure 3].
-We may just not have enough samples to adequately assess the true magnitude difference between 20 and 25 cycle communities and higher cycle numbers analyzed.
+However, there is a clear trend that suggests that the community at 20 cycles is different then the community at 25 cycles [Figure 3].
 These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
-For some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
+Further, for some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
 
@@
 It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g.
-Platinum).
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases.
 
@@
 Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014].
-In fact, from the material safety data sheet (MSDS), the differences between the two mixes is not immediately apparent.
-Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pycrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+In fact, from the material safety data sheet (MSDS), the differences between the two is not immediately apparent.
+Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
 It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.

### 2018-02-20 submission/manuscript.tex
update discussion

@@
 
-Our observations build upon previous studies (Gohl et al., 2016) by
-showing that different HiFi DNA polymerases have both significant
-differences in the number of OTUs and that changes in total OTUs
-correlate with chimeras not removed after sequence processing {[}Figure
-1-2 \& 5{]}. HiFi DNA polymerase dependent differences in total number
-of OTUs are important to consider since many diversity metrics rely on
-this metric as part of their calculations. Our observations show that
-HiFi DNA polymerase can have a noticeable affect on the OTUs generated
-and these differences are consistent across sub-sampling depth and PCR
-cycle number {[}Figure 2-4{]}. These differences were observed in high
-biomass samples, where biases introduced by such components like kit
-contamination may have less of an effect, suggesting that these
-differences may be exacerbated in low biomass samples.
+These observations build upon previous studies (Gohl et al., 2016) by
+showing that different HiFi DNA polymerases have significant differences
+in the number of OTUs and that changes in total OTUs correlate with
+chimeras not removed after sequence processing {[}Figure 1-2 \& 5{]}.
+HiFi DNA polymerase dependent differences in total number of OTUs are
+important to consider since many diversity metrics rely on this measure
+as part of their calculations. Our observations show that HiFi DNA
+polymerase can have a noticeable affect on the OTUs generated and these
+differences are consistent across sub-sampling depth and PCR cycle
+number {[}Figure 2-4{]}. Importantly, high biomass samples may
+underestimate the biases present within low biomass samples. An
+indication that this may be the case is the earlier detection of
+differences in total number of OTUs between HiFi DNA polymerases in the
+mock versus fecal samples.
 
@@
 the data suggests that there may be differences between 15 and 20 cycles
-versus higher cycle numbers, such as 30x, that are commonly used. There
-was few differences within individuals between corresponding 5-cycle
-increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a clear
-trend that suggests that 20 cycles is very different then 25 cycles
-{[}Figure 3{]}. We may just not have enough samples to adequately assess
-the true magnitude difference between 20 and 25 cycle communities and
-higher cycle numbers analyzed. These findings, in conjunction with the
-PERMANOVA results, suggest that cycle number can change bacterial
-community calculations but that these differences are minimal once 25
-cycles are reached. For some HiFI DNA polymerases, increasing the
-sub-sampling depth may reduce some of these observed community
-differences at lower cycle numbers.
+versus higher cycle numbers, such as 30 cycles, that are commonly used.
+There was few differences within individuals between corresponding
+5-cycle increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a
+clear trend that suggests that the community at 20 cycles is different
+then the community at 25 cycles {[}Figure 3{]}. These findings, in
+conjunction with the PERMANOVA results, suggest that cycle number can
+change bacterial community calculations but that these differences are
+minimal once 25 cycles are reached. Further, for some HiFI DNA
+polymerases, increasing the sub-sampling depth may reduce some of these
+observed community differences at lower cycle numbers.
 
@@
 consideration in helping to mitigate these observed changes due to high
-chimera prevalence in specific HiFi DNA polymerases (e.g.~Platinum).
+chimera prevalence in specific HiFi DNA polymerases.
 
@@
 the material safety data sheet (MSDS), the differences between the two
-mixes is not immediately apparent. Both Accuprime and Platinum contain a
-recombinant \emph{Taq} DNA polymerase, a \emph{Pycrococcus} spp GB-D
+is not immediately apparent. Both Accuprime and Platinum contain a
+recombinant \emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D
 polymerase and a platinum \emph{Taq} antibody. It is possible that

### 2018-02-20 submission/manuscript.Rmd
update conclusion

@@
 
-Our findings show that measures that rely on the number of OTUs will be dependent on both HiFi DNA polymerase and cycle number chosen.
-Due in part to these consderations, studies investigating similar areas may not be easily generalized with each other.
+Our findings show that the number of OTUs will be dependent on both HiFi DNA polymerase and cycle number chosen.
 Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.

### 2018-02-20 submission/manuscript.tex
update conclusion

@@
 
-Our findings show that measures that rely on the number of OTUs will be
-dependent on both HiFi DNA polymerase and cycle number chosen. Due in
-part to these consderations, studies investigating similar areas may not
-be easily generalized with each other. Care should be taken when
+Our findings show that the number of OTUs will be dependent on both HiFi
+DNA polymerase and cycle number chosen. Care should be taken when
 choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic

### 2018-02-20 submission/supplemental.Rmd
remove what the taq component
update with bringing the lead to the fore
update supplemental table headings

@@
 
-# What the Taq? The Influence of Different Hi-Fidelity Taq Polymerase on 16S rRNA Sequencing
+# The Influence of Different High Fidelity DNA Polymerase on 16S rRNA Sequencing
 
@@
 
-**Table S1: ANOVA Results of Number of OTUs Differences between HiFi Taq in Fecal Samples**
+**Table S1: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal Samples**
 
@@
 
-**Table S2: Tukey Post-Hoc Results of Number of OTUs Differences between HiFi Taq in Fecal Samples**
+**Table S2: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal Samples**
 
@@
 
-**Table S3: ANOVA Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
+**Table S3: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S4: Tukey Post-Hoc Results of Number of OTUs Differences between HiFi Taq in Mock Samples**
+**Table S4: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S5: Kruskal-Wallis Results of Per Base Error Rate Differences between HiFi Taq in Mock Samples**
+**Table S5: Kruskal-Wallis Results of Differences in Per Base Error Rate between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S6: Dunn's Post-Hoc Results of Per Base Error Rate between HiFi Taq in Mock Samples**
+**Table S6: Dunn's Post-Hoc Results of Differences in Per Base Error Rate between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S7: Kruskal-Wallis Results of Sequences with Error Differences between HiFi Taq in Mock Samples**
+**Table S7: Kruskal-Wallis Results of Differences in Sequences with Error between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S8: Dunn's Post-Hoc Results of Sequences with Error Differences between HiFi Taq in Mock Samples**
+**Table S8: Dunn's Post-Hoc Results of Differences in Sequences with Error between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S9: Kruskal-Wallis Results of Chimera Differences between HiFi Taq in Mock Samples**
+**Table S9: Kruskal-Wallis Results of Differences in Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
 
@@
 
-**Table S10: Dunn's Post-Hoc Results of Chimera Differences between HiFi Taq in Mock Samples**
+**Table S10: Dunn's Post-Hoc Results of Differences in Chimera Prevalence between HiFi DNA Polymerase in Mock Samples**
 

### 2018-02-23 submission/manuscript.Rmd
results updating

@@
 There was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1].
-Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1].
-All sub-sampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
-There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-Platinum HiFi DNA polymerase was the main reason for the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase versus higher cycle numbers (25, 30, and 35)  [Figure 1].
+At 35 cycles, all sub-sampling levels had significant differences between HiFi DNA polymerases (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+Some differences between HiFi DNA polymerase we identifed at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
+Most of the differences observed at 35 cycles were between Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2]
 
 This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases having high (Platinum) and low (Accuprime) number of OTUs [Figure 2 & Table S3].
-Differences between HiFi DNA polymerase were observed as early as 20 cycles and a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-Using a Tukey post-hoc test differences between Platinum and the other HiFi DNA polymerases was the major driver of the differences seen at different cycle numbers and sub-sampling depths [Table S4].
-Both fecal and mock samples consistently showed that across sub-sampling depth and cycle number the lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum [Figure 1 & 2].
-Based on these observations it is clear that HiFi DNA polymerase affects the total number of OTUs observed within a sample.
+In contrast to fecal samples, differences between HiFi DNA polymerase were observed as early as 20 cycles and as low as a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+For both different cycle numbers and sub-sampling depths, the majority of differences were between Platinum and the other HiFi DNA polymerases [Table S4].
+The lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum and this was consistent for both fecal and mock samples across sub-sampling depth and cycle number [Figure 1 & 2].
+Based on these observations it is clear that different HiFi DNA polymerase result in different total number of OTUs observed within a sample.
 
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
-For both fecal and mock samples, there was little difference between each respective 5-cycle increment at higher cycle numbers and this was consistent across the different sub-samplings [Figure 3].
-There were only two differences between 5-cycle increments that were identified.
-First, there were differences for the same fecal sample between 20 versus 25 cycles that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Overall, there were a few sub-sampling dependent differences in bacterial communities between cycle number .
+First, independent of HiFi DNA polymerase there were differences for the same fecal sample between 20 versus 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
 Second, where data is available for the mock communities, there were larger difference between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
-However, these differences between the next 5-cycle increment do not persist once 25 cycles are reached [Figure 3].
+Third, using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
+Phusion was one of only two HiFi DNA polymerases that that had fecal samples for the 1000 sub-sampling depth at 15 cycles.
+These stated differences between the next 5-cycle increment do not persist once 25 cycles are reached [Figure 3].
 
-Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Phusion was also one of only two HiFi DNA polymerases that that had samples for the 1000 sub-sampling depth at 15 cycles.
-Next, we assessed whether there were any major differences between 5-cycle increments within each sample.
-We found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+Next, we assessed whether there were any differences between 5-cycle increments within each individual or sample replicate.
+For both fecal and mock samples, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
 However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
-It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5-cycle increments within an individual.
+Similar to the PERMANOVA analysis, at higher sub-sampling depths these differences in Bray-Curtis indices disappear.
 Overall, HiFi DNA polymerase differences in Bray-Curtis index are dependent on both sub-sampling depth and cycle number.
@@
 This error rate was minimally affected by both the 'pre.cluster' step and chimera removal by VSEARCH [Figure 4].
-There were small differences in the per base error rate between the various HiFi DNA polymerase at lower cycle numbers and larger differences at higher cycle number with Platinum having the largest differences of all the HiFi DNA polymerase [Figure 4B-C and Table S6].
-
-The total sequences with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the 'pre.cluster' step [Figure S1].
-Major differences before this 'pre.cluster' step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8].
-Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1].
-However, this increased number of sequences with an error was lowered with existing bioinformatic approaches [Figure S1].
+The differences in the per base error rate between the various HiFi DNA polymerase was cycle dependent with Platinum having the largest differences versus other HiFi DNA polymerases [Figure 4B-C and Table S6].
+The total sequences with at least one error had was also cycle number dependent and mostly alleviated by the use of the 'pre.cluster' step [Figure S1].
+Differences in sequences with an error before this 'pre.cluster' step were due to large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerases [Figure S1 & Table S7 & S8].
 Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were no biases in the types of substitution made [Figure S2].
-Although sequence error is dependent on HiFi DNA polymerase it can be corrected using exsiting bioinformatic approaches.
+Although though sequence error is dependent on HiFi DNA polymerase it can be corrected using exsiting bioinformatic approaches.
 
 ***Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with Number of OTUs:***
-At all levels of sub-sampling and cycle number there were significant differences between the HiFi DNA polymerase used and chimera prevalence (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Differences between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences independent of cycle number and sub-sampling depth when using a Dunn's post-hoc test [Table S10].
+At all levels of sub-sampling and cycle numbers there were significant differences in the chimera prevalence based on HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Differences between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences and was independent of cycle number and sub-sampling depth [Table S10].
 Across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH was used [Figure 5].
@@
 For all HiFi DNA polymerases, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
-The R^2^ value between the number of OTUs and chimeric sequences did not change from the use of 'pre.cluster' and the removal of chimeras with VSEARCH [Figure 6].
+The R^2^ value between the number of OTUs and chimeric sequences did not change with the use of 'pre.cluster' or with the removal of chimeras using VSEARCH [Figure 6].
 Taken together, this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.

### 2018-02-23 submission/manuscript.tex
results updating

@@
 across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
-to result in less differences between HiFI DNA polymerase while cycle
-numbers of 25, 30, and 35 tended to have more distinct differences
-{[}Figure 1{]}. All sub-sampling levels were significantly different
-only at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There
-were differences in HiFi DNA polymerase at 25 and 30 cycles but the
-sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
-{[}Table S1{]}. Platinum HiFi DNA polymerase was the main reason for the
-differences observed across all sub-sampling depths at 35 cycles, based
-on a Tukey post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}.
+to result in less differences between HiFI DNA polymerase versus higher
+cycle numbers (25, 30, and 35) {[}Figure 1{]}. At 35 cycles, all
+sub-sampling levels had significant differences between HiFi DNA
+polymerases (P-value \textless{} 0.0001) {[}Table S1{]}. Some
+differences between HiFi DNA polymerase we identifed at 25 and 30 cycles
+but the sub-sampling depth had to be 5000 or higher (P-value \textless{}
+0.05) {[}Table S1{]}. Most of the differences observed at 35 cycles were
+between Platinum and other HiFi DNA polymerases, based on a Tukey
+post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}
 
@@
 polymerases having high (Platinum) and low (Accuprime) number of OTUs
-{[}Figure 2 \& Table S3{]}. Differences between HiFi DNA polymerase were
-observed as early as 20 cycles and a sub-sampling depth of 1000
-sequences (P-value = 0.002) {[}Table S3{]}. Using a Tukey post-hoc test
-differences between Platinum and the other HiFi DNA polymerases was the
-major driver of the differences seen at different cycle numbers and
-sub-sampling depths {[}Table S4{]}. Both fecal and mock samples
-consistently showed that across sub-sampling depth and cycle number the
-lowest number of OTUs identified was from Accuprime\textsuperscript{TM}
-while the highest was from Platinum {[}Figure 1 \& 2{]}. Based on these
-observations it is clear that HiFi DNA polymerase affects the total
-number of OTUs observed within a sample.
+{[}Figure 2 \& Table S3{]}. In contrast to fecal samples, differences
+between HiFi DNA polymerase were observed as early as 20 cycles and as
+low as a sub-sampling depth of 1000 sequences (P-value = 0.002) {[}Table
+S3{]}. For both different cycle numbers and sub-sampling depths, the
+majority of differences were between Platinum and the other HiFi DNA
+polymerases {[}Table S4{]}. The lowest number of OTUs identified was
+from Accuprime\textsuperscript{TM} while the highest was from Platinum
+and this was consistent for both fecal and mock samples across
+sub-sampling depth and cycle number {[}Figure 1 \& 2{]}. Based on these
+observations it is clear that different HiFi DNA polymerase result in
+different total number of OTUs observed within a sample.
 
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} For both fecal and mock samples, there was little difference
-between each respective 5-cycle increment at higher cycle numbers and
-this was consistent across the different sub-samplings {[}Figure 3{]}.
-There were only two differences between 5-cycle increments that were
-identified. First, there were differences for the same fecal sample
-between 20 versus 25 cycles that was independent of HiFi DNA polymerase
-but dependent on sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 -
-0.79) (median (25\% - 75\% quantile)), sub-sampled to 5000 = 0.43 (0.33
-- 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}.
-Second, where data is available for the mock communities, there were
-larger difference between 20 and 25 cycles (sub-sampled to 1000 = 0.88
-(0.42 - 0.91)) {[}Figure 3B{]}. However, these differences between the
-next 5-cycle increment do not persist once 25 cycles are reached
-{[}Figure 3{]}.
+Number:}} Overall, there were a few sub-sampling dependent differences
+in bacterial communities between cycle number . First, independent of
+HiFi DNA polymerase there were differences for the same fecal sample
+between 20 versus 25 cycles that was dependent on sub-sampling depth
+(sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median (25\% - 75\%
+quantile)), sub-sampled to 5000 = 0.43 (0.33 - 0.63), sub-sampled to
+10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Second, where data is
+available for the mock communities, there were larger difference between
+20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 - 0.91)) {[}Figure
+3B{]}. Third, using PERMANOVA to test for differences within HiFi DNA
+polymerase groups based on cycle number, only Phusion had cycle
+dependent differences at 1000 and 5000 sub-sampling depth (P-value =
+0.03 and 0.01, respectively). Phusion was one of only two HiFi DNA
+polymerases that that had fecal samples for the 1000 sub-sampling depth
+at 15 cycles. These stated differences between the next 5-cycle
+increment do not persist once 25 cycles are reached {[}Figure 3{]}.
 
-Using PERMANOVA to test for differences within HiFi DNA polymerase
-groups based on cycle number, only Phusion had cycle dependent
-differences at 1000 and 5000 sub-sampling depth (P-value = 0.03 and
-0.01, respectively). Phusion was also one of only two HiFi DNA
-polymerases that that had samples for the 1000 sub-sampling depth at 15
-cycles. Next, we assessed whether there were any major differences
-between 5-cycle increments within each sample. We found that there was
-no detectable difference in Bray-Curtis index when comparing to the
-previous 5-cycle increment (P-value \textgreater{} 0.05). However,
-Phusion at 1000 sub-sampling depth had a P-value = 0.02 before multiple
-comparison correction. It should be noted that at higher sub-sampling
-depths these differences in Bray-Curtis indices disappear for both
-differences in cycle number and within 5-cycle increments within an
-individual. Overall, HiFi DNA polymerase differences in Bray-Curtis
-index are dependent on both sub-sampling depth and cycle number.
+Next, we assessed whether there were any differences between 5-cycle
+increments within each individual or sample replicate. For both fecal
+and mock samples, we found that there was no detectable difference in
+Bray-Curtis index when comparing to the previous 5-cycle increment
+(P-value \textgreater{} 0.05). However, Phusion at 1000 sub-sampling
+depth had a P-value = 0.02 before multiple comparison correction.
+Similar to the PERMANOVA analysis, at higher sub-sampling depths these
+differences in Bray-Curtis indices disappear. Overall, HiFi DNA
+polymerase differences in Bray-Curtis index are dependent on both
+sub-sampling depth and cycle number.
 
@@
 error rate was minimally affected by both the `pre.cluster' step and
-chimera removal by VSEARCH {[}Figure 4{]}. There were small differences
-in the per base error rate between the various HiFi DNA polymerase at
-lower cycle numbers and larger differences at higher cycle number with
-Platinum having the largest differences of all the HiFi DNA polymerase
-{[}Figure 4B-C and Table S6{]}.
-
-The total sequences with at least one error had multiple differences at
-different cycle numbers and was mostly alleviated by the use of the
-`pre.cluster' step {[}Figure S1{]}. Major differences before this
-`pre.cluster' step were driven by large differences in
-Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
-polymerase tested {[}Figure S1 \& Table S7 \& S8{]}. Although
-Accuprime\textsuperscript{TM} had the lowest per base error rate it had
-the largest number of sequences with at least one error, regardless of
-cycle number or sub-sampling depth {[}Figure S1{]}. However, this
-increased number of sequences with an error was lowered with existing
-bioinformatic approaches {[}Figure S1{]}. Investigation of whether there
-were HiFi DNA polymerase dependent effects on base substitution found
-that there were no biases in the types of substitution made {[}Figure
-S2{]}. Although sequence error is dependent on HiFi DNA polymerase it
-can be corrected using exsiting bioinformatic approaches.
+chimera removal by VSEARCH {[}Figure 4{]}. The differences in the per
+base error rate between the various HiFi DNA polymerase was cycle
+dependent with Platinum having the largest differences versus other HiFi
+DNA polymerases {[}Figure 4B-C and Table S6{]}. The total sequences with
+at least one error had was also cycle number dependent and mostly
+alleviated by the use of the `pre.cluster' step {[}Figure S1{]}.
+Differences in sequences with an error before this `pre.cluster' step
+were due to large differences in Accuprime\textsuperscript{TM} and
+Platinum versus the other HiFi DNA polymerases {[}Figure S1 \& Table S7
+\& S8{]}. Investigation of whether there were HiFi DNA polymerase
+dependent effects on base substitution found that there were no biases
+in the types of substitution made {[}Figure S2{]}. Although though
+sequence error is dependent on HiFi DNA polymerase it can be corrected
+using exsiting bioinformatic approaches.
 
@@
 Correlate with Number of OTUs:}} At all levels of sub-sampling and cycle
-number there were significant differences between the HiFi DNA
-polymerase used and chimera prevalence (P-value \textless{} 0.05)
-{[}Table S9{]}. Differences between Platinum and all other HiFi DNA
-polymerases accounted for the majority of these differences independent
-of cycle number and sub-sampling depth when using a Dunn's post-hoc test
-{[}Table S10{]}. Across sub-sampling depth and cycle number
-Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
-the HiFi DNA Polymerases regardless of whether `pre.cluster' or chimera
-removal with VSEARCH was used {[}Figure 5{]}.
+numbers there were significant differences in the chimera prevalence
+based on HiFi DNA polymerase used (P-value \textless{} 0.05) {[}Table
+S9{]}. Differences between Platinum and all other HiFi DNA polymerases
+accounted for the majority of these differences and was independent of
+cycle number and sub-sampling depth {[}Table S10{]}. Across sub-sampling
+depth and cycle number Accuprime\textsuperscript{TM} had the lowest
+chimera prevalence of all the HiFi DNA Polymerases regardless of whether
+`pre.cluster' or chimera removal with VSEARCH was used {[}Figure 5{]}.
 
@@
 {[}Figure 6{]}. The R\textsuperscript{2} value between the number of
-OTUs and chimeric sequences did not change from the use of `pre.cluster'
-and the removal of chimeras with VSEARCH {[}Figure 6{]}. Taken together,
-this data suggests that a strong correlation exists between the number
-of OTUs and the prevalence of chimeric sequences.
+OTUs and chimeric sequences did not change with the use of `pre.cluster'
+or with the removal of chimeras using VSEARCH {[}Figure 6{]}. Taken
+together, this data suggests that a strong correlation exists between
+the number of OTUs and the prevalence of chimeric sequences.
 

### 2018-03-13 submission/manuscript.Rmd
intro rewrite outline and starting incorporation

@@
 
-# The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene Sequencing
+# High Fidelity DNA Polymerase can Introduce Bias into 16S rRNA Gene Sequencing Results
 
@@
 
-**Background.** Research has identified various methodological steps along the 16S rRNA gene survey pipeline that can change microbial community results.
-Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias.
+**Background.** Many of the different steps used during 16S rRNA gene sequencing can introduce bias and by extension change the observed microbial community results.
+Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
 Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream bacterial community results.
@@
 
-Recently, there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017].
-Investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant focal point of this push towards standardization [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016].
-Due to this focus, 16S rRNA gene sequencing has been much maligned for introduced biases.
-Yet, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
-Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation.
-For these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.
+It is important to differentiate between bias, reproducibility, and standardization since often time these three can be confused and used interchangeably with each other.
+
+Need to talk about sources of bias (each of these could be a small paragraph)
+
+  preservation kits (overgrowth, selective killing, DNA/RNA degradation at RT, freeze thaw)
+
+  Within kits (e.g. contamination)
+
+  PCR (16S hypervariable region primer specificity, amplification bias, error, polymerase dependent differences)
+
+  Sequencing machines themselves (non-random indels, etc.)
+
+Many of these sources of bias are common to both 16S rRNA gene and metagenomic sequencing approaches [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
+However, here we focused primarily on cycle number and high fidelity (HiFi) DNA polymerase, two areas where introduced bias might have a larger influence on 16S rRNA gene sequencing than metagenomics.
+
+*Talk about previous research on PCR bias*
+
+A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
+This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016].
+It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
+There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
+*Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
+In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 
@@
 The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
-
-A recent study in *Nature Biotechnology* found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
-This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016].
-It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
-There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
-*Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 

### 2018-03-14 submission/manuscript.Rmd
update title
introduction re-write

@@
 
-# High Fidelity DNA Polymerase can Introduce Bias into 16S rRNA Gene Sequencing Results
+# High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene Sequencing Results
 
@@
 
-It is important to differentiate between bias, reproducibility, and standardization since often time these three can be confused and used interchangeably with each other.
+It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
+Bias can change the observed results in a way that is reproducible and standardized.
+For example, DNA extraction kits used for 16S rRNA gene sequencing biases your findings versus another group not using the same brand kit but both groups have reproducible results.
+Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias.
+Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.
 
-Need to talk about sources of bias (each of these could be a small paragraph)
+A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
+Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref].
+Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref].
+Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref].
+Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014].
+Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs].
+However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method used.
 
-  preservation kits (overgrowth, selective killing, DNA/RNA degradation at RT, freeze thaw)
-
-  Within kits (e.g. contamination)
-
-  PCR (16S hypervariable region primer specificity, amplification bias, error, polymerase dependent differences)
-
-  Sequencing machines themselves (non-random indels, etc.)
-
-Many of these sources of bias are common to both 16S rRNA gene and metagenomic sequencing approaches [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].
-However, here we focused primarily on cycle number and high fidelity (HiFi) DNA polymerase, two areas where introduced bias might have a larger influence on 16S rRNA gene sequencing than metagenomics.
-
-*Talk about previous research on PCR bias*
+Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs].
+Many of these sources of biases are made worse as cycle number increases.
+The selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref].
+Both amplification error and non-specific amplification (e.g.
+incorrect amplicon size products) also increases as cycle number increases [refs].
+Further, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension [@Haas2011].
+Even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen.
+There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias.
+With respect to primers, there are different binding affinities for different bacteria that are variable region dependent [refs].
+Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014].
+Interestingly, the influence these different DNA polymerases can have on the observed results have not been well studied like the previously mentioned sources of PCR-based bias.
 
 A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
-This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016].
-It is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results.
+This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016].
+It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases.
 There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
@@
 In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
-
-The most commonly studied biases are preservation media, storage conditions, and DNA extraction kits.
-Although DNA extraction kits have consistently been shown to add bias to downstream analysis the current literature on preservation media and storage conditions has been mixed, with some studies showing biases while others do not [@contamination_Salter2014; @metagenomcis_bias_Costea_2017; @preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015].
-The study of these specific biases has become so large, aggregating them all together has become a difficult task with some researchers providing resources to actively track new findings (e.g.
-Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/).
-It is evident that these three sources of bias have been extensively covered within the literature but they are not the only sources that should be critically examined.
-The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.
 

### 2018-03-14 submission/manuscript.tex
introduction re-write

@@
 
-\section{The Influence of High Fidelity DNA Polymerase on 16S rRNA Gene
-Sequencing}\label{the-influence-of-high-fidelity-dna-polymerase-on-16s-rrna-gene-sequencing}
+\section{High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene
+Sequencing
+Results}\label{high-fidelity-dna-polymerase-introduces-bias-into-16s-rrna-gene-sequencing-results}
 
@@
 
-\textbf{Background.} Research has identified various methodological
-steps along the 16S rRNA gene survey pipeline that can change microbial
-community results. Although cycle number and high fidelity (HiFi) DNA
-polymerase are studied less often, they are still important sources of
-bias. Here, we critically examine both cycle number and HiFi DNA
-polymerase for biases that may influence downstream bacterial community
-results.
+\textbf{Background.} Many of the different steps used during 16S rRNA
+gene sequencing can introduce bias and by extension change the observed
+microbial community results. Although cycle number and high fidelity
+(HiFi) DNA polymerase are studied less often, they are still important
+sources of bias in this workflow. Here, we critically examine both cycle
+number and HiFi DNA polymerase for biases that may influence downstream
+bacterial community results.
 
@@
 
-Recently, there has been an increasing focus on standardizing
-methodological approaches in microbiota research (Kim et al., 2017;
-Hugerth \& Andersson, 2017). Investigating ways that 16S rRNA gene
-surveys can be made more reproducible has been a predominant focal point
-of this push towards standardization (Lauber et al., 2010; Salter et
-al., 2014; Song et al., 2016; Gohl et al., 2016). Due to this focus, 16S
-rRNA gene sequencing has been much maligned for introduced biases. Yet,
-many of these same considerations also affect metagenomic sequencing
-(Nayfach \& Pollard, 2016; Costea et al., 2017). Between the two
-approaches similar bias considerations include preservation media,
-storage conditions, DNA extraction kit, PCR, and sequence library
-preparation. For these overlapping considerations, biases identified for
-16S rRNA gene sequencing will also likely influence metagenomic
-sequencing results.
+It is important to differentiate between bias, reproducibility, and
+standardization since often times these three can be confused and used
+interchangeably with each other. Bias can change the observed results in
+a way that is reproducible and standardized. For example, DNA extraction
+kits used for 16S rRNA gene sequencing biases your findings versus
+another group not using the same brand kit but both groups have
+reproducible results. Therefore, standardization of 16S rRNA gene
+sequencing methods to increase reproducibility can still be problematic
+due to bias. Determining how different aspects of a 16S rRNA gene
+sequencing workflow could bias the observed results is critical for the
+interpretation of specific studies within the context of the overall
+field.
 
-The most commonly studied biases are preservation media, storage
-conditions, and DNA extraction kits. Although DNA extraction kits have
-consistently been shown to add bias to downstream analysis the current
-literature on preservation media and storage conditions has been mixed,
-with some studies showing biases while others do not (Lauber et al.,
-2010; Dominianni et al., 2014; Salter et al., 2014; Sinha et al., 2015;
-Song et al., 2016; Luo et al., 2016; Bassis et al., 2017; Costea et al.,
-2017). The study of these specific biases has become so large,
-aggregating them all together has become a difficult task with some
-researchers providing resources to actively track new findings
-(e.g.~Microbiome Digest -
-\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
-It is evident that these three sources of bias have been extensively
-covered within the literature but they are not the only sources that
-should be critically examined. The type of DNA polymerase chosen could
-also have a wide ranging affect on downstream results due to error rates
-and chimeras that may not be easily resolved using bioinformatic
-approaches.
+A typical 16S rRNA gene sequencing workflow can be divided into
+preservation, extraction, PCR, and sequencing steps. Generally, not
+using a preservation media and leaving samples at room temperature has
+been shown to cause overgrowth of low abundance members of the fecal
+bacterial community {[}ref{]}. Similarily, this overgrowth could still
+occur if the preservation media does not adequately inhibit growth
+{[}ref{]}. Reports have also shown that changes in specific community
+members might occur due to differing susceptibility to freeze thaw
+cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
+has been shown to add community members and the contribution of these
+contaminant members grows larger with lower biomass samples (Salter et
+al., 2014). Recent studies have shown that the majority of these biases
+due to either preservation or extraction tend to be smaller than the
+overall biological signal being measured {[}refs{]}. However, the
+contribution of PCR bias to this overall workflow is not well
+characterized since these studies use the same PCR approach while
+varying preservation or extraction method used.
 
-A recent study in \emph{Nature Biotechnology} found clear differences
-between normal and high fidelity (HiFi) DNA polymerase and that you
-could reduce error and chimera generation by optimizing the PCR protocol
-(Gohl et al., 2016). This study also found that, regardless of DNA
-polymerase, the number of Operational Taxonomic Units (OTUs) generated
-were not easily removed using the authors chosen bioinformatic pipeline
-(Gohl et al., 2016). It is natural to extend this line of inquiry and
-ask whether different HiFi DNA polymerase contribute different biases to
-downstream sequencing results. There is some reason to think that this
-may be the case since many of these HiFi DNA polymerase come from
-different families (e.g. \emph{Taq} belongs to the family A polymerases)
-and may intrinsically have different error rates that cannot be
-completely removed with modifications (Ishino \& Ishino, 2014). In this
-study we critically examine if any of five different HiFi DNA
-polymerases introduce significant biases into 16S rRNA gene surveys, if
-this is a cycle dependent phenomenon, and whether they can be removed
-using a standard bioinformatic pipeline.
+Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
+important because a large body of literature shows that primers used,
+cycle number, selective amplification, amplification error, non-specific
+amplification, chimeras, and polymerase dependent differences can all
+change the observed results {[}refs{]}. Many of these sources of biases
+are made worse as cycle number increases. The selective amplification of
+the dominant sequence can increase the difference between the most and
+next abundant sequence, over-emphasizing the actual difference
+{[}ref{]}. Both amplification error and non-specific amplification
+(e.g.~incorrect amplicon size products) also increases as cycle number
+increases {[}refs{]}. Further, chimeras can form from an aborted
+extension followed by subsequent priming error and secondary extension
+(Haas et al., 2011). Even if it has not been explicitly shown to
+increase with cycle number it is easy to postulate that this does
+happen. There are also intrinsic properties to primers and DNA
+polymerases chosen that can introduce bias. With respect to primers,
+there are different binding affinities for different bacteria that are
+variable region dependent {[}refs{]}. Additionally, there are multiple
+families of DNA polymerases that have their own error rate and proof
+reading capcity (Ishino \& Ishino, 2014). Interestingly, the influence
+these different DNA polymerases can have on the observed results have
+not been well studied like the previously mentioned sources of PCR-based
+bias.
+
+A recent study found clear differences between normal and high fidelity
+(HiFi) DNA polymerase and that you could reduce error and chimera
+generation by optimizing the PCR protocol (Gohl et al., 2016). This
+study also found that, regardless of DNA polymerase, the number of
+Operational Taxonomic Units (OTUs) or taxa generated were not easily
+reduced using the authors chosen bioinformatic pipeline (Gohl et al.,
+2016). It is natural to extend this line of inquiry and ask if biases in
+the number of OTUs and chimeras are also dependent on HiFi DNA
+polymerases. There is some reason to think that this may be the case
+since many of these HiFi DNA polymerase come from different families
+(e.g. \emph{Taq} belongs to the family A polymerases) and may
+intrinsically have different error rates that cannot be completely
+removed with modifications (Ishino \& Ishino, 2014). In this study we
+critically examine if any of five different HiFi DNA polymerases
+introduce significant biases into 16S rRNA gene surveys, if this is a
+cycle dependent phenomenon, and whether they can be removed using a
+standard bioinformatic pipeline.
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-storage_Bassis_2017}{}
-Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
-Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
-samples and storage conditions on bacterial community profiles.
-\emph{BMC Microbiology} 17. DOI:
-\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
-
 \hypertarget{ref-benjamini_controlling_1995}{}
@@
 \href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
-
-\hypertarget{ref-metagenomcis_bias_Costea_2017}{}
-Costea PI., Zeller G., Sunagawa S., Pelletier E., Alberti A., Levenez
-F., Tramontano M., Driessen M., Hercog R., Jung F-E., Kultima JR.,
-Hayward MR., Coelho LP., Allen-Vercoe E., Bertrand L., Blaut M., Brown
-JRM., Carton T., Cools-Portier S., Daigneault M., Derrien M., Druesne
-A., Vos WM de., Finlay BB., Flint HJ., Guarner F., Hattori M., Heilig
-H., Luna RA., Hylckama Vlieg J van., Junick J., Klymiuk I., Langella P.,
-Chatelier EL., Mai V., Manichanh C., Martin JC., Mery C., Morita H.,
-O'Toole PW., Orvain C., Patil KR., Penders J., Persson S., Pons N.,
-Popova M., Salonen A., Saulnier D., Scott KP., Singh B., Slezak K.,
-Veiga P., Versalovic J., Zhao L., Zoetendal EG., Ehrlich SD., Dore J.,
-Bork P. 2017. Towards standards for human fecal sample processing in
-metagenomic studies. \emph{Nature Biotechnology}. DOI:
-\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.
-
-\hypertarget{ref-preservation_Dominianni_2014}{}
-Dominianni C., Wu J., Hayes RB., Ahn J. 2014. Comparison of methods for
-fecal microbiome biospecimen collection. \emph{BMC Microbiology} 14:103.
-DOI:
-\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.
 
@@
 
-\hypertarget{ref-review_Hugerth_2017}{}
-Hugerth LW., Andersson AF. 2017. Analysing microbial community
-composition through amplicon sequencing: From sampling to hypothesis
-testing. \emph{Frontiers in Microbiology} 8. DOI:
-\href{https://doi.org/10.3389/fmicb.2017.01561}{10.3389/fmicb.2017.01561}.
+\hypertarget{ref-Haas2011}{}
+Haas BJ., Gevers D., Earl AM., Feldgarden M., Ward DV., Giannoukos G.,
+Ciulla D., Tabbaa D., Highlander SK., Sodergren E., Methe B., DeSantis
+TZ., Petrosino JF., Knight R., and BWB. 2011. Chimeric 16S rRNA sequence
+formation and detection in sanger and 454-pyrosequenced PCR amplicons.
+\emph{Genome Research} 21:494--504. DOI:
+\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
 
@@
 \href{https://doi.org/10.3389/fmicb.2014.00465}{10.3389/fmicb.2014.00465}.
-
-\hypertarget{ref-review_Kim_2017}{}
-Kim D., Hofstaedter CE., Zhao C., Mattei L., Tanes C., Clarke E., Lauder
-A., Sherrill-Mix S., Chehoud C., Kelsen J., Conrad M., Collman RG.,
-Baldassano R., Bushman FD., Bittinger K. 2017. Optimizing methods and
-dodging pitfalls in microbiome research. \emph{Microbiome} 5. DOI:
-\href{https://doi.org/10.1186/s40168-017-0267-5}{10.1186/s40168-017-0267-5}.
 
@@
 
-\hypertarget{ref-storage_Lauber_2010}{}
-Lauber CL., Zhou N., Gordon JI., Knight R., Fierer N. 2010. Effect of
-storage conditions on the assessment of bacterial community structure in
-soil and human-associated samples. \emph{FEMS Microbiology Letters}
-307:80--86. DOI:
-\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.
-
-\hypertarget{ref-preservation_Luo_2016}{}
-Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
-E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
-2016. Effects of specimen collection methodologies and storage
-conditions on the short-term stability of oral microbiome taxonomy.
-\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
-\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
-
 \hypertarget{ref-bc_index_Minchin1987}{}
@@
 \href{https://doi.org/10.1007/bf00038690}{10.1007/bf00038690}.
-
-\hypertarget{ref-metagenomcis_bias_Nayfach_2016}{}
-Nayfach S., Pollard KS. 2016. Toward accurate and quantitative
-comparative metagenomics. \emph{Cell} 166:1103--1116. DOI:
-\href{https://doi.org/10.1016/j.cell.2016.08.007}{10.1016/j.cell.2016.08.007}.
 
@@
 \href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
-
-\hypertarget{ref-preservation_Sinha_2015}{}
-Sinha R., Chen J., Amir A., Vogtmann E., Shi J., Inman KS., Flores R.,
-Sampson J., Knight R., Chia N. 2015. Collecting fecal samples for
-microbiome analyses in epidemiology studies. \emph{Cancer Epidemiology
-Biomarkers \& Prevention} 25:407--416. DOI:
-\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.
-
-\hypertarget{ref-preservation_Song_2016}{}
-Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
-R. 2016. Preservation methods differ in fecal microbiome stability,
-affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
-DOI:
-\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
 

### 2018-03-14 submission/manuscript.Rmd
update intro paragraph 3 and 4

@@
 
+This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
 It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
 Bias can change the observed results in a way that is reproducible and standardized.
-For example, DNA extraction kits used for 16S rRNA gene sequencing biases your findings versus another group not using the same brand kit but both groups have reproducible results.
+For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results.
 Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias.
@@
 Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs].
-However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method used.
+However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
 
 Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs].
-Many of these sources of biases are made worse as cycle number increases.
-The selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref].
+Many of these sources of biases are made worse as cycle number increases [ref].
+For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref].
 Both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) also increases as cycle number increases [refs].
-Further, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension [@Haas2011].
-Even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen.
+incorrect amplicon size products) can also increase as cycle number increases [refs].
+Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen [@Haas2011].
 There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias.
-With respect to primers, there are different binding affinities for different bacteria that are variable region dependent [refs].
+With respect to primers, there are different binding affinities for different bacteria that are dependent on variable region [refs].
 Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014].
-Interestingly, the influence these different DNA polymerases can have on the observed results have not been well studied like the previously mentioned sources of PCR-based bias.
+Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.
 
-A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016].
+A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016].
 This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].

### 2018-03-14 submission/manuscript.tex
update intro paragraph 3 and 4

@@
 
-It is important to differentiate between bias, reproducibility, and
-standardization since often times these three can be confused and used
-interchangeably with each other. Bias can change the observed results in
-a way that is reproducible and standardized. For example, DNA extraction
-kits used for 16S rRNA gene sequencing biases your findings versus
-another group not using the same brand kit but both groups have
-reproducible results. Therefore, standardization of 16S rRNA gene
-sequencing methods to increase reproducibility can still be problematic
-due to bias. Determining how different aspects of a 16S rRNA gene
-sequencing workflow could bias the observed results is critical for the
-interpretation of specific studies within the context of the overall
-field.
+This study will specifically address how cycle number and high fidelity
+(HiFi) DNA polymerases can bias observed bacterial community results
+derived from 16S rRNA gene sequencing. It is important to differentiate
+between bias, reproducibility, and standardization since often times
+these three can be confused and used interchangeably with each other.
+Bias can change the observed results in a way that is reproducible and
+standardized. For example, if one group uses one brand of DNA extraction
+kit for their 16S rRNA gene sequencing, their results may be biased
+versus another group not using the same brand kit but within their group
+they can still have reproducible results. Therefore, standardization of
+16S rRNA gene sequencing methods to increase reproducibility can still
+be problematic due to bias. Determining how different aspects of a 16S
+rRNA gene sequencing workflow could bias the observed results is
+critical for the interpretation of specific studies within the context
+of the overall field.
 
@@
 characterized since these studies use the same PCR approach while
-varying preservation or extraction method used.
+varying preservation or extraction method.
 
@@
 change the observed results {[}refs{]}. Many of these sources of biases
-are made worse as cycle number increases. The selective amplification of
-the dominant sequence can increase the difference between the most and
-next abundant sequence, over-emphasizing the actual difference
-{[}ref{]}. Both amplification error and non-specific amplification
-(e.g.~incorrect amplicon size products) also increases as cycle number
-increases {[}refs{]}. Further, chimeras can form from an aborted
-extension followed by subsequent priming error and secondary extension
-(Haas et al., 2011). Even if it has not been explicitly shown to
-increase with cycle number it is easy to postulate that this does
-happen. There are also intrinsic properties to primers and DNA
-polymerases chosen that can introduce bias. With respect to primers,
-there are different binding affinities for different bacteria that are
-variable region dependent {[}refs{]}. Additionally, there are multiple
-families of DNA polymerases that have their own error rate and proof
-reading capcity (Ishino \& Ishino, 2014). Interestingly, the influence
-these different DNA polymerases can have on the observed results have
-not been well studied like the previously mentioned sources of PCR-based
-bias.
+are made worse as cycle number increases {[}ref{]}. For example, the
+selective amplification of the dominant sequence can increase the
+difference between the most and next abundant sequence, over-emphasizing
+the actual difference {[}ref{]}. Both amplification error and
+non-specific amplification (e.g.~incorrect amplicon size products) can
+also increase as cycle number increases {[}refs{]}. Additionally,
+chimeras can form from an aborted extension followed by subsequent
+priming error and secondary extension and even if it has not been
+explicitly shown to increase with cycle number it is easy to postulate
+that this does happen (Haas et al., 2011). There are also intrinsic
+properties to primers and DNA polymerases chosen that can introduce
+bias. With respect to primers, there are different binding affinities
+for different bacteria that are dependent on variable region {[}refs{]}.
+Additionally, there are multiple families of DNA polymerases that have
+their own error rate and proof reading capcity (Ishino \& Ishino, 2014).
+Interestingly, the influence that these different DNA polymerases can
+have on the observed results have not been well studied like some of the
+other previously mentioned sources of PCR-based bias.
 
 A recent study found clear differences between normal and high fidelity
-(HiFi) DNA polymerase and that you could reduce error and chimera
-generation by optimizing the PCR protocol (Gohl et al., 2016). This
-study also found that, regardless of DNA polymerase, the number of
-Operational Taxonomic Units (OTUs) or taxa generated were not easily
-reduced using the authors chosen bioinformatic pipeline (Gohl et al.,
-2016). It is natural to extend this line of inquiry and ask if biases in
-the number of OTUs and chimeras are also dependent on HiFi DNA
-polymerases. There is some reason to think that this may be the case
-since many of these HiFi DNA polymerase come from different families
-(e.g. \emph{Taq} belongs to the family A polymerases) and may
-intrinsically have different error rates that cannot be completely
-removed with modifications (Ishino \& Ishino, 2014). In this study we
-critically examine if any of five different HiFi DNA polymerases
-introduce significant biases into 16S rRNA gene surveys, if this is a
-cycle dependent phenomenon, and whether they can be removed using a
-standard bioinformatic pipeline.
+(HiFi) DNA polymerase and that optimization of the PCR protocol could
+reduce error and chimera generation (Gohl et al., 2016). This study also
+found that, regardless of DNA polymerase, the number of Operational
+Taxonomic Units (OTUs) or taxa generated were not easily reduced using
+the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
+natural to extend this line of inquiry and ask if biases in the number
+of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
+is some reason to think that this may be the case since many of these
+HiFi DNA polymerase come from different families (e.g. \emph{Taq}
+belongs to the family A polymerases) and may intrinsically have
+different error rates that cannot be completely removed with
+modifications (Ishino \& Ishino, 2014). In this study we critically
+examine if any of five different HiFi DNA polymerases introduce
+significant biases into 16S rRNA gene surveys, if this is a cycle
+dependent phenomenon, and whether they can be removed using a standard
+bioinformatic pipeline.
 

### 2018-03-14 submission/manuscript.Rmd
further edits to intro

@@
 This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
-It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
+First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
 Bias can change the observed results in a way that is reproducible and standardized.
@@
 Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias.
-Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.
+Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the borader context of the overall field.
 
@@
 
-Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs].
+Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [refs].
 Many of these sources of biases are made worse as cycle number increases [ref].
@@
 Both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) can also increase as cycle number increases [refs].
-Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen [@Haas2011].
+incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [refs].
+Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].
+
 There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias.
-With respect to primers, there are different binding affinities for different bacteria that are dependent on variable region [refs].
+Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g.
+V1-V3 does not detect *Haemophilus Influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [refs].
 Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014].
@@
 A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016].
-This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
+This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
 It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases.
-There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g.
+There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g.
 *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+In this study, we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 

### 2018-03-14 submission/manuscript.tex
further edits to intro

@@
 (HiFi) DNA polymerases can bias observed bacterial community results
-derived from 16S rRNA gene sequencing. It is important to differentiate
-between bias, reproducibility, and standardization since often times
-these three can be confused and used interchangeably with each other.
-Bias can change the observed results in a way that is reproducible and
-standardized. For example, if one group uses one brand of DNA extraction
-kit for their 16S rRNA gene sequencing, their results may be biased
-versus another group not using the same brand kit but within their group
-they can still have reproducible results. Therefore, standardization of
-16S rRNA gene sequencing methods to increase reproducibility can still
-be problematic due to bias. Determining how different aspects of a 16S
-rRNA gene sequencing workflow could bias the observed results is
-critical for the interpretation of specific studies within the context
-of the overall field.
+derived from 16S rRNA gene sequencing. First, it is important to
+differentiate between bias, reproducibility, and standardization since
+often times these three can be confused and used interchangeably with
+each other. Bias can change the observed results in a way that is
+reproducible and standardized. For example, if one group uses one brand
+of DNA extraction kit for their 16S rRNA gene sequencing, their results
+may be biased versus another group not using the same brand kit but
+within their group they can still have reproducible results. Therefore,
+standardization of 16S rRNA gene sequencing methods to increase
+reproducibility can still be problematic due to bias. Determining how
+different aspects of a 16S rRNA gene sequencing workflow could bias the
+observed results is critical for the interpretation of specific studies
+in the borader context of the overall field.
 
@@
 Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
-important because a large body of literature shows that primers used,
-cycle number, selective amplification, amplification error, non-specific
-amplification, chimeras, and polymerase dependent differences can all
-change the observed results {[}refs{]}. Many of these sources of biases
-are made worse as cycle number increases {[}ref{]}. For example, the
-selective amplification of the dominant sequence can increase the
-difference between the most and next abundant sequence, over-emphasizing
-the actual difference {[}ref{]}. Both amplification error and
-non-specific amplification (e.g.~incorrect amplicon size products) can
-also increase as cycle number increases {[}refs{]}. Additionally,
-chimeras can form from an aborted extension followed by subsequent
-priming error and secondary extension and even if it has not been
-explicitly shown to increase with cycle number it is easy to postulate
-that this does happen (Haas et al., 2011). There are also intrinsic
-properties to primers and DNA polymerases chosen that can introduce
-bias. With respect to primers, there are different binding affinities
-for different bacteria that are dependent on variable region {[}refs{]}.
-Additionally, there are multiple families of DNA polymerases that have
-their own error rate and proof reading capcity (Ishino \& Ishino, 2014).
-Interestingly, the influence that these different DNA polymerases can
-have on the observed results have not been well studied like some of the
-other previously mentioned sources of PCR-based bias.
+important because a large body of literature shows that there are a
+variety of steps during PCR that can change the observed results
+{[}refs{]}. Many of these sources of biases are made worse as cycle
+number increases {[}ref{]}. For example, the selective amplification of
+the dominant sequence can increase the difference between the most and
+next abundant sequence, over-emphasizing the actual difference
+{[}ref{]}. Both amplification error and non-specific amplification
+(e.g.~incorrect amplicon size products) can also increase as cycle
+number increases which can drastically change commonly used diversity
+measures {[}refs{]}. Additionally, chimeras can form from an aborted
+extension followed by subsequent priming error and secondary extension
+and even if it has not been explicitly shown to increase with cycle
+number it is easy to postulate that this does happen and will also
+artificially increase community diversity (Haas et al., 2011).
+
+There are also intrinsic properties to primers and DNA polymerases
+chosen that can introduce bias. Primers have variable region dependent
+binding affinities for different bacteria and depending on the primer
+pair do not detect specific bacteria (e.g.~V1-V3 does not detect
+\emph{Haemophilus Influenzae} and V3-V5 does not detect
+\emph{Propionibacterium acnes}) {[}refs{]}. Additionally, there are
+multiple families of DNA polymerases that have their own error rate and
+proof reading capcity (Ishino \& Ishino, 2014). Interestingly, the
+influence that these different DNA polymerases can have on the observed
+results have not been well studied like some of the other previously
+mentioned sources of PCR-based bias.
 
@@
 reduce error and chimera generation (Gohl et al., 2016). This study also
-found that, regardless of DNA polymerase, the number of Operational
+found that regardless of DNA polymerase, the number of Operational
 Taxonomic Units (OTUs) or taxa generated were not easily reduced using
@@
 is some reason to think that this may be the case since many of these
-HiFi DNA polymerase come from different families (e.g. \emph{Taq}
+HiFi DNA polymerases come from different families (e.g. \emph{Taq}
 belongs to the family A polymerases) and may intrinsically have
 different error rates that cannot be completely removed with
-modifications (Ishino \& Ishino, 2014). In this study we critically
+modifications (Ishino \& Ishino, 2014). In this study, we critically
 examine if any of five different HiFi DNA polymerases introduce

### 2018-03-15 submission/manuscript.Rmd
fix capitalization error
complete adding references

@@
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref].
-Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref].
-Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref].
+Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
+Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
+Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015].
 Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014].
-Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs].
+Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017].
 However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
 
-Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [refs].
-Many of these sources of biases are made worse as cycle number increases [ref].
-For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref].
+Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000].
+Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015].
+For example, the selective amplification of AT-rich over GC-rich sequences can increase the difference between these types of 16S rRNA genes and over-emphasizing the actual difference [@polz_bias_1998].
 Both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [refs].
+incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [@BautistadelosSantos2016; @Acinas2005].
 Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].
@@
 Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g.
-V1-V3 does not detect *Haemophilus Influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [refs].
+V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
 Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014].

### 2018-03-15 submission/manuscript.tex
fix capitalization error
complete adding references

@@
 been shown to cause overgrowth of low abundance members of the fecal
-bacterial community {[}ref{]}. Similarily, this overgrowth could still
-occur if the preservation media does not adequately inhibit growth
-{[}ref{]}. Reports have also shown that changes in specific community
-members might occur due to differing susceptibility to freeze thaw
-cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
-has been shown to add community members and the contribution of these
-contaminant members grows larger with lower biomass samples (Salter et
-al., 2014). Recent studies have shown that the majority of these biases
-due to either preservation or extraction tend to be smaller than the
-overall biological signal being measured {[}refs{]}. However, the
-contribution of PCR bias to this overall workflow is not well
-characterized since these studies use the same PCR approach while
-varying preservation or extraction method.
+bacterial community (Amir et al., 2017). Similarily, this overgrowth
+could still occur if the preservation media does not adequately inhibit
+growth (Song et al., 2016; Luo et al., 2016). Reports have also shown
+that changes in specific community members might occur due to differing
+susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
+2015). Additionally, reagent contamination has been shown to add
+community members and the contribution of these contaminant members
+grows larger with lower biomass samples (Salter et al., 2014). Recent
+studies have shown that the majority of these biases due to either
+preservation or extraction tend to be smaller than the overall
+biological signal being measured (Song et al., 2016; Bassis et al.,
+2017). However, the contribution of PCR bias to this overall workflow is
+not well characterized since these studies use the same PCR approach
+while varying preservation or extraction method.
 
@@
 important because a large body of literature shows that there are a
-variety of steps during PCR that can change the observed results
-{[}refs{]}. Many of these sources of biases are made worse as cycle
-number increases {[}ref{]}. For example, the selective amplification of
-the dominant sequence can increase the difference between the most and
-next abundant sequence, over-emphasizing the actual difference
-{[}ref{]}. Both amplification error and non-specific amplification
-(e.g.~incorrect amplicon size products) can also increase as cycle
-number increases which can drastically change commonly used diversity
-measures {[}refs{]}. Additionally, chimeras can form from an aborted
-extension followed by subsequent priming error and secondary extension
-and even if it has not been explicitly shown to increase with cycle
-number it is easy to postulate that this does happen and will also
+variety of steps during PCR that can change the observed results (Eckert
+\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
+made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
+2011; Kebschull \& Zador, 2015). For example, the selective
+amplification of AT-rich over GC-rich sequences can increase the
+difference between these types of 16S rRNA genes and over-emphasizing
+the actual difference (Polz \& Cavanaugh, 1998). Both amplification
+error and non-specific amplification (e.g.~incorrect amplicon size
+products) can also increase as cycle number increases which can
+drastically change commonly used diversity measures (Acinas et al.,
+2005; Santos et al., 2016). Additionally, chimeras can form from an
+aborted extension followed by subsequent priming error and secondary
+extension and even if it has not been explicitly shown to increase with
+cycle number it is easy to postulate that this does happen and will also
 artificially increase community diversity (Haas et al., 2011).
@@
 pair do not detect specific bacteria (e.g.~V1-V3 does not detect
-\emph{Haemophilus Influenzae} and V3-V5 does not detect
-\emph{Propionibacterium acnes}) {[}refs{]}. Additionally, there are
-multiple families of DNA polymerases that have their own error rate and
-proof reading capcity (Ishino \& Ishino, 2014). Interestingly, the
-influence that these different DNA polymerases can have on the observed
-results have not been well studied like some of the other previously
-mentioned sources of PCR-based bias.
+\emph{Haemophilus influenzae} and V3-V5 does not detect
+\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
+al., 2016). Additionally, there are multiple families of DNA polymerases
+that have their own error rate and proof reading capcity (Ishino \&
+Ishino, 2014). Interestingly, the influence that these different DNA
+polymerases can have on the observed results have not been well studied
+like some of the other previously mentioned sources of PCR-based bias.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Acinas2005}{}
+Acinas SG., Sarma-Rupavtarm R., Klepac-Ceraj V., Polz MF. 2005.
+PCR-induced sequence artifacts and bias: Insights from comparison of two
+16S rRNA clone libraries constructed from the same sample. \emph{Applied
+and Environmental Microbiology} 71:8966--8969. DOI:
+\href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
+
+\hypertarget{ref-Amir2017}{}
+Amir A., McDonald D., Navas-Molina JA., Debelius J., Morton JT., Hyde
+E., Robbins-Pianka A., Knight R. 2017. Correcting for microbial blooms
+in fecal samples during room-temperature shipping. \emph{mSystems}
+2:e00199--16. DOI:
+\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.
+
+\hypertarget{ref-storage_Bassis_2017}{}
+Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
+Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
+samples and storage conditions on bacterial community profiles.
+\emph{BMC Microbiology} 17. DOI:
+\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
+
 \hypertarget{ref-benjamini_controlling_1995}{}
@@
 the Royal Statistical Society. Series B (Methodological)} 57:289--300.
+
+\hypertarget{ref-Burkardt2000}{}
+Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
+\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
+\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.
 
@@
 \href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+
+\hypertarget{ref-Eckert1991}{}
+Eckert KA., Kunkel TA. 1991. DNA polymerase fidelity and the polymerase
+chain reaction. \emph{Genome Research} 1:17--24. DOI:
+\href{https://doi.org/10.1101/gr.1.1.17}{10.1101/gr.1.1.17}.
 
@@
 
+\hypertarget{ref-Gorzelak2015}{}
+Gorzelak MA., Gill SK., Tasnim N., Ahmadi-Vand Z., Jay M., Gibson DL.
+2015. Methods for improving human gut microbiome data by reducing
+variability through sample processing and storage of stool. \emph{PLOS
+ONE} 10:e0134802. DOI:
+\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.
+
 \hypertarget{ref-Haas2011}{}
@@
 
+\hypertarget{ref-Kebschull2015}{}
+Kebschull JM., Zador AM. 2015. Sources of PCR-induced distortions in
+high-throughput sequencing data sets. \emph{Nucleic Acids
+Research}:gkv717. DOI:
+\href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
+
 \hypertarget{ref-protocol_Kozich_2013}{}
@@
 DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-preservation_Luo_2016}{}
+Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
+E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
+2016. Effects of specimen collection methodologies and storage
+conditions on the short-term stability of oral microbiome taxonomy.
+\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
+\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
+
+\hypertarget{ref-Meisel2016}{}
+Meisel JS., Hannigan GD., Tyldsley AS., SanMiguel AJ., Hodkinson BP.,
+Zheng Q., Grice EA. 2016. Skin microbiome surveys are strongly
+influenced by experimental design. \emph{Journal of Investigative
+Dermatology} 136:947--956. DOI:
+\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.
 
@@
 Szoecs E., Wagner H. 2017. \emph{Vegan: Community ecology package}.
+
+\hypertarget{ref-polz_bias_1998}{}
+Polz MF., Cavanaugh CM. 1998. Bias in template-to-product ratios in
+multitemplate PCR. \emph{Applied and Environmental Microbiology}
+64:3724--3730.
 
@@
 
+\hypertarget{ref-BautistadelosSantos2016}{}
+Santos QMB-d los., Schroeder JL., Blakemore O., Moses J., Haffey M.,
+Sloan W., Pinto AJ. 2016. The impact of sampling, PCR, and sequencing
+replication on discerning changes in drinking water bacterial community
+over diurnal time-scales. \emph{Water Research} 90:216--224. DOI:
+\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.
+
 \hypertarget{ref-mothur_schloss_2009}{}
@@
 \href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+
+\hypertarget{ref-preservation_Song_2016}{}
+Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
+R. 2016. Preservation methods differ in fecal microbiome stability,
+affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
+DOI:
+\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
+
+\hypertarget{ref-Sze2015}{}
+Sze MA., Dimitriu PA., Suzuki M., McDonough JE., Campbell JD., Brothers
+JF., Erb-Downward JR., Huffnagle GB., Hayashi S., Elliott WM., Cooper
+J., Sin DD., Lenburg ME., Spira A., Mohn WW., Hogg JC. 2015. Host
+response to the lung microbiome in chronic obstructive pulmonary
+disease. \emph{American Journal of Respiratory and Critical Care
+Medicine} 192:438--445. DOI:
+\href{https://doi.org/10.1164/rccm.201502-0223oc}{10.1164/rccm.201502-0223oc}.
+
+\hypertarget{ref-Wang1996}{}
+Wang GCY., Wang Y. 1996. The frequency of chimeric molecules as a
+consequence of PCR co-amplification of 16S rRNA genes from different
+bacterial species. \emph{Microbiology} 142:1107--1114. DOI:
+\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
 

### 2018-03-16 submission/manuscript.Rmd
a new re-write of the abstract

@@
 
-**Background.** Many of the different steps used during 16S rRNA gene sequencing can introduce bias and by extension change the observed microbial community results.
+**Background.** Using different reagents and kits at different steps of the 16S rRNA gene sequencing workflow can introduce bias by changing the observed microbial community.
 Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
-Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream bacterial community results.
+Here, we examine how both cycle number and HiFi DNA polymerase  can change the bacterial community and introduce bias to the final obtained results.
 
-**Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach.
-First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities.
+**Methods.** DNA from fecal samples (n=4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+Mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach.
+The number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities.
 Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated.
@@
 
-**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH.
-Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs and this association was not changed by chimera removal with VSEARCH.
+**Results.** When analyzing fecal samples a different number of OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+These HiFi dependent differences in the number of OTUs were identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+Chimera prevalence varied by HiFi DNA polymerase and this variation was still present after chimera removal using VSEARCH.
+The chimera prevalence was positively correlated with the number of OTUs and was also not affected by chimera removal with VSEARCH.
 
-**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have values that are not comparable across studies.
+**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult.
+Care should be exercised when choosing both HiFi DNA polymerase and cycle number to be used in 16S rRNA gene sequencing studies.
 

### 2018-03-16 submission/manuscript.tex
a new re-write of the abstract

@@
 
-\textbf{Background.} Many of the different steps used during 16S rRNA
-gene sequencing can introduce bias and by extension change the observed
-microbial community results. Although cycle number and high fidelity
-(HiFi) DNA polymerase are studied less often, they are still important
-sources of bias in this workflow. Here, we critically examine both cycle
-number and HiFi DNA polymerase for biases that may influence downstream
-bacterial community results.
+\textbf{Background.} Using different reagents and kits at different
+steps of the 16S rRNA gene sequencing workflow can introduce bias by
+changing the observed microbial community. Although cycle number and
+high fidelity (HiFi) DNA polymerase are studied less often, they are
+still important sources of bias in this workflow. Here, we examine how
+both cycle number and HiFi DNA polymerase can change the bacterial
+community and introduce bias to the final obtained results.
 
-\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
+\textbf{Methods.} DNA from fecal samples (n=4) were extracted using a
 PowerMag DNA extraction kit with a 10 minute bead beating step and
@@
 Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
-(technical replicates n = 4) consisting of previously isolated whole
+(technical replicates n=4) consisting of previously isolated whole
 genomes of 8 different bacteria were also amplified using the same
-approach. First, the number of OTUs (Operational Taxonomic Units) were
-examined for both fecal samples and mock communities. Next, Bray-Curtis
-index, the error rate, sequence error prevalence, and chimera prevalence
-were investigated. Finally, the chimera prevalence correlation with
-number of OTUs was assessed.
+approach. The number of OTUs (Operational Taxonomic Units) were examined
+for both fecal samples and mock communities. Next, Bray-Curtis index,
+the error rate, sequence error prevalence, and chimera prevalence were
+investigated. Finally, the chimera prevalence correlation with number of
+OTUs was assessed.
 
-\textbf{Results.} At 35 cycles there were significant differences
-between HiFi DNA polymerase for fecal samples (P-value \textless{}
-0.0001). These HiFi dependent differences in the number of OTUs could be
-identified as early as 20 cycles in the mock communities (P-value =
-0.002). Chimera prevalence varied by HiFi DNA polymerase and these
-differences were still observed after chimera removal using VSEARCH.
-Additionally, the chimera prevalence had a strong positive correlation
-with the number of OTUs and this association was not changed by chimera
-removal with VSEARCH.
+\textbf{Results.} When analyzing fecal samples a different number of
+OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value
+\textless{} 0.0001). These HiFi dependent differences in the number of
+OTUs were identified as early as 20 cycles in the mock communities
+(P-value = 0.002). Chimera prevalence varied by HiFi DNA polymerase and
+this variation was still present after chimera removal using VSEARCH.
+The chimera prevalence was positively correlated with the number of OTUs
+and was also not affected by chimera removal with VSEARCH.
 
-\textbf{Conclusions.} Due to HiFi DNA polymerase dependent differences
-in the number of OTUs and chimera prevalence, common diversity metrics
-could have values that are not comparable across studies.
+\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
+number of OTUs and chimera prevalence makes comparison across studies
+difficult. Care should be exercised when choosing both HiFi DNA
+polymerase and cycle number to be used in 16S rRNA gene sequencing
+studies.
 

### 2018-03-16 submission/manuscript.Rmd
fix typos in intro

@@
 Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias.
-Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the borader context of the overall field.
+Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the broader context of the overall field.
 
@@
 Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
-Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
+Similarily, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
 Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015].
@@
 Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015].
-For example, the selective amplification of AT-rich over GC-rich sequences can increase the difference between these types of 16S rRNA genes and over-emphasizing the actual difference [@polz_bias_1998].
+For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998].
 Both amplification error and non-specific amplification (e.g.
 incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
-Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].
+Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011].
 
@@
 V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
-Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014].
-Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.
+Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
+Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.
 
@@
 This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
-It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases.
+It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of HiFi DNA polymerase.
 There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g.
 *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-In this study, we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+In this study, we examine if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 

### 2018-03-16 submission/manuscript.tex
fix typos in intro

@@
 observed results is critical for the interpretation of specific studies
-in the borader context of the overall field.
+in the broader context of the overall field.
 
@@
 been shown to cause overgrowth of low abundance members of the fecal
-bacterial community (Amir et al., 2017). Similarily, this overgrowth
-could still occur if the preservation media does not adequately inhibit
-growth (Song et al., 2016; Luo et al., 2016). Reports have also shown
-that changes in specific community members might occur due to differing
+bacterial community (Amir et al., 2017). Similarily, this overgrowth can
+still occur if the preservation media does not adequately inhibit growth
+(Song et al., 2016; Luo et al., 2016). Reports have also shown that
+changes in specific community members might occur due to differing
 susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
@@
 2011; Kebschull \& Zador, 2015). For example, the selective
-amplification of AT-rich over GC-rich sequences can increase the
-difference between these types of 16S rRNA genes and over-emphasizing
-the actual difference (Polz \& Cavanaugh, 1998). Both amplification
-error and non-specific amplification (e.g.~incorrect amplicon size
-products) can also increase as cycle number increases which can
-drastically change commonly used diversity measures (Acinas et al.,
-2005; Santos et al., 2016). Additionally, chimeras can form from an
-aborted extension followed by subsequent priming error and secondary
-extension and even if it has not been explicitly shown to increase with
-cycle number it is easy to postulate that this does happen and will also
+amplification of AT-rich over GC-rich sequences can exaggerate the
+difference between 16S rRNA genes higher in AT versus those higher in GC
+(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
+amplification (e.g.~incorrect amplicon size products) can also increase
+as cycle number increases which can drastically change commonly used
+diversity measures (Acinas et al., 2005; Santos et al., 2016).
+Additionally, chimeras can form from an aborted extension step followed
+by a subsequent priming error and secondary extension and will also
 artificially increase community diversity (Haas et al., 2011).
@@
 al., 2016). Additionally, there are multiple families of DNA polymerases
-that have their own error rate and proof reading capcity (Ishino \&
+that have their own error rate and proof reading capacity (Ishino \&
 Ishino, 2014). Interestingly, the influence that these different DNA
-polymerases can have on the observed results have not been well studied
-like some of the other previously mentioned sources of PCR-based bias.
+polymerases can have on the observed 16S rRNA gene sequencing results
+have not been well studied like some of the other previously mentioned
+sources of PCR-based bias.
 
@@
 natural to extend this line of inquiry and ask if biases in the number
-of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
-is some reason to think that this may be the case since many of these
-HiFi DNA polymerases come from different families (e.g. \emph{Taq}
-belongs to the family A polymerases) and may intrinsically have
-different error rates that cannot be completely removed with
-modifications (Ishino \& Ishino, 2014). In this study, we critically
-examine if any of five different HiFi DNA polymerases introduce
+of OTUs and chimeras are also dependent on the type of HiFi DNA
+polymerase. There is some reason to think that this may be the case
+since many of these HiFi DNA polymerases come from different families
+(e.g. \emph{Taq} belongs to the family A polymerases) and may
+intrinsically have different error rates that cannot be completely
+removed with modifications (Ishino \& Ishino, 2014). In this study, we
+examine if any of five different types of HiFi DNA polymerases introduce
 significant biases into 16S rRNA gene surveys, if this is a cycle

### 2018-03-16 submission/manuscript.Rmd
add method changes --trying to make more clear

@@
 A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
-The processing and storage of these samples have been previously published [@erin_seekatz_2016].
+The processing and storage of these samples were previously published [@erin_seekatz_2016].
 Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study.
 All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
 
@@
 The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified.
-For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016].
+For Kappa and Q5, a previously published protocol was used [@taq_Gohl_2016].
 For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.
 
-Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
-The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total samples = 100).
-The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130).
+The cycle conditions for both fecal and mock samples started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
+The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total sample n=100).
+The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerases (total samples n=130).
 No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
@@
 ***Sequence Processing:***
-The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009].
-Generally, the protocol followed what has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
+The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Sequence error was determined using the 'seq.error' command on mock samples before the 'pre.cluster' command, before chimera removal, and after chimera removal [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
 
 ***Analysis Workflow:***
-The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g.
+The total number of OTUs was analyzed after sub-sampling for both the fecal and mock community samples.
+For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g.
 20 versus 25, 25 versus 30, etc.).
-For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
-Next, based on these observations we wanted to analyze potential reasons for these differences.
-First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each HiFi DNA polymerase.
-After assessing these errors, the total number of chimeras was determined after sequence processing and correlations to the number of OTUs also assessed.
+These community based measures for fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
+Based on these observations we analyzed potential reasons for these differences.
+Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal.
+Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal.
 

### 2018-03-16 submission/manuscript.tex
add method changes --trying to make more clear

@@
 Investigational Network (ERIN). The processing and storage of these
-samples have been previously published (Seekatz et al., 2016). Other
-than confirmation that none of these individuals had a \emph{Clostridium
+samples were previously published (Seekatz et al., 2016). Other than
+confirmation that none of these individuals had a \emph{Clostridium
 difficle} infection, clinical data and other types of meta data were not
@@
 (now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
-Community DNA Standard (Zymo, CA, USA) was used in this study and is
-made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
+Community DNA Standard (Zymo, CA, USA) was used for mock communities and
+was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
 \emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
@@
 The HiFi DNA polymerase activation time was 2 minutes, unless a
-different activation was specified. For Kappa and Q5, the protocol
-previously published by Gohl and colleagues was used (Gohl et al.,
-2016). For Phusion, the company defined conditions were used except for
-extension time, where the Accuprime and Platinum settings were used.
+different activation was specified. For Kappa and Q5, a previously
+published protocol was used (Gohl et al., 2016). For Phusion, the
+company defined conditions were used except for extension time, where
+the Accuprime and Platinum settings were used.
 
-Both fecal and mock samples cycle conditions started at 15 and increased
-by 5 up to 35 cycles with amplicons used at each 5-step increase for
-sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
-and 35 cycles for each HiFi DNA polymerase (total samples = 100). The
-mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10
-replicates for 30 cycles for all HiFi DNA polymerase (total samples =
-130). No mock community sample had enough PCR product at 15 cycles for
-adequate 16S rRNA gene sequencing.
+The cycle conditions for both fecal and mock samples started at 15 and
+increased by 5 up to 35 cycles with amplicons used at each 5-step
+increase for sequencing. The PCR of fecal DNA samples consisted of all 4
+samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase
+(total sample n=100). The mock communities had 4 replicates at 15, 20,
+25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA
+polymerases (total samples n=130). No mock community sample had enough
+PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
 
 \textbf{\emph{Sequence Processing:}} The mothur software program was
-utilized for all sequence processing steps (Schloss et al., 2009).
-Generally, the protocol followed what has been previously published
-(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
-major differences from the stated protocol were the use of VSEARCH
-instead of UCHIME for chimera detection and the use of the OptiClust
-algorithm instead of average neighbor for OTU generation at 97\%
-similarity (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
-Schloss, 2017). Sequence error was determined using the `seq.error'
-command on mock samples after chimera removal and classification to the
-RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
-al., 2013; Rognes et al., 2016).
+used for all sequence processing steps (Schloss et al., 2009). The
+protocol has been previously published (Kozich et al., 2013)
+(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
+from the published protocol were the use of VSEARCH instead of UCHIME
+for chimera detection and the use of the OptiClust algorithm instead of
+average neighbor for OTU generation at 97\% similarity (Edgar et al.,
+2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
+was determined using the `seq.error' command on mock samples before the
+`pre.cluster' command, before chimera removal, and after chimera removal
+(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
 
-\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
-sub-sampling was analyzed for both the fecal and mock community samples.
-Cycle dependent affects on Bray-Curtis indices were next assessed for
-the fecal samples looking at both overall cycle differences and within
-individual differences for the previous cycle (e.g.~20 versus 25, 25
-versus 30, etc.). For these community based measures, the fecal samples
-were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
-10000, and 15000) while the mock community samples were analysed at 3
-levels (1000, 5000, 10000). Next, based on these observations we wanted
-to analyze potential reasons for these differences. First, analysis of
-general sequence error rate, number of sequences with an error, and base
-substitution were assessed in the mock community for each HiFi DNA
-polymerase. After assessing these errors, the total number of chimeras
-was determined after sequence processing and correlations to the number
-of OTUs also assessed.
+\textbf{\emph{Analysis Workflow:}} The total number of OTUs was analyzed
+after sub-sampling for both the fecal and mock community samples. For
+fecal samples, cycle dependent affects on Bray-Curtis indices were
+assessed for cycle group and within individual differences from the
+previous cycle (e.g.~20 versus 25, 25 versus 30, etc.). These community
+based measures for fecal samples were analyzed at 4 different
+sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the
+mock community samples were analysed at 3 levels (1000, 5000, 10000).
+Based on these observations we analyzed potential reasons for these
+differences. Analysis of the mock community of each HiFi DNA polymerase
+for general sequence error rate, number of sequences with an error, base
+substitution, and numbers of chimeras were assessed before the
+`pre.cluster' command, before chimera removal, and after chimera
+removal. Additionally, the correlation between the number of chimeras
+and the number of OTUs was also assessed before the `pre.cluster'
+command, before chimera removal, and after chimera removal.
 

### 2018-03-16 submission/manuscript.Rmd
update and change first section of results

@@
 
-***The Number of OTUs is Dependent on HiFi DNA Polymerase:***
-There was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1].
-Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase versus higher cycle numbers (25, 30, and 35)  [Figure 1].
-At 35 cycles, all sub-sampling levels had significant differences between HiFi DNA polymerases (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
-Some differences between HiFi DNA polymerase we identifed at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-Most of the differences observed at 35 cycles were between Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2]
+***The Number of OTUs are Dependent on HiFi DNA Polymerase:***
+A consistent difference in the number of OTUs, that was dependent on the HiFi DNA polymerase used was observed regardless of sub-sampling depth for fecal samples [Figure 1].
+Additionally, there was a trend for lower cycle numbers (15-20) to result in less differences in the number of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
+For fecal samples, all sub-sampling levels had significant differences between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+Most of the differences observed at 35 cycles were between Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Differences in the number of OTUs between HiFi DNA polymerases were identifiable at earlier cycles (25 and 30) but the sub-sampling depth had to be 5000 sequences or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
 
-This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples with the same DNA polymerases having high (Platinum) and low (Accuprime) number of OTUs [Figure 2 & Table S3].
-In contrast to fecal samples, differences between HiFi DNA polymerase were observed as early as 20 cycles and as low as a sub-sampling depth of 1000 sequences (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-For both different cycle numbers and sub-sampling depths, the majority of differences were between Platinum and the other HiFi DNA polymerases [Table S4].
-The lowest number of OTUs identified was from Accuprime^TM^ while the highest was from Platinum and this was consistent for both fecal and mock samples across sub-sampling depth and cycle number [Figure 1 & 2].
-Based on these observations it is clear that different HiFi DNA polymerase result in different total number of OTUs observed within a sample.
+This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples [Figure 2].
+Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across cycle number and sub-sampling depth [Figure 1-2 & Table S1-S4].
+In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles and at as low of a sub-sampling depth as 1000 sequences in the mock community samples (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+For both cycle numbers and sub-sampling depths, the majority of  differences in the number of OTUs were between Platinum and the other HiFi DNA polymerases [Table S4].
+Based on these observations in fecal and mock communities, it is clear that different HiFi DNA polymerases result in a different total number of OTUs observed within a sample.
 

### 2018-03-16 submission/manuscript.tex
update and change first section of results

@@
 
-\textbf{\emph{The Number of OTUs is Dependent on HiFi DNA Polymerase:}}
-There was a HiFi DNA polymerase dependent difference that was consistent
-across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
-to result in less differences between HiFI DNA polymerase versus higher
-cycle numbers (25, 30, and 35) {[}Figure 1{]}. At 35 cycles, all
-sub-sampling levels had significant differences between HiFi DNA
-polymerases (P-value \textless{} 0.0001) {[}Table S1{]}. Some
-differences between HiFi DNA polymerase we identifed at 25 and 30 cycles
-but the sub-sampling depth had to be 5000 or higher (P-value \textless{}
-0.05) {[}Table S1{]}. Most of the differences observed at 35 cycles were
-between Platinum and other HiFi DNA polymerases, based on a Tukey
-post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}
+\textbf{\emph{The Number of OTUs are Dependent on HiFi DNA Polymerase:}}
+A consistent difference in the number of OTUs, that was dependent on the
+HiFi DNA polymerase used was observed regardless of sub-sampling depth
+for fecal samples {[}Figure 1{]}. Additionally, there was a trend for
+lower cycle numbers (15-20) to result in less differences in the number
+of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA
+polymerases {[}Figure 1{]}. For fecal samples, all sub-sampling levels
+had significant differences between HiFi DNA polymerases at 35 cycles
+(P-value \textless{} 0.0001) {[}Table S1{]}. Most of the differences
+observed at 35 cycles were between Platinum and other HiFi DNA
+polymerases, based on a Tukey post-hoc test (P-value \textless{} 0.05)
+{[}Table S2{]}. Differences in the number of OTUs between HiFi DNA
+polymerases were identifiable at earlier cycles (25 and 30) but the
+sub-sampling depth had to be 5000 sequences or higher (P-value
+\textless{} 0.05) {[}Table S1{]}.
 
 This HiFi DNA polymerase dependent difference in the number of OTUs was
-also observed in the mock community samples with the same DNA
-polymerases having high (Platinum) and low (Accuprime) number of OTUs
-{[}Figure 2 \& Table S3{]}. In contrast to fecal samples, differences
-between HiFi DNA polymerase were observed as early as 20 cycles and as
-low as a sub-sampling depth of 1000 sequences (P-value = 0.002) {[}Table
-S3{]}. For both different cycle numbers and sub-sampling depths, the
-majority of differences were between Platinum and the other HiFi DNA
-polymerases {[}Table S4{]}. The lowest number of OTUs identified was
-from Accuprime\textsuperscript{TM} while the highest was from Platinum
-and this was consistent for both fecal and mock samples across
-sub-sampling depth and cycle number {[}Figure 1 \& 2{]}. Based on these
-observations it is clear that different HiFi DNA polymerase result in
+also observed in the mock community samples {[}Figure 2{]}. Regardless
+if fecal or mock communities were used, the same HiFi DNA polymerases
+had high (Platinum) and low (Accuprime) number of OTUs and this was
+consistent across cycle number and sub-sampling depth {[}Figure 1-2 \&
+Table S1-S4{]}. In contrast to the results obtained with fecal samples,
+differences between HiFi DNA polymerases were observed as early as 20
+cycles and at as low of a sub-sampling depth as 1000 sequences in the
+mock community samples (P-value = 0.002) {[}Table S3{]}. For both cycle
+numbers and sub-sampling depths, the majority of differences in the
+number of OTUs were between Platinum and the other HiFi DNA polymerases
+{[}Table S4{]}. Based on these observations in fecal and mock
+communities, it is clear that different HiFi DNA polymerases result in a
 different total number of OTUs observed within a sample.

### 2018-03-16 submission/manuscript.Rmd
update first bray-curtis section results

@@
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
-Overall, there were a few sub-sampling dependent differences in bacterial communities between cycle number .
-First, independent of HiFi DNA polymerase there were differences for the same fecal sample between 20 versus 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Second, where data is available for the mock communities, there were larger difference between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
-Third, using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Phusion was one of only two HiFi DNA polymerases that that had fecal samples for the 1000 sub-sampling depth at 15 cycles.
-These stated differences between the next 5-cycle increment do not persist once 25 cycles are reached [Figure 3].
+A few small differences based on sub-sampling and cycle number were detected in bacterial communities.
+Within the same fecal sample and independent of HiFi DNA polymerases, there were differences in the community composition between 20 and 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Further, where data was available for the mock communities, there were larger observed differences between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
+Additionally, these stated community differences disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
+Finally, using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depths (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
+Phusion was one of two HiFi DNA polymerases that had enough sequences to reach a sub-sampling depth of 1000 at 15 cycles for fecal samples.
 

### 2018-03-16 submission/manuscript.tex
update first bray-curtis section results

@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} Overall, there were a few sub-sampling dependent differences
-in bacterial communities between cycle number . First, independent of
-HiFi DNA polymerase there were differences for the same fecal sample
-between 20 versus 25 cycles that was dependent on sub-sampling depth
-(sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median (25\% - 75\%
-quantile)), sub-sampled to 5000 = 0.43 (0.33 - 0.63), sub-sampled to
-10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Second, where data is
-available for the mock communities, there were larger difference between
-20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 - 0.91)) {[}Figure
-3B{]}. Third, using PERMANOVA to test for differences within HiFi DNA
-polymerase groups based on cycle number, only Phusion had cycle
-dependent differences at 1000 and 5000 sub-sampling depth (P-value =
-0.03 and 0.01, respectively). Phusion was one of only two HiFi DNA
-polymerases that that had fecal samples for the 1000 sub-sampling depth
-at 15 cycles. These stated differences between the next 5-cycle
-increment do not persist once 25 cycles are reached {[}Figure 3{]}.
+Number:}} A few small differences based on sub-sampling and cycle number
+were detected in bacterial communities. Within the same fecal sample and
+independent of HiFi DNA polymerases, there were differences in the
+community composition between 20 and 25 cycles that was dependent on
+sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median
+(25\% - 75\% quantile)), sub-sampled to 5000 = 0.43 (0.33 - 0.63),
+sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further,
+where data was available for the mock communities, there were larger
+observed differences between 20 and 25 cycles (sub-sampled to 1000 =
+0.88 (0.42 - 0.91)) {[}Figure 3B{]}. Additionally, these stated
+community differences disappear when comparing 25 to 30 cycles and do
+not persist past 25 cycles {[}Figure 3{]}. Finally, using PERMANOVA to
+test for community differences based on cycle number within HiFi DNA
+polymerases, only Phusion had cycle dependent differences at 1000 and
+5000 sub-sampling depths (P-value = 0.03 and 0.01, respectively).
+Phusion was one of two HiFi DNA polymerases that had enough sequences to
+reach a sub-sampling depth of 1000 at 15 cycles for fecal samples.
 

### 2018-03-19 submission/manuscript.Rmd
update bray-curtis results section

@@
 ***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
-A few small differences based on sub-sampling and cycle number were detected in bacterial communities.
-Within the same fecal sample and independent of HiFi DNA polymerases, there were differences in the community composition between 20 and 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Further, where data was available for the mock communities, there were larger observed differences between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
+A few small differences based on sub-sampling and cycle number were detected in overall bacterial community composition.
+Within the same fecal sample and independent of HiFi DNA polymerases, there were differences in the community composition between 20 and 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (IQR)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
+Further, when data was available for the mock communities, there were larger observed differences between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
 Additionally, these stated community differences disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
-Finally, using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depths (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively).
-Phusion was one of two HiFi DNA polymerases that had enough sequences to reach a sub-sampling depth of 1000 at 15 cycles for fecal samples.
-
-Next, we assessed whether there were any differences between 5-cycle increments within each individual or sample replicate.
-For both fecal and mock samples, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
-However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction.
-Similar to the PERMANOVA analysis, at higher sub-sampling depths these differences in Bray-Curtis indices disappear.
-Overall, HiFi DNA polymerase differences in Bray-Curtis index are dependent on both sub-sampling depth and cycle number.
+Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+Using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depths (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
+For fecal samples, Phusion was one of two HiFi DNA polymerases that had enough sequences to reach a sub-sampling depth of 1000 at 15 cycles.
+Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on both sub-sampling depth and cycle number.
 

### 2018-03-19 submission/manuscript.tex
update bray-curtis results section

@@
 Number:}} A few small differences based on sub-sampling and cycle number
-were detected in bacterial communities. Within the same fecal sample and
-independent of HiFi DNA polymerases, there were differences in the
-community composition between 20 and 25 cycles that was dependent on
-sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median
-(25\% - 75\% quantile)), sub-sampled to 5000 = 0.43 (0.33 - 0.63),
-sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further,
-where data was available for the mock communities, there were larger
-observed differences between 20 and 25 cycles (sub-sampled to 1000 =
-0.88 (0.42 - 0.91)) {[}Figure 3B{]}. Additionally, these stated
-community differences disappear when comparing 25 to 30 cycles and do
-not persist past 25 cycles {[}Figure 3{]}. Finally, using PERMANOVA to
-test for community differences based on cycle number within HiFi DNA
-polymerases, only Phusion had cycle dependent differences at 1000 and
-5000 sub-sampling depths (P-value = 0.03 and 0.01, respectively).
-Phusion was one of two HiFi DNA polymerases that had enough sequences to
-reach a sub-sampling depth of 1000 at 15 cycles for fecal samples.
-
-Next, we assessed whether there were any differences between 5-cycle
-increments within each individual or sample replicate. For both fecal
-and mock samples, we found that there was no detectable difference in
-Bray-Curtis index when comparing to the previous 5-cycle increment
-(P-value \textgreater{} 0.05). However, Phusion at 1000 sub-sampling
-depth had a P-value = 0.02 before multiple comparison correction.
-Similar to the PERMANOVA analysis, at higher sub-sampling depths these
-differences in Bray-Curtis indices disappear. Overall, HiFi DNA
-polymerase differences in Bray-Curtis index are dependent on both
+were detected in overall bacterial community composition. Within the
+same fecal sample and independent of HiFi DNA polymerases, there were
+differences in the community composition between 20 and 25 cycles that
+was dependent on sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 -
+0.79) (median (IQR)), sub-sampled to 5000 = 0.43 (0.33 - 0.63),
+sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further, when
+data was available for the mock communities, there were larger observed
+differences between 20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 -
+0.91)) {[}Figure 3B{]}. Additionally, these stated community differences
+disappear when comparing 25 to 30 cycles and do not persist past 25
+cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
+found that there was no detectable difference in Bray-Curtis index when
+comparing to the previous 5-cycle increment for both fecal and mock
+communities after multiple comparison correction (P-value \textgreater{}
+0.05). Using PERMANOVA to test for community differences based on cycle
+number within HiFi DNA polymerases, only Phusion had cycle dependent
+differences at 1000 and 5000 sub-sampling depths (P-value = 0.03 and
+0.01). For fecal samples, Phusion was one of two HiFi DNA polymerases
+that had enough sequences to reach a sub-sampling depth of 1000 at 15
+cycles. Overall, these data suggest that there are small HiFi DNA
+polymerase differences in Bray-Curtis index that are dependent on both
 sub-sampling depth and cycle number.

### 2018-03-19 submission/manuscript.Rmd
change bray section title

@@
 
-***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
+***Minimal Bray-Curtis Differences are Detected and dependent on both Cycle Number and Sub-Sampling Depth:***
 A few small differences based on sub-sampling and cycle number were detected in overall bacterial community composition.

### 2018-03-19 code/run_nmds_mock_data.R
temp code for hifi bray comparison

@@ -219,6 +219,69 @@ get_permanova <- function(depth, metaList, dataList){
 }
 
 
+# Create a function to compare within HiFi composition to between Hifi composition
+compare_hifi <- function(called_hifi, cycle_number, cycle_compare, sub_sampling, 
+                         ind_samples, metaList, distList){
+  
+  overall_metadata <- metaList[[as.character(sub_sampling)]]
+  test <- distList[[as.character(sub_sampling)]]
+  
+  test_meta <- as.data.frame(overall_metadata %>% filter(cycles == cycle_number))
+
+
+  temp_within_hifi <- test[filter(test_meta, taq == called_hifi)[, "full_name"],
+                           as.data.frame(filter(overall_metadata,
+                                                cycles %in% cycle_compare,
+                                                taq == called_hifi))[, "full_name"]]
+  
+  if(called_hifi == "ACC"){
+    
+    temp_within_hifi <- as.data.frame(temp_within_hifi) %>% 
+      rename('30x_ACC_Mock_DNA1' = Zmock_A_111716, 
+             '30x_ACC_Mock_DNA2' = Zmock_B_111716, 
+             '30x_ACC_Mock_DNA3' = Zmock_C_111716, 
+             '30x_ACC_Mock_DNA4' = Zmock_D_111716)
+  }
+
+  cycle_within_hifi <- sapply(cycle_compare,
+                              function(x)
+                                sapply(ind_samples,
+                                       function(y)
+                                         as.numeric(
+                                           temp_within_hifi[paste(cycle_number, "_", called_hifi,
+                                                                  "_Mock_", y, sep = ""),
+                                                            paste(x, "_", called_hifi,
+                                                                  "_Mock_", y, sep = "")])))
+
+
+  temp_other_hifi <- test[filter(test_meta, taq == called_hifi)[, "full_name"],
+                          filter(test_meta, taq != called_hifi, taq != "K")[, "full_name"]]
+
+  hifi_compare <- unique(overall_metadata$taq)[!unique(overall_metadata$taq) %in% c("K",called_hifi)]
+
+  other_hifi <- sapply(hifi_compare,
+                       function(x) sapply(ind_samples,
+                                          function(y)
+                                            as.numeric(temp_other_hifi[paste(cycle_number, "_", called_hifi,
+                                                                             "_Mock_", y, sep = ""),
+                                                                       paste(cycle_number, "_", x,
+                                                                             "_Mock_", y, sep = "")])))
+
[…]

### 2018-03-19 submission/manuscript.Rmd
complete fixes for result section

--- 2018-03-19 submission/manuscript.Rmd
@@
 
-***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
+***Minimal Bray-Curtis Differences are Detected and are Dependent on both Cycle Number and Sub-Sampling Depth:***
 A few small differences based on sub-sampling and cycle number were detected in overall bacterial community composition.
@@
 ***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number:***
-Differences by HiFi DNA polymerase in the median average per base error varied without a clear pattern across sub-sampling depth [Table S5].
-The highest per base median average error rates were for the Kappa HiFi DNA polymerase [Figure 4].
+Differences in the median average per base error varied by HiFi DNA polymerase without a clear pattern across sub-sampling depth [Table S5].
+The highest median average per base error rates were for the Kappa HiFi DNA polymerase [Figure 4].
 This error rate was minimally affected by both the 'pre.cluster' step and chimera removal by VSEARCH [Figure 4].
-The differences in the per base error rate between the various HiFi DNA polymerase was cycle dependent with Platinum having the largest differences versus other HiFi DNA polymerases [Figure 4B-C and Table S6].
-The total sequences with at least one error had was also cycle number dependent and mostly alleviated by the use of the 'pre.cluster' step [Figure S1].
-Differences in sequences with an error before this 'pre.cluster' step were due to large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerases [Figure S1 & Table S7 & S8].
-Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were no biases in the types of substitution made [Figure S2].
-Although though sequence error is dependent on HiFi DNA polymerase it can be corrected using exsiting bioinformatic approaches.
+The differences in the median average per base error rate between the different HiFi DNA polymerases was cycle dependent with Platinum having the largest changes versus other HiFi DNA polymerases [Figure 4B-C and Table S6].
+The total sequences with at least one error was also cycle number dependent and differences between HiFi DNA polymerases could be drastically reduced by the use of the 'pre.cluster' step [Figure S1].
+These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerases [Figure S1 & Table S7 & S8].
+Finally, we did not observe a HiFi DNA polymerase dependent difference on base substitution rate [Figure S2].
+Although sequence error is dependent on HiFi DNA polymerase some of these error dependent differences can be corrected using exsiting bioinformatic approaches.
 
-***Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with Number of OTUs:***
-At all levels of sub-sampling and cycle numbers there were significant differences in the chimera prevalence based on HiFi DNA polymerase used (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Differences between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences and was independent of cycle number and sub-sampling depth [Table S10].
-Across sub-sampling depth and cycle number Accuprime^TM^ had the lowest chimera prevalence of all the HiFi DNA Polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH was used [Figure 5].
-
-For all HiFi DNA polymerases, a positive correlation was observed between chimeric sequences and number of OTUs, with this correlation being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase [Figure 6].
+***Prevalence of Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with the Number of OTUs:***
+There were significant differences in the chimera prevalence based on HiFi DNA polymerase used at all levels of sub-sampling and cycle numbers (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Differences in chimera prevalence between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences [Table S10].
+Accuprime^TM^ had the lowest chimera prevalence of all HiFi DNA polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH was used [Figure 5].
+A positive correlation was observed between chimeric sequences and the number of OTUs for all HiFi DNA polymerases [Figure 6].
+This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion HiFi DNA Polymerases [Figure 6].
 The R^2^ value between the number of OTUs and chimeric sequences did not change with the use of 'pre.cluster' or with the removal of chimeras using VSEARCH [Figure 6].
-Taken together, this data suggests that a strong correlation exists between the number of OTUs and the prevalence of chimeric sequences.
+This data suggests that chimera prevalence depends on HiFi DNA polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 

### 2018-03-19 submission/manuscript.tex
complete fixes for result section

@@
 
-\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
-Number:}} A few small differences based on sub-sampling and cycle number
-were detected in overall bacterial community composition. Within the
-same fecal sample and independent of HiFi DNA polymerases, there were
-differences in the community composition between 20 and 25 cycles that
-was dependent on sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 -
-0.79) (median (IQR)), sub-sampled to 5000 = 0.43 (0.33 - 0.63),
-sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further, when
-data was available for the mock communities, there were larger observed
-differences between 20 and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 -
-0.91)) {[}Figure 3B{]}. Additionally, these stated community differences
-disappear when comparing 25 to 30 cycles and do not persist past 25
-cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
-found that there was no detectable difference in Bray-Curtis index when
-comparing to the previous 5-cycle increment for both fecal and mock
-communities after multiple comparison correction (P-value \textgreater{}
-0.05). Using PERMANOVA to test for community differences based on cycle
-number within HiFi DNA polymerases, only Phusion had cycle dependent
-differences at 1000 and 5000 sub-sampling depths (P-value = 0.03 and
-0.01). For fecal samples, Phusion was one of two HiFi DNA polymerases
-that had enough sequences to reach a sub-sampling depth of 1000 at 15
-cycles. Overall, these data suggest that there are small HiFi DNA
-polymerase differences in Bray-Curtis index that are dependent on both
-sub-sampling depth and cycle number.
+\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
+Dependent on both Cycle Number and Sub-Sampling Depth:}} A few small
+differences based on sub-sampling and cycle number were detected in
+overall bacterial community composition. Within the same fecal sample
+and independent of HiFi DNA polymerases, there were differences in the
+community composition between 20 and 25 cycles that was dependent on
+sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median
+(IQR)), sub-sampled to 5000 = 0.43 (0.33 - 0.63), sub-sampled to 10000 =
+0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further, when data was available for
+the mock communities, there were larger observed differences between 20
+and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 - 0.91)) {[}Figure
+3B{]}. Additionally, these stated community differences disappear when
+comparing 25 to 30 cycles and do not persist past 25 cycles {[}Figure
+3{]}. Although these trends are clearly noticeable, we found that there
+was no detectable difference in Bray-Curtis index when comparing to the
+previous 5-cycle increment for both fecal and mock communities after
+multiple comparison correction (P-value \textgreater{} 0.05). Using
+PERMANOVA to test for community differences based on cycle number within
+HiFi DNA polymerases, only Phusion had cycle dependent differences at
+1000 and 5000 sub-sampling depths (P-value = 0.03 and 0.01). For fecal
+samples, Phusion was one of two HiFi DNA polymerases that had enough
+sequences to reach a sub-sampling depth of 1000 at 15 cycles. Overall,
+these data suggest that there are small HiFi DNA polymerase differences
+in Bray-Curtis index that are dependent on both sub-sampling depth and
+cycle number.
 
 \textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
-and Cycle Number:}} Differences by HiFi DNA polymerase in the median
-average per base error varied without a clear pattern across
-sub-sampling depth {[}Table S5{]}. The highest per base median average
+and Cycle Number:}} Differences in the median average per base error
+varied by HiFi DNA polymerase without a clear pattern across
+sub-sampling depth {[}Table S5{]}. The highest median average per base
 error rates were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. This
 error rate was minimally affected by both the `pre.cluster' step and
-chimera removal by VSEARCH {[}Figure 4{]}. The differences in the per
-base error rate between the various HiFi DNA polymerase was cycle
-dependent with Platinum having the largest differences versus other HiFi
-DNA polymerases {[}Figure 4B-C and Table S6{]}. The total sequences with
-at least one error had was also cycle number dependent and mostly
-alleviated by the use of the `pre.cluster' step {[}Figure S1{]}.
-Differences in sequences with an error before this `pre.cluster' step
-were due to large differences in Accuprime\textsuperscript{TM} and
-Platinum versus the other HiFi DNA polymerases {[}Figure S1 \& Table S7
-\& S8{]}. Investigation of whether there were HiFi DNA polymerase
-dependent effects on base substitution found that there were no biases
-in the types of substitution made {[}Figure S2{]}. Although though
-sequence error is dependent on HiFi DNA polymerase it can be corrected
+chimera removal by VSEARCH {[}Figure 4{]}. The differences in the median
+average per base error rate between the different HiFi DNA polymerases
+was cycle dependent with Platinum having the largest changes versus
+other HiFi DNA polymerases {[}Figure 4B-C and Table S6{]}. The total
+sequences with at least one error was also cycle number dependent and
+differences between HiFi DNA polymerases could be drastically reduced by
+the use of the `pre.cluster' step {[}Figure S1{]}. These differences in
+sequences with at least one error were mostly due to differences in
+Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
+polymerases {[}Figure S1 \& Table S7 \& S8{]}. Finally, we did not
+observe a HiFi DNA polymerase dependent difference on base substitution
+rate {[}Figure S2{]}. Although sequence error is dependent on HiFi DNA
+polymerase some of these error dependent differences can be corrected
 using exsiting bioinformatic approaches.
 
-\textbf{\emph{Chimeric Sequences are HiFi DNA Polymerase Dependent and
-Correlate with Number of OTUs:}} At all levels of sub-sampling and cycle
-numbers there were significant differences in the chimera prevalence
-based on HiFi DNA polymerase used (P-value \textless{} 0.05) {[}Table
-S9{]}. Differences between Platinum and all other HiFi DNA polymerases
-accounted for the majority of these differences and was independent of
-cycle number and sub-sampling depth {[}Table S10{]}. Across sub-sampling
-depth and cycle number Accuprime\textsuperscript{TM} had the lowest
-chimera prevalence of all the HiFi DNA Polymerases regardless of whether
-`pre.cluster' or chimera removal with VSEARCH was used {[}Figure 5{]}.
-
-For all HiFi DNA polymerases, a positive correlation was observed
-between chimeric sequences and number of OTUs, with this correlation
-being strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
-{[}Figure 6{]}. The R\textsuperscript{2} value between the number of
-OTUs and chimeric sequences did not change with the use of `pre.cluster'
-or with the removal of chimeras using VSEARCH {[}Figure 6{]}. Taken
-together, this data suggests that a strong correlation exists between
-the number of OTUs and the prevalence of chimeric sequences.
+\textbf{\emph{Prevalence of Chimeric Sequences are HiFi DNA Polymerase
+Dependent and Correlate with the Number of OTUs:}} There were
+significant differences in the chimera prevalence based on HiFi DNA
+polymerase used at all levels of sub-sampling and cycle numbers (P-value
+\textless{} 0.05) {[}Table S9{]}. Differences in chimera prevalence
+between Platinum and all other HiFi DNA polymerases accounted for the
+majority of these differences {[}Table S10{]}.
+Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
+HiFi DNA polymerases regardless of whether `pre.cluster' or chimera
+removal with VSEARCH was used {[}Figure 5{]}. A positive correlation was
+observed between chimeric sequences and the number of OTUs for all HiFi
+DNA polymerases {[}Figure 6{]}. This positive correlation was strongest
+for Accuprime\textsuperscript{TM}, Platinum, and Phusion HiFi DNA
+Polymerases {[}Figure 6{]}. The R\textsuperscript{2} value between the
+number of OTUs and chimeric sequences did not change with the use of
+`pre.cluster' or with the removal of chimeras using VSEARCH {[}Figure
+6{]}. This data suggests that chimera prevalence depends on HiFi DNA
+polymerase used and confirms that the number of OTUs is dependent on the
+prevalence of these chimeric sequences.
 

### 2018-03-19 submission/manuscript.Rmd
discussion and conclusion updates

@@
 
-These observations build upon previous studies [@taq_Gohl_2016] by showing that different HiFi DNA polymerases have significant differences in the number of OTUs and that changes in total OTUs correlate with chimeras not removed after sequence processing [Figure 1-2 & 5].
-HiFi DNA polymerase dependent differences in total number of OTUs are important to consider since many diversity metrics rely on this measure as part of their calculations.
-Our observations show that HiFi DNA polymerase can have a noticeable affect on the OTUs generated and these differences are consistent across sub-sampling depth and PCR cycle number [Figure 2-4].
-Importantly, high biomass samples may underestimate the biases present within low biomass samples.
-An indication that this may be the case is the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples.
+In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5].
+These differences are important because many diversity metrics rely on this measure as part of their calculations (e.g.
+richness).
+Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
+Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches.
+This is important because it suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
 
-Although we did not observe strong differences in the Bray-Curtis index the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30 cycles, that are commonly used.
-There was few differences within individuals between corresponding 5-cycle increments (e.g.
-15 to 20, 20 to 25, etc.).
-However, there is a clear trend that suggests that the community at 20 cycles is different then the community at 25 cycles [Figure 3].
-These findings, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached.
-Further, for some HiFI DNA polymerases, increasing the sub-sampling depth may reduce some of these observed community differences at lower cycle numbers.
+There were few differences in bacterial community composition based on HiFi DNA polymerase.
+Additionally, the data suggests that there are no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
+One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group.
+Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
 
-Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5].
-The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations.
-However, Bray-Curtis analysis with PERMANOVA showed few differences based on HiFi DNA polymerase.
-It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases.
-
-Our observations suggest that there are clear HiFi DNA polymerase dependent differences in both per base error rate and chimera prevalence that cannot be removed using bioinformatic approaches [Figure 4 & 5].
-Although it may be a natural assumption that the variation may be due to the DNA polymerase family, the highest chimera rate, from Platinum, was a family A polymerase while the lowest, from Accuprime, was also an A family polymerase [@polymerase_Ishino_2014].
-In fact, from the material safety data sheet (MSDS), the differences between the two is not immediately apparent.
-Both Accuprime and Platinum contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-It is possible that differences in how the recombinant *Taq* was generated could be the main reason for the differences in chimera rate.
+Our observations suggest that there are clear HiFi DNA polymerase dependent differences in error rate and chimera prevalence that cannot be fully removed using bioinformatic approaches [Figure 4 & 5].
+Although this variation may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
+Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent.
+Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences in chimera rate between the different HiFi DNA polymerases.
 
@@
 
-Our findings show that the number of OTUs will be dependent on both HiFi DNA polymerase and cycle number chosen.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
+The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because intrinsic differences in HiFi DNA polymerases can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
 

### 2018-03-19 submission/manuscript.tex
discussion and conclusion updates

@@
 
-These observations build upon previous studies (Gohl et al., 2016) by
-showing that different HiFi DNA polymerases have significant differences
-in the number of OTUs and that changes in total OTUs correlate with
-chimeras not removed after sequence processing {[}Figure 1-2 \& 5{]}.
-HiFi DNA polymerase dependent differences in total number of OTUs are
-important to consider since many diversity metrics rely on this measure
-as part of their calculations. Our observations show that HiFi DNA
-polymerase can have a noticeable affect on the OTUs generated and these
-differences are consistent across sub-sampling depth and PCR cycle
-number {[}Figure 2-4{]}. Importantly, high biomass samples may
-underestimate the biases present within low biomass samples. An
-indication that this may be the case is the earlier detection of
-differences in total number of OTUs between HiFi DNA polymerases in the
-mock versus fecal samples.
+In this study we show that the number of OTUs, error rate, and chimera
+prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
+These differences are important because many diversity metrics rely on
+this measure as part of their calculations (e.g.~richness).
+Additionally, the earlier detection of differences in total number of
+OTUs between HiFi DNA polymerases in the mock versus fecal samples might
+indicate that high biomass samples may underestimate the biases present
+within low biomass samples. Importantly, many of these differences are
+due to undetected chimeras that cannot be fully removed using standard
+bioinformatic approaches. This is important because it suggests that
+some of the diversity differences between studies can be attributed to
+differences in HiFi DNA polyermase used.
 
-Although we did not observe strong differences in the Bray-Curtis index
-the data suggests that there may be differences between 15 and 20 cycles
-versus higher cycle numbers, such as 30 cycles, that are commonly used.
-There was few differences within individuals between corresponding
-5-cycle increments (e.g.~15 to 20, 20 to 25, etc.). However, there is a
-clear trend that suggests that the community at 20 cycles is different
-then the community at 25 cycles {[}Figure 3{]}. These findings, in
-conjunction with the PERMANOVA results, suggest that cycle number can
-change bacterial community calculations but that these differences are
-minimal once 25 cycles are reached. Further, for some HiFI DNA
-polymerases, increasing the sub-sampling depth may reduce some of these
-observed community differences at lower cycle numbers.
-
-Increasing the cycle number also exacerbated chimera prevalence
-differences between the different HiFi DNA polymerases {[}Figure 5{]}.
-The chimera prevalence was strongly correlated with the number of OTUs
-which is relied upon heavily for many community metric calculations.
-However, Bray-Curtis analysis with PERMANOVA showed few differences
-based on HiFi DNA polymerase. It is possible that many of the increased
-number of OTUs are not highly abundant allowing the Bray-Curtis index to
-be able to successfully down-weight these respective OTUs (Minchin,
-1987). The choice of downstream diversity metric could be an important
-consideration in helping to mitigate these observed changes due to high
-chimera prevalence in specific HiFi DNA polymerases.
+There were few differences in bacterial community composition based on
+HiFi DNA polymerase. Additionally, the data suggests that there are no
+differences in the overall bacterial community composition for
+sub-sampling depth or cycle number used. One possible reason for this
+outcome is that our study did not have enough power to detect
+differences due to low sample number in each group. Another reason is
+that many of the OTUs are likely not highly abundant, allowing the
+Bray-Curtis index to be able to successfully down-weight chimeric OTUs
+(Minchin, 1987). The choice of downstream diversity metric could be an
+important consideration in helping to mitigate these observed HiFi DNA
+polymerase dependent differences in chimera prevalence.
 
 Our observations suggest that there are clear HiFi DNA polymerase
-dependent differences in both per base error rate and chimera prevalence
-that cannot be removed using bioinformatic approaches {[}Figure 4 \&
-5{]}. Although it may be a natural assumption that the variation may be
-due to the DNA polymerase family, the highest chimera rate, from
-Platinum, was a family A polymerase while the lowest, from Accuprime,
-was also an A family polymerase (Ishino \& Ishino, 2014). In fact, from
-the material safety data sheet (MSDS), the differences between the two
-is not immediately apparent. Both Accuprime and Platinum contain a
-recombinant \emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D
-polymerase and a platinum \emph{Taq} antibody. It is possible that
-differences in how the recombinant \emph{Taq} was generated could be the
-main reason for the differences in chimera rate.
+dependent differences in error rate and chimera prevalence that cannot
+be fully removed using bioinformatic approaches {[}Figure 4 \& 5{]}.
+Although this variation may be due to the DNA polymerase family, the
+highest and lowest chimera rates both belonged to a family A polymerase
+(Platinum and Accuprime\textsuperscript{TM} respectively) (Ishino \&
+Ishino, 2014). Additionally, based on the material safety data sheet
+(MSDS) the differences between the two HiFi DNA polymerases are not
+immediately apparent. Both HiFi DNA polymerases contain a recombinant
+\emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a
+platinum \emph{Taq} antibody. With everything else being equal, it is
+possible that differences in how the recombinant \emph{Taq} was
+generated could be a contributing factor for the differences in chimera
+rate between the different HiFi DNA polymerases.
 
@@
 
-Our findings show that the number of OTUs will be dependent on both HiFi
-DNA polymerase and cycle number chosen. Care should be taken when
-choosing a HiFi DNA polymerase for 16S rRNA gene surveys since intrinsic
-differences can change the number of OTUs observed and influence
+The number of OTUs are dependent on both HiFi DNA polymerase and cycle
+number chosen. Care should be taken when choosing a HiFi DNA polymerase
+for 16S rRNA gene surveys because intrinsic differences in HiFi DNA
+polymerases can change the number of OTUs observed and influence
 diversity based metrics that do not down-weight rare observations.

### 2018-03-19 submission/manuscript.Rmd
spell check

@@
 Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
-Similarily, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
+Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
 Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015].
@@
 Finally, we did not observe a HiFi DNA polymerase dependent difference on base substitution rate [Figure S2].
-Although sequence error is dependent on HiFi DNA polymerase some of these error dependent differences can be corrected using exsiting bioinformatic approaches.
+Although sequence error is dependent on HiFi DNA polymerase some of these error dependent differences can be corrected using existing bioinformatic approaches.
 

### 2018-03-19 submission/manuscript.tex
spell check

@@
 been shown to cause overgrowth of low abundance members of the fecal
-bacterial community (Amir et al., 2017). Similarily, this overgrowth can
+bacterial community (Amir et al., 2017). Similarly, this overgrowth can
 still occur if the preservation media does not adequately inhibit growth
@@
 polymerase some of these error dependent differences can be corrected
-using exsiting bioinformatic approaches.
+using existing bioinformatic approaches.
 

### 2018-03-19 submission/manuscript.Rmd
remove extras in discussion

@@
 Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches.
-This is important because it suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
+This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
 
@@
 
-Our observations suggest that there are clear HiFi DNA polymerase dependent differences in error rate and chimera prevalence that cannot be fully removed using bioinformatic approaches [Figure 4 & 5].
-Although this variation may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
+Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
 Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent.
 Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences in chimera rate between the different HiFi DNA polymerases.
+With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases.
 

### 2018-03-19 submission/manuscript.tex
remove extras in discussion

@@
 due to undetected chimeras that cannot be fully removed using standard
-bioinformatic approaches. This is important because it suggests that
-some of the diversity differences between studies can be attributed to
-differences in HiFi DNA polyermase used.
+bioinformatic approaches. This suggests that some of the diversity
+differences between studies can be attributed to differences in HiFi DNA
+polyermase used.
 
@@
 
-Our observations suggest that there are clear HiFi DNA polymerase
-dependent differences in error rate and chimera prevalence that cannot
-be fully removed using bioinformatic approaches {[}Figure 4 \& 5{]}.
-Although this variation may be due to the DNA polymerase family, the
-highest and lowest chimera rates both belonged to a family A polymerase
-(Platinum and Accuprime\textsuperscript{TM} respectively) (Ishino \&
-Ishino, 2014). Additionally, based on the material safety data sheet
-(MSDS) the differences between the two HiFi DNA polymerases are not
-immediately apparent. Both HiFi DNA polymerases contain a recombinant
-\emph{Taq} DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a
-platinum \emph{Taq} antibody. With everything else being equal, it is
-possible that differences in how the recombinant \emph{Taq} was
-generated could be a contributing factor for the differences in chimera
-rate between the different HiFi DNA polymerases.
+Although the variation in error rate and chimera prevalence may be due
+to the DNA polymerase family, the highest and lowest chimera rates both
+belonged to a family A polymerase (Platinum and
+Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
+Additionally, based on the material safety data sheet (MSDS) the
+differences between the two HiFi DNA polymerases are not immediately
+apparent. Both HiFi DNA polymerases contain a recombinant \emph{Taq} DNA
+polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
+\emph{Taq} antibody. With everything else being equal, it is possible
+that differences in how the recombinant \emph{Taq} was generated could
+be a contributing factor for the differences observed between the HiFi
+DNA polymerases.
 

### 2018-03-19 submission/manuscript.Rmd
add further discussion points

@@
 
+- Add section on how this relates to the other components involved with 16S rRNA gene sequencing
+
+- Add section on how this relates to previous PCR bias findings
+
+- Bring everything back to the first paragraph that talks about bias, reproducibility and standardization
+
+- End on a positive that these differences are probably masked by large biological variation and so have minimal impact on findings within a study (can be part of the previous bullet point).
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}

### 2018-03-20 submission/manuscript.Rmd
add two more paragraphs to discussion

--- 2018-03-19 submission/manuscript.Rmd
@@
 
+Similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community.
+The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
+However, the chimarea prevalence for some HiFi DNA polymerases (e.g.
+Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study.
+Within the larger context of the different 16S rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used.
+
+Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach.
+However, bias can be easily reproducible and can be found in every step of the 16S rRNA gene sequencing workflow.
+This study shows that the observed microbial community consistently varies based on HiFi DNA polymerase.
+Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen () but also can artifically increase the number of OTUs observed.
+Arguably, standardizing workflow needs to be approached on a study by study basis and not every project needs to use the exact same approach.
+All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled.
+Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}

### 2018-03-20 submission/manuscript.tex
add two more paragraphs to discussion

@@
 
+Similar to using different perservation methods or different DNA
+extraction kits, the type of HiFi DNA polymerase can add bias to the
+observed bacterial community. The sequence error introduced by the HiFi
+DNA polymerase is small and likely to be smaller than the biological
+variation within a specific study, which would be consistent with
+previous findings for preservation and DNA extraction methods (Salter et
+al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
+prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
+large and might be greater than the oberseved biological variation
+within a specific study. Within the larger context of the different 16S
+rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
+important a consideration as either preservation or DNA extraction
+method used.
+
+Heavy standardization has been commonly suggested as a reasonable answer
+to finding the most reproducible approach. However, bias can be easily
+reproducible and can be found in every step of the 16S rRNA gene
+sequencing workflow. This study shows that the observed microbial
+community consistently varies based on HiFi DNA polymerase.
+Standardizing multiple workflows to one specific HiFi DNA polymerase
+could be detrimental since the PCR step not only misses entire species
+based on primer chosen () but also can artifically increase the number
+of OTUs observed. Arguably, standardizing workflow needs to be
+approached on a study by study basis and not every project needs to use
+the exact same approach. All aspects of the 16S rRNA gene sequencing
+workflow need to be customized for the specific microbial community that
+will be sampled. Although a diversity of approaches may make
+reproducibility more difficult it will help to avoid systematic biases
+from occuring due to widespread standardization of approaches.
+
 \newpage

### 2018-03-20 submission/manuscript.Rmd
discussion rewrite and reordering

@@
 In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5].
-These differences are important because many diversity metrics rely on this measure as part of their calculations (e.g.
+These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
 richness).
-Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
 Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches.
 This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
-
-There were few differences in bacterial community composition based on HiFi DNA polymerase.
-Additionally, the data suggests that there are no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
-One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group.
-Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
+Interestingly, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
 
@@
 With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases.
+
+There were few differences in bacterial community composition based on HiFi DNA polymerase.
+The data also suggests that there were no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
+One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group.
+Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
+Metrics that simply  look at presence/absence of OTUs (e.g.
+Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases.
+When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
 
@@
 Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach.
-However, bias can be easily reproducible and can be found in every step of the 16S rRNA gene sequencing workflow.
-This study shows that the observed microbial community consistently varies based on HiFi DNA polymerase.
-Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen () but also can artifically increase the number of OTUs observed.
-Arguably, standardizing workflow needs to be approached on a study by study basis and not every project needs to use the exact same approach.
+However, bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow.
+This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase.
+Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen [@Meisel2016] but also can artifically increase the number of OTUs observed.
+Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach.
 All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled.
@@
 The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because intrinsic differences in HiFi DNA polymerases can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
+Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
 

### 2018-03-20 submission/manuscript.tex
discussion rewrite and reordering

@@
 These differences are important because many diversity metrics rely on
-this measure as part of their calculations (e.g.~richness).
-Additionally, the earlier detection of differences in total number of
-OTUs between HiFi DNA polymerases in the mock versus fecal samples might
-indicate that high biomass samples may underestimate the biases present
-within low biomass samples. Importantly, many of these differences are
-due to undetected chimeras that cannot be fully removed using standard
-bioinformatic approaches. This suggests that some of the diversity
-differences between studies can be attributed to differences in HiFi DNA
-polyermase used.
-
-There were few differences in bacterial community composition based on
-HiFi DNA polymerase. Additionally, the data suggests that there are no
-differences in the overall bacterial community composition for
-sub-sampling depth or cycle number used. One possible reason for this
-outcome is that our study did not have enough power to detect
-differences due to low sample number in each group. Another reason is
-that many of the OTUs are likely not highly abundant, allowing the
-Bray-Curtis index to be able to successfully down-weight chimeric OTUs
-(Minchin, 1987). The choice of downstream diversity metric could be an
-important consideration in helping to mitigate these observed HiFi DNA
-polymerase dependent differences in chimera prevalence.
+the number of OTUs or other measures dependent on error rate and chimera
+prevalence as part of their metric calculations (e.g.~richness).
+Importantly, many of these differences are due to undetected chimeras
+that cannot be fully removed using standard bioinformatic approaches.
+This suggests that some of the diversity differences between studies can
+be attributed to differences in HiFi DNA polyermase used. Interestingly,
+the earlier detection of differences in total number of OTUs between
+HiFi DNA polymerases in the mock versus fecal samples might indicate
+that high biomass samples may underestimate the biases present within
+low biomass samples.
 
@@
 
-Similar to using different perservation methods or different DNA
+There were few differences in bacterial community composition based on
+HiFi DNA polymerase. The data also suggests that there were no
+differences in the overall bacterial community composition for
+sub-sampling depth or cycle number used. One possible reason for this
+outcome was that our study did not have enough power to detect
+differences due to low sample number in each group. Another reason was
+that many of the OTUs are likely not highly abundant, allowing the
+Bray-Curtis index to be able to successfully down-weight chimeric OTUs
+(Minchin, 1987). The choice of downstream diversity metric could be an
+important consideration in helping to mitigate these observed HiFi DNA
+polymerase dependent differences in chimera prevalence. Metrics that
+simply look at presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas,
+1996)) may be less robust to chimera prevalence and by extension total
+number of OTU differences in HiFi DNA Polymerases. When choosing a
+distance metric careful consideration of the biases introduced from the
+PCR step of the 16S rRNA gene sequencing workflow need to be taken into
+account.
+
+Similar to using different preservation methods or different DNA
 extraction kits, the type of HiFi DNA polymerase can add bias to the
@@
 to finding the most reproducible approach. However, bias can be easily
-reproducible and can be found in every step of the 16S rRNA gene
-sequencing workflow. This study shows that the observed microbial
-community consistently varies based on HiFi DNA polymerase.
-Standardizing multiple workflows to one specific HiFi DNA polymerase
-could be detrimental since the PCR step not only misses entire species
-based on primer chosen () but also can artifically increase the number
-of OTUs observed. Arguably, standardizing workflow needs to be
-approached on a study by study basis and not every project needs to use
-the exact same approach. All aspects of the 16S rRNA gene sequencing
+reproduced and can be found in every step of the 16S rRNA gene
+sequencing workflow. This study shows that specific diversity metrics
+used to measure the microbial community consistently vary based on HiFi
+DNA polymerase. Standardizing multiple workflows to one specific HiFi
+DNA polymerase could be detrimental since the PCR step not only misses
+entire species based on primer chosen (Meisel et al., 2016) but also can
+artifically increase the number of OTUs observed. Arguably, the degree
+of workflow standardization across studies and research group needs to
+be approached on a study by study basis and not every project needs to
+use the exact same approach. All aspects of the 16S rRNA gene sequencing
 workflow need to be customized for the specific microbial community that
@@
 number chosen. Care should be taken when choosing a HiFi DNA polymerase
-for 16S rRNA gene surveys because intrinsic differences in HiFi DNA
-polymerases can change the number of OTUs observed and influence
-diversity based metrics that do not down-weight rare observations.
+for 16S rRNA gene surveys because their intrinsic differences can change
+the number of OTUs observed and influence diversity based metrics that
+do not down-weight rare observations.
 
@@
 
+\hypertarget{ref-Real1996}{}
+Real R., Vargas JM. 1996. The probabilistic basis of jaccards index of
+similarity. \emph{Systematic Biology} 45:380--385. DOI:
+\href{https://doi.org/10.1093/sysbio/45.3.380}{10.1093/sysbio/45.3.380}.
+
 \hypertarget{ref-vsearch_Rognes_2016}{}

### 2018-03-20 submission/manuscript.Rmd
update acknowledgements
add pat funding

@@
 We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
-Salary support for Marc Sze came from the Canadian Institute of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.
+Salary support for Marc A.
+Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240.
+Salary support for Patrick D.
+Schloss came from NIH grant P30DK034933.
 

### 2018-03-20 submission/manuscript.tex
update acknowledgements
add pat funding

@@
 Microbiome Core Facility at the University of Michigan. Salary support
-for Marc Sze came from the Canadian Institute of Health Research and the
-Michigan Institute for Clinical and Health Research Postdoctoral
-Translational Scholar Program.
+for Marc A. Sze came from the Canadian Institute of Health Research and
+NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from
+NIH grant P30DK034933.
 

### 2018-03-23 submission/manuscript.Rmd
update acknowledgements
fix first and last intro paragraphs according to Elyse

@@
 
-This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
+Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results.
 First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
@@
 *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-In this study, we examine if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+
+Although PCR has been investigated as a source of bias in 16S rRNA gene sequencing, the contribution of specific DNA polymerases has not been.
+This study will specifically address how high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
+We will accomplish this by examining if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 
@@
 Salary support for Patrick D.
-Schloss came from NIH grant P30DK034933.
+Schloss came from NIH grants P30DK034933 and 1R01CA215574.
 

### 2018-03-23 submission/manuscript.tex
update acknowledgements
fix first and last intro paragraphs according to Elyse

@@
 
-This study will specifically address how cycle number and high fidelity
-(HiFi) DNA polymerases can bias observed bacterial community results
-derived from 16S rRNA gene sequencing. First, it is important to
-differentiate between bias, reproducibility, and standardization since
-often times these three can be confused and used interchangeably with
-each other. Bias can change the observed results in a way that is
-reproducible and standardized. For example, if one group uses one brand
-of DNA extraction kit for their 16S rRNA gene sequencing, their results
-may be biased versus another group not using the same brand kit but
-within their group they can still have reproducible results. Therefore,
-standardization of 16S rRNA gene sequencing methods to increase
-reproducibility can still be problematic due to bias. Determining how
-different aspects of a 16S rRNA gene sequencing workflow could bias the
-observed results is critical for the interpretation of specific studies
-in the broader context of the overall field.
+Accounting for bias is critical in order to reach an understanding of
+bacterial community changes using 16S rRNA gene sequencing results.
+First, it is important to differentiate between bias, reproducibility,
+and standardization since often times these three can be confused and
+used interchangeably with each other. Bias can change the observed
+results in a way that is reproducible and standardized. For example, if
+one group uses one brand of DNA extraction kit for their 16S rRNA gene
+sequencing, their results may be biased versus another group not using
+the same brand kit but within their group they can still have
+reproducible results. Therefore, standardization of 16S rRNA gene
+sequencing methods to increase reproducibility can still be problematic
+due to bias. Determining how different aspects of a 16S rRNA gene
+sequencing workflow could bias the observed results is critical for the
+interpretation of specific studies in the broader context of the overall
+field.
 
@@
 intrinsically have different error rates that cannot be completely
-removed with modifications (Ishino \& Ishino, 2014). In this study, we
-examine if any of five different types of HiFi DNA polymerases introduce
-significant biases into 16S rRNA gene surveys, if this is a cycle
-dependent phenomenon, and whether they can be removed using a standard
+removed with modifications (Ishino \& Ishino, 2014).
+
+Although PCR has been investigated as a source of bias in 16S rRNA gene
+sequencing, the contribution of specific DNA polymerases has not been.
+This study will specifically address how high fidelity (HiFi) DNA
+polymerases can bias observed bacterial community results derived from
+16S rRNA gene sequencing. We will accomplish this by examining if any of
+five different types of HiFi DNA polymerases introduce significant
+biases into 16S rRNA gene surveys, if this is a cycle dependent
+phenomenon, and whether they can be removed using a standard
 bioinformatic pipeline.
@@
 NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from
-NIH grant P30DK034933.
+NIH grants P30DK034933 and 1R01CA215574.
 

### 2018-03-26 submission/manuscript.Rmd
convert abstract to active voice

@@
 
-**Methods.** DNA from fecal samples (n=4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach.
-The number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities.
-Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated.
-Finally, the chimera prevalence correlation with number of OTUs was assessed.
+**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
+Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach.
+The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities.
+It also assessed HiFi DNA polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence.
+Finally, our analysis also examined chimera prevalence correlation with the number of OTUs.
 
-**Results.** When analyzing fecal samples a different number of OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-These HiFi dependent differences in the number of OTUs were identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-Chimera prevalence varied by HiFi DNA polymerase and this variation was still present after chimera removal using VSEARCH.
-The chimera prevalence was positively correlated with the number of OTUs and was also not affected by chimera removal with VSEARCH.
+**Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+Our analysis identified these HiFi dependent differences in the number of OTUs as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+Chimera prevalence varied by HiFi DNA polymerase and this variation persisted after chimera removal using VSEARCH.
+We also observed positive correlations between chimera prevalence and the number of OTUs which was not affected by chimera removal with VSEARCH.
 

### 2018-03-26 submission/manuscript.tex
convert abstract to active voice

@@
 
-\textbf{Methods.} DNA from fecal samples (n=4) were extracted using a
+\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
 PowerMag DNA extraction kit with a 10 minute bead beating step and
 amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
-Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
-(technical replicates n=4) consisting of previously isolated whole
-genomes of 8 different bacteria were also amplified using the same
-approach. The number of OTUs (Operational Taxonomic Units) were examined
-for both fecal samples and mock communities. Next, Bray-Curtis index,
-the error rate, sequence error prevalence, and chimera prevalence were
-investigated. Finally, the chimera prevalence correlation with number of
-OTUs was assessed.
+Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
+communities (technical replicates n=4) consisting of previously isolated
+whole genomes of 8 different bacteria used the same approach. The
+analysis initially examined the number of Operational Taxonomic Units
+(OTUs) for both fecal samples and mock communities. It also assessed
+HiFi DNA polymerase dependent differences in the Bray-Curtis index,
+error rate, sequence error prevalence, and chimera prevalence. Finally,
+our analysis also examined chimera prevalence correlation with the
+number of OTUs.
 
-\textbf{Results.} When analyzing fecal samples a different number of
-OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value
-\textless{} 0.0001). These HiFi dependent differences in the number of
-OTUs were identified as early as 20 cycles in the mock communities
-(P-value = 0.002). Chimera prevalence varied by HiFi DNA polymerase and
-this variation was still present after chimera removal using VSEARCH.
-The chimera prevalence was positively correlated with the number of OTUs
-and was also not affected by chimera removal with VSEARCH.
+\textbf{Results.} When analyzing fecal samples we observed a different
+number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
+\textless{} 0.0001). Our analysis identified these HiFi dependent
+differences in the number of OTUs as early as 20 cycles in the mock
+communities (P-value = 0.002). Chimera prevalence varied by HiFi DNA
+polymerase and this variation persisted after chimera removal using
+VSEARCH. We also observed positive correlations between chimera
+prevalence and the number of OTUs which was not affected by chimera
+removal with VSEARCH.
 

### 2018-03-27 submission/manuscript.Rmd
reoword first intro paragraph

@@
 Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results.
-First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other.
+Differentiating between bias, reproducibility, and standardization is important since often times these three can be confused and used interchangeably with each other.
 Bias can change the observed results in a way that is reproducible and standardized.
-For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results.
-Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias.
-Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the broader context of the overall field.
+For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results [@contamination_Salter2014].
+Our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
+In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.
 

### 2018-03-27 submission/manuscript.Rmd
add better transitions to intro

@@
 
+Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively.
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
+Generally, not using a preservation media and leaving samples at room temperature supports overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
 Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
 Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015].
-Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014].
-Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017].
+Additionally, reagent contamination can add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014].
+Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017].
 However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
@@
 Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011].
+Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.
 
-There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias.
+The intrinsic properties to primers and DNA polymerases chosen can also introduce bias.
 Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g.
@@
 
-Although PCR has been investigated as a source of bias in 16S rRNA gene sequencing, the contribution of specific DNA polymerases has not been.
-This study will specifically address how high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
+Although bias introduced due to differences between DNA polymerase and HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific HiFi DNA polymerases has not been.
+This study will specifically address how HiFi DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
 We will accomplish this by examining if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-27 submission/manuscript.tex
add better transitions to intro

@@
 bacterial community changes using 16S rRNA gene sequencing results.
-First, it is important to differentiate between bias, reproducibility,
-and standardization since often times these three can be confused and
-used interchangeably with each other. Bias can change the observed
-results in a way that is reproducible and standardized. For example, if
-one group uses one brand of DNA extraction kit for their 16S rRNA gene
-sequencing, their results may be biased versus another group not using
-the same brand kit but within their group they can still have
-reproducible results. Therefore, standardization of 16S rRNA gene
-sequencing methods to increase reproducibility can still be problematic
-due to bias. Determining how different aspects of a 16S rRNA gene
-sequencing workflow could bias the observed results is critical for the
-interpretation of specific studies in the broader context of the overall
-field.
+Differentiating between bias, reproducibility, and standardization is
+important since often times these three can be confused and used
+interchangeably with each other. Bias can change the observed results in
+a way that is reproducible and standardized. For example, if one group
+uses one brand of DNA extraction kit for their 16S rRNA gene sequencing,
+their results may be biased versus another group not using the same
+brand kit but within their group they can still have reproducible
+results (Salter et al., 2014). Our 16S rRNA gene sequencing methods are
+biased even when these workflows are standardized to increase
+reproducibility. In order to interpret specific studies within the
+broader context of the overall field, assessing bias at different parts
+of the 16S rRNA gene sequencing workflow is critical.
 
-A typical 16S rRNA gene sequencing workflow can be divided into
-preservation, extraction, PCR, and sequencing steps. Generally, not
-using a preservation media and leaving samples at room temperature has
-been shown to cause overgrowth of low abundance members of the fecal
-bacterial community (Amir et al., 2017). Similarly, this overgrowth can
-still occur if the preservation media does not adequately inhibit growth
-(Song et al., 2016; Luo et al., 2016). Reports have also shown that
-changes in specific community members might occur due to differing
-susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
-2015). Additionally, reagent contamination has been shown to add
+Many parts of the 16S rRNA gene sequencing workflow contribute bias to
+the results and are studied extensively. A typical 16S rRNA gene
+sequencing workflow can be divided into preservation, extraction, PCR,
+and sequencing steps. Generally, not using a preservation media and
+leaving samples at room temperature supports overgrowth of low abundance
+members of the fecal bacterial community (Amir et al., 2017). Similarly,
+this overgrowth can still occur if the preservation media does not
+adequately inhibit growth (Song et al., 2016; Luo et al., 2016). Reports
+have also shown that changes in specific community members might occur
+due to differing susceptibility to freeze thaw cycles amongst microbes
+(Gorzelak et al., 2015). Additionally, reagent contamination can add
 community members and the contribution of these contaminant members
-grows larger with lower biomass samples (Salter et al., 2014). Recent
-studies have shown that the majority of these biases due to either
-preservation or extraction tend to be smaller than the overall
-biological signal being measured (Song et al., 2016; Bassis et al.,
-2017). However, the contribution of PCR bias to this overall workflow is
-not well characterized since these studies use the same PCR approach
-while varying preservation or extraction method.
+grows larger with lower biomass samples (Salter et al., 2014). Overall,
+biases due to either preservation or extraction tend to be smaller than
+the overall biological signal being measured (Song et al., 2016; Bassis
+et al., 2017). However, the contribution of PCR bias to this overall
+workflow is not well characterized since these studies use the same PCR
+approach while varying preservation or extraction method.
 
@@
 by a subsequent priming error and secondary extension and will also
-artificially increase community diversity (Haas et al., 2011).
+artificially increase community diversity (Haas et al., 2011). Although
+these differences are not necessarily dependent on primer and DNA
+polymerase used, there are also biases that are.
 
-There are also intrinsic properties to primers and DNA polymerases
-chosen that can introduce bias. Primers have variable region dependent
-binding affinities for different bacteria and depending on the primer
-pair do not detect specific bacteria (e.g.~V1-V3 does not detect
+The intrinsic properties to primers and DNA polymerases chosen can also
+introduce bias. Primers have variable region dependent binding
+affinities for different bacteria and depending on the primer pair do
+not detect specific bacteria (e.g.~V1-V3 does not detect
 \emph{Haemophilus influenzae} and V3-V5 does not detect
@@
 
-Although PCR has been investigated as a source of bias in 16S rRNA gene
-sequencing, the contribution of specific DNA polymerases has not been.
-This study will specifically address how high fidelity (HiFi) DNA
+Although bias introduced due to differences between DNA polymerase and
+HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing,
+the bias caused due to differences between specific HiFi DNA polymerases
+has not been. This study will specifically address how HiFi DNA
 polymerases can bias observed bacterial community results derived from

### 2018-03-27 submission/manuscript.Rmd
final intro fixes and first discussion fix
update discussion transitions except last paragraph

@@
 
-In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5].
+In this study we show that the number of OTUs, error rate, and chimera prevalence depends on HiFi DNA polymerase [Figure 1-2 & 4-5].
 These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
 richness).
-Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches.
-This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
-Interestingly, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
+Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
+We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences.
+This suggests that specific diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
+Based on our observations metrics that measure within sample diversity depend on HiFi DNA polymerase but this may not be the case for metrics that assess between sample diversity.
+
+There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index.
+Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
+One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group.
+Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
+Metrics that solely use presence/absence of OTUs (e.g.
+Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases.
+When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
+With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.
 
@@
 With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases.
+We are unlikely to avoid adding HiFi dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.
 
-There were few differences in bacterial community composition based on HiFi DNA polymerase.
-The data also suggests that there were no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
-One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group.
-Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
-Metrics that simply  look at presence/absence of OTUs (e.g.
-Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases.
-When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
-
-Similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community.
+The majority of HiFi DNA polymerases we studied add a small increase in the number of OTUs and chimera prevalence that can be masked by biological differences.
 The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
@@
 Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study.
-Within the larger context of the different 16S rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used.
+The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community.
+Although avoiding HiFi DNA polymerases that yield high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
 

### 2018-03-27 submission/manuscript.Rmd
update discussion transitions

@@
 
-The majority of HiFi DNA polymerases we studied add a small increase in the number of OTUs and chimera prevalence that can be masked by biological differences.
+The majority of HiFi DNA polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences.
 The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
@@
 The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community.
-Although avoiding HiFi DNA polymerases that yield high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
+Although avoiding HiFi DNA polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
 
-Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach.
-However, bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow.
+Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations.
+Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016].
 This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase.
-Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen [@Meisel2016] but also can artifically increase the number of OTUs observed.
+Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases may work better in certain situations over others.
 Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach.

### 2018-03-27 submission/manuscript.tex
update discussion transitions

@@
 In this study we show that the number of OTUs, error rate, and chimera
-prevalence are HiFi DNA polymerase dependent {[}Figure 1-2 \& 4-5{]}.
-These differences are important because many diversity metrics rely on
-the number of OTUs or other measures dependent on error rate and chimera
+prevalence depends on HiFi DNA polymerase {[}Figure 1-2 \& 4-5{]}. These
+differences are important because many diversity metrics rely on the
+number of OTUs or other measures dependent on error rate and chimera
 prevalence as part of their metric calculations (e.g.~richness).
-Importantly, many of these differences are due to undetected chimeras
-that cannot be fully removed using standard bioinformatic approaches.
-This suggests that some of the diversity differences between studies can
-be attributed to differences in HiFi DNA polyermase used. Interestingly,
-the earlier detection of differences in total number of OTUs between
-HiFi DNA polymerases in the mock versus fecal samples might indicate
-that high biomass samples may underestimate the biases present within
-low biomass samples.
+Additionally, the earlier detection of differences in total number of
+OTUs between HiFi DNA polymerases in the mock versus fecal samples might
+indicate that high biomass samples may underestimate the biases present
+within low biomass samples. We observed that undetected chimeras that
+were not identified and removed using standard bioinformatic approaches
+cause many of these differences. This suggests that specific diversity
+differences between studies can be attributed to differences in HiFi DNA
+polyermase used. Based on our observations metrics that measure within
+sample diversity depend on HiFi DNA polymerase but this may not be the
+case for metrics that assess between sample diversity.
+
+There were few differences that depend on HiFi DNA polymerase for
+between sample diversity, as measured by the Bray-Curtis index. Using
+this metric our observations found no differences in the overall
+bacterial community composition for sub-sampling depth or cycle number
+used. One possible reason for this outcome was that our study did not
+have enough power to detect differences due to low sample number in each
+group. Another reason was that many of the OTUs are likely not highly
+abundant, allowing the Bray-Curtis index to be able to successfully
+down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
+diversity metric could be an important consideration in helping to
+mitigate these observed HiFi DNA polymerase dependent differences in
+chimera prevalence. Metrics that solely use presence/absence of OTUs
+(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
+prevalence and by extension total number of OTU differences in HiFi DNA
+Polymerases. When choosing a distance metric careful consideration of
+the biases introduced from the PCR step of the 16S rRNA gene sequencing
+workflow need to be taken into account. With differences in the number
+of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
+might be easier to avoid specific DNA polymerase families altogether.
 
@@
 be a contributing factor for the differences observed between the HiFi
-DNA polymerases.
+DNA polymerases. We are unlikely to avoid adding HiFi dependent bias to
+16S rRNA gene sequencing results, however, these differences may not be
+large enough to mask the actual biological signal.
 
-There were few differences in bacterial community composition based on
-HiFi DNA polymerase. The data also suggests that there were no
-differences in the overall bacterial community composition for
-sub-sampling depth or cycle number used. One possible reason for this
-outcome was that our study did not have enough power to detect
-differences due to low sample number in each group. Another reason was
-that many of the OTUs are likely not highly abundant, allowing the
-Bray-Curtis index to be able to successfully down-weight chimeric OTUs
-(Minchin, 1987). The choice of downstream diversity metric could be an
-important consideration in helping to mitigate these observed HiFi DNA
-polymerase dependent differences in chimera prevalence. Metrics that
-simply look at presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas,
-1996)) may be less robust to chimera prevalence and by extension total
-number of OTU differences in HiFi DNA Polymerases. When choosing a
-distance metric careful consideration of the biases introduced from the
-PCR step of the 16S rRNA gene sequencing workflow need to be taken into
-account.
-
-Similar to using different preservation methods or different DNA
-extraction kits, the type of HiFi DNA polymerase can add bias to the
-observed bacterial community. The sequence error introduced by the HiFi
-DNA polymerase is small and likely to be smaller than the biological
+The majority of HiFi DNA polymerases we studied add small increases in
+the number of OTUs and chimera prevalence that can be masked by
+biological differences. The sequence error introduced by the HiFi DNA
+polymerase is small and likely to be smaller than the biological
 variation within a specific study, which would be consistent with
@@
 large and might be greater than the oberseved biological variation
-within a specific study. Within the larger context of the different 16S
-rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
+within a specific study. The choice of HiFi DNA polymerase can be as
 important a consideration as either preservation or DNA extraction
-method used.
+method used because similar to using different preservation methods or
+different DNA extraction kits, the type of HiFi DNA polymerase can add
+bias to the observed bacterial community. Although avoiding HiFi DNA
+polymerases that yield a high number of OTUs and chimera prevalence
+might provide better standardization of results it does come at a cost.
 
-Heavy standardization has been commonly suggested as a reasonable answer
-to finding the most reproducible approach. However, bias can be easily
+Methods can be standardized but they commonly contain bias that is
+reproducible and may miss important associations. Bias can be easily
 reproduced and can be found in every step of the 16S rRNA gene
-sequencing workflow. This study shows that specific diversity metrics
-used to measure the microbial community consistently vary based on HiFi
-DNA polymerase. Standardizing multiple workflows to one specific HiFi
-DNA polymerase could be detrimental since the PCR step not only misses
-entire species based on primer chosen (Meisel et al., 2016) but also can
-artifically increase the number of OTUs observed. Arguably, the degree
+sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
+2016; Amir et al., 2017). This study shows that specific diversity
+metrics used to measure the microbial community consistently vary based
+on HiFi DNA polymerase. Standardizing multiple workflows to one specific
+HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
+may work better in certain situations over others. Arguably, the degree
 of workflow standardization across studies and research group needs to

### 2018-03-27 submission/manuscript.Rmd
improve conclusion

@@
 
-The number of OTUs are dependent on both HiFi DNA polymerase and cycle number chosen.
+Our observations fill a gap in knowledge on the bias introduced to 16S rRNA gene sequencing results due to differences in HiFi DNA polymerases.
+We found that the number of OTUs and the chimera prevalence is dependent on both HiFi DNA polymerase and cycle number chosen.
 Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
+Knowing the inherent bias associated with different HiFi DNA polymerases allows for better interpretation of the relationsip between an individual study to their respective field of research.
 

### 2018-03-27 submission/manuscript.tex
improve conclusion

@@
 
-The number of OTUs are dependent on both HiFi DNA polymerase and cycle
-number chosen. Care should be taken when choosing a HiFi DNA polymerase
-for 16S rRNA gene surveys because their intrinsic differences can change
-the number of OTUs observed and influence diversity based metrics that
-do not down-weight rare observations.
+Our observations fill a gap in knowledge on the bias introduced to 16S
+rRNA gene sequencing results due to differences in HiFi DNA polymerases.
+We found that the number of OTUs and the chimera prevalence is dependent
+on both HiFi DNA polymerase and cycle number chosen. Care should be
+taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys
+because their intrinsic differences can change the number of OTUs
+observed and influence diversity based metrics that do not down-weight
+rare observations. Knowing the inherent bias associated with different
+HiFi DNA polymerases allows for better interpretation of the relationsip
+between an individual study to their respective field of research.
 

### 2018-03-29 submission/manuscript.Rmd
add f-stat to ANOVA result reporting

@@
 Additionally, there was a trend for lower cycle numbers (15-20) to result in less differences in the number of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
-For fecal samples, all sub-sampling levels had significant differences between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
+For fecal samples, all sub-sampling levels had significant differences between HiFi DNA polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x")[, "F.value"]), digits = 2)`, P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
 Most of the differences observed at 35 cycles were between Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
-Differences in the number of OTUs between HiFi DNA polymerases were identifiable at earlier cycles (25 and 30) but the sub-sampling depth had to be 5000 sequences or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
+Differences in the number of OTUs between HiFi DNA polymerases were identifiable at earlier cycles (25 and 30) but the sub-sampling depth had to be 5000 sequences or higher (F-stat > `r round(min(filter(fs_anova_summary, cycle == "25x" | cycle == "30x", sub_sample_level != 1000)[, "F.value"]), digits = 2)`, P-value `r ifelse(max(filter(fs_anova_summary, cycle == "25x" | cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
 
@@
 Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across cycle number and sub-sampling depth [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles and at as low of a sub-sampling depth as 1000 sequences in the mock community samples (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles and at as low of a sub-sampling depth as 1000 sequences in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
 For both cycle numbers and sub-sampling depths, the majority of  differences in the number of OTUs were between Platinum and the other HiFi DNA polymerases [Table S4].

### 2018-03-29 submission/manuscript.tex
add f-stat to ANOVA result reporting

@@
 had significant differences between HiFi DNA polymerases at 35 cycles
-(P-value \textless{} 0.0001) {[}Table S1{]}. Most of the differences
-observed at 35 cycles were between Platinum and other HiFi DNA
-polymerases, based on a Tukey post-hoc test (P-value \textless{} 0.05)
-{[}Table S2{]}. Differences in the number of OTUs between HiFi DNA
-polymerases were identifiable at earlier cycles (25 and 30) but the
-sub-sampling depth had to be 5000 sequences or higher (P-value
-\textless{} 0.05) {[}Table S1{]}.
+(F-stat \textgreater{} 16.35, P-value \textless{} 0.0001) {[}Table
+S1{]}. Most of the differences observed at 35 cycles were between
+Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test
+(P-value \textless{} 0.05) {[}Table S2{]}. Differences in the number of
+OTUs between HiFi DNA polymerases were identifiable at earlier cycles
+(25 and 30) but the sub-sampling depth had to be 5000 sequences or
+higher (F-stat \textgreater{} 4.98, P-value \textless{} 0.05) {[}Table
+S1{]}.
 
@@
 cycles and at as low of a sub-sampling depth as 1000 sequences in the
-mock community samples (P-value = 0.002) {[}Table S3{]}. For both cycle
-numbers and sub-sampling depths, the majority of differences in the
-number of OTUs were between Platinum and the other HiFi DNA polymerases
-{[}Table S4{]}. Based on these observations in fecal and mock
-communities, it is clear that different HiFi DNA polymerases result in a
-different total number of OTUs observed within a sample.
+mock community samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}.
+For both cycle numbers and sub-sampling depths, the majority of
+differences in the number of OTUs were between Platinum and the other
+HiFi DNA polymerases {[}Table S4{]}. Based on these observations in
+fecal and mock communities, it is clear that different HiFi DNA
+polymerases result in a different total number of OTUs observed within a
+sample.
 

### 2018-03-29 submission/manuscript.Rmd
final re-work intro & discussion read through -- hopefully Pat likes it

@@
 
-**Background.** Using different reagents and kits at different steps of the 16S rRNA gene sequencing workflow can introduce bias by changing the observed microbial community.
+**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow.
+Commonly studied sources of bias include preservation and DNA extraction methods.
 Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
@@
 It also assessed HiFi DNA polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence.
-Finally, our analysis also examined chimera prevalence correlation with the number of OTUs.
+Our analysis also examined chimera prevalence correlation with the number of OTUs.
 
@@
 incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
-Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011].
+Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011].
 Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.
@@
 Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases.
-When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
+When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
 With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.
@@
 All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled.
-Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.
+Although a diversity of approaches may make reproducibility and replicability more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.
 

### 2018-03-29 submission/manuscript.tex
final re-work intro & discussion read through -- hopefully Pat likes it

@@
 
-\textbf{Background.} Using different reagents and kits at different
-steps of the 16S rRNA gene sequencing workflow can introduce bias by
-changing the observed microbial community. Although cycle number and
-high fidelity (HiFi) DNA polymerase are studied less often, they are
-still important sources of bias in this workflow. Here, we examine how
-both cycle number and HiFi DNA polymerase can change the bacterial
-community and introduce bias to the final obtained results.
+\textbf{Background.} Bias is introduced at different stages of the 16S
+rRNA gene sequencing workflow. Commonly studied sources of bias include
+preservation and DNA extraction methods. Although cycle number and high
+fidelity (HiFi) DNA polymerase are studied less often, they are still
+important sources of bias in this workflow. Here, we examine how both
+cycle number and HiFi DNA polymerase can change the bacterial community
+and introduce bias to the final obtained results.
 
@@
 HiFi DNA polymerase dependent differences in the Bray-Curtis index,
-error rate, sequence error prevalence, and chimera prevalence. Finally,
-our analysis also examined chimera prevalence correlation with the
-number of OTUs.
+error rate, sequence error prevalence, and chimera prevalence. Our
+analysis also examined chimera prevalence correlation with the number of
+OTUs.
 
@@
 Additionally, chimeras can form from an aborted extension step followed
-by a subsequent priming error and secondary extension and will also
+by a subsequent priming error and secondary extension and will
 artificially increase community diversity (Haas et al., 2011). Although
@@
 prevalence and by extension total number of OTU differences in HiFi DNA
-Polymerases. When choosing a distance metric careful consideration of
+Polymerases. When choosing a distance metric, careful consideration of
 the biases introduced from the PCR step of the 16S rRNA gene sequencing
@@
 will be sampled. Although a diversity of approaches may make
-reproducibility more difficult it will help to avoid systematic biases
-from occuring due to widespread standardization of approaches.
+reproducibility and replicability more difficult it will help to avoid
+systematic biases from occuring due to widespread standardization of
+approaches.
 

### 2018-05-11 code/make_mock_error_graphs.R
update to remove previous panels A and B

@@ -117,9 +117,9 @@ full_pipeline <- error_data[["mock_error"]] %>%
   theme_bw() + 
   scale_color_manual(name = "Taq Used", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = "Median Average Error Rate Per Base") + 
-  ggtitle("C") + coord_cartesian(ylim = c(0, 0.015)) + 
-  annotate("text", label = paste("Full Pipeline"), x = 2.5, y = 0.015, size = 3.5) + 
+  labs(x = "Amplification Cycles", y = "Median Error Rate") + 
+  coord_cartesian(ylim = c(0, 0.015)) + 
+  annotate("text", label = paste("Mock Data"), x = 2.5, y = 0.015, size = 3.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -132,6 +132,6 @@ full_pipeline <- error_data[["mock_error"]] %>%
 
  
  
- combined_graph <- grid.arrange(before_precluster, before_chimera_vsearch, full_pipeline, ncol = 3)
+#combined_graph <- grid.arrange(before_precluster, before_chimera_vsearch, full_pipeline, ncol = 3)
 
-ggsave("results/figures/Figure4.pdf", combined_graph, width = 11, height = 7, dpi = 300)
+ggsave("results/figures/Figure4.pdf", full_pipeline, width = 4, height = 6, dpi = 300)

### 2018-05-11 code/make_mock_chimera_count_graphs.R
update graph with pat suggestions

@@ -39,36 +39,36 @@ error_data <- sapply(error_files, function(x) read_csv(paste("data/process/table
 
 
 # Generate graph of Mock DNA samples 
-before_precluster <- error_data[["mock_precluster_error"]]  %>% 
-  filter(total_seqs >= 1000) %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
-  group_by(taq, cycles) %>% 
-  summarise(group_chimera = median(chimera_prevalence), group_iqr25 = quantile(chimera_prevalence)["25%"], 
-            group_iqr75 = quantile(chimera_prevalence)["75%"]) %>% 
-  ggplot(aes(cycles, group_chimera, color = taq, group = taq)) + 
-  geom_line(show.legend = F) + 
-  geom_errorbar(aes(ymin=group_iqr25, 
-                    ymax=group_iqr75), show.legend = F,  
-                width = 0.1, size = 0.5, alpha = 0.4) + 
-  geom_point(size = 2, alpha = 0.7, show.legend = F) + 
-  theme_bw() + 
-  scale_color_manual(name = "Taq Used", 
-                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = "Chimeric Sequences") + 
-  coord_cartesian(ylim = c(0, 0.30)) + 
-  scale_y_continuous(labels = scales::percent) + 
-  ggtitle("A") + 
-  annotate("text", label = paste("Before Pre-Cluster Step"), x = 2.5, y = 0.3, size = 3.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10), 
-        legend.position = c(0.25, 0.83), 
-        legend.title = element_blank(), 
-        legend.key = element_blank(), 
-        legend.background = element_rect(color = "black"))
+# before_precluster <- error_data[["mock_precluster_error"]]  %>% 
+#   filter(total_seqs >= 1000) %>% 
+#   mutate(taq = factor(taq, 
[…]

### 2018-05-11 code/make_correlation_graphs.R
remove non-full pipeline correlations

@@ -108,57 +108,57 @@ combined_list <- sapply(error_files,
   select(-taq.x, -taq.y)
   
 
-
-eqns_m_pc_error <- filter(combined_list, pipeline == "mock_precluster_error") %>% by(.$taq, lm_eqn)
-df_m_pc_error<- data.frame(eq = unclass(eqns_m_pc_error), taq = rownames(eqns_m_pc_error))
-
-
-mock_pc <- combined_list %>% filter(pipeline == "mock_precluster_error") %>% 
-  ggplot(aes(chimera_prevalence*100, total_otus, color = cycles.x, group = taq)) + 
-  geom_smooth(size = 1, method = "lm", se = FALSE, color = "black", show.legend = F) + 
-  geom_point(size = 2, alpha = 0.7, show.legend = F) + theme_bw() + 
-  facet_grid(taq ~.) + 
-  scale_color_manual(name = "Cycle Number", 
-                     values = c("#0000FF", "#00C957", "#CD8500", "#FF1493")) + 
-  labs(x = "Chimera Sequence Prevalence (%)", y = "Number of OTUs") + 
-  geom_text(data = df_m_pc_error, aes(x = 2.5, y = 380, label = eq), 
-            color = 'black', size = 3.5, parse = TRUE) + 
-  ggtitle("A") + coord_cartesian(ylim = c(0, 500)) + scale_x_continuous(limits = c(0, 30)) + 
-  annotate("text", label = paste("Before Precluster Step"), x = 3, y = 450, size = 2.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10), 
-        legend.position = "bottom", 
-        legend.title = element_blank(), 
-        legend.key = element_blank(), 
-        legend.background = element_rect(color = "black"))
-
-
-
-eqns_m_ch_error <- filter(combined_list, pipeline == "mock_chimera_error") %>% by(.$taq, lm_eqn)
-df_m_ch_error <- data.frame(eq = unclass(eqns_m_ch_error), taq = rownames(eqns_m_ch_error))
-
-mock_ch <- combined_list %>% filter(pipeline == "mock_chimera_error") %>% 
-  ggplot(aes(chimera_prevalence*100, total_otus, color = cycles.x, group = taq)) + 
-  geom_smooth(size = 1, method = "lm", se = FALSE, color = "black", show.legend = F) + 
[…]

### 2018-05-11 submission/manuscript.Rmd
remove old fig s1 from text

@@
 
-**Figure S1: HiFi DNA Polymerase Sequence Error Prevalence in Mock Samples.** A) Sequence error prevalence before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
-B) Sequence error prevalence before the removal of chimeras with VSEARCH.
-C) Full pipeline.
-The error bars represent the 75% interquartile range of the median.
-
-**Figure S2: HiFi DNA Polymerase Nucleotide Subsitutions in Mock Samples.**
+**Figure S1: HiFi DNA Polymerase Nucleotide Subsitutions in Mock Samples.**

### 2018-05-11 submission/manuscript.tex
remove old fig s1 from text

@@
 \textbf{\emph{Statistical Analysis:}} All analysis was done with the R
-(v 3.4.3) software package (R Core Team, 2017). Data transformation and
+(v 3.4.4) software package (R Core Team, 2017). Data transformation and
 graphing was completed using the tidyverse package (v 1.2.1) and colors
@@
 
-\textbf{Figure S1: HiFi DNA Polymerase Sequence Error Prevalence in Mock
-Samples.} A) Sequence error prevalence before the merger of sequences
-with pre.cluster and the removal of chimeras with VSEARCH. B) Sequence
-error prevalence before the removal of chimeras with VSEARCH. C) Full
-pipeline. The error bars represent the 75\% interquartile range of the
-median.
-
-\textbf{Figure S2: HiFi DNA Polymerase Nucleotide Subsitutions in Mock
+\textbf{Figure S1: HiFi DNA Polymerase Nucleotide Subsitutions in Mock
 Samples.}

### 2018-05-11 code/make_bray_distance_graphs.R
update to median

@@ -51,7 +51,7 @@ dist_data <- sapply(sub_sample_level,
                                 labels = c("15x vs 20x", "20x vs 25x", 
                                            "25x vs 30x", "30x vs 35x"))) %>% 
   group_by(taq, depth_level, cycle_compare) %>% 
-  summarise(mean_dist = mean(distance, na.rm = T), 
+  summarise(mean_dist = median(distance, na.rm = T), 
             max_dist = max(distance, na.rm = T), 
             min_dist = min(distance, na.rm = T))
   
@@ -69,7 +69,7 @@ mock_dist_data <- sapply(sub_sample_level,
                                 labels = c("15x vs 20x", "20x vs 25x", 
                                            "25x vs 30x", "30x vs 35x"))) %>% 
   group_by(taq, depth_level, cycle_compare) %>% 
-  summarise(mean_dist = mean(distance, na.rm = T), 
+  summarise(mean_dist = median(distance, na.rm = T), 
             max_dist = max(distance, na.rm = T), 
             min_dist = min(distance, na.rm = T))
 

### 2018-05-11 code/make_fecal_numOTU_graphs.R

@@ -49,7 +49,7 @@ numOTU_data <- sapply(sub_sample_level,
   
 
 summary_data <- numOTU_data %>% group_by(depth_level, taq, cycle_num) %>% 
-  summarise(taq_mean = mean(numOTUs, na.rm = T), 
+  summarise(taq_mean = median(numOTUs, na.rm = T), 
             taq_min = min(numOTUs, na.rm = T), 
             taq_max = max(numOTUs, na.rm = T))
 

### 2018-05-11 code/make_nucleotide_graphs.R
update to make message clear

@@ -53,21 +53,29 @@ nucleotide_data <- read_csv("data/process/tables/full_nucleotide_error_summary.c
   mutate(prev_subtype = str_replace(prev_subtype, "_seq_prev", ""))
 
 
-full_graph <- nucleotide_data %>% 
+summary_data <- nucleotide_data %>% 
+  group_by(sub_type, taq, cycles) %>% 
+  summarise(medain_sub_rate = median(rate, na.rm = T), 
+            max_sub_rate = max(rate, na.rm = T), 
+            min_sub_rate = min(rate, na.rm = T)) %>% 
+  ungroup()
+
+
+full_graph <- summary_data %>% 
      mutate(sub_type = factor(sub_type, 
                           levels = c("ac", "ag", "at", "ca", "cg", "ct", "ga", "gc", "gt", "ta", "tc", "tg"), 
                           labels = c("A => C", "A => G", "A => T", "C => A", "C => G", "C => T", "G => A", "G => C", 
                                      "G => T", "T => A", "T => C", "T => G"))) %>% 
-     ggplot(aes(sub_type, log10(rate + 0.000001), color = taq, group = taq)) + 
-     geom_point(size = 2, alpha = 0.2, position = position_dodge(width = 0.7), show.legend = F, color = "gray") + 
-     theme_bw() + 
-     stat_summary(aes(color = taq), fun.y = median, fun.ymin = median, fun.ymax = median, 
-                  position = position_dodge(width = 0.7), geom = "crossbar", size = 0.5, width = 0.7) + 
-     facet_grid(cycles~.) + 
-     scale_color_manual(name = "Taq Used", 
-                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
+  ggplot(aes(cycles, log10(medain_sub_rate + 0.00000001), color = sub_type, group = sub_type)) + 
+  geom_pointrange(aes(ymin = min_sub_rate, ymax = max_sub_rate), size = 0.4, 
+                  alpha = 0.9, show.legend = T, position = position_dodge(width = 0.8)) + 
+  facet_grid(taq~.) + theme_bw() + 
+  geom_vline(xintercept=seq(1.5, length(unique(summary_data$cycles))-0.5, 1), 
+             lwd=1, colour="gray") + 
+     # scale_color_manual(name = "Taq Used", 
+     #                  values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
[…]

### 2018-05-14 code/make_fecal_numOTU_graphs.R
remove extra rarefaction levels

@@ -53,28 +53,32 @@ summary_data <- numOTU_data %>% group_by(depth_level, taq, cycle_num) %>%
             taq_min = min(numOTUs, na.rm = T), 
             taq_max = max(numOTUs, na.rm = T))
 
-overall_taq_mean_data <- numOTU_data %>% group_by(taq) %>% 
-  summarise(overall_mean = mean(numOTUs, na.rm = T))
+overall_taq_mean_data <- numOTU_data %>% 
+  filter(depth_level == 1000) %>% 
+  group_by(taq) %>% 
+  summarise(overall_mean = median(numOTUs, na.rm = T))
 
 fecal_samples <- summary_data %>% 
-  ggplot(aes(factor(depth_level), taq_mean, color = cycle_num, group = cycle_num)) + 
-  geom_hline(aes(yintercept = overall_mean), overall_taq_mean_data, color = "red", size = 1, linetype = "dashed") + 
+  filter(depth_level == 1000) %>% 
+  ggplot(aes(cycle_num, taq_mean, color = taq, group = taq)) + 
+#  geom_hline(aes(yintercept = overall_mean, color = taq), overall_taq_mean_data, size = 1, 
+#             alpha = 0.5, linetype = "dashed", show.legend = F) + 
   geom_pointrange(aes(ymin = taq_min, ymax = taq_max), size = 0.4, alpha = 0.9, show.legend = T, position = position_dodge(width = 0.8)) + 
   geom_vline(xintercept=seq(1.5, length(unique(summary_data$depth_level))-0.5, 1), 
              lwd=1, colour="gray") + 
-  facet_grid(. ~ taq) + theme_bw() + 
-  scale_color_manual(name = "Cycle Number", 
+  theme_bw() + coord_cartesian(ylim = c(0, 150)) + 
+  scale_color_manual(name = "HiFi DNA Polymerase", 
                      values = c("#590059", "#006400", "#00FF7F", "#63B8FF", "#104E8B")) + 
-  labs(x = "Rarefied Depth", y = "Number of OTUs") + 
+  labs(x = "Number of Cycles", y = "Number of OTUs") + 
   theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10),
-        axis.text.x = element_text(size = 8, angle = 70, hjust = 1), 
+        axis.text.x = element_text(size = 10), 
[…]

### 2018-05-14 code/make_fecal_numOTU_graphs.R
update colors

@@ -68,7 +68,7 @@ fecal_samples <- summary_data %>%
              lwd=1, colour="gray") + 
   theme_bw() + coord_cartesian(ylim = c(0, 150)) + 
   scale_color_manual(name = "HiFi DNA Polymerase", 
-                     values = c("#590059", "#006400", "#00FF7F", "#63B8FF", "#104E8B")) + 
+                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Number of Cycles", y = "Number of OTUs") + 
   theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 

### 2018-05-14 code/make_fecal_numOTU_graphs.R
remove legend title

@@ -76,6 +76,7 @@ fecal_samples <- summary_data %>%
         axis.text.y = element_text(size = 10),
         axis.text.x = element_text(size = 10), 
         legend.key = element_blank(), 
+        legend.title = element_blank(), 
         legend.position = "bottom", 
         legend.background = element_rect(color = "black"))
   

### 2018-05-14 code/make_fecal_numOTU_graphs.R
add text annotation

@@ -70,6 +70,7 @@ fecal_samples <- summary_data %>%
   scale_color_manual(name = "HiFi DNA Polymerase", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   labs(x = "Number of Cycles", y = "Number of OTUs") + 
+  annotate("text", label = paste("Fecal Data"), x = 0.8, y = 150, size = 3.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 

### 2018-05-14 code/make_numOTU_graphs.R
update to remove rarefaction info

@@ -45,82 +45,42 @@ numOTU_data <- sapply(sub_sample_level,
   mutate(taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
+
+
+summary_data <- numOTU_data %>% 
+  group_by(depth_level, taq, cycles) %>% 
+  summarise(taq_mean = median(numOTUs, na.rm = T), 
+            taq_min = min(numOTUs, na.rm = T), 
+            taq_max = max(numOTUs, na.rm = T))
   
 
 ref_otu_num <- read_tsv("data/process/zymo_mock_only.shared") %>% 
   select(numOtus)
 
 # Generate graph of Mock DNA samples 
-thousand_graph <- numOTU_data %>% filter(depth_level == 1000) %>% 
-  ggplot(aes(cycles, log2(numOTUs), color = taq, group = taq)) + 
-  geom_smooth(size = 1, method = "lm", formula = y ~ poly(x, 2), se = FALSE, show.legend = F) + 
-  geom_point(size = 2, alpha = 0.7, show.legend = F) + 
+thousand_graph <- summary_data %>% filter(depth_level == 1000) %>% 
+  ggplot(aes(cycles, taq_mean, color = taq, group = taq)) + 
+  geom_pointrange(aes(ymin = taq_min, ymax = taq_max), size = 0.4, alpha = 0.9, show.legend = T, position = position_dodge(width = 0.8)) +
+  geom_vline(xintercept=seq(1.5, length(unique(summary_data$depth_level))-0.5, 1), 
+             lwd=1, colour="gray") + 
   geom_hline(yintercept = log2(ref_otu_num$numOtus), linetype = "dashed", size = 1) + 
   theme_bw() + 
-  scale_color_manual(name = "Taq Used", 
+  scale_color_manual(name = "HiFi DNA Polymerase", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = expression(Log["2"]~Number~of~OTUs)) + 
-  ggtitle("A") + coord_cartesian(ylim = c(2, 8)) + 
-  annotate("text", label = paste("Sub-sampled to 1000 Sequences"), x = 2.5, y = 8, size = 3.5) + 
[…]

### 2018-05-14 code/make_bray_distance_graphs.R
remove rarefaction component

@@ -54,7 +54,8 @@ dist_data <- sapply(sub_sample_level,
   summarise(mean_dist = median(distance, na.rm = T), 
             max_dist = max(distance, na.rm = T), 
             min_dist = min(distance, na.rm = T))
-  
+
+
 
 mock_dist_data <- sapply(sub_sample_level, 
                     function(x) read_csv(paste("data/process/tables/mock_bray_5_cycle_dist_", 
@@ -79,21 +80,22 @@ ann_text <- data.frame(cycle_compare = 2.2, distance = 0.95, lab = "Sub-Sampled"
 
 
 
-fecal_graph <- dist_data %>% filter(!is.na(cycle_compare)) %>% 
-  ggplot(aes(factor(depth_level), mean_dist, color = cycle_compare, group = cycle_compare)) + 
-  geom_pointrange(aes(ymin = min_dist, ymax = max_dist), size = 0.4, alpha = 0.9, show.legend = T, position = position_dodge(width = 0.8)) + 
-  geom_vline(xintercept=seq(1.5, length(unique(dist_data$depth_level))-0.5, 1), 
+fecal_graph <- dist_data %>% 
+  filter(!is.na(cycle_compare), depth_level == 1000) %>% 
+  ggplot(aes(cycle_compare, mean_dist, color = taq, group = taq)) + 
+  geom_pointrange(aes(ymin = min_dist, ymax = max_dist), size = 0.4, show.legend = T, position = position_dodge(width = 0.8)) + 
+  geom_vline(xintercept=seq(1.5, length(unique(dist_data$taq))-0.5, 1), 
              lwd=1, colour="gray") + 
-  facet_grid(~taq) + 
-  scale_color_manual(name = "Cycle Comparison", 
-                     values = c("#FF00FF", "#0000FF", "#8B1A1A", "#EE9A00")) + 
+  scale_color_manual(name = "HiFi DNA Polymerase", 
+                     values = c("#440154FF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
   theme_bw() +  coord_cartesian(ylim = c(0, 1)) + ggtitle("A") + 
-  labs(x = "Sub-sampling Depth", y = "Bray Curtis Index") + 
+  labs(x = "Cycle Comparison", y = "Bray Curtis Index") + 
+  annotate("text", label = paste("Fecal Data"), x = 4.3, y = 1.0, size = 3.5) + 
[…]

### 2018-05-14 submission/manuscript.Rmd
update figure legends

@@
 
-**Figure 1: Normalized Fecal Number of OTUs.** The x-axis represents the different sub-sampling depths used and the y-axis is the normalized within individual number of OTUs.
-The red line represents the overall mean Z-score normalized number of OTUs for each respective HiFi DNA polymerase.
-The dashed black line represents the overall Z-score normalized mean number of OTUs.
+**Figure 1: Total Number of OTUs in Fecal Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
+The legend represents the color code for the 5 different HiFi DNA polymerases used.
+The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
+This range was also larger for specific HiFi DNA polymerases.
 
-**Figure 2:  Mock Sample Variability in Number of OTUs based on HiFi DNA Polymerase.** A) Sub-sampled to 1000 reads.
-B) Sub-sampled to 5000 reads.
-C) Sub-sampled to 10000 reads.
-The dotted line represents the number of OTUs generated when the mock reference sequences are run through the pipeline.
+**Figure 2:  Total Number of OTUs in Mock Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
+The legend represents the color code for the 5 different HiFi DNA polymerases used.
+The dotted black line represents the number of OTUs detected when only the references sequences are clustered.
+The range in the number of OTUs detected in the Mock samples increased as cycle number increased.
+This range was also larger for specific HiFi DNA polymerases.
 
-**Figure 3: Community Differences by Five-Cycle Intervals and Sub-sampling Depth.** A) Fecal samples within person difference based on the next 5-cycle PCR interval.
-B) Mock samples within replicate difference based on the next 5-cycle PCR interval.
+**Figure 3: Bray-Curtis Community Differences by Five-Cycle Intervals.** A) within person differences based on the next 5-cycle PCR interval in fecal samples.
+B) Within replicate difference based on the next 5-cycle PCR interval in Mock samples.
 
-**Figure 4: HiFi DNA Polymerase Per Base Error Rate in Mock Samples.** A) Error rate before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
-B) Error rate before the removal of chimeras with VSEARCH.
-C) Full pipeline.
+**Figure 4: HiFi DNA Polymerase Error Rate in Mock Samples.** The error bars represent the 75% interquartile range of the median error rate.
+
+**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Precentage of chimeric sequences without the removal of chimeras with VSEARCH.
+C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.
+C) The total percent of chimeric sequences removed with VSEARCH by cycle number.
 The error bars represent the 75% interquartile range of the median.
 
-**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Chimera sequence percentage before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
-B) Chimera sequence percentage before the removal of chimeras with VSEARCH.
-C) Full pipeline.
-The error bars represent the 75% interquartile range of the median.
-
-**Figure 6: The Correlation between Number of OTUs and Chimeras.** A) Correlation before the merger of sequences with pre.cluster and the removal of chimeras with VSEARCH.
-B) Correlation before the removal of chimeras with VSEARCH.
-C) Correlation with full pipeline.
+**Figure 6: The Correlation between Number of OTUs and Chimeras in Mock Samples.**
 

### 2018-05-14 submission/manuscript.tex
update figure legends

@@
 
-\textbf{Figure 1: Normalized Fecal Number of OTUs.} The x-axis
-represents the different sub-sampling depths used and the y-axis is the
-normalized within individual number of OTUs. The red line represents the
-overall mean Z-score normalized number of OTUs for each respective HiFi
-DNA polymerase. The dashed black line represents the overall Z-score
-normalized mean number of OTUs.
+\textbf{Figure 1: Total Number of OTUs in Fecal Samples by Number of
+Cycles.} The x-axis represents the different number of cycles used and
+the y-axis is the number of OTUs. The legend represents the color code
+for the 5 different HiFi DNA polymerases used. The range in the number
+of OTUs detected in the different fecal samples increased as cycle
+number increased. This range was also larger for specific HiFi DNA
+polymerases.
 
-\textbf{Figure 2: Mock Sample Variability in Number of OTUs based on
-HiFi DNA Polymerase.} A) Sub-sampled to 1000 reads. B) Sub-sampled to
-5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents
-the number of OTUs generated when the mock reference sequences are run
-through the pipeline.
+\textbf{Figure 2: Total Number of OTUs in Mock Samples by Number of
+Cycles.} The x-axis represents the different number of cycles used and
+the y-axis is the number of OTUs. The legend represents the color code
+for the 5 different HiFi DNA polymerases used. The dotted black line
+represents the number of OTUs detected when only the references
+sequences are clustered. The range in the number of OTUs detected in the
+Mock samples increased as cycle number increased. This range was also
+larger for specific HiFi DNA polymerases.
 
-\textbf{Figure 3: Community Differences by Five-Cycle Intervals and
-Sub-sampling Depth.} A) Fecal samples within person difference based on
-the next 5-cycle PCR interval. B) Mock samples within replicate
-difference based on the next 5-cycle PCR interval.
+\textbf{Figure 3: Bray-Curtis Community Differences by Five-Cycle
+Intervals.} A) within person differences based on the next 5-cycle PCR
+interval in fecal samples. B) Within replicate difference based on the
+next 5-cycle PCR interval in Mock samples.
 
-\textbf{Figure 4: HiFi DNA Polymerase Per Base Error Rate in Mock
-Samples.} A) Error rate before the merger of sequences with pre.cluster
-and the removal of chimeras with VSEARCH. B) Error rate before the
-removal of chimeras with VSEARCH. C) Full pipeline. The error bars
-represent the 75\% interquartile range of the median.
+\textbf{Figure 4: HiFi DNA Polymerase Error Rate in Mock Samples.} The
+error bars represent the 75\% interquartile range of the median error
+rate.
 
 \textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
-Samples.} A) Chimera sequence percentage before the merger of sequences
-with pre.cluster and the removal of chimeras with VSEARCH. B) Chimera
-sequence percentage before the removal of chimeras with VSEARCH. C) Full
-pipeline. The error bars represent the 75\% interquartile range of the
-median.
+Samples.} A) Precentage of chimeric sequences without the removal of
+chimeras with VSEARCH. C) Percentage of chimeric sequences with the
+removal of chimeras with VSEARCH. C) The total percent of chimeric
+sequences removed with VSEARCH by cycle number. The error bars represent
+the 75\% interquartile range of the median.
 
-\textbf{Figure 6: The Correlation between Number of OTUs and Chimeras.}
-A) Correlation before the merger of sequences with pre.cluster and the
-removal of chimeras with VSEARCH. B) Correlation before the removal of
-chimeras with VSEARCH. C) Correlation with full pipeline.
+\textbf{Figure 6: The Correlation between Number of OTUs and Chimeras in
+Mock Samples.}
 

### 2018-05-14 code/make_mock_chimera_count_graphs.R
update to number of cycles on x-axis

@@ -39,38 +39,6 @@ error_data <- sapply(error_files, function(x) read_csv(paste("data/process/table
 
 
 # Generate graph of Mock DNA samples 
-# before_precluster <- error_data[["mock_precluster_error"]]  %>% 
-#   filter(total_seqs >= 1000) %>% 
-#   mutate(taq = factor(taq, 
-#                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-#                       labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
-#   group_by(taq, cycles) %>% 
-#   summarise(group_chimera = median(chimera_prevalence), group_iqr25 = quantile(chimera_prevalence)["25%"], 
-#             group_iqr75 = quantile(chimera_prevalence)["75%"]) %>% 
-#   ggplot(aes(cycles, group_chimera, color = taq, group = taq)) + 
-#   geom_line(show.legend = F) + 
-#   geom_errorbar(aes(ymin=group_iqr25, 
-#                     ymax=group_iqr75), show.legend = F,  
-#                 width = 0.1, size = 0.5, alpha = 0.4) + 
-#   geom_point(size = 2, alpha = 0.7, show.legend = F) + 
-#   theme_bw() + 
-#   scale_color_manual(name = "Taq Used", 
-#                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-#   labs(x = "Amplification Cycles", y = "Chimeric Sequences") + 
-#   coord_cartesian(ylim = c(0, 0.30)) + 
-#   scale_y_continuous(labels = scales::percent) + 
-#   ggtitle("A") + 
-#   annotate("text", label = paste("Before Pre-Cluster Step"), x = 2.5, y = 0.3, size = 3.5) + 
-#   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
-#         panel.grid.major = element_blank(), 
-#         panel.grid.minor = element_blank(), 
-#         axis.text.y = element_text(size = 10), 
-#         legend.position = c(0.25, 0.83), 
-#         legend.title = element_blank(), 
-#         legend.key = element_blank(), 
-#         legend.background = element_rect(color = "black"))
-
-
 before_chimera_vsearch <- error_data[["mock_chimera_error"]]  %>% 
[…]

### 2018-05-14 code/make_mock_error_graphs.R

@@ -40,66 +40,6 @@ error_data <- sapply(error_files, function(x) read_csv(paste("data/process/table
 
 
 # Generate graph of Mock DNA samples 
-before_precluster <- error_data[["mock_precluster_error"]] %>% 
-  filter(total_seqs >= 1000) %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
-  group_by(taq, cycles) %>% 
-  summarise(group_median = median(mean_error), group_iqr25 = quantile(mean_error)["25%"], 
-            group_iqr75 = quantile(mean_error)["75%"]) %>% 
-  ggplot(aes(cycles, group_median, color = taq, group = taq)) + 
-  geom_line(show.legend = F) + 
-  geom_errorbar(aes(ymin=group_iqr25, 
-                    ymax=group_iqr75), show.legend = F,  
-                width = 0.1, size = 0.5, alpha = 0.4) + 
-  geom_point(size = 2, alpha = 0.7, show.legend = F) + 
-  theme_bw() + 
-  scale_color_manual(name = "Taq Used", 
-                     values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  labs(x = "Amplification Cycles", y = "Median Average Error Rate Per Base") + 
-  ggtitle("A") + coord_cartesian(ylim = c(0,0.015)) + 
-  annotate("text", label = paste("Before Pre-Cluster Step"), x = 2.5, y = 0.015, size = 3.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10), 
-        legend.position = c(0.20, 0.80), 
-        legend.title = element_blank(), 
-        legend.key = element_blank(), 
-        legend.background = element_rect(color = "black"))
-
-
-before_chimera_vsearch <- error_data[["mock_chimera_error"]] %>% 
-  filter(total_seqs >= 1000) %>% 
-  mutate(taq = factor(taq, 
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
[…]

### 2018-05-14 code/make_nucleotide_graphs.R

@@ -74,7 +74,7 @@ full_graph <- summary_data %>%
              lwd=1, colour="gray") + 
      # scale_color_manual(name = "Taq Used", 
      #                  values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-   labs(x = "", y = expression(Log["10"]~Substitution~Rate)) + 
+   labs(x = "Number of Cycles", y = expression(Log["10"]~Substitution~Rate)) + 
   coord_cartesian(ylim = c(0, -6)) + 
    theme(plot.title = element_text(face="bold", hjust = -0.07, size = 10), 
          panel.grid.major = element_blank(), 

### 2018-05-15 submission/manuscript.Rmd
update first result section and methods -- remove sub-sampling

@@
 Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples before the 'pre.cluster' command, before chimera removal, and after chimera removal [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+Sequence error was determined using the 'seq.error' command on mock samples [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
 
 ***Analysis Workflow:***
-The total number of OTUs was analyzed after sub-sampling for both the fecal and mock community samples.
+All samples were rarefied to 1000 sequence.
+The total number of OTUs was analyzed for both the fecal and mock community samples.
 For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g.
 20 versus 25, 25 versus 30, etc.).
-These community based measures for fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).
 Based on these observations we analyzed potential reasons for these differences.
-Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal.
-Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed before the 'pre.cluster' command, before chimera removal, and after chimera removal.
+Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before and after chimera removal.
+Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
 
@@
 ***The Number of OTUs are Dependent on HiFi DNA Polymerase:***
-A consistent difference in the number of OTUs, that was dependent on the HiFi DNA polymerase used was observed regardless of sub-sampling depth for fecal samples [Figure 1].
-Additionally, there was a trend for lower cycle numbers (15-20) to result in less differences in the number of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
-For fecal samples, all sub-sampling levels had significant differences between HiFi DNA polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x")[, "F.value"]), digits = 2)`, P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1].
-Most of the differences observed at 35 cycles were between Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].
-Differences in the number of OTUs between HiFi DNA polymerases were identifiable at earlier cycles (25 and 30) but the sub-sampling depth had to be 5000 sequences or higher (F-stat > `r round(min(filter(fs_anova_summary, cycle == "25x" | cycle == "30x", sub_sample_level != 1000)[, "F.value"]), digits = 2)`, P-value `r ifelse(max(filter(fs_anova_summary, cycle == "25x" | cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1].
-
-This HiFi DNA polymerase dependent difference in the number of OTUs was also observed in the mock community samples [Figure 2].
+A consistent difference in the range of number of OTUs, that was dependent on the HiFi DNA polymerase used was observed for fecal samples [Figure 1].
+Additionally, there was a trend for lower cycle numbers (15-20) to result in a reduced range in the number of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
+For fecal samples, there was a significant difference between HiFi DNA polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
+No differences between the specific groups was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
+This HiFi DNA polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2].
 Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across cycle number and sub-sampling depth [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles and at as low of a sub-sampling depth as 1000 sequences in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-For both cycle numbers and sub-sampling depths, the majority of  differences in the number of OTUs were between Platinum and the other HiFi DNA polymerases [Table S4].
-Based on these observations in fecal and mock communities, it is clear that different HiFi DNA polymerases result in a different total number of OTUs observed within a sample.
+In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+For the mock community, the majority of  differences in the number of OTUs were between Kappa and Platinum and the other HiFi DNA polymerases across different number of cycles [Table S4].
+Based on these observations in fecal and mock communities, it is clear that different HiFi DNA polymerases result in a different total number of OTUs within a sample.
 

### 2018-05-15 submission/manuscript.tex
update first result section and methods -- remove sub-sampling

@@
 2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
-was determined using the `seq.error' command on mock samples before the
-`pre.cluster' command, before chimera removal, and after chimera removal
-(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+was determined using the `seq.error' command on mock samples (Schloss et
+al., 2009; Cole et al., 2013; Rognes et al., 2016).
 
-\textbf{\emph{Analysis Workflow:}} The total number of OTUs was analyzed
-after sub-sampling for both the fecal and mock community samples. For
-fecal samples, cycle dependent affects on Bray-Curtis indices were
-assessed for cycle group and within individual differences from the
-previous cycle (e.g.~20 versus 25, 25 versus 30, etc.). These community
-based measures for fecal samples were analyzed at 4 different
-sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the
-mock community samples were analysed at 3 levels (1000, 5000, 10000).
-Based on these observations we analyzed potential reasons for these
-differences. Analysis of the mock community of each HiFi DNA polymerase
-for general sequence error rate, number of sequences with an error, base
-substitution, and numbers of chimeras were assessed before the
-`pre.cluster' command, before chimera removal, and after chimera
-removal. Additionally, the correlation between the number of chimeras
-and the number of OTUs was also assessed before the `pre.cluster'
-command, before chimera removal, and after chimera removal.
+\textbf{\emph{Analysis Workflow:}} All samples were rarefied to 1000
+sequence. The total number of OTUs was analyzed for both the fecal and
+mock community samples. For fecal samples, cycle dependent affects on
+Bray-Curtis indices were assessed for cycle group and within individual
+differences from the previous cycle (e.g.~20 versus 25, 25 versus 30,
+etc.). Based on these observations we analyzed potential reasons for
+these differences. Analysis of the mock community of each HiFi DNA
+polymerase for general sequence error rate, number of sequences with an
+error, base substitution, and numbers of chimeras were assessed before
+and after chimera removal. Additionally, the correlation between the
+number of chimeras and the number of OTUs was also assessed.
 
@@
 \textbf{\emph{The Number of OTUs are Dependent on HiFi DNA Polymerase:}}
-A consistent difference in the number of OTUs, that was dependent on the
-HiFi DNA polymerase used was observed regardless of sub-sampling depth
-for fecal samples {[}Figure 1{]}. Additionally, there was a trend for
-lower cycle numbers (15-20) to result in less differences in the number
-of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA
-polymerases {[}Figure 1{]}. For fecal samples, all sub-sampling levels
-had significant differences between HiFi DNA polymerases at 35 cycles
-(F-stat \textgreater{} 16.35, P-value \textless{} 0.0001) {[}Table
-S1{]}. Most of the differences observed at 35 cycles were between
-Platinum and other HiFi DNA polymerases, based on a Tukey post-hoc test
-(P-value \textless{} 0.05) {[}Table S2{]}. Differences in the number of
-OTUs between HiFi DNA polymerases were identifiable at earlier cycles
-(25 and 30) but the sub-sampling depth had to be 5000 sequences or
-higher (F-stat \textgreater{} 4.98, P-value \textless{} 0.05) {[}Table
-S1{]}.
-
-This HiFi DNA polymerase dependent difference in the number of OTUs was
-also observed in the mock community samples {[}Figure 2{]}. Regardless
-if fecal or mock communities were used, the same HiFi DNA polymerases
-had high (Platinum) and low (Accuprime) number of OTUs and this was
-consistent across cycle number and sub-sampling depth {[}Figure 1-2 \&
-Table S1-S4{]}. In contrast to the results obtained with fecal samples,
-differences between HiFi DNA polymerases were observed as early as 20
-cycles and at as low of a sub-sampling depth as 1000 sequences in the
-mock community samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}.
-For both cycle numbers and sub-sampling depths, the majority of
-differences in the number of OTUs were between Platinum and the other
-HiFi DNA polymerases {[}Table S4{]}. Based on these observations in
-fecal and mock communities, it is clear that different HiFi DNA
-polymerases result in a different total number of OTUs observed within a
-sample.
+A consistent difference in the range of number of OTUs, that was
+dependent on the HiFi DNA polymerase used was observed for fecal samples
+{[}Figure 1{]}. Additionally, there was a trend for lower cycle numbers
+(15-20) to result in a reduced range in the number of OTUs versus higher
+cycle numbers (25, 30, and 35) between HiFi DNA polymerases {[}Figure
+1{]}. For fecal samples, there was a significant difference between HiFi
+DNA polymerases at 35 cycles (F-stat \textgreater{} 16.35, P-value =
+9.7e-05) {[}Table S1{]}. No differences between the specific groups was
+found based on a Tukey post-hoc test (P-value \textgreater{} 0.05)
+{[}Table S2{]}. This HiFi DNA polymerase dependent difference in the
+range of number of OTUs also was observed in the mock community samples
+{[}Figure 2{]}. Regardless if fecal or mock communities were used, the
+same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number
+of OTUs and this was consistent across cycle number and sub-sampling
+depth {[}Figure 1-2 \& Table S1-S4{]}. In contrast to the results
+obtained with fecal samples, differences between HiFi DNA polymerases
+were observed as early as 20 cycles in the mock community samples
+(F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
+community, the majority of differences in the number of OTUs were
+between Kappa and Platinum and the other HiFi DNA polymerases across
+different number of cycles {[}Table S4{]}. Based on these observations
+in fecal and mock communities, it is clear that different HiFi DNA
+polymerases result in a different total number of OTUs within a sample.
 

### 2018-05-15 submission/supplemental.Rmd
update first result section and methods -- remove sub-sampling

@@
 fs_anova <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv")) %>% 
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level) %>% 
   mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
@@
 
-colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle")
 
@@
   mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level) %>% 
   rename(padj = `p adj`) %>% 
   mutate(padj = as.numeric(padj)) %>% 
-  filter(padj < 0.05) %>% 
-  mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+  mutate_at(vars(padj), funs(format(., digits = 2, nsmall = 2)))
 
-colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle")
 
@@
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  filter(sub_sample_level > 500)
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level)
 
-colnames(mock_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+colnames(mock_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle")
 
@@
   mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level) %>% 
   rename(padj = `p adj`) %>% 
   mutate(padj = as.numeric(padj)) %>% 
-  filter(padj < 0.05, sub_sample_level > 500) %>% 
+  filter(padj < 0.05) %>% 
   arrange(padj) %>% 
@@
 
-colnames(mock_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+colnames(mock_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle")
 

### 2018-05-15 submission/manuscript.Rmd
make changes to bray-curtis --remove sub-sampling

@@
 
-***Minimal Bray-Curtis Differences are Detected and are Dependent on both Cycle Number and Sub-Sampling Depth:***
+***Minimal Bray-Curtis Differences are Detected and are Dependent on Cycle Number:***
 A few small differences based on sub-sampling and cycle number were detected in overall bacterial community composition.
-Within the same fecal sample and independent of HiFi DNA polymerases, there were differences in the community composition between 20 and 25 cycles that was dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (IQR)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A].
-Further, when data was available for the mock communities, there were larger observed differences between 20 and 25 cycles (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B].
-Additionally, these stated community differences disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
-Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
-Using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depths (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
-For fecal samples, Phusion was one of two HiFi DNA polymerases that had enough sequences to reach a sub-sampling depth of 1000 at 15 cycles.
-Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on both sub-sampling depth and cycle number.
+Within the same fecal sample and independent of HiFi DNA polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A].
+Additionally, when data was available for the mock communities, there were also large observed differences between 20 and 25 cycles [Figure 3B].
+These community differences within a sample disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
+Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+Using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
+For fecal samples, Phusion was one of two HiFi DNA polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
+Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on cycle number.
 

### 2018-05-15 submission/manuscript.tex
make changes to bray-curtis --remove sub-sampling

@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
-Dependent on both Cycle Number and Sub-Sampling Depth:}} A few small
-differences based on sub-sampling and cycle number were detected in
-overall bacterial community composition. Within the same fecal sample
-and independent of HiFi DNA polymerases, there were differences in the
-community composition between 20 and 25 cycles that was dependent on
-sub-sampling depth (sub-sampled to 1000 = 0.51 (0.4 - 0.79) (median
-(IQR)), sub-sampled to 5000 = 0.43 (0.33 - 0.63), sub-sampled to 10000 =
-0.4 (0.24 - 0.43)) {[}Figure 3A{]}. Further, when data was available for
-the mock communities, there were larger observed differences between 20
-and 25 cycles (sub-sampled to 1000 = 0.88 (0.42 - 0.91)) {[}Figure
-3B{]}. Additionally, these stated community differences disappear when
-comparing 25 to 30 cycles and do not persist past 25 cycles {[}Figure
-3{]}. Although these trends are clearly noticeable, we found that there
-was no detectable difference in Bray-Curtis index when comparing to the
-previous 5-cycle increment for both fecal and mock communities after
-multiple comparison correction (P-value \textgreater{} 0.05). Using
-PERMANOVA to test for community differences based on cycle number within
-HiFi DNA polymerases, only Phusion had cycle dependent differences at
-1000 and 5000 sub-sampling depths (P-value = 0.03 and 0.01). For fecal
+Dependent on Cycle Number:}} A few small differences based on
+sub-sampling and cycle number were detected in overall bacterial
+community composition. Within the same fecal sample and independent of
+HiFi DNA polymerases, there were large differences in the community
+composition between 20 and 25 cycles {[}Figure 3A{]}. Additionally, when
+data was available for the mock communities, there were also large
+observed differences between 20 and 25 cycles {[}Figure 3B{]}. These
+community differences within a sample disappear when comparing 25 to 30
+cycles and do not persist past 25 cycles {[}Figure 3{]}. Although these
+trends are clearly noticeable, we found that there was no detectable
+difference in Bray-Curtis index when comparing to the previous 5-cycle
+increment for both fecal and mock communities after multiple comparison
+correction (P-value \textgreater{} 0.05). Using PERMANOVA to test for
+community differences based on cycle number within HiFi DNA polymerases,
+only Phusion had cycle dependent differences (P-value = 0.03. For fecal
 samples, Phusion was one of two HiFi DNA polymerases that had enough
-sequences to reach a sub-sampling depth of 1000 at 15 cycles. Overall,
-these data suggest that there are small HiFi DNA polymerase differences
-in Bray-Curtis index that are dependent on both sub-sampling depth and
-cycle number.
+sequences to be rarefied to 1000 at 15 cycles. Overall, these data
+suggest that there are small HiFi DNA polymerase differences in
+Bray-Curtis index that are dependent on cycle number.
 

### 2018-05-15 submission/manuscript.Rmd
remove sub-sampling and fix cycle jargon

@@
 A consistent difference in the range of number of OTUs, that was dependent on the HiFi DNA polymerase used was observed for fecal samples [Figure 1].
-Additionally, there was a trend for lower cycle numbers (15-20) to result in a reduced range in the number of OTUs versus higher cycle numbers (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
+Additionally, there was a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
 For fecal samples, there was a significant difference between HiFi DNA polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
@@
 This HiFi DNA polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2].
-Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across cycle number and sub-sampling depth [Figure 1-2 & Table S1-S4].
+Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
 In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
@@
 ***Minimal Bray-Curtis Differences are Detected and are Dependent on Cycle Number:***
-A few small differences based on sub-sampling and cycle number were detected in overall bacterial community composition.
+A few small differences based on the number of cycles were detected in overall bacterial community composition.
 Within the same fecal sample and independent of HiFi DNA polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A].
@@
 Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
-Using PERMANOVA to test for community differences based on cycle number within HiFi DNA polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
+Using PERMANOVA to test for community differences based on the number of cycles within HiFi DNA polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
 For fecal samples, Phusion was one of two HiFi DNA polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
-Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on cycle number.
+Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on the number of cycles.
 
 ***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number:***
-Differences in the median average per base error varied by HiFi DNA polymerase without a clear pattern across sub-sampling depth [Table S5].
-The highest median average per base error rates were for the Kappa HiFi DNA polymerase [Figure 4].
-This error rate was minimally affected by both the 'pre.cluster' step and chimera removal by VSEARCH [Figure 4].
-The differences in the median average per base error rate between the different HiFi DNA polymerases was cycle dependent with Platinum having the largest changes versus other HiFi DNA polymerases [Figure 4B-C and Table S6].
-The total sequences with at least one error was also cycle number dependent and differences between HiFi DNA polymerases could be drastically reduced by the use of the 'pre.cluster' step [Figure S1].
-These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerases [Figure S1 & Table S7 & S8].
-Finally, we did not observe a HiFi DNA polymerase dependent difference on base substitution rate [Figure S2].
-Although sequence error is dependent on HiFi DNA polymerase some of these error dependent differences can be corrected using existing bioinformatic approaches.
+Differences in the median error rate varied by HiFi DNA polymerase across the number of cycles used [Table S5].
+The highest error rates were for the Kappa HiFi DNA polymerase [Figure 4].
+The differences in the median error rate between the different HiFi DNA polymerases was dependent on the number of cycles with Platinum and Kappa having the most differences versus other HiFi DNA polymerases [Figure 4 and Table S6].
+The total sequences with at least one error was also cycle number dependent [Table S7 & S8.
+These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Kappa versus the other HiFi DNA polymerases [Table S7 & S8].
+Finally, we did not observe a HiFi DNA polymerase dependent difference on base substitution rate [Figure S1].
+These results suggest that sequence error is dependent on HiFi DNA polymerase.
+We next wanted to assess whether chimeras also were dependent on HiFi DNA polymerase and whether this affected OTU number.
 

### 2018-05-15 submission/manuscript.tex
remove sub-sampling and fix cycle jargon

@@
 dependent on the HiFi DNA polymerase used was observed for fecal samples
-{[}Figure 1{]}. Additionally, there was a trend for lower cycle numbers
-(15-20) to result in a reduced range in the number of OTUs versus higher
-cycle numbers (25, 30, and 35) between HiFi DNA polymerases {[}Figure
-1{]}. For fecal samples, there was a significant difference between HiFi
-DNA polymerases at 35 cycles (F-stat \textgreater{} 16.35, P-value =
-9.7e-05) {[}Table S1{]}. No differences between the specific groups was
-found based on a Tukey post-hoc test (P-value \textgreater{} 0.05)
-{[}Table S2{]}. This HiFi DNA polymerase dependent difference in the
-range of number of OTUs also was observed in the mock community samples
-{[}Figure 2{]}. Regardless if fecal or mock communities were used, the
-same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number
-of OTUs and this was consistent across cycle number and sub-sampling
-depth {[}Figure 1-2 \& Table S1-S4{]}. In contrast to the results
+{[}Figure 1{]}. Additionally, there was a trend for lower number of
+cycles (15-20) to result in a reduced range in the number of OTUs versus
+higher number of cycles (25, 30, and 35) between HiFi DNA polymerases
+{[}Figure 1{]}. For fecal samples, there was a significant difference
+between HiFi DNA polymerases at 35 cycles (F-stat \textgreater{} 16.35,
+P-value = 9.7e-05) {[}Table S1{]}. No differences between the specific
+groups was found based on a Tukey post-hoc test (P-value \textgreater{}
+0.05) {[}Table S2{]}. This HiFi DNA polymerase dependent difference in
+the range of number of OTUs also was observed in the mock community
+samples {[}Figure 2{]}. Regardless if fecal or mock communities were
+used, the same HiFi DNA polymerases had high (Platinum) and low
+(Accuprime) number of OTUs and this was consistent across the number of
+cycles used {[}Figure 1-2 \& Table S1-S4{]}. In contrast to the results
 obtained with fecal samples, differences between HiFi DNA polymerases
@@
 \textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
-Dependent on Cycle Number:}} A few small differences based on
-sub-sampling and cycle number were detected in overall bacterial
-community composition. Within the same fecal sample and independent of
-HiFi DNA polymerases, there were large differences in the community
-composition between 20 and 25 cycles {[}Figure 3A{]}. Additionally, when
-data was available for the mock communities, there were also large
-observed differences between 20 and 25 cycles {[}Figure 3B{]}. These
-community differences within a sample disappear when comparing 25 to 30
-cycles and do not persist past 25 cycles {[}Figure 3{]}. Although these
-trends are clearly noticeable, we found that there was no detectable
-difference in Bray-Curtis index when comparing to the previous 5-cycle
-increment for both fecal and mock communities after multiple comparison
-correction (P-value \textgreater{} 0.05). Using PERMANOVA to test for
-community differences based on cycle number within HiFi DNA polymerases,
-only Phusion had cycle dependent differences (P-value = 0.03. For fecal
-samples, Phusion was one of two HiFi DNA polymerases that had enough
-sequences to be rarefied to 1000 at 15 cycles. Overall, these data
-suggest that there are small HiFi DNA polymerase differences in
-Bray-Curtis index that are dependent on cycle number.
+Dependent on Cycle Number:}} A few small differences based on the number
+of cycles were detected in overall bacterial community composition.
+Within the same fecal sample and independent of HiFi DNA polymerases,
+there were large differences in the community composition between 20 and
+25 cycles {[}Figure 3A{]}. Additionally, when data was available for the
+mock communities, there were also large observed differences between 20
+and 25 cycles {[}Figure 3B{]}. These community differences within a
+sample disappear when comparing 25 to 30 cycles and do not persist past
+25 cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
+we found that there was no detectable difference in Bray-Curtis index
+when comparing to the previous 5-cycle increment for both fecal and mock
+communities after multiple comparison correction (P-value \textgreater{}
+0.05). Using PERMANOVA to test for community differences based on the
+number of cycles within HiFi DNA polymerases, only Phusion had cycle
+dependent differences (P-value = 0.03. For fecal samples, Phusion was
+one of two HiFi DNA polymerases that had enough sequences to be rarefied
+to 1000 at 15 cycles. Overall, these data suggest that there are small
+HiFi DNA polymerase differences in Bray-Curtis index that are dependent
+on the number of cycles.
 
 \textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
-and Cycle Number:}} Differences in the median average per base error
-varied by HiFi DNA polymerase without a clear pattern across
-sub-sampling depth {[}Table S5{]}. The highest median average per base
-error rates were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. This
-error rate was minimally affected by both the `pre.cluster' step and
-chimera removal by VSEARCH {[}Figure 4{]}. The differences in the median
-average per base error rate between the different HiFi DNA polymerases
-was cycle dependent with Platinum having the largest changes versus
-other HiFi DNA polymerases {[}Figure 4B-C and Table S6{]}. The total
-sequences with at least one error was also cycle number dependent and
-differences between HiFi DNA polymerases could be drastically reduced by
-the use of the `pre.cluster' step {[}Figure S1{]}. These differences in
+and Cycle Number:}} Differences in the median error rate varied by HiFi
+DNA polymerase across the number of cycles used {[}Table S5{]}. The
+highest error rates were for the Kappa HiFi DNA polymerase {[}Figure
+4{]}. The differences in the median error rate between the different
+HiFi DNA polymerases was dependent on the number of cycles with Platinum
+and Kappa having the most differences versus other HiFi DNA polymerases
+{[}Figure 4 and Table S6{]}. The total sequences with at least one error
+was also cycle number dependent {[}Table S7 \& S8. These differences in
 sequences with at least one error were mostly due to differences in
-Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
-polymerases {[}Figure S1 \& Table S7 \& S8{]}. Finally, we did not
-observe a HiFi DNA polymerase dependent difference on base substitution
-rate {[}Figure S2{]}. Although sequence error is dependent on HiFi DNA
-polymerase some of these error dependent differences can be corrected
-using existing bioinformatic approaches.
+Accuprime\textsuperscript{TM} and Kappa versus the other HiFi DNA
+polymerases {[}Table S7 \& S8{]}. Finally, we did not observe a HiFi DNA
+polymerase dependent difference on base substitution rate {[}Figure
+S1{]}. These results suggest that sequence error is dependent on HiFi
+DNA polymerase. We next wanted to assess whether chimeras also were
+dependent on HiFi DNA polymerase and whether this affected OTU number.
 

### 2018-05-15 submission/supplemental.Rmd
remove sub-sampling and fix cycle jargon

@@
   arrange(padj) %>% 
-  mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+  mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  arrange(cycle)
 
@@
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  filter(sub_sample_level > 500)
+  select(-sub_sample_level)
 
-colnames(mock_error_rate_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+colnames(mock_error_rate_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
 
@@
   mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
-  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  filter(pvalue < 0.05) %>% 
   arrange(pvalue) %>% 
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-bh)
+  select(-bh, -sub_sample_level) %>% 
+  arrange(cycle)
 
-colnames(mock_error_rate_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+colnames(mock_error_rate_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
 
@@
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  filter(sub_sample_level > 500)
+  select(-sub_sample_level)
 
-colnames(mock_error_count_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+colnames(mock_error_count_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
 
@@
   mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
-  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  filter(pvalue < 0.05) %>% 
   arrange(pvalue) %>% 
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-bh)
+  select(-bh, -sub_sample_level) %>% 
+  arrange(cycle)
 
-colnames(mock_error_count_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+colnames(mock_error_count_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
 

### 2018-05-15 submission/manuscript.Rmd
complete removal of sub-sampling and jargon cycle stuff from results

@@
 ***Prevalence of Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with the Number of OTUs:***
-There were significant differences in the chimera prevalence based on HiFi DNA polymerase used at all levels of sub-sampling and cycle numbers (P-value `r ifelse(max(filter(kruskal_chimera_prev, sub_sample_level > 500)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+There were significant differences in the chimera prevalence based on HiFi DNA polymerase used at all numbers of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
 Differences in chimera prevalence between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences [Table S10].
-Accuprime^TM^ had the lowest chimera prevalence of all HiFi DNA polymerases regardless of whether 'pre.cluster' or chimera removal with VSEARCH was used [Figure 5].
+Accuprime^TM^ had the lowest chimera prevalence of all HiFi DNA polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
+There was a plateau in the total percent of chimeras that were removed that was similar for all HiFi DNA polymerases [Figure 5C].
 A positive correlation was observed between chimeric sequences and the number of OTUs for all HiFi DNA polymerases [Figure 6].
 This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion HiFi DNA Polymerases [Figure 6].
-The R^2^ value between the number of OTUs and chimeric sequences did not change with the use of 'pre.cluster' or with the removal of chimeras using VSEARCH [Figure 6].
 This data suggests that chimera prevalence depends on HiFi DNA polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.

### 2018-05-15 submission/manuscript.tex
complete removal of sub-sampling and jargon cycle stuff from results

@@
 significant differences in the chimera prevalence based on HiFi DNA
-polymerase used at all levels of sub-sampling and cycle numbers (P-value
-\textless{} 0.05) {[}Table S9{]}. Differences in chimera prevalence
-between Platinum and all other HiFi DNA polymerases accounted for the
-majority of these differences {[}Table S10{]}.
-Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
-HiFi DNA polymerases regardless of whether `pre.cluster' or chimera
-removal with VSEARCH was used {[}Figure 5{]}. A positive correlation was
-observed between chimeric sequences and the number of OTUs for all HiFi
-DNA polymerases {[}Figure 6{]}. This positive correlation was strongest
-for Accuprime\textsuperscript{TM}, Platinum, and Phusion HiFi DNA
-Polymerases {[}Figure 6{]}. The R\textsuperscript{2} value between the
-number of OTUs and chimeric sequences did not change with the use of
-`pre.cluster' or with the removal of chimeras using VSEARCH {[}Figure
-6{]}. This data suggests that chimera prevalence depends on HiFi DNA
-polymerase used and confirms that the number of OTUs is dependent on the
-prevalence of these chimeric sequences.
+polymerase used at all numbers of cycles used (P-value \textless{} 0.05)
+{[}Table S9{]}. Differences in chimera prevalence between Platinum and
+all other HiFi DNA polymerases accounted for the majority of these
+differences {[}Table S10{]}. Accuprime\textsuperscript{TM} had the
+lowest chimera prevalence of all HiFi DNA polymerases regardless of
+whether chimera removal with VSEARCH was used {[}Figure 5A \& 5B{]}.
+There was a plateau in the total percent of chimeras that were removed
+that was similar for all HiFi DNA polymerases {[}Figure 5C{]}. A
+positive correlation was observed between chimeric sequences and the
+number of OTUs for all HiFi DNA polymerases {[}Figure 6{]}. This
+positive correlation was strongest for Accuprime\textsuperscript{TM},
+Platinum, and Phusion HiFi DNA Polymerases {[}Figure 6{]}. This data
+suggests that chimera prevalence depends on HiFi DNA polymerase used and
+confirms that the number of OTUs is dependent on the prevalence of these
+chimeric sequences.
 

### 2018-05-15 submission/supplemental.Rmd
complete removal of sub-sampling and jargon cycle stuff from results

@@
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  filter(sub_sample_level > 500)
+  select(-sub_sample_level)
 
-colnames(mock_chimera_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Sub-Sample Depth")
+colnames(mock_chimera_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
 
@@
   mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
-  filter(sub_sample_level > 500, pvalue < 0.05) %>% 
+  filter(pvalue < 0.05) %>% 
   arrange(pvalue) %>% 
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-bh)
+  select(-bh, -sub_sample_level) %>% 
+  arrange(cycle)
 
-colnames(mock_chimera_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Sub-Sample Depth")
+colnames(mock_chimera_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
 

### 2018-05-15 submission/manuscript.tex
complete removal of sub-sampling

@@
 Bray-Curtis indices were assessed for cycle group and within individual
-differences from the previous cycle (e.g.~20 versus 25, 25 versus 30,
-etc.). Based on these observations we analyzed potential reasons for
-these differences. Analysis of the mock community of each HiFi DNA
-polymerase for general sequence error rate, number of sequences with an
-error, base substitution, and numbers of chimeras were assessed before
-and after chimera removal. Additionally, the correlation between the
-number of chimeras and the number of OTUs was also assessed.
+differences from the previous cycle (e.g.~20 versus 25, 25 versus 30).
+Based on these observations we analyzed potential reasons for these
+differences. Analysis of the mock community of each HiFi DNA polymerase
+for general sequence error rate, number of sequences with an error, base
+substitution, and numbers of chimeras were assessed before and after
+chimera removal. Additionally, the correlation between the number of
+chimeras and the number of OTUs was also assessed.
 
@@
 between sample diversity, as measured by the Bray-Curtis index. Using
-this metric our observations found no differences in the overall
-bacterial community composition for sub-sampling depth or cycle number
-used. One possible reason for this outcome was that our study did not
-have enough power to detect differences due to low sample number in each
+this metric our observations generally found no differences in the
+overall bacterial community composition for the number of cycles used.
+One possible reason for this outcome was that our study did not have
+enough power to detect differences due to low sample number in each
 group. Another reason was that many of the OTUs are likely not highly

### 2018-05-15 submission/manuscript.Rmd
complete removal of sub-sampling
fix cycle number jargon

@@
 Commonly studied sources of bias include preservation and DNA extraction methods.
-Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
+Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
 Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.
@@
 **Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult.
-Care should be exercised when choosing both HiFi DNA polymerase and cycle number to be used in 16S rRNA gene sequencing studies.
+Care should be exercised when choosing both HiFi DNA polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
 
@@
 Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000].
-Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015].
+Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015].
 For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998].
 Both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
+incorrect amplicon size products) can also increase as the number of cycles increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
 Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011].
@@
 For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g.
-20 versus 25, 25 versus 30, etc.).
+20 versus 25, 25 versus 30).
 Based on these observations we analyzed potential reasons for these differences.
@@
 There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index.
-Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used.
+Using this metric our observations generally found no differences in the overall bacterial community composition for the number of cycles used.
 One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group.

### 2018-05-15 submission/manuscript.tex
fix cycle number jargon

@@
 rRNA gene sequencing workflow. Commonly studied sources of bias include
-preservation and DNA extraction methods. Although cycle number and high
-fidelity (HiFi) DNA polymerase are studied less often, they are still
-important sources of bias in this workflow. Here, we examine how both
-cycle number and HiFi DNA polymerase can change the bacterial community
-and introduce bias to the final obtained results.
+preservation and DNA extraction methods. Although the number of cycles
+and high fidelity (HiFi) DNA polymerase are studied less often, they are
+still important sources of bias in this workflow. Here, we examine how
+both cycle number and HiFi DNA polymerase can change the bacterial
+community and introduce bias to the final obtained results.
 
@@
 difficult. Care should be exercised when choosing both HiFi DNA
-polymerase and cycle number to be used in 16S rRNA gene sequencing
+polymerase and number of cycles to be used in 16S rRNA gene sequencing
 studies.
@@
 \& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
-made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
-2011; Kebschull \& Zador, 2015). For example, the selective
+made worse as the number of cycles increases (Wang \& Wang, 1996; Haas
+et al., 2011; Kebschull \& Zador, 2015). For example, the selective
 amplification of AT-rich over GC-rich sequences can exaggerate the
@@
 amplification (e.g.~incorrect amplicon size products) can also increase
-as cycle number increases which can drastically change commonly used
-diversity measures (Acinas et al., 2005; Santos et al., 2016).
+as the number of cycles increases which can drastically change commonly
+used diversity measures (Acinas et al., 2005; Santos et al., 2016).
 Additionally, chimeras can form from an aborted extension step followed

### 2018-05-16 submission/manuscript.Rmd
remove copious mention of HiFi DNA from manuscript

@@
 Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
-Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.
+Here, we examine how both cycle number and polymerase can change the bacterial community and introduce bias to the final obtained results.
 
@@
 The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities.
-It also assessed HiFi DNA polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence.
+It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence.
 Our analysis also examined chimera prevalence correlation with the number of OTUs.
@@
 **Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-Our analysis identified these HiFi dependent differences in the number of OTUs as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-Chimera prevalence varied by HiFi DNA polymerase and this variation persisted after chimera removal using VSEARCH.
+Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH.
 We also observed positive correlations between chimera prevalence and the number of OTUs which was not affected by chimera removal with VSEARCH.
@@
 **Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult.
-Care should be exercised when choosing both HiFi DNA polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
+Care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
 
@@
 
-A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016].
+A recent study found clear differences between normal and high fidelity DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016].
 This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
-It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of HiFi DNA polymerase.
+It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase.
 There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g.
@@
 
-Although bias introduced due to differences between DNA polymerase and HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific HiFi DNA polymerases has not been.
-This study will specifically address how HiFi DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
-We will accomplish this by examining if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been.
+This study will specifically address how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
+We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
 
@@
 ***PCR Protocol:***
-The five different HiFi DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
 The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified.
+The polymerases activation time was 2 minutes, unless a different activation was specified.
 For Kappa and Q5, a previously published protocol was used [@taq_Gohl_2016].
@@
 The cycle conditions for both fecal and mock samples started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
-The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total sample n=100).
-The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerases (total samples n=130).
+The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
+The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130).
 No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
@@
 Based on these observations we analyzed potential reasons for these differences.
-Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before and after chimera removal.
+Analysis of the mock community of each polymerase for sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
 Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
@@
 
-***The Number of OTUs are Dependent on HiFi DNA Polymerase:***
-A consistent difference in the range of number of OTUs, that was dependent on the HiFi DNA polymerase used was observed for fecal samples [Figure 1].
-Additionally, there was a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) between HiFi DNA polymerases [Figure 1].
-For fecal samples, there was a significant difference between HiFi DNA polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
+***The Number of OTUs are Dependent on Polymerase:***
+A consistent difference in the range of number of OTUs, that was dependent on the polymerase used was observed for fecal samples [Figure 1].
+Additionally, there was a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) between polymerases [Figure 1].
+For fecal samples, there was a significant difference between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
 No differences between the specific groups was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
-This HiFi DNA polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2].
-Regardless if fecal or mock communities were used, the same HiFi DNA polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between HiFi DNA polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-For the mock community, the majority of  differences in the number of OTUs were between Kappa and Platinum and the other HiFi DNA polymerases across different number of cycles [Table S4].
-Based on these observations in fecal and mock communities, it is clear that different HiFi DNA polymerases result in a different total number of OTUs within a sample.
+This polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2].
+Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
+In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+For the mock community, the majority of  differences in the number of OTUs were between Kappa and Platinum and the other polymerases across different number of cycles [Table S4].
+Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.
 
@@
 A few small differences based on the number of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of HiFi DNA polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A].
+Within the same fecal sample and independent of polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A].
 Additionally, when data was available for the mock communities, there were also large observed differences between 20 and 25 cycles [Figure 3B].
@@
 Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
-Using PERMANOVA to test for community differences based on the number of cycles within HiFi DNA polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
-For fecal samples, Phusion was one of two HiFi DNA polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
-Overall, these data suggest that there are small HiFi DNA polymerase differences in Bray-Curtis index that are dependent on the number of cycles.
+Using PERMANOVA to test for community differences based on the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
+For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
+Overall, these data suggest that there are small polymerase differences in Bray-Curtis index that are dependent on the number of cycles.
 
-***Sequence Error is Dependent on both HiFi DNA Polymerase and Cycle Number:***
-Differences in the median error rate varied by HiFi DNA polymerase across the number of cycles used [Table S5].
+***Sequence Error is Dependent on both Polymerase and Cycle Number:***
+Differences in the median error rate varied by polymerase across the number of cycles used [Table S5].
 The highest error rates were for the Kappa HiFi DNA polymerase [Figure 4].
-The differences in the median error rate between the different HiFi DNA polymerases was dependent on the number of cycles with Platinum and Kappa having the most differences versus other HiFi DNA polymerases [Figure 4 and Table S6].
+The differences in the median error rate between the different polymerases was dependent on the number of cycles with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6].
 The total sequences with at least one error was also cycle number dependent [Table S7 & S8.
-These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Kappa versus the other HiFi DNA polymerases [Table S7 & S8].
-Finally, we did not observe a HiFi DNA polymerase dependent difference on base substitution rate [Figure S1].
-These results suggest that sequence error is dependent on HiFi DNA polymerase.
-We next wanted to assess whether chimeras also were dependent on HiFi DNA polymerase and whether this affected OTU number.
+These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8].
+Finally, we did not observe a polymerase dependent difference on base substitution rate [Figure S1].
+These results suggest that sequence error is dependent on polymerase.
+We next wanted to assess whether chimeras also were dependent on polymerase and whether this affected OTU number.
 
-***Prevalence of Chimeric Sequences are HiFi DNA Polymerase Dependent and Correlate with the Number of OTUs:***
-There were significant differences in the chimera prevalence based on HiFi DNA polymerase used at all numbers of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Differences in chimera prevalence between Platinum and all other HiFi DNA polymerases accounted for the majority of these differences [Table S10].
-Accuprime^TM^ had the lowest chimera prevalence of all HiFi DNA polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
-There was a plateau in the total percent of chimeras that were removed that was similar for all HiFi DNA polymerases [Figure 5C].
-A positive correlation was observed between chimeric sequences and the number of OTUs for all HiFi DNA polymerases [Figure 6].
-This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion HiFi DNA Polymerases [Figure 6].
-This data suggests that chimera prevalence depends on HiFi DNA polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
+***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
+There were significant differences in the chimera prevalence based on polymerase used at all numbers of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S10].
+Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
+There was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
+A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6].
+This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6].
+This data suggests that chimera prevalence depends on polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 
@@
 
-In this study we show that the number of OTUs, error rate, and chimera prevalence depends on HiFi DNA polymerase [Figure 1-2 & 4-5].
+In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase [Figure 1-2 & 4-5].
 These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
 richness).
-Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
+Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
 We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences.
-This suggests that specific diversity differences between studies can be attributed to differences in HiFi DNA polyermase used.
-Based on our observations metrics that measure within sample diversity depend on HiFi DNA polymerase but this may not be the case for metrics that assess between sample diversity.
+This suggests that specific diversity differences between studies can be attributed to differences in polyermase used.
+Based on our observations metrics that measure within sample diversity depend on polymerase but this may not be the case for metrics that assess between sample diversity.
 
-There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index.
+There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index.
 Using this metric our observations generally found no differences in the overall bacterial community composition for the number of cycles used.
@@
 Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence.
+The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence.
 Metrics that solely use presence/absence of OTUs (e.g.
-Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases.
+Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in polymerases.
 When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
-With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.
+With differences in the number of OTUs and chimera prevalence depending on polymerase used, it might be easier to avoid specific DNA polymerase families altogether.
 
-Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
-Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent.
-Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases.
-We are unlikely to avoid adding HiFi dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.
+Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, this is unlikely to be the only contributor.
+For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
+Additionally, based on the material safety data sheet (MSDS) the differences between the two polymerases are not immediately apparent.
+Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the polymerases.
+We are unlikely to avoid adding polymerase dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.
 
-The majority of HiFi DNA polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences.
-The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
-However, the chimarea prevalence for some HiFi DNA polymerases (e.g.
+The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences.
+The sequence error introduced by the polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
+However, the chimarea prevalence for some polymerases (e.g.
 Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study.
-The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community.
-Although avoiding HiFi DNA polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
+The choice of polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of polymerase can add bias to the observed bacterial community.
+Although avoiding polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
 
@@
 Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016].
-This study shows that specific diversity metrics used to measure the microbial community consistently vary based on HiFi DNA polymerase.
-Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases may work better in certain situations over others.
+This study shows that specific diversity metrics used to measure the microbial community consistently vary based on polymerase.
+Standardizing multiple workflows to one specific polymerase could be detrimental since some polymerases may work better in certain situations over others.
 Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach.
@@
 
-Our observations fill a gap in knowledge on the bias introduced to 16S rRNA gene sequencing results due to differences in HiFi DNA polymerases.
-We found that the number of OTUs and the chimera prevalence is dependent on both HiFi DNA polymerase and cycle number chosen.
-Care should be taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
-Knowing the inherent bias associated with different HiFi DNA polymerases allows for better interpretation of the relationsip between an individual study to their respective field of research.
+Our observations fill a gap in knowledge on the bias introduced to 16S rRNA gene sequencing results due to differences in polymerases.
+We found that the number of OTUs and the chimera prevalence is dependent on both polymerase and cycle number chosen.
+Care should be taken when choosing a polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
+Knowing the inherent bias associated with different polymerases allows for better interpretation of the relationsip between an individual study to their respective field of research.
 

### 2018-05-16 submission/manuscript.tex
remove copious mention of HiFi DNA from manuscript

@@
 still important sources of bias in this workflow. Here, we examine how
-both cycle number and HiFi DNA polymerase can change the bacterial
-community and introduce bias to the final obtained results.
+both cycle number and polymerase can change the bacterial community and
+introduce bias to the final obtained results.
 
@@
 (OTUs) for both fecal samples and mock communities. It also assessed
-HiFi DNA polymerase dependent differences in the Bray-Curtis index,
-error rate, sequence error prevalence, and chimera prevalence. Our
-analysis also examined chimera prevalence correlation with the number of
-OTUs.
+polymerase dependent differences in the Bray-Curtis index, error rate,
+sequence error prevalence, and chimera prevalence. Our analysis also
+examined chimera prevalence correlation with the number of OTUs.
 
@@
 number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
-\textless{} 0.0001). Our analysis identified these HiFi dependent
+\textless{} 0.0001). Our analysis identified these polymerase dependent
 differences in the number of OTUs as early as 20 cycles in the mock
-communities (P-value = 0.002). Chimera prevalence varied by HiFi DNA
-polymerase and this variation persisted after chimera removal using
-VSEARCH. We also observed positive correlations between chimera
-prevalence and the number of OTUs which was not affected by chimera
-removal with VSEARCH.
+communities (P-value = 0.002). Chimera prevalence varied by polymerase
+and this variation persisted after chimera removal using VSEARCH. We
+also observed positive correlations between chimera prevalence and the
+number of OTUs which was not affected by chimera removal with VSEARCH.
 
@@
 number of OTUs and chimera prevalence makes comparison across studies
-difficult. Care should be exercised when choosing both HiFi DNA
-polymerase and number of cycles to be used in 16S rRNA gene sequencing
-studies.
+difficult. Care should be exercised when choosing the polymerase and
+number of cycles to be used in 16S rRNA gene sequencing studies.
 
@@
 A recent study found clear differences between normal and high fidelity
-(HiFi) DNA polymerase and that optimization of the PCR protocol could
-reduce error and chimera generation (Gohl et al., 2016). This study also
-found that regardless of DNA polymerase, the number of Operational
-Taxonomic Units (OTUs) or taxa generated were not easily reduced using
-the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
-natural to extend this line of inquiry and ask if biases in the number
-of OTUs and chimeras are also dependent on the type of HiFi DNA
-polymerase. There is some reason to think that this may be the case
-since many of these HiFi DNA polymerases come from different families
-(e.g. \emph{Taq} belongs to the family A polymerases) and may
-intrinsically have different error rates that cannot be completely
-removed with modifications (Ishino \& Ishino, 2014).
+DNA polymerase and that optimization of the PCR protocol could reduce
+error and chimera generation (Gohl et al., 2016). This study also found
+that regardless of DNA polymerase, the number of Operational Taxonomic
+Units (OTUs) or taxa generated were not easily reduced using the authors
+chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
+extend this line of inquiry and ask if biases in the number of OTUs and
+chimeras are also dependent on the type of high fidelity DNA polymerase.
+There is some reason to think that this may be the case since many of
+these HiFi DNA polymerases come from different families (e.g. \emph{Taq}
+belongs to the family A polymerases) and may intrinsically have
+different error rates that cannot be completely removed with
+modifications (Ishino \& Ishino, 2014).
 
 Although bias introduced due to differences between DNA polymerase and
-HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing,
-the bias caused due to differences between specific HiFi DNA polymerases
-has not been. This study will specifically address how HiFi DNA
-polymerases can bias observed bacterial community results derived from
-16S rRNA gene sequencing. We will accomplish this by examining if any of
-five different types of HiFi DNA polymerases introduce significant
-biases into 16S rRNA gene surveys, if this is a cycle dependent
-phenomenon, and whether they can be removed using a standard
-bioinformatic pipeline.
+high fidelity DNA polymerase has been investigated for 16S rRNA gene
+sequencing, the bias caused due to differences between specific high
+fidelity DNA polymerases has not been. This study will specifically
+address how high fidelity DNA polymerases can bias observed bacterial
+community results derived from 16S rRNA gene sequencing. We will
+accomplish this by examining if any of five different types of high
+fidelity DNA polymerases introduce significant biases into 16S rRNA gene
+surveys, if this is a cycle dependent phenomenon, and whether they can
+be removed using a standard bioinformatic pipeline.
 
@@
 
-\textbf{\emph{PCR Protocol:}} The five different HiFi DNA polymerases
-that were tested included AccuPrime\textsuperscript{TM} (ThermoFisher,
-MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA),
-Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-The PCR cycle conditions for Platinum and Accuprime followed a
-previously published protocol (Kozich et al., 2013)
+\textbf{\emph{PCR Protocol:}} The five different high fidelity DNA
+polymerases (hereto referred to as polymerases) that were tested
+included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
+HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
+(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR
+cycle conditions for Platinum and Accuprime followed a previously
+published protocol (Kozich et al., 2013)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
-The HiFi DNA polymerase activation time was 2 minutes, unless a
-different activation was specified. For Kappa and Q5, a previously
-published protocol was used (Gohl et al., 2016). For Phusion, the
-company defined conditions were used except for extension time, where
-the Accuprime and Platinum settings were used.
+The polymerases activation time was 2 minutes, unless a different
+activation was specified. For Kappa and Q5, a previously published
+protocol was used (Gohl et al., 2016). For Phusion, the company defined
+conditions were used except for extension time, where the Accuprime and
+Platinum settings were used.
 
@@
 increase for sequencing. The PCR of fecal DNA samples consisted of all 4
-samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase
-(total sample n=100). The mock communities had 4 replicates at 15, 20,
-25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA
-polymerases (total samples n=130). No mock community sample had enough
-PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
+samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total
+sample n=100). The mock communities had 4 replicates at 15, 20, 25, and
+35 cycles and 10 replicates for 30 cycles for all polymerases (total
+samples n=130). No mock community sample had enough PCR product at 15
+cycles for adequate 16S rRNA gene sequencing.
 
@@
 Based on these observations we analyzed potential reasons for these
-differences. Analysis of the mock community of each HiFi DNA polymerase
-for general sequence error rate, number of sequences with an error, base
-substitution, and numbers of chimeras were assessed before and after
-chimera removal. Additionally, the correlation between the number of
-chimeras and the number of OTUs was also assessed.
+differences. Analysis of the mock community of each polymerase for
+sequence error rate, number of sequences with an error, base
+substitution, and numbers of chimeras before and after chimera removal
+with VSEARCH was assessed. Additionally, the correlation between the
+number of chimeras and the number of OTUs was also assessed.
 
@@
 
-\textbf{\emph{The Number of OTUs are Dependent on HiFi DNA Polymerase:}}
-A consistent difference in the range of number of OTUs, that was
-dependent on the HiFi DNA polymerase used was observed for fecal samples
-{[}Figure 1{]}. Additionally, there was a trend for lower number of
-cycles (15-20) to result in a reduced range in the number of OTUs versus
-higher number of cycles (25, 30, and 35) between HiFi DNA polymerases
-{[}Figure 1{]}. For fecal samples, there was a significant difference
-between HiFi DNA polymerases at 35 cycles (F-stat \textgreater{} 16.35,
-P-value = 9.7e-05) {[}Table S1{]}. No differences between the specific
-groups was found based on a Tukey post-hoc test (P-value \textgreater{}
-0.05) {[}Table S2{]}. This HiFi DNA polymerase dependent difference in
-the range of number of OTUs also was observed in the mock community
-samples {[}Figure 2{]}. Regardless if fecal or mock communities were
-used, the same HiFi DNA polymerases had high (Platinum) and low
-(Accuprime) number of OTUs and this was consistent across the number of
-cycles used {[}Figure 1-2 \& Table S1-S4{]}. In contrast to the results
-obtained with fecal samples, differences between HiFi DNA polymerases
-were observed as early as 20 cycles in the mock community samples
-(F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
+\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}} A
+consistent difference in the range of number of OTUs, that was dependent
+on the polymerase used was observed for fecal samples {[}Figure 1{]}.
+Additionally, there was a trend for lower number of cycles (15-20) to
+result in a reduced range in the number of OTUs versus higher number of
+cycles (25, 30, and 35) between polymerases {[}Figure 1{]}. For fecal
+samples, there was a significant difference between polymerases at 35
+cycles (F-stat \textgreater{} 16.35, P-value = 9.7e-05) {[}Table S1{]}.
+No differences between the specific groups was found based on a Tukey
+post-hoc test (P-value \textgreater{} 0.05) {[}Table S2{]}. This
+polymerase dependent difference in the range of number of OTUs also was
+observed in the mock community samples {[}Figure 2{]}. Regardless if
+fecal or mock communities were used, the same polymerases had high
+(Platinum) and low (Accuprime) number of OTUs and this was consistent
+across the number of cycles used {[}Figure 1-2 \& Table S1-S4{]}. In
+contrast to the results obtained with fecal samples, differences between
+polymerases were observed as early as 20 cycles in the mock community
+samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
 community, the majority of differences in the number of OTUs were
-between Kappa and Platinum and the other HiFi DNA polymerases across
-different number of cycles {[}Table S4{]}. Based on these observations
-in fecal and mock communities, it is clear that different HiFi DNA
-polymerases result in a different total number of OTUs within a sample.
+between Kappa and Platinum and the other polymerases across different
+number of cycles {[}Table S4{]}. Based on these observations in fecal
+and mock communities, it is clear that different polymerases result in a
+different total number of OTUs within a sample.
 
@@
 of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of HiFi DNA polymerases,
-there were large differences in the community composition between 20 and
-25 cycles {[}Figure 3A{]}. Additionally, when data was available for the
-mock communities, there were also large observed differences between 20
-and 25 cycles {[}Figure 3B{]}. These community differences within a
-sample disappear when comparing 25 to 30 cycles and do not persist past
-25 cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
-we found that there was no detectable difference in Bray-Curtis index
-when comparing to the previous 5-cycle increment for both fecal and mock
+Within the same fecal sample and independent of polymerases, there were
+large differences in the community composition between 20 and 25 cycles
+{[}Figure 3A{]}. Additionally, when data was available for the mock
+communities, there were also large observed differences between 20 and
+25 cycles {[}Figure 3B{]}. These community differences within a sample
+disappear when comparing 25 to 30 cycles and do not persist past 25
+cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
+found that there was no detectable difference in Bray-Curtis index when
+comparing to the previous 5-cycle increment for both fecal and mock
 communities after multiple comparison correction (P-value \textgreater{}
 0.05). Using PERMANOVA to test for community differences based on the
-number of cycles within HiFi DNA polymerases, only Phusion had cycle
-dependent differences (P-value = 0.03. For fecal samples, Phusion was
-one of two HiFi DNA polymerases that had enough sequences to be rarefied
-to 1000 at 15 cycles. Overall, these data suggest that there are small
-HiFi DNA polymerase differences in Bray-Curtis index that are dependent
-on the number of cycles.
+number of cycles within polymerases, only Phusion had cycle dependent
+differences (P-value = 0.03. For fecal samples, Phusion was one of two
+polymerases that had enough sequences to be rarefied to 1000 at 15
+cycles. Overall, these data suggest that there are small polymerase
+differences in Bray-Curtis index that are dependent on the number of
+cycles.
 
-\textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
-and Cycle Number:}} Differences in the median error rate varied by HiFi
-DNA polymerase across the number of cycles used {[}Table S5{]}. The
-highest error rates were for the Kappa HiFi DNA polymerase {[}Figure
-4{]}. The differences in the median error rate between the different
-HiFi DNA polymerases was dependent on the number of cycles with Platinum
-and Kappa having the most differences versus other HiFi DNA polymerases
-{[}Figure 4 and Table S6{]}. The total sequences with at least one error
-was also cycle number dependent {[}Table S7 \& S8. These differences in
-sequences with at least one error were mostly due to differences in
-Accuprime\textsuperscript{TM} and Kappa versus the other HiFi DNA
-polymerases {[}Table S7 \& S8{]}. Finally, we did not observe a HiFi DNA
-polymerase dependent difference on base substitution rate {[}Figure
-S1{]}. These results suggest that sequence error is dependent on HiFi
-DNA polymerase. We next wanted to assess whether chimeras also were
-dependent on HiFi DNA polymerase and whether this affected OTU number.
+\textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
+Number:}} Differences in the median error rate varied by polymerase
+across the number of cycles used {[}Table S5{]}. The highest error rates
+were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. The differences
+in the median error rate between the different polymerases was dependent
+on the number of cycles with Platinum and Kappa having the most
+differences versus other polymerases {[}Figure 4 and Table S6{]}. The
+total sequences with at least one error was also cycle number dependent
+{[}Table S7 \& S8. These differences in sequences with at least one
+error were mostly due to differences in Accuprime\textsuperscript{TM}
+and Kappa versus the other polymerases {[}Table S7 \& S8{]}. Finally, we
+did not observe a polymerase dependent difference on base substitution
+rate {[}Figure S1{]}. These results suggest that sequence error is
+dependent on polymerase. We next wanted to assess whether chimeras also
+were dependent on polymerase and whether this affected OTU number.
 
-\textbf{\emph{Prevalence of Chimeric Sequences are HiFi DNA Polymerase
-Dependent and Correlate with the Number of OTUs:}} There were
-significant differences in the chimera prevalence based on HiFi DNA
-polymerase used at all numbers of cycles used (P-value \textless{} 0.05)
-{[}Table S9{]}. Differences in chimera prevalence between Platinum and
-all other HiFi DNA polymerases accounted for the majority of these
-differences {[}Table S10{]}. Accuprime\textsuperscript{TM} had the
-lowest chimera prevalence of all HiFi DNA polymerases regardless of
-whether chimera removal with VSEARCH was used {[}Figure 5A \& 5B{]}.
-There was a plateau in the total percent of chimeras that were removed
-that was similar for all HiFi DNA polymerases {[}Figure 5C{]}. A
-positive correlation was observed between chimeric sequences and the
-number of OTUs for all HiFi DNA polymerases {[}Figure 6{]}. This
-positive correlation was strongest for Accuprime\textsuperscript{TM},
-Platinum, and Phusion HiFi DNA Polymerases {[}Figure 6{]}. This data
-suggests that chimera prevalence depends on HiFi DNA polymerase used and
-confirms that the number of OTUs is dependent on the prevalence of these
-chimeric sequences.
+\textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
+and Correlate with the Number of OTUs:}} There were significant
+differences in the chimera prevalence based on polymerase used at all
+numbers of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
+Differences in chimera prevalence between Platinum and all other
+polymerases accounted for the majority of these differences {[}Table
+S10{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+of all polymerases regardless of whether chimera removal with VSEARCH
+was used {[}Figure 5A \& 5B{]}. There was a plateau in the total percent
+of chimeras that were removed that was similar for all polymerases
+{[}Figure 5C{]}. A positive correlation was observed between chimeric
+sequences and the number of OTUs for all polymerases {[}Figure 6{]}.
+This positive correlation was strongest for
+Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
+This data suggests that chimera prevalence depends on polymerase used
+and confirms that the number of OTUs is dependent on the prevalence of
+these chimeric sequences.
 
@@
 In this study we show that the number of OTUs, error rate, and chimera
-prevalence depends on HiFi DNA polymerase {[}Figure 1-2 \& 4-5{]}. These
+prevalence depends on polymerase {[}Figure 1-2 \& 4-5{]}. These
 differences are important because many diversity metrics rely on the
@@
 Additionally, the earlier detection of differences in total number of
-OTUs between HiFi DNA polymerases in the mock versus fecal samples might
-indicate that high biomass samples may underestimate the biases present
-within low biomass samples. We observed that undetected chimeras that
-were not identified and removed using standard bioinformatic approaches
-cause many of these differences. This suggests that specific diversity
-differences between studies can be attributed to differences in HiFi DNA
+OTUs between polymerases in the mock versus fecal samples might indicate
+that high biomass samples may underestimate the biases present within
+low biomass samples. We observed that undetected chimeras that were not
+identified and removed using standard bioinformatic approaches cause
+many of these differences. This suggests that specific diversity
+differences between studies can be attributed to differences in
 polyermase used. Based on our observations metrics that measure within
-sample diversity depend on HiFi DNA polymerase but this may not be the
-case for metrics that assess between sample diversity.
+sample diversity depend on polymerase but this may not be the case for
+metrics that assess between sample diversity.
 
-There were few differences that depend on HiFi DNA polymerase for
-between sample diversity, as measured by the Bray-Curtis index. Using
-this metric our observations generally found no differences in the
-overall bacterial community composition for the number of cycles used.
-One possible reason for this outcome was that our study did not have
-enough power to detect differences due to low sample number in each
-group. Another reason was that many of the OTUs are likely not highly
-abundant, allowing the Bray-Curtis index to be able to successfully
-down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
-diversity metric could be an important consideration in helping to
-mitigate these observed HiFi DNA polymerase dependent differences in
-chimera prevalence. Metrics that solely use presence/absence of OTUs
-(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
-prevalence and by extension total number of OTU differences in HiFi DNA
-Polymerases. When choosing a distance metric, careful consideration of
-the biases introduced from the PCR step of the 16S rRNA gene sequencing
-workflow need to be taken into account. With differences in the number
-of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
-might be easier to avoid specific DNA polymerase families altogether.
+There were few differences that depend on polymerase for between sample
+diversity, as measured by the Bray-Curtis index. Using this metric our
+observations generally found no differences in the overall bacterial
+community composition for the number of cycles used. One possible reason
+for this outcome was that our study did not have enough power to detect
+differences due to low sample number in each group. Another reason was
+that many of the OTUs are likely not highly abundant, allowing the
+Bray-Curtis index to be able to successfully down-weight chimeric OTUs
+(Minchin, 1987). The choice of downstream diversity metric could be an
+important consideration in helping to mitigate these observed polymerase
+dependent differences in chimera prevalence. Metrics that solely use
+presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
+less robust to chimera prevalence and by extension total number of OTU
+differences in polymerases. When choosing a distance metric, careful
+consideration of the biases introduced from the PCR step of the 16S rRNA
+gene sequencing workflow need to be taken into account. With differences
+in the number of OTUs and chimera prevalence depending on polymerase
+used, it might be easier to avoid specific DNA polymerase families
+altogether.
 
 Although the variation in error rate and chimera prevalence may be due
-to the DNA polymerase family, the highest and lowest chimera rates both
+to the DNA polymerase family, this is unlikely to be the only
+contributor. For example, the highest and lowest chimera rates both
 belonged to a family A polymerase (Platinum and
@@
 Additionally, based on the material safety data sheet (MSDS) the
-differences between the two HiFi DNA polymerases are not immediately
-apparent. Both HiFi DNA polymerases contain a recombinant \emph{Taq} DNA
-polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
-\emph{Taq} antibody. With everything else being equal, it is possible
-that differences in how the recombinant \emph{Taq} was generated could
-be a contributing factor for the differences observed between the HiFi
-DNA polymerases. We are unlikely to avoid adding HiFi dependent bias to
-16S rRNA gene sequencing results, however, these differences may not be
-large enough to mask the actual biological signal.
+differences between the two polymerases are not immediately apparent.
+Both polymerases contain a recombinant \emph{Taq} DNA polymerase, a
+\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
+antibody. With everything else being equal, it is possible that
+differences in how the recombinant \emph{Taq} was generated could be a
+contributing factor for the differences observed between the
+polymerases. We are unlikely to avoid adding polymerase dependent bias
+to 16S rRNA gene sequencing results, however, these differences may not
+be large enough to mask the actual biological signal.
 
-The majority of HiFi DNA polymerases we studied add small increases in
-the number of OTUs and chimera prevalence that can be masked by
-biological differences. The sequence error introduced by the HiFi DNA
-polymerase is small and likely to be smaller than the biological
-variation within a specific study, which would be consistent with
-previous findings for preservation and DNA extraction methods (Salter et
-al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
-prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
-large and might be greater than the oberseved biological variation
-within a specific study. The choice of HiFi DNA polymerase can be as
-important a consideration as either preservation or DNA extraction
-method used because similar to using different preservation methods or
-different DNA extraction kits, the type of HiFi DNA polymerase can add
-bias to the observed bacterial community. Although avoiding HiFi DNA
-polymerases that yield a high number of OTUs and chimera prevalence
-might provide better standardization of results it does come at a cost.
+The majority of polymerases we studied add small increases in the number
+of OTUs and chimera prevalence that can be masked by biological
+differences. The sequence error introduced by the polymerase is small
+and likely to be smaller than the biological variation within a specific
+study, which would be consistent with previous findings for preservation
+and DNA extraction methods (Salter et al., 2014; Song et al., 2016; Luo
+et al., 2016). However, the chimarea prevalence for some polymerases
+(e.g.~Platinum) are relatively large and might be greater than the
+oberseved biological variation within a specific study. The choice of
+polymerase can be as important a consideration as either preservation or
+DNA extraction method used because similar to using different
+preservation methods or different DNA extraction kits, the type of
+polymerase can add bias to the observed bacterial community. Although
+avoiding polymerases that yield a high number of OTUs and chimera
+prevalence might provide better standardization of results it does come
+at a cost.
 
@@
 metrics used to measure the microbial community consistently vary based
-on HiFi DNA polymerase. Standardizing multiple workflows to one specific
-HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
-may work better in certain situations over others. Arguably, the degree
-of workflow standardization across studies and research group needs to
-be approached on a study by study basis and not every project needs to
-use the exact same approach. All aspects of the 16S rRNA gene sequencing
-workflow need to be customized for the specific microbial community that
-will be sampled. Although a diversity of approaches may make
-reproducibility and replicability more difficult it will help to avoid
-systematic biases from occuring due to widespread standardization of
-approaches.
+on polymerase. Standardizing multiple workflows to one specific
+polymerase could be detrimental since some polymerases may work better
+in certain situations over others. Arguably, the degree of workflow
+standardization across studies and research group needs to be approached
+on a study by study basis and not every project needs to use the exact
+same approach. All aspects of the 16S rRNA gene sequencing workflow need
+to be customized for the specific microbial community that will be
+sampled. Although a diversity of approaches may make reproducibility and
+replicability more difficult it will help to avoid systematic biases
+from occuring due to widespread standardization of approaches.
 
@@
 Our observations fill a gap in knowledge on the bias introduced to 16S
-rRNA gene sequencing results due to differences in HiFi DNA polymerases.
-We found that the number of OTUs and the chimera prevalence is dependent
-on both HiFi DNA polymerase and cycle number chosen. Care should be
-taken when choosing a HiFi DNA polymerase for 16S rRNA gene surveys
-because their intrinsic differences can change the number of OTUs
-observed and influence diversity based metrics that do not down-weight
-rare observations. Knowing the inherent bias associated with different
-HiFi DNA polymerases allows for better interpretation of the relationsip
-between an individual study to their respective field of research.
+rRNA gene sequencing results due to differences in polymerases. We found
+that the number of OTUs and the chimera prevalence is dependent on both
+polymerase and cycle number chosen. Care should be taken when choosing a
+polymerase for 16S rRNA gene surveys because their intrinsic differences
+can change the number of OTUs observed and influence diversity based
+metrics that do not down-weight rare observations. Knowing the inherent
+bias associated with different polymerases allows for better
+interpretation of the relationsip between an individual study to their
+respective field of research.
 

### 2018-05-16 submission/manuscript.Rmd
remove extraction kit from abstract
update intro 2 paragraph --remove emphasis on preservation and extraction

@@
 **Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow.
-Commonly studied sources of bias include preservation and DNA extraction methods.
 Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
@@
 Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results.
-Differentiating between bias, reproducibility, and standardization is important since often times these three can be confused and used interchangeably with each other.
 Bias can change the observed results in a way that is reproducible and standardized.
-For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results [@contamination_Salter2014].
-Our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
+Within this context, our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
 In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.
@@
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-Generally, not using a preservation media and leaving samples at room temperature supports overgrowth of low abundance members of the fecal bacterial community [@Amir2017].
-Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016].
-Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015].
-Additionally, reagent contamination can add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014].
 Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017].
-However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
+Although the observed biological signal is larger, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
 
@@
 Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
-Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.
+Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.
 
@@
 It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase.
-There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g.
+There is some reason to think that this may be the case since many of these high fidelity DNA polymerases come from different families (e.g.
 *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
@@
 Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been.
-This study will specifically address how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
+This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
 We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-05-16 submission/manuscript.tex
remove extraction kit from abstract
update intro 2 paragraph --remove emphasis on preservation and extraction

@@
 \textbf{Background.} Bias is introduced at different stages of the 16S
-rRNA gene sequencing workflow. Commonly studied sources of bias include
-preservation and DNA extraction methods. Although the number of cycles
-and high fidelity (HiFi) DNA polymerase are studied less often, they are
-still important sources of bias in this workflow. Here, we examine how
-both cycle number and polymerase can change the bacterial community and
+rRNA gene sequencing workflow. Although the number of cycles and high
+fidelity (HiFi) DNA polymerase are studied less often, they are still
+important sources of bias in this workflow. Here, we examine how both
+cycle number and polymerase can change the bacterial community and
 introduce bias to the final obtained results.
@@
 Accounting for bias is critical in order to reach an understanding of
-bacterial community changes using 16S rRNA gene sequencing results.
-Differentiating between bias, reproducibility, and standardization is
-important since often times these three can be confused and used
-interchangeably with each other. Bias can change the observed results in
-a way that is reproducible and standardized. For example, if one group
-uses one brand of DNA extraction kit for their 16S rRNA gene sequencing,
-their results may be biased versus another group not using the same
-brand kit but within their group they can still have reproducible
-results (Salter et al., 2014). Our 16S rRNA gene sequencing methods are
-biased even when these workflows are standardized to increase
+bacterial community changes using 16S rRNA gene sequencing results. Bias
+can change the observed results in a way that is reproducible and
+standardized. Within this context, our 16S rRNA gene sequencing methods
+are biased even when these workflows are standardized to increase
 reproducibility. In order to interpret specific studies within the
@@
 sequencing workflow can be divided into preservation, extraction, PCR,
-and sequencing steps. Generally, not using a preservation media and
-leaving samples at room temperature supports overgrowth of low abundance
-members of the fecal bacterial community (Amir et al., 2017). Similarly,
-this overgrowth can still occur if the preservation media does not
-adequately inhibit growth (Song et al., 2016; Luo et al., 2016). Reports
-have also shown that changes in specific community members might occur
-due to differing susceptibility to freeze thaw cycles amongst microbes
-(Gorzelak et al., 2015). Additionally, reagent contamination can add
-community members and the contribution of these contaminant members
-grows larger with lower biomass samples (Salter et al., 2014). Overall,
-biases due to either preservation or extraction tend to be smaller than
-the overall biological signal being measured (Song et al., 2016; Bassis
-et al., 2017). However, the contribution of PCR bias to this overall
-workflow is not well characterized since these studies use the same PCR
-approach while varying preservation or extraction method.
+and sequencing steps. Overall, biases due to either preservation or
+extraction tend to be smaller than the overall biological signal being
+measured (Song et al., 2016; Bassis et al., 2017). Although the observed
+biological signal is larger, the contribution of PCR bias to this
+overall workflow is not well characterized since these studies use the
+same PCR approach while varying preservation or extraction method.
 
@@
 polymerases can have on the observed 16S rRNA gene sequencing results
-have not been well studied like some of the other previously mentioned
-sources of PCR-based bias.
+have not been well studied like some of the other sources of PCR-based
+bias.
 
@@
 There is some reason to think that this may be the case since many of
-these HiFi DNA polymerases come from different families (e.g. \emph{Taq}
-belongs to the family A polymerases) and may intrinsically have
-different error rates that cannot be completely removed with
+these high fidelity DNA polymerases come from different families (e.g.
+\emph{Taq} belongs to the family A polymerases) and may intrinsically
+have different error rates that cannot be completely removed with
 modifications (Ishino \& Ishino, 2014).
@@
 sequencing, the bias caused due to differences between specific high
-fidelity DNA polymerases has not been. This study will specifically
-address how high fidelity DNA polymerases can bias observed bacterial
-community results derived from 16S rRNA gene sequencing. We will
-accomplish this by examining if any of five different types of high
-fidelity DNA polymerases introduce significant biases into 16S rRNA gene
-surveys, if this is a cycle dependent phenomenon, and whether they can
-be removed using a standard bioinformatic pipeline.
+fidelity DNA polymerases has not been. This study will investigate how
+high fidelity DNA polymerases can bias observed bacterial community
+results derived from 16S rRNA gene sequencing. We will accomplish this
+by examining if any of five different types of high fidelity DNA
+polymerases introduce significant biases into 16S rRNA gene surveys, if
+this is a cycle dependent phenomenon, and whether they can be removed
+using a standard bioinformatic pipeline.
 
@@
 
-\hypertarget{ref-Gorzelak2015}{}
-Gorzelak MA., Gill SK., Tasnim N., Ahmadi-Vand Z., Jay M., Gibson DL.
-2015. Methods for improving human gut microbiome data by reducing
-variability through sample processing and storage of stool. \emph{PLOS
-ONE} 10:e0134802. DOI:
-\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.
-
 \hypertarget{ref-Haas2011}{}

### 2018-05-16 submission/manuscript.Rmd
update to downplay specific bias call outs in discussion

@@
 
-In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase [Figure 1-2 & 4-5].
+In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-2 & 4-5].
 These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
@@
 The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences.
-The sequence error introduced by the polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
-However, the chimarea prevalence for some polymerases (e.g.
-Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study.
-The choice of polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of polymerase can add bias to the observed bacterial community.
-Although avoiding polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
+The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
+The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations.
+Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.
 

### 2018-05-16 submission/manuscript.tex
update to downplay specific bias call outs in discussion

@@
 In this study we show that the number of OTUs, error rate, and chimera
-prevalence depends on polymerase {[}Figure 1-2 \& 4-5{]}. These
+prevalence depends on polymerase used {[}Figure 1-2 \& 4-5{]}. These
 differences are important because many diversity metrics rely on the
@@
 of OTUs and chimera prevalence that can be masked by biological
-differences. The sequence error introduced by the polymerase is small
-and likely to be smaller than the biological variation within a specific
-study, which would be consistent with previous findings for preservation
-and DNA extraction methods (Salter et al., 2014; Song et al., 2016; Luo
-et al., 2016). However, the chimarea prevalence for some polymerases
-(e.g.~Platinum) are relatively large and might be greater than the
-oberseved biological variation within a specific study. The choice of
-polymerase can be as important a consideration as either preservation or
-DNA extraction method used because similar to using different
-preservation methods or different DNA extraction kits, the type of
-polymerase can add bias to the observed bacterial community. Although
-avoiding polymerases that yield a high number of OTUs and chimera
-prevalence might provide better standardization of results it does come
-at a cost.
+differences. The sequence error introduced by the polymerase is also
+small and likely to be smaller than the biological variation within a
+specific study, which would be consistent with previous findings for
+preservation and DNA extraction methods (Salter et al., 2014; Song et
+al., 2016; Luo et al., 2016). The choice of polymerase should be an
+important consideration in the creation of a 16S rRNA gene sequencing
+workflow because using different polymerases can add error and bias to
+the downstream observations. Although standardization of the workflow
+may partially solve this problem by introducing a consistent bias to all
+samples, it does come at a cost.
 

### 2018-05-16 submission/manuscript.Rmd
re-write results section

@@
 ***The Number of OTUs are Dependent on Polymerase:***
-A consistent difference in the range of number of OTUs, that was dependent on the polymerase used was observed for fecal samples [Figure 1].
-Additionally, there was a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) between polymerases [Figure 1].
-For fecal samples, there was a significant difference between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
-No differences between the specific groups was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
-This polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2].
+Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used [Figure 1].
+Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases [Figure 1].
+There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
+No differences between the polymerases was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
+The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2].
 Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-For the mock community, the majority of  differences in the number of OTUs were between Kappa and Platinum and the other polymerases across different number of cycles [Table S4].
+In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
+The majority of differences for the number of OTUs detected in the Mock community is largely due to Kappa and Platinum differences to the other polymerases across the different number of cycles [Table S4].
 Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.
@@
 A few small differences based on the number of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A].
-Additionally, when data was available for the mock communities, there were also large observed differences between 20 and 25 cycles [Figure 3B].
+Within the same fecal sample and independent of polymerases, there are large differences in the community composition between 20 and 25 cycles [Figure 3A].
+Additionally, when data was available for the mock communities, there are also large observed differences between 20 and 25 cycles [Figure 3B].
 These community differences within a sample disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
-Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
+Although these trends are clearly noticeable, there is no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
 Using PERMANOVA to test for community differences based on the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
@@
 ***Sequence Error is Dependent on both Polymerase and Cycle Number:***
-Differences in the median error rate varied by polymerase across the number of cycles used [Table S5].
-The highest error rates were for the Kappa HiFi DNA polymerase [Figure 4].
-The differences in the median error rate between the different polymerases was dependent on the number of cycles with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6].
-The total sequences with at least one error was also cycle number dependent [Table S7 & S8.
-These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8].
-Finally, we did not observe a polymerase dependent difference on base substitution rate [Figure S1].
-These results suggest that sequence error is dependent on polymerase.
-We next wanted to assess whether chimeras also were dependent on polymerase and whether this affected OTU number.
+Differences in the median error rate varied by polymerase, with Kappa being the highest [Figure 4 & Table S5].
+The differences in the median error rate between the polymerases also is dependent on the number of cycles, with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6].
+The total sequences with at least one error also is dependent on the number of cycles [Table S7 & S8].
+These differences in sequences with at least one error is mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8].
+The differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
+These results suggest that sequence error is dependent on polymerase and the number of cycles, but is not attributable to a specific type.
+Based on these results, we examined whether chimeras also were dependent on polymerase and whether this also could affect the number of OTUs.
 
 ***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
-There were significant differences in the chimera prevalence based on polymerase used at all numbers of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
+There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
 Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S10].
 Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
-There was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
+Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
 A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6].
@@
 
-In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-2 & 4-5].
+In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-5].
 These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
@@
 We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences.
-This suggests that specific diversity differences between studies can be attributed to differences in polyermase used.
-Based on our observations metrics that measure within sample diversity depend on polymerase but this may not be the case for metrics that assess between sample diversity.
+This suggests that some of the specific diversity differences between studies can be attributed to differences in polyermase used.
+Based on these observations, metrics that measure within sample diversity like richness depend on polymerase but this may not be the case for metrics that assess between sample diversity.
 
 There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index.
-Using this metric our observations generally found no differences in the overall bacterial community composition for the number of cycles used.
-One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group.
-Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
+Our observations generally found no differences in the overall bacterial community composition for the number of cycles used.
+One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group.
+Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
 The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence.
@@
 For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
-Additionally, based on the material safety data sheet (MSDS) the differences between the two polymerases are not immediately apparent.
+Additionally, based on the material safety data sheet the differences between the two polymerases are not immediately apparent.
 Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
@@
 
-The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences.
+The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence and may be masked by biological differences.
 The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].

### 2018-05-16 submission/manuscript.tex
re-write results section

@@
 
-\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}} A
-consistent difference in the range of number of OTUs, that was dependent
-on the polymerase used was observed for fecal samples {[}Figure 1{]}.
-Additionally, there was a trend for lower number of cycles (15-20) to
+\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}}
+Differences in the range of the number of OTUs detected for fecal
+samples is dependent on the polymerase used {[}Figure 1{]}.
+Additionally, there is a trend for lower number of cycles (15-20) to
 result in a reduced range in the number of OTUs versus higher number of
-cycles (25, 30, and 35) between polymerases {[}Figure 1{]}. For fecal
-samples, there was a significant difference between polymerases at 35
-cycles (F-stat \textgreater{} 16.35, P-value = 9.7e-05) {[}Table S1{]}.
-No differences between the specific groups was found based on a Tukey
-post-hoc test (P-value \textgreater{} 0.05) {[}Table S2{]}. This
-polymerase dependent difference in the range of number of OTUs also was
-observed in the mock community samples {[}Figure 2{]}. Regardless if
-fecal or mock communities were used, the same polymerases had high
-(Platinum) and low (Accuprime) number of OTUs and this was consistent
-across the number of cycles used {[}Figure 1-2 \& Table S1-S4{]}. In
-contrast to the results obtained with fecal samples, differences between
-polymerases were observed as early as 20 cycles in the mock community
-samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
-community, the majority of differences in the number of OTUs were
-between Kappa and Platinum and the other polymerases across different
-number of cycles {[}Table S4{]}. Based on these observations in fecal
-and mock communities, it is clear that different polymerases result in a
-different total number of OTUs within a sample.
+cycles (25, 30, and 35) for all polymerases {[}Figure 1{]}. There is an
+overall difference in the number of OTUs detected within fecal samples
+between polymerases at 35 cycles (F-stat \textgreater{} 16.35, P-value =
+9.7e-05) {[}Table S1{]}. No differences between the polymerases was
+found based on a Tukey post-hoc test (P-value \textgreater{} 0.05)
+{[}Table S2{]}. The polymerase dependent difference in the range of the
+number of OTUs also was observed in the mock community samples {[}Figure
+2{]}. Regardless if fecal or mock communities were used, the same
+polymerases had high (Platinum) and low (Accuprime) number of OTUs and
+this was consistent across the number of cycles used {[}Figure 1-2 \&
+Table S1-S4{]}. In contrast to the results obtained with fecal samples,
+differences between polymerases were observed as early as 20 cycles in
+the mock community (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. The
+majority of differences for the number of OTUs detected in the Mock
+community is largely due to Kappa and Platinum differences to the other
+polymerases across the different number of cycles {[}Table S4{]}. Based
+on these observations in fecal and mock communities, it is clear that
+different polymerases result in a different total number of OTUs within
+a sample.
 
@@
 of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of polymerases, there were
+Within the same fecal sample and independent of polymerases, there are
 large differences in the community composition between 20 and 25 cycles
 {[}Figure 3A{]}. Additionally, when data was available for the mock
-communities, there were also large observed differences between 20 and
-25 cycles {[}Figure 3B{]}. These community differences within a sample
+communities, there are also large observed differences between 20 and 25
+cycles {[}Figure 3B{]}. These community differences within a sample
 disappear when comparing 25 to 30 cycles and do not persist past 25
-cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
-found that there was no detectable difference in Bray-Curtis index when
-comparing to the previous 5-cycle increment for both fecal and mock
-communities after multiple comparison correction (P-value \textgreater{}
-0.05). Using PERMANOVA to test for community differences based on the
-number of cycles within polymerases, only Phusion had cycle dependent
-differences (P-value = 0.03. For fecal samples, Phusion was one of two
-polymerases that had enough sequences to be rarefied to 1000 at 15
-cycles. Overall, these data suggest that there are small polymerase
-differences in Bray-Curtis index that are dependent on the number of
-cycles.
+cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
+there is no detectable difference in Bray-Curtis index when comparing to
+the previous 5-cycle increment for both fecal and mock communities after
+multiple comparison correction (P-value \textgreater{} 0.05). Using
+PERMANOVA to test for community differences based on the number of
+cycles within polymerases, only Phusion had cycle dependent differences
+(P-value = 0.03. For fecal samples, Phusion was one of two polymerases
+that had enough sequences to be rarefied to 1000 at 15 cycles. Overall,
+these data suggest that there are small polymerase differences in
+Bray-Curtis index that are dependent on the number of cycles.
 
 \textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
-Number:}} Differences in the median error rate varied by polymerase
-across the number of cycles used {[}Table S5{]}. The highest error rates
-were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. The differences
-in the median error rate between the different polymerases was dependent
-on the number of cycles with Platinum and Kappa having the most
+Number:}} Differences in the median error rate varied by polymerase,
+with Kappa being the highest {[}Figure 4 \& Table S5{]}. The differences
+in the median error rate between the polymerases also is dependent on
+the number of cycles, with Platinum and Kappa having the most
 differences versus other polymerases {[}Figure 4 and Table S6{]}. The
-total sequences with at least one error was also cycle number dependent
-{[}Table S7 \& S8. These differences in sequences with at least one
-error were mostly due to differences in Accuprime\textsuperscript{TM}
-and Kappa versus the other polymerases {[}Table S7 \& S8{]}. Finally, we
-did not observe a polymerase dependent difference on base substitution
-rate {[}Figure S1{]}. These results suggest that sequence error is
-dependent on polymerase. We next wanted to assess whether chimeras also
-were dependent on polymerase and whether this affected OTU number.
+total sequences with at least one error also is dependent on the number
+of cycles {[}Table S7 \& S8{]}. These differences in sequences with at
+least one error is mostly due to differences in
+Accuprime\textsuperscript{TM} and Kappa versus the other polymerases
+{[}Table S7 \& S8{]}. The differences in error rates were not due to
+polymerase dependent differences in base substitution rate {[}Figure
+S1{]}. These results suggest that sequence error is dependent on
+polymerase and the number of cycles, but is not attributable to a
+specific type. Based on these results, we examined whether chimeras also
+were dependent on polymerase and whether this also could affect the
+number of OTUs.
 
 \textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
-and Correlate with the Number of OTUs:}} There were significant
-differences in the chimera prevalence based on polymerase used at all
-numbers of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
+and Correlate with the Number of OTUs:}} There is significant
+differences in the chimera prevalence based on polymerase at all the
+number of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
 Differences in chimera prevalence between Platinum and all other
@@
 of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 5A \& 5B{]}. There was a plateau in the total percent
-of chimeras that were removed that was similar for all polymerases
-{[}Figure 5C{]}. A positive correlation was observed between chimeric
-sequences and the number of OTUs for all polymerases {[}Figure 6{]}.
-This positive correlation was strongest for
+was used {[}Figure 5A \& 5B{]}. Additionally, there was a plateau in the
+total percent of chimeras that were removed that was similar for all
+polymerases {[}Figure 5C{]}. A positive correlation was observed between
+chimeric sequences and the number of OTUs for all polymerases {[}Figure
+6{]}. This positive correlation was strongest for
 Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
@@
 In this study we show that the number of OTUs, error rate, and chimera
-prevalence depends on polymerase used {[}Figure 1-2 \& 4-5{]}. These
+prevalence depends on polymerase used {[}Figure 1-5{]}. These
 differences are important because many diversity metrics rely on the
@@
 identified and removed using standard bioinformatic approaches cause
-many of these differences. This suggests that specific diversity
-differences between studies can be attributed to differences in
-polyermase used. Based on our observations metrics that measure within
-sample diversity depend on polymerase but this may not be the case for
-metrics that assess between sample diversity.
+many of these differences. This suggests that some of the specific
+diversity differences between studies can be attributed to differences
+in polyermase used. Based on these observations, metrics that measure
+within sample diversity like richness depend on polymerase but this may
+not be the case for metrics that assess between sample diversity.
 
 There were few differences that depend on polymerase for between sample
-diversity, as measured by the Bray-Curtis index. Using this metric our
-observations generally found no differences in the overall bacterial
-community composition for the number of cycles used. One possible reason
-for this outcome was that our study did not have enough power to detect
-differences due to low sample number in each group. Another reason was
+diversity, as measured by the Bray-Curtis index. Our observations
+generally found no differences in the overall bacterial community
+composition for the number of cycles used. One possible reason for this
+outcome is that our study did not have enough power to detect
+differences due to low sample number in each group. Another reason is
 that many of the OTUs are likely not highly abundant, allowing the
@@
 Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
-Additionally, based on the material safety data sheet (MSDS) the
-differences between the two polymerases are not immediately apparent.
-Both polymerases contain a recombinant \emph{Taq} DNA polymerase, a
+Additionally, based on the material safety data sheet the differences
+between the two polymerases are not immediately apparent. Both
+polymerases contain a recombinant \emph{Taq} DNA polymerase, a
 \emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
@@
 The majority of polymerases we studied add small increases in the number
-of OTUs and chimera prevalence that can be masked by biological
+of OTUs and chimera prevalence and may be masked by biological
 differences. The sequence error introduced by the polymerase is also

### 2018-05-16 submission/manuscript.Rmd
re-write abstract

@@
 Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
-Here, we examine how both cycle number and polymerase can change the bacterial community and introduce bias to the final obtained results.
+Here, we examine how both cycle number and polymerase can change the bacterial community and introduce both error and bias to the final obtained results.
 
@@
 Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach.
-The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities.
-It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence.
-Our analysis also examined chimera prevalence correlation with the number of OTUs.
+The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities.
+It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and chimera prevalence correlation with the number of OTUs.
 
 **Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
 Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs which was not affected by chimera removal with VSEARCH.
+We also observed positive correlations between chimera prevalence and the number of OTUs.
 

### 2018-05-16 submission/manuscript.tex
re-write abstract

@@
 cycle number and polymerase can change the bacterial community and
-introduce bias to the final obtained results.
+introduce both error and bias to the final obtained results.
 
@@
 analysis initially examined the number of Operational Taxonomic Units
-(OTUs) for both fecal samples and mock communities. It also assessed
+(OTUs) for fecal samples and mock communities. It also assessed
 polymerase dependent differences in the Bray-Curtis index, error rate,
-sequence error prevalence, and chimera prevalence. Our analysis also
-examined chimera prevalence correlation with the number of OTUs.
+sequence error prevalence, chimera prevalence, and chimera prevalence
+correlation with the number of OTUs.
 
@@
 \textless{} 0.0001). Our analysis identified these polymerase dependent
-differences in the number of OTUs as early as 20 cycles in the mock
+differences in the number of OTUs as early as 20 cycles in mock
 communities (P-value = 0.002). Chimera prevalence varied by polymerase
@@
 also observed positive correlations between chimera prevalence and the
-number of OTUs which was not affected by chimera removal with VSEARCH.
+number of OTUs.
 

### 2018-05-24 submission/manuscript.Rmd
merge tables together and update manuscript

@@
 csl: peerj.csl
-fontsize: 12pt
+fontsize: 11pt
 output:
@@
 Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3].
-The majority of differences for the number of OTUs detected in the Mock community is largely due to Kappa and Platinum differences to the other polymerases across the different number of cycles [Table S4].
+In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1].
+The majority of differences for the number of OTUs detected in the Mock community is largely due to Kappa and Platinum differences to the other polymerases across the different number of cycles [Table S2].
 Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.
@@
 ***Sequence Error is Dependent on both Polymerase and Cycle Number:***
-Differences in the median error rate varied by polymerase, with Kappa being the highest [Figure 4 & Table S5].
-The differences in the median error rate between the polymerases also is dependent on the number of cycles, with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6].
-The total sequences with at least one error also is dependent on the number of cycles [Table S7 & S8].
-These differences in sequences with at least one error is mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8].
+Differences in the median error rate varied by polymerase, with Kappa being the highest [Figure 4 & Table S3].
+The differences in the median error rate between the polymerases also is dependent on the number of cycles, with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S4].
+The total sequences with at least one error also is dependent on the number of cycles [Table S3 & S4].
+These differences in sequences with at least one error is mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S3 & S4].
 The differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
@@
 ***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
-There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9].
-Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S10].
+There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
+Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S4].
 Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].

### 2018-05-24 submission/manuscript.tex
merge tables together and update manuscript

@@
-\documentclass[12pt,]{article}
+\documentclass[11pt,]{article}
 \usepackage{lmodern}
@@
 differences between polymerases were observed as early as 20 cycles in
-the mock community (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. The
+the mock community (F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. The
 majority of differences for the number of OTUs detected in the Mock
 community is largely due to Kappa and Platinum differences to the other
-polymerases across the different number of cycles {[}Table S4{]}. Based
+polymerases across the different number of cycles {[}Table S2{]}. Based
 on these observations in fecal and mock communities, it is clear that
@@
 Number:}} Differences in the median error rate varied by polymerase,
-with Kappa being the highest {[}Figure 4 \& Table S5{]}. The differences
+with Kappa being the highest {[}Figure 4 \& Table S3{]}. The differences
 in the median error rate between the polymerases also is dependent on
 the number of cycles, with Platinum and Kappa having the most
-differences versus other polymerases {[}Figure 4 and Table S6{]}. The
+differences versus other polymerases {[}Figure 4 and Table S4{]}. The
 total sequences with at least one error also is dependent on the number
-of cycles {[}Table S7 \& S8{]}. These differences in sequences with at
+of cycles {[}Table S3 \& S4{]}. These differences in sequences with at
 least one error is mostly due to differences in
 Accuprime\textsuperscript{TM} and Kappa versus the other polymerases
-{[}Table S7 \& S8{]}. The differences in error rates were not due to
+{[}Table S3 \& S4{]}. The differences in error rates were not due to
 polymerase dependent differences in base substitution rate {[}Figure
@@
 differences in the chimera prevalence based on polymerase at all the
-number of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
+number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
 Differences in chimera prevalence between Platinum and all other
 polymerases accounted for the majority of these differences {[}Table
-S10{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+S4{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
 of all polymerases regardless of whether chimera removal with VSEARCH

### 2018-05-24 submission/supplemental.Rmd
merge tables together and update manuscript

@@
 
-**Table S1: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal Samples**
+**Table S1: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal and Mock Samples**
 
@@
   mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  mutate(sample_type = "Fecal Sample") %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv")) %>% 
+    mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    filter(sub_sample_level == 1000) %>% 
+    select(-sub_sample_level) %>% 
+    mutate(sample_type = "Mock Sample"))
 
-colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle")
+colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sample Type")
 
@@
 
-**Table S2: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal Samples**
+**Table S2: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal and Mock Samples**
 
@@
   mutate(padj = as.numeric(padj)) %>% 
-  mutate_at(vars(padj), funs(format(., digits = 2, nsmall = 2)))
-
-colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle")
-
-kable(fs_tukey, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S3: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Mock Samples**
-
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_anova <- as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv")) %>% 
-  mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  filter(sub_sample_level == 1000) %>% 
-  select(-sub_sample_level)
-
-colnames(mock_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle")
-
-kable(mock_anova, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S4: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Mock Samples**
-
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_tukey <- as.data.frame(read_csv("../data/process/tables/mock_overall_tukey_results.csv")) %>% 
+  mutate_at(vars(padj), funs(format(., digits = 2, nsmall = 2))) %>% 
+  mutate(sample_type = "Fecal Sample") %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_overall_tukey_results.csv")) %>% 
     mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
@@
     mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  arrange(cycle)
+    arrange(cycle) %>% 
+    mutate(sample_type = "Mock Sample"))
 
-colnames(mock_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle")
+colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sample Type")
 
-kable(mock_tukey, format = "markdown", align = 'c', padding = 2)
+kable(fs_tukey, format = "markdown", align = 'c', padding = 2)
 
@@
 
-**Table S5: Kruskal-Wallis Results of Differences in Per Base Error Rate between HiFi DNA Polymerases in Mock Samples**
+**Table S3: Kruskal-Wallis Results of Differences in Error Rate, Sequences with an Error, and Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
 
@@
   mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-sub_sample_level)
+  select(-sub_sample_level) %>% 
+  mutate(test_type = "Error Rate") %>% 
+  bind_rows(
+    as.data.frame(
+    read_csv("../data/process/tables/mock_error_count_overall_kruskal_results.csv")) %>% 
+    mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    select(-sub_sample_level) %>% 
+    mutate(test_type = "Sequences with Error")) %>% 
+  bind_rows(
+    as.data.frame(
+    read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv")) %>% 
+    mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    select(-sub_sample_level) %>% 
+    mutate(test_type = "Chimera Prevalence"))
 
-colnames(mock_error_rate_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
+colnames(mock_error_rate_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle", "Test Type")
 
@@
 
-**Table S6: Dunn's Post-Hoc Results of Differences in Per Base Error Rate between HiFi DNA Polymerases in Mock Samples**
+**Table S4: Dunn's Post-Hoc Results of Differences in Error Rate, Sequences with an Error, and Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
 
@@
   select(-bh, -sub_sample_level) %>% 
-  arrange(cycle)
+  arrange(cycle) %>% 
+  mutate(test_type = "Error Rate") %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_error_count_overall_dunn_results.csv")) %>% 
+    mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
+    filter(pvalue < 0.05) %>% 
+    arrange(pvalue) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    select(-bh, -sub_sample_level) %>% 
+    arrange(cycle) %>% 
+    mutate(test_type = "Sequences with Error")) %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")) %>% 
+    mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
+    filter(pvalue < 0.05) %>% 
+    arrange(pvalue) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    select(-bh, -sub_sample_level) %>% 
+    arrange(cycle) %>% 
+    mutate(test_type = "Chimera Prevalence"))
 
-colnames(mock_error_rate_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
+colnames(mock_error_rate_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle", "Test Type")
 
@@
 \newpage
-
-**Table S7: Kruskal-Wallis Results of Differences in Sequences with Error between HiFi DNA Polymerases in Mock Samples**
-
-```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_error_count_kruskal <- as.data.frame(
-  read_csv("../data/process/tables/mock_error_count_overall_kruskal_results.csv")) %>% 
-  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-sub_sample_level)
-
-colnames(mock_error_count_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
-
-kable(mock_error_count_kruskal, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S8: Dunn's Post-Hoc Results of Differences in Sequences with Error between HiFi DNA Polymerases in Mock Samples**
-
-```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_error_count_dunn <- as.data.frame(read_csv("../data/process/tables/mock_error_count_overall_dunn_results.csv")) %>% 
-  mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
-  filter(pvalue < 0.05) %>% 
-  arrange(pvalue) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-bh, -sub_sample_level) %>% 
-  arrange(cycle)
-
-colnames(mock_error_count_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
-
-kable(mock_error_count_dunn, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S9: Kruskal-Wallis Results of Differences in Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
-
-```{r supplemental_table9, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_chimera_kruskal <- as.data.frame(
-  read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv")) %>% 
-  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-sub_sample_level)
-
-colnames(mock_chimera_kruskal) <- c("Rank Sum Statistic", "DF", "P-value", "BH", "Cycle")
-
-kable(mock_chimera_kruskal, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S10: Dunn's Post-Hoc Results of Differences in Chimera Prevalence between HiFi DNA Polymerase in Mock Samples**
-
-```{r supplemental_table10, echo=FALSE, message=FALSE, warning=FALSE}
-
-mock_chimera_dunn <- as.data.frame(read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")) %>% 
-  mutate_at(vars(chi2:Z), funs(round(., digits = 2))) %>% 
-  filter(pvalue < 0.05) %>% 
-  arrange(pvalue) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  select(-bh, -sub_sample_level) %>% 
-  arrange(cycle)
-
-colnames(mock_chimera_dunn) <- c("Chi Squared", "Z Statistic", "P Adjusted", "Comparison", "Cycle")
-
-kable(mock_chimera_dunn, format = "markdown", align = 'c', padding = 2)
-
-```

### 2018-05-25 code/make_correlation_graphs.R
add correlation table write out

@@ -189,8 +189,23 @@ mock_error <- combined_list %>% filter(pipeline == "mock_error") %>%
 ###########################################################################################################################
 
 
-
+# Write out figure
 ggsave("results/figures/Figure6.pdf", mock_error, width = 8, height = 8, dpi = 300)
+
+# Write out regression data
+regression_table <- df_m_error %>% separate(eq, c("eq", "r2", "x1", "x2", "x3")) %>% 
+  select(taq, x3) %>% 
+  rename(r2 = x3) %>% 
+  mutate(r2 = as.numeric(r2), 
+         r2 = case_when(r2 < 100 ~ r2*10, 
+                        TRUE ~ r2), 
+         r2 = r2/1000, 
+         taq = as.character(taq))
+
+write_csv(regression_table, "data/process/tables/r2_otu_vs_chimera_summary.csv")
+
+
+
 
 
 

### 2018-05-25 submission/manuscript.Rmd
make changes to title and abstract

@@
 
+# Read in fecal count data
+fecal_numOTU_data <- read_csv("../data/process/tables/fecal_sub_sample_1000_count_table.csv") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")), 
+         cycles = factor(cycles, 
+                            levels = c("15x", "20x", "25x", "30x", "35x"), 
+                            labels = c("15x", "20x", "25x", "30x", "35x"))) %>% 
+  group_by(taq, cycles) %>% 
+  summarise(taq_median = median(numOTUs, na.rm = T), 
+            taq_min = min(numOTUs, na.rm = T), 
+            taq_max = max(numOTUs, na.rm = T)) %>% 
+  mutate(range_diff = taq_max - taq_min)
+
 # Read in fecal sample number of OTU results
@@
   rename(padj = `p adj`)
+
+# Read in the Mock count data
+mock_num_otu_data <- read_csv("../data/process/tables/mock_sub_sample_1000_count_table.csv") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
+  group_by(taq, cycles) %>% 
+  summarise(taq_median = median(numOTUs, na.rm = T), 
+            taq_min = min(numOTUs, na.rm = T), 
+            taq_max = max(numOTUs, na.rm = T)) %>% 
+  mutate(range_diff = taq_max - taq_min)
+
+ref_mock_otu_num <- read_tsv("../data/process/zymo_mock_only.shared") %>% 
+  select(numOtus)
 
@@
 
+# Read in actual error data
+error_data <- read_csv("../data/process/tables/mock_chimera_error_summary.csv") %>% 
+                       filter(total_seqs >= 1000)
+
+error_summary <- error_data %>% 
+  group_by(taq, cycles) %>% 
+  summarise_at(vars(mean_error, chimera_prevalence, seq_error_prevalence), funs(median, min, max)) %>% 
+  select(taq, cycles, contains("chimera"), contains("mean_error"), contains("seq_error"))
+
 # Vector of sub samples used
@@
 
+# Read in the regression data
+r2_table <- read_csv("../data/process/tables/r2_otu_vs_chimera_summary.csv")
+
 ```
 
-# High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene Sequencing Results
+# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used
 
@@
 
-**Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs.
+**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
+Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+Variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
+Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
+We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).
 
-**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult.
-Care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
+**Conclusions.** Although the variation in the number of OTUs observed in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for this difference.
+In mock community samples, the divergence in the number of OTUs between the polymerases which can be explained by chimera prevalence suggests that this is the main reason for differences between the polymerases.
+Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
 

### 2018-05-25 submission/manuscript.tex
make changes to title and abstract

@@
 
-\section{High Fidelity DNA Polymerase Introduces Bias into 16S rRNA Gene
-Sequencing
-Results}\label{high-fidelity-dna-polymerase-introduces-bias-into-16s-rrna-gene-sequencing-results}
+\section{Error Introduced into 16S rRNA Gene Sequencing Results Varies
+by High Fidelity DNA Polymerase
+Used}\label{error-introduced-into-16s-rrna-gene-sequencing-results-varies-by-high-fidelity-dna-polymerase-used}
 
@@
 
-\textbf{Results.} When analyzing fecal samples we observed a different
-number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
-\textless{} 0.0001). Our analysis identified these polymerase dependent
-differences in the number of OTUs as early as 20 cycles in mock
-communities (P-value = 0.002). Chimera prevalence varied by polymerase
+\textbf{Results.} When analyzing fecal samples we observed that the
+range in the number of OTUs detected was not consistent between HiFi DNA
+polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
+Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
+HiFi DNA polymerase used (P-value \textless{} 0.0001). Variations in the
+number of OTUs detected by DNA polymerase was observable as early as 20
+cycles in mock communities (P-value = 0.002). There also was a large
+range in the number of OTUs by HiFi DNA polymerase at 35 cycles
+(Accuprime = 15 - 20 versus Phusion = 14 - 73). Chimera prevalence in
+mock communities varied by polymerase with differences being most
+notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
 and this variation persisted after chimera removal using VSEARCH. We
 also observed positive correlations between chimera prevalence and the
-number of OTUs.
+number of OTUs with Platinum having the highest (R\^{}2 = 0.974) and
+Kappa having the worse (R\^{}2 = 0.478).
 
-\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
-number of OTUs and chimera prevalence makes comparison across studies
-difficult. Care should be exercised when choosing the polymerase and
-number of cycles to be used in 16S rRNA gene sequencing studies.
+\textbf{Conclusions.} Although the variation in the number of OTUs
+observed in fecal samples could be due to certain polymerases capturing
+the biological variability better than others, this is unlikely to be
+the main reason for this difference. In mock community samples, the
+divergence in the number of OTUs between the polymerases which can be
+explained by chimera prevalence suggests that this is the main reason
+for differences between the polymerases. Ultimately, this variation
+makes comparison across studies difficult and care should be exercised
+when choosing the polymerase and number of cycles to be used in 16S rRNA
+gene sequencing studies.
 

### 2018-05-25 submission/manuscript.Rmd
complete re-write of the abstract

@@
 
-**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow.
-Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow.
-Here, we examine how both cycle number and polymerase can change the bacterial community and introduce both error and bias to the final obtained results.
+**Background.** It is challenging to compare 16S rRNA gene sequencing data across studies and one of the reasons for this is due to error.
+There are many different places throughout the workflow where error can be introduced into the pipeline.
+Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error into this workflow because of less investigation into this area.
+By varying both cycle number and polymerase we assess the changes that occur to the bacterial community in human fecal and mock community samples.
 
@@
 The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities.
-It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and chimera prevalence correlation with the number of OTUs.
+It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.
 
@@
 Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-Variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+When analyzing mock community samples the variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
 There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
@@
 
-**Conclusions.** Although the variation in the number of OTUs observed in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for this difference.
-In mock community samples, the divergence in the number of OTUs between the polymerases which can be explained by chimera prevalence suggests that this is the main reason for differences between the polymerases.
-Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.
+**Conclusions.** Although the variation in the number of OTUs in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for our observed differences.
+In mock community samples, the strong correlation between chimera prevalence and the number of OTUs suggests that this is the main reason for differences between the polymerases.
+Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles in 16S rRNA gene sequencing studies.
 

### 2018-05-25 submission/manuscript.tex
complete re-write of the abstract

@@
 
-\textbf{Background.} Bias is introduced at different stages of the 16S
-rRNA gene sequencing workflow. Although the number of cycles and high
-fidelity (HiFi) DNA polymerase are studied less often, they are still
-important sources of bias in this workflow. Here, we examine how both
-cycle number and polymerase can change the bacterial community and
-introduce both error and bias to the final obtained results.
+\textbf{Background.} It is challenging to compare 16S rRNA gene
+sequencing data across studies and one of the reasons for this is due to
+error. There are many different places throughout the workflow where
+error can be introduced into the pipeline. Here, we focus on studying
+how the number of cycles and high fidelity (HiFi) DNA polymerase
+introduce error into this workflow because of less investigation into
+this area. By varying both cycle number and polymerase we assess the
+changes that occur to the bacterial community in human fecal and mock
+community samples.
 
@@
 polymerase dependent differences in the Bray-Curtis index, error rate,
-sequence error prevalence, chimera prevalence, and chimera prevalence
-correlation with the number of OTUs.
+sequence error prevalence, chimera prevalence, and the correlation
+between chimera prevalence and number of OTUs.
 
@@
 Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
-HiFi DNA polymerase used (P-value \textless{} 0.0001). Variations in the
-number of OTUs detected by DNA polymerase was observable as early as 20
-cycles in mock communities (P-value = 0.002). There also was a large
-range in the number of OTUs by HiFi DNA polymerase at 35 cycles
-(Accuprime = 15 - 20 versus Phusion = 14 - 73). Chimera prevalence in
-mock communities varied by polymerase with differences being most
-notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
-and this variation persisted after chimera removal using VSEARCH. We
-also observed positive correlations between chimera prevalence and the
-number of OTUs with Platinum having the highest (R\^{}2 = 0.974) and
-Kappa having the worse (R\^{}2 = 0.478).
+HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
+mock community samples the variations in the number of OTUs detected by
+DNA polymerase was observable as early as 20 cycles (P-value = 0.002).
+There also was a large range in the number of OTUs by HiFi DNA
+polymerase at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
+Chimera prevalence in mock communities varied by polymerase with
+differences being most notable at 35 cycles (Kappa = 5.71\% (median)
+versus Platinum = 26.62\%) and this variation persisted after chimera
+removal using VSEARCH. We also observed positive correlations between
+chimera prevalence and the number of OTUs with Platinum having the
+highest (R\^{}2 = 0.974) and Kappa having the worse (R\^{}2 = 0.478).
 
-\textbf{Conclusions.} Although the variation in the number of OTUs
-observed in fecal samples could be due to certain polymerases capturing
-the biological variability better than others, this is unlikely to be
-the main reason for this difference. In mock community samples, the
-divergence in the number of OTUs between the polymerases which can be
-explained by chimera prevalence suggests that this is the main reason
-for differences between the polymerases. Ultimately, this variation
-makes comparison across studies difficult and care should be exercised
-when choosing the polymerase and number of cycles to be used in 16S rRNA
-gene sequencing studies.
+\textbf{Conclusions.} Although the variation in the number of OTUs in
+fecal samples could be due to certain polymerases capturing the
+biological variability better than others, this is unlikely to be the
+main reason for our observed differences. In mock community samples, the
+strong correlation between chimera prevalence and the number of OTUs
+suggests that this is the main reason for differences between the
+polymerases. Ultimately, this variation makes comparison across studies
+difficult and care should be exercised when choosing the polymerase and
+number of cycles in 16S rRNA gene sequencing studies.
 

### 2018-05-25 submission/manuscript.Rmd
make changes inspired by Marcy to intro

@@
 There are many different places throughout the workflow where error can be introduced into the pipeline.
-Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error into this workflow because of less investigation into this area.
-By varying both cycle number and polymerase we assess the changes that occur to the bacterial community in human fecal and mock community samples.
+Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error by varying cycle number and polymerase used to amplify 16S rRNA genes in human fecal and mock community samples.
 
@@
 Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-When analyzing mock community samples the variations in the number of OTUs detected by DNA polymerase was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-There also was a large range in the number of OTUs by HiFi DNA polymerase at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
+When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
 Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worse (R^2 = `r min(pull(r2_table, r2))`).
+We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2 = `r min(pull(r2_table, r2))`).
 
@@
 
-Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results.
-Bias can change the observed results in a way that is reproducible and standardized.
-Within this context, our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
-In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.
+The bacterial community is reported to vary between case and control for a number of diseases [].
+However, for diseases like obesity, the taxa identified have varied widely depending on the study [].
+Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow.
+Yet, standardizing a 16S rRNA gene sequencing workflow will ultimately result in a standardized and reproducible bias due to choices made on the methods used for preservation, extraction, PCR, and sequencing.
+Within this context, all 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
+In order to interpret specific studies within the broader context of the overall field, assessing error at different parts of the 16S rRNA gene sequencing workflow is critical.
 
-Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively.
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017].
-Although the observed biological signal is larger, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.
+The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [].
+For preservation and extraction stages of the workflow, it has been consistently found that there are biases based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
+Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR bias to this overall workflow is not well characterized.
 
-Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000].
+There is a large body of literature that shows there are different biases within each of the steps in the PCR stage of 16S rRNA gene sequencing.
+One source of error is from primers chosen for the 16S rRNA gene sequencing.
+There are variable region dependent binding affinities for different bacteria, and depending on the primer pair used they do not detect specific bacteria (e.g.
+V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
+Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998].
 Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015].
-For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998].
-Both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) can also increase as the number of cycles increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
-Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011].
-Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.
-
-The intrinsic properties to primers and DNA polymerases chosen can also introduce bias.
-Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g.
-V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
-Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
+For example, both amplification error and non-specific amplification (e.g.
+incorrect amplicon size products) also can increase as the number of cycles increases.
+This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
+Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011].
+These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011].
+In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
 Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.
 
-A recent study found clear differences between normal and high fidelity DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016].
-This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016].
-It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase.
-There is some reason to think that this may be the case since many of these high fidelity DNA polymerases come from different families (e.g.
-*Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].
-
-Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been.
+A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
+The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
+Yet, within this study there was no comparison made between different high fidelity DNA polymerases.
+Due to this gap, it is natural to extend this line of inquiry and test if biases in the number of OTUs and chimeras also are dependent on the type of high fidelity DNA polymerase.
 This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
-We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.
+We will accomplish this by examining the number of OTUs, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases
 

### 2018-05-25 submission/manuscript.tex
make changes inspired by Marcy to intro

@@
 how the number of cycles and high fidelity (HiFi) DNA polymerase
-introduce error into this workflow because of less investigation into
-this area. By varying both cycle number and polymerase we assess the
-changes that occur to the bacterial community in human fecal and mock
-community samples.
+introduce error by varying cycle number and polymerase used to amplify
+16S rRNA genes in human fecal and mock community samples.
 
@@
 HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
-mock community samples the variations in the number of OTUs detected by
-DNA polymerase was observable as early as 20 cycles (P-value = 0.002).
-There also was a large range in the number of OTUs by HiFi DNA
-polymerase at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
+mock community samples the variation in the number of OTUs detected by
+the polymerases was observable as early as 20 cycles (P-value = 0.002).
+There also was a large range in the number of OTUs amplified by the
+polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
 Chimera prevalence in mock communities varied by polymerase with
@@
 chimera prevalence and the number of OTUs with Platinum having the
-highest (R\^{}2 = 0.974) and Kappa having the worse (R\^{}2 = 0.478).
+highest (R\^{}2 = 0.974) and Kappa having the worst (R\^{}2 = 0.478).
 
@@
 
-Accounting for bias is critical in order to reach an understanding of
-bacterial community changes using 16S rRNA gene sequencing results. Bias
-can change the observed results in a way that is reproducible and
-standardized. Within this context, our 16S rRNA gene sequencing methods
-are biased even when these workflows are standardized to increase
-reproducibility. In order to interpret specific studies within the
-broader context of the overall field, assessing bias at different parts
-of the 16S rRNA gene sequencing workflow is critical.
+The bacterial community is reported to vary between case and control for
+a number of diseases {[}{]}. However, for diseases like obesity, the
+taxa identified have varied widely depending on the study {[}{]}. Some
+of this variation could be due to error introduced during the 16S rRNA
+gene sequencing workflow. Yet, standardizing a 16S rRNA gene sequencing
+workflow will ultimately result in a standardized and reproducible bias
+due to choices made on the methods used for preservation, extraction,
+PCR, and sequencing. Within this context, all 16S rRNA gene sequencing
+methods are biased even when these workflows are standardized to
+increase reproducibility. In order to interpret specific studies within
+the broader context of the overall field, assessing error at different
+parts of the 16S rRNA gene sequencing workflow is critical.
 
-Many parts of the 16S rRNA gene sequencing workflow contribute bias to
-the results and are studied extensively. A typical 16S rRNA gene
-sequencing workflow can be divided into preservation, extraction, PCR,
-and sequencing steps. Overall, biases due to either preservation or
-extraction tend to be smaller than the overall biological signal being
-measured (Song et al., 2016; Bassis et al., 2017). Although the observed
-biological signal is larger, the contribution of PCR bias to this
-overall workflow is not well characterized since these studies use the
-same PCR approach while varying preservation or extraction method.
+A typical 16S rRNA gene sequencing workflow can be divided into
+preservation, extraction, PCR, and sequencing steps. The preservation
+and extraction stages of the 16S rRNA gene sequencing workflow have been
+the most extensively studied {[}{]}. For preservation and extraction
+stages of the workflow, it has been consistently found that there are
+biases based on the kits used, but that these differences are smaller
+than the overall biological difference measured between samples with
+different kits (Song et al., 2016; Bassis et al., 2017). Since these
+studies use the same PCR approach while varying preservation or
+extraction method, the contribution of PCR bias to this overall workflow
+is not well characterized.
 
-Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
-important because a large body of literature shows that there are a
-variety of steps during PCR that can change the observed results (Eckert
-\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
-made worse as the number of cycles increases (Wang \& Wang, 1996; Haas
-et al., 2011; Kebschull \& Zador, 2015). For example, the selective
-amplification of AT-rich over GC-rich sequences can exaggerate the
-difference between 16S rRNA genes higher in AT versus those higher in GC
-(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
-amplification (e.g.~incorrect amplicon size products) can also increase
-as the number of cycles increases which can drastically change commonly
-used diversity measures (Acinas et al., 2005; Santos et al., 2016).
-Additionally, chimeras can form from an aborted extension step followed
-by a subsequent priming error and secondary extension and will
-artificially increase community diversity (Haas et al., 2011). Although
-these differences are not necessarily dependent on primer and DNA
-polymerase used, there are also biases that are.
-
-The intrinsic properties to primers and DNA polymerases chosen can also
-introduce bias. Primers have variable region dependent binding
-affinities for different bacteria and depending on the primer pair do
+There is a large body of literature that shows there are different
+biases within each of the steps in the PCR stage of 16S rRNA gene
+sequencing. One source of error is from primers chosen for the 16S rRNA
+gene sequencing. There are variable region dependent binding affinities
+for different bacteria, and depending on the primer pair used they do
 not detect specific bacteria (e.g.~V1-V3 does not detect
@@
 \emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
-al., 2016). Additionally, there are multiple families of DNA polymerases
-that have their own error rate and proof reading capacity (Ishino \&
-Ishino, 2014). Interestingly, the influence that these different DNA
-polymerases can have on the observed 16S rRNA gene sequencing results
-have not been well studied like some of the other sources of PCR-based
-bias.
+al., 2016). Another source of error is the selective amplification of
+AT-rich over GC-rich sequences which exaggerate the difference between
+16S rRNA genes higher in AT versus those higher in GC content (Polz \&
+Cavanaugh, 1998). Many of these sources of biases are made worse as the
+number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
+Kebschull \& Zador, 2015). For example, both amplification error and
+non-specific amplification (e.g.~incorrect amplicon size products) also
+can increase as the number of cycles increases. This will increase the
+number of Operational Taxonomic Units (OTUs) observed and drastically
+change the values obtained from commonly used diversity measures (Acinas
+et al., 2005; Santos et al., 2016). Additionally, as the number of
+cycles increases more chimeras can form from an aborted extension step
+that causes a priming error and subsequent secondary extension (Haas et
+al., 2011). These chimeras will artificially increase community
+diversity by increasing the number of OTUs that are observed (Haas et
+al., 2011). In addition to these sources of errors, there also are
+multiple families of DNA polymerases that have their own error rate and
+proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
+influence that these different DNA polymerases can have on the observed
+16S rRNA gene sequencing results have not been well studied like some of
+the other sources of PCR-based bias.
 
-A recent study found clear differences between normal and high fidelity
-DNA polymerase and that optimization of the PCR protocol could reduce
-error and chimera generation (Gohl et al., 2016). This study also found
-that regardless of DNA polymerase, the number of Operational Taxonomic
-Units (OTUs) or taxa generated were not easily reduced using the authors
-chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
-extend this line of inquiry and ask if biases in the number of OTUs and
-chimeras are also dependent on the type of high fidelity DNA polymerase.
-There is some reason to think that this may be the case since many of
-these high fidelity DNA polymerases come from different families (e.g.
-\emph{Taq} belongs to the family A polymerases) and may intrinsically
-have different error rates that cannot be completely removed with
-modifications (Ishino \& Ishino, 2014).
-
-Although bias introduced due to differences between DNA polymerase and
-high fidelity DNA polymerase has been investigated for 16S rRNA gene
-sequencing, the bias caused due to differences between specific high
-fidelity DNA polymerases has not been. This study will investigate how
-high fidelity DNA polymerases can bias observed bacterial community
-results derived from 16S rRNA gene sequencing. We will accomplish this
-by examining if any of five different types of high fidelity DNA
-polymerases introduce significant biases into 16S rRNA gene surveys, if
-this is a cycle dependent phenomenon, and whether they can be removed
-using a standard bioinformatic pipeline.
+A recent study found differences in the number of OTUS and chimeras
+between normal and high fidelity DNA polymerases (Gohl et al., 2016).
+The authors could reduce the difference between the two polymerases by
+optimizing the annealing and extension steps within the PCR protocol
+(Gohl et al., 2016). Yet, within this study there was no comparison made
+between different high fidelity DNA polymerases. Due to this gap, it is
+natural to extend this line of inquiry and test if biases in the number
+of OTUs and chimeras also are dependent on the type of high fidelity DNA
+polymerase. This study will investigate how high fidelity DNA
+polymerases can bias observed bacterial community results derived from
+16S rRNA gene sequencing. We will accomplish this by examining the
+number of OTUs, error rate, number of sequences with an error, and
+chimera prevalence at varying number of cycles in five different high
+fidelity DNA polymerases
 
@@
 
-\hypertarget{ref-Burkardt2000}{}
-Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
-\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
-\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.
-
 \hypertarget{ref-rdp_Cole_2013}{}
@@
 \href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
-
-\hypertarget{ref-Eckert1991}{}
-Eckert KA., Kunkel TA. 1991. DNA polymerase fidelity and the polymerase
-chain reaction. \emph{Genome Research} 1:17--24. DOI:
-\href{https://doi.org/10.1101/gr.1.1.17}{10.1101/gr.1.1.17}.
 

### 2018-05-25 submission/manuscript.Rmd
add references

@@
 
-The bacterial community is reported to vary between case and control for a number of diseases [].
-However, for diseases like obesity, the taxa identified have varied widely depending on the study [].
+The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017].
+However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012].
 Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow.
@@
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [].
+The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017].
 For preservation and extraction stages of the workflow, it has been consistently found that there are biases based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].

### 2018-05-25 submission/manuscript.tex
add references

@@
 The bacterial community is reported to vary between case and control for
-a number of diseases {[}{]}. However, for diseases like obesity, the
-taxa identified have varied widely depending on the study {[}{]}. Some
-of this variation could be due to error introduced during the 16S rRNA
-gene sequencing workflow. Yet, standardizing a 16S rRNA gene sequencing
-workflow will ultimately result in a standardized and reproducible bias
-due to choices made on the methods used for preservation, extraction,
-PCR, and sequencing. Within this context, all 16S rRNA gene sequencing
-methods are biased even when these workflows are standardized to
-increase reproducibility. In order to interpret specific studies within
-the broader context of the overall field, assessing error at different
-parts of the 16S rRNA gene sequencing workflow is critical.
+a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
+et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
+the taxa identified have varied widely depending on the study (Turnbaugh
+et al., 2008; Zupancic et al., 2012). Some of this variation could be
+due to error introduced during the 16S rRNA gene sequencing workflow.
+Yet, standardizing a 16S rRNA gene sequencing workflow will ultimately
+result in a standardized and reproducible bias due to choices made on
+the methods used for preservation, extraction, PCR, and sequencing.
+Within this context, all 16S rRNA gene sequencing methods are biased
+even when these workflows are standardized to increase reproducibility.
+In order to interpret specific studies within the broader context of the
+overall field, assessing error at different parts of the 16S rRNA gene
+sequencing workflow is critical.
 
@@
 and extraction stages of the 16S rRNA gene sequencing workflow have been
-the most extensively studied {[}{]}. For preservation and extraction
+the most extensively studied (Salter et al., 2014; Song et al., 2016;
+Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
 stages of the workflow, it has been consistently found that there are
@@
 
+\hypertarget{ref-Baxter2016}{}
+Baxter NT., Ruffin MT., Rogers MAM., Schloss PD. 2016. Microbiota-based
+model improves the sensitivity of fecal immunochemical test for
+detecting colonic lesions. \emph{Genome Medicine} 8. DOI:
+\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
 \hypertarget{ref-benjamini_controlling_1995}{}
@@
 the Royal Statistical Society. Series B (Methodological)} 57:289--300.
+
+\hypertarget{ref-Bonfili2017}{}
+Bonfili L., Cecarini V., Berardi S., Scarpona S., Suchodolski JS.,
+Nasuti C., Fiorini D., Boarelli MC., Rossi G., Eleuteri AM. 2017.
+Microbiota modulation counteracts alzheimer's disease progression
+influencing neuronal proteolysis and gut hormones plasma levels.
+\emph{Scientific Reports} 7. DOI:
+\href{https://doi.org/10.1038/s41598-017-02587-2}{10.1038/s41598-017-02587-2}.
 
@@
 \href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
+
+\hypertarget{ref-review_Kim_2017}{}
+Kim D., Hofstaedter CE., Zhao C., Mattei L., Tanes C., Clarke E., Lauder
+A., Sherrill-Mix S., Chehoud C., Kelsen J., Conrad M., Collman RG.,
+Baldassano R., Bushman FD., Bittinger K. 2017. Optimizing methods and
+dodging pitfalls in microbiome research. \emph{Microbiome} 5. DOI:
+\href{https://doi.org/10.1186/s40168-017-0267-5}{10.1186/s40168-017-0267-5}.
 
@@
 
+\hypertarget{ref-Turnbaugh2008}{}
+Turnbaugh PJ., Hamady M., Yatsunenko T., Cantarel BL., Duncan A., Ley
+RE., Sogin ML., Jones WJ., Roe BA., Affourtit JP., Egholm M., Henrissat
+B., Heath AC., Knight R., Gordon JI. 2008. A core gut microbiome in
+obese and lean twins. \emph{Nature} 457:480--484. DOI:
+\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.
+
 \hypertarget{ref-Wang1996}{}
@@
 packages}.
+
+\hypertarget{ref-Zupancic2012}{}
+Zupancic ML., Cantarel BL., Liu Z., Drabek EF., Ryan KA., Cirimotich S.,
+Jones C., Knight R., Walters WA., Knights D., Mongodin EF., Horenstein
+RB., Mitchell BD., Steinle N., Snitker S., Shuldiner AR., Fraser CM.
+2012. Analysis of the gut microbiota in the old order amish and its
+relation to the metabolic syndrome. \emph{PLoS ONE} 7:e43052. DOI:
+\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
 

### 2018-05-29 submission/manuscript.Rmd
make intro and method changes

@@
 
-There is a large body of literature that shows there are different biases within each of the steps in the PCR stage of 16S rRNA gene sequencing.
-One source of error is from primers chosen for the 16S rRNA gene sequencing.
-There are variable region dependent binding affinities for different bacteria, and depending on the primer pair used they do not detect specific bacteria (e.g.
+There is a large body of literature that shows there are biases due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000].
+Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g.
 V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
@@
 ***Human and Mock Samples:***
-A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
+Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
 The processing and storage of these samples were previously published [@erin_seekatz_2016].
@@
 The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-The polymerases activation time was 2 minutes, unless a different activation was specified.
-For Kappa and Q5, a previously published protocol was used [@taq_Gohl_2016].
-For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.
+The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer.
+The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
+For Kappa and Q5, the annealing and extension time were from a previously published protocol [@taq_Gohl_2016].
+For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
 
-The cycle conditions for both fecal and mock samples started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing.
+The number of cycles in the PCR for fecal and mock samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
 The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
@@
 Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
 
 ***Analysis Workflow:***
-All samples were rarefied to 1000 sequence.
+To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
 The total number of OTUs was analyzed for both the fecal and mock community samples.

### 2018-05-29 submission/manuscript.tex
make intro and method changes

@@
 
-There is a large body of literature that shows there are different
-biases within each of the steps in the PCR stage of 16S rRNA gene
-sequencing. One source of error is from primers chosen for the 16S rRNA
-gene sequencing. There are variable region dependent binding affinities
-for different bacteria, and depending on the primer pair used they do
-not detect specific bacteria (e.g.~V1-V3 does not detect
+There is a large body of literature that shows there are biases due to
+primer and number of cycles chosen for the PCR stage of 16S rRNA gene
+sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
+variable region dependent binding affinities which causes an inability
+to detect specific bacteria (e.g.~V1-V3 does not detect
 \emph{Haemophilus influenzae} and V3-V5 does not detect
@@
 
-\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
-obtained from 4 individuals who were part of the Enterics Research
-Investigational Network (ERIN). The processing and storage of these
-samples were previously published (Seekatz et al., 2016). Other than
-confirmation that none of these individuals had a \emph{Clostridium
-difficle} infection, clinical data and other types of meta data were not
-utilized or accessed for this study. All samples were extracted using
-the MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
+\textbf{\emph{Human and Mock Samples:}} Fecal samples were obtained from
+4 individuals who were part of the Enterics Research Investigational
+Network (ERIN). The processing and storage of these samples were
+previously published (Seekatz et al., 2016). Other than confirmation
+that none of these individuals had a \emph{Clostridium difficle}
+infection, clinical data and other types of meta data were not utilized
+or accessed for this study. All samples were extracted using the
+MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
 (now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
@@
 HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
-(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR
-cycle conditions for Platinum and Accuprime followed a previously
-published protocol (Kozich et al., 2013)
+(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
+polymerases activation time was 2 minutes, unless a different activation
+was specified by the manufacturer. The annealing and extension time for
+Platinum and Accuprime followed a previously published protocol (Kozich
+et al., 2013)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
-The polymerases activation time was 2 minutes, unless a different
-activation was specified. For Kappa and Q5, a previously published
-protocol was used (Gohl et al., 2016). For Phusion, the company defined
-conditions were used except for extension time, where the Accuprime and
-Platinum settings were used.
+For Kappa and Q5, the annealing and extension time were from a
+previously published protocol (Gohl et al., 2016). For Phusion, the
+company defined activation and annealing times were used while the
+extension time followed the Accuprime and Platinum settings.
 
-The cycle conditions for both fecal and mock samples started at 15 and
-increased by 5 up to 35 cycles with amplicons used at each 5-step
+The number of cycles in the PCR for fecal and mock samples started at 15
+and increased by 5 up to 35 cycles, with amplicons used at each 5-step
 increase for sequencing. The PCR of fecal DNA samples consisted of all 4
@@
 2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
-was determined using the `seq.error' command on mock samples (Schloss et
-al., 2009; Cole et al., 2013; Rognes et al., 2016).
+was determined using the `seq.error' command on mock samples to compare
+back to the reference 16S sequences (Schloss et al., 2009; Cole et al.,
+2013; Rognes et al., 2016).
 
-\textbf{\emph{Analysis Workflow:}} All samples were rarefied to 1000
-sequence. The total number of OTUs was analyzed for both the fecal and
-mock community samples. For fecal samples, cycle dependent affects on
-Bray-Curtis indices were assessed for cycle group and within individual
-differences from the previous cycle (e.g.~20 versus 25, 25 versus 30).
-Based on these observations we analyzed potential reasons for these
-differences. Analysis of the mock community of each polymerase for
-sequence error rate, number of sequences with an error, base
-substitution, and numbers of chimeras before and after chimera removal
-with VSEARCH was assessed. Additionally, the correlation between the
-number of chimeras and the number of OTUs was also assessed.
+\textbf{\emph{Analysis Workflow:}} To adjust for unequal sequencing, all
+samples were rarefied to 1000 sequences for downstream analysis. The
+total number of OTUs was analyzed for both the fecal and mock community
+samples. For fecal samples, cycle dependent affects on Bray-Curtis
+indices were assessed for cycle group and within individual differences
+from the previous cycle (e.g.~20 versus 25, 25 versus 30). Based on
+these observations we analyzed potential reasons for these differences.
+Analysis of the mock community of each polymerase for sequence error
+rate, number of sequences with an error, base substitution, and numbers
+of chimeras before and after chimera removal with VSEARCH was assessed.
+Additionally, the correlation between the number of chimeras and the
+number of OTUs was also assessed.
 
@@
 
+\hypertarget{ref-Burkardt2000}{}
+Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
+\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
+\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.
+
 \hypertarget{ref-rdp_Cole_2013}{}

### 2018-05-29 submission/manuscript.Rmd
update first result section

@@
 ***The Number of OTUs are Dependent on Polymerase:***
-Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used [Figure 1].
-Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases [Figure 1].
+Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g.
+Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1].
+Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g.
+Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1].
 There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
-No differences between the polymerases was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
+Using a Tukey post-hoc test to identify which polymerase groups were different from each other at 35 cycles, no difference was found (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
 The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2].
-Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4].
-In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1].
-The majority of differences for the number of OTUs detected in the Mock community is largely due to Kappa and Platinum differences to the other polymerases across the different number of cycles [Table S2].
-Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.
+The closest the polymerases came to the total of `r pull(ref_mock_otu_num, numOtus)` OTUs created by the mock reference 16S sequences was at 25 and 30 cycles [Figure 2].
+Regardless of if fecal or mock communities were used, the same polymerases generated high and low number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S2].
+In contrast to the results obtained with fecal samples, differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1].
+Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community is largely due to Kappa and Platinum differences versus the other polymerases across the different number of cycles [Table S2].
+Based on these observations in fecal and mock communities, it is clear that using different polymerases result in a different total number of OTUs within a sample.
 
@@
 **Figure 1: Total Number of OTUs in Fecal Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
-The legend represents the color code for the 5 different HiFi DNA polymerases used.
+The points represent the median value of all four fecal samples.
+The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
 The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
-This range was also larger for specific HiFi DNA polymerases.
+This increased range also was larger for specific HiFi DNA polymerases.
 
 **Figure 2:  Total Number of OTUs in Mock Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
-The legend represents the color code for the 5 different HiFi DNA polymerases used.
-The dotted black line represents the number of OTUs detected when only the references sequences are clustered.
-The range in the number of OTUs detected in the Mock samples increased as cycle number increased.
+The points represent the median value of all four fecal samples.
+The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
+The dotted black line represents the number of OTUs detected when only the references sequences for the mock community are clustered.
+The range in the number of OTUs detected in the mock samples increased as cycle number increased.
 This range was also larger for specific HiFi DNA polymerases.

### 2018-05-29 submission/manuscript.tex
update first result section

@@
 Differences in the range of the number of OTUs detected for fecal
-samples is dependent on the polymerase used {[}Figure 1{]}.
+samples is dependent on the polymerase used (e.g.~Accuprime at 35 cycles
+= 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
 Additionally, there is a trend for lower number of cycles (15-20) to
 result in a reduced range in the number of OTUs versus higher number of
-cycles (25, 30, and 35) for all polymerases {[}Figure 1{]}. There is an
-overall difference in the number of OTUs detected within fecal samples
-between polymerases at 35 cycles (F-stat \textgreater{} 16.35, P-value =
-9.7e-05) {[}Table S1{]}. No differences between the polymerases was
-found based on a Tukey post-hoc test (P-value \textgreater{} 0.05)
-{[}Table S2{]}. The polymerase dependent difference in the range of the
-number of OTUs also was observed in the mock community samples {[}Figure
-2{]}. Regardless if fecal or mock communities were used, the same
-polymerases had high (Platinum) and low (Accuprime) number of OTUs and
-this was consistent across the number of cycles used {[}Figure 1-2 \&
-Table S1-S4{]}. In contrast to the results obtained with fecal samples,
-differences between polymerases were observed as early as 20 cycles in
-the mock community (F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. The
-majority of differences for the number of OTUs detected in the Mock
-community is largely due to Kappa and Platinum differences to the other
-polymerases across the different number of cycles {[}Table S2{]}. Based
-on these observations in fecal and mock communities, it is clear that
-different polymerases result in a different total number of OTUs within
-a sample.
+cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15 cycles =
+10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}. There is
+an overall difference in the number of OTUs detected within fecal
+samples between polymerases at 35 cycles (F-stat \textgreater{} 16.35,
+P-value = 9.7e-05) {[}Table S1{]}. Using a Tukey post-hoc test to
+identify which polymerase groups were different from each other at 35
+cycles, no difference was found (P-value \textgreater{} 0.05) {[}Table
+S2{]}. The polymerase dependent difference in the range of the number of
+OTUs also was observed in the mock community samples {[}Figure 2{]}. The
+closest the polymerases came to the total of 8 OTUs created by the mock
+reference 16S sequences was at 25 and 30 cycles {[}Figure 2{]}.
+Regardless of if fecal or mock communities were used, the same
+polymerases generated high and low number of OTUs and this was
+consistent across the number of cycles used {[}Figure 1-2 \& Table
+S1-S2{]}. In contrast to the results obtained with fecal samples,
+differences between polymerases for the number of OTUs created were
+observed as early as 20 cycles in the mock community (F-stat = 15.82,
+P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc test, the
+majority of differences for the number of OTUs detected in the mock
+community is largely due to Kappa and Platinum differences versus the
+other polymerases across the different number of cycles {[}Table S2{]}.
+Based on these observations in fecal and mock communities, it is clear
+that using different polymerases result in a different total number of
+OTUs within a sample.
 
@@
 Cycles.} The x-axis represents the different number of cycles used and
-the y-axis is the number of OTUs. The legend represents the color code
-for the 5 different HiFi DNA polymerases used. The range in the number
-of OTUs detected in the different fecal samples increased as cycle
-number increased. This range was also larger for specific HiFi DNA
-polymerases.
+the y-axis is the number of OTUs. The points represent the median value
+of all four fecal samples. The lines represent the range of the minimum
+and maximum number of OTUs detected within the four fecal samples. The
+range in the number of OTUs detected in the different fecal samples
+increased as cycle number increased. This increased range also was
+larger for specific HiFi DNA polymerases.
 
@@
 Cycles.} The x-axis represents the different number of cycles used and
-the y-axis is the number of OTUs. The legend represents the color code
-for the 5 different HiFi DNA polymerases used. The dotted black line
-represents the number of OTUs detected when only the references
-sequences are clustered. The range in the number of OTUs detected in the
-Mock samples increased as cycle number increased. This range was also
-larger for specific HiFi DNA polymerases.
+the y-axis is the number of OTUs. The points represent the median value
+of all four fecal samples. The lines represent the range of the minimum
+and maximum number of OTUs detected within the four fecal samples. The
+dotted black line represents the number of OTUs detected when only the
+references sequences for the mock community are clustered. The range in
+the number of OTUs detected in the mock samples increased as cycle
+number increased. This range was also larger for specific HiFi DNA
+polymerases.
 

### 2018-05-29 submission/manuscript.Rmd
update second section of results

@@
 
-loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr"))
+loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr", "FSA"))
 
@@
 # Vector of sub samples used
-sub_sample_level <- c("1000", "5000", "10000")
+dist_data <- read_csv("../data/process/tables/bray_5_cycle_dist_1000_data.csv") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
+  group_by(taq, cycle_compare) %>% 
+  summarise(dist_median = median(distance, na.rm = T), 
+            dist_min = min(distance, na.rm = T), 
+            dist_max = max(distance, na.rm = T)) %>% 
+  mutate(range_diff = dist_max - dist_min) %>% 
+  ungroup() %>% 
+  filter(cycle_compare != "25to35")
 
-dist_data <- sapply(sub_sample_level, 
-                    function(x) as.data.frame(read_csv(paste("../data/process/tables/bray_5_cycle_dist_",                                                x, "_data.csv", sep = ""))), simplify = F)
+fecal_kruskal_test_by_cycle_compare_summary <- dist_data %>% 
+  nest() %>% 
+  mutate(kruskal_analysis = map(data, ~kruskal.test(dist_median ~ factor(cycle_compare), data = .x)), 
+         summary_data = map(kruskal_analysis, broom::tidy)) %>% 
+  select(summary_data) %>% unnest(summary_data)
 
-mock_dist_data <- sapply(sub_sample_level, 
-                    function(x) as.data.frame(read_csv(paste("../data/process/tables/mock_bray_5_cycle_dist_", 
-                                               x, "_data.csv", sep = ""))), simplify = F)
+dunns_test_by_cycle_compare_summary <- dist_data %>% 
+  nest() %>% 
+  mutate(dunn_analysis = map(data, ~dunnTest(dist_median ~ factor(cycle_compare), data = .x, method = "bh")[["res"]])) %>% 
+  select(dunn_analysis) %>% unnest() %>% 
+  arrange(Comparison)
+
+mock_dist_data <- read_csv("../data/process/tables/mock_bray_5_cycle_dist_1000_data.csv") %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
+  filter(!is.na(distance)) %>% 
+  group_by(taq, cycle_compare) %>% 
+  summarise(dist_median = median(distance, na.rm = T), 
+            dist_min = min(distance, na.rm = T), 
+            dist_max = max(distance, na.rm = T)) %>% 
+  mutate(range_diff = dist_max - dist_min) %>% 
+  ungroup() %>% 
+  filter(cycle_compare != "25to35")
+
+mock_kruskal_test_by_cycle_compare_summary <- mock_dist_data %>% 
+  nest() %>% 
+  mutate(kruskal_analysis = map(data, ~kruskal.test(dist_median ~ factor(cycle_compare), data = .x)), 
+         summary_data = map(kruskal_analysis, broom::tidy)) %>% 
+  select(summary_data) %>% unnest(summary_data)
 
@@
 
-***The Number of OTUs are Dependent on Polymerase:***
+***The number of OTUs are dependent on polymerase:***
 Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g.
@@
 
-***Minimal Bray-Curtis Differences are Detected and are Dependent on Cycle Number:***
-A few small differences based on the number of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of polymerases, there are large differences in the community composition between 20 and 25 cycles [Figure 3A].
-Additionally, when data was available for the mock communities, there are also large observed differences between 20 and 25 cycles [Figure 3B].
-These community differences within a sample disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3].
-Although these trends are clearly noticeable, there is no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05).
-Using PERMANOVA to test for community differences based on the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
+***The bacterial community is similar within polymerase and varies by number of cycles:***
+Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3].
+Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
 For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
-Overall, these data suggest that there are small polymerase differences in Bray-Curtis index that are dependent on the number of cycles.
+There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
+the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A].
+Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curits index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
+The mock community has similar trends to the obeservations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
+Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
 
@@
 
-**Figure 1: Total Number of OTUs in Fecal Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
-The points represent the median value of all four fecal samples.
+**Figure 1: Total Number of OTUs in Fecal Samples by Number of Cycles.** The points represent the median number of OTUs of all four fecal samples.
 The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
@@
 
-**Figure 2:  Total Number of OTUs in Mock Samples by Number of Cycles.** The x-axis represents the different number of cycles used and the y-axis is the number of OTUs.
-The points represent the median value of all four fecal samples.
+**Figure 2:  Total Number of OTUs in Mock Samples by Number of Cycles.** The points represent the median number of OTUs for the mock samples.
 The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
@@
 **Figure 3: Bray-Curtis Community Differences by Five-Cycle Intervals.** A) within person differences based on the next 5-cycle PCR interval in fecal samples.
-B) Within replicate difference based on the next 5-cycle PCR interval in Mock samples.
+B) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
+The points represent the median Bray-Curtis index for the samples.
+The lines represent the range of the minimum and maximum Bray-Curits index value for each PCR 5-cycle increment comparison.
+The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.
 

### 2018-05-29 submission/manuscript.tex
update second section of results

@@
 
-\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}}
+\textbf{\emph{The number of OTUs are dependent on polymerase:}}
 Differences in the range of the number of OTUs detected for fecal
@@
 
-\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
-Dependent on Cycle Number:}} A few small differences based on the number
-of cycles were detected in overall bacterial community composition.
-Within the same fecal sample and independent of polymerases, there are
-large differences in the community composition between 20 and 25 cycles
-{[}Figure 3A{]}. Additionally, when data was available for the mock
-communities, there are also large observed differences between 20 and 25
-cycles {[}Figure 3B{]}. These community differences within a sample
-disappear when comparing 25 to 30 cycles and do not persist past 25
-cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
-there is no detectable difference in Bray-Curtis index when comparing to
-the previous 5-cycle increment for both fecal and mock communities after
-multiple comparison correction (P-value \textgreater{} 0.05). Using
-PERMANOVA to test for community differences based on the number of
-cycles within polymerases, only Phusion had cycle dependent differences
-(P-value = 0.03. For fecal samples, Phusion was one of two polymerases
-that had enough sequences to be rarefied to 1000 at 15 cycles. Overall,
-these data suggest that there are small polymerase differences in
-Bray-Curtis index that are dependent on the number of cycles.
+\textbf{\emph{The bacterial community is similar within polymerase and
+varies by number of cycles:}} Within each respective 5-cycle increment
+comparison there was no difference in Bray-Curtis index between the
+polymerases (P-value \textgreater{} 0.05) {[}Figure 3{]}. Using
+PERMANOVA to test for community differences based on any of the number
+of cycles within polymerases, only Phusion had cycle dependent
+differences (P-value = 0.03. For fecal samples, Phusion was one of two
+polymerases that had enough sequences to be rarefied to 1000 at 15
+cycles. There was an overall decrease in Bray-Curtis index by cycle
+comparison group (e.g.~the 15 cycle versus 20 cycle group compared to
+the 30 cycle to 35 cycle group) (P-value \textless{} 0.01) {[}Figure
+3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
+cycle versus 25 cycle comparison groups had a higher Bray-Curits index
+than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
+The mock community has similar trends to the obeservations reported for
+the fecal bacterial community but none were significant \textgreater{}
+0.05) {[}Figure 3B{]}. Overall, these data suggest that the number of
+cycles can change the bacterial community independent of the polymerase
+used to generate the sequences.
 
@@
 \textbf{Figure 1: Total Number of OTUs in Fecal Samples by Number of
-Cycles.} The x-axis represents the different number of cycles used and
-the y-axis is the number of OTUs. The points represent the median value
-of all four fecal samples. The lines represent the range of the minimum
-and maximum number of OTUs detected within the four fecal samples. The
-range in the number of OTUs detected in the different fecal samples
-increased as cycle number increased. This increased range also was
-larger for specific HiFi DNA polymerases.
+Cycles.} The points represent the median number of OTUs of all four
+fecal samples. The lines represent the range of the minimum and maximum
+number of OTUs detected within the four fecal samples. The range in the
+number of OTUs detected in the different fecal samples increased as
+cycle number increased. This increased range also was larger for
+specific HiFi DNA polymerases.
 
 \textbf{Figure 2: Total Number of OTUs in Mock Samples by Number of
-Cycles.} The x-axis represents the different number of cycles used and
-the y-axis is the number of OTUs. The points represent the median value
-of all four fecal samples. The lines represent the range of the minimum
-and maximum number of OTUs detected within the four fecal samples. The
-dotted black line represents the number of OTUs detected when only the
-references sequences for the mock community are clustered. The range in
-the number of OTUs detected in the mock samples increased as cycle
-number increased. This range was also larger for specific HiFi DNA
-polymerases.
+Cycles.} The points represent the median number of OTUs for the mock
+samples. The lines represent the range of the minimum and maximum number
+of OTUs detected within the four fecal samples. The dotted black line
+represents the number of OTUs detected when only the references
+sequences for the mock community are clustered. The range in the number
+of OTUs detected in the mock samples increased as cycle number
+increased. This range was also larger for specific HiFi DNA polymerases.
 
@@
 interval in fecal samples. B) Within replicate difference based on the
-next 5-cycle PCR interval in Mock samples.
+next 5-cycle PCR interval in mock samples. The points represent the
+median Bray-Curtis index for the samples. The lines represent the range
+of the minimum and maximum Bray-Curits index value for each PCR 5-cycle
+increment comparison. The closer a sample is to a Bray-Curtis index of
+1.00 the more dissimilar the bacterial community is of the two compared
+number of cycles.
 

### 2018-05-29 submission/manuscript.Rmd
redo third section of results

@@
 
-***The number of OTUs are dependent on polymerase:***
+***The number of OTUs generated are dependent on polymerase used:***
 Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g.
@@
 
-***Sequence Error is Dependent on both Polymerase and Cycle Number:***
-Differences in the median error rate varied by polymerase, with Kappa being the highest [Figure 4 & Table S3].
-The differences in the median error rate between the polymerases also is dependent on the number of cycles, with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S4].
-The total sequences with at least one error also is dependent on the number of cycles [Table S3 & S4].
-These differences in sequences with at least one error is mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S3 & S4].
-The differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
-These results suggest that sequence error is dependent on polymerase and the number of cycles, but is not attributable to a specific type.
-Based on these results, we examined whether chimeras also were dependent on polymerase and whether this also could affect the number of OTUs.
+***Sequence error varies by polymerase and is consistent across the number of cycles used:***
+The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].
+The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 and Table S4].
+The total sequences with at least one error is also polymerase dependent with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
+These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
+Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base subsitution.
 
-***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
+***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
 There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].

### 2018-05-29 submission/manuscript.tex
redo third section of results

@@
 
-\textbf{\emph{The number of OTUs are dependent on polymerase:}}
-Differences in the range of the number of OTUs detected for fecal
-samples is dependent on the polymerase used (e.g.~Accuprime at 35 cycles
-= 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
-Additionally, there is a trend for lower number of cycles (15-20) to
-result in a reduced range in the number of OTUs versus higher number of
-cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15 cycles =
-10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}. There is
-an overall difference in the number of OTUs detected within fecal
-samples between polymerases at 35 cycles (F-stat \textgreater{} 16.35,
-P-value = 9.7e-05) {[}Table S1{]}. Using a Tukey post-hoc test to
+\textbf{\emph{The number of OTUs generated are dependent on polymerase
+used:}} Differences in the range of the number of OTUs detected for
+fecal samples is dependent on the polymerase used (e.g.~Accuprime at 35
+cycles = 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure
+1{]}. Additionally, there is a trend for lower number of cycles (15-20)
+to result in a reduced range in the number of OTUs versus higher number
+of cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15
+cycles = 10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
+There is an overall difference in the number of OTUs detected within
+fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
+16.35, P-value = 9.7e-05) {[}Table S1{]}. Using a Tukey post-hoc test to
 identify which polymerase groups were different from each other at 35
@@
 
-\textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
-Number:}} Differences in the median error rate varied by polymerase,
-with Kappa being the highest {[}Figure 4 \& Table S3{]}. The differences
-in the median error rate between the polymerases also is dependent on
-the number of cycles, with Platinum and Kappa having the most
-differences versus other polymerases {[}Figure 4 and Table S4{]}. The
-total sequences with at least one error also is dependent on the number
-of cycles {[}Table S3 \& S4{]}. These differences in sequences with at
-least one error is mostly due to differences in
-Accuprime\textsuperscript{TM} and Kappa versus the other polymerases
-{[}Table S3 \& S4{]}. The differences in error rates were not due to
-polymerase dependent differences in base substitution rate {[}Figure
-S1{]}. These results suggest that sequence error is dependent on
-polymerase and the number of cycles, but is not attributable to a
-specific type. Based on these results, we examined whether chimeras also
-were dependent on polymerase and whether this also could affect the
-number of OTUs.
+\textbf{\emph{Sequence error varies by polymerase and is consistent
+across the number of cycles used:}} The median error rate varied by
+polymerase, with Kappa having the highest error rate of all the
+polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
+majority of the differences across the number of cycles was between
+Kappa and the other polymerases {[}Figure 4 and Table S4{]}. The total
+sequences with at least one error is also polymerase dependent with the
+majority of differences being between Kappa or Accuprime and the other
+polymerases {[}Table S3 \& S4{]}. These differences in error rates were
+not due to polymerase dependent differences in base substitution rate
+{[}Figure S1{]}. Collectively, the results suggest that sequence error
+is dependent on polymerase, persists across the number of cycles used,
+and are not due to any bias towards a specific base subsitution.
 
 \textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
-and Correlate with the Number of OTUs:}} There is significant
+and Correlate with the Number of OTUs:}} Based on the previous results,
+we examined whether chimeras also were dependent on polymerase and
+whether this could affect the number of OTUs. There is significant
 differences in the chimera prevalence based on polymerase at all the

### 2018-05-29 submission/manuscript.Rmd
redo final section of results

@@
 
-***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
-There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
+***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
+We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
 Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S4].
 Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
+The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 5A & 5B].
 Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
@@
 This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6].
-This data suggests that chimera prevalence depends on polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
+This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 

### 2018-05-29 submission/manuscript.tex
redo final section of results

@@
 
-\textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
-and Correlate with the Number of OTUs:}} Based on the previous results,
+\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
+and correlate with the number of OTUs:}} Based on the previous results,
 we examined whether chimeras also were dependent on polymerase and
-whether this could affect the number of OTUs. There is significant
-differences in the chimera prevalence based on polymerase at all the
+whether this could affect the number of OTUs. We observed significant
+differences in the chimera prevalence based on polymerase across the
 number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
@@
 of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 5A \& 5B{]}. Additionally, there was a plateau in the
-total percent of chimeras that were removed that was similar for all
-polymerases {[}Figure 5C{]}. A positive correlation was observed between
-chimeric sequences and the number of OTUs for all polymerases {[}Figure
-6{]}. This positive correlation was strongest for
+was used {[}Figure 5A \& 5B{]}. The number of cycles used clearly
+increased chimera prevalence for all polymerases but the rate of
+increase differed {[}Figure 5A \& 5B{]}. Additionally, there was a
+plateau in the total percent of chimeras that were removed that was
+similar for all polymerases {[}Figure 5C{]}. A positive correlation was
+observed between chimeric sequences and the number of OTUs for all
+polymerases {[}Figure 6{]}. This positive correlation was strongest for
 Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
-This data suggests that chimera prevalence depends on polymerase used
-and confirms that the number of OTUs is dependent on the prevalence of
-these chimeric sequences.
+This data suggests that chimera prevalence depends on polymerase, is
+made worse by increasing the number of cycles, and confirms that the
+number of OTUs is dependent on the prevalence of these chimeric
+sequences.
 

### 2018-05-29 code/make_bray_distance_graphs.R
add kappa to figure

@@ -36,60 +36,44 @@ read_data <- function(pathing, start_name, end_name, differentiator){
 ###########################################################################################################################
 
 # Vector of sub samples used
-sub_sample_level <- c("1000", "5000", "10000")
-
-dist_data <- sapply(sub_sample_level, 
-                    function(x) read_csv(paste("data/process/tables/bray_5_cycle_dist_", 
-                                               x, "_data.csv", sep = "")) %>% 
-                      mutate(depth_level = as.numeric(x)), simplify = F) %>% bind_rows() %>% 
-  filter(taq != "K") %>% 
+dist_data <- read_csv("data/process/tables/bray_5_cycle_dist_1000_data.csv") %>% 
+  filter(cycle_compare != "25to35") %>% 
   mutate(taq = factor(taq, 
-                      levels = c("ACC", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Phusion", "Platinum", "Q5")), 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")),
          cycle_compare = factor(cycle_compare, 
                                 levels = c("15to20", "20to25", "25to30", "30to35"), 
                                 labels = c("15x vs 20x", "20x vs 25x", 
                                            "25x vs 30x", "30x vs 35x"))) %>% 
-  group_by(taq, depth_level, cycle_compare) %>% 
+  group_by(taq, cycle_compare) %>% 
   summarise(mean_dist = median(distance, na.rm = T), 
             max_dist = max(distance, na.rm = T), 
             min_dist = min(distance, na.rm = T))
 
 
-
-mock_dist_data <- sapply(sub_sample_level, 
-                    function(x) read_csv(paste("data/process/tables/mock_bray_5_cycle_dist_", 
-                                               x, "_data.csv", sep = "")) %>% 
-                      mutate(depth_level = as.numeric(x)), simplify = F) %>% bind_rows() %>% 
-  filter(taq != "K") %>% 
+mock_dist_data <- read_csv("data/process/tables/mock_bray_5_cycle_dist_1000_data.csv") %>%  
[…]

### 2018-05-30 submission/manuscript.Rmd
update first discussion paragraph

@@
 
-In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-5].
-These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g.
-richness).
-Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.
-We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences.
-This suggests that some of the specific diversity differences between studies can be attributed to differences in polyermase used.
-Based on these observations, metrics that measure within sample diversity like richness depend on polymerase but this may not be the case for metrics that assess between sample diversity.
+Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-5].
+Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences.
+In addition to the reported polymerase dependent differences, the bacterial community also varied by the number of cycles used.
+These differences are important because many diversity metrics rely on calculations (e.g.
+richness) that are dependent on the number of OTUs.
+Since the number of OTUs are dependent on error rate and chimera prevalence the diversity metrics can vary simply by changing the polymerase used.
+Based on these observations, metrics that measure diversity using calculations such as richness depend on the polymerase, but this may not be the case for all diversity metrics.
 

### 2018-05-30 submission/manuscript.tex
update first discussion paragraph

@@
 
-In this study we show that the number of OTUs, error rate, and chimera
-prevalence depends on polymerase used {[}Figure 1-5{]}. These
-differences are important because many diversity metrics rely on the
-number of OTUs or other measures dependent on error rate and chimera
-prevalence as part of their metric calculations (e.g.~richness).
-Additionally, the earlier detection of differences in total number of
-OTUs between polymerases in the mock versus fecal samples might indicate
-that high biomass samples may underestimate the biases present within
-low biomass samples. We observed that undetected chimeras that were not
-identified and removed using standard bioinformatic approaches cause
-many of these differences. This suggests that some of the specific
-diversity differences between studies can be attributed to differences
-in polyermase used. Based on these observations, metrics that measure
-within sample diversity like richness depend on polymerase but this may
-not be the case for metrics that assess between sample diversity.
+Our observations support that the number of OTUs, error rate, and
+chimera prevalence depends on polymerase used and that chimera
+prevalence is highly correlated with the number of OTUs {[}Figure
+1-5{]}. Our observations also suggest that undetected chimeras, not
+removed using standard bioinformatic approaches, cause many of these
+differences. In addition to the reported polymerase dependent
+differences, the bacterial community also varied by the number of cycles
+used. These differences are important because many diversity metrics
+rely on calculations (e.g.~richness) that are dependent on the number of
+OTUs. Since the number of OTUs are dependent on error rate and chimera
+prevalence the diversity metrics can vary simply by changing the
+polymerase used. Based on these observations, metrics that measure
+diversity using calculations such as richness depend on the polymerase,
+but this may not be the case for all diversity metrics.
 

### 2018-05-30 submission/manuscript.Rmd
complete changes to the discussion

@@
 
-Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-5].
+Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-2 & 4-6].
 Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences.
@@
 
-There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index.
-Our observations generally found no differences in the overall bacterial community composition for the number of cycles used.
-One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group.
-Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence.
+Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community.
+Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 3].
+However, the distance between cycle comparison groups (e.g.
+20 cycles versus 25 cycles group compared to 30 cycles versus 35 cycles group) was different across polymerase used.
+These results support the idea that cycle number and not the specific polymerase used may have a larger affect on metrics such as Bray-Curtis.
+One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987].
+The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence.
 Metrics that solely use presence/absence of OTUs (e.g.
-Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in polymerases.
+Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence.
 When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
-With differences in the number of OTUs and chimera prevalence depending on polymerase used, it might be easier to avoid specific DNA polymerase families altogether.
+One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.
 
-Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, this is unlikely to be the only contributor.
-For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
-Additionally, based on the material safety data sheet the differences between the two polymerases are not immediately apparent.
-Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the polymerases.
-We are unlikely to avoid adding polymerase dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.
+Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family because of the different binding affinities and error correction capacity [@polymerase_Ishino_2014], this is unlikely to be the only contributor.
+Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
+Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent.
+Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence.
+Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.
 
-The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence and may be masked by biological differences.
-The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014].
-The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations.
-Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.
-
-Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations.
-Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016].
-This study shows that specific diversity metrics used to measure the microbial community consistently vary based on polymerase.
-Standardizing multiple workflows to one specific polymerase could be detrimental since some polymerases may work better in certain situations over others.
-Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach.
-All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled.
-Although a diversity of approaches may make reproducibility and replicability more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.
+Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples.
+However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community.
+Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis.
+All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered.
+Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occuring.
 

### 2018-05-30 submission/manuscript.tex
complete changes to the discussion

@@
 chimera prevalence depends on polymerase used and that chimera
-prevalence is highly correlated with the number of OTUs {[}Figure
-1-5{]}. Our observations also suggest that undetected chimeras, not
+prevalence is highly correlated with the number of OTUs {[}Figure 1-2 \&
+4-6{]}. Our observations also suggest that undetected chimeras, not
 removed using standard bioinformatic approaches, cause many of these
@@
 
-There were few differences that depend on polymerase for between sample
-diversity, as measured by the Bray-Curtis index. Our observations
-generally found no differences in the overall bacterial community
-composition for the number of cycles used. One possible reason for this
-outcome is that our study did not have enough power to detect
-differences due to low sample number in each group. Another reason is
-that many of the OTUs are likely not highly abundant, allowing the
-Bray-Curtis index to be able to successfully down-weight chimeric OTUs
-(Minchin, 1987). The choice of downstream diversity metric could be an
-important consideration in helping to mitigate these observed polymerase
-dependent differences in chimera prevalence. Metrics that solely use
-presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
-less robust to chimera prevalence and by extension total number of OTU
-differences in polymerases. When choosing a distance metric, careful
+Based on polymerase used, our observations generally found no difference
+when using the Bray-Curtis index to measure the bacterial community.
+Specifically, there was no difference in distance between successive
+5-cycle increments within samples between the polymerases when using the
+Bray-Curtis index {[}Figure 3{]}. However, the distance between cycle
+comparison groups (e.g.~20 cycles versus 25 cycles group compared to 30
+cycles versus 35 cycles group) was different across polymerase used.
+These results support the idea that cycle number and not the specific
+polymerase used may have a larger affect on metrics such as Bray-Curtis.
+One possible reason for this outcome is that many of the OTUs generated
+by polymerase dependent error rates and chimera prevalence are likely
+not highly abundant, allowing the Bray-Curtis index to be able to
+successfully down-weight these OTUs (Minchin, 1987). The choice of
+downstream diversity metric could be an important consideration in
+helping to mitigate some of the observed polymerase dependent
+differences in error rate and chimera prevalence. Metrics that solely
+use presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996),
+richness) may be less robust to polymerase dependent error rates and
+chimera prevalence. When choosing a distance metric, careful
 consideration of the biases introduced from the PCR step of the 16S rRNA
-gene sequencing workflow need to be taken into account. With differences
-in the number of OTUs and chimera prevalence depending on polymerase
-used, it might be easier to avoid specific DNA polymerase families
-altogether.
+gene sequencing workflow need to be taken into account. One possible way
+to better choose polymerases may be based on the DNA polymerase family
+used in the PCR mixture.
 
 Although the variation in error rate and chimera prevalence may be due
-to the DNA polymerase family, this is unlikely to be the only
-contributor. For example, the highest and lowest chimera rates both
-belonged to a family A polymerase (Platinum and
+to the DNA polymerase family because of the different binding affinities
+and error correction capacity (Ishino \& Ishino, 2014), this is unlikely
+to be the only contributor. Within our study the highest and lowest
+chimera prevalence both belonged to a family A polymerase (Platinum and
 Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
-Additionally, based on the material safety data sheet the differences
-between the two polymerases are not immediately apparent. Both
-polymerases contain a recombinant \emph{Taq} DNA polymerase, a
-\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
-antibody. With everything else being equal, it is possible that
-differences in how the recombinant \emph{Taq} was generated could be a
-contributing factor for the differences observed between the
-polymerases. We are unlikely to avoid adding polymerase dependent bias
-to 16S rRNA gene sequencing results, however, these differences may not
-be large enough to mask the actual biological signal.
+Additionally, based on the information supplied by the respective
+manufacturers, the differences between the two PCR mixtures are not
+immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
+DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
+\emph{Taq} antibody. Since it is not possible to know everything about
+the mixture beyond what was willingly provided by the manufacturer, it
+is possible that differences in how the recombinant \emph{Taq} was
+generated or other compounds within the PCR mixture could be a
+contributing factor for the differences in error rate and chimera
+prevalence. Beyond the choice of the type of polymerase, there may be
+other ways to reduce the affect of polymerase dependent error rates and
+chimera prevalence on the downstream results.
 
-The majority of polymerases we studied add small increases in the number
-of OTUs and chimera prevalence and may be masked by biological
-differences. The sequence error introduced by the polymerase is also
-small and likely to be smaller than the biological variation within a
-specific study, which would be consistent with previous findings for
-preservation and DNA extraction methods (Salter et al., 2014; Song et
-al., 2016; Luo et al., 2016). The choice of polymerase should be an
-important consideration in the creation of a 16S rRNA gene sequencing
-workflow because using different polymerases can add error and bias to
-the downstream observations. Although standardization of the workflow
-may partially solve this problem by introducing a consistent bias to all
-samples, it does come at a cost.
-
-Methods can be standardized but they commonly contain bias that is
-reproducible and may miss important associations. Bias can be easily
-reproduced and can be found in every step of the 16S rRNA gene
-sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
-2016; Amir et al., 2017). This study shows that specific diversity
-metrics used to measure the microbial community consistently vary based
-on polymerase. Standardizing multiple workflows to one specific
-polymerase could be detrimental since some polymerases may work better
-in certain situations over others. Arguably, the degree of workflow
-standardization across studies and research group needs to be approached
-on a study by study basis and not every project needs to use the exact
-same approach. All aspects of the 16S rRNA gene sequencing workflow need
-to be customized for the specific microbial community that will be
-sampled. Although a diversity of approaches may make reproducibility and
-replicability more difficult it will help to avoid systematic biases
-from occuring due to widespread standardization of approaches.
+Standardization of the 16S rRNA gene sequencing workflow may partially
+solve this problem by introducing a consistent bias to all samples.
+However, as mentioned in the introduction, many of the choices made to
+standardize the workflow will result in missing important members of the
+microbial community. Therefore, the degree of workflow standardization
+across studies and research groups need to be approached on a study by
+study basis. All aspects of the 16S rRNA gene sequencing workflow should
+be customized for the specific scientific questions that wants to be
+answered. Although a diversity of approaches may make reproducibility
+and replicability more difficult it will help reduce systematic biases
+from occuring.
 
@@
 \href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
-
-\hypertarget{ref-Amir2017}{}
-Amir A., McDonald D., Navas-Molina JA., Debelius J., Morton JT., Hyde
-E., Robbins-Pianka A., Knight R. 2017. Correcting for microbial blooms
-in fecal samples during room-temperature shipping. \emph{mSystems}
-2:e00199--16. DOI:
-\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.
 
@@
 DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
-
-\hypertarget{ref-preservation_Luo_2016}{}
-Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
-E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
-2016. Effects of specimen collection methodologies and storage
-conditions on the short-term stability of oral microbiome taxonomy.
-\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
-\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
 

### 2018-05-30 submission/manuscript.Rmd
update conclusion and acknowledgements

@@
 **Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+Additionally, the median number of OTUs varied by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
 When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
@@
 the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A].
-Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curits index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
-The mock community has similar trends to the obeservations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
+Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
+The mock community has similar trends to the observations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
 Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
@@
 These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
-Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base subsitution.
+Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.
 
@@
 All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered.
-Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occuring.
+Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.
 
@@
 
-Our observations fill a gap in knowledge on the bias introduced to 16S rRNA gene sequencing results due to differences in polymerases.
-We found that the number of OTUs and the chimera prevalence is dependent on both polymerase and cycle number chosen.
-Care should be taken when choosing a polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations.
-Knowing the inherent bias associated with different polymerases allows for better interpretation of the relationsip between an individual study to their respective field of research.
+Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing results.
+Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics.
+By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.
 
@@
 We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
+Additional thanks to members of the Schloss lab and Dr.
+Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques.
 Salary support for Marc A.
@@
 The points represent the median Bray-Curtis index for the samples.
-The lines represent the range of the minimum and maximum Bray-Curits index value for each PCR 5-cycle increment comparison.
+The lines represent the range of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle increment comparison.
 The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.
@@
 
-**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Precentage of chimeric sequences without the removal of chimeras with VSEARCH.
+**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
 C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.

### 2018-05-30 submission/manuscript.tex
update conclusion and acknowledgements

@@
 polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
-Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
+Phusion = 84 - 136). Additionally, the median number of OTUs varied by
 HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
@@
 3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
-cycle versus 25 cycle comparison groups had a higher Bray-Curits index
+cycle versus 25 cycle comparison groups had a higher Bray-Curtis index
 than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
-The mock community has similar trends to the obeservations reported for
+The mock community has similar trends to the observations reported for
 the fecal bacterial community but none were significant \textgreater{}
@@
 is dependent on polymerase, persists across the number of cycles used,
-and are not due to any bias towards a specific base subsitution.
+and are not due to any bias towards a specific base substitution.
 
@@
 and replicability more difficult it will help reduce systematic biases
-from occuring.
+from occurring.
 
@@
 
-Our observations fill a gap in knowledge on the bias introduced to 16S
-rRNA gene sequencing results due to differences in polymerases. We found
-that the number of OTUs and the chimera prevalence is dependent on both
-polymerase and cycle number chosen. Care should be taken when choosing a
-polymerase for 16S rRNA gene surveys because their intrinsic differences
-can change the number of OTUs observed and influence diversity based
-metrics that do not down-weight rare observations. Knowing the inherent
-bias associated with different polymerases allows for better
-interpretation of the relationsip between an individual study to their
-respective field of research.
+Our observations show that errors dependent on either high fidelity DNA
+polymerases used or the number of cycles chosen can change diversity
+metrics used in 16S rRNA gene sequencing results. Care should be taken
+when choosing a polymerase for 16S rRNA gene surveys because they can
+influence diversity-based metrics. By knowing the inherent bias
+associated with different polymerases it allows for more informed
+choices to be made on how to reduce 16S rRNA gene sequencing-based
+error.
 
@@
 Cockburn for their effort in sequencing the samples as part of the
-Microbiome Core Facility at the University of Michigan. Salary support
-for Marc A. Sze came from the Canadian Institute of Health Research and
-NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from
-NIH grants P30DK034933 and 1R01CA215574.
+Microbiome Core Facility at the University of Michigan. Additional
+thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
+earlier drafts of the manuscript and providing helpful critiques. Salary
+support for Marc A. Sze came from the Canadian Institute of Health
+Research and NIH grant UL1TR002240. Salary support for Patrick D.
+Schloss came from NIH grants P30DK034933 and 1R01CA215574.
 
@@
 median Bray-Curtis index for the samples. The lines represent the range
-of the minimum and maximum Bray-Curits index value for each PCR 5-cycle
+of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle
 increment comparison. The closer a sample is to a Bray-Curtis index of
@@
 \textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
-Samples.} A) Precentage of chimeric sequences without the removal of
+Samples.} A) Percentage of chimeric sequences without the removal of
 chimeras with VSEARCH. C) Percentage of chimeric sequences with the

### 2018-05-30 submission/manuscript.Rmd
fix superscript in intro

@@
 Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2 = `r min(pull(r2_table, r2))`).
+We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).
 

### 2018-05-30 submission/manuscript.tex
fix superscript in intro

@@
 chimera prevalence and the number of OTUs with Platinum having the
-highest (R\^{}2 = 0.974) and Kappa having the worst (R\^{}2 = 0.478).
+highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
+(R\textsuperscript{2} = 0.478).
 

### 2018-06-01 submission/manuscript.Rmd
update figure legends

@@
 
-**Figure 1: Total Number of OTUs in Fecal Samples by Number of Cycles.** The points represent the median number of OTUs of all four fecal samples.
+**Figure 1: The number and range of OTUs in fecal samples varies by polymerase.** The points represent the median number of OTUs of all four fecal samples.
 The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
@@
 
-**Figure 2:  Total Number of OTUs in Mock Samples by Number of Cycles.** The points represent the median number of OTUs for the mock samples.
+**Figure 2: The number and range of OTUs in mock samples varies by polymerase.** The points represent the median number of OTUs for the mock samples.
 The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
@@
 
-**Figure 3: Bray-Curtis Community Differences by Five-Cycle Intervals.** A) within person differences based on the next 5-cycle PCR interval in fecal samples.
+**Figure 3: Lower five-cycle intervals have larger differences than higher five-cycle intervals using Bray-Curtis.** A) within person differences based on the next 5-cycle PCR interval in fecal samples.
 B) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
@@
 
-**Figure 4: HiFi DNA Polymerase Error Rate in Mock Samples.** The error bars represent the 75% interquartile range of the median error rate.
+**Figure 4: Variation in error rate by polymerase in mock community samples are similar across the number of cycles used.** The error bars represent the 75% interquartile range of the median error rate.
 
-**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
+**Figure 5: Chimera prevalence varies by polymerase and increases with higher number of cycles used.** A) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
 C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.
@@
 
-**Figure 6: The Correlation between Number of OTUs and Chimeras in Mock Samples.**
+**Figure 6: Regardless of polymerase used the higher the chimera prevalence the higher the observed number of OTUs in mock community samples.**
 
@@
 
-**Figure S1: HiFi DNA Polymerase Nucleotide Subsitutions in Mock Samples.**
+**Figure S1: No preference for specific base subsitutions was observed across the polymerases in mock community samples.**

### 2018-06-01 submission/manuscript.tex
update figure legends

@@
 
-\textbf{Figure 1: Total Number of OTUs in Fecal Samples by Number of
-Cycles.} The points represent the median number of OTUs of all four
-fecal samples. The lines represent the range of the minimum and maximum
-number of OTUs detected within the four fecal samples. The range in the
-number of OTUs detected in the different fecal samples increased as
-cycle number increased. This increased range also was larger for
+\textbf{Figure 1: The number and range of OTUs in fecal samples varies
+by polymerase.} The points represent the median number of OTUs of all
+four fecal samples. The lines represent the range of the minimum and
+maximum number of OTUs detected within the four fecal samples. The range
+in the number of OTUs detected in the different fecal samples increased
+as cycle number increased. This increased range also was larger for
 specific HiFi DNA polymerases.
 
-\textbf{Figure 2: Total Number of OTUs in Mock Samples by Number of
-Cycles.} The points represent the median number of OTUs for the mock
+\textbf{Figure 2: The number and range of OTUs in mock samples varies by
+polymerase.} The points represent the median number of OTUs for the mock
 samples. The lines represent the range of the minimum and maximum number
@@
 
-\textbf{Figure 3: Bray-Curtis Community Differences by Five-Cycle
-Intervals.} A) within person differences based on the next 5-cycle PCR
-interval in fecal samples. B) Within replicate difference based on the
-next 5-cycle PCR interval in mock samples. The points represent the
-median Bray-Curtis index for the samples. The lines represent the range
-of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle
-increment comparison. The closer a sample is to a Bray-Curtis index of
-1.00 the more dissimilar the bacterial community is of the two compared
-number of cycles.
+\textbf{Figure 3: Lower five-cycle intervals have larger differences
+than higher five-cycle intervals using Bray-Curtis.} A) within person
+differences based on the next 5-cycle PCR interval in fecal samples. B)
+Within replicate difference based on the next 5-cycle PCR interval in
+mock samples. The points represent the median Bray-Curtis index for the
+samples. The lines represent the range of the minimum and maximum
+Bray-Curtis index value for each PCR 5-cycle increment comparison. The
+closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar
+the bacterial community is of the two compared number of cycles.
 
-\textbf{Figure 4: HiFi DNA Polymerase Error Rate in Mock Samples.} The
+\textbf{Figure 4: Variation in error rate by polymerase in mock
+community samples are similar across the number of cycles used.} The
 error bars represent the 75\% interquartile range of the median error
@@
 
-\textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
-Samples.} A) Percentage of chimeric sequences without the removal of
-chimeras with VSEARCH. C) Percentage of chimeric sequences with the
-removal of chimeras with VSEARCH. C) The total percent of chimeric
-sequences removed with VSEARCH by cycle number. The error bars represent
-the 75\% interquartile range of the median.
+\textbf{Figure 5: Chimera prevalence varies by polymerase and increases
+with higher number of cycles used.} A) Percentage of chimeric sequences
+without the removal of chimeras with VSEARCH. C) Percentage of chimeric
+sequences with the removal of chimeras with VSEARCH. C) The total
+percent of chimeric sequences removed with VSEARCH by cycle number. The
+error bars represent the 75\% interquartile range of the median.
 
-\textbf{Figure 6: The Correlation between Number of OTUs and Chimeras in
-Mock Samples.}
+\textbf{Figure 6: Regardless of polymerase used the higher the chimera
+prevalence the higher the observed number of OTUs in mock community
+samples.}
 
@@
 
-\textbf{Figure S1: HiFi DNA Polymerase Nucleotide Subsitutions in Mock
-Samples.}
+\textbf{Figure S1: No preference for specific base subsitutions was
+observed across the polymerases in mock community samples.}
 

### 2018-06-01 submission/supplemental.Rmd
make changes to title and abstract
update table titles

@@
 
-# The Influence of Different High Fidelity DNA Polymerase on 16S rRNA Sequencing
+# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used
 
@@
 
-**Table S1: ANOVA Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal and Mock Samples**
+**Table S1: ANOVA results of differences in number of OTUs between polymerases in fecal and mock samples**
 
@@
 
-**Table S2: Tukey Post-Hoc Results of Differences in Number of OTUs between HiFi DNA Polymerases in Fecal and Mock Samples**
+**Table S2: Tukey post-hoc results of differences in number of OTUs between polymerases in fecal and mock samples**
 
@@
 
-**Table S3: Kruskal-Wallis Results of Differences in Error Rate, Sequences with an Error, and Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
+**Table S3: Kruskal-Wallis results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 
@@
 
-**Table S4: Dunn's Post-Hoc Results of Differences in Error Rate, Sequences with an Error, and Chimera Prevalence between HiFi DNA Polymerases in Mock Samples**
+**Table S4: Dunn's Post-Hoc results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 

### 2018-06-01 submission/manuscript.Rmd
updates from a final read through

@@
 
-# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used
+# Error Introduced into 16S rRNA Gene Sequencing Varies by High Fidelity DNA Polymerase Used
 
@@
 The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities.
-It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.
+It also assessed polymerase and number of cycle dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.
 
-**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-Additionally, the median number of OTUs varied by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg.
+at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
+Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
 When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
@@
 This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
-We will accomplish this by examining the number of OTUs, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases
+We will accomplish this by examining the number of OTUs, Bray-Curtis index, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases
 
@@
 
-***Human and Mock Samples:***
+***Human and mock samples:***
 Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
@@
 
-***PCR Protocol:***
+***PCR protocol:***
 The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
@@
 The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-For Kappa and Q5, the annealing and extension time were from a previously published protocol [@taq_Gohl_2016].
+For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016].
 For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
@@
 
-***Sequence Processing:***
+***Sequence processing:***
 The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
@@
 Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P.
+aeruginosa*, *E.
+coli*, *S.
+enterica*, *L.
+fermentum*, *E.
+faecalis*, *S.
+aureus*, *L.
+monocytogenes*, and *B.
+subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
 
-***Analysis Workflow:***
+***Analysis workflow:***
 To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
@@
 20 versus 25, 25 versus 30).
-Based on these observations we analyzed potential reasons for these differences.
+Based on these observations we analyzed potential reasons for these differences using the mock community samples.
 Analysis of the mock community of each polymerase for sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
@@
 
-***Statistical Analysis:***
+***Statistical analysis:***
 All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
@@
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
-Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth.
+For the comparison in fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
+Bray-Curtis distance matrices were generated using mothur.
 The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
 For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
-Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
+Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
 
-***Reproducible Methods:***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number SRP132931.
+***Reproducible methods:***
+The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
 
@@
 Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1].
-There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1].
-Using a Tukey post-hoc test to identify which polymerase groups were different from each other at 35 cycles, no difference was found (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
+There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1], but a Tukey post-hoc test failed to identify which specific polymerase groups were different (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
 The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2].
@@
 In contrast to the results obtained with fecal samples, differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1].
-Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community is largely due to Kappa and Platinum differences versus the other polymerases across the different number of cycles [Table S2].
+Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S2].
 Based on these observations in fecal and mock communities, it is clear that using different polymerases result in a different total number of OTUs within a sample.
 
-***The bacterial community is similar within polymerase and varies by number of cycles:***
+***The bacterial communities generated by the polymerases were similar but varied by number of cycles:***
 Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3].
-Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`.
-For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles.
+Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
+For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
 There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
-the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A].
-Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
-The mock community has similar trends to the observations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
+the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A].
+Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
+Similar trends were observed in the mock community but none were significant (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
 Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
@@
 The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].
-The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 and Table S4].
-The total sequences with at least one error is also polymerase dependent with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
+The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4].
+The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
 These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
@@
 Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis.
-All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered.
+All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that want to be answered.
 Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.
@@
 
-Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing results.
+Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing.
 Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics.
@@
 The authors would like to thank all the study participants in ERIN whose samples were utilized.
-We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
+We also would like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
 Additional thanks to members of the Schloss lab and Dr.

### 2018-06-01 submission/manuscript.tex
updates from a final read through

@@
 
-\section{Error Introduced into 16S rRNA Gene Sequencing Results Varies
-by High Fidelity DNA Polymerase
-Used}\label{error-introduced-into-16s-rrna-gene-sequencing-results-varies-by-high-fidelity-dna-polymerase-used}
+\section{Error Introduced into 16S rRNA Gene Sequencing Varies by High
+Fidelity DNA Polymerase
+Used}\label{error-introduced-into-16s-rrna-gene-sequencing-varies-by-high-fidelity-dna-polymerase-used}
 
@@
 (OTUs) for fecal samples and mock communities. It also assessed
-polymerase dependent differences in the Bray-Curtis index, error rate,
-sequence error prevalence, chimera prevalence, and the correlation
-between chimera prevalence and number of OTUs.
+polymerase and number of cycle dependent differences in the Bray-Curtis
+index, error rate, sequence error prevalence, chimera prevalence, and
+the correlation between chimera prevalence and number of OTUs.
 
@@
 range in the number of OTUs detected was not consistent between HiFi DNA
-polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
+polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
 Phusion = 84 - 136). Additionally, the median number of OTUs varied by
-HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
-mock community samples the variation in the number of OTUs detected by
-the polymerases was observable as early as 20 cycles (P-value = 0.002).
-There also was a large range in the number of OTUs amplified by the
-polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
-Chimera prevalence in mock communities varied by polymerase with
-differences being most notable at 35 cycles (Kappa = 5.71\% (median)
-versus Platinum = 26.62\%) and this variation persisted after chimera
-removal using VSEARCH. We also observed positive correlations between
-chimera prevalence and the number of OTUs with Platinum having the
-highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
-(R\textsuperscript{2} = 0.478).
+HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001). When
+analyzing mock community samples the variation in the number of OTUs
+detected by the polymerases was observable as early as 20 cycles
+(P-value = 0.002). There also was a large range in the number of OTUs
+amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
+Phusion = 14 - 73). Chimera prevalence in mock communities varied by
+polymerase with differences being most notable at 35 cycles (Kappa =
+5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
+after chimera removal using VSEARCH. We also observed positive
+correlations between chimera prevalence and the number of OTUs with
+Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
+having the worst (R\textsuperscript{2} = 0.478).
 
@@
 16S rRNA gene sequencing. We will accomplish this by examining the
-number of OTUs, error rate, number of sequences with an error, and
-chimera prevalence at varying number of cycles in five different high
-fidelity DNA polymerases
+number of OTUs, Bray-Curtis index, error rate, number of sequences with
+an error, and chimera prevalence at varying number of cycles in five
+different high fidelity DNA polymerases
 
@@
 
-\textbf{\emph{Human and Mock Samples:}} Fecal samples were obtained from
+\textbf{\emph{Human and mock samples:}} Fecal samples were obtained from
 4 individuals who were part of the Enterics Research Investigational
@@
 
-\textbf{\emph{PCR Protocol:}} The five different high fidelity DNA
+\textbf{\emph{PCR protocol:}} The five different high fidelity DNA
 polymerases (hereto referred to as polymerases) that were tested
@@
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
-For Kappa and Q5, the annealing and extension time were from a
+For Kappa and Q5, the annealing and extension time also were from a
 previously published protocol (Gohl et al., 2016). For Phusion, the
@@
 
-\textbf{\emph{Sequence Processing:}} The mothur software program was
+\textbf{\emph{Sequence processing:}} The mothur software program was
 used for all sequence processing steps (Schloss et al., 2009). The
@@
 was determined using the `seq.error' command on mock samples to compare
-back to the reference 16S sequences (Schloss et al., 2009; Cole et al.,
-2013; Rognes et al., 2016).
+back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
+coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
+\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
+(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
 
-\textbf{\emph{Analysis Workflow:}} To adjust for unequal sequencing, all
+\textbf{\emph{Analysis workflow:}} To adjust for unequal sequencing, all
 samples were rarefied to 1000 sequences for downstream analysis. The
@@
 from the previous cycle (e.g.~20 versus 25, 25 versus 30). Based on
-these observations we analyzed potential reasons for these differences.
-Analysis of the mock community of each polymerase for sequence error
-rate, number of sequences with an error, base substitution, and numbers
-of chimeras before and after chimera removal with VSEARCH was assessed.
-Additionally, the correlation between the number of chimeras and the
-number of OTUs was also assessed.
+these observations we analyzed potential reasons for these differences
+using the mock community samples. Analysis of the mock community of each
+polymerase for sequence error rate, number of sequences with an error,
+base substitution, and numbers of chimeras before and after chimera
+removal with VSEARCH was assessed. Additionally, the correlation between
+the number of chimeras and the number of OTUs was also assessed.
 
-\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
+\textbf{\emph{Statistical analysis:}} All analysis was done with the R
 (v 3.4.4) software package (R Core Team, 2017). Data transformation and
@@
 2017). Differences in the total number of OTUs were analyzed using an
-ANOVA with a tukey post-hoc test. For the fecal samples the data was
-normalized to each individual by cycle number to account for the
-biological variation between people. Bray-Curtis distance matrices were
-generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
-15000 sequence depth. The distance matrix data was analyzed using
-PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
-Kruskal-Wallis tests within R. For both error and chimera analysis,
-samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
-Where applicable correction for multiple comparison utilized the
+ANOVA with a tukey post-hoc test. For the comparison in fecal samples
+the data was normalized to each individual by cycle number to account
+for the biological variation between people. Bray-Curtis distance
+matrices were generated using mothur. The distance matrix data was
+analyzed using PERMANOVA with the vegan package (v 2.4.5) (Oksanen et
+al., 2017) and Kruskal-Wallis tests within R. For both error and chimera
+analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc
+test. Where applicable, correction for multiple comparison utilized the
 Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
 
-\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
+\textbf{\emph{Reproducible methods:}} The code and analysis can be found
 here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
-The raw sequences can be found in the SRA at the following accession
+The raw sequences can be found on the SRA at the following accession
 number SRP132931.
@@
 fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
-16.35, P-value = 9.7e-05) {[}Table S1{]}. Using a Tukey post-hoc test to
-identify which polymerase groups were different from each other at 35
-cycles, no difference was found (P-value \textgreater{} 0.05) {[}Table
-S2{]}. The polymerase dependent difference in the range of the number of
-OTUs also was observed in the mock community samples {[}Figure 2{]}. The
-closest the polymerases came to the total of 8 OTUs created by the mock
-reference 16S sequences was at 25 and 30 cycles {[}Figure 2{]}.
-Regardless of if fecal or mock communities were used, the same
-polymerases generated high and low number of OTUs and this was
-consistent across the number of cycles used {[}Figure 1-2 \& Table
-S1-S2{]}. In contrast to the results obtained with fecal samples,
-differences between polymerases for the number of OTUs created were
-observed as early as 20 cycles in the mock community (F-stat = 15.82,
-P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc test, the
-majority of differences for the number of OTUs detected in the mock
-community is largely due to Kappa and Platinum differences versus the
-other polymerases across the different number of cycles {[}Table S2{]}.
-Based on these observations in fecal and mock communities, it is clear
-that using different polymerases result in a different total number of
-OTUs within a sample.
+16.35, P-value = 9.7e-05) {[}Table S1{]}, but a Tukey post-hoc test
+failed to identify which specific polymerase groups were different
+(P-value \textgreater{} 0.05) {[}Table S2{]}. The polymerase dependent
+difference in the range of the number of OTUs also was observed in the
+mock community samples {[}Figure 2{]}. The closest the polymerases came
+to the total of 8 OTUs created by the mock reference 16S sequences was
+at 25 and 30 cycles {[}Figure 2{]}. Regardless of if fecal or mock
+communities were used, the same polymerases generated high and low
+number of OTUs and this was consistent across the number of cycles used
+{[}Figure 1-2 \& Table S1-S2{]}. In contrast to the results obtained
+with fecal samples, differences between polymerases for the number of
+OTUs created were observed as early as 20 cycles in the mock community
+(F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc
+test, the majority of differences for the number of OTUs detected in the
+mock community was largely due to Kappa and Platinum versus the other
+polymerases across the number of cycles used {[}Table S2{]}. Based on
+these observations in fecal and mock communities, it is clear that using
+different polymerases result in a different total number of OTUs within
+a sample.
 
-\textbf{\emph{The bacterial community is similar within polymerase and
-varies by number of cycles:}} Within each respective 5-cycle increment
-comparison there was no difference in Bray-Curtis index between the
-polymerases (P-value \textgreater{} 0.05) {[}Figure 3{]}. Using
-PERMANOVA to test for community differences based on any of the number
-of cycles within polymerases, only Phusion had cycle dependent
-differences (P-value = 0.03. For fecal samples, Phusion was one of two
-polymerases that had enough sequences to be rarefied to 1000 at 15
-cycles. There was an overall decrease in Bray-Curtis index by cycle
-comparison group (e.g.~the 15 cycle versus 20 cycle group compared to
-the 30 cycle to 35 cycle group) (P-value \textless{} 0.01) {[}Figure
-3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
-cycle versus 25 cycle comparison groups had a higher Bray-Curtis index
-than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
-The mock community has similar trends to the observations reported for
-the fecal bacterial community but none were significant \textgreater{}
-0.05) {[}Figure 3B{]}. Overall, these data suggest that the number of
-cycles can change the bacterial community independent of the polymerase
-used to generate the sequences.
+\textbf{\emph{The bacterial communities generated by the polymerases
+were similar but varied by number of cycles:}} Within each respective
+5-cycle increment comparison there was no difference in Bray-Curtis
+index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
+3{]}. Using PERMANOVA to test for community differences based on any of
+the number of cycles within polymerases, only Phusion had cycle
+dependent differences (P-value = 0.03). For fecal samples, Phusion was
+one of two polymerases that had enough sequences to be rarefied to 1000
+sequences at 15 cycles. There was an overall decrease in Bray-Curtis
+index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
+compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
+0.01) {[}Figure 3A{]}. Using a Dunn's post-hoc test the 15 cycle versus
+20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
+Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
+(P-value \textless{} 0.05). Similar trends were observed in the mock
+community but none were significant (P-value \textgreater{} 0.05)
+{[}Figure 3B{]}. Overall, these data suggest that the number of cycles
+can change the bacterial community independent of the polymerase used to
+generate the sequences.
 
@@
 majority of the differences across the number of cycles was between
-Kappa and the other polymerases {[}Figure 4 and Table S4{]}. The total
-sequences with at least one error is also polymerase dependent with the
-majority of differences being between Kappa or Accuprime and the other
-polymerases {[}Table S3 \& S4{]}. These differences in error rates were
-not due to polymerase dependent differences in base substitution rate
-{[}Figure S1{]}. Collectively, the results suggest that sequence error
-is dependent on polymerase, persists across the number of cycles used,
-and are not due to any bias towards a specific base substitution.
+Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
+sequences with at least one error was also polymerase dependent, with
+the majority of differences being between Kappa or Accuprime and the
+other polymerases {[}Table S3 \& S4{]}. These differences in error rates
+were not due to polymerase dependent differences in base substitution
+rate {[}Figure S1{]}. Collectively, the results suggest that sequence
+error is dependent on polymerase, persists across the number of cycles
+used, and are not due to any bias towards a specific base substitution.
 
@@
 study basis. All aspects of the 16S rRNA gene sequencing workflow should
-be customized for the specific scientific questions that wants to be
+be customized for the specific scientific questions that want to be
 answered. Although a diversity of approaches may make reproducibility
@@
 polymerases used or the number of cycles chosen can change diversity
-metrics used in 16S rRNA gene sequencing results. Care should be taken
-when choosing a polymerase for 16S rRNA gene surveys because they can
+metrics used in 16S rRNA gene sequencing. Care should be taken when
+choosing a polymerase for 16S rRNA gene surveys because they can
 influence diversity-based metrics. By knowing the inherent bias
@@
 The authors would like to thank all the study participants in ERIN whose
-samples were utilized. We would also like to thank Judy Opp and April
+samples were utilized. We also would like to thank Judy Opp and April
 Cockburn for their effort in sequencing the samples as part of the

### 2018-07-03 submission/manuscript.Rmd
change title
update abstract - background and methods

@@
 
-# Error Introduced into 16S rRNA Gene Sequencing Varies by High Fidelity DNA Polymerase Used
+# Assessing Differences in High Fidelity DNA Polymerase for 16S rRNA Gene Sequencing
 
@@
 
-**Background.** It is challenging to compare 16S rRNA gene sequencing data across studies and one of the reasons for this is due to error.
-There are many different places throughout the workflow where error can be introduced into the pipeline.
-Here, we focus on studying how the number of cycles and high fidelity (HiFi) DNA polymerase introduce error by varying cycle number and polymerase used to amplify 16S rRNA genes in human fecal and mock community samples.
+**Background.** A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, amplification, and sequencing steps.
+At each of these stages error can be introduced that will change the underlying bacterial community composition results.
+In this study we focus on the amplification step's contribution to this overall error.
+To accomplish this we assessed 16S rRNA gene sequencing results in human fecal and mock community samples after using different high fidelity (HiFi) DNA polymerases and number of amplification cycles.
 
@@
 Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach.
-The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities.
-It also assessed polymerase and number of cycle dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and the  correlation between chimera prevalence and number of OTUs.
+We first assessed error rate, sequence error prevalence, chimera prevalence, correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs), and GC dependent differences by polymerase and number of cycles.
+Next, differences in the number of OTUs and taxa was examined based on the polymerase and number of cycles used.
+Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles.
+Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor.
+We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models also were the most important in the model for individual donors.
 

### 2018-07-03 submission/manuscript.Rmd
update abstract

@@
 
+# Read in GC data
+gc_data <- read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv")
+
+# Read in RF model data
+dunn_rf_data <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv")
+
 ```
@@
 Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach.
-We first assessed error rate, sequence error prevalence, chimera prevalence, correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs), and GC dependent differences by polymerase and number of cycles.
+We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles.
 Next, differences in the number of OTUs and taxa was examined based on the polymerase and number of cycles used.
@@
 
-**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg.
+**Results.** Although the total counts for specific taxa varied by polymerase, we found no noticeable differences in total counts based on high and low GC content (P-value $\geqslant$ `r round(min(pull(gc_data, p.value)), digits = 2)`).
+Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
+We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).
+When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
+There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
+When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg.
 at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
 Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).
+Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
+Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor sample model.
 
-**Conclusions.** Although the variation in the number of OTUs in fecal samples could be due to certain polymerases capturing the biological variability better than others, this is unlikely to be the main reason for our observed differences.
-In mock community samples, the strong correlation between chimera prevalence and the number of OTUs suggests that this is the main reason for differences between the polymerases.
-Ultimately, this variation makes comparison across studies difficult and care should be exercised when choosing the polymerase and number of cycles in 16S rRNA gene sequencing studies.
+**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between indivdiuals.
+Additionally, our observations were not able to identify GC-based differences in amplification based on the number of cycles used.
+Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing, should be detectable regardless of polymerase and number of cycles used.
 

### 2018-07-03 submission/manuscript.tex
update abstract

@@
 
-\section{Error Introduced into 16S rRNA Gene Sequencing Varies by High
-Fidelity DNA Polymerase
-Used}\label{error-introduced-into-16s-rrna-gene-sequencing-varies-by-high-fidelity-dna-polymerase-used}
+\section{Assessing Differences in High Fidelity DNA Polymerase for 16S
+rRNA Gene
+Sequencing}\label{assessing-differences-in-high-fidelity-dna-polymerase-for-16s-rrna-gene-sequencing}
 
@@
 
-\textbf{Background.} It is challenging to compare 16S rRNA gene
-sequencing data across studies and one of the reasons for this is due to
-error. There are many different places throughout the workflow where
-error can be introduced into the pipeline. Here, we focus on studying
-how the number of cycles and high fidelity (HiFi) DNA polymerase
-introduce error by varying cycle number and polymerase used to amplify
-16S rRNA genes in human fecal and mock community samples.
+\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
+divided into preservation, extraction, amplification, and sequencing
+steps. At each of these stages error can be introduced that will change
+the underlying bacterial community composition results. In this study we
+focus on the amplification step's contribution to this overall error. To
+accomplish this we assessed 16S rRNA gene sequencing results in human
+fecal and mock community samples after using different high fidelity
+(HiFi) DNA polymerases and number of amplification cycles.
 
@@
 communities (technical replicates n=4) consisting of previously isolated
-whole genomes of 8 different bacteria used the same approach. The
-analysis initially examined the number of Operational Taxonomic Units
-(OTUs) for fecal samples and mock communities. It also assessed
-polymerase and number of cycle dependent differences in the Bray-Curtis
-index, error rate, sequence error prevalence, chimera prevalence, and
-the correlation between chimera prevalence and number of OTUs.
+whole genomes of 8 different bacteria used the same approach. We first
+assessed GC dependent differences, error rate, sequence error
+prevalence, chimera prevalence, and correlation between chimera
+prevalence and number of Operational Taxonomic Units (OTUs) by
+polymerase and number of cycles. Next, differences in the number of OTUs
+and taxa was examined based on the polymerase and number of cycles used.
+Additionally, differences in the bacterial community composition by the
+Bray-Curtis index also was assessed based on polymerase and number of
+cycles. Finally, Random Forest models were created to test whether the
+bacterial community was better at classifying polymerases, number of
+cycles, or individual donor. We also assessed whether the most important
+taxa in the polymerase and number of cycle Random Forest models also
+were the most important in the model for individual donors.
 
-\textbf{Results.} When analyzing fecal samples we observed that the
-range in the number of OTUs detected was not consistent between HiFi DNA
-polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
-Phusion = 84 - 136). Additionally, the median number of OTUs varied by
-HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001). When
+\textbf{Results.} Although the total counts for specific taxa varied by
+polymerase, we found no noticeable differences in total counts based on
+high and low GC content (P-value \(\geqslant\) 0.23). Chimera prevalence
+in mock communities varied by polymerase with differences being most
+notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
+and this variation persisted after chimera removal using VSEARCH. We
+also observed positive correlations between chimera prevalence and the
+number of OTUs with Platinum having the highest (R\textsuperscript{2} =
+0.974) and Kappa having the worst (R\textsuperscript{2} = 0.478). When
 analyzing mock community samples the variation in the number of OTUs
@@
 amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
-Phusion = 14 - 73). Chimera prevalence in mock communities varied by
-polymerase with differences being most notable at 35 cycles (Kappa =
-5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
-after chimera removal using VSEARCH. We also observed positive
-correlations between chimera prevalence and the number of OTUs with
-Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
-having the worst (R\textsuperscript{2} = 0.478).
+Phusion = 14 - 73). When analyzing fecal samples we observed that the
+range in the number of OTUs detected was not consistent between HiFi DNA
+polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
+Phusion = 84 - 136). Additionally, the median number of OTUs varied by
+HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001).
+Random Forest models were most successful at classifying individual
+donor samples rather than polymerase or number of cycles used (P-value
+\(\leqslant\) 5.49e-07). Additionally, the most important OTUs in the
+polymerase and number of cycle models were not the most important in the
+individual donor sample model.
 
-\textbf{Conclusions.} Although the variation in the number of OTUs in
-fecal samples could be due to certain polymerases capturing the
-biological variability better than others, this is unlikely to be the
-main reason for our observed differences. In mock community samples, the
-strong correlation between chimera prevalence and the number of OTUs
-suggests that this is the main reason for differences between the
-polymerases. Ultimately, this variation makes comparison across studies
-difficult and care should be exercised when choosing the polymerase and
-number of cycles in 16S rRNA gene sequencing studies.
+\textbf{Conclusions.} Although there are 16S rRNA gene sequencing
+differences based on polymerase and number of cycles used, they are
+small with respect to the biological differences between indivdiuals.
+Additionally, our observations were not able to identify GC-based
+differences in amplification based on the number of cycles used.
+Collectively, these results provide evidence that a real biological
+difference between groups, based on 16S rRNA gene sequencing, should be
+detectable regardless of polymerase and number of cycles used.
 

### 2018-07-03 submission/manuscript.tex
change title

@@
 
-\section{Assessing Differences in High Fidelity DNA Polymerase for 16S
-rRNA Gene
-Sequencing}\label{assessing-differences-in-high-fidelity-dna-polymerase-for-16s-rrna-gene-sequencing}
+\section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
+High Fidelity DNA
+Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}
 

### 2018-07-05 code/make_rf_fecal_imp_vars.graph.R
change axis label

@@ -42,7 +42,7 @@ model_graph <- graph_data %>%
   mutate(model_type = factor(model_type, 
                              levels = c("fecal_sample_group", "fecal_polymerase", "fecal_cycles", 
                                         "mock_polymerase", "mock_cycles"), 
-                             labels = c("Fecal Sample Group Model", "Fecal Polymerase Model", 
+                             labels = c("Fecal Individual Model", "Fecal Polymerase Model", 
                                         "Fecal Cycles Model", "Mock Polymerase Model", "Mock Cycles Model"))) %>% 
   ggplot(aes(model_type, probs)) + 
     stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 

### 2018-07-05 submission/manuscript.Rmd
change title
update methods

@@
 
-# Assessing Differences in High Fidelity DNA Polymerase for 16S rRNA Gene Sequencing
+# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase
 
@@
 To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
-The total number of OTUs was analyzed for both the fecal and mock community samples.
-For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g.
-20 versus 25, 25 versus 30).
-Based on these observations we analyzed potential reasons for these differences using the mock community samples.
-Analysis of the mock community of each polymerase for sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
+Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
 Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
+The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples.
+Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data.
+Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).
 
@@
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the comparison in fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people.
+For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation.
 Bray-Curtis distance matrices were generated using mothur.
 The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
+The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation].
+A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value.
+The probability of a correct call was obtained from this Logloss value by taking the negative natural logrithim.
 For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.

### 2018-07-05 submission/manuscript.tex
update methods

@@
 \textbf{\emph{Analysis workflow:}} To adjust for unequal sequencing, all
-samples were rarefied to 1000 sequences for downstream analysis. The
-total number of OTUs was analyzed for both the fecal and mock community
-samples. For fecal samples, cycle dependent affects on Bray-Curtis
-indices were assessed for cycle group and within individual differences
-from the previous cycle (e.g.~20 versus 25, 25 versus 30). Based on
-these observations we analyzed potential reasons for these differences
-using the mock community samples. Analysis of the mock community of each
-polymerase for sequence error rate, number of sequences with an error,
-base substitution, and numbers of chimeras before and after chimera
-removal with VSEARCH was assessed. Additionally, the correlation between
-the number of chimeras and the number of OTUs was also assessed.
+samples were rarefied to 1000 sequences for downstream analysis.
+Analysis of the mock community of each polymerase for GC-based
+amplification differences, sequence error rate, number of sequences with
+an error, base substitution, and numbers of chimeras before and after
+chimera removal with VSEARCH was assessed. Additionally, the correlation
+between the number of chimeras and the number of OTUs was also assessed.
+The total number of OTUs, taxa differences, and Bray-Curtis indices were
+analyzed for both the fecal and mock community samples. Finally, Random
+Forest models were created to assess whether classification of
+polymerase, cycles, or individual could be performed best using 16S rRNA
+gene sequencing data. Additionally, overlap between the most important
+OTUs to the three models was assessed using mean decrease in accuracy
+(MDA).
 
@@
 2017). Differences in the total number of OTUs were analyzed using an
-ANOVA with a tukey post-hoc test. For the comparison in fecal samples
-the data was normalized to each individual by cycle number to account
-for the biological variation between people. Bray-Curtis distance
-matrices were generated using mothur. The distance matrix data was
-analyzed using PERMANOVA with the vegan package (v 2.4.5) (Oksanen et
-al., 2017) and Kruskal-Wallis tests within R. For both error and chimera
-analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc
-test. Where applicable, correction for multiple comparison utilized the
+ANOVA with a tukey post-hoc test. For the comparison of the number of
+OTUS in fecal samples the data was normalized to each individual by
+cycle number to account for the biological variation. Bray-Curtis
+distance matrices were generated using mothur. The distance matrix data
+was analyzed using PERMANOVA with the vegan package (v 2.4.5) (Oksanen
+et al., 2017) and Kruskal-Wallis tests within R. The Random Forest
+models were run using the caret package (v 6.0.78) (Jed Wing et al.,
+2017). A total of 100 10-fold CV runs on different 80/20 splits of the
+data was run to generate a range of the Logloss value. The probability
+of a correct call was obtained from this Logloss value by taking the
+negative natural logrithim. For both error and chimera analysis, samples
+were tested using Kruskal-Wallis with a Dunns post-hoc test. Where
+applicable, correction for multiple comparison utilized the
 Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
@@
 
+\hypertarget{ref-caret_citation}{}
+Jed Wing MKC from., Weston S., Williams A., Keefer C., Engelhardt A.,
+Cooper T., Mayer Z., Kenkel B., R Core Team., Benesty M., Lescarbeau R.,
+Ziem A., Scrucca L., Tang Y., Candan C., Hunt. T. 2017. \emph{Caret:
+Classification and regression training}.
+
 \hypertarget{ref-Kebschull2015}{}

### 2018-07-05 code/make_gc_content_graph.R
update gc data output

@@ -9,13 +9,14 @@ source("code/functions.R")
 loadLibs(c("tidyverse", "gridExtra", "scales", "viridis", "RColorBrewer", "ggrepel"))
 
 # Load in needed data
-summary_data <- read_csv("data/process/tables/gc_content_amp_summary.csv")
+summary_data <- read_csv("data/process/tables/gc_content_amp_summary.csv") %>% 
+  filter(cycles != "fifteen")
 
 
 gc_graph <- summary_data %>% 
   mutate(cycles = factor(cycles, 
-                         levels = c("fifteen", "twenty", "twenty_five", "thirty", "thirty_five"), 
-                         labels = c("15x", "20x", "25x", "30x", "35x")), 
+                         levels = c("twenty", "twenty_five", "thirty", "thirty_five"), 
+                         labels = c("20x", "25x", "30x", "35x")), 
          higher_gc = factor(higher_gc, 
                             levels = c("yes", "no"), 
                             labels = c("Yes", "No")), 

### 2018-07-05 code/run_mock_gc_content_analysis.R

@@ -61,6 +61,7 @@ summarized_table <- error_summary %>%
            whole_genome_gc_percent <= wg_median_cutoff ~ "no", 
            TRUE ~ "uh oh"))
 
+write_csv(summarized_table, "data/process/tables/gc_summary_table.csv")
 
 
 # Summarise the counts from each sequence by polymerase and cycle

### 2018-07-05 code/run_mock_gc_content_analysis.R
update code

@@ -46,7 +46,7 @@ summarized_table <- error_summary %>%
   group_by(reference) %>% 
   filter(weight == max(weight)) %>% 
   ungroup() %>% 
-  select(query, reference, AA, TT, CC, GG, total) %>% 
+  select(query, reference, weight, AA, TT, CC, GG, total) %>% 
   mutate(
     whole_genome_gc_percent = c(32.7, 38.04, 50.6, 52.1, 66.6, 38, 49.7, 43.61), 
     at_percent = (AA + TT)/total * 100, 
@@ -61,8 +61,82 @@ summarized_table <- error_summary %>%
            whole_genome_gc_percent <= wg_median_cutoff ~ "no", 
            TRUE ~ "uh oh"))
 
-write_csv(summarized_table, "data/process/tables/gc_summary_table.csv")
 
+seq_compare <- error_table %>% 
+  filter(cycles != "fifteen") %>% 
+  left_join(select(error_summary, query, reference), by = "query") %>% 
+  group_by(taq, cycles, sample_name.y, reference) %>% 
+  summarise(new_total = sum(total_seqs)) %>% 
+  ungroup() %>% 
+  group_by(taq, cycles, sample_name.y) %>% 
+  mutate(group_total = sum(new_total)) %>% 
+  ungroup() %>% 
+  group_by(taq, cycles, reference) %>% 
+  summarise(median_abund = median(new_total/group_total), 
+            iqr25 = quantile(new_total/group_total)["25%"],
+            iqr75 = quantile(new_total/group_total)["75%"],
+            min_abund = min(new_total/group_total),
+            max_abund = max(new_total/group_total))
+
+write_csv(seq_compare, "data/process/tables/gc_summary_table.csv")
+
+
+# Test based on taxa generally
+taxa_gc_test <- error_table %>% 
+  filter(cycles != "fifteen") %>% 
+  left_join(select(error_summary, query, reference), by = "query") %>% 
+  group_by(taq, cycles, sample_name.y, reference) %>% 
+  summarise(new_total = sum(total_seqs)) %>% 
+  mutate(group_total = sum(new_total), 
+         corrected_values = new_total/group_total) %>% 
+  ungroup() %>% 
+  group_by(taq, cycles) %>% 
+  nest() %>% 
+  mutate(test = map(data, ~kruskal.test(corrected_values ~ factor(reference), data = .x)), 
[…]

### 2018-07-05 code/run_mock_gc_content_analysis.R
update with total seq correction

@@ -140,30 +140,42 @@ write_csv(summary_test_data, "data/process/tables/gc_taxa_dunn_summary.csv")
 
 # Summarise the counts from each sequence by polymerase and cycle
 seq_counts_table <- error_table %>% 
-  group_by(query, taq, cycles) %>% 
-  summarise(median_abund = median(total_seqs), 
-            iqr25 = quantile(total_seqs)["25%"],
-            iqr75 = quantile(total_seqs)["75%"],
-            min_abund = min(total_seqs),
-            max_abund = max(total_seqs))
+  group_by(taq, cycles, sample_name.y) %>% 
+  mutate(overall_total = sum(total_seqs)) %>% 
+  ungroup() %>% 
+  mutate(rel_abund = total_seqs/overall_total) %>% 
+  group_by(taq, cycles) %>% 
+  summarise(median_abund = median(rel_abund), 
+            iqr25 = quantile(rel_abund)["25%"],
+            iqr75 = quantile(rel_abund)["75%"],
+            min_abund = min(rel_abund),
+            max_abund = max(rel_abund))
   
-test <- seq_counts_table %>% 
+test <- error_table %>% 
+  group_by(taq, cycles, sample_name.y) %>% 
+  mutate(overall_total = sum(total_seqs)) %>% 
+  ungroup() %>% 
+  mutate(rel_abund = total_seqs/overall_total) %>% 
   inner_join(select(summarized_table, query, reference, higher_gc), by = "query") %>% 
-  group_by(higher_gc, cycles, taq) %>% 
-  summarise(median_of_median_abund = median(median_abund), 
-            iqr25 = quantile(median_abund)["25%"],
-            iqr75 = quantile(median_abund)["75%"],
-            min_abund = min(median_abund),
-            max_abund = max(median_abund))
+  group_by(taq, cycles, higher_gc) %>% 
+  summarise(median_abund = median(rel_abund), 
+            iqr25 = quantile(rel_abund)["25%"],
+            iqr75 = quantile(rel_abund)["75%"],
+            min_abund = min(rel_abund),
+            max_abund = max(rel_abund))
 
-whole_test <- seq_counts_table %>% 
+whole_test <- error_table %>% 
+  group_by(taq, cycles, sample_name.y) %>% 
+  mutate(overall_total = sum(total_seqs)) %>% 
+  ungroup() %>% 
[…]

### 2018-07-05 code/make_gc_content_graph.R
change first figure

@@ -1,4 +1,4 @@
-### Plot RF results MDA for classifying samples
+### Plot GC content graph
 # Marc Sze
 
 
@@ -9,32 +9,41 @@ source("code/functions.R")
 loadLibs(c("tidyverse", "gridExtra", "scales", "viridis", "RColorBrewer", "ggrepel"))
 
 # Load in needed data
-summary_data <- read_csv("data/process/tables/gc_content_amp_summary.csv") %>% 
-  filter(cycles != "fifteen")
+summary_data <- read_csv("data/process/tables/gc_summary_table.csv")
 
+# summary_data <- read_csv("data/process/tables/gc_content_amp_summary.csv") %>% 
+#   filter(cycles != "fifteen")
 
 gc_graph <- summary_data %>% 
   mutate(cycles = factor(cycles, 
                          levels = c("twenty", "twenty_five", "thirty", "thirty_five"), 
                          labels = c("20x", "25x", "30x", "35x")), 
-         higher_gc = factor(higher_gc, 
-                            levels = c("yes", "no"), 
-                            labels = c("Yes", "No")), 
          taq = factor(taq, 
                       levels = c("ACC", "K", "PHU", "PL", "Q5"), 
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>% 
-  ggplot(aes(cycles, log10(median_of_median_abund+1), color = higher_gc)) + 
-  geom_pointrange(aes(ymin = log10(min_abund+1), ymax = log10(max_abund+1)), position = position_dodge(width = 0.7)) + 
-  facet_wrap(~taq) + theme_bw() + 
-  labs(x = "", y = expression(Log["10"]~median~counts)) + 
-  scale_color_manual(name = "Higher 16S V4\nGC Content", values = c('#FF8000', '#8B5A2B')) + 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")), 
+         reference = factor(reference, 
+                            levels = c("Staphylococcus_aureus_16S", "Listeria_monocytogenes_16S", 
+                                       "Enterococcus_faecalis_16S", "Bacillus_subtilis_16S", 
+                                       "Lactobacillus_fermentum_16S", "Escherichia_coli_16S_1", 
+                                       "Salmonella_enterica_16S_1", "Pseudomonas_aeruginosa_16S"), 
[…]

### 2018-07-05 submission/manuscript.Rmd
fix abstract

@@
 # Read in GC data
-gc_data <- read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv")
+gc_data <- read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv") %>% 
+  filter(cycles != "fifteen") %>% 
+  mutate(bh = p.adjust(p.value, method = "BH"))
+
+gc_genome_data <- read_csv("../data/process/tables/whole_genome_gc_wilcox_test_summary.csv") %>% 
+  filter(cycles != "fifteen") %>% 
+  mutate(bh = p.adjust(p.value, method = "BH"))
+
+gc_summary <- read_csv("../data/process/tables/gc_summary_table.csv")
 
@@
 
-**Results.** Although the total counts for specific taxa varied by polymerase, we found no noticeable differences in total counts based on high and low GC content (P-value $\geqslant$ `r round(min(pull(gc_data, p.value)), digits = 2)`).
+**Results.** In addition to the total counts for specific taxa varying by polymerase, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`).
 Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
@@
 **Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between indivdiuals.
-Additionally, our observations were not able to identify GC-based differences in amplification based on the number of cycles used.
 Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing, should be detectable regardless of polymerase and number of cycles used.
@@
 Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
+High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community.
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
@@
 ## Results
+
+***16S rRNA gene sequencing counts do not differ by GC content:*** Although the 16S rRNA gene sequencing counts varied by polymerase used, there was no difference between V4 16S rRNA genes with high and low GC content at every amplification cycle [Figure 1 & Table S1].
+Additionally, there was also no difference in 16S rRNA gene sequencing counts between high and low GC content genomes [Table S2].
+
+***Sequence error varies by polymerase and is consistent across the number of cycles used:***
+The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].
+The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4].
+The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
+These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
+Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.
+
+***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
+We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
+Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S4].
+Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
+The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 5A & 5B].
+Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
+A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6].
+This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6].
+This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 
@@
 Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
-
-***Sequence error varies by polymerase and is consistent across the number of cycles used:***
-The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].
-The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4].
-The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
-These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
-Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.
-
-***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
-We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
-Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S4].
-Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
-The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 5A & 5B].
-Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
-A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6].
-This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6].
-This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 

### 2018-07-05 submission/manuscript.tex
fix abstract

@@
 
-\textbf{Results.} Although the total counts for specific taxa varied by
-polymerase, we found no noticeable differences in total counts based on
-high and low GC content (P-value \(\geqslant\) 0.23). Chimera prevalence
-in mock communities varied by polymerase with differences being most
-notable at 35 cycles (Kappa = 5.71\% (median) versus Platinum = 26.62\%)
-and this variation persisted after chimera removal using VSEARCH. We
-also observed positive correlations between chimera prevalence and the
-number of OTUs with Platinum having the highest (R\textsuperscript{2} =
-0.974) and Kappa having the worst (R\textsuperscript{2} = 0.478). When
-analyzing mock community samples the variation in the number of OTUs
-detected by the polymerases was observable as early as 20 cycles
-(P-value = 0.002). There also was a large range in the number of OTUs
-amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
-Phusion = 14 - 73). When analyzing fecal samples we observed that the
-range in the number of OTUs detected was not consistent between HiFi DNA
-polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
-Phusion = 84 - 136). Additionally, the median number of OTUs varied by
-HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001).
-Random Forest models were most successful at classifying individual
-donor samples rather than polymerase or number of cycles used (P-value
-\(\leqslant\) 5.49e-07). Additionally, the most important OTUs in the
-polymerase and number of cycle models were not the most important in the
-individual donor sample model.
+\textbf{Results.} In addition to the total counts for specific taxa
+varying by polymerase, we found noticeable differences in relative
+abundance based on high and low GC content (P-value \(\leqslant\) 0.04).
+Chimera prevalence in mock communities varied by polymerase with
+differences being most notable at 35 cycles (Kappa = 5.71\% (median)
+versus Platinum = 26.62\%) and this variation persisted after chimera
+removal using VSEARCH. We also observed positive correlations between
+chimera prevalence and the number of OTUs with Platinum having the
+highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
+(R\textsuperscript{2} = 0.478). When analyzing mock community samples
+the variation in the number of OTUs detected by the polymerases was
+observable as early as 20 cycles (P-value = 0.002). There also was a
+large range in the number of OTUs amplified by the polymerases at 35
+cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73). When analyzing
+fecal samples we observed that the range in the number of OTUs detected
+was not consistent between HiFi DNA polymerases (eg. at 35 cycles
+Accuprime = 84 - 106 (min - max) versus Phusion = 84 - 136).
+Additionally, the median number of OTUs varied by HiFi DNA polymerase
+used at 35 cycles (P-value \textless{} 0.0001). Random Forest models
+were most successful at classifying individual donor samples rather than
+polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
+Additionally, the most important OTUs in the polymerase and number of
+cycle models were not the most important in the individual donor sample
+model.
 
@@
 small with respect to the biological differences between indivdiuals.
-Additionally, our observations were not able to identify GC-based
-differences in amplification based on the number of cycles used.
 Collectively, these results provide evidence that a real biological
@@
 selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
-2017). Differences in the total number of OTUs were analyzed using an
-ANOVA with a tukey post-hoc test. For the comparison of the number of
-OTUS in fecal samples the data was normalized to each individual by
-cycle number to account for the biological variation. Bray-Curtis
-distance matrices were generated using mothur. The distance matrix data
-was analyzed using PERMANOVA with the vegan package (v 2.4.5) (Oksanen
-et al., 2017) and Kruskal-Wallis tests within R. The Random Forest
-models were run using the caret package (v 6.0.78) (Jed Wing et al.,
-2017). A total of 100 10-fold CV runs on different 80/20 splits of the
-data was run to generate a range of the Logloss value. The probability
-of a correct call was obtained from this Logloss value by taking the
-negative natural logrithim. For both error and chimera analysis, samples
-were tested using Kruskal-Wallis with a Dunns post-hoc test. Where
-applicable, correction for multiple comparison utilized the
-Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).
+2017). High and low GC content was determined based on the median GC
+percentage of either the V4 region or the whole genome of the bacterial
+species used in the mock community. Differences in the total number of
+OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
+comparison of the number of OTUS in fecal samples the data was
+normalized to each individual by cycle number to account for the
+biological variation. Bray-Curtis distance matrices were generated using
+mothur. The distance matrix data was analyzed using PERMANOVA with the
+vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
+within R. The Random Forest models were run using the caret package (v
+6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
+different 80/20 splits of the data was run to generate a range of the
+Logloss value. The probability of a correct call was obtained from this
+Logloss value by taking the negative natural logrithim. For both error
+and chimera analysis, samples were tested using Kruskal-Wallis with a
+Dunns post-hoc test. Where applicable, correction for multiple
+comparison utilized the Benjamini-Hochberg method (Benjamini \&
+Hochberg, 1995).
 
@@
 \subsection{Results}\label{results}
+
+***16S rRNA gene sequencing counts do not differ by GC \url{content:***}
+Although the 16S rRNA gene sequencing counts varied by polymerase used,
+there was no difference between V4 16S rRNA genes with high and low GC
+content at every amplification cycle {[}Figure 1 \& Table S1{]}.
+Additionally, there was also no difference in 16S rRNA gene sequencing
+counts between high and low GC content genomes {[}Table S2{]}.
+
+\textbf{\emph{Sequence error varies by polymerase and is consistent
+across the number of cycles used:}} The median error rate varied by
+polymerase, with Kappa having the highest error rate of all the
+polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
+majority of the differences across the number of cycles was between
+Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
+sequences with at least one error was also polymerase dependent, with
+the majority of differences being between Kappa or Accuprime and the
+other polymerases {[}Table S3 \& S4{]}. These differences in error rates
+were not due to polymerase dependent differences in base substitution
+rate {[}Figure S1{]}. Collectively, the results suggest that sequence
+error is dependent on polymerase, persists across the number of cycles
+used, and are not due to any bias towards a specific base substitution.
+
+\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
+and correlate with the number of OTUs:}} Based on the previous results,
+we examined whether chimeras also were dependent on polymerase and
+whether this could affect the number of OTUs. We observed significant
+differences in the chimera prevalence based on polymerase across the
+number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
+Differences in chimera prevalence between Platinum and all other
+polymerases accounted for the majority of these differences {[}Table
+S4{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+of all polymerases regardless of whether chimera removal with VSEARCH
+was used {[}Figure 5A \& 5B{]}. The number of cycles used clearly
+increased chimera prevalence for all polymerases but the rate of
+increase differed {[}Figure 5A \& 5B{]}. Additionally, there was a
+plateau in the total percent of chimeras that were removed that was
+similar for all polymerases {[}Figure 5C{]}. A positive correlation was
+observed between chimeric sequences and the number of OTUs for all
+polymerases {[}Figure 6{]}. This positive correlation was strongest for
+Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
+This data suggests that chimera prevalence depends on polymerase, is
+made worse by increasing the number of cycles, and confirms that the
+number of OTUs is dependent on the prevalence of these chimeric
+sequences.
 
@@
 generate the sequences.
-
-\textbf{\emph{Sequence error varies by polymerase and is consistent
-across the number of cycles used:}} The median error rate varied by
-polymerase, with Kappa having the highest error rate of all the
-polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
-majority of the differences across the number of cycles was between
-Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
-sequences with at least one error was also polymerase dependent, with
-the majority of differences being between Kappa or Accuprime and the
-other polymerases {[}Table S3 \& S4{]}. These differences in error rates
-were not due to polymerase dependent differences in base substitution
-rate {[}Figure S1{]}. Collectively, the results suggest that sequence
-error is dependent on polymerase, persists across the number of cycles
-used, and are not due to any bias towards a specific base substitution.
-
-\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
-and correlate with the number of OTUs:}} Based on the previous results,
-we examined whether chimeras also were dependent on polymerase and
-whether this could affect the number of OTUs. We observed significant
-differences in the chimera prevalence based on polymerase across the
-number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
-Differences in chimera prevalence between Platinum and all other
-polymerases accounted for the majority of these differences {[}Table
-S4{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
-of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 5A \& 5B{]}. The number of cycles used clearly
-increased chimera prevalence for all polymerases but the rate of
-increase differed {[}Figure 5A \& 5B{]}. Additionally, there was a
-plateau in the total percent of chimeras that were removed that was
-similar for all polymerases {[}Figure 5C{]}. A positive correlation was
-observed between chimeric sequences and the number of OTUs for all
-polymerases {[}Figure 6{]}. This positive correlation was strongest for
-Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
-This data suggests that chimera prevalence depends on polymerase, is
-made worse by increasing the number of cycles, and confirms that the
-number of OTUs is dependent on the prevalence of these chimeric
-sequences.
 

### 2018-07-05 submission/manuscript.Rmd
update supplemental with gc table

@@
 
-***16S rRNA gene sequencing counts do not differ by GC content:*** Although the 16S rRNA gene sequencing counts varied by polymerase used, there was no difference between V4 16S rRNA genes with high and low GC content at every amplification cycle [Figure 1 & Table S1].
-Additionally, there was also no difference in 16S rRNA gene sequencing counts between high and low GC content genomes [Table S2].
+***16S rRNA gene sequencing counts differ by GC content:*** There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1, Table S1 & S2].
 

### 2018-07-05 submission/supplemental.Rmd
change title
update supplemental with gc table

@@
 
-# Error Introduced into 16S rRNA Gene Sequencing Results Varies by High Fidelity DNA Polymerase Used
+# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase
 
@@
 
-**Table S1: ANOVA results of differences in number of OTUs between polymerases in fecal and mock samples**
+**Table S1: Wilcoxson Rank Sum Results of High versus Low GC groups based on the median percent of the V4 hypervariable region or the whole genome**
 
 ```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}
+
+mock_gc_wlcox <- as.data.frame(read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv")) %>% 
+  arrange(p.value) %>% 
+  mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(p.value:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  mutate(test_type = "V4 Region") %>% 
+  bind_rows(
+    as.data.frame(
+    read_csv("../data/process/tables/whole_genome_gc_wilcox_test_summary.csv")) %>% 
+    arrange(p.value) %>% 
+    mutate_at(vars(statistic), funs(round(., digits = 2))) %>% 
+    mutate_at(vars(p.value:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    mutate(test_type = "Whole Genome")) %>% 
+  select(cycles, taq, statistic, p.value, bh, test_type) %>% 
+  filter(cycles != "fifteen") %>% 
+  mutate(cycles = case_when(
+    cycles == "twenty" ~ "20x", 
+    cycles == "twenty_five" ~ "25x", 
+    cycles == "thirty" ~ "30x", 
+    cycles == "thirty_five" ~ "35x", 
+    TRUE ~ "uh oh"))
+
+colnames(mock_gc_wlcox) <- c("Number of Cycles", "Polymerase", "Rank Sum Statistic", "P-value", "BH", "Test Type")
+
+kable(mock_gc_wlcox, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S2: ANOVA results of differences in number of OTUs between polymerases in fecal and mock samples**
+
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-**Table S2: Tukey post-hoc results of differences in number of OTUs between polymerases in fecal and mock samples**
+**Table S3: Tukey post-hoc results of differences in number of OTUs between polymerases in fecal and mock samples**
 
-```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-**Table S3: Kruskal-Wallis results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
+**Table S4: Kruskal-Wallis results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 
@@
 
-**Table S4: Dunn's Post-Hoc results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
+**Table S5: Dunn's Post-Hoc results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 

### 2018-07-05 submission/manuscript.Rmd
update first part of results with gc

@@
 
-***Human and mock samples:***
+***Human and mock samples.***
 Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
@@
 
-***PCR protocol:***
+***PCR protocol.***
 The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
@@
 
-***Sequence processing:***
+***Sequence processing.***
 The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
@@
 
-***Analysis workflow:***
+***Analysis workflow.***
 To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
@@
 
-***Statistical analysis:***
+***Statistical analysis.***
 All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
@@
 
-***Reproducible methods:***
+***Reproducible methods.***
 The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
@@
 
-***16S rRNA gene sequencing counts differ by GC content:*** There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1, Table S1 & S2].
+***Observed relative abundance differs by GC content.***
+There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1 & Table S1].
 
-***Sequence error varies by polymerase and is consistent across the number of cycles used:***
+***Sequence error varies by polymerase and is consistent across the number of cycles used.***
 The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].

### 2018-07-05 submission/manuscript.tex
update first part of results with gc

@@
 
-\textbf{\emph{Human and mock samples:}} Fecal samples were obtained from
+\textbf{\emph{Human and mock samples.}} Fecal samples were obtained from
 4 individuals who were part of the Enterics Research Investigational
@@
 
-\textbf{\emph{PCR protocol:}} The five different high fidelity DNA
+\textbf{\emph{PCR protocol.}} The five different high fidelity DNA
 polymerases (hereto referred to as polymerases) that were tested
@@
 
-\textbf{\emph{Sequence processing:}} The mothur software program was
+\textbf{\emph{Sequence processing.}} The mothur software program was
 used for all sequence processing steps (Schloss et al., 2009). The
@@
 
-\textbf{\emph{Analysis workflow:}} To adjust for unequal sequencing, all
+\textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
 samples were rarefied to 1000 sequences for downstream analysis.
@@
 
-\textbf{\emph{Statistical analysis:}} All analysis was done with the R
+\textbf{\emph{Statistical analysis.}} All analysis was done with the R
 (v 3.4.4) software package (R Core Team, 2017). Data transformation and
@@
 
-\textbf{\emph{Reproducible methods:}} The code and analysis can be found
+\textbf{\emph{Reproducible methods.}} The code and analysis can be found
 here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
@@
 
-***16S rRNA gene sequencing counts do not differ by GC \url{content:***}
-Although the 16S rRNA gene sequencing counts varied by polymerase used,
-there was no difference between V4 16S rRNA genes with high and low GC
-content at every amplification cycle {[}Figure 1 \& Table S1{]}.
-Additionally, there was also no difference in 16S rRNA gene sequencing
-counts between high and low GC content genomes {[}Table S2{]}.
+\textbf{\emph{Observed relative abundance differs by GC content.}} There
+was a significant difference in relative abundance between high/low GC
+content based on either the V4 16S rRNA gene region or the whole genome
+{[}Figure 1 \& Table S1{]}.
 
 \textbf{\emph{Sequence error varies by polymerase and is consistent
-across the number of cycles used:}} The median error rate varied by
+across the number of cycles used.}} The median error rate varied by
 polymerase, with Kappa having the highest error rate of all the

### 2018-07-06 submission/supplemental.Rmd
change table orders

@@
 
-**Table S2: ANOVA results of differences in number of OTUs between polymerases in fecal and mock samples**
+**Table S2: Kruskal-Wallis results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
-
-fs_anova <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv")) %>% 
-  filter(sub_sample_level == 1000) %>% 
-  select(-sub_sample_level) %>% 
-  mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
-  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-  mutate(sample_type = "Fecal Sample") %>% 
-  bind_rows(
-    as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv")) %>% 
-    mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
-    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-    filter(sub_sample_level == 1000) %>% 
-    select(-sub_sample_level) %>% 
-    mutate(sample_type = "Mock Sample"))
-
-colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sample Type")
-
-kable(fs_anova, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S3: Tukey post-hoc results of differences in number of OTUs between polymerases in fecal and mock samples**
-
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
-
-fs_tukey <- as.data.frame(read_csv("../data/process/tables/fecal_overall_tukey_results.csv")) %>% 
-  mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
-  filter(sub_sample_level == 1000) %>% 
-  select(-sub_sample_level) %>% 
-  rename(padj = `p adj`) %>% 
-  mutate(padj = as.numeric(padj)) %>% 
-  mutate_at(vars(padj), funs(format(., digits = 2, nsmall = 2))) %>% 
-  mutate(sample_type = "Fecal Sample") %>% 
-  bind_rows(
-    as.data.frame(read_csv("../data/process/tables/mock_overall_tukey_results.csv")) %>% 
-    mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
-    filter(sub_sample_level == 1000) %>% 
-    select(-sub_sample_level) %>% 
-    rename(padj = `p adj`) %>% 
-    mutate(padj = as.numeric(padj)) %>% 
-    filter(padj < 0.05) %>% 
-    arrange(padj) %>% 
-    mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
-    arrange(cycle) %>% 
-    mutate(sample_type = "Mock Sample"))
-
-colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sample Type")
-
-kable(fs_tukey, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S4: Kruskal-Wallis results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-**Table S5: Dunn's Post-Hoc results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
+**Table S3: Dunn's Post-Hoc results of differences in error rate, sequences with an error, and chimera prevalence between polymerases in mock samples**
 
-```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 \newpage
+
+**Table S4: ANOVA results of differences in number of OTUs between polymerases in fecal and mock samples**
+
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+
+fs_anova <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv")) %>% 
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level) %>% 
+  mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
+  mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  mutate(sample_type = "Fecal Sample") %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv")) %>% 
+    mutate_at(vars(Sum.Sq:F.value), funs(round(., digits = 2))) %>% 
+    mutate_at(vars(pvalue:bh), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    filter(sub_sample_level == 1000) %>% 
+    select(-sub_sample_level) %>% 
+    mutate(sample_type = "Mock Sample"))
+
+colnames(fs_anova) <- c("DF", "Sum Squares", "Mean Squares", "F value", "P-value", "BH", "Cycle", "Sample Type")
+
+kable(fs_anova, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S5: Tukey post-hoc results of differences in number of OTUs between polymerases in fecal and mock samples**
+
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+
+fs_tukey <- as.data.frame(read_csv("../data/process/tables/fecal_overall_tukey_results.csv")) %>% 
+  mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+  filter(sub_sample_level == 1000) %>% 
+  select(-sub_sample_level) %>% 
+  rename(padj = `p adj`) %>% 
+  mutate(padj = as.numeric(padj)) %>% 
+  mutate_at(vars(padj), funs(format(., digits = 2, nsmall = 2))) %>% 
+  mutate(sample_type = "Fecal Sample") %>% 
+  bind_rows(
+    as.data.frame(read_csv("../data/process/tables/mock_overall_tukey_results.csv")) %>% 
+    mutate_at(vars(diff:upr), funs(round(., digits = 2))) %>% 
+    filter(sub_sample_level == 1000) %>% 
+    select(-sub_sample_level) %>% 
+    rename(padj = `p adj`) %>% 
+    mutate(padj = as.numeric(padj)) %>% 
+    filter(padj < 0.05) %>% 
+    arrange(padj) %>% 
+    mutate_at(vars(padj), funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+    arrange(cycle) %>% 
+    mutate(sample_type = "Mock Sample"))
+
+colnames(fs_tukey) <- c("Difference", "Lower", "Upper", "P Adjusted", "Comparison", "Cycle", "Sample Type")
+
+kable(fs_tukey, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage

### 2018-07-06 code/make_bray_distance_graphs.R
change fecal/mock order

@@ -72,7 +72,7 @@ fecal_graph <- dist_data %>%
              lwd=1, colour="gray") + 
   scale_color_manual(name = "HiFi DNA Polymerase", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  theme_bw() +  coord_cartesian(ylim = c(0, 1)) + ggtitle("A") + 
+  theme_bw() +  coord_cartesian(ylim = c(0, 1)) + ggtitle("B") + 
   labs(x = "Cycle Comparison", y = "Bray-Curtis Index") + 
   annotate("text", label = paste("Fecal Data"), x = 4.3, y = 1.0, size = 3.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
@@ -94,7 +94,7 @@ mock_graph <- mock_dist_data %>%
              lwd=1, colour="gray") + 
   scale_color_manual(name = "HiFi DNA Polymerase", 
                      values = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")) + 
-  theme_bw() +  coord_cartesian(ylim = c(0, 1)) + ggtitle("B") + 
+  theme_bw() +  coord_cartesian(ylim = c(0, 1)) + ggtitle("A") + 
   labs(x = "Cycle Comparison", y = "Bray-Curtis Index") + 
   annotate("text", label = paste("Mock Data"), x = 3.35, y = 1.0, size = 3.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
@@ -108,6 +108,6 @@ mock_graph <- mock_dist_data %>%
         legend.key = element_blank(), 
         legend.background = element_rect(color = "black"))
 
-combined_graph <- grid.arrange(fecal_graph, mock_graph, ncol = 1, nrow = 2)
+combined_graph <- grid.arrange(mock_graph, fecal_graph, ncol = 1, nrow = 2)
 
 ggsave("results/figures/Figure4.pdf", combined_graph, width = 6, height = 6.5, dpi = 300)

### 2018-07-06 submission/supplemental.Rmd
add new table

@@
 \newpage
+
+**Table S6: Dunns post-hoc results of differences in model classification success using 16S rRNA gene sequencing data**
+
+```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+
+rf_sig_table <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv") %>% 
+  mutate_at(vars(chi2:bh),  funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  separate(var1, c("sample_type1", "model1"), sep = "_") %>% 
+  separate(var2, c("sample_type2", "model2"), sep = "_") %>% 
+  mutate(model1 = case_when(
+    model1 == "sample" ~ "individual", 
+    TRUE ~ model1), 
+    model2 = case_when(
+      model2 == "sample" ~ "individual", 
+      TRUE ~ model2))
+
+colnames(rf_sig_table) <- c("Chi Sq", "Z", "P-value", "BH", "Sample Type 1", "Model 1", "Sample Type 2", "Model 2")
+
+kable(rf_sig_table, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage

### 2018-07-06 submission/manuscript.Rmd
add result updates

@@
 
+# Read in the taxa data
+fecal_otu_tests <- read_csv("../data/process/tables/kruskal_otu_polymerase_fecal_sample_diffs_summary.csv") %>% 
+  arrange(p.value)
+
+mock_otu_tests_kruskal <- read_csv("../data/process/tables/kruskal_otu_polymerase_mock_diffs_summary.csv") %>% 
+  arrange(bh)
+
+mock_otu_tests_dunns <- read_csv("../data/process/tables/dunns_sig_otu_polymerase_mock_diffs_summary.csv")
+
 # Read in RF model data
@@
 At each of these stages error can be introduced that will change the underlying bacterial community composition results.
-In this study we focus on the amplification step's contribution to this overall error.
+In this study, we focus on the amplification step's contribution to this overall error.
 To accomplish this we assessed 16S rRNA gene sequencing results in human fecal and mock community samples after using different high fidelity (HiFi) DNA polymerases and number of amplification cycles.
@@
 We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles.
-Next, differences in the number of OTUs and taxa was examined based on the polymerase and number of cycles used.
+Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used.
 Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles.
+Based on these findings individual taxa differences based on polymerase and number of cycles was investigated.
 Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor.
-We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models also were the most important in the model for individual donors.
+We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models were also important in the model for individual donors.
 
-**Results.** In addition to the total counts for specific taxa varying by polymerase, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`).
+**Results.** Predictably, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`).
 Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
@@
 There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases (eg.
-at 35 cycles Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-Additionally, the median number of OTUs varied by HiFi DNA polymerase used at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+When analyzing fecal samples we observed similar differences in the number of OTUs.
+However, the median number of OTUs only varied by HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
 Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
@@
 
-***Observed relative abundance differs by GC content.***
+***Mock community relative abundance differs by GC content.***
 There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1 & Table S1].
+However, using only the GC content of the V4 region resulted in less differences than when using the whole genome [Table S1].
+The differences between high and low GC content groups were significant as early as 20 cycles in all polymerases except Q5 [Table S1].
+The highest (*Staphylococcus*) and lowest (*Pseudomonas*) GC content bacteria were the most divergent  from the expected relative abundance of 0.12 [Figure 1].
+These observations suggest that the high fidelity DNA polymerases tested do not prevent GC-based amplification error.
 
 ***Sequence error varies by polymerase and is consistent across the number of cycles used.***
-The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3].
-The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4].
-The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4].
+The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 2A & Table S2].
+Thus, the majority of the differences observed across the number of cycles was between Kappa and other polymerases [Figure 2A & Table S3].
+The total sequences that contained at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S2 & S3].
 These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
@@
 
-***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs:***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the number of OTUs.
-We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S3].
-Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S4].
-Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B].
-The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 5A & 5B].
-Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C].
-A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6].
-This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6].
+***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs.***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
+We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2].
+Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3].
+Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C].
+The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C].
+Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D].
+A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E].
+This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E].
 This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 
-***The number of OTUs generated are dependent on polymerase used:***
-Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used (e.g.
-Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1].
-Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g.
-Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 1].
-There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1], but a Tukey post-hoc test failed to identify which specific polymerase groups were different (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2].
-The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2].
-The closest the polymerases came to the total of `r pull(ref_mock_otu_num, numOtus)` OTUs created by the mock reference 16S sequences was at 25 and 30 cycles [Figure 2].
-Regardless of if fecal or mock communities were used, the same polymerases generated high and low number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S2].
-In contrast to the results obtained with fecal samples, differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S1].
-Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S2].
-Based on these observations in fecal and mock communities, it is clear that using different polymerases result in a different total number of OTUs within a sample.
+***The number of OTUs generated are dependent on polymerase used.***
+In addition to the sequence error and chimera differences between polymerases, we also investigated whether there were polymerase or number of cycle dependent effects on the observed number of OTUs.
+There was a polymerase dependent difference in the range of the number of OTUs in the mock community samples [Figure 3A].
+The closest any of the polymerases came to a total of `r pull(ref_mock_otu_num, numOtus)` OTUs, created by running the mock reference 16S sequences through mothur processing, was at 25 and 30 cycles [Figure 3A].
+Differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S4].
+Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S5].
+Similarily, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
+Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
+There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g.
+Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
+Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerase groups (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5].
+Overall, mock communities showed a noticeably larger difference between polymerases in the number of OTUs than actual fecal samples [Figure 3 & Table S4-S5].
+This was consistent across the number of amplification cycles tested [Figure 3].
+Based on these observations, it is clear that larger differences in total number of OTUs based on polymerases are found in mock versus fecal communities.
 
-***The bacterial communities generated by the polymerases were similar but varied by number of cycles:***
-Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3].
+***The bacterial communities generated by the polymerases were similar but varied by number of cycles.***
+Next, we investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
+We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A].
+With respect to fecal samples, within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
 Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
-For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
+Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
 There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
-the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A].
+the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B].
 Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Similar trends were observed in the mock community but none were significant (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
 Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
+
+***Polymerase dependent differences in taxa are more pronounced in mock versus fecal samples.***
+Finding minimal differences in overall bacterial community composision, we next examined if specific taxa vary depending on polymerase and number of cycles.
+Further investigation of mock community samples found that OTUs taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
+These specific OTUs were all members of the mock community [Table S7].
+The majority of the differences between taxa were due to differences between Kappa and the other polymerases [Table S7].
+When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
+These results suggest that taxa specific differences in mock community samples do not carry over to actual fecal community samples.
+
+***Variation due to polymerase and number of cycles is dwarfed by the biological variation of an individual.***
+Using 16S rRNA gene sequencing data there was a large disparity in a model's ability to correctly classify samples based on polymerase used, number of cycles, and individual [Figure 5A].
+The model classification success for an individual was significantly better than all other models tested [Figure 5A & Table S8].
+Additionally, the top 10 most impotant variables to the polymerase and number of cycle models were not the most important variables in the model classifying individuals [Figure 5B].
+This observation shows that OTUs that change due to the number of cycles and polymerase used are not the same as OTUs that differentiate between indvidiuals [Figure 5B].
+Although there are a variety of errors associated with the amplification process, they are smaller than the bacterial community variation between indviduals.
 

### 2018-07-06 submission/manuscript.tex
add result updates

@@
 steps. At each of these stages error can be introduced that will change
-the underlying bacterial community composition results. In this study we
-focus on the amplification step's contribution to this overall error. To
-accomplish this we assessed 16S rRNA gene sequencing results in human
+the underlying bacterial community composition results. In this study,
+we focus on the amplification step's contribution to this overall error.
+To accomplish this we assessed 16S rRNA gene sequencing results in human
 fecal and mock community samples after using different high fidelity
@@
 polymerase and number of cycles. Next, differences in the number of OTUs
-and taxa was examined based on the polymerase and number of cycles used.
+was examined based on the polymerase and number of cycles used.
 Additionally, differences in the bacterial community composition by the
 Bray-Curtis index also was assessed based on polymerase and number of
-cycles. Finally, Random Forest models were created to test whether the
-bacterial community was better at classifying polymerases, number of
-cycles, or individual donor. We also assessed whether the most important
-taxa in the polymerase and number of cycle Random Forest models also
-were the most important in the model for individual donors.
+cycles. Based on these findings individual taxa differences based on
+polymerase and number of cycles was investigated. Finally, Random Forest
+models were created to test whether the bacterial community was better
+at classifying polymerases, number of cycles, or individual donor. We
+also assessed whether the most important taxa in the polymerase and
+number of cycle Random Forest models were also important in the model
+for individual donors.
 
-\textbf{Results.} In addition to the total counts for specific taxa
-varying by polymerase, we found noticeable differences in relative
-abundance based on high and low GC content (P-value \(\leqslant\) 0.04).
-Chimera prevalence in mock communities varied by polymerase with
-differences being most notable at 35 cycles (Kappa = 5.71\% (median)
-versus Platinum = 26.62\%) and this variation persisted after chimera
-removal using VSEARCH. We also observed positive correlations between
-chimera prevalence and the number of OTUs with Platinum having the
-highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
-(R\textsuperscript{2} = 0.478). When analyzing mock community samples
-the variation in the number of OTUs detected by the polymerases was
-observable as early as 20 cycles (P-value = 0.002). There also was a
-large range in the number of OTUs amplified by the polymerases at 35
-cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73). When analyzing
-fecal samples we observed that the range in the number of OTUs detected
-was not consistent between HiFi DNA polymerases (eg. at 35 cycles
-Accuprime = 84 - 106 (min - max) versus Phusion = 84 - 136).
-Additionally, the median number of OTUs varied by HiFi DNA polymerase
-used at 35 cycles (P-value \textless{} 0.0001). Random Forest models
-were most successful at classifying individual donor samples rather than
+\textbf{Results.} Predictably, we found noticeable differences in
+relative abundance based on high and low GC content (P-value
+\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
+polymerase with differences being most notable at 35 cycles (Kappa =
+5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
+after chimera removal using VSEARCH. We also observed positive
+correlations between chimera prevalence and the number of OTUs with
+Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
+having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
+community samples the variation in the number of OTUs detected by the
+polymerases was observable as early as 20 cycles (P-value = 0.002).
+There also was a large range in the number of OTUs amplified by the
+polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
+When analyzing fecal samples we observed similar differences in the
+number of OTUs. However, the median number of OTUs only varied by HiFi
+DNA polymerase used at 35 cycles and not at 20 cycles like the mock
+community (P-value \textless{} 0.0001). Random Forest models were most
+successful at classifying individual donor samples rather than
 polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
@@
 
-\textbf{\emph{Observed relative abundance differs by GC content.}} There
-was a significant difference in relative abundance between high/low GC
-content based on either the V4 16S rRNA gene region or the whole genome
-{[}Figure 1 \& Table S1{]}.
+\textbf{\emph{Mock community relative abundance differs by GC content.}}
+There was a significant difference in relative abundance between
+high/low GC content based on either the V4 16S rRNA gene region or the
+whole genome {[}Figure 1 \& Table S1{]}. However, using only the GC
+content of the V4 region resulted in less differences than when using
+the whole genome {[}Table S1{]}. The differences between high and low GC
+content groups were significant as early as 20 cycles in all polymerases
+except Q5 {[}Table S1{]}. The highest (\emph{Staphylococcus}) and lowest
+(\emph{Pseudomonas}) GC content bacteria were the most divergent from
+the expected relative abundance of 0.12 {[}Figure 1{]}. These
+observations suggest that the high fidelity DNA polymerases tested do
+not prevent GC-based amplification error.
 
@@
 polymerase, with Kappa having the highest error rate of all the
-polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
-majority of the differences across the number of cycles was between
-Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
-sequences with at least one error was also polymerase dependent, with
-the majority of differences being between Kappa or Accuprime and the
-other polymerases {[}Table S3 \& S4{]}. These differences in error rates
-were not due to polymerase dependent differences in base substitution
-rate {[}Figure S1{]}. Collectively, the results suggest that sequence
-error is dependent on polymerase, persists across the number of cycles
-used, and are not due to any bias towards a specific base substitution.
+polymerases across the number of cycles {[}Figure 2A \& Table S2{]}.
+Thus, the majority of the differences observed across the number of
+cycles was between Kappa and other polymerases {[}Figure 2A \& Table
+S3{]}. The total sequences that contained at least one error was also
+polymerase dependent, with the majority of differences being between
+Kappa or Accuprime and the other polymerases {[}Table S2 \& S3{]}. These
+differences in error rates were not due to polymerase dependent
+differences in base substitution rate {[}Figure S1{]}. Collectively, the
+results suggest that sequence error is dependent on polymerase, persists
+across the number of cycles used, and are not due to any bias towards a
+specific base substitution.
 
 \textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
-and correlate with the number of OTUs:}} Based on the previous results,
+and correlate with the number of OTUs.}} Based on the previous results,
 we examined whether chimeras also were dependent on polymerase and
-whether this could affect the number of OTUs. We observed significant
-differences in the chimera prevalence based on polymerase across the
-number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
-Differences in chimera prevalence between Platinum and all other
+whether this could affect the total number of OTUs. We observed
+significant differences in the chimera prevalence based on polymerase
+across the number of cycles used (P-value \textless{} 0.05) {[}Table
+S2{]}. Differences in chimera prevalence between Platinum and all other
 polymerases accounted for the majority of these differences {[}Table
-S4{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
 of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 5A \& 5B{]}. The number of cycles used clearly
+was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
 increased chimera prevalence for all polymerases but the rate of
-increase differed {[}Figure 5A \& 5B{]}. Additionally, there was a
+increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
 plateau in the total percent of chimeras that were removed that was
-similar for all polymerases {[}Figure 5C{]}. A positive correlation was
+similar for all polymerases {[}Figure 2D{]}. A positive correlation was
 observed between chimeric sequences and the number of OTUs for all
-polymerases {[}Figure 6{]}. This positive correlation was strongest for
-Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
+polymerases {[}Figure 2E{]}. This positive correlation was strongest for
+Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
 This data suggests that chimera prevalence depends on polymerase, is
@@
 \textbf{\emph{The number of OTUs generated are dependent on polymerase
-used:}} Differences in the range of the number of OTUs detected for
-fecal samples is dependent on the polymerase used (e.g.~Accuprime at 35
-cycles = 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure
-1{]}. Additionally, there is a trend for lower number of cycles (15-20)
-to result in a reduced range in the number of OTUs versus higher number
-of cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15
-cycles = 10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
-There is an overall difference in the number of OTUs detected within
-fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
-16.35, P-value = 9.7e-05) {[}Table S1{]}, but a Tukey post-hoc test
-failed to identify which specific polymerase groups were different
-(P-value \textgreater{} 0.05) {[}Table S2{]}. The polymerase dependent
-difference in the range of the number of OTUs also was observed in the
-mock community samples {[}Figure 2{]}. The closest the polymerases came
-to the total of 8 OTUs created by the mock reference 16S sequences was
-at 25 and 30 cycles {[}Figure 2{]}. Regardless of if fecal or mock
-communities were used, the same polymerases generated high and low
-number of OTUs and this was consistent across the number of cycles used
-{[}Figure 1-2 \& Table S1-S2{]}. In contrast to the results obtained
-with fecal samples, differences between polymerases for the number of
-OTUs created were observed as early as 20 cycles in the mock community
-(F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc
-test, the majority of differences for the number of OTUs detected in the
-mock community was largely due to Kappa and Platinum versus the other
-polymerases across the number of cycles used {[}Table S2{]}. Based on
-these observations in fecal and mock communities, it is clear that using
-different polymerases result in a different total number of OTUs within
-a sample.
+used.}} In addition to the sequence error and chimera differences
+between polymerases, we also investigated whether there were polymerase
+or number of cycle dependent effects on the observed number of OTUs.
+There was a polymerase dependent difference in the range of the number
+of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
+of the polymerases came to a total of 8 OTUs, created by running the
+mock reference 16S sequences through mothur processing, was at 25 and 30
+cycles {[}Figure 3A{]}. Differences between polymerases for the number
+of OTUs created were observed as early as 20 cycles in the mock
+community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
+Tukey post-hoc test, the majority of differences for the number of OTUs
+detected in the mock community was largely due to Kappa and Platinum
+versus the other polymerases across the number of cycles used {[}Table
+S5{]}. Similarily, differences in the range of the number of OTUs
+detected for fecal samples was dependent on the polymerase used
+(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
+- 136) {[}Figure 3B{]}. There was also a trend for lower number of
+cycles (15-20) to result in a reduced range in the number of OTUs versus
+higher number of cycles (25, 30, and 35) for all polymerases
+(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
+136) {[}Figure 3B{]}. Unlike mock community samples, an overall
+difference in the number of OTUs was only detected between polymerases
+at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
+9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
+any differences between specific polymerase groups (P-value
+\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
+noticeably larger difference between polymerases in the number of OTUs
+than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
+consistent across the number of amplification cycles tested {[}Figure
+3{]}. Based on these observations, it is clear that larger differences
+in total number of OTUs based on polymerases are found in mock versus
+fecal communities.
 
 \textbf{\emph{The bacterial communities generated by the polymerases
-were similar but varied by number of cycles:}} Within each respective
+were similar but varied by number of cycles.}} Next, we investigated the
+effect polymerases and number of cycles had on the bacterial community
+composition in more detail by using the Bray-Curtis index. We observed
+no difference in the bacterial community composition between any measure
+tested when using the mock community (P-value \textgreater{} 0.05)
+{[}Figure 4A{]}. With respect to fecal samples, within each respective
 5-cycle increment comparison there was no difference in Bray-Curtis
 index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
-3{]}. Using PERMANOVA to test for community differences based on any of
+4B{]}. Using PERMANOVA to test for community differences based on any of
 the number of cycles within polymerases, only Phusion had cycle
-dependent differences (P-value = 0.03). For fecal samples, Phusion was
-one of two polymerases that had enough sequences to be rarefied to 1000
+dependent differences (P-value = 0.03). Additionally, Phusion was one of
+two polymerases that had enough sequences to be rarefied to 1000
 sequences at 15 cycles. There was an overall decrease in Bray-Curtis
@@
 compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
-0.01) {[}Figure 3A{]}. Using a Dunn's post-hoc test the 15 cycle versus
+0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus
 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
 Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
-(P-value \textless{} 0.05). Similar trends were observed in the mock
-community but none were significant (P-value \textgreater{} 0.05)
-{[}Figure 3B{]}. Overall, these data suggest that the number of cycles
-can change the bacterial community independent of the polymerase used to
-generate the sequences.
+(P-value \textless{} 0.05). Overall, these data suggest that the number
+of cycles can change the bacterial community independent of the
+polymerase used to generate the sequences.
+
+\textbf{\emph{Polymerase dependent differences in taxa are more
+pronounced in mock versus fecal samples.}} Finding minimal differences
+in overall bacterial community composision, we next examined if specific
+taxa vary depending on polymerase and number of cycles. Further
+investigation of mock community samples found that OTUs taxonomically
+classified to \emph{Salmonella}, \emph{Escherichia},
+\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
+polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
+members of the mock community {[}Table S7{]}. The majority of the
+differences between taxa were due to differences between Kappa and the
+other polymerases {[}Table S7{]}. When differences in OTUs based on
+polymerase were investigated for fecal samples, no differences could be
+found (P-value \(\geqslant\) 0.01, corrected P-value = 1). These results
+suggest that taxa specific differences in mock community samples do not
+carry over to actual fecal community samples.
+
+\textbf{\emph{Variation due to polymerase and number of cycles is
+dwarfed by the biological variation of an individual.}} Using 16S rRNA
+gene sequencing data there was a large disparity in a model's ability to
+correctly classify samples based on polymerase used, number of cycles,
+and individual {[}Figure 5A{]}. The model classification success for an
+individual was significantly better than all other models tested
+{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most impotant
+variables to the polymerase and number of cycle models were not the most
+important variables in the model classifying individuals {[}Figure
+5B{]}. This observation shows that OTUs that change due to the number of
+cycles and polymerase used are not the same as OTUs that differentiate
+between indvidiuals {[}Figure 5B{]}. Although there are a variety of
+errors associated with the amplification process, they are smaller than
+the bacterial community variation between indviduals.
 

### 2018-07-06 submission/supplemental.Rmd
add result updates

@@
 
-**Table S6: Dunns post-hoc results of differences in model classification success using 16S rRNA gene sequencing data**
+**Table S6: Kruskal-Wallis results of significant differences in OTU in mock community samples**
 
 ```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+
+kruskal_otus <- read_csv("../data/process/tables/kruskal_otu_polymerase_mock_diffs_summary.csv") %>% 
+  mutate(sample_type = "mock") %>% 
+  filter(bh < 0.05) %>% 
+  arrange(cycles, p.value) %>% 
+  mutate_at(vars(p.value:bh),  funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
+  mutate(genus = str_replace_all(genus, "_unclassified", ""))
+
+colnames(kruskal_otus) <- c("Cycles", "OTU", "Lowest Taxa ID", "Statistic", "P-value", "BH", "Sample Type")
+
+kable(kruskal_otus, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage
+
+**Table S7: Significant Dunns post-hoc results of differences between polymerases for specific taxa**
+
+\footnotesize
+
+```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}
+
+dunn_otus <- read_csv("../data/process/tables/dunns_sig_otu_polymerase_mock_diffs_summary.csv") %>% 
+  arrange(bh) %>% 
+  mutate_at(vars(chi2:bh),  funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))
+
+colnames(dunn_otus) <- c("Chi Sq", "Z", "P-value", "BH", "Comparison", "Cycles", "OTU", "Taxa", "Part of Mock")
+
+kable(dunn_otus, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\normalsize
+
+\newpage
+
+**Table S8: Dunns post-hoc results of differences in model classification success using 16S rRNA gene sequencing data**
+
+```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}
 

### 2018-07-06 submission/manuscript.Rmd
fix typo and update sentence

@@
 ***Polymerase dependent differences in taxa are more pronounced in mock versus fecal samples.***
-Finding minimal differences in overall bacterial community composision, we next examined if specific taxa vary depending on polymerase and number of cycles.
+Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles.
 Further investigation of mock community samples found that OTUs taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
-These specific OTUs were all members of the mock community [Table S7].
+These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7].
 The majority of the differences between taxa were due to differences between Kappa and the other polymerases [Table S7].
-When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
+When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found at any number of cycles (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
 These results suggest that taxa specific differences in mock community samples do not carry over to actual fecal community samples.

### 2018-07-06 submission/manuscript.tex
fix typo and update sentence

@@
 pronounced in mock versus fecal samples.}} Finding minimal differences
-in overall bacterial community composision, we next examined if specific
+in overall bacterial community composition, we next examined if specific
 taxa vary depending on polymerase and number of cycles. Further
@@
 polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
-members of the mock community {[}Table S7{]}. The majority of the
-differences between taxa were due to differences between Kappa and the
-other polymerases {[}Table S7{]}. When differences in OTUs based on
-polymerase were investigated for fecal samples, no differences could be
-found (P-value \(\geqslant\) 0.01, corrected P-value = 1). These results
-suggest that taxa specific differences in mock community samples do not
-carry over to actual fecal community samples.
+members of the mock community and the differences were most pronounced
+at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
+between taxa were due to differences between Kappa and the other
+polymerases {[}Table S7{]}. When differences in OTUs based on polymerase
+were investigated for fecal samples, no differences could be found at
+any number of cycles (P-value \(\geqslant\) 0.01, corrected P-value =
+1). These results suggest that taxa specific differences in mock
+community samples do not carry over to actual fecal community samples.
 

### 2018-07-10 submission/manuscript.Rmd
update discussion

@@
 
-Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-2 & 4-6].
-Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences.
-In addition to the reported polymerase dependent differences, the bacterial community also varied by the number of cycles used.
-These differences are important because many diversity metrics rely on calculations (e.g.
-richness) that are dependent on the number of OTUs.
-Since the number of OTUs are dependent on error rate and chimera prevalence the diversity metrics can vary simply by changing the polymerase used.
-Based on these observations, metrics that measure diversity using calculations such as richness depend on the polymerase, but this may not be the case for all diversity metrics.
+Previous studies using regular DNA polymerase have reported GC dependent amplification differences [@polz_bias_1998] and we confirm that these differences still persist in high fidelity DNA polymerase when analyzing mock community samples [Figure 1].
+Additionally, these specific polymerases also have different sequence error rate and chimera prevalence in these samples [Figure 2].
+Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a sample [Figure 2E].
+The total number of OTUs also varied by polymerase used for mock community samples [Fiugre 3A].
+The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for mock community and fecal samples [Figure 3].
+However, these differences did not translate to large bacterial community composition changes based on polymerase [Figure 4].
+Even though there were bacterial community composition changes based on the number of cycles they occured independently of the polymerase used [Figure 4].
+Additionally, there were no differences observed within the number of cycles between any of the polymerases for fecal samples.
+When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing the model performance for individuals was significantly greater than all other models tested [Figure 5A].
+In addition to the better model performance of individuals the most important OTUs to this model were not the most important OTUs in the other models [Figure 5B].
+Our observations show that differences between high fidelity DNA polymerases can be found.
+However, when using 16S rRNA gene sequencing these technical differences are dwarfed by biological differences between individuals.
+
+The small differences observed due to error rate and chimera prevalence may be due to the actual DNA polymerase family being used within the PCR mix.
+DNA polymerases from distinct families have different binding affinities and error correction capacity [@polymerase_Ishino_2014].
+However, based on the observations within our study this is unlikely to be the only contributor.
+Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
+Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent.
+Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence.
+Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.
 
 Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community.
-Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 3].
-However, the distance between cycle comparison groups (e.g.
-20 cycles versus 25 cycles group compared to 30 cycles versus 35 cycles group) was different across polymerase used.
-These results support the idea that cycle number and not the specific polymerase used may have a larger affect on metrics such as Bray-Curtis.
+Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4].
 One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987].
@@
 
-Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family because of the different binding affinities and error correction capacity [@polymerase_Ishino_2014], this is unlikely to be the only contributor.
-Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
-Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent.
-Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
-Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence.
-Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.
-
-Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples.
-However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community.
-Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis.
-All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that want to be answered.
-Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.
+Collectively, our data shows that using distinct high fidelity DNA polymerases will result in a different number of OTUs being detected.
+The differences in the number of OTUs are primarily due to chimera prevalence and sequence error rates that are distinct to the specific polymerase.
+Despite this variation in the number of OTUs, our Random Forest models were able to correctly classify individuals regardless of the number of cycles and polymerase used.
+These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to bacterial community composition differences between individuals will still be consistent.
 
@@
 
-Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing.
-Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics.
-By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.
+Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation between individuals.
+Therefore, if the biological signal of interest is similar to differences in bacterial community composition found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
 

### 2018-07-10 submission/manuscript.tex
update discussion

@@
 
-Our observations support that the number of OTUs, error rate, and
-chimera prevalence depends on polymerase used and that chimera
-prevalence is highly correlated with the number of OTUs {[}Figure 1-2 \&
-4-6{]}. Our observations also suggest that undetected chimeras, not
-removed using standard bioinformatic approaches, cause many of these
-differences. In addition to the reported polymerase dependent
-differences, the bacterial community also varied by the number of cycles
-used. These differences are important because many diversity metrics
-rely on calculations (e.g.~richness) that are dependent on the number of
-OTUs. Since the number of OTUs are dependent on error rate and chimera
-prevalence the diversity metrics can vary simply by changing the
-polymerase used. Based on these observations, metrics that measure
-diversity using calculations such as richness depend on the polymerase,
-but this may not be the case for all diversity metrics.
+Previous studies using regular DNA polymerase have reported GC dependent
+amplification differences (Polz \& Cavanaugh, 1998) and we confirm that
+these differences still persist in high fidelity DNA polymerase when
+analyzing mock community samples {[}Figure 1{]}. Additionally, these
+specific polymerases also have different sequence error rate and chimera
+prevalence in these samples {[}Figure 2{]}. Chimera prevalence in
+particular was strongly associated with the total number of OTUs
+detected in a sample {[}Figure 2E{]}. The total number of OTUs also
+varied by polymerase used for mock community samples {[}Fiugre 3A{]}.
+The range in the number of OTUs observed increased with the number of
+cycles and the change in range was dependent on the polymerase for mock
+community and fecal samples {[}Figure 3{]}. However, these differences
+did not translate to large bacterial community composition changes based
+on polymerase {[}Figure 4{]}. Even though there were bacterial community
+composition changes based on the number of cycles they occured
+independently of the polymerase used {[}Figure 4{]}. Additionally, there
+were no differences observed within the number of cycles between any of
+the polymerases for fecal samples. When Random Forest models were
+created to classify number of cycles, polymerase, and individual based
+on the 16S rRNA gene sequencing the model performance for individuals
+was significantly greater than all other models tested {[}Figure 5A{]}.
+In addition to the better model performance of individuals the most
+important OTUs to this model were not the most important OTUs in the
+other models {[}Figure 5B{]}. Our observations show that differences
+between high fidelity DNA polymerases can be found. However, when using
+16S rRNA gene sequencing these technical differences are dwarfed by
+biological differences between individuals.
 
-Based on polymerase used, our observations generally found no difference
-when using the Bray-Curtis index to measure the bacterial community.
-Specifically, there was no difference in distance between successive
-5-cycle increments within samples between the polymerases when using the
-Bray-Curtis index {[}Figure 3{]}. However, the distance between cycle
-comparison groups (e.g.~20 cycles versus 25 cycles group compared to 30
-cycles versus 35 cycles group) was different across polymerase used.
-These results support the idea that cycle number and not the specific
-polymerase used may have a larger affect on metrics such as Bray-Curtis.
-One possible reason for this outcome is that many of the OTUs generated
-by polymerase dependent error rates and chimera prevalence are likely
-not highly abundant, allowing the Bray-Curtis index to be able to
-successfully down-weight these OTUs (Minchin, 1987). The choice of
-downstream diversity metric could be an important consideration in
-helping to mitigate some of the observed polymerase dependent
-differences in error rate and chimera prevalence. Metrics that solely
-use presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996),
-richness) may be less robust to polymerase dependent error rates and
-chimera prevalence. When choosing a distance metric, careful
-consideration of the biases introduced from the PCR step of the 16S rRNA
-gene sequencing workflow need to be taken into account. One possible way
-to better choose polymerases may be based on the DNA polymerase family
-used in the PCR mixture.
-
-Although the variation in error rate and chimera prevalence may be due
-to the DNA polymerase family because of the different binding affinities
-and error correction capacity (Ishino \& Ishino, 2014), this is unlikely
-to be the only contributor. Within our study the highest and lowest
-chimera prevalence both belonged to a family A polymerase (Platinum and
+The small differences observed due to error rate and chimera prevalence
+may be due to the actual DNA polymerase family being used within the PCR
+mix. DNA polymerases from distinct families have different binding
+affinities and error correction capacity (Ishino \& Ishino, 2014).
+However, based on the observations within our study this is unlikely to
+be the only contributor. Within our study the highest and lowest chimera
+prevalence both belonged to a family A polymerase (Platinum and
 Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
@@
 
-Standardization of the 16S rRNA gene sequencing workflow may partially
-solve this problem by introducing a consistent bias to all samples.
-However, as mentioned in the introduction, many of the choices made to
-standardize the workflow will result in missing important members of the
-microbial community. Therefore, the degree of workflow standardization
-across studies and research groups need to be approached on a study by
-study basis. All aspects of the 16S rRNA gene sequencing workflow should
-be customized for the specific scientific questions that want to be
-answered. Although a diversity of approaches may make reproducibility
-and replicability more difficult it will help reduce systematic biases
-from occurring.
+Based on polymerase used, our observations generally found no difference
+when using the Bray-Curtis index to measure the bacterial community.
+Specifically, there was no difference in distance between successive
+5-cycle increments within samples between the polymerases when using the
+Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
+is that many of the OTUs generated by polymerase dependent error rates
+and chimera prevalence are likely not highly abundant, allowing the
+Bray-Curtis index to be able to successfully down-weight these OTUs
+(Minchin, 1987). The choice of downstream diversity metric could be an
+important consideration in helping to mitigate some of the observed
+polymerase dependent differences in error rate and chimera prevalence.
+Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
+Vargas, 1996), richness) may be less robust to polymerase dependent
+error rates and chimera prevalence. When choosing a distance metric,
+careful consideration of the biases introduced from the PCR step of the
+16S rRNA gene sequencing workflow need to be taken into account. One
+possible way to better choose polymerases may be based on the DNA
+polymerase family used in the PCR mixture.
+
+Collectively, our data shows that using distinct high fidelity DNA
+polymerases will result in a different number of OTUs being detected.
+The differences in the number of OTUs are primarily due to chimera
+prevalence and sequence error rates that are distinct to the specific
+polymerase. Despite this variation in the number of OTUs, our Random
+Forest models were able to correctly classify individuals regardless of
+the number of cycles and polymerase used. These results suggest that
+even though there is a large variation in the high fidelity DNA
+polymerases used across studies, biological variation similar in effect
+to bacterial community composition differences between individuals will
+still be consistent.
 
@@
 
-Our observations show that errors dependent on either high fidelity DNA
-polymerases used or the number of cycles chosen can change diversity
-metrics used in 16S rRNA gene sequencing. Care should be taken when
-choosing a polymerase for 16S rRNA gene surveys because they can
-influence diversity-based metrics. By knowing the inherent bias
-associated with different polymerases it allows for more informed
-choices to be made on how to reduce 16S rRNA gene sequencing-based
-error.
+Although care should always be taken when choosing a polymerase for 16S
+rRNA gene sequencing, our observations show that the differences between
+a variety of polymerases are dwarfed by the actual biological variation
+between individuals. Therefore, if the biological signal of interest is
+similar to differences in bacterial community composition found between
+individuals, then the type of high fidelity DNA polymerase used will
+only minimally change the results.
 

### 2018-07-10 submission/manuscript.Rmd
fix up discussion first paragraph

@@
 
-Previous studies using regular DNA polymerase have reported GC dependent amplification differences [@polz_bias_1998] and we confirm that these differences still persist in high fidelity DNA polymerase when analyzing mock community samples [Figure 1].
-Additionally, these specific polymerases also have different sequence error rate and chimera prevalence in these samples [Figure 2].
-Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a sample [Figure 2E].
-The total number of OTUs also varied by polymerase used for mock community samples [Fiugre 3A].
-The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for mock community and fecal samples [Figure 3].
+Previous studies have reported GC dependent amplification differences [@polz_bias_1998] and we confirm that these differences still persist in high fidelity DNA polymerase when analyzing mock community samples [Figure 1].
+Additionally, these specific polymerases have different sequence error rates and chimera prevalence [Figure 2].
+Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a mock community [Figure 2E].
+This total number of OTUs also varied by polymerase used for these mock community samples [Fiugre 3A].
+The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for both mock community and fecal samples [Figure 3].
 However, these differences did not translate to large bacterial community composition changes based on polymerase [Figure 4].
@@
 Additionally, there were no differences observed within the number of cycles between any of the polymerases for fecal samples.
-When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing the model performance for individuals was significantly greater than all other models tested [Figure 5A].
-In addition to the better model performance of individuals the most important OTUs to this model were not the most important OTUs in the other models [Figure 5B].
+When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing, the model performance for individuals was significantly greater than all other models tested [Figure 5A].
+In addition to the better model performance of individuals, the most important OTUs to this model were not the most important OTUs for the other models tested [Figure 5B].
 Our observations show that differences between high fidelity DNA polymerases can be found.

### 2018-07-10 submission/manuscript.tex
fix up discussion first paragraph

@@
 
-Previous studies using regular DNA polymerase have reported GC dependent
-amplification differences (Polz \& Cavanaugh, 1998) and we confirm that
-these differences still persist in high fidelity DNA polymerase when
-analyzing mock community samples {[}Figure 1{]}. Additionally, these
-specific polymerases also have different sequence error rate and chimera
-prevalence in these samples {[}Figure 2{]}. Chimera prevalence in
-particular was strongly associated with the total number of OTUs
-detected in a sample {[}Figure 2E{]}. The total number of OTUs also
-varied by polymerase used for mock community samples {[}Fiugre 3A{]}.
-The range in the number of OTUs observed increased with the number of
-cycles and the change in range was dependent on the polymerase for mock
-community and fecal samples {[}Figure 3{]}. However, these differences
-did not translate to large bacterial community composition changes based
-on polymerase {[}Figure 4{]}. Even though there were bacterial community
-composition changes based on the number of cycles they occured
-independently of the polymerase used {[}Figure 4{]}. Additionally, there
-were no differences observed within the number of cycles between any of
-the polymerases for fecal samples. When Random Forest models were
-created to classify number of cycles, polymerase, and individual based
-on the 16S rRNA gene sequencing the model performance for individuals
-was significantly greater than all other models tested {[}Figure 5A{]}.
-In addition to the better model performance of individuals the most
-important OTUs to this model were not the most important OTUs in the
-other models {[}Figure 5B{]}. Our observations show that differences
-between high fidelity DNA polymerases can be found. However, when using
-16S rRNA gene sequencing these technical differences are dwarfed by
-biological differences between individuals.
+Previous studies have reported GC dependent amplification differences
+(Polz \& Cavanaugh, 1998) and we confirm that these differences still
+persist in high fidelity DNA polymerase when analyzing mock community
+samples {[}Figure 1{]}. Additionally, these specific polymerases have
+different sequence error rates and chimera prevalence {[}Figure 2{]}.
+Chimera prevalence in particular was strongly associated with the total
+number of OTUs detected in a mock community {[}Figure 2E{]}. This total
+number of OTUs also varied by polymerase used for these mock community
+samples {[}Fiugre 3A{]}. The range in the number of OTUs observed
+increased with the number of cycles and the change in range was
+dependent on the polymerase for both mock community and fecal samples
+{[}Figure 3{]}. However, these differences did not translate to large
+bacterial community composition changes based on polymerase {[}Figure
+4{]}. Even though there were bacterial community composition changes
+based on the number of cycles they occured independently of the
+polymerase used {[}Figure 4{]}. Additionally, there were no differences
+observed within the number of cycles between any of the polymerases for
+fecal samples. When Random Forest models were created to classify number
+of cycles, polymerase, and individual based on the 16S rRNA gene
+sequencing, the model performance for individuals was significantly
+greater than all other models tested {[}Figure 5A{]}. In addition to the
+better model performance of individuals, the most important OTUs to this
+model were not the most important OTUs for the other models tested
+{[}Figure 5B{]}. Our observations show that differences between high
+fidelity DNA polymerases can be found. However, when using 16S rRNA gene
+sequencing these technical differences are dwarfed by biological
+differences between individuals.
 

### 2018-07-11 submission/manuscript.Rmd
add missing noun in first paragraph
fix figure legends

@@
 Even though there were bacterial community composition changes based on the number of cycles they occured independently of the polymerase used [Figure 4].
-Additionally, there were no differences observed within the number of cycles between any of the polymerases for fecal samples.
+Additionally, there were no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.
 When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing, the model performance for individuals was significantly greater than all other models tested [Figure 5A].
@@
 
-**Figure 1: The number and range of OTUs in fecal samples varies by polymerase.** The points represent the median number of OTUs of all four fecal samples.
+**Figure 1: Relative abundance differences due to GC is consistent across polymerases.** At each number of cycle the points and lines represent the median and with the minimum and maximum relative abundance respectively.
+Regardless of the number of cycle used a consistent difference between high and low GC bacteria was observed.
+The dotted line represents the actual relative abundance that each bacterium should be at within the mock community sample.
+
+**Figure 2: Error rate and chimera prevalence vary by polymerase and affect the number of observed OTUs.** A) The error bars represent the 75% interquartile range of the median error rate.
+B) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
+C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.
+D) The total percent of chimeric sequences removed with VSEARCH by cycle number.
+The error bars represent the 75% interquartile range of the median.
+E) Chimera prevalence and the the observed number of OTUs are strongly correlated in mock community samples.
+
+**Figure 3: The number and range of OTUs in mock and fecal samples varies by polymerase.** The points represent the median number of OTUs for the mock samples.
 The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
-The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
+The dotted black line represents the number of OTUs detected when only the references sequences for the mock community are clustered.
+A) The range in the number of OTUs detected in the mock samples increased as cycle number increased.
+This range was also larger for specific HiFi DNA polymerases.
+B) The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
 This increased range also was larger for specific HiFi DNA polymerases.
 
-**Figure 2: The number and range of OTUs in mock samples varies by polymerase.** The points represent the median number of OTUs for the mock samples.
-The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
-The dotted black line represents the number of OTUs detected when only the references sequences for the mock community are clustered.
-The range in the number of OTUs detected in the mock samples increased as cycle number increased.
-This range was also larger for specific HiFi DNA polymerases.
-
-**Figure 3: Lower five-cycle intervals have larger differences than higher five-cycle intervals using Bray-Curtis.** A) within person differences based on the next 5-cycle PCR interval in fecal samples.
-B) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
+**Figure 4: Lower five-cycle intervals have larger differences than higher five-cycle intervals using Bray-Curtis.** A) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
+B) Within person differences based on the next 5-cycle PCR interval in fecal samples.
 The points represent the median Bray-Curtis index for the samples.
@@
 
-**Figure 4: Variation in error rate by polymerase in mock community samples are similar across the number of cycles used.** The error bars represent the 75% interquartile range of the median error rate.
-
-**Figure 5: Chimera prevalence varies by polymerase and increases with higher number of cycles used.** A) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
-C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.
-C) The total percent of chimeric sequences removed with VSEARCH by cycle number.
-The error bars represent the 75% interquartile range of the median.
-
-**Figure 6: Regardless of polymerase used the higher the chimera prevalence the higher the observed number of OTUs in mock community samples.**
+**Figure 5: Biological variation in bacterial community composition due to individual was larger than number of cycles and polymerase used.** A) Using 16S rRNA gene sequencing, the classification of individuals was higher than any other model tested.
+The lines represent the median probability of correct classification and the gray points represent the individual probability from one run out of the 100 10-fold CV runs.
+B) The most important OTUs to the model classifying individuals are not the same as those that are most important for the number of cycles or polymerase used.
 

### 2018-07-11 submission/manuscript.tex
add missing noun in first paragraph
fix figure legends

@@
 based on the number of cycles they occured independently of the
-polymerase used {[}Figure 4{]}. Additionally, there were no differences
-observed within the number of cycles between any of the polymerases for
-fecal samples. When Random Forest models were created to classify number
-of cycles, polymerase, and individual based on the 16S rRNA gene
-sequencing, the model performance for individuals was significantly
-greater than all other models tested {[}Figure 5A{]}. In addition to the
-better model performance of individuals, the most important OTUs to this
-model were not the most important OTUs for the other models tested
-{[}Figure 5B{]}. Our observations show that differences between high
-fidelity DNA polymerases can be found. However, when using 16S rRNA gene
-sequencing these technical differences are dwarfed by biological
-differences between individuals.
+polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
+differences observed within the number of cycles between any of the
+polymerases for fecal samples. When Random Forest models were created to
+classify number of cycles, polymerase, and individual based on the 16S
+rRNA gene sequencing, the model performance for individuals was
+significantly greater than all other models tested {[}Figure 5A{]}. In
+addition to the better model performance of individuals, the most
+important OTUs to this model were not the most important OTUs for the
+other models tested {[}Figure 5B{]}. Our observations show that
+differences between high fidelity DNA polymerases can be found. However,
+when using 16S rRNA gene sequencing these technical differences are
+dwarfed by biological differences between individuals.
 
@@
 
-\textbf{Figure 1: The number and range of OTUs in fecal samples varies
-by polymerase.} The points represent the median number of OTUs of all
-four fecal samples. The lines represent the range of the minimum and
-maximum number of OTUs detected within the four fecal samples. The range
-in the number of OTUs detected in the different fecal samples increased
-as cycle number increased. This increased range also was larger for
-specific HiFi DNA polymerases.
+\textbf{Figure 1: Relative abundance differences due to GC is consistent
+across polymerases.} At each number of cycle the points and lines
+represent the median and with the minimum and maximum relative abundance
+respectively. Regardless of the number of cycle used a consistent
+difference between high and low GC bacteria was observed. The dotted
+line represents the actual relative abundance that each bacterium should
+be at within the mock community sample.
 
-\textbf{Figure 2: The number and range of OTUs in mock samples varies by
-polymerase.} The points represent the median number of OTUs for the mock
-samples. The lines represent the range of the minimum and maximum number
-of OTUs detected within the four fecal samples. The dotted black line
-represents the number of OTUs detected when only the references
-sequences for the mock community are clustered. The range in the number
-of OTUs detected in the mock samples increased as cycle number
-increased. This range was also larger for specific HiFi DNA polymerases.
+\textbf{Figure 2: Error rate and chimera prevalence vary by polymerase
+and affect the number of observed OTUs.} A) The error bars represent the
+75\% interquartile range of the median error rate. B) Percentage of
+chimeric sequences without the removal of chimeras with VSEARCH. C)
+Percentage of chimeric sequences with the removal of chimeras with
+VSEARCH. D) The total percent of chimeric sequences removed with VSEARCH
+by cycle number. The error bars represent the 75\% interquartile range
+of the median. E) Chimera prevalence and the the observed number of OTUs
+are strongly correlated in mock community samples.
 
-\textbf{Figure 3: Lower five-cycle intervals have larger differences
-than higher five-cycle intervals using Bray-Curtis.} A) within person
-differences based on the next 5-cycle PCR interval in fecal samples. B)
-Within replicate difference based on the next 5-cycle PCR interval in
-mock samples. The points represent the median Bray-Curtis index for the
+\textbf{Figure 3: The number and range of OTUs in mock and fecal samples
+varies by polymerase.} The points represent the median number of OTUs
+for the mock samples. The lines represent the range of the minimum and
+maximum number of OTUs detected within the four fecal samples. The
+dotted black line represents the number of OTUs detected when only the
+references sequences for the mock community are clustered. A) The range
+in the number of OTUs detected in the mock samples increased as cycle
+number increased. This range was also larger for specific HiFi DNA
+polymerases. B) The range in the number of OTUs detected in the
+different fecal samples increased as cycle number increased. This
+increased range also was larger for specific HiFi DNA polymerases.
+
+\textbf{Figure 4: Lower five-cycle intervals have larger differences
+than higher five-cycle intervals using Bray-Curtis.} A) Within replicate
+difference based on the next 5-cycle PCR interval in mock samples. B)
+Within person differences based on the next 5-cycle PCR interval in
+fecal samples. The points represent the median Bray-Curtis index for the
 samples. The lines represent the range of the minimum and maximum
@@
 
-\textbf{Figure 4: Variation in error rate by polymerase in mock
-community samples are similar across the number of cycles used.} The
-error bars represent the 75\% interquartile range of the median error
-rate.
-
-\textbf{Figure 5: Chimera prevalence varies by polymerase and increases
-with higher number of cycles used.} A) Percentage of chimeric sequences
-without the removal of chimeras with VSEARCH. C) Percentage of chimeric
-sequences with the removal of chimeras with VSEARCH. C) The total
-percent of chimeric sequences removed with VSEARCH by cycle number. The
-error bars represent the 75\% interquartile range of the median.
-
-\textbf{Figure 6: Regardless of polymerase used the higher the chimera
-prevalence the higher the observed number of OTUs in mock community
-samples.}
+\textbf{Figure 5: Biological variation in bacterial community
+composition due to individual was larger than number of cycles and
+polymerase used.} A) Using 16S rRNA gene sequencing, the classification
+of individuals was higher than any other model tested. The lines
+represent the median probability of correct classification and the gray
+points represent the individual probability from one run out of the 100
+10-fold CV runs. B) The most important OTUs to the model classifying
+individuals are not the same as those that are most important for the
+number of cycles or polymerase used.
 

### 2018-07-11 submission/manuscript.Rmd
add fecal to biological variation

@@
 Our observations show that differences between high fidelity DNA polymerases can be found.
-However, when using 16S rRNA gene sequencing these technical differences are dwarfed by biological differences between individuals.
+However, when using 16S rRNA gene sequencing these technical differences are dwarfed by biological differences between fecal bacterial communities within individuals.
 
@@
 Despite this variation in the number of OTUs, our Random Forest models were able to correctly classify individuals regardless of the number of cycles and polymerase used.
-These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to bacterial community composition differences between individuals will still be consistent.
+These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to fecal bacterial community composition differences between individuals will still be consistent.
 
@@
 
-Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation between individuals.
-Therefore, if the biological signal of interest is similar to differences in bacterial community composition found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
+Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation in fecal communities between individuals.
+Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
 

### 2018-07-11 submission/manuscript.tex
add fecal to biological variation

@@
 when using 16S rRNA gene sequencing these technical differences are
-dwarfed by biological differences between individuals.
+dwarfed by biological differences between fecal bacterial communities
+within individuals.
 
@@
 polymerases used across studies, biological variation similar in effect
-to bacterial community composition differences between individuals will
-still be consistent.
+to fecal bacterial community composition differences between individuals
+will still be consistent.
 
@@
 a variety of polymerases are dwarfed by the actual biological variation
-between individuals. Therefore, if the biological signal of interest is
-similar to differences in bacterial community composition found between
-individuals, then the type of high fidelity DNA polymerase used will
-only minimally change the results.
+in fecal communities between individuals. Therefore, if the biological
+signal of interest is similar to differences in fecal bacterial
+communities found between individuals, then the type of high fidelity
+DNA polymerase used will only minimally change the results.
 

### 2018-07-11 submission/manuscript.Rmd
add disclaimer to abstract that effect needs to match fecal samples

@@
 
-**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between indivdiuals.
-Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing, should be detectable regardless of polymerase and number of cycles used.
+**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between fecal bacterial communities within indivdiuals.
+Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing of fecal samples, should be detectable regardless of polymerase and number of cycles used.
 

### 2018-07-11 submission/manuscript.tex
add disclaimer to abstract that effect needs to match fecal samples

@@
 differences based on polymerase and number of cycles used, they are
-small with respect to the biological differences between indivdiuals.
-Collectively, these results provide evidence that a real biological
-difference between groups, based on 16S rRNA gene sequencing, should be
-detectable regardless of polymerase and number of cycles used.
+small with respect to the biological differences between fecal bacterial
+communities within indivdiuals. Collectively, these results provide
+evidence that a real biological difference between groups, based on 16S
+rRNA gene sequencing of fecal samples, should be detectable regardless
+of polymerase and number of cycles used.
 

### 2018-07-11 submission/manuscript.Rmd
first rearrangement of introduction

@@
 Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow.
-Yet, standardizing a 16S rRNA gene sequencing workflow will ultimately result in a standardized and reproducible bias due to choices made on the methods used for preservation, extraction, PCR, and sequencing.
-Within this context, all 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility.
-In order to interpret specific studies within the broader context of the overall field, assessing error at different parts of the 16S rRNA gene sequencing workflow is critical.
-
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
 The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017].
-For preservation and extraction stages of the workflow, it has been consistently found that there are biases based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
-Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR bias to this overall workflow is not well characterized.
+For preservation and extraction stages of the workflow, it has been consistently found that there are errors based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
+Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not well characterized.
 
-There is a large body of literature that shows there are biases due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000].
+There is a large body of literature that shows there are errors due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000].
 Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g.
@@
 Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998].
-Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015].
+Many of these sources of error are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015].
 For example, both amplification error and non-specific amplification (e.g.
@@
 Yet, within this study there was no comparison made between different high fidelity DNA polymerases.
-Due to this gap, it is natural to extend this line of inquiry and test if biases in the number of OTUs and chimeras also are dependent on the type of high fidelity DNA polymerase.
-This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing.
-We will accomplish this by examining the number of OTUs, Bray-Curtis index, error rate, number of sequences with an error, and chimera prevalence at varying number of cycles in five different high fidelity DNA polymerases
+Additionally, the more complex fecal samples that were analyzed had different methodolgies applied to the two different groups tested, which could exagerate the actual biological differences detected.
+Due to this gap, it is natural to extend this line of inquiry.
+First, for detecting whether there are high fidelity polymerase and number of cycle dependent error.
+Second, for examining whether these differences are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
+Thus, this study will investigate whether differences in high fidelity DNA polymerases are due to error and whether these differences obscure the biological variation in the fecal bacterial community composition between individuals.
+We will accomplish this by examining GC-based amplification error, sequence error (e.g.
+subsitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases.
+Additionally, we also used Random Forest models to classify number of cycles, polymerase, and individuals to assess which model was best when using 16S rRNA gene sequencing data.
 
@@
 subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g.
+higher GC content genomes, base subsitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
 

### 2018-07-11 submission/manuscript.tex
first rearrangement of introduction

@@
 et al., 2008; Zupancic et al., 2012). Some of this variation could be
-due to error introduced during the 16S rRNA gene sequencing workflow.
-Yet, standardizing a 16S rRNA gene sequencing workflow will ultimately
-result in a standardized and reproducible bias due to choices made on
-the methods used for preservation, extraction, PCR, and sequencing.
-Within this context, all 16S rRNA gene sequencing methods are biased
-even when these workflows are standardized to increase reproducibility.
-In order to interpret specific studies within the broader context of the
-overall field, assessing error at different parts of the 16S rRNA gene
-sequencing workflow is critical.
-
-A typical 16S rRNA gene sequencing workflow can be divided into
+due to error introduced during the 16S rRNA gene sequencing workflow. A
+typical 16S rRNA gene sequencing workflow can be divided into
 preservation, extraction, PCR, and sequencing steps. The preservation
@@
 stages of the workflow, it has been consistently found that there are
-biases based on the kits used, but that these differences are smaller
+errors based on the kits used, but that these differences are smaller
 than the overall biological difference measured between samples with
@@
 studies use the same PCR approach while varying preservation or
-extraction method, the contribution of PCR bias to this overall workflow
-is not well characterized.
+extraction method, the contribution of PCR error to this overall
+workflow is not well characterized.
 
-There is a large body of literature that shows there are biases due to
+There is a large body of literature that shows there are errors due to
 primer and number of cycles chosen for the PCR stage of 16S rRNA gene
@@
 16S rRNA genes higher in AT versus those higher in GC content (Polz \&
-Cavanaugh, 1998). Many of these sources of biases are made worse as the
+Cavanaugh, 1998). Many of these sources of error are made worse as the
 number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
@@
 (Gohl et al., 2016). Yet, within this study there was no comparison made
-between different high fidelity DNA polymerases. Due to this gap, it is
-natural to extend this line of inquiry and test if biases in the number
-of OTUs and chimeras also are dependent on the type of high fidelity DNA
-polymerase. This study will investigate how high fidelity DNA
-polymerases can bias observed bacterial community results derived from
-16S rRNA gene sequencing. We will accomplish this by examining the
-number of OTUs, Bray-Curtis index, error rate, number of sequences with
-an error, and chimera prevalence at varying number of cycles in five
-different high fidelity DNA polymerases
+between different high fidelity DNA polymerases. Additionally, the more
+complex fecal samples that were analyzed had different methodolgies
+applied to the two different groups tested, which could exagerate the
+actual biological differences detected. Due to this gap, it is natural
+to extend this line of inquiry. First, for detecting whether there are
+high fidelity polymerase and number of cycle dependent error. Second,
+for examining whether these differences are enough to obscure smaller
+biological signals than what has been previously reported (Gohl et al.,
+2016). Thus, this study will investigate whether differences in high
+fidelity DNA polymerases are due to error and whether these differences
+obscure the biological variation in the fecal bacterial community
+composition between individuals. We will accomplish this by examining
+GC-based amplification error, sequence error (e.g.~subsitutions),
+chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
+differences at varying number of cycles in five different high fidelity
+DNA polymerases. Additionally, we also used Random Forest models to
+classify number of cycles, polymerase, and individuals to assess which
+model was best when using 16S rRNA gene sequencing data.
 
@@
 \emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
-(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016). For
+simplicity, the use of the term `error' within the context of this study
+moving forward encompasses changes in the downstream sequencing results
+due to the amplification step (e.g.~higher GC content genomes, base
+subsitutions, deletions, insertions, non-specific amplification,
+chimeras, etc.).
 

### 2018-07-11 submission/manuscript.Rmd
intro edits

@@
 The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017].
-For preservation and extraction stages of the workflow, it has been consistently found that there are errors based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
+For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
 Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not well characterized.
@@
 In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
-Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.
+Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error
 
 A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
-The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
-Yet, within this study there was no comparison made between different high fidelity DNA polymerases.
-Additionally, the more complex fecal samples that were analyzed had different methodolgies applied to the two different groups tested, which could exagerate the actual biological differences detected.
-Due to this gap, it is natural to extend this line of inquiry.
-First, for detecting whether there are high fidelity polymerase and number of cycle dependent error.
-Second, for examining whether these differences are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
+The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
+Yet, within this study there was no comparison made between different high fidelity DNA polymerases on the same samples.
+Additionally, the fecal samples that were analyzed, as a model of a more complex community, had different processing methodolgies applied to the two different treatment groups tested, which could exagerate the observed biological differences.
+Due to these gaps, it is natural to extend the previous line of inquiry to include a more detailed examination of two specific areas.
+First, for the investigation of whether differences in PCR error are detected between high fidelity polymerases and number of cycles used.
+Second, for examining whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
 Thus, this study will investigate whether differences in high fidelity DNA polymerases are due to error and whether these differences obscure the biological variation in the fecal bacterial community composition between individuals.
-We will accomplish this by examining GC-based amplification error, sequence error (e.g.
+We accomplish this by examining GC-based amplification error, sequence error (e.g.
 subsitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases.
-Additionally, we also used Random Forest models to classify number of cycles, polymerase, and individuals to assess which model was best when using 16S rRNA gene sequencing data.
+Additionally, we also created Random Forest models to classify number of cycles, polymerase, and individuals to assess which model had the best classification success when using 16S rRNA gene sequencing data.
 

### 2018-07-11 submission/manuscript.tex
intro edits

@@
 stages of the workflow, it has been consistently found that there are
-errors based on the kits used, but that these differences are smaller
-than the overall biological difference measured between samples with
-different kits (Song et al., 2016; Bassis et al., 2017). Since these
-studies use the same PCR approach while varying preservation or
-extraction method, the contribution of PCR error to this overall
-workflow is not well characterized.
+errors in the observed bacterial community composition based on the kits
+used, but that these differences are smaller than the overall biological
+difference measured between samples with different kits (Song et al.,
+2016; Bassis et al., 2017). Since these studies use the same PCR
+approach while varying preservation or extraction method, the
+contribution of PCR error to this overall workflow is not well
+characterized.
 
@@
 16S rRNA gene sequencing results have not been well studied like some of
-the other sources of PCR-based bias.
+the other sources of PCR-based error
 
@@
 between normal and high fidelity DNA polymerases (Gohl et al., 2016).
-The authors could reduce the difference between the two polymerases by
-optimizing the annealing and extension steps within the PCR protocol
-(Gohl et al., 2016). Yet, within this study there was no comparison made
-between different high fidelity DNA polymerases. Additionally, the more
-complex fecal samples that were analyzed had different methodolgies
-applied to the two different groups tested, which could exagerate the
-actual biological differences detected. Due to this gap, it is natural
-to extend this line of inquiry. First, for detecting whether there are
-high fidelity polymerase and number of cycle dependent error. Second,
-for examining whether these differences are enough to obscure smaller
-biological signals than what has been previously reported (Gohl et al.,
-2016). Thus, this study will investigate whether differences in high
-fidelity DNA polymerases are due to error and whether these differences
-obscure the biological variation in the fecal bacterial community
-composition between individuals. We will accomplish this by examining
-GC-based amplification error, sequence error (e.g.~subsitutions),
-chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
+The authors of this study could reduce the difference between two
+polymerases by optimizing the annealing and extension steps within the
+PCR protocol (Gohl et al., 2016). Yet, within this study there was no
+comparison made between different high fidelity DNA polymerases on the
+same samples. Additionally, the fecal samples that were analyzed, as a
+model of a more complex community, had different processing methodolgies
+applied to the two different treatment groups tested, which could
+exagerate the observed biological differences. Due to these gaps, it is
+natural to extend the previous line of inquiry to include a more
+detailed examination of two specific areas. First, for the investigation
+of whether differences in PCR error are detected between high fidelity
+polymerases and number of cycles used. Second, for examining whether
+these differences in error are enough to obscure smaller biological
+signals than what has been previously reported (Gohl et al., 2016).
+Thus, this study will investigate whether differences in high fidelity
+DNA polymerases are due to error and whether these differences obscure
+the biological variation in the fecal bacterial community composition
+between individuals. We accomplish this by examining GC-based
+amplification error, sequence error (e.g.~subsitutions), chimera
+prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
 differences at varying number of cycles in five different high fidelity
-DNA polymerases. Additionally, we also used Random Forest models to
+DNA polymerases. Additionally, we also created Random Forest models to
 classify number of cycles, polymerase, and individuals to assess which
-model was best when using 16S rRNA gene sequencing data.
+model had the best classification success when using 16S rRNA gene
+sequencing data.
 

### 2018-07-13 submission/manuscript.Rmd
fix intro with lab suggestions

@@
 
-**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between fecal bacterial communities within indivdiuals.
+**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between fecal bacterial communities within individuals.
 Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing of fecal samples, should be detectable regardless of polymerase and number of cycles used.
@@
 For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
-Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not well characterized.
+Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not as well characterized.
 
@@
 In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
-Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error
+For simplicity, the use of the term 'error' within the context of this study will be used as a broad term for changes in the downstream sequencing results due to the amplification step (e.g.
+higher GC content genomes, base substitutions, chimeras, changes from the expected relative abundance, etc.).
 
+Interestingly, the influence different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error.
 A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
@@
 Yet, within this study there was no comparison made between different high fidelity DNA polymerases on the same samples.
-Additionally, the fecal samples that were analyzed, as a model of a more complex community, had different processing methodolgies applied to the two different treatment groups tested, which could exagerate the observed biological differences.
-Due to these gaps, it is natural to extend the previous line of inquiry to include a more detailed examination of two specific areas.
-First, for the investigation of whether differences in PCR error are detected between high fidelity polymerases and number of cycles used.
-Second, for examining whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
-Thus, this study will investigate whether differences in high fidelity DNA polymerases are due to error and whether these differences obscure the biological variation in the fecal bacterial community composition between individuals.
-We accomplish this by examining GC-based amplification error, sequence error (e.g.
-subsitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases.
-Additionally, we also created Random Forest models to classify number of cycles, polymerase, and individuals to assess which model had the best classification success when using 16S rRNA gene sequencing data.
+The authors instead focused their efforts on testing specific types of amplification methods.
+Additionally, the fecal samples that were analyzed, as a model of a more complex community, may overestimate the biological variation in human samples due to the comparison of captive versus semi-captive red-shanked doucs.
+Due to these gaps, it makes sense to extend the previous line of inquiry to include a more detailed examination of two specific areas.
+First, whether differences in PCR error are detected based on number of cycles used and high fidelity polymerases.
+Second, whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
+
+This study will investigate whether there are error differences that are dependent on high fidelity DNA polymerases and whether these differences obscure the biological variation in the fecal bacterial communities between individuals.
+We accomplish this by first examining GC-based amplification error, sequence error (e.g.
+substitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases in mock communities.
+In addition to these metrics we also created Random Forest models of the mock community to assess how successful classification based on 16S rRNA gene sequencing of number of cycles and type of polymerase could be.
+We then assessed if similar differences in the number of OTUs, Bray-Curtis index, and OTUs could be detected in human fecal samples.
+Additionally, we also created Random Forest models to assess how successful they could be at classifying fecal samples by number of cycles, polymerase used, or individual.
+Collectively, our observations suggest that although detectable differences occur based on number of cycles or high fidelity DNA polymerase used, they are smaller than the biological differences between individuals.
 
@@
 For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g.
-higher GC content genomes, base subsitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
+higher GC content genomes, base substitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
 
@@
 A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value.
-The probability of a correct call was obtained from this Logloss value by taking the negative natural logrithim.
+The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm.
 For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
@@
 Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S5].
-Similarily, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
+Similarly, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
 Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
@@
 The model classification success for an individual was significantly better than all other models tested [Figure 5A & Table S8].
-Additionally, the top 10 most impotant variables to the polymerase and number of cycle models were not the most important variables in the model classifying individuals [Figure 5B].
-This observation shows that OTUs that change due to the number of cycles and polymerase used are not the same as OTUs that differentiate between indvidiuals [Figure 5B].
-Although there are a variety of errors associated with the amplification process, they are smaller than the bacterial community variation between indviduals.
+Additionally, the top 10 most important variables to the polymerase and number of cycle models were not the most important variables in the model classifying individuals [Figure 5B].
+This observation shows that OTUs that change due to the number of cycles and polymerase used are not the same as OTUs that differentiate between individuals [Figure 5B].
+Although there are a variety of errors associated with the amplification process, they are smaller than the bacterial community variation between individuals.
 
@@
 Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a mock community [Figure 2E].
-This total number of OTUs also varied by polymerase used for these mock community samples [Fiugre 3A].
+This total number of OTUs also varied by polymerase used for these mock community samples [Figure 3A].
 The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for both mock community and fecal samples [Figure 3].
 However, these differences did not translate to large bacterial community composition changes based on polymerase [Figure 4].
-Even though there were bacterial community composition changes based on the number of cycles they occured independently of the polymerase used [Figure 4].
+Even though there were bacterial community composition changes based on the number of cycles they occurred independently of the polymerase used [Figure 4].
 Additionally, there were no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.

### 2018-07-13 submission/manuscript.tex
fix intro with lab suggestions

@@
 small with respect to the biological differences between fecal bacterial
-communities within indivdiuals. Collectively, these results provide
+communities within individuals. Collectively, these results provide
 evidence that a real biological difference between groups, based on 16S
@@
 approach while varying preservation or extraction method, the
-contribution of PCR error to this overall workflow is not well
+contribution of PCR error to this overall workflow is not as well
 characterized.
@@
 multiple families of DNA polymerases that have their own error rate and
-proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
-influence that these different DNA polymerases can have on the observed
-16S rRNA gene sequencing results have not been well studied like some of
-the other sources of PCR-based error
+proof reading capacity (Ishino \& Ishino, 2014). For simplicity, the use
+of the term `error' within the context of this study will be used as a
+broad term for changes in the downstream sequencing results due to the
+amplification step (e.g.~higher GC content genomes, base substitutions,
+chimeras, changes from the expected relative abundance, etc.).
 
-A recent study found differences in the number of OTUS and chimeras
-between normal and high fidelity DNA polymerases (Gohl et al., 2016).
-The authors of this study could reduce the difference between two
-polymerases by optimizing the annealing and extension steps within the
-PCR protocol (Gohl et al., 2016). Yet, within this study there was no
-comparison made between different high fidelity DNA polymerases on the
-same samples. Additionally, the fecal samples that were analyzed, as a
-model of a more complex community, had different processing methodolgies
-applied to the two different treatment groups tested, which could
-exagerate the observed biological differences. Due to these gaps, it is
-natural to extend the previous line of inquiry to include a more
-detailed examination of two specific areas. First, for the investigation
-of whether differences in PCR error are detected between high fidelity
-polymerases and number of cycles used. Second, for examining whether
-these differences in error are enough to obscure smaller biological
-signals than what has been previously reported (Gohl et al., 2016).
-Thus, this study will investigate whether differences in high fidelity
-DNA polymerases are due to error and whether these differences obscure
-the biological variation in the fecal bacterial community composition
-between individuals. We accomplish this by examining GC-based
-amplification error, sequence error (e.g.~subsitutions), chimera
+Interestingly, the influence different DNA polymerases can have on the
+observed 16S rRNA gene sequencing results have not been well studied
+like some of the other sources of PCR-based error. A recent study found
+differences in the number of OTUS and chimeras between normal and high
+fidelity DNA polymerases (Gohl et al., 2016). The authors of this study
+could reduce the difference between two polymerases by optimizing the
+annealing and extension steps within the PCR protocol (Gohl et al.,
+2016). Yet, within this study there was no comparison made between
+different high fidelity DNA polymerases on the same samples. The authors
+instead focused their efforts on testing specific types of amplification
+methods. Additionally, the fecal samples that were analyzed, as a model
+of a more complex community, may overestimate the biological variation
+in human samples due to the comparison of captive versus semi-captive
+red-shanked doucs. Due to these gaps, it makes sense to extend the
+previous line of inquiry to include a more detailed examination of two
+specific areas. First, whether differences in PCR error are detected
+based on number of cycles used and high fidelity polymerases. Second,
+whether these differences in error are enough to obscure smaller
+biological signals than what has been previously reported (Gohl et al.,
+2016).
+
+This study will investigate whether there are error differences that are
+dependent on high fidelity DNA polymerases and whether these differences
+obscure the biological variation in the fecal bacterial communities
+between individuals. We accomplish this by first examining GC-based
+amplification error, sequence error (e.g.~substitutions), chimera
 prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
 differences at varying number of cycles in five different high fidelity
-DNA polymerases. Additionally, we also created Random Forest models to
-classify number of cycles, polymerase, and individuals to assess which
-model had the best classification success when using 16S rRNA gene
-sequencing data.
+DNA polymerases in mock communities. In addition to these metrics we
+also created Random Forest models of the mock community to assess how
+successful classification based on 16S rRNA gene sequencing of number of
+cycles and type of polymerase could be. We then assessed if similar
+differences in the number of OTUs, Bray-Curtis index, and OTUs could be
+detected in human fecal samples. Additionally, we also created Random
+Forest models to assess how successful they could be at classifying
+fecal samples by number of cycles, polymerase used, or individual.
+Collectively, our observations suggest that although detectable
+differences occur based on number of cycles or high fidelity DNA
+polymerase used, they are smaller than the biological differences
+between individuals.
 
@@
 due to the amplification step (e.g.~higher GC content genomes, base
-subsitutions, deletions, insertions, non-specific amplification,
+substitutions, deletions, insertions, non-specific amplification,
 chimeras, etc.).
@@
 Logloss value. The probability of a correct call was obtained from this
-Logloss value by taking the negative natural logrithim. For both error
+Logloss value by taking the negative natural logarithm. For both error
 and chimera analysis, samples were tested using Kruskal-Wallis with a
@@
 versus the other polymerases across the number of cycles used {[}Table
-S5{]}. Similarily, differences in the range of the number of OTUs
+S5{]}. Similarly, differences in the range of the number of OTUs
 detected for fecal samples was dependent on the polymerase used
@@
 individual was significantly better than all other models tested
-{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most impotant
+{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most important
 variables to the polymerase and number of cycle models were not the most
@@
 cycles and polymerase used are not the same as OTUs that differentiate
-between indvidiuals {[}Figure 5B{]}. Although there are a variety of
+between individuals {[}Figure 5B{]}. Although there are a variety of
 errors associated with the amplification process, they are smaller than
-the bacterial community variation between indviduals.
+the bacterial community variation between individuals.
 
@@
 number of OTUs also varied by polymerase used for these mock community
-samples {[}Fiugre 3A{]}. The range in the number of OTUs observed
+samples {[}Figure 3A{]}. The range in the number of OTUs observed
 increased with the number of cycles and the change in range was
@@
 4{]}. Even though there were bacterial community composition changes
-based on the number of cycles they occured independently of the
+based on the number of cycles they occurred independently of the
 polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based

### 2018-07-13 submission/manuscript.Rmd
update manuscirpt order and results order (rough template

@@
 # Read in RF model data
+mock_poly_rf_summary <- read_csv("../data/process/tables/polymerase_group_mock_rf_model_summary.csv")
+mock_cycles_rf_summary <- read_csv("../data/process/tables/cycles_group_mock_rf_model_summary.csv")
+
 dunn_rf_data <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv")
@@
 
-## Materials & Methods
-
-***Human and mock samples.***
-Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
-The processing and storage of these samples were previously published [@erin_seekatz_2016].
-Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study.
-All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
-
-***PCR protocol.***
-The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer.
-The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016].
-For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
-
-The number of cycles in the PCR for fecal and mock samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
-The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
-The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130).
-No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
-
-***Sequence processing.***
-The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
-The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P.
-aeruginosa*, *E.
-coli*, *S.
-enterica*, *L.
-fermentum*, *E.
-faecalis*, *S.
-aureus*, *L.
-monocytogenes*, and *B.
-subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
-For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g.
-higher GC content genomes, base substitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
-
-***Analysis workflow.***
-To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
-Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
-Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
-The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples.
-Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data.
-Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).
-
-***Statistical analysis.***
-All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
-High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community.
-Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation.
-Bray-Curtis distance matrices were generated using mothur.
-The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
-The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation].
-A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value.
-The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm.
-For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
-Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
-
-***Reproducible methods.***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
-
-\newpage
-
 ## Results
 
-***Mock community relative abundance differs by GC content.***
+***Error differences due to number of cycles or polymerase used are detectable in mock communities.***
 There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1 & Table S1].
@@
 
-***Sequence error varies by polymerase and is consistent across the number of cycles used.***
 The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 2A & Table S2].
@@
 
-***Prevalence of chimeric sequences are polymerase dependent and correlate with the number of OTUs.***  Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
+Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
 We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2].
@@
 
-***The number of OTUs generated are dependent on polymerase used.***
-In addition to the sequence error and chimera differences between polymerases, we also investigated whether there were polymerase or number of cycle dependent effects on the observed number of OTUs.
-There was a polymerase dependent difference in the range of the number of OTUs in the mock community samples [Figure 3A].
+***Bacterial community composition differences based on number of cycles or polymerase used are less obvious in mock community samples.***
+Due to the close relationship between chimera prevalence and the number of OTUs, we also observed a polymerase dependent difference in the range of the number of OTUs in the mock community samples based on number of cycles and polymerase [Figure 3A].
 The closest any of the polymerases came to a total of `r pull(ref_mock_otu_num, numOtus)` OTUs, created by running the mock reference 16S sequences through mothur processing, was at 25 and 30 cycles [Figure 3A].
@@
 Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S5].
+
+Next, we investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
+We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A].
+Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles.
+We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
+These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7].
+The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases [Table S7].
+
+When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`].
+Although there was a difference, neither model was successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)).
+Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are subtle.
+
+***Few differences based on number of cycles and polymerased used were identified in fecal samples.***
 Similarly, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
@@
 
-***The bacterial communities generated by the polymerases were similar but varied by number of cycles.***
-Next, we investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
-We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A].
 With respect to fecal samples, within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
@@
 
-***Polymerase dependent differences in taxa are more pronounced in mock versus fecal samples.***
-Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles.
-Further investigation of mock community samples found that OTUs taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
-These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7].
-The majority of the differences between taxa were due to differences between Kappa and the other polymerases [Table S7].
 When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found at any number of cycles (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
@@
 
-***Variation due to polymerase and number of cycles is dwarfed by the biological variation of an individual.***
 Using 16S rRNA gene sequencing data there was a large disparity in a model's ability to correctly classify samples based on polymerase used, number of cycles, and individual [Figure 5A].
@@
 
+## Materials & Methods
+
+***Human and mock samples.***
+Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
+The processing and storage of these samples were previously published [@erin_seekatz_2016].
+Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study.
+All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+
+***PCR protocol.***
+The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer.
+The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
+For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016].
+For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
+
+The number of cycles in the PCR for fecal and mock samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
+The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
+The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130).
+No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
+
+***Sequence processing.***
+The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
+The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
+Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P.
+aeruginosa*, *E.
+coli*, *S.
+enterica*, *L.
+fermentum*, *E.
+faecalis*, *S.
+aureus*, *L.
+monocytogenes*, and *B.
+subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
+For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g.
+higher GC content genomes, base substitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
+
+***Analysis workflow.***
+To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
+Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
+Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
+The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples.
+Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data.
+Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).
+
+***Statistical analysis.***
+All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
+Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
+High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community.
+Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
+For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation.
+Bray-Curtis distance matrices were generated using mothur.
+The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
+The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation].
+A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value.
+The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm.
+For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
+Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
+
+***Reproducible methods.***
+The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
+
+\newpage
+
 ## Acknowledgements

### 2018-07-13 submission/manuscript.tex
update manuscirpt order and results order (rough template

@@
 
+\subsection{Results}\label{results}
+
+\textbf{\emph{Error differences due to number of cycles or polymerase
+used are detectable in mock communities.}} There was a significant
+difference in relative abundance between high/low GC content based on
+either the V4 16S rRNA gene region or the whole genome {[}Figure 1 \&
+Table S1{]}. However, using only the GC content of the V4 region
+resulted in less differences than when using the whole genome {[}Table
+S1{]}. The differences between high and low GC content groups were
+significant as early as 20 cycles in all polymerases except Q5 {[}Table
+S1{]}. The highest (\emph{Staphylococcus}) and lowest
+(\emph{Pseudomonas}) GC content bacteria were the most divergent from
+the expected relative abundance of 0.12 {[}Figure 1{]}. These
+observations suggest that the high fidelity DNA polymerases tested do
+not prevent GC-based amplification error.
+
+The median error rate varied by polymerase, with Kappa having the
+highest error rate of all the polymerases across the number of cycles
+{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
+observed across the number of cycles was between Kappa and other
+polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
+contained at least one error was also polymerase dependent, with the
+majority of differences being between Kappa or Accuprime and the other
+polymerases {[}Table S2 \& S3{]}. These differences in error rates were
+not due to polymerase dependent differences in base substitution rate
+{[}Figure S1{]}. Collectively, the results suggest that sequence error
+is dependent on polymerase, persists across the number of cycles used,
+and are not due to any bias towards a specific base substitution.
+
+Based on the previous results, we examined whether chimeras also were
+dependent on polymerase and whether this could affect the total number
+of OTUs. We observed significant differences in the chimera prevalence
+based on polymerase across the number of cycles used (P-value
+\textless{} 0.05) {[}Table S2{]}. Differences in chimera prevalence
+between Platinum and all other polymerases accounted for the majority of
+these differences {[}Table S3{]}. Accuprime\textsuperscript{TM} had the
+lowest chimera prevalence of all polymerases regardless of whether
+chimera removal with VSEARCH was used {[}Figure 2B \& 2C{]}. The number
+of cycles used clearly increased chimera prevalence for all polymerases
+but the rate of increase differed {[}Figure 2B \& 2C{]}. Additionally,
+there was a plateau in the total percent of chimeras that were removed
+that was similar for all polymerases {[}Figure 2D{]}. A positive
+correlation was observed between chimeric sequences and the number of
+OTUs for all polymerases {[}Figure 2E{]}. This positive correlation was
+strongest for Accuprime\textsuperscript{TM}, Platinum, and Phusion
+{[}Figure 2E{]}. This data suggests that chimera prevalence depends on
+polymerase, is made worse by increasing the number of cycles, and
+confirms that the number of OTUs is dependent on the prevalence of these
+chimeric sequences.
+
+\textbf{\emph{Bacterial community composition differences based on
+number of cycles or polymerase used are less obvious in mock community
+samples.}} Due to the close relationship between chimera prevalence and
+the number of OTUs, we also observed a polymerase dependent difference
+in the range of the number of OTUs in the mock community samples based
+on number of cycles and polymerase {[}Figure 3A{]}. The closest any of
+the polymerases came to a total of 8 OTUs, created by running the mock
+reference 16S sequences through mothur processing, was at 25 and 30
+cycles {[}Figure 3A{]}. Differences between polymerases for the number
+of OTUs created were observed as early as 20 cycles in the mock
+community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
+Tukey post-hoc test, the majority of differences for the number of OTUs
+detected in the mock community was largely due to Kappa and Platinum
+versus the other polymerases across the number of cycles used {[}Table
+S5{]}.
+
+Next, we investigated the effect polymerases and number of cycles had on
+the bacterial community composition in more detail by using the
+Bray-Curtis index. We observed no difference in the bacterial community
+composition between any measure tested when using the mock community
+(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
+differences in overall bacterial community composition, we next examined
+if specific taxa vary depending on polymerase and number of cycles. We
+found that OTUs that taxonomically classified to \emph{Salmonella},
+\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
+differed based on polymerase used {[}Table S6 \& S7{]}. These specific
+OTUs were all members of the mock community and the differences were
+most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
+differences between taxa were due to differences between a single
+polymerase (Kappa) and the other polymerases {[}Table S7{]}.
+
+When assessing the Random Forest models there was a difference between
+models built to classify number of cycles versus those built to classify
+polymerase used {[}P-value = 5.06e-07{]}. Although there was a
+difference, neither model was successful at correctly classifying
+samples based on 16S rRNA gene sequencing data (cycles model,
+probability of correct classification = 0.42 (0.4 - 0.43)(min - max)
+versus polymerase model, probability of correct classification = 0.39
+(0.37 - 0.4)). Our observations in mock communities suggest that
+although differences are detected based on number of cycles or
+polymerase used, the overall community-wide differences are subtle.
+
+\textbf{\emph{Few differences based on number of cycles and polymerased
+used were identified in fecal samples.}} Similarly, differences in the
+range of the number of OTUs detected for fecal samples was dependent on
+the polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus
+Phusion at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend
+for lower number of cycles (15-20) to result in a reduced range in the
+number of OTUs versus higher number of cycles (25, 30, and 35) for all
+polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
+cycles = 84 - 136) {[}Figure 3B{]}. Unlike mock community samples, an
+overall difference in the number of OTUs was only detected between
+polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
+P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
+identify any differences between specific polymerase groups (P-value
+\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
+noticeably larger difference between polymerases in the number of OTUs
+than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
+consistent across the number of amplification cycles tested {[}Figure
+3{]}. Based on these observations, it is clear that larger differences
+in total number of OTUs based on polymerases are found in mock versus
+fecal communities.
+
+With respect to fecal samples, within each respective 5-cycle increment
+comparison there was no difference in Bray-Curtis index between the
+polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}. Using
+PERMANOVA to test for community differences based on any of the number
+of cycles within polymerases, only Phusion had cycle dependent
+differences (P-value = 0.03). Additionally, Phusion was one of two
+polymerases that had enough sequences to be rarefied to 1000 sequences
+at 15 cycles. There was an overall decrease in Bray-Curtis index by
+cycle comparison group (e.g.~the 15 cycle versus 20 cycle group compared
+to the 30 cycle versus 35 cycle group) (P-value \textless{} 0.01)
+{[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus 20
+cycle and 20 cycle versus 25 cycle comparison groups had a higher
+Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
+(P-value \textless{} 0.05). Overall, these data suggest that the number
+of cycles can change the bacterial community independent of the
+polymerase used to generate the sequences.
+
+When differences in OTUs based on polymerase were investigated for fecal
+samples, no differences could be found at any number of cycles (P-value
+\(\geqslant\) 0.01, corrected P-value = 1). These results suggest that
+taxa specific differences in mock community samples do not carry over to
+actual fecal community samples.
+
+Using 16S rRNA gene sequencing data there was a large disparity in a
+model's ability to correctly classify samples based on polymerase used,
+number of cycles, and individual {[}Figure 5A{]}. The model
+classification success for an individual was significantly better than
+all other models tested {[}Figure 5A \& Table S8{]}. Additionally, the
+top 10 most important variables to the polymerase and number of cycle
+models were not the most important variables in the model classifying
+individuals {[}Figure 5B{]}. This observation shows that OTUs that
+change due to the number of cycles and polymerase used are not the same
+as OTUs that differentiate between individuals {[}Figure 5B{]}. Although
+there are a variety of errors associated with the amplification process,
+they are smaller than the bacterial community variation between
+individuals.
+
+\newpage
+
+\subsection{Discussion}\label{discussion}
+
+Previous studies have reported GC dependent amplification differences
+(Polz \& Cavanaugh, 1998) and we confirm that these differences still
+persist in high fidelity DNA polymerase when analyzing mock community
+samples {[}Figure 1{]}. Additionally, these specific polymerases have
+different sequence error rates and chimera prevalence {[}Figure 2{]}.
+Chimera prevalence in particular was strongly associated with the total
+number of OTUs detected in a mock community {[}Figure 2E{]}. This total
+number of OTUs also varied by polymerase used for these mock community
+samples {[}Figure 3A{]}. The range in the number of OTUs observed
+increased with the number of cycles and the change in range was
+dependent on the polymerase for both mock community and fecal samples
+{[}Figure 3{]}. However, these differences did not translate to large
+bacterial community composition changes based on polymerase {[}Figure
+4{]}. Even though there were bacterial community composition changes
+based on the number of cycles they occurred independently of the
+polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
+differences observed within the number of cycles between any of the
+polymerases for fecal samples. When Random Forest models were created to
+classify number of cycles, polymerase, and individual based on the 16S
+rRNA gene sequencing, the model performance for individuals was
+significantly greater than all other models tested {[}Figure 5A{]}. In
+addition to the better model performance of individuals, the most
+important OTUs to this model were not the most important OTUs for the
+other models tested {[}Figure 5B{]}. Our observations show that
+differences between high fidelity DNA polymerases can be found. However,
+when using 16S rRNA gene sequencing these technical differences are
+dwarfed by biological differences between fecal bacterial communities
+within individuals.
+
+The small differences observed due to error rate and chimera prevalence
+may be due to the actual DNA polymerase family being used within the PCR
+mix. DNA polymerases from distinct families have different binding
+affinities and error correction capacity (Ishino \& Ishino, 2014).
+However, based on the observations within our study this is unlikely to
+be the only contributor. Within our study the highest and lowest chimera
+prevalence both belonged to a family A polymerase (Platinum and
+Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
+Additionally, based on the information supplied by the respective
+manufacturers, the differences between the two PCR mixtures are not
+immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
+DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
+\emph{Taq} antibody. Since it is not possible to know everything about
+the mixture beyond what was willingly provided by the manufacturer, it
+is possible that differences in how the recombinant \emph{Taq} was
+generated or other compounds within the PCR mixture could be a
+contributing factor for the differences in error rate and chimera
+prevalence. Beyond the choice of the type of polymerase, there may be
+other ways to reduce the affect of polymerase dependent error rates and
+chimera prevalence on the downstream results.
+
+Based on polymerase used, our observations generally found no difference
+when using the Bray-Curtis index to measure the bacterial community.
+Specifically, there was no difference in distance between successive
+5-cycle increments within samples between the polymerases when using the
+Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
+is that many of the OTUs generated by polymerase dependent error rates
+and chimera prevalence are likely not highly abundant, allowing the
+Bray-Curtis index to be able to successfully down-weight these OTUs
+(Minchin, 1987). The choice of downstream diversity metric could be an
+important consideration in helping to mitigate some of the observed
+polymerase dependent differences in error rate and chimera prevalence.
+Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
+Vargas, 1996), richness) may be less robust to polymerase dependent
+error rates and chimera prevalence. When choosing a distance metric,
+careful consideration of the biases introduced from the PCR step of the
+16S rRNA gene sequencing workflow need to be taken into account. One
+possible way to better choose polymerases may be based on the DNA
+polymerase family used in the PCR mixture.
+
+Collectively, our data shows that using distinct high fidelity DNA
+polymerases will result in a different number of OTUs being detected.
+The differences in the number of OTUs are primarily due to chimera
+prevalence and sequence error rates that are distinct to the specific
+polymerase. Despite this variation in the number of OTUs, our Random
+Forest models were able to correctly classify individuals regardless of
+the number of cycles and polymerase used. These results suggest that
+even though there is a large variation in the high fidelity DNA
+polymerases used across studies, biological variation similar in effect
+to fecal bacterial community composition differences between individuals
+will still be consistent.
+
+\newpage
+
+\subsection{Conclusion}\label{conclusion}
+
+Although care should always be taken when choosing a polymerase for 16S
+rRNA gene sequencing, our observations show that the differences between
+a variety of polymerases are dwarfed by the actual biological variation
+in fecal communities between individuals. Therefore, if the biological
+signal of interest is similar to differences in fecal bacterial
+communities found between individuals, then the type of high fidelity
+DNA polymerase used will only minimally change the results.
+
+\newpage
+
 \subsection{Materials \& Methods}\label{materials-methods}
@@
 number SRP132931.
-
-\newpage
-
-\subsection{Results}\label{results}
-
-\textbf{\emph{Mock community relative abundance differs by GC content.}}
-There was a significant difference in relative abundance between
-high/low GC content based on either the V4 16S rRNA gene region or the
-whole genome {[}Figure 1 \& Table S1{]}. However, using only the GC
-content of the V4 region resulted in less differences than when using
-the whole genome {[}Table S1{]}. The differences between high and low GC
-content groups were significant as early as 20 cycles in all polymerases
-except Q5 {[}Table S1{]}. The highest (\emph{Staphylococcus}) and lowest
-(\emph{Pseudomonas}) GC content bacteria were the most divergent from
-the expected relative abundance of 0.12 {[}Figure 1{]}. These
-observations suggest that the high fidelity DNA polymerases tested do
-not prevent GC-based amplification error.
-
-\textbf{\emph{Sequence error varies by polymerase and is consistent
-across the number of cycles used.}} The median error rate varied by
-polymerase, with Kappa having the highest error rate of all the
-polymerases across the number of cycles {[}Figure 2A \& Table S2{]}.
-Thus, the majority of the differences observed across the number of
-cycles was between Kappa and other polymerases {[}Figure 2A \& Table
-S3{]}. The total sequences that contained at least one error was also
-polymerase dependent, with the majority of differences being between
-Kappa or Accuprime and the other polymerases {[}Table S2 \& S3{]}. These
-differences in error rates were not due to polymerase dependent
-differences in base substitution rate {[}Figure S1{]}. Collectively, the
-results suggest that sequence error is dependent on polymerase, persists
-across the number of cycles used, and are not due to any bias towards a
-specific base substitution.
-
-\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
-and correlate with the number of OTUs.}} Based on the previous results,
-we examined whether chimeras also were dependent on polymerase and
-whether this could affect the total number of OTUs. We observed
-significant differences in the chimera prevalence based on polymerase
-across the number of cycles used (P-value \textless{} 0.05) {[}Table
-S2{]}. Differences in chimera prevalence between Platinum and all other
-polymerases accounted for the majority of these differences {[}Table
-S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
-of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
-increased chimera prevalence for all polymerases but the rate of
-increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
-plateau in the total percent of chimeras that were removed that was
-similar for all polymerases {[}Figure 2D{]}. A positive correlation was
-observed between chimeric sequences and the number of OTUs for all
-polymerases {[}Figure 2E{]}. This positive correlation was strongest for
-Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
-This data suggests that chimera prevalence depends on polymerase, is
-made worse by increasing the number of cycles, and confirms that the
-number of OTUs is dependent on the prevalence of these chimeric
-sequences.
-
-\textbf{\emph{The number of OTUs generated are dependent on polymerase
-used.}} In addition to the sequence error and chimera differences
-between polymerases, we also investigated whether there were polymerase
-or number of cycle dependent effects on the observed number of OTUs.
-There was a polymerase dependent difference in the range of the number
-of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
-of the polymerases came to a total of 8 OTUs, created by running the
-mock reference 16S sequences through mothur processing, was at 25 and 30
-cycles {[}Figure 3A{]}. Differences between polymerases for the number
-of OTUs created were observed as early as 20 cycles in the mock
-community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
-Tukey post-hoc test, the majority of differences for the number of OTUs
-detected in the mock community was largely due to Kappa and Platinum
-versus the other polymerases across the number of cycles used {[}Table
-S5{]}. Similarly, differences in the range of the number of OTUs
-detected for fecal samples was dependent on the polymerase used
-(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
-- 136) {[}Figure 3B{]}. There was also a trend for lower number of
-cycles (15-20) to result in a reduced range in the number of OTUs versus
-higher number of cycles (25, 30, and 35) for all polymerases
-(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
-136) {[}Figure 3B{]}. Unlike mock community samples, an overall
-difference in the number of OTUs was only detected between polymerases
-at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
-9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
-any differences between specific polymerase groups (P-value
-\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
-noticeably larger difference between polymerases in the number of OTUs
-than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
-consistent across the number of amplification cycles tested {[}Figure
-3{]}. Based on these observations, it is clear that larger differences
-in total number of OTUs based on polymerases are found in mock versus
-fecal communities.
-
-\textbf{\emph{The bacterial communities generated by the polymerases
-were similar but varied by number of cycles.}} Next, we investigated the
-effect polymerases and number of cycles had on the bacterial community
-composition in more detail by using the Bray-Curtis index. We observed
-no difference in the bacterial community composition between any measure
-tested when using the mock community (P-value \textgreater{} 0.05)
-{[}Figure 4A{]}. With respect to fecal samples, within each respective
-5-cycle increment comparison there was no difference in Bray-Curtis
-index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
-4B{]}. Using PERMANOVA to test for community differences based on any of
-the number of cycles within polymerases, only Phusion had cycle
-dependent differences (P-value = 0.03). Additionally, Phusion was one of
-two polymerases that had enough sequences to be rarefied to 1000
-sequences at 15 cycles. There was an overall decrease in Bray-Curtis
-index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
-compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
-0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus
-20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
-Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
-(P-value \textless{} 0.05). Overall, these data suggest that the number
-of cycles can change the bacterial community independent of the
-polymerase used to generate the sequences.
-
-\textbf{\emph{Polymerase dependent differences in taxa are more
-pronounced in mock versus fecal samples.}} Finding minimal differences
-in overall bacterial community composition, we next examined if specific
-taxa vary depending on polymerase and number of cycles. Further
-investigation of mock community samples found that OTUs taxonomically
-classified to \emph{Salmonella}, \emph{Escherichia},
-\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
-polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
-members of the mock community and the differences were most pronounced
-at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
-between taxa were due to differences between Kappa and the other
-polymerases {[}Table S7{]}. When differences in OTUs based on polymerase
-were investigated for fecal samples, no differences could be found at
-any number of cycles (P-value \(\geqslant\) 0.01, corrected P-value =
-1). These results suggest that taxa specific differences in mock
-community samples do not carry over to actual fecal community samples.
-
-\textbf{\emph{Variation due to polymerase and number of cycles is
-dwarfed by the biological variation of an individual.}} Using 16S rRNA
-gene sequencing data there was a large disparity in a model's ability to
-correctly classify samples based on polymerase used, number of cycles,
-and individual {[}Figure 5A{]}. The model classification success for an
-individual was significantly better than all other models tested
-{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most important
-variables to the polymerase and number of cycle models were not the most
-important variables in the model classifying individuals {[}Figure
-5B{]}. This observation shows that OTUs that change due to the number of
-cycles and polymerase used are not the same as OTUs that differentiate
-between individuals {[}Figure 5B{]}. Although there are a variety of
-errors associated with the amplification process, they are smaller than
-the bacterial community variation between individuals.
-
-\newpage
-
-\subsection{Discussion}\label{discussion}
-
-Previous studies have reported GC dependent amplification differences
-(Polz \& Cavanaugh, 1998) and we confirm that these differences still
-persist in high fidelity DNA polymerase when analyzing mock community
-samples {[}Figure 1{]}. Additionally, these specific polymerases have
-different sequence error rates and chimera prevalence {[}Figure 2{]}.
-Chimera prevalence in particular was strongly associated with the total
-number of OTUs detected in a mock community {[}Figure 2E{]}. This total
-number of OTUs also varied by polymerase used for these mock community
-samples {[}Figure 3A{]}. The range in the number of OTUs observed
-increased with the number of cycles and the change in range was
-dependent on the polymerase for both mock community and fecal samples
-{[}Figure 3{]}. However, these differences did not translate to large
-bacterial community composition changes based on polymerase {[}Figure
-4{]}. Even though there were bacterial community composition changes
-based on the number of cycles they occurred independently of the
-polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
-differences observed within the number of cycles between any of the
-polymerases for fecal samples. When Random Forest models were created to
-classify number of cycles, polymerase, and individual based on the 16S
-rRNA gene sequencing, the model performance for individuals was
-significantly greater than all other models tested {[}Figure 5A{]}. In
-addition to the better model performance of individuals, the most
-important OTUs to this model were not the most important OTUs for the
-other models tested {[}Figure 5B{]}. Our observations show that
-differences between high fidelity DNA polymerases can be found. However,
-when using 16S rRNA gene sequencing these technical differences are
-dwarfed by biological differences between fecal bacterial communities
-within individuals.
-
-The small differences observed due to error rate and chimera prevalence
-may be due to the actual DNA polymerase family being used within the PCR
-mix. DNA polymerases from distinct families have different binding
-affinities and error correction capacity (Ishino \& Ishino, 2014).
-However, based on the observations within our study this is unlikely to
-be the only contributor. Within our study the highest and lowest chimera
-prevalence both belonged to a family A polymerase (Platinum and
-Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
-Additionally, based on the information supplied by the respective
-manufacturers, the differences between the two PCR mixtures are not
-immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
-DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
-\emph{Taq} antibody. Since it is not possible to know everything about
-the mixture beyond what was willingly provided by the manufacturer, it
-is possible that differences in how the recombinant \emph{Taq} was
-generated or other compounds within the PCR mixture could be a
-contributing factor for the differences in error rate and chimera
-prevalence. Beyond the choice of the type of polymerase, there may be
-other ways to reduce the affect of polymerase dependent error rates and
-chimera prevalence on the downstream results.
-
-Based on polymerase used, our observations generally found no difference
-when using the Bray-Curtis index to measure the bacterial community.
-Specifically, there was no difference in distance between successive
-5-cycle increments within samples between the polymerases when using the
-Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
-is that many of the OTUs generated by polymerase dependent error rates
-and chimera prevalence are likely not highly abundant, allowing the
-Bray-Curtis index to be able to successfully down-weight these OTUs
-(Minchin, 1987). The choice of downstream diversity metric could be an
-important consideration in helping to mitigate some of the observed
-polymerase dependent differences in error rate and chimera prevalence.
-Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
-Vargas, 1996), richness) may be less robust to polymerase dependent
-error rates and chimera prevalence. When choosing a distance metric,
-careful consideration of the biases introduced from the PCR step of the
-16S rRNA gene sequencing workflow need to be taken into account. One
-possible way to better choose polymerases may be based on the DNA
-polymerase family used in the PCR mixture.
-
-Collectively, our data shows that using distinct high fidelity DNA
-polymerases will result in a different number of OTUs being detected.
-The differences in the number of OTUs are primarily due to chimera
-prevalence and sequence error rates that are distinct to the specific
-polymerase. Despite this variation in the number of OTUs, our Random
-Forest models were able to correctly classify individuals regardless of
-the number of cycles and polymerase used. These results suggest that
-even though there is a large variation in the high fidelity DNA
-polymerases used across studies, biological variation similar in effect
-to fecal bacterial community composition differences between individuals
-will still be consistent.
-
-\newpage
-
-\subsection{Conclusion}\label{conclusion}
-
-Although care should always be taken when choosing a polymerase for 16S
-rRNA gene sequencing, our observations show that the differences between
-a variety of polymerases are dwarfed by the actual biological variation
-in fecal communities between individuals. Therefore, if the biological
-signal of interest is similar to differences in fecal bacterial
-communities found between individuals, then the type of high fidelity
-DNA polymerase used will only minimally change the results.
 

### 2018-07-13 submission/manuscript.Rmd
revamp results section

@@
 
+fecal_sample_rf_summary <- read_csv("../data/process/tables/sample_group_rf_model_summary.csv") 
+fecal_poly_rf_summary <- read_csv("../data/process/tables/polymerase_group_rf_model_summary.csv")
+fecal_cycles_rf_summary <- read_csv("../data/process/tables/cycles_group_rf_model_summary.csv")
+
 dunn_rf_data <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv")
@@
 These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
-Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.
+Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific type of base substitution.
 
-Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
+We next examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
 We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2].
@@
 This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E].
-This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.
+This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms previous reports that the number of OTUs is dependent on the prevalence of these chimeric sequences.
 
@@
 
-Next, we investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
+We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
 We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A].
@@
 These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7].
-The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases [Table S7].
+The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases tested [Table S7].
 
+Based on these observations, we created Random Forest models to classify samples based on number of cycles or polymerase used.
 When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`].
@@
 
-***Few differences based on number of cycles and polymerased used were identified in fecal samples.***
-Similarly, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
+***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
+Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5].
+However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
 Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
@@
 Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
-Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerase groups (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5].
-Overall, mock communities showed a noticeably larger difference between polymerases in the number of OTUs than actual fecal samples [Figure 3 & Table S4-S5].
-This was consistent across the number of amplification cycles tested [Figure 3].
-Based on these observations, it is clear that larger differences in total number of OTUs based on polymerases are found in mock versus fecal communities.
+Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.
 
-With respect to fecal samples, within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
-Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
+When using the Bray-Curtis index, we observed that the affect polymerases and number of cycles had on the bacterial community composition in fecal samples was minimal.
+Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
+When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
 Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
-There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
+There was also an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
 the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B].
 Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.
+In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
+Overall, these results suggest that in fecal samples the number of cycles and polymerase used only minimally change the bacterial community composition.
 
-When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found at any number of cycles (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
-These results suggest that taxa specific differences in mock community samples do not carry over to actual fecal community samples.
-
-Using 16S rRNA gene sequencing data there was a large disparity in a model's ability to correctly classify samples based on polymerase used, number of cycles, and individual [Figure 5A].
-The model classification success for an individual was significantly better than all other models tested [Figure 5A & Table S8].
-Additionally, the top 10 most important variables to the polymerase and number of cycle models were not the most important variables in the model classifying individuals [Figure 5B].
-This observation shows that OTUs that change due to the number of cycles and polymerase used are not the same as OTUs that differentiate between individuals [Figure 5B].
-Although there are a variety of errors associated with the amplification process, they are smaller than the bacterial community variation between individuals.
+Finally, we built Random Forest models to classify samples for the number of cycles, polymerase used, or individual it was obtained from.
+Using 16S rRNA gene sequencing data there was a large disparity in the model's ability to correctly classify samples based on polymerase used, number of cycles, and individual (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal") & str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
+There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)).
+Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5B].
+This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5B].
+Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.
 

### 2018-07-13 submission/manuscript.tex
revamp results section

@@
 is dependent on polymerase, persists across the number of cycles used,
-and are not due to any bias towards a specific base substitution.
+and are not due to any bias towards a specific type of base
+substitution.
 
-Based on the previous results, we examined whether chimeras also were
-dependent on polymerase and whether this could affect the total number
-of OTUs. We observed significant differences in the chimera prevalence
-based on polymerase across the number of cycles used (P-value
-\textless{} 0.05) {[}Table S2{]}. Differences in chimera prevalence
-between Platinum and all other polymerases accounted for the majority of
-these differences {[}Table S3{]}. Accuprime\textsuperscript{TM} had the
-lowest chimera prevalence of all polymerases regardless of whether
-chimera removal with VSEARCH was used {[}Figure 2B \& 2C{]}. The number
-of cycles used clearly increased chimera prevalence for all polymerases
-but the rate of increase differed {[}Figure 2B \& 2C{]}. Additionally,
-there was a plateau in the total percent of chimeras that were removed
-that was similar for all polymerases {[}Figure 2D{]}. A positive
-correlation was observed between chimeric sequences and the number of
-OTUs for all polymerases {[}Figure 2E{]}. This positive correlation was
-strongest for Accuprime\textsuperscript{TM}, Platinum, and Phusion
-{[}Figure 2E{]}. This data suggests that chimera prevalence depends on
-polymerase, is made worse by increasing the number of cycles, and
-confirms that the number of OTUs is dependent on the prevalence of these
+We next examined whether chimeras also were dependent on polymerase and
+whether this could affect the total number of OTUs. We observed
+significant differences in the chimera prevalence based on polymerase
+across the number of cycles used (P-value \textless{} 0.05) {[}Table
+S2{]}. Differences in chimera prevalence between Platinum and all other
+polymerases accounted for the majority of these differences {[}Table
+S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
+of all polymerases regardless of whether chimera removal with VSEARCH
+was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
+increased chimera prevalence for all polymerases but the rate of
+increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
+plateau in the total percent of chimeras that were removed that was
+similar for all polymerases {[}Figure 2D{]}. A positive correlation was
+observed between chimeric sequences and the number of OTUs for all
+polymerases {[}Figure 2E{]}. This positive correlation was strongest for
+Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
+This data suggests that chimera prevalence depends on polymerase, is
+made worse by increasing the number of cycles, and confirms previous
+reports that the number of OTUs is dependent on the prevalence of these
 chimeric sequences.
@@
 
-Next, we investigated the effect polymerases and number of cycles had on
+We next investigated the effect polymerases and number of cycles had on
 the bacterial community composition in more detail by using the
@@
 differences between taxa were due to differences between a single
-polymerase (Kappa) and the other polymerases {[}Table S7{]}.
+polymerase (Kappa) and the other polymerases tested {[}Table S7{]}.
 
-When assessing the Random Forest models there was a difference between
-models built to classify number of cycles versus those built to classify
-polymerase used {[}P-value = 5.06e-07{]}. Although there was a
-difference, neither model was successful at correctly classifying
-samples based on 16S rRNA gene sequencing data (cycles model,
-probability of correct classification = 0.42 (0.4 - 0.43)(min - max)
-versus polymerase model, probability of correct classification = 0.39
-(0.37 - 0.4)). Our observations in mock communities suggest that
-although differences are detected based on number of cycles or
-polymerase used, the overall community-wide differences are subtle.
+Based on these observations, we created Random Forest models to classify
+samples based on number of cycles or polymerase used. When assessing the
+Random Forest models there was a difference between models built to
+classify number of cycles versus those built to classify polymerase used
+{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
+was successful at correctly classifying samples based on 16S rRNA gene
+sequencing data (cycles model, probability of correct classification =
+0.42 (0.4 - 0.43)(min - max) versus polymerase model, probability of
+correct classification = 0.39 (0.37 - 0.4)). Our observations in mock
+communities suggest that although differences are detected based on
+number of cycles or polymerase used, the overall community-wide
+differences are subtle.
 
-\textbf{\emph{Few differences based on number of cycles and polymerased
-used were identified in fecal samples.}} Similarly, differences in the
-range of the number of OTUs detected for fecal samples was dependent on
-the polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus
-Phusion at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend
-for lower number of cycles (15-20) to result in a reduced range in the
+\textbf{\emph{Few differences based on number of cycles and polymerase
+used were identified in fecal samples.}} Unlike mock community samples,
+an overall difference in the number of OTUs was only detected between
+polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
+P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
+identify any differences between specific polymerases used (P-value
+\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
+of the number of OTUs detected for fecal samples was dependent on the
+polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
+at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend for
+lower number of cycles (15-20) to result in a reduced range in the
 number of OTUs versus higher number of cycles (25, 30, and 35) for all
 polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
-cycles = 84 - 136) {[}Figure 3B{]}. Unlike mock community samples, an
-overall difference in the number of OTUs was only detected between
-polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
-P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
-identify any differences between specific polymerase groups (P-value
-\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
-noticeably larger difference between polymerases in the number of OTUs
-than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
-consistent across the number of amplification cycles tested {[}Figure
-3{]}. Based on these observations, it is clear that larger differences
-in total number of OTUs based on polymerases are found in mock versus
-fecal communities.
+cycles = 84 - 136) {[}Figure 3B{]}. Based on these observations, there
+are only small differences in the total number of OTUs detected in fecal
+samples that are based on number of cycles or polymerases used.
 
-With respect to fecal samples, within each respective 5-cycle increment
-comparison there was no difference in Bray-Curtis index between the
-polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}. Using
-PERMANOVA to test for community differences based on any of the number
-of cycles within polymerases, only Phusion had cycle dependent
-differences (P-value = 0.03). Additionally, Phusion was one of two
-polymerases that had enough sequences to be rarefied to 1000 sequences
-at 15 cycles. There was an overall decrease in Bray-Curtis index by
-cycle comparison group (e.g.~the 15 cycle versus 20 cycle group compared
-to the 30 cycle versus 35 cycle group) (P-value \textless{} 0.01)
-{[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus 20
-cycle and 20 cycle versus 25 cycle comparison groups had a higher
-Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
-(P-value \textless{} 0.05). Overall, these data suggest that the number
-of cycles can change the bacterial community independent of the
-polymerase used to generate the sequences.
+When using the Bray-Curtis index, we observed that the affect
+polymerases and number of cycles had on the bacterial community
+composition in fecal samples was minimal. Within each respective 5-cycle
+increment comparison there was no difference in Bray-Curtis index
+between the polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}.
+When using PERMANOVA to test for community differences based on any of
+the number of cycles within polymerases, only Phusion had cycle
+dependent differences (P-value = 0.03). Additionally, Phusion was one of
+two polymerases that had enough sequences to be rarefied to 1000
+sequences at 15 cycles. There was also an overall decrease in
+Bray-Curtis index by cycle comparison group (e.g.~the 15 cycle versus 20
+cycle group compared to the 30 cycle versus 35 cycle group) (P-value
+\textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15
+cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had
+a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison
+group (P-value \textless{} 0.05). In addition to assessing community
+composition by Bray-Curtis index, we also investigated differences in
+OTUs based on polymerase at each number of cycle used and no significant
+differences could be found (P-value \(\geqslant\) 0.01, corrected
+P-value = 1). Overall, these results suggest that in fecal samples the
+number of cycles and polymerase used only minimally change the bacterial
+community composition.
 
-When differences in OTUs based on polymerase were investigated for fecal
-samples, no differences could be found at any number of cycles (P-value
-\(\geqslant\) 0.01, corrected P-value = 1). These results suggest that
-taxa specific differences in mock community samples do not carry over to
-actual fecal community samples.
-
-Using 16S rRNA gene sequencing data there was a large disparity in a
+Finally, we built Random Forest models to classify samples for the
+number of cycles, polymerase used, or individual it was obtained from.
+Using 16S rRNA gene sequencing data there was a large disparity in the
 model's ability to correctly classify samples based on polymerase used,
-number of cycles, and individual {[}Figure 5A{]}. The model
-classification success for an individual was significantly better than
-all other models tested {[}Figure 5A \& Table S8{]}. Additionally, the
-top 10 most important variables to the polymerase and number of cycle
-models were not the most important variables in the model classifying
-individuals {[}Figure 5B{]}. This observation shows that OTUs that
-change due to the number of cycles and polymerase used are not the same
-as OTUs that differentiate between individuals {[}Figure 5B{]}. Although
-there are a variety of errors associated with the amplification process,
-they are smaller than the bacterial community variation between
-individuals.
+number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
+was a clear difference in the success of the model used to classify
+individuals (probability of correct classification = 0.87 (0.86 - 0.9))
+versus the models created to classify number of cycles or polymerase
+(cycles model, probability of correct classification = 0.26 (0.25 -
+0.27)(min - max) and polymerase model, probability of correct
+classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
+important OTUs to the polymerase and number of cycle models were not the
+most important OTUs to the model that classified individuals {[}Figure
+5B{]}. This suggests that OTUs that change the most due to the number of
+cycles and polymerase in fecal samples are not the same as OTUs that are
+part of the biological variation between individuals {[}Figure 5B{]}.
+Collectively, our observations show that despite a variety of errors
+associated with the amplification process, it is still possible to
+detect the bacterial community variation between individuals.
 

### 2018-07-13 code/figure3.R
add updated figs

@@ -1,0 +1,123 @@
+### Create mock specific community graph
+### Mock Community only
+### Marc Sze
+
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "viridis", "gridExtra"))
+
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = ""))
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+# Vector of sub samples used
+sub_sample_level <- c("50", "100", "500", "1000", "5000", "10000")
+
+# Read in the count data
+numOTU_data <- sapply(sub_sample_level, 
+                      function(x) 
+                        read_data("data/process/tables/", 
+                                  "mock_sub_sample_", "_count_table.csv", x) %>% 
+                        mutate(depth_level = as.numeric(x)), 
+                      simplify = F) %>% bind_rows() %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")))
+
+
+summary_data <- numOTU_data %>% 
+  group_by(depth_level, taq, cycles) %>% 
+  summarise(taq_mean = median(numOTUs, na.rm = T), 
[…]

### 2018-07-13 code/figure4.R

@@ -1,0 +1,123 @@
+### Create fecal specific community graph
+### Fecal Community only
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+loadLibs(c("tidyverse", "stringr", "viridis", "gridExtra"))
+
+###########################################################################################################################
+############################### List of functions to make analysis work ###################################################
+###########################################################################################################################
+
+# Function to read in the data
+read_data <- function(pathing, start_name, end_name, differentiator){
+  # pathing is the variable that controls the path
+  # start_name is the beginning of the file
+  # end_name is how the file ends (e.g. .csv)
+  # differentiator is the unique component of the file name
+  
+  # reads in the respective file 
+  tempData <- read_csv(paste(pathing, start_name, differentiator, end_name, sep = "")) %>% 
+    filter(sample_name %in% c("DA10001", "DA10016", "DA10029", "DA10040")) %>% 
+    mutate(cycle_num = as.numeric(str_replace(cycles, "x", "")))
+  
+  # returns the file to the global environment
+  return(tempData)
+  
+}
+
+###########################################################################################################################
+############################### Run actual analysis programs  #############################################################
+###########################################################################################################################
+
+# Vector of sub samples used
+sub_sample_level <- c("1000", "5000", "10000", "15000", "20000")
+
+# Read in the count data
+numOTU_data <- sapply(sub_sample_level, 
+                      function(x) read_data("data/process/tables/", "fecal_sub_sample_", "_count_table.csv", x) %>% 
+                        mutate(depth_level = as.numeric(x)), 
+                      simplify = F) %>% bind_rows() %>% 
+  mutate(taq = factor(taq, 
+                      levels = c("ACC", "K", "PHU", "PL", "Q5"), 
+                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")), 
+         cycle_num = factor(cycle_num, 
[…]

### 2018-07-13 code/make_rf_fecal_imp_vars.graph.R
modify rf fig

@@ -78,7 +78,7 @@ imp_taxa_graph <- rf_data %>%
                             labels = c("Cycles", "Polymerase", "Cycles & Polymerase"))) %>% 
   ggplot(aes(otu, median_mda, color = other_model, shape = other_model, fill = other_model)) + 
   # geom_hline(yintercept = 0, linetype = "dashed") + 
-  geom_point(size = 3) + ggtitle("B") + coord_cartesian(ylim = c(0, 7)) + 
+  geom_point(size = 3) + coord_cartesian(ylim = c(0, 7)) + 
   theme_bw() + labs(x = "", y = "Median MDA") + 
   scale_shape_manual(name = "", values = c(1, 21, 21, 21)) + 
   scale_color_manual(name = "", values = c('#8B8386', '#B0171F', '#0000FF', '#308014')) + 
@@ -112,6 +112,6 @@ imp_taxa_graph <- rf_data %>%
 
 test <- grid.arrange(model_graph, imp_taxa_graph, layout_matrix = rbind(c(1, 1, 2, 2, 2)))
 
-ggsave("results/figures/Figure5.pdf", test, width = 11, height = 5, dpi = 300)
+ggsave("results/figures/Figure5.pdf", imp_taxa_graph, width = 6.5, height = 5, dpi = 300)
 
              

### 2018-07-13 submission/manuscript.Rmd
update figure numbers

@@
 The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C].
-Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D].
-A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E].
+There was also a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D].
+Additionally, a positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E].
 This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E].
@@
 We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
-We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A].
+We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
 Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles.
@@
 However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
-Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
+Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A].
 There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g.
-Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B].
+Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A].
 Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.
@@
 There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)).
-Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5B].
-This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5B].
+Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5].
+This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5].
 Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.

### 2018-07-13 submission/manuscript.tex
update figure numbers

@@
 increased chimera prevalence for all polymerases but the rate of
-increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
-plateau in the total percent of chimeras that were removed that was
-similar for all polymerases {[}Figure 2D{]}. A positive correlation was
+increase differed {[}Figure 2B \& 2C{]}. There was also a plateau in the
+total percent of chimeras that were removed that was similar for all
+polymerases {[}Figure 2D{]}. Additionally, a positive correlation was
 observed between chimeric sequences and the number of OTUs for all
@@
 composition between any measure tested when using the mock community
-(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
+(P-value \textgreater{} 0.05) {[}Figure 3B{]}. Finding minimal
 differences in overall bacterial community composition, we next examined
@@
 polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
-at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend for
+at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
 lower number of cycles (15-20) to result in a reduced range in the
@@
 polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
-cycles = 84 - 136) {[}Figure 3B{]}. Based on these observations, there
+cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
 are only small differences in the total number of OTUs detected in fecal
@@
 most important OTUs to the model that classified individuals {[}Figure
-5B{]}. This suggests that OTUs that change the most due to the number of
+5{]}. This suggests that OTUs that change the most due to the number of
 cycles and polymerase in fecal samples are not the same as OTUs that are
-part of the biological variation between individuals {[}Figure 5B{]}.
+part of the biological variation between individuals {[}Figure 5{]}.
 Collectively, our observations show that despite a variety of errors

### 2018-07-13 submission/manuscript.Rmd
update figure legends

@@
 
-**Figure 3: The number and range of OTUs in mock and fecal samples varies by polymerase.** The points represent the median number of OTUs for the mock samples.
-The lines represent the range of the minimum and maximum number of OTUs detected within the four fecal samples.
+**Figure 3: Subtle differences  based on number of cycles and polymerase used are detected in bacterial community composition of mock samples.** A) The range in the number of OTUs detected in the mock samples increased as cycle number increased.
+This range was larger for specific HiFi DNA polymerases.
+The points represent the median number of OTUs and the lines represent the range from the minimum to maximum number of OTUs detected within the four technical replicates.
 The dotted black line represents the number of OTUs detected when only the references sequences for the mock community are clustered.
-A) The range in the number of OTUs detected in the mock samples increased as cycle number increased.
-This range was also larger for specific HiFi DNA polymerases.
-B) The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
-This increased range also was larger for specific HiFi DNA polymerases.
+A) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
+The lines represent the range of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle increment comparison.
+The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.
 
-**Figure 4: Lower five-cycle intervals have larger differences than higher five-cycle intervals using Bray-Curtis.** A) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
+**Figure 4: Subtle differences based on number of cycles and polymerase used are detected in bacterial community composition of fecal samples.** A) The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
+This range was larger for specific HiFi DNA polymerases.
+The points represent the median number of OTUs and the lines represent the range from the minimum to maximum number of OTUs detected within the four fecal samples.
 B) Within person differences based on the next 5-cycle PCR interval in fecal samples.
@@
 
-**Figure 5: Biological variation in bacterial community composition due to individual was larger than number of cycles and polymerase used.** A) Using 16S rRNA gene sequencing, the classification of individuals was higher than any other model tested.
-The lines represent the median probability of correct classification and the gray points represent the individual probability from one run out of the 100 10-fold CV runs.
-B) The most important OTUs to the model classifying individuals are not the same as those that are most important for the number of cycles or polymerase used.
+**Figure 5: Important OTUs for the number of cycles and polymerase models are not the most important OTUs in the model classifying individuals.** Color highlights the top 10 OTUs in number of cycles and polymerase used and where they fall in the model used to classify individuals.
+The majority of these top 10 OTUs have 0 importance to the model used to classify individuals.
 

### 2018-07-13 submission/manuscript.tex
update figure legends

@@
 
-\textbf{Figure 3: The number and range of OTUs in mock and fecal samples
-varies by polymerase.} The points represent the median number of OTUs
-for the mock samples. The lines represent the range of the minimum and
-maximum number of OTUs detected within the four fecal samples. The
-dotted black line represents the number of OTUs detected when only the
-references sequences for the mock community are clustered. A) The range
-in the number of OTUs detected in the mock samples increased as cycle
-number increased. This range was also larger for specific HiFi DNA
-polymerases. B) The range in the number of OTUs detected in the
-different fecal samples increased as cycle number increased. This
-increased range also was larger for specific HiFi DNA polymerases.
-
-\textbf{Figure 4: Lower five-cycle intervals have larger differences
-than higher five-cycle intervals using Bray-Curtis.} A) Within replicate
-difference based on the next 5-cycle PCR interval in mock samples. B)
-Within person differences based on the next 5-cycle PCR interval in
-fecal samples. The points represent the median Bray-Curtis index for the
+\textbf{Figure 3: Subtle differences based on number of cycles and
+polymerase used are detected in bacterial community composition of mock
+samples.} A) The range in the number of OTUs detected in the mock
+samples increased as cycle number increased. This range was larger for
+specific HiFi DNA polymerases. The points represent the median number of
+OTUs and the lines represent the range from the minimum to maximum
+number of OTUs detected within the four technical replicates. The dotted
+black line represents the number of OTUs detected when only the
+references sequences for the mock community are clustered. A) Within
+replicate difference based on the next 5-cycle PCR interval in mock
 samples. The lines represent the range of the minimum and maximum
@@
 
-\textbf{Figure 5: Biological variation in bacterial community
-composition due to individual was larger than number of cycles and
-polymerase used.} A) Using 16S rRNA gene sequencing, the classification
-of individuals was higher than any other model tested. The lines
-represent the median probability of correct classification and the gray
-points represent the individual probability from one run out of the 100
-10-fold CV runs. B) The most important OTUs to the model classifying
-individuals are not the same as those that are most important for the
-number of cycles or polymerase used.
+\textbf{Figure 4: Subtle differences based on number of cycles and
+polymerase used are detected in bacterial community composition of fecal
+samples.} A) The range in the number of OTUs detected in the different
+fecal samples increased as cycle number increased. This range was larger
+for specific HiFi DNA polymerases. The points represent the median
+number of OTUs and the lines represent the range from the minimum to
+maximum number of OTUs detected within the four fecal samples. B) Within
+person differences based on the next 5-cycle PCR interval in fecal
+samples. The points represent the median Bray-Curtis index for the
+samples. The lines represent the range of the minimum and maximum
+Bray-Curtis index value for each PCR 5-cycle increment comparison. The
+closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar
+the bacterial community is of the two compared number of cycles.
+
+\textbf{Figure 5: Important OTUs for the number of cycles and polymerase
+models are not the most important OTUs in the model classifying
+individuals.} Color highlights the top 10 OTUs in number of cycles and
+polymerase used and where they fall in the model used to classify
+individuals. The majority of these top 10 OTUs have 0 importance to the
+model used to classify individuals.
 

### 2018-07-16 code/run_nmds_mock_data.R
update permanova with a nested design analysis

@@ -340,6 +340,21 @@ permanova_results <- sapply(sub_sample_level,
                             function(x) get_permanova(x, sep_meta_list, sep_red_bray_dist), simplify = F) %>% 
   bind_rows()
 
+# Run the PERMANOVA with both variables included
+samples_to_keep <- meta_list[["1000"]]$full_name
+bray_all_1000 <- braycurtis_dist[["1000"]][samples_to_keep, samples_to_keep]
+meta_1000_only <- meta_list[["1000"]]
+
+set.seed(12345)
+combo_test <- adonis(as.dist(bray_all_1000) ~ 
+#                 factor(meta_1000_only$cycles) + factor(meta_1000_only$taq) + 
+                  factor(meta_1000_only$taq)*factor(meta_1000_only$cycles), permutations = 9999)$aov.tab %>% 
+  mutate(x1 = case_when(
+    str_detect(rownames(.), "cycles") & str_detect(rownames(.), "taq") == F ~ "cycles", 
+    str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") == F ~ "polymerase", 
+    str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") ~ "interaction", 
+    TRUE ~ rownames(.)))
+
 # Write out results 
 write_csv(kruskal_test_data, "data/process/tables/mock_bray_sim_to_prev_cycle_kruskal_results.csv")
 write_csv(permanova_results, "data/process/tables/mock_bray_permanova_by_taq_results.csv")
@@ -349,4 +364,4 @@ sapply(c(1:length(finalized_tables)),
                              paste("data/process/tables/mock_bray_5_cycle_dist_", sub_sample_level[x], "_data.csv", sep = "")))
 
 
-
+write_csv(combo_test, "data/process/tables/mock_bray_combined_permanova_results.csv")

### 2018-07-16 code/run_nmds_mock_data.R
update permanova formula

@@ -347,12 +347,11 @@ meta_1000_only <- meta_list[["1000"]]
 
 set.seed(12345)
 combo_test <- adonis(as.dist(bray_all_1000) ~ 
-#                 factor(meta_1000_only$cycles) + factor(meta_1000_only$taq) + 
-                  factor(meta_1000_only$taq)*factor(meta_1000_only$cycles), permutations = 9999)$aov.tab %>% 
+                  factor(meta_1000_only$taq)/factor(meta_1000_only$cycles), permutations = 9999)$aov.tab %>% 
   mutate(x1 = case_when(
     str_detect(rownames(.), "cycles") & str_detect(rownames(.), "taq") == F ~ "cycles", 
     str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") == F ~ "polymerase", 
-    str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") ~ "interaction", 
+    str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") ~ "interaction_polymerase_cycles", 
     TRUE ~ rownames(.)))
 
 # Write out results 

### 2018-07-16 code/run_nmds_fecal_data.R
update permanova comparison

@@ -270,6 +270,25 @@ permanova_results <- sapply(sub_sample_level,
                function(x) get_permanova(x, sep_meta_list, sep_red_bray_dist), simplify = F) %>% 
   bind_rows()
 
+# Run the PERMANOVA with both variables included
+meta_1000_only <- meta_list[["1000"]]
+samples_to_keep <- meta_1000_only$full_name
+bray_all_1000 <- braycurtis_dist[["1000"]][samples_to_keep, samples_to_keep]
+
+
+set.seed(12345)
+combo_test <- adonis(as.dist(bray_all_1000) ~ 
+                       factor(meta_1000_only$sample_name) * factor(meta_1000_only$taq)/factor(meta_1000_only$cycles), 
+                     permutations = 9999)$aov.tab %>% 
+  mutate(x1 = case_when(
+    str_detect(rownames(.), "sample_name") & str_detect(rownames(.), "taq") == F ~ "individual", 
+    str_detect(rownames(.), "taq") & str_detect(rownames(.), "cycles") == F & 
+      str_detect(rownames(.), "sample_name") == F ~ "polymerase", 
+    str_detect(rownames(.), "sample_name") & str_detect(rownames(.), "cycles") ~ "interaction_ind_poly_cycle",
+    str_detect(rownames(.), "sample_name") & str_detect(rownames(.), "cycles") == F ~ "interaction_ind_poly", 
+    TRUE ~ rownames(.)))
+
+
 # Write out results 
 write_csv(kruskal_test_data, "data/process/tables/bray_sim_to_prev_cycle_kruskal_results.csv")
 write_csv(permanova_results, "data/process/tables/bray_permanova_by_taq_results.csv")
@@ -279,4 +298,4 @@ sapply(c(1:length(finalized_tables)),
                              paste("data/process/tables/bray_5_cycle_dist_", sub_sample_level[x], "_data.csv", sep = "")))
 
 
-
+write_csv(combo_test, "data/process/tables/fecal_bray_combined_permanova_results.csv")

### 2018-07-16 submission/manuscript.Rmd
update results

@@
 bray_permanova_by_taq_results <- as.data.frame(read_csv("../data/process/tables/bray_permanova_by_taq_results.csv"))
+
+bray_mock_combined <- read_csv("../data/process/tables/mock_bray_combined_permanova_results.csv")
+bray_fecal_combined <- read_csv("../data/process/tables/fecal_bray_combined_permanova_results.csv")
 
@@
 There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-When analyzing fecal samples we observed similar differences in the number of OTUs.
-However, the median number of OTUs only varied by HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+When analyzing fecal samples we observed smaller differences in the number of OTUs.
+Additionally, the median number of OTUs only varied by HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
 Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
@@
 
-**Conclusions.** Although there are 16S rRNA gene sequencing differences based on polymerase and number of cycles used, they are small with respect to the biological differences between fecal bacterial communities within individuals.
-Collectively, these results provide evidence that a real biological difference between groups, based on 16S rRNA gene sequencing of fecal samples, should be detectable regardless of polymerase and number of cycles used.
+**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences are mostly only detectable in mock communities.
+Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of fecal samples, can be consistently detected regardless of polymerase and number of cycles used.
 
@@
 We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
-We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
-Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles.
+We observed clear differences in the mock community based on polymerase used (P-value = `r format(filter(bray_mock_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), scientific = T, digits = 2)`) as well as differences based on polymerase and number of cycles (P-value = `r format(filter(bray_mock_combined, x1 == "interaction_polymerase_cycles") %>% select("Pr(>F)") %>% pull(), scientific = T, digits = 2)`).
+However, when investigating specific cycles within each polymerase we observed no difference in the bacterial community composition (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
+Finding specific differences in overall bacterial community composition based on number of cycles and polymerase, we next examined if specific taxa vary depending on polymerase and number of cycles.
 We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
@@
 
-When using the Bray-Curtis index, we observed that the affect polymerases and number of cycles had on the bacterial community composition in fecal samples was minimal.
-Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
+When using the Bray-Curtis index, we observed that polymerases and number of cycles had no affect on the bacterial community composition in fecal samples (P-value = `r round(filter(bray_fecal_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), digits = 2)`).
+There was also no interaction between individuals, number of cycles, and polymerase used (P-value = `r round(filter(bray_fecal_combined, x1 == "interaction_ind_poly_cycle") %>% select("Pr(>F)") %>% pull(), digits = 2)`).
 When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
 Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
-There was also an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
+Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
+However, there was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
 the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B].

### 2018-07-16 submission/manuscript.tex
update results

@@
 polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
-When analyzing fecal samples we observed similar differences in the
-number of OTUs. However, the median number of OTUs only varied by HiFi
-DNA polymerase used at 35 cycles and not at 20 cycles like the mock
+When analyzing fecal samples we observed smaller differences in the
+number of OTUs. Additionally, the median number of OTUs only varied by
+HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock
 community (P-value \textless{} 0.0001). Random Forest models were most
@@
 
-\textbf{Conclusions.} Although there are 16S rRNA gene sequencing
-differences based on polymerase and number of cycles used, they are
-small with respect to the biological differences between fecal bacterial
-communities within individuals. Collectively, these results provide
-evidence that a real biological difference between groups, based on 16S
-rRNA gene sequencing of fecal samples, should be detectable regardless
-of polymerase and number of cycles used.
+\textbf{Conclusions.} Although there were 16S rRNA gene sequencing
+differences based on polymerase and number of cycles used, bacterial
+community composition differences are mostly only detectable in mock
+communities. Collectively, these results provide evidence that smaller
+biological differences between groups, based on 16S rRNA gene sequencing
+of fecal samples, can be consistently detected regardless of polymerase
+and number of cycles used.
 
@@
 the bacterial community composition in more detail by using the
-Bray-Curtis index. We observed no difference in the bacterial community
-composition between any measure tested when using the mock community
-(P-value \textgreater{} 0.05) {[}Figure 3B{]}. Finding minimal
-differences in overall bacterial community composition, we next examined
-if specific taxa vary depending on polymerase and number of cycles. We
-found that OTUs that taxonomically classified to \emph{Salmonella},
-\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
-differed based on polymerase used {[}Table S6 \& S7{]}. These specific
-OTUs were all members of the mock community and the differences were
-most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
-differences between taxa were due to differences between a single
-polymerase (Kappa) and the other polymerases tested {[}Table S7{]}.
+Bray-Curtis index. We observed clear differences in the mock community
+based on polymerase used (P-value = 1e-04) as well as differences based
+on polymerase and number of cycles (P-value = 1e-04). However, when
+investigating specific cycles within each polymerase we observed no
+difference in the bacterial community composition (P-value
+\textgreater{} 0.05) {[}Figure 3B{]}. Finding specific differences in
+overall bacterial community composition based on number of cycles and
+polymerase, we next examined if specific taxa vary depending on
+polymerase and number of cycles. We found that OTUs that taxonomically
+classified to \emph{Salmonella}, \emph{Escherichia},
+\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
+polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
+members of the mock community and the differences were most pronounced
+at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
+between taxa were due to differences between a single polymerase (Kappa)
+and the other polymerases tested {[}Table S7{]}.
 
@@
 
-When using the Bray-Curtis index, we observed that the affect
-polymerases and number of cycles had on the bacterial community
-composition in fecal samples was minimal. Within each respective 5-cycle
-increment comparison there was no difference in Bray-Curtis index
-between the polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}.
-When using PERMANOVA to test for community differences based on any of
-the number of cycles within polymerases, only Phusion had cycle
-dependent differences (P-value = 0.03). Additionally, Phusion was one of
-two polymerases that had enough sequences to be rarefied to 1000
-sequences at 15 cycles. There was also an overall decrease in
-Bray-Curtis index by cycle comparison group (e.g.~the 15 cycle versus 20
-cycle group compared to the 30 cycle versus 35 cycle group) (P-value
-\textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15
-cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had
-a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison
-group (P-value \textless{} 0.05). In addition to assessing community
-composition by Bray-Curtis index, we also investigated differences in
-OTUs based on polymerase at each number of cycle used and no significant
-differences could be found (P-value \(\geqslant\) 0.01, corrected
-P-value = 1). Overall, these results suggest that in fecal samples the
-number of cycles and polymerase used only minimally change the bacterial
-community composition.
+When using the Bray-Curtis index, we observed that polymerases and
+number of cycles had no affect on the bacterial community composition in
+fecal samples (P-value = 1). There was also no interaction between
+individuals, number of cycles, and polymerase used (P-value = 1). When
+using PERMANOVA to test for community differences based on any of the
+number of cycles within polymerases, only Phusion had cycle dependent
+differences (P-value = 0.03). Additionally, Phusion was one of two
+polymerases that had enough sequences to be rarefied to 1000 sequences
+at 15 cycles. Within each respective 5-cycle increment comparison there
+was no difference in Bray-Curtis index between the polymerases (P-value
+\textgreater{} 0.05) {[}Figure 4B{]}. However, there was an overall
+decrease in Bray-Curtis index by cycle comparison group (e.g.~the 15
+cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle
+group) (P-value \textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's
+post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle
+comparison groups had a higher Bray-Curtis index than the 30 cycle
+versus 35 cycle comparison group (P-value \textless{} 0.05). In addition
+to assessing community composition by Bray-Curtis index, we also
+investigated differences in OTUs based on polymerase at each number of
+cycle used and no significant differences could be found (P-value
+\(\geqslant\) 0.01, corrected P-value = 1). Overall, these results
+suggest that in fecal samples the number of cycles and polymerase used
+only minimally change the bacterial community composition.
 

### 2018-07-17 submission/manuscript.Rmd
update discussion and methods

@@
 Additionally, these specific polymerases have different sequence error rates and chimera prevalence [Figure 2].
-Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a mock community [Figure 2E].
-This total number of OTUs also varied by polymerase used for these mock community samples [Figure 3A].
-The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for both mock community and fecal samples [Figure 3].
-However, these differences did not translate to large bacterial community composition changes based on polymerase [Figure 4].
-Even though there were bacterial community composition changes based on the number of cycles they occurred independently of the polymerase used [Figure 4].
-Additionally, there were no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.
-When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing, the model performance for individuals was significantly greater than all other models tested [Figure 5A].
-In addition to the better model performance of individuals, the most important OTUs to this model were not the most important OTUs for the other models tested [Figure 5B].
-Our observations show that differences between high fidelity DNA polymerases can be found.
-However, when using 16S rRNA gene sequencing these technical differences are dwarfed by biological differences between fecal bacterial communities within individuals.
+Chimera prevalence in particular was strongly associated with the total number of OTUs detected in our mock community and vaired by polymerase  [Figure 2E & 3A].
+Additionally, the bacterial community composition varied by number of cycles and polymerase in these mock community samples.
+Clearly, it is possible to identify a large number of differences based on number of cycles and polymerase used when analyzing mock community samples.
+However, these differences may not be a major concern when analyzing more complex community samples.
+
+Fecal samples showed a similar increase in median and range for the number of OTUs, based on number of cycles and polymerase used, as that found when analyzing mock community samples [Figure 4A].
+However, there were few differences between the polymerases within each of the amplification cycles [Figure 4A].
+Additionally, these differences did not translate to large bacterial community composition changes based on number of cycles or polymerase.
+The differences that were identifed were small and occured based on the number of cycles, independent of the polymerase used [Figure 4B].
+There was also no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.
+Interestingly, Random Forest models built on this OTU data had the highest classification success for individuals versus number of cycles or polymerase.
+In addition to the better model performance of Random Forest for individuals, the most important OTUs to this model were not the most important OTUs for the other models tested [Figure 5].
+Collectively, these observations show that differences based on the number of cycles and between high fidelity DNA polymerases can be found.
+However, when using 16S rRNA gene sequencing these technical differences can be smaller than the biological differences between fecal bacterial communities within individuals.
 
@@
 To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
-Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
+The analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
 Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
 The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples.
-Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data.
+Finally, Random Forest models were created to assess whether classification of polymerase, number of cycles, or individual could be performed best using 16S rRNA gene sequencing data.
+For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual labels were used.
+Both number of cycles and polymerase models included all samples but only the number of cycles or polymerase label was used for each respective model.
 Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).

### 2018-07-17 submission/manuscript.tex
update discussion and methods

@@
 Chimera prevalence in particular was strongly associated with the total
-number of OTUs detected in a mock community {[}Figure 2E{]}. This total
-number of OTUs also varied by polymerase used for these mock community
-samples {[}Figure 3A{]}. The range in the number of OTUs observed
-increased with the number of cycles and the change in range was
-dependent on the polymerase for both mock community and fecal samples
-{[}Figure 3{]}. However, these differences did not translate to large
-bacterial community composition changes based on polymerase {[}Figure
-4{]}. Even though there were bacterial community composition changes
-based on the number of cycles they occurred independently of the
-polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
-differences observed within the number of cycles between any of the
-polymerases for fecal samples. When Random Forest models were created to
-classify number of cycles, polymerase, and individual based on the 16S
-rRNA gene sequencing, the model performance for individuals was
-significantly greater than all other models tested {[}Figure 5A{]}. In
-addition to the better model performance of individuals, the most
-important OTUs to this model were not the most important OTUs for the
-other models tested {[}Figure 5B{]}. Our observations show that
-differences between high fidelity DNA polymerases can be found. However,
-when using 16S rRNA gene sequencing these technical differences are
-dwarfed by biological differences between fecal bacterial communities
-within individuals.
+number of OTUs detected in our mock community and vaired by polymerase
+{[}Figure 2E \& 3A{]}. Additionally, the bacterial community composition
+varied by number of cycles and polymerase in these mock community
+samples. Clearly, it is possible to identify a large number of
+differences based on number of cycles and polymerase used when analyzing
+mock community samples. However, these differences may not be a major
+concern when analyzing more complex community samples.
+
+Fecal samples showed a similar increase in median and range for the
+number of OTUs, based on number of cycles and polymerase used, as that
+found when analyzing mock community samples {[}Figure 4A{]}. However,
+there were few differences between the polymerases within each of the
+amplification cycles {[}Figure 4A{]}. Additionally, these differences
+did not translate to large bacterial community composition changes based
+on number of cycles or polymerase. The differences that were identifed
+were small and occured based on the number of cycles, independent of the
+polymerase used {[}Figure 4B{]}. There was also no OTU-based differences
+observed within the number of cycles between any of the polymerases for
+fecal samples. Interestingly, Random Forest models built on this OTU
+data had the highest classification success for individuals versus
+number of cycles or polymerase. In addition to the better model
+performance of Random Forest for individuals, the most important OTUs to
+this model were not the most important OTUs for the other models tested
+{[}Figure 5{]}. Collectively, these observations show that differences
+based on the number of cycles and between high fidelity DNA polymerases
+can be found. However, when using 16S rRNA gene sequencing these
+technical differences can be smaller than the biological differences
+between fecal bacterial communities within individuals.
 
@@
 \textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
-samples were rarefied to 1000 sequences for downstream analysis.
-Analysis of the mock community of each polymerase for GC-based
+samples were rarefied to 1000 sequences for downstream analysis. The
+analysis of the mock community of each polymerase for GC-based
 amplification differences, sequence error rate, number of sequences with
@@
 Forest models were created to assess whether classification of
-polymerase, cycles, or individual could be performed best using 16S rRNA
-gene sequencing data. Additionally, overlap between the most important
-OTUs to the three models was assessed using mean decrease in accuracy
-(MDA).
+polymerase, number of cycles, or individual could be performed best
+using 16S rRNA gene sequencing data. For the model used to classify
+individuals, all samples were included from all number of cycles and
+polymerases but only the individual labels were used. Both number of
+cycles and polymerase models included all samples but only the number of
+cycles or polymerase label was used for each respective model.
+Additionally, overlap between the most important OTUs to the three
+models was assessed using mean decrease in accuracy (MDA).
 

### 2018-07-19 submission/manuscript.Rmd
update manuscript from full read through

@@
 In this study, we focus on the amplification step's contribution to this overall error.
-To accomplish this we assessed 16S rRNA gene sequencing results in human fecal and mock community samples after using different high fidelity (HiFi) DNA polymerases and number of amplification cycles.
+To accomplish this we assessed 16S rRNA gene sequencing results in mock community and human fecal samples after using different  amplification cycles and high fidelity DNA polymerases.
 
-**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase.
-Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach.
+**Methods.** Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
+For fecal samples (n=4) we first extracted DNA using a PowerMag DNA extraction kit with a 10 minute bead beating step and then used the same approach as the mock community for the PCR amplification step.
 We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles.
 Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used.
-Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles.
-Based on these findings individual taxa differences based on polymerase and number of cycles was investigated.
+Additionally, differences in the bacterial community composition was assessed based on polymerase and number of cycles using the Bray-Curtis index.
+Based on these findings individual taxa differences based on polymerase and number of cycles was also investigated.
 Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor.
@@
 When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`).
-When analyzing fecal samples we observed smaller differences in the number of OTUs.
-Additionally, the median number of OTUs only varied by HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
+When analyzing fecal samples we observed smaller differences in the number of OTUs between polymerases.
+The median number of OTUs only varied by high fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
 Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
-Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor sample model.
+Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor model.
 
-**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences are mostly only detectable in mock communities.
-Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of fecal samples, can be consistently detected regardless of polymerase and number of cycles used.
+**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences were mostly only detectable in mock communities.
+Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of human fecal samples, can be consistently detected regardless of polymerase and number of cycles used.
 
@@
 However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012].
-Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow.
+Although some of the differences could be attributed to study design, error introduced during the 16S rRNA gene sequencing workflow could also be another source of variation.
 A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
@@
 When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`].
-Although there was a difference, neither model was successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)).
-Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are subtle.
+Although there was a difference, neither model was overly successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)).
+Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are not as large.
 
@@
 Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A].
-Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.
+Based on these observations, there are small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.
 
@@
 In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
-Overall, these results suggest that in fecal samples the number of cycles and polymerase used only minimally change the bacterial community composition.
+Overall, these results suggest that in fecal samples the number of cycles and polymerase used minimally change the bacterial community composition.
 
@@
 Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent.
-Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody.
+Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase, and a platinum *Taq* antibody.
 Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence.
@@
 Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence.
-When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
-One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.
+When choosing a distance metric, careful consideration of the errors introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
 
@@
 
-Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation in fecal communities between individuals.
+Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are smaller than the actual biological variation in fecal communities between individuals.
 Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
@@
 
-***Human and mock samples.***
+***Mock community and human samples.***
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+The theoretical abundance of 0.12 used as a comparison point for our relative abundance was based on this equal genomic DNA composition of the mock community.
 Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
@@
 All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
 
 ***PCR protocol.***
-The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+The five different high fidelity DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
 The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer.
 The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016].
+For Kappa and Q5, the annealing and extension time also were from a previously published optimized protocol [@taq_Gohl_2016].
 For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
 
-The number of cycles in the PCR for fecal and mock samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
-The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
+The number of cycles in the PCR for mock and fecal samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
 The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130).
 No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
+The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
 
@@
 subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
-For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g.
-higher GC content genomes, base substitutions, deletions, insertions, non-specific amplification, chimeras, etc.).
 
@@
 Finally, Random Forest models were created to assess whether classification of polymerase, number of cycles, or individual could be performed best using 16S rRNA gene sequencing data.
-For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual labels were used.
+For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual label was used.
 Both number of cycles and polymerase models included all samples but only the number of cycles or polymerase label was used for each respective model.
@@
 Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
-High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community.
+High and low GC content was determined based on the median GC percentage of either the V4 region or the average GC content of the whole genome of the bacterial species used in the mock community.
 Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
@@
 The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation].
-A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value.
+A total of 100 10-fold CV runs on different 80/20 splits of the data were run to generate a range of the Logloss value.
 The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm.
@@
 ***Reproducible methods.***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
+The code and analysis for this study can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
 

### 2018-07-19 submission/manuscript.tex
update manuscript from full read through

@@
 we focus on the amplification step's contribution to this overall error.
-To accomplish this we assessed 16S rRNA gene sequencing results in human
-fecal and mock community samples after using different high fidelity
-(HiFi) DNA polymerases and number of amplification cycles.
+To accomplish this we assessed 16S rRNA gene sequencing results in mock
+community and human fecal samples after using different amplification
+cycles and high fidelity DNA polymerases.
 
-\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
-PowerMag DNA extraction kit with a 10 minute bead beating step and
-amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
-Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
-communities (technical replicates n=4) consisting of previously isolated
-whole genomes of 8 different bacteria used the same approach. We first
+\textbf{Methods.} Amplification of mock communities (technical
+replicates n=4) consisting of previously isolated whole genomes of 8
+different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using
+Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
+For fecal samples (n=4) we first extracted DNA using a PowerMag DNA
+extraction kit with a 10 minute bead beating step and then used the same
+approach as the mock community for the PCR amplification step. We first
 assessed GC dependent differences, error rate, sequence error
@@
 was examined based on the polymerase and number of cycles used.
-Additionally, differences in the bacterial community composition by the
-Bray-Curtis index also was assessed based on polymerase and number of
-cycles. Based on these findings individual taxa differences based on
-polymerase and number of cycles was investigated. Finally, Random Forest
-models were created to test whether the bacterial community was better
-at classifying polymerases, number of cycles, or individual donor. We
-also assessed whether the most important taxa in the polymerase and
-number of cycle Random Forest models were also important in the model
-for individual donors.
+Additionally, differences in the bacterial community composition was
+assessed based on polymerase and number of cycles using the Bray-Curtis
+index. Based on these findings individual taxa differences based on
+polymerase and number of cycles was also investigated. Finally, Random
+Forest models were created to test whether the bacterial community was
+better at classifying polymerases, number of cycles, or individual
+donor. We also assessed whether the most important taxa in the
+polymerase and number of cycle Random Forest models were also important
+in the model for individual donors.
 
@@
 community samples the variation in the number of OTUs detected by the
-polymerases was observable as early as 20 cycles (P-value = 0.002).
-There also was a large range in the number of OTUs amplified by the
-polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
-When analyzing fecal samples we observed smaller differences in the
-number of OTUs. Additionally, the median number of OTUs only varied by
-HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock
-community (P-value \textless{} 0.0001). Random Forest models were most
-successful at classifying individual donor samples rather than
+polymerases was observable as early as 20 cycles (P-value = 0.002). When
+analyzing fecal samples we observed smaller differences in the number of
+OTUs between polymerases. The median number of OTUs only varied by high
+fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the
+mock community (P-value \textless{} 0.0001). Random Forest models were
+most successful at classifying individual donor samples rather than
 polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
 Additionally, the most important OTUs in the polymerase and number of
-cycle models were not the most important in the individual donor sample
-model.
+cycle models were not the most important in the individual donor model.
 
@@
 differences based on polymerase and number of cycles used, bacterial
-community composition differences are mostly only detectable in mock
+community composition differences were mostly only detectable in mock
 communities. Collectively, these results provide evidence that smaller
 biological differences between groups, based on 16S rRNA gene sequencing
-of fecal samples, can be consistently detected regardless of polymerase
-and number of cycles used.
+of human fecal samples, can be consistently detected regardless of
+polymerase and number of cycles used.
 
@@
 the taxa identified have varied widely depending on the study (Turnbaugh
-et al., 2008; Zupancic et al., 2012). Some of this variation could be
-due to error introduced during the 16S rRNA gene sequencing workflow. A
-typical 16S rRNA gene sequencing workflow can be divided into
+et al., 2008; Zupancic et al., 2012). Although some of the differences
+could be attributed to study design, error introduced during the 16S
+rRNA gene sequencing workflow could also be another source of variation.
+A typical 16S rRNA gene sequencing workflow can be divided into
 preservation, extraction, PCR, and sequencing steps. The preservation
@@
 {[}P-value = 5.06e-07{]}. Although there was a difference, neither model
-was successful at correctly classifying samples based on 16S rRNA gene
-sequencing data (cycles model, probability of correct classification =
-0.42 (0.4 - 0.43)(min - max) versus polymerase model, probability of
-correct classification = 0.39 (0.37 - 0.4)). Our observations in mock
-communities suggest that although differences are detected based on
-number of cycles or polymerase used, the overall community-wide
-differences are subtle.
+was overly successful at correctly classifying samples based on 16S rRNA
+gene sequencing data (cycles model, probability of correct
+classification = 0.42 (0.4 - 0.43)(min - max) versus polymerase model,
+probability of correct classification = 0.39 (0.37 - 0.4)). Our
+observations in mock communities suggest that although differences are
+detected based on number of cycles or polymerase used, the overall
+community-wide differences are not as large.
 
@@
 cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
-are only small differences in the total number of OTUs detected in fecal
+are small differences in the total number of OTUs detected in fecal
 samples that are based on number of cycles or polymerases used.
@@
 suggest that in fecal samples the number of cycles and polymerase used
-only minimally change the bacterial community composition.
+minimally change the bacterial community composition.
 
@@
 immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
-DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
+DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase, and a platinum
 \emph{Taq} antibody. Since it is not possible to know everything about
@@
 error rates and chimera prevalence. When choosing a distance metric,
-careful consideration of the biases introduced from the PCR step of the
-16S rRNA gene sequencing workflow need to be taken into account. One
-possible way to better choose polymerases may be based on the DNA
-polymerase family used in the PCR mixture.
+careful consideration of the errors introduced from the PCR step of the
+16S rRNA gene sequencing workflow need to be taken into account.
 
@@
 rRNA gene sequencing, our observations show that the differences between
-a variety of polymerases are dwarfed by the actual biological variation
-in fecal communities between individuals. Therefore, if the biological
-signal of interest is similar to differences in fecal bacterial
-communities found between individuals, then the type of high fidelity
-DNA polymerase used will only minimally change the results.
+a variety of polymerases are smaller than the actual biological
+variation in fecal communities between individuals. Therefore, if the
+biological signal of interest is similar to differences in fecal
+bacterial communities found between individuals, then the type of high
+fidelity DNA polymerase used will only minimally change the results.
 
@@
 
-\textbf{\emph{Human and mock samples.}} Fecal samples were obtained from
-4 individuals who were part of the Enterics Research Investigational
-Network (ERIN). The processing and storage of these samples were
-previously published (Seekatz et al., 2016). Other than confirmation
-that none of these individuals had a \emph{Clostridium difficle}
-infection, clinical data and other types of meta data were not utilized
-or accessed for this study. All samples were extracted using the
-MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
-(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
-Community DNA Standard (Zymo, CA, USA) was used for mock communities and
-was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
+\textbf{\emph{Mock community and human samples.}} The
+ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
+CA, USA) was used for mock communities and the bacterial component was
+made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
 \emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
@@
 (\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
+The theoretical abundance of 0.12 used as a comparison point for our
+relative abundance was based on this equal genomic DNA composition of
+the mock community. Fecal samples were obtained from 4 individuals who
+were part of the Enterics Research Investigational Network (ERIN). The
+processing and storage of these samples were previously published
+(Seekatz et al., 2016). Other than confirmation that none of these
+individuals had a \emph{Clostridium difficle} infection, clinical data
+and other types of meta data were not utilized or accessed for this
+study. All samples were extracted using the MOBIO\textsuperscript{TM}
+PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
 
 \textbf{\emph{PCR protocol.}} The five different high fidelity DNA
-polymerases (hereto referred to as polymerases) that were tested
-included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
-HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
-(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
-polymerases activation time was 2 minutes, unless a different activation
-was specified by the manufacturer. The annealing and extension time for
-Platinum and Accuprime followed a previously published protocol (Kozich
-et al., 2013)
+polymerases that were tested included AccuPrime\textsuperscript{TM}
+(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
+(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
+England Biolabs, MA, USA). The polymerases activation time was 2
+minutes, unless a different activation was specified by the
+manufacturer. The annealing and extension time for Platinum and
+Accuprime followed a previously published protocol (Kozich et al., 2013)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
 For Kappa and Q5, the annealing and extension time also were from a
-previously published protocol (Gohl et al., 2016). For Phusion, the
-company defined activation and annealing times were used while the
-extension time followed the Accuprime and Platinum settings.
+previously published optimized protocol (Gohl et al., 2016). For
+Phusion, the company defined activation and annealing times were used
+while the extension time followed the Accuprime and Platinum settings.
 
-The number of cycles in the PCR for fecal and mock samples started at 15
+The number of cycles in the PCR for mock and fecal samples started at 15
 and increased by 5 up to 35 cycles, with amplicons used at each 5-step
-increase for sequencing. The PCR of fecal DNA samples consisted of all 4
-samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total
-sample n=100). The mock communities had 4 replicates at 15, 20, 25, and
-35 cycles and 10 replicates for 30 cycles for all polymerases (total
-samples n=130). No mock community sample had enough PCR product at 15
-cycles for adequate 16S rRNA gene sequencing.
+increase for sequencing. The mock communities had 4 replicates at 15,
+20, 25, and 35 cycles and 10 replicates for 30 cycles for all
+polymerases (total samples n=130). No mock community sample had enough
+PCR product at 15 cycles for adequate 16S rRNA gene sequencing. The PCR
+of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and
+35 cycles for each polymerase (total sample n=100).
 
@@
 \emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
-(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016). For
-simplicity, the use of the term `error' within the context of this study
-moving forward encompasses changes in the downstream sequencing results
-due to the amplification step (e.g.~higher GC content genomes, base
-substitutions, deletions, insertions, non-specific amplification,
-chimeras, etc.).
+(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
 
@@
 individuals, all samples were included from all number of cycles and
-polymerases but only the individual labels were used. Both number of
+polymerases but only the individual label was used. Both number of
 cycles and polymerase models included all samples but only the number of
@@
 2017). High and low GC content was determined based on the median GC
-percentage of either the V4 region or the whole genome of the bacterial
-species used in the mock community. Differences in the total number of
-OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
-comparison of the number of OTUS in fecal samples the data was
-normalized to each individual by cycle number to account for the
-biological variation. Bray-Curtis distance matrices were generated using
-mothur. The distance matrix data was analyzed using PERMANOVA with the
-vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
-within R. The Random Forest models were run using the caret package (v
-6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
-different 80/20 splits of the data was run to generate a range of the
-Logloss value. The probability of a correct call was obtained from this
-Logloss value by taking the negative natural logarithm. For both error
-and chimera analysis, samples were tested using Kruskal-Wallis with a
-Dunns post-hoc test. Where applicable, correction for multiple
-comparison utilized the Benjamini-Hochberg method (Benjamini \&
-Hochberg, 1995).
+percentage of either the V4 region or the average GC content of the
+whole genome of the bacterial species used in the mock community.
+Differences in the total number of OTUs were analyzed using an ANOVA
+with a tukey post-hoc test. For the comparison of the number of OTUS in
+fecal samples the data was normalized to each individual by cycle number
+to account for the biological variation. Bray-Curtis distance matrices
+were generated using mothur. The distance matrix data was analyzed using
+PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
+Kruskal-Wallis tests within R. The Random Forest models were run using
+the caret package (v 6.0.78) (Jed Wing et al., 2017). A total of 100
+10-fold CV runs on different 80/20 splits of the data were run to
+generate a range of the Logloss value. The probability of a correct call
+was obtained from this Logloss value by taking the negative natural
+logarithm. For both error and chimera analysis, samples were tested
+using Kruskal-Wallis with a Dunns post-hoc test. Where applicable,
+correction for multiple comparison utilized the Benjamini-Hochberg
+method (Benjamini \& Hochberg, 1995).
 
-\textbf{\emph{Reproducible methods.}} The code and analysis can be found
-here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
-The raw sequences can be found on the SRA at the following accession
-number SRP132931.
+\textbf{\emph{Reproducible methods.}} The code and analysis for this
+study can be found here
+\url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}. The raw
+sequences can be found on the SRA at the following accession number
+SRP132931.
 

### 2018-11-08 submission/manuscript.Rmd
Early edits

@@
 
+# Influence of DNA Polymerase on PCR bias, chimera formation, and errors when sequencing 16S rRNA genes from complex communities
+
 \begin{center}
@@
 
+16S rRNA gene sequencing is a widely used tool for surveying the structure of microbial communities
+* Numerous sources of bias ranging from sample preservation, extraction, primers, PCR, and sequencing
+* PCR and sequencing errors can also influence our ability to accurately describe the structure of the communities
+* Within a study, it is unlikely that samples would be differentially affected by these sources of bias; however, if we want to compare the results of one study to another, the results are likely to be confounded by the differences in these biases and sources of error
+* There has been great interest in developing bioinformatics tools to account for errors and limited interest in accounting for bias
+* There has also been considerable interest in considering how laboratory methods influence biases and errors.
+* One area that has not received extensive attention has been the influence of the polymerase on bias and error.
+
+Factors known to influence PCR bias
+* Fragment length
+* GC content
+* Flanking DNA sequence
+* Polymerase?
+* Complex communities?
+* Number of cycles
+
+Factors known to influence chimera rate
+* Fragmentation of DNA
+* DNA concentration
+* Similarity of potential parents
+* PCR extension time
+* Number of cycles
+
+Factors known to influence polymerase error rates
+* Number of cycles
+* Proof-reading
+* Enzymology
+
+Early work focused on simple mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation.
+* Have a poor understanding of how polymerase impacts these problems
+* Have a poor understanding of how these problems manifest in more complex communities
+
 The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017].
@@
 ## Results
+
+Quantifying the problems...
+
+* Polymerases' effects on bias
+	* Different polymerases and cycle numbers
+	* Mock community w/ perfect chimera removal
+	* Relative abundance
+	* Variation in Shannon/Inverse Simpson diversity across cycle numbers
+	* Variation in Bray-Curtis across cycle numbers
+	* Impact of fragment length on bias (are they all the same length)
+	* Impact of fragment GC content on bias
+
+* Polymerases' effects on error rates
+	* Different polymerases and cycle numbers
+	* Mock community w/ perfect chimera removal
+
+* Polymerases' effects on chimera rate
+	* Different polymerases and cycle numbers
+	* Mock community w/ perfect chimera removal
+
+Characterizing the effects...
+
+* Polymerases' effects on bias
+	* Different polymerases and cycle numbers
+	* Human communities w/ uchime chimera removal
+	* Variation in Shannon/Inverse Simpson diversity across cycle numbers
+	* Variation in Bray-Curtis across cycle numbers
+	* Which factor has biggest influence on beta-diversity - polymerase, cycle number, sample
+	* Impact of fragment length on bias (are they all the same length)
+	* Impact of fragment GC content on bias
+
+* Detection of chimeras
+	* Different polymerases and cycle numbers
+	* Use Mock community data
+	* Use uchime to remove chimeras to get residual chimera rate
+	* Error rate vs residual chimera rate
 

### 2018-11-29 submission/manuscript.Rmd
Rough in Results

@@
 ---
-
-```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-source('../code/functions.R')
-
-loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr", "FSA"))
-
-```
-
-```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-#Read in the meta data file
-metadata <- read_csv("../data/process/tables/meta_data.csv")
-
-# Read in fecal count data
-fecal_numOTU_data <- read_csv("../data/process/tables/fecal_sub_sample_1000_count_table.csv") %>%
-  mutate(taq = factor(taq,
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"),
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5")),
-         cycles = factor(cycles,
-                            levels = c("15x", "20x", "25x", "30x", "35x"),
-                            labels = c("15x", "20x", "25x", "30x", "35x"))) %>%
-  group_by(taq, cycles) %>%
-  summarise(taq_median = median(numOTUs, na.rm = T),
-            taq_min = min(numOTUs, na.rm = T),
-            taq_max = max(numOTUs, na.rm = T)) %>%
-  mutate(range_diff = taq_max - taq_min)
-
-# Read in fecal sample number of OTU results
-fs_anova_summary <- as.data.frame(read_csv("../data/process/tables/fecal_overall_anova_results.csv"))
-fs_tukey_summary <- as.data.frame(read_csv("../data/process/tables/fecal_overall_tukey_results.csv")) %>%
-  rename(padj = `p adj`)
-
-# Read in the Mock count data
-mock_num_otu_data <- read_csv("../data/process/tables/mock_sub_sample_1000_count_table.csv") %>%
-  mutate(taq = factor(taq,
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"),
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>%
-  group_by(taq, cycles) %>%
-  summarise(taq_median = median(numOTUs, na.rm = T),
-            taq_min = min(numOTUs, na.rm = T),
-            taq_max = max(numOTUs, na.rm = T)) %>%
-  mutate(range_diff = taq_max - taq_min)
-
-ref_mock_otu_num <- read_tsv("../data/process/zymo_mock_only.shared") %>%
-  select(numOtus)
-
-# Read in mock sample number of OTU results
-mock_anova_summary <- as.data.frame(read_csv("../data/process/tables/mock_overall_anova_results.csv"))
-mock_tukey_summary <- read_csv("../data/process/tables/mock_overall_tukey_results.csv")
-
-# Read in error rate results
-krsukal_error_rate <- read_csv("../data/process/tables/mock_error_overall_kruskal_results.csv")
-dunn_error_rate <- read_csv("../data/process/tables/mock_error_overall_dunn_results.csv")
-
-# Read in sequence error count results
-kruskal_seq_error_count <- read_csv("../data/process/tables/mock_error_count_overall_kruskal_results.csv")
-dunn_seq_error_count <- read_csv("../data/process/tables/mock_error_count_overall_dunn_results.csv")
-
-# Read in chimera prevalence results
-kruskal_chimera_prev <- as.data.frame(read_csv("../data/process/tables/mock_chimera_overall_kruskal_results.csv"))
-dunn_chimera_prev <- read_csv("../data/process/tables/mock_chimera_overall_dunn_results.csv")
-
-# Read in results from PERMANOVA and bray analysis
-bray_sim_to_prev_cycle_kruskal_results <- as.data.frame(read_csv("../data/process/tables/bray_sim_to_prev_cycle_kruskal_results.csv"))
-bray_permanova_by_taq_results <- as.data.frame(read_csv("../data/process/tables/bray_permanova_by_taq_results.csv"))
-
-bray_mock_combined <- read_csv("../data/process/tables/mock_bray_combined_permanova_results.csv")
-bray_fecal_combined <- read_csv("../data/process/tables/fecal_bray_combined_permanova_results.csv")
-
-# Read in actual error data
-error_data <- read_csv("../data/process/tables/mock_chimera_error_summary.csv") %>%
-                       filter(total_seqs >= 1000)
-
-error_summary <- error_data %>%
-  group_by(taq, cycles) %>%
-  summarise_at(vars(mean_error, chimera_prevalence, seq_error_prevalence), funs(median, min, max)) %>%
-  select(taq, cycles, contains("chimera"), contains("mean_error"), contains("seq_error"))
-
-# Vector of sub samples used
-dist_data <- read_csv("../data/process/tables/bray_5_cycle_dist_1000_data.csv") %>%
-  mutate(taq = factor(taq,
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"),
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>%
-  group_by(taq, cycle_compare) %>%
-  summarise(dist_median = median(distance, na.rm = T),
-            dist_min = min(distance, na.rm = T),
-            dist_max = max(distance, na.rm = T)) %>%
-  mutate(range_diff = dist_max - dist_min) %>%
-  ungroup() %>%
-  filter(cycle_compare != "25to35")
-
-fecal_kruskal_test_by_cycle_compare_summary <- dist_data %>%
-  nest() %>%
-  mutate(kruskal_analysis = map(data, ~kruskal.test(dist_median ~ factor(cycle_compare), data = .x)),
-         summary_data = map(kruskal_analysis, broom::tidy)) %>%
-  select(summary_data) %>% unnest(summary_data)
-
-dunns_test_by_cycle_compare_summary <- dist_data %>%
-  nest() %>%
-  mutate(dunn_analysis = map(data, ~dunnTest(dist_median ~ factor(cycle_compare), data = .x, method = "bh")[["res"]])) %>%
-  select(dunn_analysis) %>% unnest() %>%
-  arrange(Comparison)
-
-mock_dist_data <- read_csv("../data/process/tables/mock_bray_5_cycle_dist_1000_data.csv") %>%
-  mutate(taq = factor(taq,
-                      levels = c("ACC", "K", "PHU", "PL", "Q5"),
-                      labels = c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"))) %>%
-  filter(!is.na(distance)) %>%
-  group_by(taq, cycle_compare) %>%
-  summarise(dist_median = median(distance, na.rm = T),
-            dist_min = min(distance, na.rm = T),
-            dist_max = max(distance, na.rm = T)) %>%
-  mutate(range_diff = dist_max - dist_min) %>%
-  ungroup() %>%
-  filter(cycle_compare != "25to35")
-
-mock_kruskal_test_by_cycle_compare_summary <- mock_dist_data %>%
-  nest() %>%
-  mutate(kruskal_analysis = map(data, ~kruskal.test(dist_median ~ factor(cycle_compare), data = .x)),
-         summary_data = map(kruskal_analysis, broom::tidy)) %>%
-  select(summary_data) %>% unnest(summary_data)
-
-# Read in the regression data
-r2_table <- read_csv("../data/process/tables/r2_otu_vs_chimera_summary.csv")
-
-# Read in GC data
-gc_data <- read_csv("../data/process/tables/v4_gc_wilcox_test_summary.csv") %>%
-  filter(cycles != "fifteen") %>%
-  mutate(bh = p.adjust(p.value, method = "BH"))
-
-gc_genome_data <- read_csv("../data/process/tables/whole_genome_gc_wilcox_test_summary.csv") %>%
-  filter(cycles != "fifteen") %>%
-  mutate(bh = p.adjust(p.value, method = "BH"))
-
-gc_summary <- read_csv("../data/process/tables/gc_summary_table.csv")
-
-# Read in the taxa data
-fecal_otu_tests <- read_csv("../data/process/tables/kruskal_otu_polymerase_fecal_sample_diffs_summary.csv") %>%
-  arrange(p.value)
-
-mock_otu_tests_kruskal <- read_csv("../data/process/tables/kruskal_otu_polymerase_mock_diffs_summary.csv") %>%
-  arrange(bh)
-
-mock_otu_tests_dunns <- read_csv("../data/process/tables/dunns_sig_otu_polymerase_mock_diffs_summary.csv")
-
-# Read in RF model data
-mock_poly_rf_summary <- read_csv("../data/process/tables/polymerase_group_mock_rf_model_summary.csv")
-mock_cycles_rf_summary <- read_csv("../data/process/tables/cycles_group_mock_rf_model_summary.csv")
-
-fecal_sample_rf_summary <- read_csv("../data/process/tables/sample_group_rf_model_summary.csv")
-fecal_poly_rf_summary <- read_csv("../data/process/tables/polymerase_group_rf_model_summary.csv")
-fecal_cycles_rf_summary <- read_csv("../data/process/tables/cycles_group_rf_model_summary.csv")
-
-dunn_rf_data <- read_csv("../data/process/tables/dunn_rf_model_test_summary.csv")
-
-```
 
@@
 
+## pat's final paragraph
+We addressed X potential areas where the type of PCR polymerase was hypothesized to impact the results of a microbiome analysis.
+These included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process.
+To characterize these factors we sequenced a mock community with known sequences and community structure and human fecal samples with unknown sequences and community structures.
+
+Three factors can impact whether an organism's 16S rRNA gene sequence is correctly represented.
+These include mis-incorporated bases during the PCR amplification phase, mis-incorporated bases during sequencing, and creation of chimeras during the PCR amplification phase.
+
+We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform.
+This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol.
+The mock community is commercially available and was made up of genomic DNA from 8 bacteria.
+The benefit of using this mock community is that the sequences of the 16S rRNA genes from these bacteria is known.
+This allowed us to confidently identify any chimeras and to calculate the sequencing error rate.
+
+We selected five high fidelity DNA polymerases to amplify the V4 region of the 16S rRNA gene from a mock community containing genomic DNA from 8 previously sequenced bacteria.
+The genomic DNA from the mock community was amplified using 15, 20, 25, 30, and 35 rounds of amplification using one of five different polymerases.
+Each condition was replicated four times and reads for each condition were pooled.
+Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur.
+Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
+The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur.
+This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
+The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.
+
 \newpage
@@
 
-Quantifying the problems...
+***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs).
+More importantly, sequencing errors themselves can alter the representation of the community.
+Therefore, it is important to minimize the number of sequencing errors in an analysis.
+Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community.
+We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [ref].
+Regardless of the polymerase, the error rate increased with the number of rounds of amplification.
+Using 30 rounds of PCR is a common approach across diverse types of samples.
+Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e.
+0.XXX%) followed by the Platinum (i.e.
+0.XXX%), Phusion (i.e.
+0.XXX%), Kappa (i.e.
+0.XXX%), and Q5 (i.e.
+0.XXX%) polymerases (Figure 1).
+The order of the error rates using these polymerases is consistent with previously documented reports of their per cycle error rates.
+When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e.
+0.XXX%) followed by the Accuprime (i.e.
+0.XXX%), Phusion (i.e.
+0.XXX%), Kappa (i.e.
+0.XXX%), and Q5 (i.e.
+0.XXX%) polymerases (Figure 1).
+Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary is a good guide to minimizing the impact of polymerase on the error rate.
+Furthermore, using a polymerase with the lowest possible error rate is preferred.
 
-* Polymerases' effects on bias
-	* Different polymerases and cycle numbers
-	* Mock community w/ perfect chimera removal
-	* Relative abundance
-	* Variation in Shannon/Inverse Simpson diversity across cycle numbers
-	* Variation in Bray-Curtis across cycle numbers
-	* Impact of fragment length on bias (are they all the same length)
-	* Impact of fragment GC content on bias
+***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses.
+Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors.
+We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence.
+As expected from previous studies, the number of chimeras increased with rounds of amplification.
+Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used.
+After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e.
+XX%) followed by the Q5 (i.e.
+XX%), Phusion (i.e.
+XX%), Kappa (i.e.
+XX%), and Accuprime (i.e.
+XX%) polymerases.
+We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity.
+For all polymerases, the specificity was above XX% and showed a slight association with the number of cycles used.
+There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community.
+This suggests that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
+In general, the sensitivity of UChime varied between 50 and 90%.
+These results could be affected by the use of a single mock community with limited genetic diversity.
+Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric.
+These results followed those from the mock communities - additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
+Although it is difficult to identify a feature of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
 
-* Polymerases' effects on error rates
-	* Different polymerases and cycle numbers
-	* Mock community w/ perfect chimera removal
+***The bias of PCR amplification is inconsistent and subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses.
+The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria.
+The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community.
+We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase.
+Interestingly, for some bacteria, there representation became less biased with additional rounds of PCR (e.g.
+*L.
+fermentum*), while others became more biased (e.g.
+*E.
+faecalis*), and others had little change (e.g.
+*B.
+subtilis*).
+The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
+Across the bacteria there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed.
+One of the bacteria represented in the mock community, *S.
+enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide.
+We used the sequence data to calculate the ratio of the dominant to rare variants from *S.
+enterica* (Figure.
+XXX).
+With the exception of the Kappa polymerase, the other four polymerases converged to a ratio of 5.XXX.
+Based on the relative abundance data from the mock community it was not possible to discern the factors that result in amplification bias.
+As an additional test, we quantified the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification.
+For the mock community data, the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units.
+The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX).
+In contrast, when we characterized the variation between amplification rounds using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases.
+In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases.
+The data generated using the Platinum polymerase was consistent across rounds, but it was more biased than these polymerases.
+Considering the average distance across the four samples varied between 0.XX and 0.XX, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
+It is difficult to develop a specific recommendation based on the level of bias across rounds of PCR or polymerases; however, the general suggestion is to use as few rounds of amplification as possible.
 
-* Polymerases' effects on chimera rate
-	* Different polymerases and cycle numbers
-	* Mock community w/ perfect chimera removal
+***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift.
+To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification.
+Using the mock community data there were no obvious trends.
+The average Bray-Curtis distance within a set of conditions varied by 0.XX to 0.XX units.
+Although we did not obtain replicates of each of the stool samples, the intra-sample variation for each set of conditions was consistent and varied between 0.XX and 0.XX units.
+These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.
 
-Characterizing the effects...
+***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels appeared to be small relative to the overall differences between samples.
+To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
+We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor.
+Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
+We also performed a principle coordinates analysis of the Bray-Curtis distances showed distinct clusters by donor.
+Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
+Both of these results indicate that for these samples, the differences were larger than the difference between conditions.
+These results suggest that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important.
 
-* Polymerases' effects on bias
-	* Different polymerases and cycle numbers
-	* Human communities w/ uchime chimera removal
-	* Variation in Shannon/Inverse Simpson diversity across cycle numbers
-	* Variation in Bray-Curtis across cycle numbers
-	* Which factor has biggest influence on beta-diversity - polymerase, cycle number, sample
-	* Impact of fragment length on bias (are they all the same length)
-	* Impact of fragment GC content on bias
+Generate PCoA diagrams
+Alpha diversity figures
 
-* Detection of chimeras
-	* Different polymerases and cycle numbers
-	* Use Mock community data
-	* Use uchime to remove chimeras to get residual chimera rate
-	* Error rate vs residual chimera rate
-
-***Error differences due to number of cycles or polymerase used are detectable in mock communities.***
-There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1 & Table S1].
-However, using only the GC content of the V4 region resulted in less differences than when using the whole genome [Table S1].
-The differences between high and low GC content groups were significant as early as 20 cycles in all polymerases except Q5 [Table S1].
-The highest (*Staphylococcus*) and lowest (*Pseudomonas*) GC content bacteria were the most divergent  from the expected relative abundance of 0.12 [Figure 1].
-These observations suggest that the high fidelity DNA polymerases tested do not prevent GC-based amplification error.
-
-The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 2A & Table S2].
-Thus, the majority of the differences observed across the number of cycles was between Kappa and other polymerases [Figure 2A & Table S3].
-The total sequences that contained at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S2 & S3].
-These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1].
-Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific type of base substitution.
-
-We next examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs.
-We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2].
-Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3].
-Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C].
-The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C].
-There was also a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D].
-Additionally, a positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E].
-This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E].
-This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms previous reports that the number of OTUs is dependent on the prevalence of these chimeric sequences.
-
-***Bacterial community composition differences based on number of cycles or polymerase used are less obvious in mock community samples.***
-Due to the close relationship between chimera prevalence and the number of OTUs, we also observed a polymerase dependent difference in the range of the number of OTUs in the mock community samples based on number of cycles and polymerase [Figure 3A].
-The closest any of the polymerases came to a total of `r pull(ref_mock_otu_num, numOtus)` OTUs, created by running the mock reference 16S sequences through mothur processing, was at 25 and 30 cycles [Figure 3A].
-Differences between polymerases for the number of OTUs created were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S4].
-Using a Tukey post-hoc test, the majority of differences for the number of OTUs detected in the mock community was largely due to Kappa and Platinum versus the other polymerases across the number of cycles used [Table S5].
-
-We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index.
-We observed clear differences in the mock community based on polymerase used (P-value = `r format(filter(bray_mock_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), scientific = T, digits = 2)`) as well as differences based on polymerase and number of cycles (P-value = `r format(filter(bray_mock_combined, x1 == "interaction_polymerase_cycles") %>% select("Pr(>F)") %>% pull(), scientific = T, digits = 2)`).
-However, when investigating specific cycles within each polymerase we observed no difference in the bacterial community composition (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B].
-Finding specific differences in overall bacterial community composition based on number of cycles and polymerase, we next examined if specific taxa vary depending on polymerase and number of cycles.
-We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7].
-These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7].
-The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases tested [Table S7].
-
-Based on these observations, we created Random Forest models to classify samples based on number of cycles or polymerase used.
-When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`].
-Although there was a difference, neither model was overly successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)).
-Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are not as large.
-
-***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
-Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5].
-However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g.
-Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A].
-There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g.
-Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A].
-Based on these observations, there are small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.
-
-When using the Bray-Curtis index, we observed that polymerases and number of cycles had no affect on the bacterial community composition in fecal samples (P-value = `r round(filter(bray_fecal_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), digits = 2)`).
-There was also no interaction between individuals, number of cycles, and polymerase used (P-value = `r round(filter(bray_fecal_combined, x1 == "interaction_ind_poly_cycle") %>% select("Pr(>F)") %>% pull(), digits = 2)`).
-When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`).
-Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles.
-Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B].
-However, there was an overall decrease in Bray-Curtis index by cycle comparison group (e.g.
-the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B].
-Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05).
-In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`).
-Overall, these results suggest that in fecal samples the number of cycles and polymerase used minimally change the bacterial community composition.
-
-Finally, we built Random Forest models to classify samples for the number of cycles, polymerase used, or individual it was obtained from.
-Using 16S rRNA gene sequencing data there was a large disparity in the model's ability to correctly classify samples based on polymerase used, number of cycles, and individual (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal") & str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
-There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)).
-Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5].
-This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5].
-Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.
+https://www.hindawi.com/journals/mbi/2014/287430/
+https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169774
+https://academic.oup.com/nar/article/24/18/3546/2359477
 
@@
 Additionally, these specific polymerases have different sequence error rates and chimera prevalence [Figure 2].
-Chimera prevalence in particular was strongly associated with the total number of OTUs detected in our mock community and vaired by polymerase  [Figure 2E & 3A].
+Chimera prevalence in particular was strongly associated with the total number of OTUs detected in our mock community and varied by polymerase  [Figure 2E & 3A].
 Additionally, the bacterial community composition varied by number of cycles and polymerase in these mock community samples.
@@
 Additionally, these differences did not translate to large bacterial community composition changes based on number of cycles or polymerase.
-The differences that were identifed were small and occured based on the number of cycles, independent of the polymerase used [Figure 4B].
+The differences that were identified were small and occurred based on the number of cycles, independent of the polymerase used [Figure 4B].
 There was also no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.

### 2018-11-29 code/bias_analysis.R
Re-run bias analysis

@@ -92,7 +92,7 @@ inner_join(rel_abundances, mock, by="species") %>%
 		theme(strip.text=element_text(face="italic"),
 					strip.text.x=element_text(hjust=0, size=9),
 					strip.background = element_rect(color = 'white', fill = 'white', size = 10)) +
-		ggsave("bias.pdf")
+		ggsave("results/figures/mock_bias.pdf")
 
 
 
@@ -119,4 +119,4 @@ ggplot(salmonella_ratio, aes(x=rounds, y=ratio, group=polymerase, color=polymera
 	geom_hline(aes(yintercept=6), col="gray") +
 	geom_line() +
 	theme_classic() +
-	ggsave("salmonella_ratio.pdf")
+	ggsave("results/figures/mock_salmonella_ratio.pdf")

### 2018-11-29 code/plot_mock_pcoa.R
Generate PCoA figures

@@ -1,0 +1,15 @@
+library(tidyverse)
+
+args <- commandArgs(trailingOnly=TRUE)
+file_name <- args[1]
+
+# file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.vsearch.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes"
+
+read_tsv(file_name) %>%
+	select(group, axis1, axis2) %>%
+	separate(group, into=c("rounds", "polymerase", "mock")) %>%
+	mutate(rounds = str_replace(rounds, "x", "")) %>%
+	ggplot(aes(x=axis1, y=axis2, color=rounds, shape=polymerase)) +
+		geom_point() +
+		stat_ellipse(aes(group=rounds)) +
+		ggsave("results/figures/mock_pcoa.pdf")

### 2018-11-29 code/plot_stool_pcoa.R

@@ -1,0 +1,16 @@
+library(tidyverse)
+
+args <- commandArgs(trailingOnly=TRUE)
+file_name <- args[1]
+
+# file_name <- "data/mothur/stool.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes"
+
+read_tsv(file_name) %>%
+	select(group, axis1, axis2) %>%
+	mutate(group = str_replace(group, "_PMM_FS", "")) %>%
+	separate(group, into=c("rounds", "polymerase", "subject")) %>%
+	mutate(rounds = str_replace(rounds, "x", "")) %>%
+	ggplot(aes(x=axis1, y=axis2, color=polymerase, shape=rounds)) +
+		geom_point() +
+		stat_ellipse(aes(group=subject), type="norm", color="gray", fill="gray") +
+		ggsave("results/figures/stool_pcoa.pdf")

### 2018-12-05 submission/manuscript.Rmd
Remove 15 cycle data

@@
 
+https://www.hindawi.com/journals/mbi/2014/287430/
+https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169774
+https://academic.oup.com/nar/article/24/18/3546/2359477
+
 ## Introduction
 
-16S rRNA gene sequencing is a widely used tool for surveying the structure of microbial communities
-* Numerous sources of bias ranging from sample preservation, extraction, primers, PCR, and sequencing
-* PCR and sequencing errors can also influence our ability to accurately describe the structure of the communities
-* Within a study, it is unlikely that samples would be differentially affected by these sources of bias; however, if we want to compare the results of one study to another, the results are likely to be confounded by the differences in these biases and sources of error
-* There has been great interest in developing bioinformatics tools to account for errors and limited interest in accounting for bias
-* There has also been considerable interest in considering how laboratory methods influence biases and errors.
-* One area that has not received extensive attention has been the influence of the polymerase on bias and error.
+16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities.
+This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample.
+Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples, storage conditions, extraction methods, amplification conditions, sequencing method, and analytical pipeline.
+The increased sampling depth that is now available relative to previous Sanger-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.
 
-Factors known to influence PCR bias
-* Fragment length
-* GC content
-* Flanking DNA sequence
-* Polymerase?
-* Complex communities?
-* Number of cycles
+One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification.
+Factors such as the choice of primers have an obvious impact on which populations will be amplified.
+However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras, misincorporation of nucleotides, preferential amplification of populations leading to bias, and accumulation of random amplification events leading to PCR drift.
+Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs.
+Others have attempted to account for PCR bias using modeling approaches; however, these have been developed for idealized situations.
+Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability.
+To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift.
+Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community.
+Early investigations of the factors that lead to the formation of PCR artifacts focused on analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation.
+Although these studies were instrumental in forcing researchers to be cautious about the interpretation of their results, we have a poor understanding of how these factors affect the formation of PCR artifacts in more complex communities.
 
-Factors known to influence chimera rate
-* Fragmentation of DNA
-* DNA concentration
-* Similarity of potential parents
-* PCR extension time
-* Number of cycles
+The influence that the choice of DNA polymerase has on the formation of PCR artifacts has not been well studied.
+There has been a lot of recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate; however, these studies use the same DNA polymerase in the PCR step and assume that the rate of misincorporation of nucleotides due to PCR are significantly smaller than those from the sequencing phase.
+There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras.
+A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
+The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
+Yet this optimization was specific for the community they were analyzing (i.e.
+captive and semi-captive red-shanked doucs) and assumes that if the two polymerases generate the same community structure that the community structure is correct.
+In fact, the community structure generated by both methods is not free of artifacts, but has the same artifacts.
+A challenge in these types of experiments is a priori knowledge of the true community representation.
+A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias; however, mock communities have a limited phylogenetic diversity relative to natural communities.
+Natural communities, in contrast, have an unknown community composition making absolute measurements impossible.
+They can be used to validate results from mock communities and to understand the relative impacts of artifacts on the ability to differentiate biological and methodological sources of variation.
+Given the large number of DNA polymerases available to researchers it is unlikely that a specific recommendation is possible.
+Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are desired.
 
-Factors known to influence polymerase error rates
-* Number of cycles
-* Proof-reading
-* Enzymology
-
-Early work focused on simple mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation.
-* Have a poor understanding of how polymerase impacts these problems
-* Have a poor understanding of how these problems manifest in more complex communities
-
-The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017].
-However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012].
-Although some of the differences could be attributed to study design, error introduced during the 16S rRNA gene sequencing workflow could also be another source of variation.
-A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps.
-The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017].
-For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017].
-Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not as well characterized.
-
-There is a large body of literature that shows there are errors due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000].
-Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g.
-V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)].
-Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998].
-Many of these sources of error are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015].
-For example, both amplification error and non-specific amplification (e.g.
-incorrect amplicon size products) also can increase as the number of cycles increases.
-This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005].
-Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011].
-These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011].
-In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014].
-For simplicity, the use of the term 'error' within the context of this study will be used as a broad term for changes in the downstream sequencing results due to the amplification step (e.g.
-higher GC content genomes, base substitutions, chimeras, changes from the expected relative abundance, etc.).
-
-Interestingly, the influence different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error.
-A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
-The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
-Yet, within this study there was no comparison made between different high fidelity DNA polymerases on the same samples.
-The authors instead focused their efforts on testing specific types of amplification methods.
-Additionally, the fecal samples that were analyzed, as a model of a more complex community, may overestimate the biological variation in human samples due to the comparison of captive versus semi-captive red-shanked doucs.
-Due to these gaps, it makes sense to extend the previous line of inquiry to include a more detailed examination of two specific areas.
-First, whether differences in PCR error are detected based on number of cycles used and high fidelity polymerases.
-Second, whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].
-
-This study will investigate whether there are error differences that are dependent on high fidelity DNA polymerases and whether these differences obscure the biological variation in the fecal bacterial communities between individuals.
-We accomplish this by first examining GC-based amplification error, sequence error (e.g.
-substitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases in mock communities.
-In addition to these metrics we also created Random Forest models of the mock community to assess how successful classification based on 16S rRNA gene sequencing of number of cycles and type of polymerase could be.
-We then assessed if similar differences in the number of OTUs, Bray-Curtis index, and OTUs could be detected in human fecal samples.
-Additionally, we also created Random Forest models to assess how successful they could be at classifying fecal samples by number of cycles, polymerase used, or individual.
-Collectively, our observations suggest that although detectable differences occur based on number of cycles or high fidelity DNA polymerase used, they are smaller than the biological differences between individuals.
-
-## pat's final paragraph
-We addressed X potential areas where the type of PCR polymerase was hypothesized to impact the results of a microbiome analysis.
-These included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process.
-To characterize these factors we sequenced a mock community with known sequences and community structure and human fecal samples with unknown sequences and community structures.
-
-Three factors can impact whether an organism's 16S rRNA gene sequence is correctly represented.
-These include mis-incorporated bases during the PCR amplification phase, mis-incorporated bases during sequencing, and creation of chimeras during the PCR amplification phase.
-
+This study investigated the impact of choice of high fidelity DNA polymerase and the number of rounds of amplification on the formation of PCR artifacts using a mock community and human stool samples.
+These PCR artifacts included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process.
+To characterize these factors we sequenced a mock community of 8 organisms with known sequences and community structure and human fecal samples with unknown sequences and community structures.
 We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform.
 This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol.
-The mock community is commercially available and was made up of genomic DNA from 8 bacteria.
-The benefit of using this mock community is that the sequences of the 16S rRNA genes from these bacteria is known.
-This allowed us to confidently identify any chimeras and to calculate the sequencing error rate.
-
-We selected five high fidelity DNA polymerases to amplify the V4 region of the 16S rRNA gene from a mock community containing genomic DNA from 8 previously sequenced bacteria.
-The genomic DNA from the mock community was amplified using 15, 20, 25, 30, and 35 rounds of amplification using one of five different polymerases.
-Each condition was replicated four times and reads for each condition were pooled.
-Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur.
-Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
-The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur.
-This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
-The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.
+To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 15, 20, 25, 30, and 35 rounds of amplification (Table 1).
+Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.
 
@@
 Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
-Both of these results indicate that for these samples, the differences were larger than the difference between conditions.
-These results suggest that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important.
-
-Generate PCoA diagrams
-Alpha diversity figures
-
-https://www.hindawi.com/journals/mbi/2014/287430/
-https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169774
-https://academic.oup.com/nar/article/24/18/3546/2359477
+However, when we performed the same analysis using the mock community, there were clear shifts in the represented community structures as the number of rounds of PCR increased.
+Using adonis, there were statistically significant differences in the community structures for the mock and stool community data (all P<0.001).
+Within the mock community data, the number of rounds of PCR explained the most variation (R^2^=0.XXX) followed by the choice of polymerase (R^2^=0.XXX).
+Similarly, within the stool community data, the subject explained the most variation (R^2^=0.XXX), followed by the number of rounds of PCR (R^2^=0.XXX) and the choice of polymerase (R^2^=0.XXX).
+Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
 
@@
 ## Discussion
+
+* Something like...
+our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals
+
+* be consistent within a study; do not interpret values like richness in absolute, but relative terms; take care in comparing results across studies
+
+* GC, sequence length non-issues
+
+* Other factors to investigate include impact of phylogenetic diversity, shearing, extension time, interactions of factors probably precludes ability to make definitive decisions.
+
+* Interpret results in light of polymerase characteristics - Accuprime and Platinum have error rates that are ~10-times lower than that of Taq whereas Phusion, Q5, and Kappa are >100 times lower.
+These error rates track well with what we saw for sequencing error rates
+
+* Minimize number of rounds of amplification - bias could be due to mis-determination of relative abundances by zymo
+
+* Use multiple methods to triangulate on reality/validate claims
 
@@
 
-Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community.
-Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4].
-One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987].
-The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence.
-Metrics that solely use presence/absence of OTUs (e.g.
-Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence.
-When choosing a distance metric, careful consideration of the errors introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.
-
-Collectively, our data shows that using distinct high fidelity DNA polymerases will result in a different number of OTUs being detected.
-The differences in the number of OTUs are primarily due to chimera prevalence and sequence error rates that are distinct to the specific polymerase.
-Despite this variation in the number of OTUs, our Random Forest models were able to correctly classify individuals regardless of the number of cycles and polymerase used.
-These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to fecal bacterial community composition differences between individuals will still be consistent.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ## Materials & Methods
+
+#Each condition was replicated four times and reads for each condition were pooled.
+Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur.
+Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
+The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur.
+This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
+The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.
 

### 2018-12-05 code/vegan_analysis.R
Rough in vegan analysis

@@ -1,0 +1,67 @@
+library(tidyverse)
+library(vegan)
+
+####################################################################################################
+
+read_dist <- function(dist_file_name){
+
+	linear_data <- scan(dist_file_name, what="character", sep="\n", quiet=TRUE)
+
+	n_samples <- as.numeric(linear_data[1])
+	linear_data <- linear_data[-1]
+
+	samples <- str_replace(linear_data, "\t.*", "")
+	linear_data <- str_replace(linear_data, "[^\t]*\t", "")
+	linear_data <- linear_data[-1]
+
+	distance_matrix <- matrix(0, nrow=n_samples, ncol=n_samples)
+
+	for(i in 1:(n_samples-1)){
+		row <- as.numeric(unlist(str_split(linear_data[i], "\t")))
+		distance_matrix[i+1,1:length(row)] <- row
+	}
+
+	distance_matrix <- distance_matrix + t(distance_matrix)
+	rownames(distance_matrix) <- samples
+
+	as.dist(distance_matrix)
+}
+
+
+####################################################################################################
+
+pmock_file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.perfect.opti_mcc.braycurtis.0.03.lt.ave.dist"
+
+pmock_dist <- read_dist(pmock_file_name)
+
+pmock_samples <- tibble(names = attr(pmock_dist, "Labels")) %>% mutate(names=str_replace(names, "_Mock_.*", "")) %>% separate(names, into=c("rounds", "polymerase"))
+
+pmock_adonis <- adonis(pmock_dist~pmock_samples$rounds+pmock_samples$polymerase, permutations=9999)
+pmock_r2 <- pmock_adonis$aov.tab$R2[1:2]
+names(pmock_r2) <- c("rounds", "polymerase")
+
+####################################################################################################
+
+vmock_file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.vsearch.opti_mcc.braycurtis.0.03.lt.ave.dist"
+
+vmock_dist <- read_dist(vmock_file_name)
+
+vmock_samples <- tibble(names = attr(vmock_dist, "Labels")) %>% mutate(names=str_replace(names, "_Mock_.*", "")) %>% separate(names, into=c("rounds", "polymerase"))
+
+vmock_adonis <- adonis(vmock_dist~vmock_samples$rounds+vmock_samples$polymerase, permutations=9999)
[…]

### 2018-12-10 submission/manuscript.Rmd
Working on Discussion

@@
 
-* Something like...
-our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals
+Our results suggest that the number of rounds of PCR and to a lesser degree the choice of DNA polymerase impact the analysis of 16S rRNA gene sequence data from bacterial communities.
+Although it was not possible to make direct connections between PCR conditions and specific sources of bias, we were able to identify general recommendations that reduce the amount of error, chimera formation, and bias.
+Researchers should strive to minimize the number of rounds of PCR and should use a high fidelity polymerase.
+Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between individuals.
+Based on these observations, within a study, amplicons must be generated by consistent protocols to yield meaningful comparisons.
+When comparing across studies, values like richness, diversity, and relative abundances must be made in relative and not absolute terms.
 
-* be consistent within a study; do not interpret values like richness in absolute, but relative terms; take care in comparing results across studies
+The observed sequencing error rates and alpha diversity metrics followed the manufacturers' measurements of their polymerases' fidelity.
+Accuprime and Platinum have fidelity that are approximately 10-times higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa are more than 100 times higher.
+Among these polymerases, the Kappa polymerase resulted in the the lowest error rate, lowest chimera rate, and least bias across rounds of PCR.
+Considering polymerase development is an active area of commercial development and it likely that new polymerases will come to market, it is important for researchers to understand how changing the polymerase impacts downstream analyses for their type of samples.
 
+### PCR bias
 * GC, sequence length non-issues
-
-* Other factors to investigate include impact of phylogenetic diversity, shearing, extension time, interactions of factors probably precludes ability to make definitive decisions.
-
-* Interpret results in light of polymerase characteristics - Accuprime and Platinum have error rates that are ~10-times lower than that of Taq whereas Phusion, Q5, and Kappa are >100 times lower.
-These error rates track well with what we saw for sequencing error rates
-
+* Other factors to investigate include impact of phylogenetic diversity, shearing, interactions of factors probably precludes ability to make definitive decisions.
 * Minimize number of rounds of amplification - bias could be due to mis-determination of relative abundances by zymo
-
-* Use multiple methods to triangulate on reality/validate claims
 
@@
 
-Fecal samples showed a similar increase in median and range for the number of OTUs, based on number of cycles and polymerase used, as that found when analyzing mock community samples [Figure 4A].
-However, there were few differences between the polymerases within each of the amplification cycles [Figure 4A].
-Additionally, these differences did not translate to large bacterial community composition changes based on number of cycles or polymerase.
-The differences that were identified were small and occurred based on the number of cycles, independent of the polymerase used [Figure 4B].
-There was also no OTU-based differences observed within the number of cycles between any of the polymerases for fecal samples.
-Interestingly, Random Forest models built on this OTU data had the highest classification success for individuals versus number of cycles or polymerase.
-In addition to the better model performance of Random Forest for individuals, the most important OTUs to this model were not the most important OTUs for the other models tested [Figure 5].
-Collectively, these observations show that differences based on the number of cycles and between high fidelity DNA polymerases can be found.
-However, when using 16S rRNA gene sequencing these technical differences can be smaller than the biological differences between fecal bacterial communities within individuals.
-
-The small differences observed due to error rate and chimera prevalence may be due to the actual DNA polymerase family being used within the PCR mix.
-DNA polymerases from distinct families have different binding affinities and error correction capacity [@polymerase_Ishino_2014].
-However, based on the observations within our study this is unlikely to be the only contributor.
-Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014].
-Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent.
-Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase, and a platinum *Taq* antibody.
-Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence.
-Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#Results have an impact on downstream diversity measurements due to effects on total OTUs
-
-#Need to cover that different types of polymerase may affect outcome
-#- Accuprime HiFi from recombinant A-Family Taq DNA polymerase, Pyrococcus spp GB-D polymerase (proof-reading), platinum Taq Antibody (inhibits room temp activity)
-
-#- Kappa HiFi from B-family DNA polymerase (proof-reading baked in), propietary antibody (inhibit room temp activity)
-
-#- Q5 HiFi from B-family thermostable DNA polymerase with 3´→ 5´ exonuclease activity, fused to a processivity-enhancing Sso7d domain to support robust DNA amplification
-
-#-Phusion HiFi B-family DNA polymerase a novel Pyrococcus-like enzyme fused with a processivity-enhancing domain, generate blunt-ended products.
-
-#- Platinum HiFi recombinant A-Family Taq DNA polymerase, Pyrococcus species GB-D polymeras, Platinum Taq Antibody
-
-```
-
-\newpage
-
-## Conclusion
+### Conclusion
+* Use multiple methods to triangulate on reality/validate claims
 

### 2018-12-11 submission/manuscript.Rmd
Work on Discussion and Materials & Methods

@@
 
-### PCR bias
-* GC, sequence length non-issues
-* Other factors to investigate include impact of phylogenetic diversity, shearing, interactions of factors probably precludes ability to make definitive decisions.
-* Minimize number of rounds of amplification - bias could be due to mis-determination of relative abundances by zymo
-
-Previous studies have reported GC dependent amplification differences [@polz_bias_1998] and we confirm that these differences still persist in high fidelity DNA polymerase when analyzing mock community samples [Figure 1].
-Additionally, these specific polymerases have different sequence error rates and chimera prevalence [Figure 2].
-Chimera prevalence in particular was strongly associated with the total number of OTUs detected in our mock community and varied by polymerase  [Figure 2E & 3A].
-Additionally, the bacterial community composition varied by number of cycles and polymerase in these mock community samples.
-Clearly, it is possible to identify a large number of differences based on number of cycles and polymerase used when analyzing mock community samples.
-However, these differences may not be a major concern when analyzing more complex community samples.
-
-### Conclusion
-* Use multiple methods to triangulate on reality/validate claims
-
+Over the past 20 years, a large literature has attempted to document various PCR biases and underscores the fact that data based on amplification of DNA from a mixed community are not a true representation of the actual community.
+In addition to obvious biases imposed by primer selection, other factors inherent in PCR can influence the representation of communities.
+Factors that can lead to preferential amplification of one fragment over another have included guanine and cytosine composition [@polz_bias_1998], length [@XXXXX], and flanking DNA composition [@XXXXX].
+Other factors including the degree that the DNA has been sheared [@XXXXX], contaminants [@XXXXX], and number of rounds of PCR [@XXXXX] are known to impact the generation of artifacts by PCR.
+Less well understood is the effect of phylogenetic diversity on bias and chimera formation.
+Communities with low phylogenetic diversity may be more prone to chimera formation since chimeras are more likely to form among closely related sequences [@XXXXX].
+The interaction of these various influences on PCR artifacts are complex and difficult to tease apart.
+Minimizing the level of DNA shearing and using the fewest number of rounds of PCR with a polymerase that has the highest possible fidelity are strategies that can be employed to minimize the formation of chimeras.
 Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are smaller than the actual biological variation in fecal communities between individuals.
 Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
+
+[@Eckert1991; @Burkardt2000] [@Wang1996; @Haas2011; @Kebschull2015] [@BautistadelosSantos2016; @Acinas2005]
+
+Even with these strategies it is impossible to remove all PCR artifacts.
+Beyond the imperfections of the best polymerases, sometimes difficult to lyse organisms require stringent lysis steps and low biomass samples require additional rounds of PCR.
+A host of bioinformatics tools are available for removing residual sequencing errors [@XXXXX] and chimeras [@XXXXX].
+Others recommend removing any read that has an abundance below a specified threshold (e.g.
+removing all sequences that only appear once).
+This must be approached with caution as such approaches are likely to introduce a different bias of the community representation and ignore the fact that artifacts may be quite abundant.
+Ultimately, researchers must test their hypotheses with multiple methods to validate the claims they reach with any one method [@XXXXX].
+All methods have biases and limitations and we must use complementary methods to obtain robust results.
 
@@
 
-#Each condition was replicated four times and reads for each condition were pooled.
-Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur.
-Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
-The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur.
-This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
-The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.
+***Mock community.***
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+The actual relative abundance for each bacterium was obtained from Zymo's certificate of analysis for the lot (Lot: ZRC187325), which they determined using shotgun metagenomic sequencing (https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017/data/references/ZRC187325.pdf).
 
-***Mock community and human samples.***
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
-The theoretical abundance of 0.12 used as a comparison point for our relative abundance was based on this equal genomic DNA composition of the mock community.
-Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN).
-The processing and storage of these samples were previously published [@erin_seekatz_2016].
-Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study.
-All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
+***Human samples.***
+Fecal samples were obtained from 4 individuals who were part of an earlier study [@erin_seekatz_2016].
+These samples were collected using a protocol approved by the University of Michigan Institutional Review Board.
+For this study, the samples were de-identified.
+DNA was extracted from the fecal samples using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
 
 ***PCR protocol.***
-The five different high fidelity DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer.
-The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md).
-For Kappa and Q5, the annealing and extension time also were from a previously published optimized protocol [@taq_Gohl_2016].
-For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.
+Five high fidelity DNA polymerases were tested including AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+Manufacturer recommendations were followed except for the annealing and extension times, which were selected based on previously published protocols [@protocol_Kozich_2013; @taq_Gohl_2016].
+Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@protocol_Kozich_2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md].
+The number of rounds of PCR used for each sample and polymerase 15 and increased by 5 rounds up to 35 cycles.
+Insufficient PCR product was generated using 15 rounds and has not been included in our analysis.
 
-The number of cycles in the PCR for mock and fecal samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing.
-The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130).
-No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.
-The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).
+***Library generation and sequencing.***
+Each PCR condition (i.e.
+combination of polymerase and number of rounds of PCR) were replicated four times for the mock community and one time for each fecal sample.
+Libraries were generated as described previously [@protocol_Kozich_2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md].
+The libraries were sequenced using the Illumina MiSeq sequencing platform to generate paired 250-nt reads.
 
 ***Sequence processing.***
-The mothur software program was used for all sequence processing steps [@mothur_schloss_2009].
+The mothur software program (v 1.41) was used for all sequence processing steps [@mothur_schloss_2009].
 The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017].
-Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P.
-aeruginosa*, *E.
-coli*, *S.
-enterica*, *L.
-fermentum*, *E.
-faecalis*, *S.
-aureus*, *L.
-monocytogenes*, and *B.
-subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].
-
-***Analysis workflow.***
-To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis.
-The analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed.
-Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.
-The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples.
-Finally, Random Forest models were created to assess whether classification of polymerase, number of cycles, or individual could be performed best using 16S rRNA gene sequencing data.
-For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual label was used.
-Both number of cycles and polymerase models included all samples but only the number of cycles or polymerase label was used for each respective model.
-Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).
+Briefly, paired reads were assembled using mothur's make.contigs command to correct errors introduced by sequencing [@XXXXX].
+Any assembled contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
+Sequences were further denoised using mothur's pre.cluster command to merge the counts of sequences that were within 2 nt of a more abundant sequence.
+The VSEARCH implementation of UCHIME was used to screen for chimeras [@vsearch_Rognes_2016; @uchime_Edgar_2011].
+At various stages in the sequence processing pipeline for the mock community data, the mothur seq.error command was used to quantify the sequencing error rate as well as the true chimera rate.
+This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
+The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence and reports the error rate without the inclusion of chimeric sequences [@XXXXX].
+The reference sequences and operon copy number for each bacterium were obtaind from the ZymoBIOMICS^TM^ Microbial Community DNA Standard protocol (https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf).
+Sequences were assigned to operational taxonomic units (OTUs) at a threshold of 3% dissimilarity using the OptiClust algorithm [@XXXXX].
+To adjust for unequal sequencing when measuring alpha and beta diversity, all samples were rarefied to 1,000 sequences 1,000 times for downstream analysis.
+The Good's coverage for the samples at this level of sampling was routinely greater than 95%.
 
@@
 All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
-High and low GC content was determined based on the median GC percentage of either the V4 region or the average GC content of the whole genome of the bacterial species used in the mock community.
-Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test.
-For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation.
-Bray-Curtis distance matrices were generated using mothur.
-The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R.
-The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation].
-A total of 100 10-fold CV runs on different 80/20 splits of the data were run to generate a range of the Logloss value.
-The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm.
-For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
-Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].
+Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
+The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation].
 
 ***Reproducible methods.***
-The code and analysis for this study can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
+The data analysis code for this study can be found at https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences are available at the SRA (Accession SRP132931).
 

### 2018-12-13 submission/manuscript.Rmd
Edit results section

@@
 ## Abstract
-
-**Background.** A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, amplification, and sequencing steps.
-At each of these stages error can be introduced that will change the underlying bacterial community composition results.
-In this study, we focus on the amplification step's contribution to this overall error.
-To accomplish this we assessed 16S rRNA gene sequencing results in mock community and human fecal samples after using different  amplification cycles and high fidelity DNA polymerases.
-
-**Methods.** Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
-For fecal samples (n=4) we first extracted DNA using a PowerMag DNA extraction kit with a 10 minute bead beating step and then used the same approach as the mock community for the PCR amplification step.
-We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles.
-Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used.
-Additionally, differences in the bacterial community composition was assessed based on polymerase and number of cycles using the Bray-Curtis index.
-Based on these findings individual taxa differences based on polymerase and number of cycles was also investigated.
-Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor.
-We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models were also important in the model for individual donors.
-
-**Results.** Predictably, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`).
-Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH.
-We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`).
-When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`).
-When analyzing fecal samples we observed smaller differences in the number of OTUs between polymerases.
-The median number of OTUs only varied by high fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001).
-Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`).
-Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor model.
-
-**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences were mostly only detectable in mock communities.
-Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of human fecal samples, can be consistently detected regardless of polymerase and number of cycles used.
 
@@
 More importantly, sequencing errors themselves can alter the representation of the community.
-Therefore, it is important to minimize the number of sequencing errors in an analysis.
+Therefore, it is important to minimize the number of sequencing errors.
 Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community.
@@
 0.XXX%) polymerases (Figure 1).
-The order of the error rates using these polymerases is consistent with previously documented reports of their per cycle error rates.
 When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e.
@@
 0.XXX%) polymerases (Figure 1).
-Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary is a good guide to minimizing the impact of polymerase on the error rate.
-Furthermore, using a polymerase with the lowest possible error rate is preferred.
+Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.
 
@@
 We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity.
-For all polymerases, the specificity was above XX% and showed a slight association with the number of cycles used.
+For all polymerases, the specificity was above XX% and showed a weak association with the number of cycles used.
 There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community.
-This suggests that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
-In general, the sensitivity of UChime varied between 50 and 90%.
-These results could be affected by the use of a single mock community with limited genetic diversity.
+This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
+In general, the sensitivity of UChime varied between XX and XX%.
+The generalizability of these results is limited because we used a single mock community with limited genetic diversity.
 Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric.
-These results followed those from the mock communities - additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
-Although it is difficult to identify a feature of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
+These results followed those from the mock communities; additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
+Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
 
-***The bias of PCR amplification is inconsistent and subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses.
+***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses.
 The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria.
@@
 We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase.
-Interestingly, for some bacteria, there representation became less biased with additional rounds of PCR (e.g.
+Interestingly, for some bacteria, their representation became less biased with additional rounds of PCR (e.g.
 *L.
@@
 The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
-Across the bacteria there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed.
+Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed.
 One of the bacteria represented in the mock community, *S.
 enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide.
+The dominant copy had a thymidine and the rare copy had a guanine.
 We used the sequence data to calculate the ratio of the dominant to rare variants from *S.
@@
 XXX).
-With the exception of the Kappa polymerase, the other four polymerases converged to a ratio of 5.XXX.
-Based on the relative abundance data from the mock community it was not possible to discern the factors that result in amplification bias.
-As an additional test, we quantified the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification.
-For the mock community data, the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units.
+The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of PCR.
+Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias and the previous observation regarding the guanine and cytosine relative abundance was not observed with our data.
+
+***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline.
+First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors.
+This gave us a community distribution that reflected the distribution following PCR without any artifacts.
+Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased.
+These data suggest that PCR had the effect of making the community distribution more even than it was originally.
+Next, we used the true sequence diversity, but removed chimeras based on mapping reads to all possible chimeras between reference sequences, and clustered the reads to OTUs.
+The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher.
+Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras.
+Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher.
+These comparisons demonstrated that although the bias at the sequence level was subtle, PCR does have a bias that is exacerbated by errors and chimeras.
+When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units.
 The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX).
-In contrast, when we characterized the variation between amplification rounds using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases.
+Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR.
+This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=0.XX, P=0.XXXX) was a larger factor than the choice of polymerase (R^2^=0.XX, P=0.XXXX).
+These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure.
+This variation underscores the importance of only comparing sequence data that have been generated using the same PCR conditions.
+
+***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels using mock community data appeared to be small relative to the expected differences between biological samples.
+To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
+We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor.
+Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
+When we characterized the variation between rounds of amplification using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases.
 In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases.
@@
 Considering the average distance across the four samples varied between 0.XX and 0.XX, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
+In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by donor.
+Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
+However, our statistical analysis using adonis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=0.XXX, P=0.XXXX), followed by the number of rounds of PCR (R^2^=0.XXX, P=0.XXXX) and the choice of polymerase (R^2^=0.XXX, P=0.XXXX).
+Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
 It is difficult to develop a specific recommendation based on the level of bias across rounds of PCR or polymerases; however, the general suggestion is to use as few rounds of amplification as possible.
@@
 These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.
-
-***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels appeared to be small relative to the overall differences between samples.
-To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
-We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor.
-Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
-We also performed a principle coordinates analysis of the Bray-Curtis distances showed distinct clusters by donor.
-Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
-However, when we performed the same analysis using the mock community, there were clear shifts in the represented community structures as the number of rounds of PCR increased.
-Using adonis, there were statistically significant differences in the community structures for the mock and stool community data (all P<0.001).
-Within the mock community data, the number of rounds of PCR explained the most variation (R^2^=0.XXX) followed by the choice of polymerase (R^2^=0.XXX).
-Similarly, within the stool community data, the subject explained the most variation (R^2^=0.XXX), followed by the number of rounds of PCR (R^2^=0.XXX) and the choice of polymerase (R^2^=0.XXX).
-Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
 

### 2018-12-13 code/plot_stool_pcoa.R
Fix stool pcoa diagram

@@ -11,6 +11,6 @@ read_tsv(file_name) %>%
 	separate(group, into=c("rounds", "polymerase", "subject")) %>%
 	mutate(rounds = str_replace(rounds, "x", "")) %>%
 	ggplot(aes(x=axis1, y=axis2, color=polymerase, shape=rounds)) +
+		stat_ellipse(aes(group=subject), type="norm", geom="polygon", color="gray", fill="lightgray") +
 		geom_point() +
-		stat_ellipse(aes(group=subject), type="norm", color="gray", fill="gray") +
 		ggsave("results/figures/stool_pcoa.pdf")

### 2018-12-21 submission/manuscript.Rmd
Fix references

@@
 title: ''
-csl: peerj.csl
+csl: mbio.csl
 fontsize: 11pt
@@
 
-https://www.hindawi.com/journals/mbi/2014/287430/
-https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169774
-https://academic.oup.com/nar/article/24/18/3546/2359477
-
 ## Introduction
 
-16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities.
+16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities [@Gilbert2018; @HMP2012; @Schloss2016].
 This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample.
-Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples, storage conditions, extraction methods, amplification conditions, sequencing method, and analytical pipeline.
+Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017].
 The increased sampling depth that is now available relative to previous Sanger-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.
 
-One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification.
-Factors such as the choice of primers have an obvious impact on which populations will be amplified.
-However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras, misincorporation of nucleotides, preferential amplification of populations leading to bias, and accumulation of random amplification events leading to PCR drift.
-Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs.
-Others have attempted to account for PCR bias using modeling approaches; however, these have been developed for idealized situations.
-Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability.
-To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift.
-Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community.
+One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification [@BautistadelosSantos2016; @Haas2011; @Sinha2017].
+Factors such as the choice of primers have an obvious impact on which populations will be amplified [@Parada2015; @Kozich2013].
+However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras [@Haas2011; @Wang1996; @Potapov2017; @Kebschull2015], misincorporation of nucleotides [@Potapov2017; @McInerney2014; @Cline1996], preferential amplification of populations leading to bias [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998], and accumulation of random amplification events leading to PCR drift [@Kebschull2015; @Kennedy2014; @Acinas2005; @Wagner1994].
+Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011].
+Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015]; however, these have been developed for idealized situations.
+Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017].
+To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift [@Kennedy2014].
+Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community [@Polz1998; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998].
 Early investigations of the factors that lead to the formation of PCR artifacts focused on analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation.
@@
 The influence that the choice of DNA polymerase has on the formation of PCR artifacts has not been well studied.
-There has been a lot of recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate; however, these studies use the same DNA polymerase in the PCR step and assume that the rate of misincorporation of nucleotides due to PCR are significantly smaller than those from the sequencing phase.
-There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras.
-A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016].
-The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016].
+There has been recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017]; however, these studies use the same DNA polymerase in the PCR step and implicitly assume that the rate of nucleotide misincorporation from PCR are significantly smaller than those from the sequencing phase.
+There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras [@Potapov2017; @Gohl2016].
+A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@Gohl2016].
+The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@Gohl2016].
 Yet this optimization was specific for the community they were analyzing (i.e.
-captive and semi-captive red-shanked doucs) and assumes that if the two polymerases generate the same community structure that the community structure is correct.
-In fact, the community structure generated by both methods is not free of artifacts, but has the same artifacts.
-A challenge in these types of experiments is a priori knowledge of the true community representation.
-A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias; however, mock communities have a limited phylogenetic diversity relative to natural communities.
+captive and semi-captive red-shanked doucs) and assumed that if the two polymerases generate the same community structure that the community structure was correct.
+In fact, the community structure generated by both methods was not free of artifacts, but has the same artifacts.
+A challenge in these types of experiments is having a priori knowledge of the true community representation.
+A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias [@Schloss2011]; however, mock communities have a limited phylogenetic diversity relative to natural communities.
 Natural communities, in contrast, have an unknown community composition making absolute measurements impossible.
@@
 In addition to obvious biases imposed by primer selection, other factors inherent in PCR can influence the representation of communities.
-Factors that can lead to preferential amplification of one fragment over another have included guanine and cytosine composition [@polz_bias_1998], length [@XXXXX], and flanking DNA composition [@XXXXX].
-Other factors including the degree that the DNA has been sheared [@XXXXX], contaminants [@XXXXX], and number of rounds of PCR [@XXXXX] are known to impact the generation of artifacts by PCR.
+Factors that can lead to preferential amplification of one fragment over another have included guanine and cytosine composition, length, flanking DNA composition, amount of DNA shearing, and number of rounds of PCR [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998].
+In addition, environmental and reagent contaminants can also have a significant impact on the analysis of low biomass samples [@Salter2014].
 Less well understood is the effect of phylogenetic diversity on bias and chimera formation.
-Communities with low phylogenetic diversity may be more prone to chimera formation since chimeras are more likely to form among closely related sequences [@XXXXX].
+Communities with low phylogenetic diversity may be more prone to chimera formation since chimeras are more likely to form among closely related sequences [@Haas2011; @Edgar2011].
 The interaction of these various influences on PCR artifacts are complex and difficult to tease apart.
@@
 
-[@Eckert1991; @Burkardt2000] [@Wang1996; @Haas2011; @Kebschull2015] [@BautistadelosSantos2016; @Acinas2005]
-
 Even with these strategies it is impossible to remove all PCR artifacts.
 Beyond the imperfections of the best polymerases, sometimes difficult to lyse organisms require stringent lysis steps and low biomass samples require additional rounds of PCR.
-A host of bioinformatics tools are available for removing residual sequencing errors [@XXXXX] and chimeras [@XXXXX].
+A host of bioinformatics tools are available for removing residual sequencing errors [@Kozich2013; @Callahan2016; @Edgar2016; @Amir2017a].
+These tools struggle to correctly differentiate true biological diversity (i.e.
+Amplicon Sequencing Variants or ASVs) and PCR or sequencing errors.
+Although these methods remove residual errors, they also risk splitting intragenomic variants into separate ASVs, merging 16S rRNA gene sequences from different taxa into the same ASV.
+Other tools are available for removing chimeras [@Edgar2011; @Haas2011] where there is a trade off between the sensitivity of detecting chimeras and the specificity of correctly calling a sequence a chimera.
+In recent years, parameters for these algorithms have been changed to increase their sensitivity with little evaluation of the effects on the specificity of the algorithms [@Callahan2016; @Amir2017a].
 Others recommend removing any read that has an abundance below a specified threshold (e.g.
-removing all sequences that only appear once).
-This must be approached with caution as such approaches are likely to introduce a different bias of the community representation and ignore the fact that artifacts may be quite abundant.
-Ultimately, researchers must test their hypotheses with multiple methods to validate the claims they reach with any one method [@XXXXX].
+removing all sequences that only appear once) [@Bokulich2012; @Callahan2016; @Edgar2016; @Amir2017a].
+This method must be approached with caution as such approaches are likely to introduce a different bias of the community representation and ignore the fact that artifacts may be quite abundant.
+Ultimately, researchers must test their hypotheses with multiple methods to validate the claims they reach with any one method [@Schloss2018].
 All methods have biases and limitations and we must use complementary methods to obtain robust results.
@@
 ***Mock community.***
-The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
-The actual relative abundance for each bacterium was obtained from Zymo's certificate of analysis for the lot (Lot: ZRC187325), which they determined using shotgun metagenomic sequencing (https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017/data/references/ZRC187325.pdf).
+The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (https://web.archive.org/web/20171217151108/http://www.zymoresearch.com:80/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).
+The actual relative abundance for each bacterium was obtained from Zymo's certificate of analysis for the lot (Lot: ZRC187325), which they determined using shotgun metagenomic sequencing (https://github.com/SchlossLab/Sze_PCRSeqEffects_mSphere_2019/data/references/ZRC187325.pdf).
 
 ***Human samples.***
-Fecal samples were obtained from 4 individuals who were part of an earlier study [@erin_seekatz_2016].
+Fecal samples were obtained from 4 individuals who were part of an earlier study [@Seekatz2016].
 These samples were collected using a protocol approved by the University of Michigan Institutional Review Board.
@@
 Five high fidelity DNA polymerases were tested including AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
-Manufacturer recommendations were followed except for the annealing and extension times, which were selected based on previously published protocols [@protocol_Kozich_2013; @taq_Gohl_2016].
-Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@protocol_Kozich_2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md].
+Manufacturer recommendations were followed except for the annealing and extension times, which were selected based on previously published protocols [@Kozich2013; @Gohl2016].
+Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@Kozich2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/].
 The number of rounds of PCR used for each sample and polymerase 15 and increased by 5 rounds up to 35 cycles.
@@
 combination of polymerase and number of rounds of PCR) were replicated four times for the mock community and one time for each fecal sample.
-Libraries were generated as described previously [@protocol_Kozich_2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md].
+Libraries were generated as described previously [@Kozich2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/].
 The libraries were sequenced using the Illumina MiSeq sequencing platform to generate paired 250-nt reads.
@@
 ***Sequence processing.***
-The mothur software program (v 1.41) was used for all sequence processing steps [@mothur_schloss_2009].
-The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP).
-Briefly, paired reads were assembled using mothur's make.contigs command to correct errors introduced by sequencing [@XXXXX].
+The mothur software program (v 1.41) was used for all sequence processing steps [@Schloss2009].
+The protocol has been previously published [@Kozich2013] (https://www.mothur.org/wiki/MiSeq_SOP).
+Briefly, paired reads were assembled using mothur's make.contigs command to correct errors introduced by sequencing [@Kozich2013].
 Any assembled contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses.
 Sequences were further denoised using mothur's pre.cluster command to merge the counts of sequences that were within 2 nt of a more abundant sequence.
-The VSEARCH implementation of UCHIME was used to screen for chimeras [@vsearch_Rognes_2016; @uchime_Edgar_2011].
+The VSEARCH implementation of UCHIME was used to screen for chimeras [@Rognes2016; @Edgar2011].
 At various stages in the sequence processing pipeline for the mock community data, the mothur seq.error command was used to quantify the sequencing error rate as well as the true chimera rate.
 This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence.
-The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence and reports the error rate without the inclusion of chimeric sequences [@XXXXX].
-The reference sequences and operon copy number for each bacterium were obtaind from the ZymoBIOMICS^TM^ Microbial Community DNA Standard protocol (https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf).
-Sequences were assigned to operational taxonomic units (OTUs) at a threshold of 3% dissimilarity using the OptiClust algorithm [@XXXXX].
+The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence and reports the error rate without the inclusion of chimeric sequences [@Schloss2011].
+The reference sequences and operon copy number for each bacterium were obtained from the ZymoBIOMICS^TM^ Microbial Community DNA Standard protocol (https://web.archive.org/web/20181221151905/https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf).
+Sequences were assigned to operational taxonomic units (OTUs) at a threshold of 3% dissimilarity using the OptiClust algorithm [@Westcott2017].
 To adjust for unequal sequencing when measuring alpha and beta diversity, all samples were rarefied to 1,000 sequences 1,000 times for downstream analysis.
@@
 ***Statistical analysis.***
-All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017].
-The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation].
+All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2018].
+Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2017].
+The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`) [@Vegan2018].
 
 ***Reproducible methods.***
-The data analysis code for this study can be found at https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences are available at the SRA (Accession SRP132931).
+The data analysis code for this study can be found at https://github.com/SchlossLab/Sze_PCRSeqEffects_mSphere_2019. The raw sequences are available at the SRA (Accession SRP132931).
 
@@
 
-**Figure 1: Relative abundance differences due to GC is consistent across polymerases.** At each number of cycle the points and lines represent the median and with the minimum and maximum relative abundance respectively.
-Regardless of the number of cycle used a consistent difference between high and low GC bacteria was observed.
-The dotted line represents the actual relative abundance that each bacterium should be at within the mock community sample.
-
-**Figure 2: Error rate and chimera prevalence vary by polymerase and affect the number of observed OTUs.** A) The error bars represent the 75% interquartile range of the median error rate.
-B) Percentage of chimeric sequences without the removal of chimeras with VSEARCH.
-C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH.
-D) The total percent of chimeric sequences removed with VSEARCH by cycle number.
-The error bars represent the 75% interquartile range of the median.
-E) Chimera prevalence and the the observed number of OTUs are strongly correlated in mock community samples.
-
-**Figure 3: Subtle differences  based on number of cycles and polymerase used are detected in bacterial community composition of mock samples.** A) The range in the number of OTUs detected in the mock samples increased as cycle number increased.
-This range was larger for specific HiFi DNA polymerases.
-The points represent the median number of OTUs and the lines represent the range from the minimum to maximum number of OTUs detected within the four technical replicates.
-The dotted black line represents the number of OTUs detected when only the references sequences for the mock community are clustered.
-A) Within replicate difference based on the next 5-cycle PCR interval in mock samples.
-The lines represent the range of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle increment comparison.
-The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.
-
-**Figure 4: Subtle differences based on number of cycles and polymerase used are detected in bacterial community composition of fecal samples.** A) The range in the number of OTUs detected in the different fecal samples increased as cycle number increased.
-This range was larger for specific HiFi DNA polymerases.
-The points represent the median number of OTUs and the lines represent the range from the minimum to maximum number of OTUs detected within the four fecal samples.
-B) Within person differences based on the next 5-cycle PCR interval in fecal samples.
-The points represent the median Bray-Curtis index for the samples.
-The lines represent the range of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle increment comparison.
-The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.
-
-**Figure 5: Important OTUs for the number of cycles and polymerase models are not the most important OTUs in the model classifying individuals.** Color highlights the top 10 OTUs in number of cycles and polymerase used and where they fall in the model used to classify individuals.
-The majority of these top 10 OTUs have 0 importance to the model used to classify individuals.
+**Figure 1: XXXXXXXX.**
 
@@
 
-**Figure S1: No preference for specific base subsitutions was observed across the polymerases in mock community samples.**
+**Figure S1: XXXXXXXX.**

### 2018-12-21 submission/manuscript.tex
Fix references

@@
 \setlength{\droptitle}{-2em}
+
   \title{}
@@
 
+\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
+errors when sequencing 16S rRNA genes from complex
+communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}
+
 \begin{center}
@@
 
-\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
-divided into preservation, extraction, amplification, and sequencing
-steps. At each of these stages error can be introduced that will change
-the underlying bacterial community composition results. In this study,
-we focus on the amplification step's contribution to this overall error.
-To accomplish this we assessed 16S rRNA gene sequencing results in mock
-community and human fecal samples after using different amplification
-cycles and high fidelity DNA polymerases.
-
-\textbf{Methods.} Amplification of mock communities (technical
-replicates n=4) consisting of previously isolated whole genomes of 8
-different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using
-Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
-For fecal samples (n=4) we first extracted DNA using a PowerMag DNA
-extraction kit with a 10 minute bead beating step and then used the same
-approach as the mock community for the PCR amplification step. We first
-assessed GC dependent differences, error rate, sequence error
-prevalence, chimera prevalence, and correlation between chimera
-prevalence and number of Operational Taxonomic Units (OTUs) by
-polymerase and number of cycles. Next, differences in the number of OTUs
-was examined based on the polymerase and number of cycles used.
-Additionally, differences in the bacterial community composition was
-assessed based on polymerase and number of cycles using the Bray-Curtis
-index. Based on these findings individual taxa differences based on
-polymerase and number of cycles was also investigated. Finally, Random
-Forest models were created to test whether the bacterial community was
-better at classifying polymerases, number of cycles, or individual
-donor. We also assessed whether the most important taxa in the
-polymerase and number of cycle Random Forest models were also important
-in the model for individual donors.
-
-\textbf{Results.} Predictably, we found noticeable differences in
-relative abundance based on high and low GC content (P-value
-\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
-polymerase with differences being most notable at 35 cycles (Kappa =
-5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
-after chimera removal using VSEARCH. We also observed positive
-correlations between chimera prevalence and the number of OTUs with
-Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
-having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
-community samples the variation in the number of OTUs detected by the
-polymerases was observable as early as 20 cycles (P-value = 0.002). When
-analyzing fecal samples we observed smaller differences in the number of
-OTUs between polymerases. The median number of OTUs only varied by high
-fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the
-mock community (P-value \textless{} 0.0001). Random Forest models were
-most successful at classifying individual donor samples rather than
-polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
-Additionally, the most important OTUs in the polymerase and number of
-cycle models were not the most important in the individual donor model.
-
-\textbf{Conclusions.} Although there were 16S rRNA gene sequencing
-differences based on polymerase and number of cycles used, bacterial
-community composition differences were mostly only detectable in mock
-communities. Collectively, these results provide evidence that smaller
-biological differences between groups, based on 16S rRNA gene sequencing
-of human fecal samples, can be consistently detected regardless of
-polymerase and number of cycles used.
-
 \newpage
@@
 
-The bacterial community is reported to vary between case and control for
-a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
-et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
-the taxa identified have varied widely depending on the study (Turnbaugh
-et al., 2008; Zupancic et al., 2012). Although some of the differences
-could be attributed to study design, error introduced during the 16S
-rRNA gene sequencing workflow could also be another source of variation.
-A typical 16S rRNA gene sequencing workflow can be divided into
-preservation, extraction, PCR, and sequencing steps. The preservation
-and extraction stages of the 16S rRNA gene sequencing workflow have been
-the most extensively studied (Salter et al., 2014; Song et al., 2016;
-Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
-stages of the workflow, it has been consistently found that there are
-errors in the observed bacterial community composition based on the kits
-used, but that these differences are smaller than the overall biological
-difference measured between samples with different kits (Song et al.,
-2016; Bassis et al., 2017). Since these studies use the same PCR
-approach while varying preservation or extraction method, the
-contribution of PCR error to this overall workflow is not as well
-characterized.
+16S rRNA gene sequencing is a powerful and widely used tool for
+surveying the structure of microbial communities (1--3). This approach
+has exploded in popularity with the advent of high throughput sequencing
+where it is possible to characterize numerous samples with thousands of
+sequences per sample. Numerous factors can impact how a natural
+community is represented by the sequencing data including the method of
+acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
+methods (13), amplification conditions (8, 14, 15), sequencing method
+(15--17), and analytical pipeline (15, 18--20). The increased sampling
+depth that is now available relative to previous Sanger-based methods is
+expected to compound the impacts of an investigator's choices and the
+interpretation of their results.
 
-There is a large body of literature that shows there are errors due to
-primer and number of cycles chosen for the PCR stage of 16S rRNA gene
-sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
-variable region dependent binding affinities which causes an inability
-to detect specific bacteria (e.g.~V1-V3 does not detect
-\emph{Haemophilus influenzae} and V3-V5 does not detect
-\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
-al., 2016). Another source of error is the selective amplification of
-AT-rich over GC-rich sequences which exaggerate the difference between
-16S rRNA genes higher in AT versus those higher in GC content (Polz \&
-Cavanaugh, 1998). Many of these sources of error are made worse as the
-number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
-Kebschull \& Zador, 2015). For example, both amplification error and
-non-specific amplification (e.g.~incorrect amplicon size products) also
-can increase as the number of cycles increases. This will increase the
-number of Operational Taxonomic Units (OTUs) observed and drastically
-change the values obtained from commonly used diversity measures (Acinas
-et al., 2005; Santos et al., 2016). Additionally, as the number of
-cycles increases more chimeras can form from an aborted extension step
-that causes a priming error and subsequent secondary extension (Haas et
-al., 2011). These chimeras will artificially increase community
-diversity by increasing the number of OTUs that are observed (Haas et
-al., 2011). In addition to these sources of errors, there also are
-multiple families of DNA polymerases that have their own error rate and
-proof reading capacity (Ishino \& Ishino, 2014). For simplicity, the use
-of the term `error' within the context of this study will be used as a
-broad term for changes in the downstream sequencing results due to the
-amplification step (e.g.~higher GC content genomes, base substitutions,
-chimeras, changes from the expected relative abundance, etc.).
+One step in the generation of 16S rRNA gene sequence data that has been
+long known to have a significant impact on the description of microbial
+communities is the choice of conditions for PCR amplification (8, 14,
+15). Factors such as the choice of primers have an obvious impact on
+which populations will be amplified (18, 21). However, a variety of PCR
+artifacts can also impact the perception of a community including the
+formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
+25, 26), preferential amplification of populations leading to bias (24,
+27--33), and accumulation of random amplification events leading to PCR
+drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
+identify chimeras, there are significant sensitivity and specificity
+tradeoffs (14, 35). Others have attempted to account for PCR bias using
+modeling approaches (29, 36); however, these have been developed for
+idealized situations. Laboratory-based solutions to minimize chimera
+formation have also been proposed such as minimizing the amount of
+template DNA in the PCR, minimizing the number of rounds of PCR,
+minimizing the amount of shearing in the template DNA, and using DNA
+polymerases that have a proof-reading ability (14, 23). To minimize PCR
+drift, some investigators will pool technical replicate PCRs hoping to
+average out the drift (34). Other factors that have been shown to impact
+the formation of PCR artifacts are outside the control of an
+investigator including the fraction of DNA bases that are guanines or
+cytosines, the variation in the length of the targeted region across the
+community, the sequence of the DNA that flanks the template, and the
+genetic diversity of the community (28, 30--33). Early investigations of
+the factors that lead to the formation of PCR artifacts focused on
+analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
+explore PCR biases and chimera formation. Although these studies were
+instrumental in forcing researchers to be cautious about the
+interpretation of their results, we have a poor understanding of how
+these factors affect the formation of PCR artifacts in more complex
+communities.
 
-Interestingly, the influence different DNA polymerases can have on the
-observed 16S rRNA gene sequencing results have not been well studied
-like some of the other sources of PCR-based error. A recent study found
-differences in the number of OTUS and chimeras between normal and high
-fidelity DNA polymerases (Gohl et al., 2016). The authors of this study
-could reduce the difference between two polymerases by optimizing the
-annealing and extension steps within the PCR protocol (Gohl et al.,
-2016). Yet, within this study there was no comparison made between
-different high fidelity DNA polymerases on the same samples. The authors
-instead focused their efforts on testing specific types of amplification
-methods. Additionally, the fecal samples that were analyzed, as a model
-of a more complex community, may overestimate the biological variation
-in human samples due to the comparison of captive versus semi-captive
-red-shanked doucs. Due to these gaps, it makes sense to extend the
-previous line of inquiry to include a more detailed examination of two
-specific areas. First, whether differences in PCR error are detected
-based on number of cycles used and high fidelity polymerases. Second,
-whether these differences in error are enough to obscure smaller
-biological signals than what has been previously reported (Gohl et al.,
-2016).
+The influence that the choice of DNA polymerase has on the formation of
+PCR artifacts has not been well studied. There has been recent interest
+in how the choice of the hypervariable region and data analysis
+pipelines impact the sequencing error rate (15, 18--20); however, these
+studies use the same DNA polymerase in the PCR step and implicitly
+assume that the rate of nucleotide misincorporation from PCR are
+significantly smaller than those from the sequencing phase. There has
+been more limited interest in the impact that DNA polymerase choice has
+on the formation of chimeras (23, 37). A recent study found differences
+in the number of OTUs and chimeras between normal and high fidelity DNA
+polymerases (37). The authors of this study could reduce the difference
+between two polymerases by optimizing the annealing and extension steps
+within the PCR protocol (37). Yet this optimization was specific for the
+community they were analyzing (i.e.~captive and semi-captive red-shanked
+doucs) and assumed that if the two polymerases generate the same
+community structure that the community structure was correct. In fact,
+the community structure generated by both methods was not free of
+artifacts, but has the same artifacts. A challenge in these types of
+experiments is having a priori knowledge of the true community
+representation. A mock community with known composition allows
+researchers to quantify the sequencing error rate, fraction of chimeras,
+and bias (19); however, mock communities have a limited phylogenetic
+diversity relative to natural communities. Natural communities, in
+contrast, have an unknown community composition making absolute
+measurements impossible. They can be used to validate results from mock
+communities and to understand the relative impacts of artifacts on the
+ability to differentiate biological and methodological sources of
+variation. Given the large number of DNA polymerases available to
+researchers it is unlikely that a specific recommendation is possible.
+Rather, the development of general best practices and understanding the
+impact of PCR artifacts on an analysis are desired.
 
-This study will investigate whether there are error differences that are
-dependent on high fidelity DNA polymerases and whether these differences
-obscure the biological variation in the fecal bacterial communities
-between individuals. We accomplish this by first examining GC-based
-amplification error, sequence error (e.g.~substitutions), chimera
-prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
-differences at varying number of cycles in five different high fidelity
-DNA polymerases in mock communities. In addition to these metrics we
-also created Random Forest models of the mock community to assess how
-successful classification based on 16S rRNA gene sequencing of number of
-cycles and type of polymerase could be. We then assessed if similar
-differences in the number of OTUs, Bray-Curtis index, and OTUs could be
-detected in human fecal samples. Additionally, we also created Random
-Forest models to assess how successful they could be at classifying
-fecal samples by number of cycles, polymerase used, or individual.
-Collectively, our observations suggest that although detectable
-differences occur based on number of cycles or high fidelity DNA
-polymerase used, they are smaller than the biological differences
-between individuals.
+This study investigated the impact of choice of high fidelity DNA
+polymerase and the number of rounds of amplification on the formation of
+PCR artifacts using a mock community and human stool samples. These PCR
+artifacts included (i) the effect of the polymerase on the error rate of
+the bases represented in the final sequences, (ii) the fraction of
+sequences that appeared to be chimeras and the ability to detect those
+chimeras, (iii) the bias of preferentially amplifying one fragment over
+another in a mixed pool of templates, and (iv) inter-sample variation in
+community structure of samples amplified with the same polymerase across
+the amplification process. To characterize these factors we sequenced a
+mock community of 8 organisms with known sequences and community
+structure and human fecal samples with unknown sequences and community
+structures. We sequenced the V4 region of the 16S rRNA genes from a mock
+community by generating paired 250 nt reads on the Illumina MiSeq
+platform. This region and sequencing approach was used because it has
+been shown to result in a relatively low sequencing error rate and is a
+widely used protocol. To better understand the impact of DNA polymerase
+choice on PCR artifacts, we selected five high fidelity DNA polymerases
+and amplified the communities using 15, 20, 25, 30, and 35 rounds of
+amplification (Table 1). Collectively, our results suggest that the
+number of rounds and to a lesser degree the choice of DNA polymerase
+used in PCR impact the sequence data, the effects are consistent and are
+smaller than the biological differences between individuals.
 
@@
 
-\textbf{\emph{Error differences due to number of cycles or polymerase
-used are detectable in mock communities.}} There was a significant
-difference in relative abundance between high/low GC content based on
-either the V4 16S rRNA gene region or the whole genome {[}Figure 1 \&
-Table S1{]}. However, using only the GC content of the V4 region
-resulted in less differences than when using the whole genome {[}Table
-S1{]}. The differences between high and low GC content groups were
-significant as early as 20 cycles in all polymerases except Q5 {[}Table
-S1{]}. The highest (\emph{Staphylococcus}) and lowest
-(\emph{Pseudomonas}) GC content bacteria were the most divergent from
-the expected relative abundance of 0.12 {[}Figure 1{]}. These
-observations suggest that the high fidelity DNA polymerases tested do
-not prevent GC-based amplification error.
+\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
+polymerase used in PCR.}} The presence of sequence errors can confound
+the ability to accurately classify 16S rRNA gene sequences and group
+sequences into operational taxonomic units (OTUs). More importantly,
+sequencing errors themselves can alter the representation of the
+community. Therefore, it is important to minimize the number of
+sequencing errors. Using a widely-used approach that generates the
+lowest reported error rate, we quantified the error rate by sequencing
+the V4 region of the 16S rRNA genes from an 8 member mock community. We
+also removed any contigs that were at least three bases more similar to
+a chimera of two references than to a single reference sequence
+{[}ref{]}. Regardless of the polymerase, the error rate increased with
+the number of rounds of amplification. Using 30 rounds of PCR is a
+common approach across diverse types of samples. Among the data
+generated using 30 rounds of PCR the Accuprime polymerase had the
+highest error rate (i.e.~0.XXX\%) followed by the Platinum
+(i.e.~0.XXX\%), Phusion (i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5
+(i.e.~0.XXX\%) polymerases (Figure 1). When we applied the
+pre.clustering denoising step, which merges the counts of reads within 2
+nt of a more abundant sequence, the error rates dropped considerably
+such that the Platinum polymerase had the highest error rate
+(i.e.~0.XXX\%) followed by the Accuprime (i.e.~0.XXX\%), Phusion
+(i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5 (i.e.~0.XXX\%) polymerases
+(Figure 1). Although specific recommendations are difficult to make
+because the impact of the actual community structure and concentration
+of the initial DNA template are likely to have an impact on the results,
+it is clear that using as few PCR cycles as necessary and a polymerase
+with the lowest possible error rate is a good guide to minimizing the
+impact of polymerase on the error rate.
 
-The median error rate varied by polymerase, with Kappa having the
-highest error rate of all the polymerases across the number of cycles
-{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
-observed across the number of cycles was between Kappa and other
-polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
-contained at least one error was also polymerase dependent, with the
-majority of differences being between Kappa or Accuprime and the other
-polymerases {[}Table S2 \& S3{]}. These differences in error rates were
-not due to polymerase dependent differences in base substitution rate
-{[}Figure S1{]}. Collectively, the results suggest that sequence error
-is dependent on polymerase, persists across the number of cycles used,
-and are not due to any bias towards a specific type of base
-substitution.
+\textbf{\emph{The fraction of sequences identified as being chimeric
+varies by the number of cycles and the DNA polymerase used in PCR.}}
+Chimeric PCR products can significantly confound downstream analyses.
+Although numerous bioinformatic tools exist to identify and remove
+chimeric sequences with high specificity, their sensitivity is
+relatively low and can be reduced by the presence of sequencing errors.
+We identified those contigs that were at least three bases more similar
+to a chimera of two references than to a single reference sequence. As
+expected from previous studies, the number of chimeras increased with
+rounds of amplification. Interestingly, the fraction of chimeric
+sequences from the mock community varied by the type of polymerase used.
+After 30 rounds of PCR, the Platinum polymerase had the highest chimera
+rate (i.e.~XX\%) followed by the Q5 (i.e.~XX\%), Phusion (i.e.~XX\%),
+Kappa (i.e.~XX\%), and Accuprime (i.e.~XX\%) polymerases. We used the
+UChime algorithm to detect chimeras in our pre-clustered mock community
+data and calculate the algorithm's sensitivity and specificity. For all
+polymerases, the specificity was above XX\% and showed a weak
+association with the number of cycles used. There was considerable
+inter-polymerase and inter-round of amplification variation in the
+sensitivity of UChime to detect the chimeras from the mock community.
+This suggested that the residual error rate after pre-clustering the
+sequence data did not compromise the sensitivity. In general, the
+sensitivity of UChime varied between XX and XX\%. The generalizability
+of these results is limited because we used a single mock community with
+limited genetic diversity. Although we did not know the true chimera
+rate for our four human stool samples, we were able to calculate the
+fraction of sequences that UChime identified as being chimeric. These
+results followed those from the mock communities; additional rounds of
+amplification significantly increased the rate of chimeras and there was
+variation between the polymerases that we used. Although it was not
+possible to identify the features of a polymerase that resulted in
+higher rates of chimeras, it is clear that using the smallest number of
+PCR cycles possible will minimize the impacts of chimeras.
 
-We next examined whether chimeras also were dependent on polymerase and
-whether this could affect the total number of OTUs. We observed
-significant differences in the chimera prevalence based on polymerase
-across the number of cycles used (P-value \textless{} 0.05) {[}Table
-S2{]}. Differences in chimera prevalence between Platinum and all other
-polymerases accounted for the majority of these differences {[}Table
-S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
-of all polymerases regardless of whether chimera removal with VSEARCH
-was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
-increased chimera prevalence for all polymerases but the rate of
-increase differed {[}Figure 2B \& 2C{]}. There was also a plateau in the
-total percent of chimeras that were removed that was similar for all
-polymerases {[}Figure 2D{]}. Additionally, a positive correlation was
-observed between chimeric sequences and the number of OTUs for all
-polymerases {[}Figure 2E{]}. This positive correlation was strongest for
-Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
-This data suggests that chimera prevalence depends on polymerase, is
-made worse by increasing the number of cycles, and confirms previous
-reports that the number of OTUs is dependent on the prevalence of these
-chimeric sequences.
+\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
+Since researchers began using PCR to sequence 16S rRNA genes there has
+been concern that amplifying fragments from a mixed template pool could
+lead to a biased representation in the pool of products and would
+confound downstream analyses. The mock community was generated by mixing
+equal amounts of genomic DNA from each of the 8 bacteria resulting in
+uneven representation of the \emph{rrn} operons across the bacteria. The
+vendor of the mock community subjects each lot of genomic DNA to shotgun
+sequencing to more accurately quantify the actual abundance of each
+organism in the community. We compared the expected relative abundance
+of the 16S rRNA genes from each bacterium in the mock community to the
+data we generated across rounds of amplification and polymerase.
+Interestingly, for some bacteria, their representation became less
+biased with additional rounds of PCR (e.g. \emph{L. fermentum}), while
+others became more biased (e.g. \emph{E. faecalis}), and others had
+little change (e.g. \emph{B. subtilis}). The percentage of bases in the
+V4 region that were guanines or cytosines was not predictive of the
+amount of bias. Across the strains there was no variation in the length
+of their V4 regions and they each had the same sequence in the region
+that the primers annealed. One of the bacteria represented in the mock
+community, \emph{S. enterica}, had 6 identical copies of the V4 region
+and 1 copy that differed from those by one nucleotide. The dominant copy
+had a thymidine and the rare copy had a guanine. We used the sequence
+data to calculate the ratio of the dominant to rare variants from
+\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
+Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
+Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
+PCR. Given the subtle nature of the variation in the relative abundances
+of each 16S rRNA gene fragment, it was not possible to create
+generalizable rules that would explain the bias and the previous
+observation regarding the guanine and cytosine relative abundance was
+not observed with our data.
 
-\textbf{\emph{Bacterial community composition differences based on
-number of cycles or polymerase used are less obvious in mock community
-samples.}} Due to the close relationship between chimera prevalence and
-the number of OTUs, we also observed a polymerase dependent difference
-in the range of the number of OTUs in the mock community samples based
-on number of cycles and polymerase {[}Figure 3A{]}. The closest any of
-the polymerases came to a total of 8 OTUs, created by running the mock
-reference 16S sequences through mothur processing, was at 25 and 30
-cycles {[}Figure 3A{]}. Differences between polymerases for the number
-of OTUs created were observed as early as 20 cycles in the mock
-community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
-Tukey post-hoc test, the majority of differences for the number of OTUs
-detected in the mock community was largely due to Kappa and Platinum
-versus the other polymerases across the number of cycles used {[}Table
-S5{]}.
+\textbf{\emph{At the community level, the effects of PCR amplification
+bias grow with additional rounds of PCR.}} Because the distance between
+samples and across rounds of PCR could be artificially inflated due to
+sequencing errors and chimeras, we used analyzed the alpha and beta
+diversity of the mock community data at different phases of the sequence
+curation pipeline. First, we removed the chimeras from the mock
+community data as described above and mapped the individual reads to the
+OTUs that the 16S rRNA gene fragments would cluster into if there were
+no sequencing errors. This gave us a community distribution that
+reflected the distribution following PCR without any artifacts. Alpha
+diversity metrics such as richness and Shannon and Inverse Simpson
+diversity metrics increased with the number of rounds of PCR for all
+polymerases except for the Kappa polymerase, where the richness remained
+constant but the diversity decreased. These data suggest that PCR had
+the effect of making the community distribution more even than it was
+originally. Next, we used the true sequence diversity, but removed
+chimeras based on mapping reads to all possible chimeras between
+reference sequences, and clustered the reads to OTUs. The same trends
+were observed as with perfect sequence data except that the richness and
+diversity metrics trended higher. Finally, we used the observed sequence
+data and used the UCHIME algorithm to identify chimeras. Again, we
+observed the same trends as with the other analyses except the richness
+and diversity metrics trended higher. These comparisons demonstrated
+that although the bias at the sequence level was subtle, PCR does have a
+bias that is exacerbated by errors and chimeras. When we measured the
+Bray-Curtis distance between the communities observed after 25 rounds of
+amplification and those at 30 and 35 rounds of amplification the
+distances between 25 and 35 rounds were higher than between 25 and 30
+rounds for each of the polymerases by an average of 0.XX units. The
+Platinum polymerase varied the most across rounds of amplification (25
+vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). Although the distances
+between samples were small, the ordination of these distances showed a
+clear change in community structure with increasing rounds of PCR. This
+observation was supported by our statistical analysis, which revealed
+that the number of rounds of PCR (R\textsuperscript{2}=0.XX, P=0.XXXX)
+was a larger factor than the choice of polymerase
+(R\textsuperscript{2}=0.XX, P=0.XXXX). These results demonstrate that
+subtle differences in relative abundances can have an impact on overall
+community structure. This variation underscores the importance of only
+comparing sequence data that have been generated using the same PCR
+conditions.
 
-We next investigated the effect polymerases and number of cycles had on
-the bacterial community composition in more detail by using the
-Bray-Curtis index. We observed clear differences in the mock community
-based on polymerase used (P-value = 1e-04) as well as differences based
-on polymerase and number of cycles (P-value = 1e-04). However, when
-investigating specific cycles within each polymerase we observed no
-difference in the bacterial community composition (P-value
-\textgreater{} 0.05) {[}Figure 3B{]}. Finding specific differences in
-overall bacterial community composition based on number of cycles and
-polymerase, we next examined if specific taxa vary depending on
-polymerase and number of cycles. We found that OTUs that taxonomically
-classified to \emph{Salmonella}, \emph{Escherichia},
-\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
-polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
-members of the mock community and the differences were most pronounced
-at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
-between taxa were due to differences between a single polymerase (Kappa)
-and the other polymerases tested {[}Table S7{]}.
+\textbf{\emph{The choice of polymerase or the number of rounds of
+amplification have little impact on the relative interpretation of
+community-wide metrics of diversity.}} The biases that we observed at
+the population and community levels using mock community data appeared
+to be small relative to the expected differences between biological
+samples. To study this further, we calculated alpha and beta-diversity
+metrics using the human stool samples for each of the polymerases and
+rounds of amplification. We calculated the number of observed OTUs,
+Shannon diversity, inverse Simpson diversity index for each condition
+and donor. Although there were clear differences between conditions, the
+relative ordering of the stool samples did not meaningfully vary across
+conditions. When we characterized the variation between rounds of
+amplification using human stool samples, the distance between the 25 and
+30 rounds and 25 and 35 rounds varied considerably between samples and
+polymerases. In general the inter-round variation was lowest for the
+data generated using the Kappa and Accuprime polymerases. The data
+generated using the Platinum polymerase was consistent across rounds,
+but it was more biased than these polymerases. Considering the average
+distance across the four samples varied between 0.XX and 0.XX,
+regardless of the polymerases and number of rounds of amplification, any
+bias due to amplification is unlikely to obscure community-wide
+differences between samples. In support of this was our principle
+coordinates analysis of the Bray-Curtis distances, which revealed
+distinct clusters by donor. Within each cluster there were no obvious
+patterns related to the polymerase or number of rounds of PCR. However,
+our statistical analysis using adonis revealed statistically significant
+differences in the community structures with the subject explaining the
+most variation (R\textsuperscript{2}=0.XXX, P=0.XXXX), followed by the
+number of rounds of PCR (R\textsuperscript{2}=0.XXX, P=0.XXXX) and the
+choice of polymerase (R\textsuperscript{2}=0.XXX, P=0.XXXX). Together,
+these results indicate that for a coarse analysis of communities, the
+choice of number of rounds of amplification or polymerase are not
+important, but that they must be consistent across samples. It is
+difficult to develop a specific recommendation based on the level of
+bias across rounds of PCR or polymerases; however, the general
+suggestion is to use as few rounds of amplification as possible.
 
-Based on these observations, we created Random Forest models to classify
-samples based on number of cycles or polymerase used. When assessing the
-Random Forest models there was a difference between models built to
-classify number of cycles versus those built to classify polymerase used
-{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
-was overly successful at correctly classifying samples based on 16S rRNA
-gene sequencing data (cycles model, probability of correct
-classification = 0.42 (0.4 - 0.43)(min - max) versus polymerase model,
-probability of correct classification = 0.39 (0.37 - 0.4)). Our
-observations in mock communities suggest that although differences are
-detected based on number of cycles or polymerase used, the overall
-community-wide differences are not as large.
-
-\textbf{\emph{Few differences based on number of cycles and polymerase
-used were identified in fecal samples.}} Unlike mock community samples,
-an overall difference in the number of OTUs was only detected between
-polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
-P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
-identify any differences between specific polymerases used (P-value
-\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
-of the number of OTUs detected for fecal samples was dependent on the
-polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
-at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
-lower number of cycles (15-20) to result in a reduced range in the
-number of OTUs versus higher number of cycles (25, 30, and 35) for all
-polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
-cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
-are small differences in the total number of OTUs detected in fecal
-samples that are based on number of cycles or polymerases used.
-
-When using the Bray-Curtis index, we observed that polymerases and
-number of cycles had no affect on the bacterial community composition in
-fecal samples (P-value = 1). There was also no interaction between
-individuals, number of cycles, and polymerase used (P-value = 1). When
-using PERMANOVA to test for community differences based on any of the
-number of cycles within polymerases, only Phusion had cycle dependent
-differences (P-value = 0.03). Additionally, Phusion was one of two
-polymerases that had enough sequences to be rarefied to 1000 sequences
-at 15 cycles. Within each respective 5-cycle increment comparison there
-was no difference in Bray-Curtis index between the polymerases (P-value
-\textgreater{} 0.05) {[}Figure 4B{]}. However, there was an overall
-decrease in Bray-Curtis index by cycle comparison group (e.g.~the 15
-cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle
-group) (P-value \textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's
-post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle
-comparison groups had a higher Bray-Curtis index than the 30 cycle
-versus 35 cycle comparison group (P-value \textless{} 0.05). In addition
-to assessing community composition by Bray-Curtis index, we also
-investigated differences in OTUs based on polymerase at each number of
-cycle used and no significant differences could be found (P-value
-\(\geqslant\) 0.01, corrected P-value = 1). Overall, these results
-suggest that in fecal samples the number of cycles and polymerase used
-minimally change the bacterial community composition.
-
-Finally, we built Random Forest models to classify samples for the
-number of cycles, polymerase used, or individual it was obtained from.
-Using 16S rRNA gene sequencing data there was a large disparity in the
-model's ability to correctly classify samples based on polymerase used,
-number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
-was a clear difference in the success of the model used to classify
-individuals (probability of correct classification = 0.87 (0.86 - 0.9))
-versus the models created to classify number of cycles or polymerase
-(cycles model, probability of correct classification = 0.26 (0.25 -
-0.27)(min - max) and polymerase model, probability of correct
-classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
-important OTUs to the polymerase and number of cycle models were not the
-most important OTUs to the model that classified individuals {[}Figure
-5{]}. This suggests that OTUs that change the most due to the number of
-cycles and polymerase in fecal samples are not the same as OTUs that are
-part of the biological variation between individuals {[}Figure 5{]}.
-Collectively, our observations show that despite a variety of errors
-associated with the amplification process, it is still possible to
-detect the bacterial community variation between individuals.
+\textbf{\emph{There is little evidence of a relationship between
+polymerase or number of rounds of amplification on PCR drift.}} There
+have been concerns that the same template DNA subjected to the same PCR
+conditions could result in different representations of communities
+because of random drift. To test this, we determined the average
+Bray-Curtis distance between replicate reactions using the same
+polymerase and number of rounds of amplification. Using the mock
+community data there were no obvious trends. The average Bray-Curtis
+distance within a set of conditions varied by 0.XX to 0.XX units.
+Although we did not obtain replicates of each of the stool samples, the
+intra-sample variation for each set of conditions was consistent and
+varied between 0.XX and 0.XX units. These data suggest that amplicon
+sequencing is robust to random variation in amplification and that any
+differences are likely to be smaller than what is considered
+biologically relevant.
 
@@
 
-Previous studies have reported GC dependent amplification differences
-(Polz \& Cavanaugh, 1998) and we confirm that these differences still
-persist in high fidelity DNA polymerase when analyzing mock community
-samples {[}Figure 1{]}. Additionally, these specific polymerases have
-different sequence error rates and chimera prevalence {[}Figure 2{]}.
-Chimera prevalence in particular was strongly associated with the total
-number of OTUs detected in our mock community and vaired by polymerase
-{[}Figure 2E \& 3A{]}. Additionally, the bacterial community composition
-varied by number of cycles and polymerase in these mock community
-samples. Clearly, it is possible to identify a large number of
-differences based on number of cycles and polymerase used when analyzing
-mock community samples. However, these differences may not be a major
-concern when analyzing more complex community samples.
+Our results suggest that the number of rounds of PCR and to a lesser
+degree the choice of DNA polymerase impact the analysis of 16S rRNA gene
+sequence data from bacterial communities. Although it was not possible
+to make direct connections between PCR conditions and specific sources
+of bias, we were able to identify general recommendations that reduce
+the amount of error, chimera formation, and bias. Researchers should
+strive to minimize the number of rounds of PCR and should use a high
+fidelity polymerase. Although specific PCR conditions impact the precise
+interpretation of the data, the effects were consistent and were smaller
+than the biological differences between individuals. Based on these
+observations, within a study, amplicons must be generated by consistent
+protocols to yield meaningful comparisons. When comparing across
+studies, values like richness, diversity, and relative abundances must
+be made in relative and not absolute terms.
 
-Fecal samples showed a similar increase in median and range for the
-number of OTUs, based on number of cycles and polymerase used, as that
-found when analyzing mock community samples {[}Figure 4A{]}. However,
-there were few differences between the polymerases within each of the
-amplification cycles {[}Figure 4A{]}. Additionally, these differences
-did not translate to large bacterial community composition changes based
-on number of cycles or polymerase. The differences that were identifed
-were small and occured based on the number of cycles, independent of the
-polymerase used {[}Figure 4B{]}. There was also no OTU-based differences
-observed within the number of cycles between any of the polymerases for
-fecal samples. Interestingly, Random Forest models built on this OTU
-data had the highest classification success for individuals versus
-number of cycles or polymerase. In addition to the better model
-performance of Random Forest for individuals, the most important OTUs to
-this model were not the most important OTUs for the other models tested
-{[}Figure 5{]}. Collectively, these observations show that differences
-based on the number of cycles and between high fidelity DNA polymerases
-can be found. However, when using 16S rRNA gene sequencing these
-technical differences can be smaller than the biological differences
-between fecal bacterial communities within individuals.
+The observed sequencing error rates and alpha diversity metrics followed
+the manufacturers' measurements of their polymerases' fidelity.
+Accuprime and Platinum have fidelity that are approximately 10-times
+higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
+are more than 100 times higher. Among these polymerases, the Kappa
+polymerase resulted in the the lowest error rate, lowest chimera rate,
+and least bias across rounds of PCR. Considering polymerase development
+is an active area of commercial development and it likely that new
+polymerases will come to market, it is important for researchers to
+understand how changing the polymerase impacts downstream analyses for
+their type of samples.
 
-The small differences observed due to error rate and chimera prevalence
-may be due to the actual DNA polymerase family being used within the PCR
-mix. DNA polymerases from distinct families have different binding
-affinities and error correction capacity (Ishino \& Ishino, 2014).
-However, based on the observations within our study this is unlikely to
-be the only contributor. Within our study the highest and lowest chimera
-prevalence both belonged to a family A polymerase (Platinum and
-Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
-Additionally, based on the information supplied by the respective
-manufacturers, the differences between the two PCR mixtures are not
-immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
-DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase, and a platinum
-\emph{Taq} antibody. Since it is not possible to know everything about
-the mixture beyond what was willingly provided by the manufacturer, it
-is possible that differences in how the recombinant \emph{Taq} was
-generated or other compounds within the PCR mixture could be a
-contributing factor for the differences in error rate and chimera
-prevalence. Beyond the choice of the type of polymerase, there may be
-other ways to reduce the affect of polymerase dependent error rates and
-chimera prevalence on the downstream results.
+Over the past 20 years, a large literature has attempted to document
+various PCR biases and underscores the fact that data based on
+amplification of DNA from a mixed community are not a true
+representation of the actual community. In addition to obvious biases
+imposed by primer selection, other factors inherent in PCR can influence
+the representation of communities. Factors that can lead to preferential
+amplification of one fragment over another have included guanine and
+cytosine composition, length, flanking DNA composition, amount of DNA
+shearing, and number of rounds of PCR (24, 27--33). In addition,
+environmental and reagent contaminants can also have a significant
+impact on the analysis of low biomass samples (38). Less well understood
+is the effect of phylogenetic diversity on bias and chimera formation.
+Communities with low phylogenetic diversity may be more prone to chimera
+formation since chimeras are more likely to form among closely related
+sequences (14, 35). The interaction of these various influences on PCR
+artifacts are complex and difficult to tease apart. Minimizing the level
+of DNA shearing and using the fewest number of rounds of PCR with a
+polymerase that has the highest possible fidelity are strategies that
+can be employed to minimize the formation of chimeras. Although care
+should always be taken when choosing a polymerase for 16S rRNA gene
+sequencing, our observations show that the differences between a variety
+of polymerases are smaller than the actual biological variation in fecal
+communities between individuals. Therefore, if the biological signal of
+interest is similar to differences in fecal bacterial communities found
+between individuals, then the type of high fidelity DNA polymerase used
+will only minimally change the results.
 
-Based on polymerase used, our observations generally found no difference
-when using the Bray-Curtis index to measure the bacterial community.
-Specifically, there was no difference in distance between successive
-5-cycle increments within samples between the polymerases when using the
-Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
-is that many of the OTUs generated by polymerase dependent error rates
-and chimera prevalence are likely not highly abundant, allowing the
-Bray-Curtis index to be able to successfully down-weight these OTUs
-(Minchin, 1987). The choice of downstream diversity metric could be an
-important consideration in helping to mitigate some of the observed
-polymerase dependent differences in error rate and chimera prevalence.
-Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
-Vargas, 1996), richness) may be less robust to polymerase dependent
-error rates and chimera prevalence. When choosing a distance metric,
-careful consideration of the errors introduced from the PCR step of the
-16S rRNA gene sequencing workflow need to be taken into account.
-
-Collectively, our data shows that using distinct high fidelity DNA
-polymerases will result in a different number of OTUs being detected.
-The differences in the number of OTUs are primarily due to chimera
-prevalence and sequence error rates that are distinct to the specific
-polymerase. Despite this variation in the number of OTUs, our Random
-Forest models were able to correctly classify individuals regardless of
-the number of cycles and polymerase used. These results suggest that
-even though there is a large variation in the high fidelity DNA
-polymerases used across studies, biological variation similar in effect
-to fecal bacterial community composition differences between individuals
-will still be consistent.
-
-\newpage
-
-\subsection{Conclusion}\label{conclusion}
-
-Although care should always be taken when choosing a polymerase for 16S
-rRNA gene sequencing, our observations show that the differences between
-a variety of polymerases are smaller than the actual biological
-variation in fecal communities between individuals. Therefore, if the
-biological signal of interest is similar to differences in fecal
-bacterial communities found between individuals, then the type of high
-fidelity DNA polymerase used will only minimally change the results.
+Even with these strategies it is impossible to remove all PCR artifacts.
+Beyond the imperfections of the best polymerases, sometimes difficult to
+lyse organisms require stringent lysis steps and low biomass samples
+require additional rounds of PCR. A host of bioinformatics tools are
+available for removing residual sequencing errors (18, 39--41). These
+tools struggle to correctly differentiate true biological diversity
+(i.e.~Amplicon Sequencing Variants or ASVs) and PCR or sequencing
+errors. Although these methods remove residual errors, they also risk
+splitting intragenomic variants into separate ASVs, merging 16S rRNA
+gene sequences from different taxa into the same ASV. Other tools are
+available for removing chimeras (14, 35) where there is a trade off
+between the sensitivity of detecting chimeras and the specificity of
+correctly calling a sequence a chimera. In recent years, parameters for
+these algorithms have been changed to increase their sensitivity with
+little evaluation of the effects on the specificity of the algorithms
+(39, 41). Others recommend removing any read that has an abundance below
+a specified threshold (e.g.~removing all sequences that only appear
+once) (20, 39--41). This method must be approached with caution as such
+approaches are likely to introduce a different bias of the community
+representation and ignore the fact that artifacts may be quite abundant.
+Ultimately, researchers must test their hypotheses with multiple methods
+to validate the claims they reach with any one method (42). All methods
+have biases and limitations and we must use complementary methods to
+obtain robust results.
 
@@
 
-\textbf{\emph{Mock community and human samples.}} The
-ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
-CA, USA) was used for mock communities and the bacterial component was
-made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
-\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
-\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
-\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
-genomic DNA abundance
-(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
-The theoretical abundance of 0.12 used as a comparison point for our
-relative abundance was based on this equal genomic DNA composition of
-the mock community. Fecal samples were obtained from 4 individuals who
-were part of the Enterics Research Investigational Network (ERIN). The
-processing and storage of these samples were previously published
-(Seekatz et al., 2016). Other than confirmation that none of these
-individuals had a \emph{Clostridium difficle} infection, clinical data
-and other types of meta data were not utilized or accessed for this
-study. All samples were extracted using the MOBIO\textsuperscript{TM}
-PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).
+\textbf{\emph{Mock community.}} The ZymoBIOMICS\textsuperscript{TM}
+Microbial Community DNA Standard (Zymo, CA, USA) was used for mock
+communities and the bacterial component was made up of \emph{Pseudomonas
+aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
+\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
+\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
+\emph{Bacillus subtilis} at equal genomic DNA abundance
+(\url{https://web.archive.org/web/20171217151108/http://www.zymoresearch.com:80/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
+The actual relative abundance for each bacterium was obtained from
+Zymo's certificate of analysis for the lot (Lot: ZRC187325), which they
+determined using shotgun metagenomic sequencing
+(\url{https://github.com/SchlossLab/Sze_PCRSeqEffects_mSphere_2019/data/references/ZRC187325.pdf}).
 
-\textbf{\emph{PCR protocol.}} The five different high fidelity DNA
-polymerases that were tested included AccuPrime\textsuperscript{TM}
-(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
-(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
-England Biolabs, MA, USA). The polymerases activation time was 2
-minutes, unless a different activation was specified by the
-manufacturer. The annealing and extension time for Platinum and
-Accuprime followed a previously published protocol (Kozich et al., 2013)
-(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
-For Kappa and Q5, the annealing and extension time also were from a
-previously published optimized protocol (Gohl et al., 2016). For
-Phusion, the company defined activation and annealing times were used
-while the extension time followed the Accuprime and Platinum settings.
+\textbf{\emph{Human samples.}} Fecal samples were obtained from 4
+individuals who were part of an earlier study (43). These samples were
+collected using a protocol approved by the University of Michigan
+Institutional Review Board. For this study, the samples were
+de-identified. DNA was extracted from the fecal samples using the
+MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
+(now Qiagen, MD, USA).
 
-The number of cycles in the PCR for mock and fecal samples started at 15
-and increased by 5 up to 35 cycles, with amplicons used at each 5-step
-increase for sequencing. The mock communities had 4 replicates at 15,
-20, 25, and 35 cycles and 10 replicates for 30 cycles for all
-polymerases (total samples n=130). No mock community sample had enough
-PCR product at 15 cycles for adequate 16S rRNA gene sequencing. The PCR
-of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and
-35 cycles for each polymerase (total sample n=100).
+\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
+tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
+KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
+Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
+Manufacturer recommendations were followed except for the annealing and
+extension times, which were selected based on previously published
+protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
+were used with modifications to generate MiSeq amplicon libraries
+{[}(18); \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The
+number of rounds of PCR used for each sample and polymerase 15 and
+increased by 5 rounds up to 35 cycles. Insufficient PCR product was
+generated using 15 rounds and has not been included in our analysis.
 
-\textbf{\emph{Sequence processing.}} The mothur software program was
-used for all sequence processing steps (Schloss et al., 2009). The
-protocol has been previously published (Kozich et al., 2013)
-(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
-from the published protocol were the use of VSEARCH instead of UCHIME
-for chimera detection and the use of the OptiClust algorithm instead of
-average neighbor for OTU generation at 97\% similarity (Edgar et al.,
-2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
-was determined using the `seq.error' command on mock samples to compare
-back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
-coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
-\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
-(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).
+\textbf{\emph{Library generation and sequencing.}} Each PCR condition
+(i.e.~combination of polymerase and number of rounds of PCR) were
+replicated four times for the mock community and one time for each fecal
+sample. Libraries were generated as described previously {[}(18);
+\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
+were sequenced using the Illumina MiSeq sequencing platform to generate
+paired 250-nt reads.
 
-\textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
-samples were rarefied to 1000 sequences for downstream analysis. The
-analysis of the mock community of each polymerase for GC-based
-amplification differences, sequence error rate, number of sequences with
-an error, base substitution, and numbers of chimeras before and after
-chimera removal with VSEARCH was assessed. Additionally, the correlation
-between the number of chimeras and the number of OTUs was also assessed.
-The total number of OTUs, taxa differences, and Bray-Curtis indices were
-analyzed for both the fecal and mock community samples. Finally, Random
-Forest models were created to assess whether classification of
-polymerase, number of cycles, or individual could be performed best
-using 16S rRNA gene sequencing data. For the model used to classify
-individuals, all samples were included from all number of cycles and
-polymerases but only the individual label was used. Both number of
-cycles and polymerase models included all samples but only the number of
-cycles or polymerase label was used for each respective model.
-Additionally, overlap between the most important OTUs to the three
-models was assessed using mean decrease in accuracy (MDA).
+\textbf{\emph{Sequence processing.}} The mothur software program (v
+1.41) was used for all sequence processing steps (44). The protocol has
+been previously published (18)
+(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Briefly, paired reads
+were assembled using mothur's make.contigs command to correct errors
+introduced by sequencing (18). Any assembled contigs that contained an
+ambiguous base call, mapped to the incorrect region of the 16S rRNA
+gene, or appeared to be a contaminant were removed from subsequent
+analyses. Sequences were further denoised using mothur's pre.cluster
+command to merge the counts of sequences that were within 2 nt of a more
+abundant sequence. The VSEARCH implementation of UCHIME was used to
+screen for chimeras (35, 45). At various stages in the sequence
+processing pipeline for the mock community data, the mothur seq.error
+command was used to quantify the sequencing error rate as well as the
+true chimera rate. This command uses the true sequences from the mock
+community to generate all possible chimeras and removes any contigs that
+were at least three bases more similar to a chimera than to a reference
+sequence. The command then counts the number of substitutions,
+insertions, and deletions in the contig relative to the reference
+sequence and reports the error rate without the inclusion of chimeric
+sequences (19). The reference sequences and operon copy number for each
+bacterium were obtained from the ZymoBIOMICS\textsuperscript{TM}
+Microbial Community DNA Standard protocol
+(\url{https://web.archive.org/web/20181221151905/https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf}).
+Sequences were assigned to operational taxonomic units (OTUs) at a
+threshold of 3\% dissimilarity using the OptiClust algorithm (46). To
+adjust for unequal sequencing when measuring alpha and beta diversity,
+all samples were rarefied to 1,000 sequences 1,000 times for downstream
+analysis. The Good's coverage for the samples at this level of sampling
+was routinely greater than 95\%.
 
 \textbf{\emph{Statistical analysis.}} All analysis was done with the R
-(v 3.4.4) software package (R Core Team, 2017). Data transformation and
-graphing was completed using the tidyverse package (v 1.2.1) and colors
-selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
-2017). High and low GC content was determined based on the median GC
-percentage of either the V4 region or the average GC content of the
-whole genome of the bacterial species used in the mock community.
-Differences in the total number of OTUs were analyzed using an ANOVA
-with a tukey post-hoc test. For the comparison of the number of OTUS in
-fecal samples the data was normalized to each individual by cycle number
-to account for the biological variation. Bray-Curtis distance matrices
-were generated using mothur. The distance matrix data was analyzed using
-PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
-Kruskal-Wallis tests within R. The Random Forest models were run using
-the caret package (v 6.0.78) (Jed Wing et al., 2017). A total of 100
-10-fold CV runs on different 80/20 splits of the data were run to
-generate a range of the Logloss value. The probability of a correct call
-was obtained from this Logloss value by taking the negative natural
-logarithm. For both error and chimera analysis, samples were tested
-using Kruskal-Wallis with a Dunns post-hoc test. Where applicable,
-correction for multiple comparison utilized the Benjamini-Hochberg
-method (Benjamini \& Hochberg, 1995).
+(v 3.5.0) software package (47). Data transformation and graphing were
+completed using the tidyverse package (v 1.2.1) (48). The distance
+matrix data was analyzed using the adonis function within the vegan
+package (v 2.5.2) (49).
 
-\textbf{\emph{Reproducible methods.}} The code and analysis for this
-study can be found here
-\url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}. The raw
-sequences can be found on the SRA at the following accession number
-SRP132931.
+\textbf{\emph{Reproducible methods.}} The data analysis code for this
+study can be found at
+\url{https://github.com/SchlossLab/Sze_PCRSeqEffects_mSphere_2019}. The
+raw sequences are available at the SRA (Accession SRP132931).
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Acinas2005}{}
-Acinas SG., Sarma-Rupavtarm R., Klepac-Ceraj V., Polz MF. 2005.
-PCR-induced sequence artifacts and bias: Insights from comparison of two
-16S rRNA clone libraries constructed from the same sample. \emph{Applied
-and Environmental Microbiology} 71:8966--8969. DOI:
-\href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
+\hypertarget{ref-Gilbert2018}{}
+1. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2018.
+Earth microbiome project and global systems biology. mSystems
+\textbf{3}.
+doi:\href{https://doi.org/10.1128/msystems.00217-17}{10.1128/msystems.00217-17}.
 
-\hypertarget{ref-storage_Bassis_2017}{}
-Bassis CM., Nicholas M. Moore., Lolans K., Seekatz AM., Weinstein RA.,
-Young VB., Hayden MK. 2017. Comparison of stool versus rectal swab
-samples and storage conditions on bacterial community profiles.
-\emph{BMC Microbiology} 17. DOI:
-\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
+\hypertarget{ref-HMP2012}{}
+2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
+the healthy human microbiome. Nature \textbf{486}:207--214.
+doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
 
-\hypertarget{ref-Baxter2016}{}
-Baxter NT., Ruffin MT., Rogers MAM., Schloss PD. 2016. Microbiota-based
-model improves the sensitivity of fecal immunochemical test for
-detecting colonic lesions. \emph{Genome Medicine} 8. DOI:
-\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+\hypertarget{ref-Schloss2016}{}
+3. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
+\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
+bacterial census: An update. mBio \textbf{7}.
+doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.
 
-\hypertarget{ref-benjamini_controlling_1995}{}
-Benjamini Y., Hochberg Y. 1995. Controlling the false discovery rate: A
-practical and powerful approach to multiple testing. \emph{Journal of
-the Royal Statistical Society. Series B (Methodological)} 57:289--300.
+\hypertarget{ref-Luo2016}{}
+4. \textbf{Luo T}, \textbf{Srinivasan U}, \textbf{Ramadugu K},
+\textbf{Shedden KA}, \textbf{Neiswanger K}, \textbf{Trumble E},
+\textbf{Li JJ}, \textbf{McNeil DW}, \textbf{Crout RJ}, \textbf{Weyant
+RJ}, \textbf{Marazita ML}, \textbf{Foxman B}. 2016. Effects of specimen
+collection methodologies and storage conditions on the short-term
+stability of oral microbiome taxonomy. Applied and Environmental
+Microbiology \textbf{82}:5519--5529.
+doi:\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
 
-\hypertarget{ref-Bonfili2017}{}
-Bonfili L., Cecarini V., Berardi S., Scarpona S., Suchodolski JS.,
-Nasuti C., Fiorini D., Boarelli MC., Rossi G., Eleuteri AM. 2017.
-Microbiota modulation counteracts alzheimer's disease progression
-influencing neuronal proteolysis and gut hormones plasma levels.
-\emph{Scientific Reports} 7. DOI:
-\href{https://doi.org/10.1038/s41598-017-02587-2}{10.1038/s41598-017-02587-2}.
+\hypertarget{ref-Bassis2017}{}
+5. \textbf{Bassis CM}, \textbf{Nicholas M. Moore}, \textbf{Lolans K},
+\textbf{Seekatz AM}, \textbf{Weinstein RA}, \textbf{Young VB},
+\textbf{Hayden MK}. 2017. Comparison of stool versus rectal swab samples
+and storage conditions on bacterial community profiles. BMC Microbiology
+\textbf{17}.
+doi:\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
 
-\hypertarget{ref-Burkardt2000}{}
-Burkardt H-J. 2000. Standardization and quality control of PCR analyses.
-\emph{Clinical Chemistry and Laboratory Medicine} 38. DOI:
-\href{https://doi.org/10.1515/cclm.2000.014}{10.1515/cclm.2000.014}.
+\hypertarget{ref-Gorzelak2015}{}
+6. \textbf{Gorzelak MA}, \textbf{Gill SK}, \textbf{Tasnim N},
+\textbf{Ahmadi-Vand Z}, \textbf{Jay M}, \textbf{Gibson DL}. 2015.
+Methods for improving human gut microbiome data by reducing variability
+through sample processing and storage of stool. PLOS ONE
+\textbf{10}:e0134802.
+doi:\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.
 
-\hypertarget{ref-rdp_Cole_2013}{}
-Cole JR., Wang Q., Fish JA., Chai B., McGarrell DM., Sun Y., Brown CT.,
-Porras-Alfaro A., Kuske CR., Tiedje JM. 2013. Ribosomal database
-project: Data and tools for high throughput rRNA analysis. \emph{Nucleic
-Acids Research} 42:D633--D642. DOI:
-\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+\hypertarget{ref-Dominianni2014}{}
+7. \textbf{Dominianni C}, \textbf{Wu J}, \textbf{Hayes RB}, \textbf{Ahn
+J}. 2014. Comparison of methods for fecal microbiome biospecimen
+collection. BMC Microbiology \textbf{14}:103.
+doi:\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.
 
-\hypertarget{ref-Eckert1991}{}
-Eckert KA., Kunkel TA. 1991. DNA polymerase fidelity and the polymerase
-chain reaction. \emph{Genome Research} 1:17--24. DOI:
-\href{https://doi.org/10.1101/gr.1.1.17}{10.1101/gr.1.1.17}.
+\hypertarget{ref-BautistadelosSantos2016}{}
+8. \textbf{Santos QMB-d los}, \textbf{Schroeder JL}, \textbf{Blakemore
+O}, \textbf{Moses J}, \textbf{Haffey M}, \textbf{Sloan W}, \textbf{Pinto
+AJ}. 2016. The impact of sampling, PCR, and sequencing replication on
+discerning changes in drinking water bacterial community over diurnal
+time-scales. Water Research \textbf{90}:216--224.
+doi:\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.
 
-\hypertarget{ref-uchime_Edgar_2011}{}
-Edgar RC., Haas BJ., Clemente JC., Quince C., Knight R. 2011. UCHIME
-improves sensitivity and speed of chimera detection.
-\emph{Bioinformatics} 27:2194--2200. DOI:
-\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+\hypertarget{ref-Sinha2015}{}
+9. \textbf{Sinha R}, \textbf{Chen J}, \textbf{Amir A}, \textbf{Vogtmann
+E}, \textbf{Shi J}, \textbf{Inman KS}, \textbf{Flores R},
+\textbf{Sampson J}, \textbf{Knight R}, \textbf{Chia N}. 2015. Collecting
+fecal samples for microbiome analyses in epidemiology studies. Cancer
+Epidemiology Biomarkers \& Prevention \textbf{25}:407--416.
+doi:\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.
 
-\hypertarget{ref-viridis_citation_2017}{}
-Garnier S. 2017. \emph{Viridis: Default color maps from 'matplotlib'}.
+\hypertarget{ref-Amir2017b}{}
+10. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
+\textbf{Debelius J}, \textbf{Morton JT}, \textbf{Hyde E},
+\textbf{Robbins-Pianka A}, \textbf{Knight R}. 2017. Correcting for
+microbial blooms in fecal samples during room-temperature shipping.
+mSystems \textbf{2}:e00199--16.
+doi:\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.
 
-\hypertarget{ref-taq_Gohl_2016}{}
-Gohl DM., Vangay P., Garbe J., MacLean A., Hauge A., Becker A., Gould
-TJ., Clayton JB., Johnson TJ., Hunter R., Knights D., Beckman KB. 2016.
-Systematic improvement of amplicon marker gene methods for increased
-accuracy in microbiome studies. \emph{Nature Biotechnology} 34:942--949.
-DOI: \href{https://doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
+\hypertarget{ref-Lauber2010}{}
+11. \textbf{Lauber CL}, \textbf{Zhou N}, \textbf{Gordon JI},
+\textbf{Knight R}, \textbf{Fierer N}. 2010. Effect of storage conditions
+on the assessment of bacterial community structure in soil and
+human-associated samples. FEMS Microbiology Letters \textbf{307}:80--86.
+doi:\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.
+
+\hypertarget{ref-Song2016}{}
+12. \textbf{Song SJ}, \textbf{Amir A}, \textbf{Metcalf JL},
+\textbf{Amato KR}, \textbf{Xu ZZ}, \textbf{Humphrey G}, \textbf{Knight
+R}. 2016. Preservation methods differ in fecal microbiome stability,
+affecting suitability for field studies. mSystems \textbf{1}:e00021--16.
+doi:\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
+
+\hypertarget{ref-Costea2017}{}
+13. \textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa S},
+\textbf{Pelletier E}, \textbf{Alberti A}, \textbf{Levenez F},
+\textbf{Tramontano M}, \textbf{Driessen M}, \textbf{Hercog R},
+\textbf{Jung F-E}, \textbf{Kultima JR}, \textbf{Hayward MR},
+\textbf{Coelho LP}, \textbf{Allen-Vercoe E}, \textbf{Bertrand L},
+\textbf{Blaut M}, \textbf{Brown JRM}, \textbf{Carton T},
+\textbf{Cools-Portier S}, \textbf{Daigneault M}, \textbf{Derrien M},
+\textbf{Druesne A}, \textbf{Vos WM de}, \textbf{Finlay BB},
+\textbf{Flint HJ}, \textbf{Guarner F}, \textbf{Hattori M},
+\textbf{Heilig H}, \textbf{Luna RA}, \textbf{Hylckama Vlieg J van},
+\textbf{Junick J}, \textbf{Klymiuk I}, \textbf{Langella P},
+\textbf{Chatelier EL}, \textbf{Mai V}, \textbf{Manichanh C},
+\textbf{Martin JC}, \textbf{Mery C}, \textbf{Morita H}, \textbf{O'Toole
+PW}, \textbf{Orvain C}, \textbf{Patil KR}, \textbf{Penders J},
+\textbf{Persson S}, \textbf{Pons N}, \textbf{Popova M}, \textbf{Salonen
+A}, \textbf{Saulnier D}, \textbf{Scott KP}, \textbf{Singh B},
+\textbf{Slezak K}, \textbf{Veiga P}, \textbf{Versalovic J}, \textbf{Zhao
+L}, \textbf{Zoetendal EG}, \textbf{Ehrlich SD}, \textbf{Dore J},
+\textbf{Bork P}. 2017. Towards standards for human fecal sample
+processing in metagenomic studies. Nature Biotechnology.
+doi:\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.
 
 \hypertarget{ref-Haas2011}{}
-Haas BJ., Gevers D., Earl AM., Feldgarden M., Ward DV., Giannoukos G.,
-Ciulla D., Tabbaa D., Highlander SK., Sodergren E., Methe B., DeSantis
-TZ., Petrosino JF., Knight R., and BWB. 2011. Chimeric 16S rRNA sequence
-formation and detection in sanger and 454-pyrosequenced PCR amplicons.
-\emph{Genome Research} 21:494--504. DOI:
-\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
+14. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
+\textbf{Feldgarden M}, \textbf{Ward DV}, \textbf{Giannoukos G},
+\textbf{Ciulla D}, \textbf{Tabbaa D}, \textbf{Highlander SK},
+\textbf{Sodergren E}, \textbf{Methe B}, \textbf{DeSantis TZ},
+\textbf{Petrosino JF}, \textbf{Knight R}, \textbf{and BWB}. 2011.
+Chimeric 16S rRNA sequence formation and detection in sanger and
+454-pyrosequenced PCR amplicons. Genome Research \textbf{21}:494--504.
+doi:\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
 
-\hypertarget{ref-polymerase_Ishino_2014}{}
-Ishino S., Ishino Y. 2014. DNA polymerases as useful reagents for
-biotechnology â the history of developmental research in the field.
-\emph{Frontiers in Microbiology} 5. DOI:
-\href{https://doi.org/10.3389/fmicb.2014.00465}{10.3389/fmicb.2014.00465}.
+\hypertarget{ref-Sinha2017}{}
+15. \textbf{Sinha R}, \textbf{Abu-Ali G}, \textbf{Vogtmann E},
+\textbf{Fodor AA}, \textbf{Ren B}, \textbf{Amir A}, \textbf{Schwager E},
+\textbf{Crabtree J}, \textbf{Ma S}, \textbf{Abnet CC}, \textbf{Knight
+R}, \textbf{White O}, \textbf{Huttenhower C}. 2017. Assessment of
+variation in microbial community amplicon sequencing by the microbiome
+quality control (MBQC) project consortium. Nature Biotechnology.
+doi:\href{https://doi.org/10.1038/nbt.3981}{10.1038/nbt.3981}.
 
-\hypertarget{ref-caret_citation}{}
-Jed Wing MKC from., Weston S., Williams A., Keefer C., Engelhardt A.,
-Cooper T., Mayer Z., Kenkel B., R Core Team., Benesty M., Lescarbeau R.,
-Ziem A., Scrucca L., Tang Y., Candan C., Hunt. T. 2017. \emph{Caret:
-Classification and regression training}.
+\hypertarget{ref-Meisel2016}{}
+16. \textbf{Meisel JS}, \textbf{Hannigan GD}, \textbf{Tyldsley AS},
+\textbf{SanMiguel AJ}, \textbf{Hodkinson BP}, \textbf{Zheng Q},
+\textbf{Grice EA}. 2016. Skin microbiome surveys are strongly influenced
+by experimental design. Journal of Investigative Dermatology
+\textbf{136}:947--956.
+doi:\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.
+
+\hypertarget{ref-Caporaso2010}{}
+17. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
+\textbf{Berg-Lyons D}, \textbf{Lozupone CA}, \textbf{Turnbaugh PJ},
+\textbf{Fierer N}, \textbf{Knight R}. 2010. Global patterns of 16S rRNA
+diversity at a depth of millions of sequences per sample. Proceedings of
+the National Academy of Sciences \textbf{108}:4516--4522.
+doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.
+
+\hypertarget{ref-Kozich2013}{}
+18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-Schloss2011}{}
+19. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+
+\hypertarget{ref-Bokulich2012}{}
+20. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
+\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
+DA}, \textbf{Caporaso JG}. 2012. Quality-filtering vastly improves
+diversity estimates from illumina amplicon sequencing. Nature Methods
+\textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
+
+\hypertarget{ref-Parada2015}{}
+21. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
+Every base matters: Assessing small subunit rRNA primers for marine
+microbiomes with mock communities, time series and global field samples.
+Environmental Microbiology \textbf{18}:1403--1414.
+doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.
+
+\hypertarget{ref-Wang1996}{}
+22. \textbf{Wang GCY}, \textbf{Wang Y}. 1996. The frequency of chimeric
+molecules as a consequence of PCR co-amplification of 16S rRNA genes
+from different bacterial species. Microbiology \textbf{142}:1107--1114.
+doi:\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
+
+\hypertarget{ref-Potapov2017}{}
+23. \textbf{Potapov V}, \textbf{Ong JL}. 2017. Examining sources of
+error in PCR by single-molecule sequencing. PLOS ONE
+\textbf{12}:e0169774.
+doi:\href{https://doi.org/10.1371/journal.pone.0169774}{10.1371/journal.pone.0169774}.
 
 \hypertarget{ref-Kebschull2015}{}
-Kebschull JM., Zador AM. 2015. Sources of PCR-induced distortions in
-high-throughput sequencing data sets. \emph{Nucleic Acids
-Research}:gkv717. DOI:
-\href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
+24. \textbf{Kebschull JM}, \textbf{Zador AM}. 2015. Sources of
+PCR-induced distortions in high-throughput sequencing data sets. Nucleic
+Acids Research gkv717.
+doi:\href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
 
-\hypertarget{ref-review_Kim_2017}{}
-Kim D., Hofstaedter CE., Zhao C., Mattei L., Tanes C., Clarke E., Lauder
-A., Sherrill-Mix S., Chehoud C., Kelsen J., Conrad M., Collman RG.,
-Baldassano R., Bushman FD., Bittinger K. 2017. Optimizing methods and
-dodging pitfalls in microbiome research. \emph{Microbiome} 5. DOI:
-\href{https://doi.org/10.1186/s40168-017-0267-5}{10.1186/s40168-017-0267-5}.
+\hypertarget{ref-McInerney2014}{}
+25. \textbf{McInerney P}, \textbf{Adams P}, \textbf{Hadi MZ}. 2014.
+Error rate comparison during polymerase chain reaction by DNA
+polymerase. Molecular Biology International \textbf{2014}:1--8.
+doi:\href{https://doi.org/10.1155/2014/287430}{10.1155/2014/287430}.
 
-\hypertarget{ref-protocol_Kozich_2013}{}
-Kozich JJ., Westcott SL., Baxter NT., Highlander SK., Schloss PD. 2013.
-Development of a dual-index sequencing strategy and curation pipeline
-for analyzing amplicon sequence data on the MiSeq illumina sequencing
-platform. \emph{Applied and Environmental Microbiology} 79:5112--5120.
-DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+\hypertarget{ref-Cline1996}{}
+26. \textbf{Cline J}. 1996. PCR fidelity of pfu DNA polymerase and other
+thermostable DNA polymerases. Nucleic Acids Research
+\textbf{24}:3546--3551.
+doi:\href{https://doi.org/10.1093/nar/24.18.3546}{10.1093/nar/24.18.3546}.
 
-\hypertarget{ref-Meisel2016}{}
-Meisel JS., Hannigan GD., Tyldsley AS., SanMiguel AJ., Hodkinson BP.,
-Zheng Q., Grice EA. 2016. Skin microbiome surveys are strongly
-influenced by experimental design. \emph{Journal of Investigative
-Dermatology} 136:947--956. DOI:
-\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.
+\hypertarget{ref-Acinas2005}{}
+27. \textbf{Acinas SG}, \textbf{Sarma-Rupavtarm R}, \textbf{Klepac-Ceraj
+V}, \textbf{Polz MF}. 2005. PCR-induced sequence artifacts and bias:
+Insights from comparison of two 16S rRNA clone libraries constructed
+from the same sample. Applied and Environmental Microbiology
+\textbf{71}:8966--8969.
+doi:\href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
 
-\hypertarget{ref-bc_index_Minchin1987}{}
-Minchin PR. 1987. An evaluation of the relative robustness of techniques
-for ecological ordination. \emph{Vegetatio} 69:89--107. DOI:
-\href{https://doi.org/10.1007/bf00038690}{10.1007/bf00038690}.
+\hypertarget{ref-Polz1998}{}
+28. \textbf{Polz MF}, \textbf{Cavanaugh CM}. 1998. Bias in
+template-to-product ratios in multitemplate PCR. Applied and
+Environmental Microbiology \textbf{64}:3724--3730.
 
-\hypertarget{ref-vegan_citation}{}
-Oksanen J., Blanchet FG., Friendly M., Kindt R., Legendre P., McGlinn
-D., Minchin PR., O'Hara RB., Simpson GL., Solymos P., Stevens MHH.,
-Szoecs E., Wagner H. 2017. \emph{Vegan: Community ecology package}.
+\hypertarget{ref-Brooks2015}{}
+29. \textbf{Brooks JP}, \textbf{David J Edwards}, \textbf{Harwich MD},
+\textbf{Rivera MC}, \textbf{Fettweis JM}, \textbf{Serrano MG},
+\textbf{Reris RA}, \textbf{Sheth NU}, \textbf{Huang B}, \textbf{Girerd
+P}, \textbf{Strauss JF}, \textbf{Jefferson KK}, \textbf{Buck GA}. 2015.
+The truth about metagenomics: Quantifying and counteracting bias in 16S
+rRNA studies. BMC Microbiology \textbf{15}.
+doi:\href{https://doi.org/10.1186/s12866-015-0351-6}{10.1186/s12866-015-0351-6}.
 
-\hypertarget{ref-polz_bias_1998}{}
-Polz MF., Cavanaugh CM. 1998. Bias in template-to-product ratios in
-multitemplate PCR. \emph{Applied and Environmental Microbiology}
-64:3724--3730.
+\hypertarget{ref-Suzuki1996}{}
+30. \textbf{Suzuki MT}, \textbf{Giovannoni SJ}. 1996. Bias caused by
+template annealing in the amplification of mixtures of 16S rRNA genes by
+pcr. Applied and environmental microbiology \textbf{62}:625--630.
 
-\hypertarget{ref-r_citation_2017}{}
-R Core Team. 2017. \emph{R: A language and environment for statistical
-computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\hypertarget{ref-Chandler1997}{}
+31. \textbf{Chandler D}, \textbf{Fredrickson J}, \textbf{Brockman F}.
+1997. Effect of pcr template concentration on the composition and
+distribution of total community 16S rDNA clone libraries. Molecular
+Ecology \textbf{6}:475--482.
 
-\hypertarget{ref-Real1996}{}
-Real R., Vargas JM. 1996. The probabilistic basis of jaccards index of
-similarity. \emph{Systematic Biology} 45:380--385. DOI:
-\href{https://doi.org/10.1093/sysbio/45.3.380}{10.1093/sysbio/45.3.380}.
+\hypertarget{ref-Wagner1994}{}
+32. \textbf{Wagner A}, \textbf{Blackstone N}, \textbf{Cartwright P},
+\textbf{Dick M}, \textbf{Misof B}, \textbf{Snow P}, \textbf{Wagner GP},
+\textbf{Bartels J}, \textbf{Murtha M}, \textbf{Pendleton J}. 1994.
+Surveys of gene families using polymerase chain reaction: PCR selection
+and pcr drift. Systematic Biology \textbf{43}:250--261.
 
-\hypertarget{ref-vsearch_Rognes_2016}{}
-Rognes T., Flouri T., Nichols B., Quince C., Mahé F. 2016. VSEARCH: A
-versatile open source tool for metagenomics. \emph{PeerJ} 4:e2584. DOI:
-\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+\hypertarget{ref-Hansen1998}{}
+33. \textbf{Hansen MC}, \textbf{Tolker-Nielsen T}, \textbf{Givskov M},
+\textbf{Molin S}. 1998. Biased 16S rDNA pcr amplification caused by
+interference from dna flanking the template region. FEMS Microbiology
+Ecology \textbf{26}:141--149.
 
-\hypertarget{ref-contamination_Salter2014}{}
-Salter SJ., Cox MJ., Turek EM., Calus ST., Cookson WO., Moffatt MF.,
-Turner P., Parkhill J., Loman NJ., Walker AW. 2014. Reagent and
-laboratory contamination can critically impact sequence-based microbiome
-analyses. \emph{BMC Biology} 12. DOI:
-\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
+\hypertarget{ref-Kennedy2014}{}
+34. \textbf{Kennedy K}, \textbf{Hall MW}, \textbf{Lynch MDJ},
+\textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}. 2014. Evaluating bias
+of illumina-based bacterial 16S rRNA gene profiles. Applied and
+Environmental Microbiology \textbf{80}:5717--5722.
+doi:\href{https://doi.org/10.1128/aem.01451-14}{10.1128/aem.01451-14}.
 
-\hypertarget{ref-BautistadelosSantos2016}{}
-Santos QMB-d los., Schroeder JL., Blakemore O., Moses J., Haffey M.,
-Sloan W., Pinto AJ. 2016. The impact of sampling, PCR, and sequencing
-replication on discerning changes in drinking water bacterial community
-over diurnal time-scales. \emph{Water Research} 90:216--224. DOI:
-\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.
+\hypertarget{ref-Edgar2011}{}
+35. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
+and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
 
-\hypertarget{ref-mothur_schloss_2009}{}
-Schloss PD., Westcott SL., Ryabin T., Hall JR., Hartmann M., Hollister
-EB., Lesniewski RA., Oakley BB., Parks DH., Robinson CJ., Sahl JW.,
-Stres B., Thallinger GG., Horn DJV., Weber CF. 2009. Introducing mothur:
-Open-source, platform-independent, community-supported software for
-describing and comparing microbial communities. \emph{Applied and
-Environmental Microbiology} 75:7537--7541. DOI:
-\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+\hypertarget{ref-Edgar2017}{}
+36. \textbf{Edgar RC}. 2017. UNBIAS: An attempt to correct abundance
+bias in 16S sequencing, with limited success.
+doi:\href{https://doi.org/10.1101/124149}{10.1101/124149}.
 
-\hypertarget{ref-erin_seekatz_2016}{}
-Seekatz AM., Rao K., Santhosh K., Young VB. 2016. Dynamics of the fecal
-microbiome in patients with recurrent and nonrecurrent clostridium
-difficile infection. \emph{Genome Medicine} 8. DOI:
-\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+\hypertarget{ref-Gohl2016}{}
+37. \textbf{Gohl DM}, \textbf{Vangay P}, \textbf{Garbe J},
+\textbf{MacLean A}, \textbf{Hauge A}, \textbf{Becker A}, \textbf{Gould
+TJ}, \textbf{Clayton JB}, \textbf{Johnson TJ}, \textbf{Hunter R},
+\textbf{Knights D}, \textbf{Beckman KB}. 2016. Systematic improvement of
+amplicon marker gene methods for increased accuracy in microbiome
+studies. Nature Biotechnology \textbf{34}:942--949.
+doi:\href{https://doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
 
-\hypertarget{ref-preservation_Song_2016}{}
-Song SJ., Amir A., Metcalf JL., Amato KR., Xu ZZ., Humphrey G., Knight
-R. 2016. Preservation methods differ in fecal microbiome stability,
-affecting suitability for field studies. \emph{mSystems} 1:e00021--16.
-DOI:
-\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
+\hypertarget{ref-Salter2014}{}
+38. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
+\textbf{Calus ST}, \textbf{Cookson WO}, \textbf{Moffatt MF},
+\textbf{Turner P}, \textbf{Parkhill J}, \textbf{Loman NJ},
+\textbf{Walker AW}. 2014. Reagent and laboratory contamination can
+critically impact sequence-based microbiome analyses. BMC Biology
+\textbf{12}.
+doi:\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
 
-\hypertarget{ref-Sze2015}{}
-Sze MA., Dimitriu PA., Suzuki M., McDonough JE., Campbell JD., Brothers
-JF., Erb-Downward JR., Huffnagle GB., Hayashi S., Elliott WM., Cooper
-J., Sin DD., Lenburg ME., Spira A., Mohn WW., Hogg JC. 2015. Host
-response to the lung microbiome in chronic obstructive pulmonary
-disease. \emph{American Journal of Respiratory and Critical Care
-Medicine} 192:438--445. DOI:
-\href{https://doi.org/10.1164/rccm.201502-0223oc}{10.1164/rccm.201502-0223oc}.
+\hypertarget{ref-Callahan2016}{}
+39. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
+\textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes SP}. 2016. DADA2:
+High-resolution sample inference from illumina amplicon data. Nature
+Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
 
-\hypertarget{ref-Turnbaugh2008}{}
-Turnbaugh PJ., Hamady M., Yatsunenko T., Cantarel BL., Duncan A., Ley
-RE., Sogin ML., Jones WJ., Roe BA., Affourtit JP., Egholm M., Henrissat
-B., Heath AC., Knight R., Gordon JI. 2008. A core gut microbiome in
-obese and lean twins. \emph{Nature} 457:480--484. DOI:
-\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.
+\hypertarget{ref-Edgar2016}{}
+40. \textbf{Edgar RC}. 2016. UNOISE2: Improved error-correction for
+illumina 16S and ITS amplicon sequencing.
+doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.
 
-\hypertarget{ref-Wang1996}{}
-Wang GCY., Wang Y. 1996. The frequency of chimeric molecules as a
-consequence of PCR co-amplification of 16S rRNA genes from different
-bacterial species. \emph{Microbiology} 142:1107--1114. DOI:
-\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
+\hypertarget{ref-Amir2017a}{}
+41. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
+\textbf{Kopylova E}, \textbf{Morton JT}, \textbf{Xu ZZ},
+\textbf{Kightley EP}, \textbf{Thompson LR}, \textbf{Hyde ER},
+\textbf{Gonzalez A}, \textbf{Knight R}. 2017. Deblur rapidly resolves
+single-nucleotide community sequence patterns. mSystems \textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.
 
-\hypertarget{ref-opticlust_Westcott_2017}{}
-Westcott SL., Schloss PD. 2017. OptiClust, an improved method for
-assigning amplicon-based sequence data to operational taxonomic units.
-\emph{mSphere} 2:e00073--17. DOI:
-\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+\hypertarget{ref-Schloss2018}{}
+42. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
+reproducibility, replicability, robustness, and generalizability in
+microbiome research. mBio \textbf{9}.
+doi:\href{https://doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
 
-\hypertarget{ref-tidyverse_2017}{}
-Wickham H. 2017. \emph{Tidyverse: Easily install and load 'tidyverse'
-packages}.
+\hypertarget{ref-Seekatz2016}{}
+43. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+\textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
+with recurrent and nonrecurrent clostridium difficile infection. Genome
+Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
 
-\hypertarget{ref-Zupancic2012}{}
-Zupancic ML., Cantarel BL., Liu Z., Drabek EF., Ryan KA., Cirimotich S.,
-Jones C., Knight R., Walters WA., Knights D., Mongodin EF., Horenstein
-RB., Mitchell BD., Steinle N., Snitker S., Shuldiner AR., Fraser CM.
-2012. Analysis of the gut microbiota in the old order amish and its
-relation to the metabolic syndrome. \emph{PLoS ONE} 7:e43052. DOI:
-\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
+\hypertarget{ref-Schloss2009}{}
+44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\hypertarget{ref-Rognes2016}{}
+45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-Westcott2017}{}
+46. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+improved method for assigning amplicon-based sequence data to
+operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+
+\hypertarget{ref-r_citation_2018}{}
+47. \textbf{R Core Team}. 2018. R: A language and environment for
+statistical computing. R Foundation for Statistical Computing, Vienna,
+Austria.
+
+\hypertarget{ref-Tidyverse2017}{}
+48. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
+'tidyverse'.
+
+\hypertarget{ref-Vegan2018}{}
+49. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
+\textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
+\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Simpson GL},
+\textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
+\textbf{Wagner H}. 2018. Vegan: Community ecology package.
 
@@
 
-\textbf{Figure 1: Relative abundance differences due to GC is consistent
-across polymerases.} At each number of cycle the points and lines
-represent the median and with the minimum and maximum relative abundance
-respectively. Regardless of the number of cycle used a consistent
-difference between high and low GC bacteria was observed. The dotted
-line represents the actual relative abundance that each bacterium should
-be at within the mock community sample.
-
-\textbf{Figure 2: Error rate and chimera prevalence vary by polymerase
-and affect the number of observed OTUs.} A) The error bars represent the
-75\% interquartile range of the median error rate. B) Percentage of
-chimeric sequences without the removal of chimeras with VSEARCH. C)
-Percentage of chimeric sequences with the removal of chimeras with
-VSEARCH. D) The total percent of chimeric sequences removed with VSEARCH
-by cycle number. The error bars represent the 75\% interquartile range
-of the median. E) Chimera prevalence and the the observed number of OTUs
-are strongly correlated in mock community samples.
-
-\textbf{Figure 3: Subtle differences based on number of cycles and
-polymerase used are detected in bacterial community composition of mock
-samples.} A) The range in the number of OTUs detected in the mock
-samples increased as cycle number increased. This range was larger for
-specific HiFi DNA polymerases. The points represent the median number of
-OTUs and the lines represent the range from the minimum to maximum
-number of OTUs detected within the four technical replicates. The dotted
-black line represents the number of OTUs detected when only the
-references sequences for the mock community are clustered. A) Within
-replicate difference based on the next 5-cycle PCR interval in mock
-samples. The lines represent the range of the minimum and maximum
-Bray-Curtis index value for each PCR 5-cycle increment comparison. The
-closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar
-the bacterial community is of the two compared number of cycles.
-
-\textbf{Figure 4: Subtle differences based on number of cycles and
-polymerase used are detected in bacterial community composition of fecal
-samples.} A) The range in the number of OTUs detected in the different
-fecal samples increased as cycle number increased. This range was larger
-for specific HiFi DNA polymerases. The points represent the median
-number of OTUs and the lines represent the range from the minimum to
-maximum number of OTUs detected within the four fecal samples. B) Within
-person differences based on the next 5-cycle PCR interval in fecal
-samples. The points represent the median Bray-Curtis index for the
-samples. The lines represent the range of the minimum and maximum
-Bray-Curtis index value for each PCR 5-cycle increment comparison. The
-closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar
-the bacterial community is of the two compared number of cycles.
-
-\textbf{Figure 5: Important OTUs for the number of cycles and polymerase
-models are not the most important OTUs in the model classifying
-individuals.} Color highlights the top 10 OTUs in number of cycles and
-polymerase used and where they fall in the model used to classify
-individuals. The majority of these top 10 OTUs have 0 importance to the
-model used to classify individuals.
+\textbf{Figure 1: XXXXXXXX.}
 
@@
 
-\textbf{Figure S1: No preference for specific base subsitutions was
-observed across the polymerases in mock community samples.}
+\textbf{Figure S1: XXXXXXXX.}
 

### 2019-01-14 code/plot_chimera_rate.R
Fix stylizing of figures

@@ -1,8 +1,9 @@
 library(tidyverse)
-library(viridis)
 library(RColorBrewer)
-
 library(cowplot)
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 facet_names <- c(
 									'percentage' = "Total chimeras (%)",
@@ -24,7 +25,7 @@ mock <- read_tsv("data/process/error_chimera_rates.tsv") %>%
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
+												values=polymerase_colors) +
 		theme_classic() +
 		theme(
 			panel.spacing=unit(0.5,"cm"),

### 2019-01-14 code/plot_drift.R

@@ -1,6 +1,9 @@
 library(tidyverse)
 library(cowplot)
-library(viridis)
+library(RColorBrewer)
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 
 labels <- c('mock' = 'Intra-replicate mock community\nBray-Curtis distances',
@@ -32,7 +35,7 @@ bind_rows(mock, stool) %>%
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
+												values=polymerase_colors) +
 		labs(y=NULL, x="Number of rounds of PCR") +
 		theme_classic() +
 		theme(

### 2019-01-14 code/plot_error_rate.R

@@ -1,6 +1,8 @@
 library(tidyverse)
-library(viridis)
-# library(RColorBrewer)
+library(RColorBrewer)
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 facet_names <- c(
 									'contig_error' = "Raw contigs",
@@ -17,7 +19,7 @@ read_tsv("data/process/error_chimera_rates.tsv") %>%
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
+												values=polymerase_colors) +
 		theme_classic() +
 		theme(
 			panel.spacing=unit(0.5,"cm"),

### 2019-01-14 code/plot_species_bias.R

@@ -1,7 +1,10 @@
 library(tidyverse)
-library(viridis)
+library(RColorBrewer)
 
 mock <- read_tsv("data/process/mock_bias_species.tsv")
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 species_order <- mock %>% arrange(percent_gc) %>% pull(species) %>% str_replace("_", " ") %>% unique
 
@@ -13,18 +16,20 @@ mock %>%
 	ggplot(aes(x=rounds, y=obs_rel_abund, group=polymerase, color=polymerase)) +
 		facet_wrap(~species)+#, labeller = label_parsed) +
 		geom_hline(aes(yintercept=actual_rel_abund), col="gray") +
-		geom_line() +
-		geom_text(aes(x=20.75, y=25, label=format(percent_gc, nsmall=1L)), size=3, color="black") +
+		geom_line(size=1) +
+		geom_text(aes(x=20.75, y=0.5, label=format(percent_gc, nsmall=1L)), size=3, color="black") +
 		coord_cartesian(ylim=c(0,25), xlim=c(20,35)) +
 		labs(y="Relative abundance (%)", x="Number of rounds of PCR") +
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
+												values=polymerase_colors) +
 		theme_classic() +
 		theme(strip.text=element_text(face="italic"),
 					strip.text.x=element_text(hjust=0, size=9),
 					strip.background = element_rect(color = 'white', fill = 'white', size = 10),
-					legend.position=c(0.85, 0.1))
+					legend.position=c(0.85, 0.1),
+					legend.key.height = unit(0.8, "line")
+				)
 
 ggsave("results/figures/species_bias.pdf", width=6.875, height=6)

### 2019-01-14 code/plot_drift.R
Fix stylizing of figures

@@ -1,10 +1,13 @@
 library(tidyverse)
 library(cowplot)
-library(viridis)
+library(RColorBrewer)
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 
-labels <- c('mock' = 'Intra-replicate mock community\nBray-Curtis distances',
-						'stool' = 'Inter-stool sample\nBray-Curtis distances')
+labels <- c('mock' = 'Mean intra-replicate Bray-Curtis\ndistances for the mock community',
+						'stool' = 'Mean inter-sample Bray-Curtis\ndistances for the stool samples')
 
 mock <- read_csv("data/process/mock_beta_drift.csv") %>%
 	filter(method == "vsearch") %>%
@@ -32,7 +35,7 @@ bind_rows(mock, stool) %>%
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
+												values=polymerase_colors) +
 		labs(y=NULL, x="Number of rounds of PCR") +
 		theme_classic() +
 		theme(
@@ -42,4 +45,4 @@ bind_rows(mock, stool) %>%
 			legend.key.height = unit(0.8, "line")
 		)
 
-ggsave("results/figures/drift.pdf", width=6, height=3)
+ggsave("results/figures/drift.pdf", width=6.5, height=3)

### 2019-01-14 code/plot_mock_community.R

@@ -58,12 +58,12 @@ rel_distance <- read_tsv("data/process/mock_beta_diversity.tsv") %>%
 	mutate(col_rounds = str_replace(col_rounds, "x", "")) %>%
 	ggplot(aes(x=polymerase, y=distances, fill=col_rounds))+
 		geom_col(position=position_dodge()) +
-		labs(x=NULL, y="Bray-Curtis Distance") +
+		labs(x=NULL, y="Bray-Curtis distance relative\nto 25 rounds of PCR") +
 		scale_x_discrete(labels=polymerase_names) +
 		scale_fill_manual(name=NULL, breaks=c(30, 35), labels=c("30 rounds", "35 rounds"), values=c("blue", "red")) +
 		theme_classic() +
 		theme(
-			legend.position=c(0.22, 0.90),
+			legend.position=c(0.23, 0.90),
 			axis.text.x=element_text(size=8)
 		)
 

### 2019-01-14 code/plot_salmonella_bias.R

@@ -1,16 +1,22 @@
 library(tidyverse)
-library(viridis)
+library(RColorBrewer)
+
+polymerase_colors <- brewer.pal(5, "RdBu")
+polymerase_colors[3] <- "darkgray"
 
 read_tsv("data/process/mock_bias_salmonella.tsv") %>%
 	ggplot(aes(x=rounds, y=ratio, group=polymerase, color=polymerase)) +
 		geom_hline(aes(yintercept=6), col="gray") +
-		geom_line() +
+		geom_line(size=1) +
 		coord_cartesian(ylim=c(4,8)) +
-		labs(y="Ratio of relative abundances for each operon", x="Number of rounds of PCR") +
+		labs(y="Ratio of relative abundances\nfor each operon", x="Number of rounds of PCR") +
 		scale_color_manual(name=NULL,
 												breaks=c("ACC", "K", "PHU", "PL", "Q5"),
 												labels=c("Accuprime", "Kappa", "Phusion", "Platinum", "Q5"),
-												values=viridis(5)) +
-		theme_classic()
+												values=polymerase_colors) +
+		theme_classic() +
+		theme(
+			legend.key.height = unit(0.8, "line")
+		)
 
-ggsave("results/figures/salmonella_bias.pdf", width=6.875, height=6)
+ggsave("results/figures/salmonella_bias.pdf", width=4.4, height=3)

### 2019-01-14 code/plot_stool_community.R

@@ -56,7 +56,7 @@ rel_distance <- read_tsv("data/process/stool_beta_diversity.tsv") %>%
 	ggplot(aes(x=polymerase, y=distances, col=col_rounds, group=paste(subject, polymerase))) +
 		geom_line(position=position_dodge(width=0.5), color="black")+
 		geom_point(position=position_dodge(width=0.5)) +
-		labs(x=NULL, y="Bray-Curtis Distance") +
+		labs(x=NULL, y="Bray-Curtis distance relative\nto 25 rounds of PCR") +
 		scale_x_discrete(labels=polymerase_names) +
 		scale_color_manual(name=NULL, breaks=c(30, 35), labels=c("30 rounds", "35 rounds"), values=c("blue", "red")) +
 		coord_cartesian(ylim=c(0,0.35)) +

### 2019-01-14 submission/manuscript.tex
Add figure legends

@@
-\documentclass[11pt,]{article}
+\documentclass[11,]{article}
 \usepackage{lmodern}
@@
     \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
 \fi
@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 \begin{itemize}
-\tightlist
+\itemsep1pt\parskip0pt\parsep0pt
 \item
-  \href{mailto:marcsze@med.umich.edu}{\nolinkurl{marcsze@med.umich.edu}}
+  \href{mailto:marcsze@med.umich.edu}{marcsze@med.umich.edu}
 \end{itemize}
@@
 \newpage
-
 \linenumbers
@@
 were used with modifications to generate MiSeq amplicon libraries
-{[}(18); \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The
-number of rounds of PCR used for each sample and polymerase 15 and
-increased by 5 rounds up to 35 cycles. Insufficient PCR product was
-generated using 15 rounds and has not been included in our analysis.
+{[}Reference 18;
+\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The number of
+rounds of PCR used for each sample and polymerase 15 and increased by 5
+rounds up to 35 cycles. Insufficient PCR product was generated using 15
+rounds and has not been included in our analysis.
 
@@
 replicated four times for the mock community and one time for each fecal
-sample. Libraries were generated as described previously {[}(18);
-\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
-were sequenced using the Illumina MiSeq sequencing platform to generate
-paired 250-nt reads.
+sample. Libraries were generated as described previously {[}Reference
+18; \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The
+libraries were sequenced using the Illumina MiSeq sequencing platform to
+generate paired 250-nt reads.
 
@@
 matrix data was analyzed using the adonis function within the vegan
-package (v 2.5.2) (49).
+package (v 2.5.3) (49).
 
@@
 
-The authors would like to thank all the study participants in ERIN whose
-samples were utilized. We also would like to thank Judy Opp and April
-Cockburn for their effort in sequencing the samples as part of the
-Microbiome Core Facility at the University of Michigan. Additional
-thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
-earlier drafts of the manuscript and providing helpful critiques. Salary
-support for Marc A. Sze came from the Canadian Institute of Health
-Research and NIH grant UL1TR002240. Salary support for Patrick D.
-Schloss came from NIH grants P30DK034933 and 1R01CA215574.
+The authors thank the study participants in ERIN whose samples were
+utilized. We also would like to thank Judy Opp and April Cockburn for
+their effort in sequencing the samples as part of the Microbiome Core
+Facility at the University of Michigan. Additional thanks to members of
+the Schloss lab and Dr.~Marcy Balunas for reading earlier drafts of the
+manuscript and providing helpful critiques. Salary support for Marc A.
+Sze came from the Canadian Institute of Health Research and NIH grant
+UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants
+P30DK034933 and 1R01CA215574.
+
+\subsection{References}\label{references}
+
+\textbf{Figure 1. The error rate of assembled sequence reads increases
+with the number of rounds of PCR used and follows the relative error
+rates provided by the manufacturers; however, much of this error is
+mediated by denoising.}
+
+\textbf{Figure 2. The fraction of all denoised sequences that were
+identified as being chimeric increases with the number of rounds of PCR
+used and varied between polymerases.} (A) Sequencing of a mock community
+allowed us to identify the total fraction of sequences that were
+chimeric as well as the sensitivity and specificity of UCHIME to detect
+those chimeras. (B) Sequencing of four human stool samples after using
+one of five different polymerases again demonstrated increased rate of
+chimera formation with increasing number of rounds of PCR and variation
+across polymerases.
+
+\textbf{Figure 3. The relative abundances of reads mapped to reference
+sequences differed subtly from the expected relative abundances as
+determined by shotgun metagenomic sequencing.} Bias did not increase
+with number of rounds of PCR or vary by polymerase or the guanine and
+cytosine content of the fragment. The expected relative abundance of
+each organism is indicated by the horizontal gray line. The percentage
+of bases that were guanines or cytosines within the V4 region of the 16S
+rRNA genes in each organism is indicated by the number in the lower left
+corner of each panel.
+
+\textbf{Figure 4. Despite evidence of subtle PCR bias at the genome
+level, there was significant evidence of bias using community-wide
+metrics that grew with the number of rounds of PCR when using a mock
+community.} (A) With the exception of data collected using the Kappa
+polymerase, the OTU richness and Shannon diversity values increased with
+number of rounds of PCR and the inclusion of residual sequencing errors
+and chimeras. The horizontal black line indicates the expected richness
+and diversity if there were no bias, errors, or chimeras. (B) Relative
+to the mock community sampled after 25 rounds of PCR, the Bray-Curtis
+distance to the communities sampled after 30 and 35 rounds of PCR
+increased for all polymerases. (C) The variation between samples
+collected after 20, 25, 30, and 35 rounds of PCR for the five
+polymerases demonstrated a significant change in the community driven by
+the number of rounds of PCR and the polymerase used.
+
+\textbf{Figure 5. Sequencing of human stool samples indicated clear
+increase in bias with number of rounds of PCR, however, the bias
+appeared to be consistent within each sample.} (A) With the exception of
+data collected using the Kappa polymerase, the OTU richness and Shannon
+diversity values increased with number of rounds of PCR. (B) Relative to
+the stool communities sampled after 25 rounds of PCR, the Bray-Curtis
+distance to the stool communities sampled after 30 and 35 rounds of PCR
+was inconsistent and there was little difference in variation for data
+collected using the Kappa polymerase. (C) The variation between stool
+samples was larger than the amount of variation introduced by varying
+the number of rounds of PCR or polymerase.
+
+\textbf{Figure 6. There is little evidence for PCR drift adversely
+impacting the results of amplicon studies.} The average distance between
+replicates of sequencing the same mock community or between the the
+human stool samples did not vary by number of rounds of PCR or by
+polymerase.
 
@@
 
-\subsection{References}\label{references}
+\textbf{Figure S1: With the exception of the seuqence data generated
+using the Kappa polymerase, the ratio of the two \emph{Salmonella
+enterica} V4 sequences to a value was lower than the expected ratio of
+6:1.} The two \emph{S. enterica} V4 sequences differed by a single base.
 
-\hypertarget{refs}{}
-\hypertarget{ref-Gilbert2018}{}
+\newpage
+
 1. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2018.
@@
 \textbf{3}.
-doi:\href{https://doi.org/10.1128/msystems.00217-17}{10.1128/msystems.00217-17}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00217-17}{10.1128/msystems.00217-17}.
 
-\hypertarget{ref-HMP2012}{}
 2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
 the healthy human microbiome. Nature \textbf{486}:207--214.
-doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
+doi:\href{http://dx.doi.org/10.1038/nature11234}{10.1038/nature11234}.
 
-\hypertarget{ref-Schloss2016}{}
 3. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
 \textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
-bacterial census: An update. mBio \textbf{7}.
-doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.
+bacterial census: an update. mBio \textbf{7}.
+doi:\href{http://dx.doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.
 
-\hypertarget{ref-Luo2016}{}
 4. \textbf{Luo T}, \textbf{Srinivasan U}, \textbf{Ramadugu K},
@@
 Microbiology \textbf{82}:5519--5529.
-doi:\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
+doi:\href{http://dx.doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.
 
-\hypertarget{ref-Bassis2017}{}
 5. \textbf{Bassis CM}, \textbf{Nicholas M. Moore}, \textbf{Lolans K},
@@
 \textbf{17}.
-doi:\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
+doi:\href{http://dx.doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.
 
-\hypertarget{ref-Gorzelak2015}{}
 6. \textbf{Gorzelak MA}, \textbf{Gill SK}, \textbf{Tasnim N},
@@
 \textbf{10}:e0134802.
-doi:\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.
+doi:\href{http://dx.doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.
 
-\hypertarget{ref-Dominianni2014}{}
 7. \textbf{Dominianni C}, \textbf{Wu J}, \textbf{Hayes RB}, \textbf{Ahn
@@
 collection. BMC Microbiology \textbf{14}:103.
-doi:\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.
+doi:\href{http://dx.doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.
 
-\hypertarget{ref-BautistadelosSantos2016}{}
 8. \textbf{Santos QMB-d los}, \textbf{Schroeder JL}, \textbf{Blakemore
@@
 time-scales. Water Research \textbf{90}:216--224.
-doi:\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.
+doi:\href{http://dx.doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.
 
-\hypertarget{ref-Sinha2015}{}
 9. \textbf{Sinha R}, \textbf{Chen J}, \textbf{Amir A}, \textbf{Vogtmann
@@
 Epidemiology Biomarkers \& Prevention \textbf{25}:407--416.
-doi:\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.
+doi:\href{http://dx.doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.
 
-\hypertarget{ref-Amir2017b}{}
 10. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
@@
 mSystems \textbf{2}:e00199--16.
-doi:\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.
 
-\hypertarget{ref-Lauber2010}{}
 11. \textbf{Lauber CL}, \textbf{Zhou N}, \textbf{Gordon JI},
@@
 human-associated samples. FEMS Microbiology Letters \textbf{307}:80--86.
-doi:\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.
+doi:\href{http://dx.doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.
 
-\hypertarget{ref-Song2016}{}
 12. \textbf{Song SJ}, \textbf{Amir A}, \textbf{Metcalf JL},
@@
 affecting suitability for field studies. mSystems \textbf{1}:e00021--16.
-doi:\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.
 
-\hypertarget{ref-Costea2017}{}
 13. \textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa S},
@@
 processing in metagenomic studies. Nature Biotechnology.
-doi:\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.
+doi:\href{http://dx.doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.
 
-\hypertarget{ref-Haas2011}{}
 14. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
@@
 454-pyrosequenced PCR amplicons. Genome Research \textbf{21}:494--504.
-doi:\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
+doi:\href{http://dx.doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
 
-\hypertarget{ref-Sinha2017}{}
 15. \textbf{Sinha R}, \textbf{Abu-Ali G}, \textbf{Vogtmann E},
@@
 quality control (MBQC) project consortium. Nature Biotechnology.
-doi:\href{https://doi.org/10.1038/nbt.3981}{10.1038/nbt.3981}.
+doi:\href{http://dx.doi.org/10.1038/nbt.3981}{10.1038/nbt.3981}.
 
-\hypertarget{ref-Meisel2016}{}
 16. \textbf{Meisel JS}, \textbf{Hannigan GD}, \textbf{Tyldsley AS},
@@
 \textbf{136}:947--956.
-doi:\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.
+doi:\href{http://dx.doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.
 
-\hypertarget{ref-Caporaso2010}{}
 17. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
@@
 the National Academy of Sciences \textbf{108}:4516--4522.
-doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.
+doi:\href{http://dx.doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.
 
-\hypertarget{ref-Kozich2013}{}
 18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+doi:\href{http://dx.doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
 
-\hypertarget{ref-Schloss2011}{}
 19. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+doi:\href{http://dx.doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
 
-\hypertarget{ref-Bokulich2012}{}
 20. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
@@
 \textbf{10}:57--59.
-doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
+doi:\href{http://dx.doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
 
-\hypertarget{ref-Parada2015}{}
 21. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
-Every base matters: Assessing small subunit rRNA primers for marine
+Every base matters: assessing small subunit rRNA primers for marine
 microbiomes with mock communities, time series and global field samples.
 Environmental Microbiology \textbf{18}:1403--1414.
-doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.
+doi:\href{http://dx.doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.
 
-\hypertarget{ref-Wang1996}{}
 22. \textbf{Wang GCY}, \textbf{Wang Y}. 1996. The frequency of chimeric
@@
 from different bacterial species. Microbiology \textbf{142}:1107--1114.
-doi:\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
+doi:\href{http://dx.doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
 
-\hypertarget{ref-Potapov2017}{}
 23. \textbf{Potapov V}, \textbf{Ong JL}. 2017. Examining sources of
@@
 \textbf{12}:e0169774.
-doi:\href{https://doi.org/10.1371/journal.pone.0169774}{10.1371/journal.pone.0169774}.
+doi:\href{http://dx.doi.org/10.1371/journal.pone.0169774}{10.1371/journal.pone.0169774}.
 
-\hypertarget{ref-Kebschull2015}{}
 24. \textbf{Kebschull JM}, \textbf{Zador AM}. 2015. Sources of
@@
 Acids Research gkv717.
-doi:\href{https://doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
+doi:\href{http://dx.doi.org/10.1093/nar/gkv717}{10.1093/nar/gkv717}.
 
-\hypertarget{ref-McInerney2014}{}
 25. \textbf{McInerney P}, \textbf{Adams P}, \textbf{Hadi MZ}. 2014.
@@
 polymerase. Molecular Biology International \textbf{2014}:1--8.
-doi:\href{https://doi.org/10.1155/2014/287430}{10.1155/2014/287430}.
+doi:\href{http://dx.doi.org/10.1155/2014/287430}{10.1155/2014/287430}.
 
-\hypertarget{ref-Cline1996}{}
 26. \textbf{Cline J}. 1996. PCR fidelity of pfu DNA polymerase and other
@@
 \textbf{24}:3546--3551.
-doi:\href{https://doi.org/10.1093/nar/24.18.3546}{10.1093/nar/24.18.3546}.
+doi:\href{http://dx.doi.org/10.1093/nar/24.18.3546}{10.1093/nar/24.18.3546}.
 
-\hypertarget{ref-Acinas2005}{}
 27. \textbf{Acinas SG}, \textbf{Sarma-Rupavtarm R}, \textbf{Klepac-Ceraj
@@
 \textbf{71}:8966--8969.
-doi:\href{https://doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
+doi:\href{http://dx.doi.org/10.1128/aem.71.12.8966-8969.2005}{10.1128/aem.71.12.8966-8969.2005}.
 
-\hypertarget{ref-Polz1998}{}
 28. \textbf{Polz MF}, \textbf{Cavanaugh CM}. 1998. Bias in
@@
 
-\hypertarget{ref-Brooks2015}{}
 29. \textbf{Brooks JP}, \textbf{David J Edwards}, \textbf{Harwich MD},
@@
 P}, \textbf{Strauss JF}, \textbf{Jefferson KK}, \textbf{Buck GA}. 2015.
-The truth about metagenomics: Quantifying and counteracting bias in 16S
+The truth about metagenomics: quantifying and counteracting bias in 16S
 rRNA studies. BMC Microbiology \textbf{15}.
-doi:\href{https://doi.org/10.1186/s12866-015-0351-6}{10.1186/s12866-015-0351-6}.
+doi:\href{http://dx.doi.org/10.1186/s12866-015-0351-6}{10.1186/s12866-015-0351-6}.
 
-\hypertarget{ref-Suzuki1996}{}
 30. \textbf{Suzuki MT}, \textbf{Giovannoni SJ}. 1996. Bias caused by
 template annealing in the amplification of mixtures of 16S rRNA genes by
-pcr. Applied and environmental microbiology \textbf{62}:625--630.
+pCR. Applied and environmental microbiology \textbf{62}:625--630.
 
-\hypertarget{ref-Chandler1997}{}
 31. \textbf{Chandler D}, \textbf{Fredrickson J}, \textbf{Brockman F}.
-1997. Effect of pcr template concentration on the composition and
+1997. Effect of pCR template concentration on the composition and
 distribution of total community 16S rDNA clone libraries. Molecular
@@
 
-\hypertarget{ref-Wagner1994}{}
 32. \textbf{Wagner A}, \textbf{Blackstone N}, \textbf{Cartwright P},
@@
 Surveys of gene families using polymerase chain reaction: PCR selection
-and pcr drift. Systematic Biology \textbf{43}:250--261.
+and pCR drift. Systematic Biology \textbf{43}:250--261.
 
-\hypertarget{ref-Hansen1998}{}
 33. \textbf{Hansen MC}, \textbf{Tolker-Nielsen T}, \textbf{Givskov M},
-\textbf{Molin S}. 1998. Biased 16S rDNA pcr amplification caused by
-interference from dna flanking the template region. FEMS Microbiology
+\textbf{Molin S}. 1998. Biased 16S rDNA pCR amplification caused by
+interference from dNA flanking the template region. FEMS Microbiology
 Ecology \textbf{26}:141--149.
 
-\hypertarget{ref-Kennedy2014}{}
 34. \textbf{Kennedy K}, \textbf{Hall MW}, \textbf{Lynch MDJ},
@@
 Environmental Microbiology \textbf{80}:5717--5722.
-doi:\href{https://doi.org/10.1128/aem.01451-14}{10.1128/aem.01451-14}.
+doi:\href{http://dx.doi.org/10.1128/aem.01451-14}{10.1128/aem.01451-14}.
 
-\hypertarget{ref-Edgar2011}{}
 35. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+doi:\href{http://dx.doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
 
-\hypertarget{ref-Edgar2017}{}
 36. \textbf{Edgar RC}. 2017. UNBIAS: An attempt to correct abundance
 bias in 16S sequencing, with limited success.
-doi:\href{https://doi.org/10.1101/124149}{10.1101/124149}.
+doi:\href{http://dx.doi.org/10.1101/124149}{10.1101/124149}.
 
-\hypertarget{ref-Gohl2016}{}
 37. \textbf{Gohl DM}, \textbf{Vangay P}, \textbf{Garbe J},
@@
 studies. Nature Biotechnology \textbf{34}:942--949.
-doi:\href{https://doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
+doi:\href{http://dx.doi.org/10.1038/nbt.3601}{10.1038/nbt.3601}.
 
-\hypertarget{ref-Salter2014}{}
 38. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
@@
 \textbf{12}.
-doi:\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
+doi:\href{http://dx.doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
 
-\hypertarget{ref-Callahan2016}{}
 39. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
@@
 Methods \textbf{13}:581--583.
-doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
+doi:\href{http://dx.doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
 
-\hypertarget{ref-Edgar2016}{}
-40. \textbf{Edgar RC}. 2016. UNOISE2: Improved error-correction for
+40. \textbf{Edgar RC}. 2016. UNOISE2: improved error-correction for
 illumina 16S and ITS amplicon sequencing.
-doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.
+doi:\href{http://dx.doi.org/10.1101/081257}{10.1101/081257}.
 
-\hypertarget{ref-Amir2017a}{}
 41. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
@@
 single-nucleotide community sequence patterns. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.
 
-\hypertarget{ref-Schloss2018}{}
 42. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
@@
 microbiome research. mBio \textbf{9}.
-doi:\href{https://doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
+doi:\href{http://dx.doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
 
-\hypertarget{ref-Seekatz2016}{}
 43. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
@@
 Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+doi:\href{http://dx.doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
 
-\hypertarget{ref-Schloss2009}{}
 44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 \textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+doi:\href{http://dx.doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
 
-\hypertarget{ref-Rognes2016}{}
 45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
-\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+\textbf{Quince C}, \textbf{Mah{é} F}. 2016. VSEARCH: a versatile open
 source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+doi:\href{http://dx.doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
 
-\hypertarget{ref-Westcott2017}{}
 46. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
@@
 operational taxonomic units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
+doi:\href{http://dx.doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
 
-\hypertarget{ref-r_citation_2018}{}
 47. \textbf{R Core Team}. 2018. R: A language and environment for
-statistical computing. R Foundation for Statistical Computing, Vienna,
+statistical computing. R Foundation for Statistical ComputingVienna,
 Austria.
 
-\hypertarget{ref-Tidyverse2017}{}
-48. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
+48. \textbf{Wickham H}. 2017. tidyverse: Easily install and load the
 'tidyverse'.
 
-\hypertarget{ref-Vegan2018}{}
 49. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
@@
 \textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
-\textbf{Wagner H}. 2018. Vegan: Community ecology package.
-
-\newpage
-
-\textbf{Figure 1: XXXXXXXX.}
-
-\newpage
-
-\textbf{Figure S1: XXXXXXXX.}
+\textbf{Wagner H}. 2018. vegan: Community ecology package.
 

### 2019-01-14 submission/manuscript.Rmd
Fix legend placement

@@
     keep_tex: yes
-  word_document: default
 geometry: margin=1.0in
@@
 
-The authors would like to thank all the study participants in ERIN whose samples were utilized.
+The authors thank the study participants in ERIN whose samples were utilized.
 We also would like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
@@
 
-**Figure 1: XXXXXXXX.**
+## Figure Lengends
 
-\newpage
+**Figure 1.
+The error rate of assembled sequence reads increases with the number of rounds of PCR used and follows the relative error rates provided by the manufacturers; however, much of this error is mediated by denoising.**
 
-**Figure S1: XXXXXXXX.**
+**Figure 2.
+The fraction of all denoised sequences that were identified as being chimeric increases with the number of rounds of PCR used and varied between polymerases.** (A) Sequencing of a mock community allowed us to identify the total fraction of sequences that were chimeric as well as the sensitivity and specificity of UCHIME to detect those chimeras.
+(B) Sequencing of four human stool samples after using one of five different polymerases again demonstrated increased rate of chimera formation with increasing number of rounds of PCR and variation across polymerases.
+
+**Figure 3.
+The relative abundances of reads mapped to reference sequences differed subtly from the expected relative abundances as determined by shotgun metagenomic sequencing.** Bias did not increase with number of rounds of PCR or vary by polymerase or the guanine and cytosine content of the fragment.
+The expected relative abundance of each organism is indicated by the horizontal gray line.
+The percentage of bases that were guanines or cytosines within the V4 region of the 16S rRNA genes in each organism is indicated by the number in the lower left corner of each panel.
+
+**Figure 4.
+Despite evidence of subtle PCR bias at the genome level, there was significant evidence of bias using community-wide metrics that grew with the number of rounds of PCR when using a mock community.** (A) With the exception of data collected using the Kappa polymerase, the OTU richness and Shannon diversity values increased with number of rounds of PCR and the inclusion of residual sequencing errors and chimeras.
+The horizontal black line indicates the expected richness and diversity if there were no bias, errors, or chimeras.
+(B) Relative to the mock community sampled after 25 rounds of PCR, the Bray-Curtis distance to the communities sampled after 30 and 35 rounds of PCR increased for all polymerases.
+(C) The variation between samples collected after 20, 25, 30, and 35 rounds of PCR for the five polymerases demonstrated a significant change in the community driven by the number of rounds of PCR and the polymerase used.
+
+**Figure 5.
+Sequencing of human stool samples indicated clear increase in bias with number of rounds of PCR, however, the bias appeared to be consistent within each sample.** (A) With the exception of data collected using the Kappa polymerase, the OTU richness and Shannon diversity values increased with number of rounds of PCR.
+(B) Relative to the stool communities sampled after 25 rounds of PCR, the Bray-Curtis distance to the stool communities sampled after 30 and 35 rounds of PCR was inconsistent and there was little difference in variation for data collected using the Kappa polymerase.
+(C) The variation between stool samples was larger than the amount of variation introduced by varying the number of rounds of PCR or polymerase.
+
+**Figure 6.
+There is little evidence for PCR drift adversely impacting the results of amplicon studies.** The average distance between replicates of sequencing the same mock community or between the the human stool samples did not vary by number of rounds of PCR or by polymerase.
+
+**Figure S1: With the exception of the seuqence data generated using the Kappa polymerase, the ratio of the two *Salmonella enterica* V4 sequences to a value was lower than the expected ratio of 6:1.** The two *S.
+enterica* V4 sequences differed by a single base.

### 2019-01-14 submission/manuscript.tex
Render manuscript
Fix legend placement

--- 2018-12-21 submission/manuscript.tex
@@
 
+\hypertarget{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}{%
 \section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
 High Fidelity DNA
-Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}
-
+Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}}
+
+\hypertarget{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}{%
 \section{Influence of DNA Polymerase on PCR bias, chimera formation, and
 errors when sequencing 16S rRNA genes from complex
-communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}
+communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}}
 
@@
 \newpage
-
 \linenumbers
 
-\subsection{Abstract}\label{abstract}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
@@
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
@@
 
-\subsection{Results}\label{results}
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
@@
 Interestingly, for some bacteria, their representation became less
-biased with additional rounds of PCR (e.g. \emph{L. fermentum}), while
-others became more biased (e.g. \emph{E. faecalis}), and others had
-little change (e.g. \emph{B. subtilis}). The percentage of bases in the
+biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
+others became more biased (e.g.~\emph{E. faecalis}), and others had
+little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
 V4 region that were guanines or cytosines was not predictive of the
@@
 
-\subsection{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 
-\subsection{Materials \& Methods}\label{materials-methods}
+\hypertarget{materials-methods}{%
+\subsection{Materials \& Methods}\label{materials-methods}}
 
@@
 were used with modifications to generate MiSeq amplicon libraries
-{[}(18); \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The
-number of rounds of PCR used for each sample and polymerase 15 and
-increased by 5 rounds up to 35 cycles. Insufficient PCR product was
-generated using 15 rounds and has not been included in our analysis.
+{[}@Kozich2013;
+\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The number of
+rounds of PCR used for each sample and polymerase 15 and increased by 5
+rounds up to 35 cycles. Insufficient PCR product was generated using 15
+rounds and has not been included in our analysis.
 
@@
 replicated four times for the mock community and one time for each fecal
-sample. Libraries were generated as described previously {[}(18);
+sample. Libraries were generated as described previously {[}@Kozich2013;
 \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
-
-The authors would like to thank all the study participants in ERIN whose
-samples were utilized. We also would like to thank Judy Opp and April
-Cockburn for their effort in sequencing the samples as part of the
-Microbiome Core Facility at the University of Michigan. Additional
-thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
-earlier drafts of the manuscript and providing helpful critiques. Salary
-support for Marc A. Sze came from the Canadian Institute of Health
-Research and NIH grant UL1TR002240. Salary support for Patrick D.
-Schloss came from NIH grants P30DK034933 and 1R01CA215574.
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
+
+The authors thank the study participants in ERIN whose samples were
+utilized. We also would like to thank Judy Opp and April Cockburn for
+their effort in sequencing the samples as part of the Microbiome Core
+Facility at the University of Michigan. Additional thanks to members of
+the Schloss lab and Dr.~Marcy Balunas for reading earlier drafts of the
+manuscript and providing helpful critiques. Salary support for Marc A.
+Sze came from the Canadian Institute of Health Research and NIH grant
+UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants
+P30DK034933 and 1R01CA215574.
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-Gilbert2018}{}
+\leavevmode\hypertarget{ref-Gilbert2018}{}%
 1. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2018.
@@
 
-\hypertarget{ref-HMP2012}{}
+\leavevmode\hypertarget{ref-HMP2012}{}%
 2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
@@
 
-\hypertarget{ref-Schloss2016}{}
+\leavevmode\hypertarget{ref-Schloss2016}{}%
 3. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
@@
 
-\hypertarget{ref-Luo2016}{}
+\leavevmode\hypertarget{ref-Luo2016}{}%
 4. \textbf{Luo T}, \textbf{Srinivasan U}, \textbf{Ramadugu K},
@@
 
-\hypertarget{ref-Bassis2017}{}
+\leavevmode\hypertarget{ref-Bassis2017}{}%
 5. \textbf{Bassis CM}, \textbf{Nicholas M. Moore}, \textbf{Lolans K},
@@
 
-\hypertarget{ref-Gorzelak2015}{}
+\leavevmode\hypertarget{ref-Gorzelak2015}{}%
 6. \textbf{Gorzelak MA}, \textbf{Gill SK}, \textbf{Tasnim N},
@@
 
-\hypertarget{ref-Dominianni2014}{}
+\leavevmode\hypertarget{ref-Dominianni2014}{}%
 7. \textbf{Dominianni C}, \textbf{Wu J}, \textbf{Hayes RB}, \textbf{Ahn
@@
 
-\hypertarget{ref-BautistadelosSantos2016}{}
+\leavevmode\hypertarget{ref-BautistadelosSantos2016}{}%
 8. \textbf{Santos QMB-d los}, \textbf{Schroeder JL}, \textbf{Blakemore
@@
 
-\hypertarget{ref-Sinha2015}{}
+\leavevmode\hypertarget{ref-Sinha2015}{}%
 9. \textbf{Sinha R}, \textbf{Chen J}, \textbf{Amir A}, \textbf{Vogtmann
@@
 
-\hypertarget{ref-Amir2017b}{}
+\leavevmode\hypertarget{ref-Amir2017b}{}%
 10. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
@@
 
-\hypertarget{ref-Lauber2010}{}
+\leavevmode\hypertarget{ref-Lauber2010}{}%
 11. \textbf{Lauber CL}, \textbf{Zhou N}, \textbf{Gordon JI},
@@
 
-\hypertarget{ref-Song2016}{}
+\leavevmode\hypertarget{ref-Song2016}{}%
 12. \textbf{Song SJ}, \textbf{Amir A}, \textbf{Metcalf JL},
@@
 
-\hypertarget{ref-Costea2017}{}
+\leavevmode\hypertarget{ref-Costea2017}{}%
 13. \textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa S},
@@
 
-\hypertarget{ref-Haas2011}{}
+\leavevmode\hypertarget{ref-Haas2011}{}%
 14. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
@@
 
-\hypertarget{ref-Sinha2017}{}
+\leavevmode\hypertarget{ref-Sinha2017}{}%
 15. \textbf{Sinha R}, \textbf{Abu-Ali G}, \textbf{Vogtmann E},
@@
 
-\hypertarget{ref-Meisel2016}{}
+\leavevmode\hypertarget{ref-Meisel2016}{}%
 16. \textbf{Meisel JS}, \textbf{Hannigan GD}, \textbf{Tyldsley AS},
@@
 
-\hypertarget{ref-Caporaso2010}{}
+\leavevmode\hypertarget{ref-Caporaso2010}{}%
 17. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
@@
 
-\hypertarget{ref-Kozich2013}{}
+\leavevmode\hypertarget{ref-Kozich2013}{}%
 18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
-\hypertarget{ref-Schloss2011}{}
+\leavevmode\hypertarget{ref-Schloss2011}{}%
 19. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 
-\hypertarget{ref-Bokulich2012}{}
+\leavevmode\hypertarget{ref-Bokulich2012}{}%
 20. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
@@
 
-\hypertarget{ref-Parada2015}{}
+\leavevmode\hypertarget{ref-Parada2015}{}%
 21. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
@@
 
-\hypertarget{ref-Wang1996}{}
+\leavevmode\hypertarget{ref-Wang1996}{}%
 22. \textbf{Wang GCY}, \textbf{Wang Y}. 1996. The frequency of chimeric
@@
 
-\hypertarget{ref-Potapov2017}{}
+\leavevmode\hypertarget{ref-Potapov2017}{}%
 23. \textbf{Potapov V}, \textbf{Ong JL}. 2017. Examining sources of
@@
 
-\hypertarget{ref-Kebschull2015}{}
+\leavevmode\hypertarget{ref-Kebschull2015}{}%
 24. \textbf{Kebschull JM}, \textbf{Zador AM}. 2015. Sources of
@@
 
-\hypertarget{ref-McInerney2014}{}
+\leavevmode\hypertarget{ref-McInerney2014}{}%
 25. \textbf{McInerney P}, \textbf{Adams P}, \textbf{Hadi MZ}. 2014.
@@
 
-\hypertarget{ref-Cline1996}{}
+\leavevmode\hypertarget{ref-Cline1996}{}%
 26. \textbf{Cline J}. 1996. PCR fidelity of pfu DNA polymerase and other
@@
 
-\hypertarget{ref-Acinas2005}{}
+\leavevmode\hypertarget{ref-Acinas2005}{}%
 27. \textbf{Acinas SG}, \textbf{Sarma-Rupavtarm R}, \textbf{Klepac-Ceraj
@@
 
-\hypertarget{ref-Polz1998}{}
+\leavevmode\hypertarget{ref-Polz1998}{}%
 28. \textbf{Polz MF}, \textbf{Cavanaugh CM}. 1998. Bias in
@@
 
-\hypertarget{ref-Brooks2015}{}
+\leavevmode\hypertarget{ref-Brooks2015}{}%
 29. \textbf{Brooks JP}, \textbf{David J Edwards}, \textbf{Harwich MD},
@@
 
-\hypertarget{ref-Suzuki1996}{}
+\leavevmode\hypertarget{ref-Suzuki1996}{}%
 30. \textbf{Suzuki MT}, \textbf{Giovannoni SJ}. 1996. Bias caused by
@@
 
-\hypertarget{ref-Chandler1997}{}
+\leavevmode\hypertarget{ref-Chandler1997}{}%
 31. \textbf{Chandler D}, \textbf{Fredrickson J}, \textbf{Brockman F}.
@@
 
-\hypertarget{ref-Wagner1994}{}
+\leavevmode\hypertarget{ref-Wagner1994}{}%
 32. \textbf{Wagner A}, \textbf{Blackstone N}, \textbf{Cartwright P},
@@
 
-\hypertarget{ref-Hansen1998}{}
+\leavevmode\hypertarget{ref-Hansen1998}{}%
 33. \textbf{Hansen MC}, \textbf{Tolker-Nielsen T}, \textbf{Givskov M},
@@
 
-\hypertarget{ref-Kennedy2014}{}
+\leavevmode\hypertarget{ref-Kennedy2014}{}%
 34. \textbf{Kennedy K}, \textbf{Hall MW}, \textbf{Lynch MDJ},
@@
 
-\hypertarget{ref-Edgar2011}{}
+\leavevmode\hypertarget{ref-Edgar2011}{}%
 35. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 
-\hypertarget{ref-Edgar2017}{}
+\leavevmode\hypertarget{ref-Edgar2017}{}%
 36. \textbf{Edgar RC}. 2017. UNBIAS: An attempt to correct abundance
@@
 
-\hypertarget{ref-Gohl2016}{}
+\leavevmode\hypertarget{ref-Gohl2016}{}%
 37. \textbf{Gohl DM}, \textbf{Vangay P}, \textbf{Garbe J},
@@
 
-\hypertarget{ref-Salter2014}{}
+\leavevmode\hypertarget{ref-Salter2014}{}%
 38. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
@@
 
-\hypertarget{ref-Callahan2016}{}
+\leavevmode\hypertarget{ref-Callahan2016}{}%
 39. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
@@
 
-\hypertarget{ref-Edgar2016}{}
+\leavevmode\hypertarget{ref-Edgar2016}{}%
 40. \textbf{Edgar RC}. 2016. UNOISE2: Improved error-correction for
@@
 
-\hypertarget{ref-Amir2017a}{}
+\leavevmode\hypertarget{ref-Amir2017a}{}%
 41. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
@@
 
-\hypertarget{ref-Schloss2018}{}
+\leavevmode\hypertarget{ref-Schloss2018}{}%
 42. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
@@
 
-\hypertarget{ref-Seekatz2016}{}
+\leavevmode\hypertarget{ref-Seekatz2016}{}%
 43. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
@@
 
-\hypertarget{ref-Schloss2009}{}
+\leavevmode\hypertarget{ref-Schloss2009}{}%
 44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 
-\hypertarget{ref-Rognes2016}{}
+\leavevmode\hypertarget{ref-Rognes2016}{}%
 45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
@@
 
-\hypertarget{ref-Westcott2017}{}
+\leavevmode\hypertarget{ref-Westcott2017}{}%
 46. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
@@
 
-\hypertarget{ref-r_citation_2018}{}
+\leavevmode\hypertarget{ref-r_citation_2018}{}%
 47. \textbf{R Core Team}. 2018. R: A language and environment for
@@
 
-\hypertarget{ref-Tidyverse2017}{}
+\leavevmode\hypertarget{ref-Tidyverse2017}{}%
 48. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
@@
 
-\hypertarget{ref-Vegan2018}{}
+\leavevmode\hypertarget{ref-Vegan2018}{}%
 49. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
@@
 
-\textbf{Figure 1: XXXXXXXX.}
-
-\newpage
-
-\textbf{Figure S1: XXXXXXXX.}
+\hypertarget{figure-lengends}{%
+\subsection{Figure Lengends}\label{figure-lengends}}
+
+\textbf{Figure 1. The error rate of assembled sequence reads increases
+with the number of rounds of PCR used and follows the relative error
+rates provided by the manufacturers; however, much of this error is
+mediated by denoising.}
+
+\textbf{Figure 2. The fraction of all denoised sequences that were
+identified as being chimeric increases with the number of rounds of PCR
+used and varied between polymerases.} (A) Sequencing of a mock community
+allowed us to identify the total fraction of sequences that were
+chimeric as well as the sensitivity and specificity of UCHIME to detect
+those chimeras. (B) Sequencing of four human stool samples after using
+one of five different polymerases again demonstrated increased rate of
+chimera formation with increasing number of rounds of PCR and variation
+across polymerases.
+
+\textbf{Figure 3. The relative abundances of reads mapped to reference
+sequences differed subtly from the expected relative abundances as
+determined by shotgun metagenomic sequencing.} Bias did not increase
+with number of rounds of PCR or vary by polymerase or the guanine and
+cytosine content of the fragment. The expected relative abundance of
+each organism is indicated by the horizontal gray line. The percentage
+of bases that were guanines or cytosines within the V4 region of the 16S
+rRNA genes in each organism is indicated by the number in the lower left
+corner of each panel.
+
+\textbf{Figure 4. Despite evidence of subtle PCR bias at the genome
+level, there was significant evidence of bias using community-wide
+metrics that grew with the number of rounds of PCR when using a mock
+community.} (A) With the exception of data collected using the Kappa
+polymerase, the OTU richness and Shannon diversity values increased with
+number of rounds of PCR and the inclusion of residual sequencing errors
+and chimeras. The horizontal black line indicates the expected richness
+and diversity if there were no bias, errors, or chimeras. (B) Relative
+to the mock community sampled after 25 rounds of PCR, the Bray-Curtis
+distance to the communities sampled after 30 and 35 rounds of PCR
+increased for all polymerases. (C) The variation between samples
+collected after 20, 25, 30, and 35 rounds of PCR for the five
+polymerases demonstrated a significant change in the community driven by
+the number of rounds of PCR and the polymerase used.
+
+\textbf{Figure 5. Sequencing of human stool samples indicated clear
+increase in bias with number of rounds of PCR, however, the bias
+appeared to be consistent within each sample.} (A) With the exception of
+data collected using the Kappa polymerase, the OTU richness and Shannon
+diversity values increased with number of rounds of PCR. (B) Relative to
+the stool communities sampled after 25 rounds of PCR, the Bray-Curtis
+distance to the stool communities sampled after 30 and 35 rounds of PCR
+was inconsistent and there was little difference in variation for data
+collected using the Kappa polymerase. (C) The variation between stool
+samples was larger than the amount of variation introduced by varying
+the number of rounds of PCR or polymerase.
+
+\textbf{Figure 6. There is little evidence for PCR drift adversely
+impacting the results of amplicon studies.} The average distance between
+replicates of sequencing the same mock community or between the the
+human stool samples did not vary by number of rounds of PCR or by
+polymerase.
+
+\textbf{Figure S1: With the exception of the seuqence data generated
+using the Kappa polymerase, the ratio of the two \emph{Salmonella
+enterica} V4 sequences to a value was lower than the expected ratio of
+6:1.} The two \emph{S. enterica} V4 sequences differed by a single base.
 

### 2019-01-15 submission/manuscript.Rmd
Add inline R code and refs to figures

@@
 ---
+
+```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
+options(tidyverse.quiet = TRUE)
+
+library(tidyverse)
+library(knitr)
+
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("message" = FALSE)
+opts_chunk$set("cache" = FALSE)
+
+inline_hook <- function(x){
+	print(x)
+	if(is.list(x)){
+		x <- unlist(x)
+	}
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=2, nsmall=3, scientific=FALSE))
+		}
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
+```
 
@@
 
+```{r errors}
+error <- read_tsv("../data/process/error_chimera_rates.tsv") %>%
+	filter(rounds == 30) %>%
+	select(polymerase, contig_error, pc_error) %>%
+	mutate(contig_error = 100 * contig_error,
+				pc_error = 100 * pc_error)
+```
+
 ***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs).
@@
 Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community.
-We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [ref].
-Regardless of the polymerase, the error rate increased with the number of rounds of amplification.
+We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [@Kozich2013; @Schloss2011; @Quince2011].
+Regardless of the polymerase, the error rate increased with the number of rounds of amplification (Figure 1).
 Using 30 rounds of PCR is a common approach across diverse types of samples.
 Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e.
-0.XXX%) followed by the Platinum (i.e.
-0.XXX%), Phusion (i.e.
-0.XXX%), Kappa (i.e.
-0.XXX%), and Q5 (i.e.
-0.XXX%) polymerases (Figure 1).
+`r filter(error, polymerase=="ACC") %>% select(contig_error)`%) followed by the Platinum (i.e.
+`r filter(error, polymerase=="PL") %>% select(contig_error)`%), Phusion (i.e.
+`r filter(error, polymerase=="PHU") %>% select(contig_error)`%), Kappa (i.e.
+`r filter(error, polymerase=="K") %>% select(contig_error)`%), and Q5 (i.e.
+`r filter(error, polymerase=="Q5") %>% select(contig_error)`%) polymerases (Figure 1).
 When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e.
-0.XXX%) followed by the Accuprime (i.e.
-0.XXX%), Phusion (i.e.
-0.XXX%), Kappa (i.e.
-0.XXX%), and Q5 (i.e.
-0.XXX%) polymerases (Figure 1).
+`r filter(error, polymerase=="PL") %>% select(pc_error)`%) followed by the Accuprime (i.e.
+`r filter(error, polymerase=="ACC") %>% select(pc_error)`%), Q5 (i.e.
+`r filter(error, polymerase=="Q5") %>% select(pc_error)`%), Phusion (i.e.
+`r filter(error, polymerase=="PHU") %>% select(pc_error)`%), and Kappa (i.e.
+`r filter(error, polymerase=="K") %>% select(pc_error)`%) polymerases (Figure 1).
 Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.
+
+```{r chimeras}
+chimera <- read_tsv("../data/process/error_chimera_rates.tsv") %>%
+	filter(rounds == 30) %>%
+	select(polymerase, pc_frac_chimera) %>%
+	mutate(pc_frac_chimera = 100 * pc_frac_chimera)
+
+min_specificity <- read_tsv("../data/process/error_chimera_rates.tsv") %>%
+	filter(rounds == 30) %>%
+	select(vsearch_sens, vsearch_spec) %>%
+	summarize(min_spec = min(100 * vsearch_spec))
+
+sensitivity <- read_tsv("../data/process/error_chimera_rates.tsv") %>%
+	filter(rounds == 30) %>%
+	select(vsearch_sens, vsearch_spec) %>%
+	summarize(min = min(100*vsearch_sens), max=max(100*vsearch_sens))
+
+inline_hook <- function(x){
+	print(x)
+	if(is.list(x)){
+		x <- unlist(x)
+	}
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
+		}
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
+
+```
 
@@
 We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence.
-As expected from previous studies, the number of chimeras increased with rounds of amplification.
+As expected from previous studies, the number of chimeras increased with rounds of amplification (Figure 2A).
 Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used.
 After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e.
-XX%) followed by the Q5 (i.e.
-XX%), Phusion (i.e.
-XX%), Kappa (i.e.
-XX%), and Accuprime (i.e.
-XX%) polymerases.
-We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity.
-For all polymerases, the specificity was above XX% and showed a weak association with the number of cycles used.
+`r filter(chimera, polymerase=="PL") %>% select(pc_frac_chimera)`%) followed by the Q5 (i.e.
+`r filter(chimera, polymerase=="Q5") %>% select(pc_frac_chimera)`%), Phusion (i.e.
+`r filter(chimera, polymerase=="PHU") %>% select(pc_frac_chimera)`%), Kappa (i.e.
+`r filter(chimera, polymerase=="K") %>% select(pc_frac_chimera)`%), and Accuprime (i.e.
+`r filter(chimera, polymerase=="ACC") %>% select(pc_frac_chimera)`%) polymerases.
+We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity (Figure 2A).
+For all polymerases, the specificity was above `r min_specificity`% and showed a weak association with the number of cycles used.
 There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community.
 This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
-In general, the sensitivity of UChime varied between XX and XX%.
+The sensitivity of UChime varied between `r sensitivity["min"]` and `r sensitivity["max"]`%.
 The generalizability of these results is limited because we used a single mock community with limited genetic diversity.
-Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric.
+Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric  (Figure 2B).
 These results followed those from the mock communities; additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
 Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
+
+```{r species_bias}
+salmonella_data <- read_tsv("../data/process/mock_bias_salmonella.tsv") %>%	filter(rounds == 35)
+
+others <- salmonella_data %>% filter(polymerase != "K") %>% summarize(mean = mean(ratio))
+
+kappa <- read_tsv("../data/process/mock_bias_salmonella.tsv") %>% filter(polymerase == "K", rounds >=25, rounds <= 35) %>% select(ratio) %>% summarize(min = min(ratio), max = max(ratio))
+```
 
@@
 The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community.
-We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase.
+We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase (Figure 3).
 Interestingly, for some bacteria, their representation became less biased with additional rounds of PCR (e.g.
@@
 We used the sequence data to calculate the ratio of the dominant to rare variants from *S.
-enterica* (Figure.
-XXX).
-The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of PCR.
+enterica* (Figure S1).
+The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of `r others`; however, the ratio for the Kappa polymerase varied between `r kappa["min"]` and `r kappa["max"]` for 25 to 35 rounds of PCR.
 Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias and the previous observation regarding the guanine and cytosine relative abundance was not observed with our data.
 
-***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline.
+```{r community_bias}
+mock_beta <- read_tsv("../data/process/mock_beta_diversity.tsv") %>%
+	separate(rows, into=c("row_round", "row_poly", row="row_mock")) %>%
+	separate(columns, into=c("col_round", "col_poly", row="col_mock")) %>%
+	filter(row_poly == col_poly, row_round == "25x", method == "vsearch") %>%
+	select(col_round, row_poly, distances)
+
+mean_diff <- mock_beta %>%
+	group_by(row_poly) %>%
+	spread(key = col_round, value = distances) %>%
+	mutate(diff = `35x` - `30x`) %>%
+	ungroup() %>%
+	summarize(mean = mean(diff))
+
+pl <- mock_beta %>% filter(row_poly == "PL")
+
+mock_vegan <- read_tsv("../data/process/vegan_mock.tsv")
+stopifnot(mock_vegan$p < 0.001)
+
+inline_hook <- function(x){
+	print(x)
+	if(is.list(x)){
+		x <- unlist(x)
+	}
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
+		}
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
+```
+
+***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline (Figure 4).
 First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors.
 This gave us a community distribution that reflected the distribution following PCR without any artifacts.
-Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased.
+Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased (Figure 4A).
 These data suggest that PCR had the effect of making the community distribution more even than it was originally.
 Next, we used the true sequence diversity, but removed chimeras based on mapping reads to all possible chimeras between reference sequences, and clustered the reads to OTUs.
-The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher.
+The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher (Figure 4A).
 Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras.
-Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher.
+Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher (Figure 4A).
 These comparisons demonstrated that although the bias at the sequence level was subtle, PCR does have a bias that is exacerbated by errors and chimeras.
-When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units.
-The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX).
-Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR.
-This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=0.XX, P=0.XXXX) was a larger factor than the choice of polymerase (R^2^=0.XX, P=0.XXXX).
+When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B).
+The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: `r filter(pl, col_round=="30x") %>% select(distances)`; 25 vs 35 rounds: `r filter(pl, col_round=="35x") %>% select(distances)`).
+Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR (Figure 4C).
+This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=`r filter(mock_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) was a larger factor than the choice of polymerase (R^2^=`r filter(mock_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
 These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure.
@@
 
+```{r stool_community}
+inter_diversity <- read_tsv("../data/process/stool_beta_diversity.tsv") %>%
+	separate(rows, into=c("row_round", "row_poly", row="row_sample")) %>%
+	separate(columns, into=c("col_round", "col_poly", row="col_sample")) %>%
+	filter(row_poly == col_poly, row_round == col_round) %>%
+	group_by(row_poly,row_round) %>%
+	summarize(mean = mean(distances)) %>%
+	ungroup() %>%
+	summarize(min=min(mean), max=max(mean))
+
+stool_vegan <- read_tsv("../data/process/vegan_stool.tsv")
+stopifnot(stool_vegan$p < 0.001)
+```
+
 ***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels using mock community data appeared to be small relative to the expected differences between biological samples.
 To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
-We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor.
+We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor (Figure 5A).
 Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
-When we characterized the variation between rounds of amplification using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases.
+When we characterized the variation between rounds of amplification using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases (Figure 5B).
 In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases.
 The data generated using the Platinum polymerase was consistent across rounds, but it was more biased than these polymerases.
-Considering the average distance across the four samples varied between 0.XX and 0.XX, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
-In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by donor.
+Considering the average distance across the four samples varied between `r inter_diversity["min"]` and `r inter_diversity["max"]`, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
+In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by donor (Figure 5C).
 Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
-However, our statistical analysis using adonis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=0.XXX, P=0.XXXX), followed by the number of rounds of PCR (R^2^=0.XXX, P=0.XXXX) and the choice of polymerase (R^2^=0.XXX, P=0.XXXX).
+However, our statistical analysis using adonis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=`r filter(stool_vegan, effects=="subject") %>% select(r_sq)`, P<0.001), followed by the number of rounds of PCR (R^2^=`r filter(stool_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) and the choice of polymerase (R^2^=`r filter(stool_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
 Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
@@
 
+```{r drift}
+mock_drift <- read_csv("../data/process/mock_beta_drift.csv") %>% filter(method == "vsearch") %>% group_by(rounds, polymerase) %>% summarize(mean = mean(ave_dist)) %>% ungroup() %>% summarize(min = min(mean), max = max(mean))
+
+stool_drift <- read_tsv("../data/process/stool_beta_diversity.tsv") %>%
+	separate(rows, into=c("rounds", "polymerase", "row_subject")) %>%
+	separate(columns, into=c("col_rounds", "col_polymerase", "col_subject")) %>%
+	filter(rounds == col_rounds, polymerase == col_polymerase) %>%
+	select(rounds, polymerase, row_subject, col_subject, distances) %>%
+	group_by(rounds, polymerase) %>%
+	summarize(ave_dist = mean(distances), n = n()) %>%
+	ungroup() %>%
+	filter(n == 6) %>%
+	summarize(min = min(ave_dist), max = max(ave_dist))
+```
+
 ***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift.
-To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification.
+To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification (Figure 6).
 Using the mock community data there were no obvious trends.
-The average Bray-Curtis distance within a set of conditions varied by 0.XX to 0.XX units.
-Although we did not obtain replicates of each of the stool samples, the intra-sample variation for each set of conditions was consistent and varied between 0.XX and 0.XX units.
+The average Bray-Curtis distance within a set of conditions varied by `r mock_drift["min"]` to `r mock_drift["max"]` units.
+Although we did not obtain replicates of each of the stool samples, the intra-sample variation for each set of conditions was consistent and varied between `r stool_drift["min"]` and `r stool_drift["max"]` units.
 These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.
@@
 
-The observed sequencing error rates and alpha diversity metrics followed the manufacturers' measurements of their polymerases' fidelity.
+The observed sequencing error rates and alpha diversity metrics followed the manufacturers' measurements of their polymerases' fidelity (Figure 1).
 Accuprime and Platinum have fidelity that are approximately 10-times higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa are more than 100 times higher.
@@
 Manufacturer recommendations were followed except for the annealing and extension times, which were selected based on previously published protocols [@Kozich2013; @Gohl2016].
-Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@Kozich2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/].
+Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@Kozich2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/).
 The number of rounds of PCR used for each sample and polymerase 15 and increased by 5 rounds up to 35 cycles.
@@
 combination of polymerase and number of rounds of PCR) were replicated four times for the mock community and one time for each fecal sample.
-Libraries were generated as described previously [@Kozich2013; https://github.com/SchlossLab/MiSeq_WetLab_SOP/].
+Libraries were generated as described previously [@Kozich2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/).
 The libraries were sequenced using the Illumina MiSeq sequencing platform to generate paired 250-nt reads.

### 2019-01-15 submission/manuscript.tex
Add inline R code and refs to figures

@@
 generated using 30 rounds of PCR the Accuprime polymerase had the
-highest error rate (i.e.~0.XXX\%) followed by the Platinum
-(i.e.~0.XXX\%), Phusion (i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5
-(i.e.~0.XXX\%) polymerases (Figure 1). When we applied the
+highest error rate (i.e.~0.124\%) followed by the Platinum
+(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
+(i.e.~0.060\%) polymerases (Figure 1). When we applied the
 pre.clustering denoising step, which merges the counts of reads within 2
@@
 such that the Platinum polymerase had the highest error rate
-(i.e.~0.XXX\%) followed by the Accuprime (i.e.~0.XXX\%), Phusion
-(i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5 (i.e.~0.XXX\%) polymerases
-(Figure 1). Although specific recommendations are difficult to make
-because the impact of the actual community structure and concentration
-of the initial DNA template are likely to have an impact on the results,
-it is clear that using as few PCR cycles as necessary and a polymerase
-with the lowest possible error rate is a good guide to minimizing the
-impact of polymerase on the error rate.
+(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
+(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
+polymerases (Figure 1). Although specific recommendations are difficult
+to make because the impact of the actual community structure and
+concentration of the initial DNA template are likely to have an impact
+on the results, it is clear that using as few PCR cycles as necessary
+and a polymerase with the lowest possible error rate is a good guide to
+minimizing the impact of polymerase on the error rate.
 
@@
 After 30 rounds of PCR, the Platinum polymerase had the highest chimera
-rate (i.e.~XX\%) followed by the Q5 (i.e.~XX\%), Phusion (i.e.~XX\%),
-Kappa (i.e.~XX\%), and Accuprime (i.e.~XX\%) polymerases. We used the
-UChime algorithm to detect chimeras in our pre-clustered mock community
-data and calculate the algorithm's sensitivity and specificity. For all
-polymerases, the specificity was above XX\% and showed a weak
-association with the number of cycles used. There was considerable
-inter-polymerase and inter-round of amplification variation in the
-sensitivity of UChime to detect the chimeras from the mock community.
-This suggested that the residual error rate after pre-clustering the
-sequence data did not compromise the sensitivity. In general, the
-sensitivity of UChime varied between XX and XX\%. The generalizability
-of these results is limited because we used a single mock community with
-limited genetic diversity. Although we did not know the true chimera
-rate for our four human stool samples, we were able to calculate the
-fraction of sequences that UChime identified as being chimeric. These
-results followed those from the mock communities; additional rounds of
-amplification significantly increased the rate of chimeras and there was
-variation between the polymerases that we used. Although it was not
-possible to identify the features of a polymerase that resulted in
-higher rates of chimeras, it is clear that using the smallest number of
-PCR cycles possible will minimize the impacts of chimeras.
+rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
+(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
+polymerases. We used the UChime algorithm to detect chimeras in our
+pre-clustered mock community data and calculate the algorithm's
+sensitivity and specificity. For all polymerases, the specificity was
+above 95.8\% and showed a weak association with the number of cycles
+used. There was considerable inter-polymerase and inter-round of
+amplification variation in the sensitivity of UChime to detect the
+chimeras from the mock community. This suggested that the residual error
+rate after pre-clustering the sequence data did not compromise the
+sensitivity. The sensitivity of UChime varied between 50 and 84.8\%. The
+generalizability of these results is limited because we used a single
+mock community with limited genetic diversity. Although we did not know
+the true chimera rate for our four human stool samples, we were able to
+calculate the fraction of sequences that UChime identified as being
+chimeric. These results followed those from the mock communities;
+additional rounds of amplification significantly increased the rate of
+chimeras and there was variation between the polymerases that we used.
+Although it was not possible to identify the features of a polymerase
+that resulted in higher rates of chimeras, it is clear that using the
+smallest number of PCR cycles possible will minimize the impacts of
+chimeras.
 
@@
 data to calculate the ratio of the dominant to rare variants from
-\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
-Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
-Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
-PCR. Given the subtle nature of the variation in the relative abundances
-of each 16S rRNA gene fragment, it was not possible to create
-generalizable rules that would explain the bias and the previous
-observation regarding the guanine and cytosine relative abundance was
-not observed with our data.
+\emph{S. enterica} (Figure XXX). The Accuprime, Phusion, Platinum, and
+Q5 polymerases converged to a ratio of 5.4; however, the ratio for the
+Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
+Given the subtle nature of the variation in the relative abundances of
+each 16S rRNA gene fragment, it was not possible to create generalizable
+rules that would explain the bias and the previous observation regarding
+the guanine and cytosine relative abundance was not observed with our
+data.
 
@@
 distances between 25 and 35 rounds were higher than between 25 and 30
-rounds for each of the polymerases by an average of 0.XX units. The
+rounds for each of the polymerases by an average of 0.023 units. The
 Platinum polymerase varied the most across rounds of amplification (25
-vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). Although the distances
+vs 30 rounds: 0.13; 25 vs 35 rounds: 0.16). Although the distances
 between samples were small, the ordination of these distances showed a
@@
 observation was supported by our statistical analysis, which revealed
-that the number of rounds of PCR (R\textsuperscript{2}=0.XX, P=0.XXXX)
-was a larger factor than the choice of polymerase
-(R\textsuperscript{2}=0.XX, P=0.XXXX). These results demonstrate that
-subtle differences in relative abundances can have an impact on overall
-community structure. This variation underscores the importance of only
-comparing sequence data that have been generated using the same PCR
-conditions.
+that the number of rounds of PCR (R\textsuperscript{2}=0.20,
+P\textless{}0.001) was a larger factor than the choice of polymerase
+(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
+demonstrate that subtle differences in relative abundances can have an
+impact on overall community structure. This variation underscores the
+importance of only comparing sequence data that have been generated
+using the same PCR conditions.
 
@@
 but it was more biased than these polymerases. Considering the average
-distance across the four samples varied between 0.XX and 0.XX,
+distance across the four samples varied between 0.39 and 0.56,
 regardless of the polymerases and number of rounds of amplification, any
@@
 differences in the community structures with the subject explaining the
-most variation (R\textsuperscript{2}=0.XXX, P=0.XXXX), followed by the
-number of rounds of PCR (R\textsuperscript{2}=0.XXX, P=0.XXXX) and the
-choice of polymerase (R\textsuperscript{2}=0.XXX, P=0.XXXX). Together,
-these results indicate that for a coarse analysis of communities, the
-choice of number of rounds of amplification or polymerase are not
-important, but that they must be consistent across samples. It is
-difficult to develop a specific recommendation based on the level of
-bias across rounds of PCR or polymerases; however, the general
-suggestion is to use as few rounds of amplification as possible.
+most variation (R\textsuperscript{2}=0.79, P\textless{}0.001), followed
+by the number of rounds of PCR (R\textsuperscript{2}=0.044,
+P\textless{}0.001) and the choice of polymerase
+(R\textsuperscript{2}=0.033, P\textless{}0.001). Together, these results
+indicate that for a coarse analysis of communities, the choice of number
+of rounds of amplification or polymerase are not important, but that
+they must be consistent across samples. It is difficult to develop a
+specific recommendation based on the level of bias across rounds of PCR
+or polymerases; however, the general suggestion is to use as few rounds
+of amplification as possible.
 
@@
 community data there were no obvious trends. The average Bray-Curtis
-distance within a set of conditions varied by 0.XX to 0.XX units.
+distance within a set of conditions varied by 0.063 to 0.11 units.
 Although we did not obtain replicates of each of the stool samples, the
 intra-sample variation for each set of conditions was consistent and
-varied between 0.XX and 0.XX units. These data suggest that amplicon
+varied between 0.50 and 0.56 units. These data suggest that amplicon
 sequencing is robust to random variation in amplification and that any
@@
 protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
-were used with modifications to generate MiSeq amplicon libraries
-{[}@Kozich2013;
-\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The number of
+were used with modifications to generate MiSeq amplicon libraries (18)
+(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The number of
 rounds of PCR used for each sample and polymerase 15 and increased by 5
@@
 replicated four times for the mock community and one time for each fecal
-sample. Libraries were generated as described previously {[}@Kozich2013;
-\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
+sample. Libraries were generated as described previously (18)
+(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The libraries
 were sequenced using the Illumina MiSeq sequencing platform to generate

### 2019-01-15 code/vegan_analysis.R

@@ -27,30 +27,20 @@ read_dist <- function(dist_file_name){
 	as.dist(distance_matrix)
 }
 
-
 ####################################################################################################
 
-pmock_file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.perfect.opti_mcc.braycurtis.0.03.lt.ave.dist"
+mock_file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.vsearch.opti_mcc.braycurtis.0.03.lt.ave.dist"
 
-pmock_dist <- read_dist(pmock_file_name)
+mock_dist <- read_dist(mock_file_name)
 
-pmock_samples <- tibble(names = attr(pmock_dist, "Labels")) %>% mutate(names=str_replace(names, "_Mock_.*", "")) %>% separate(names, into=c("rounds", "polymerase"))
+mock_samples <- tibble(names = attr(mock_dist, "Labels")) %>% mutate(names=str_replace(names, "_Mock_.*", "")) %>% separate(names, into=c("rounds", "polymerase"))
 
-pmock_adonis <- adonis(pmock_dist~pmock_samples$rounds+pmock_samples$polymerase, permutations=9999)
-pmock_r2 <- pmock_adonis$aov.tab$R2[1:2]
-names(pmock_r2) <- c("rounds", "polymerase")
+mock_adonis <- adonis(mock_dist~mock_samples$rounds+mock_samples$polymerase, permutations=9999)
 
-####################################################################################################
-
-vmock_file_name <- "data/mothur/mock.trim.contigs.good.unique.good.filter.unique.pick.pick.precluster.vsearch.opti_mcc.braycurtis.0.03.lt.ave.dist"
-
-vmock_dist <- read_dist(vmock_file_name)
-
-vmock_samples <- tibble(names = attr(vmock_dist, "Labels")) %>% mutate(names=str_replace(names, "_Mock_.*", "")) %>% separate(names, into=c("rounds", "polymerase"))
-
-vmock_adonis <- adonis(vmock_dist~vmock_samples$rounds+vmock_samples$polymerase, permutations=9999)
-vmock_r2 <- vmock_adonis$aov.tab$R2[1:2]
-names(vmock_r2) <- c("rounds", "polymerase")
+tibble(effects = c("rounds", "polymerase"),
+			r_sq = mock_adonis$aov.tab$R2[1:2],
+			p = mock_adonis$aov.tab$Pr[1:2]) %>%
[…]

### 2019-01-15 submission/manuscript.Rmd
Fresh make write.paper

@@
 
-# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase
+\vspace{25mm}
+\begin{center}
 
-# Influence of DNA Polymerase on PCR bias, chimera formation, and errors when sequencing 16S rRNA genes from complex communities
+\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}
 
-\begin{center}
+\vspace{50mm}
+
+\textbf{Running title:} Quantifying the choice of DNA polymerase
+
 \vspace{25mm}
@@
 
-Co-author e-mails:
-
-* marcsze@med.umich.edu
-
 \newpage
@@
 ## Abstract
+
+PCR amplification of 16S rRNA genes is a critical, yet under appreciated step in the generation of sequence data to describe the taxonomic composition of microbial communities.
+Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample.
+The choice of PCR cycling conditions are known to impact the outcome of PCR and downstream inferences.
+We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community with known sequences and community structure and human stool samples.
+Although it is impossible to derive specific recommendations, we did observe general trends.
+Namely, using a polymerase with the highest possible fidelity and minimizing the number of rounds of PCR reduced the sequencing error rate, fraction of chimeric sequences, and bias.
+Evidence of bias at the sequence level was subtle and could not be ascribed to the fragments' fraction of bases that were guanines or cytosines.
+When analyzing the mock community data, the amount that the community deviated from the expected composition increased with rounds of PCR.
+The effect was inconsistent for human stool samples.
+Overall the results underscore the difficulty of comparing sequence data that are generated by different PCR protocols; however, the results indicate that the variation in human stool samples is generally larger than that introduced by the choice of polymerase or number of rounds of PCR.
+
+### Importance
+
+The growth in interest in characterizing microbial communities from diverse communities has exploded with the steep decline in sequencing costs.
+A significant amount of effort has gone into understanding the error profiles of DNA sequencers; however, little work has been done to understand the effects of choices that researchers make to generate the PCR amplicons that are sequenced.
+Here we quantified the effects of the choice of polymerase and number of PCR cycles on the quality of downstream data.
+Overall, we found that these choices can have a profound impact on the way that a microbial community is represented in the sequence data.
+Although the effects are relatively small compared to the variation in human stool samples, care should be taken to use polymerases with the highest possible fidelity and to minimize the number of rounds of PCR.
 
@@
 
-## Figure Lengends
-
 **Figure 1.
 The error rate of assembled sequence reads increases with the number of rounds of PCR used and follows the relative error rates provided by the manufacturers; however, much of this error is mediated by denoising.**
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/mock_error.pdf}
+
+\newpage
 
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/chimera_plots.pdf}
+
+\newpage
+
 **Figure 3.
@@
 The percentage of bases that were guanines or cytosines within the V4 region of the 16S rRNA genes in each organism is indicated by the number in the lower left corner of each panel.
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/species_bias.pdf}
+
+\newpage
 
@@
 
+\includegraphics[width=0.8\columnwidth]{../results/figures/mock_community.pdf}
+
+\newpage
+
 **Figure 5.
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/stool_community.pdf}
+
+\newpage
+
 **Figure 6.
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/drift.pdf}
+
+\newpage
+
 **Figure S1: With the exception of the seuqence data generated using the Kappa polymerase, the ratio of the two *Salmonella enterica* V4 sequences to a value was lower than the expected ratio of 6:1.** The two *S.
 enterica* V4 sequences differed by a single base.
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/salmonella_bias.pdf}

### 2019-01-15 submission/manuscript.tex
Fresh make write.paper

@@
 
-\hypertarget{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}{%
-\section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
-High Fidelity DNA
-Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}}
-
-\hypertarget{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}{%
-\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
-errors when sequencing 16S rRNA genes from complex
-communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}}
-
+\vspace{25mm}
 \begin{center}
+
+\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}
+
+\vspace{50mm}
+
+\textbf{Running title:} Quantifying the choice of DNA polymerase
+
 \vspace{25mm}
@@
 \end{center}
-
-Co-author e-mails:
-
-\begin{itemize}
-\tightlist
-\item
-  \href{mailto:marcsze@med.umich.edu}{\nolinkurl{marcsze@med.umich.edu}}
-\end{itemize}
 
@@
 \subsection{Abstract}\label{abstract}}
+
+PCR amplification of 16S rRNA genes is a critical, yet under appreciated
+step in the generation of sequence data to describe the taxonomic
+composition of microbial communities. Numerous factors in the design of
+PCR can impact the sequencing error rate, the abundance of chimeric
+sequences, and the degree to which the fragments in the product
+represent their abundance in the original sample. The choice of PCR
+cycling conditions are known to impact the outcome of PCR and downstream
+inferences. We compared the performance of high fidelity polymerases and
+varying number of rounds of amplification when amplifying a mock
+community with known sequences and community structure and human stool
+samples. Although it is impossible to derive specific recommendations,
+we did observe general trends. Namely, using a polymerase with the
+highest possible fidelity and minimizing the number of rounds of PCR
+reduced the sequencing error rate, fraction of chimeric sequences, and
+bias. Evidence of bias at the sequence level was subtle and could not be
+ascribed to the fragments' fraction of bases that were guanines or
+cytosines. When analyzing the mock community data, the amount that the
+community deviated from the expected composition increased with rounds
+of PCR. The effect was inconsistent for human stool samples. Overall the
+results underscore the difficulty of comparing sequence data that are
+generated by different PCR protocols; however, the results indicate that
+the variation in human stool samples is generally larger than that
+introduced by the choice of polymerase or number of rounds of PCR.
+
+\hypertarget{importance}{%
+\subsubsection{Importance}\label{importance}}
+
+The growth in interest in characterizing microbial communities from
+diverse communities has exploded with the steep decline in sequencing
+costs. A significant amount of effort has gone into understanding the
+error profiles of DNA sequencers; however, little work has been done to
+understand the effects of choices that researchers make to generate the
+PCR amplicons that are sequenced. Here we quantified the effects of the
+choice of polymerase and number of PCR cycles on the quality of
+downstream data. Overall, we found that these choices can have a
+profound impact on the way that a microbial community is represented in
+the sequence data. Although the effects are relatively small compared to
+the variation in human stool samples, care should be taken to use
+polymerases with the highest possible fidelity and to minimize the
+number of rounds of PCR.
 
@@
 also removed any contigs that were at least three bases more similar to
-a chimera of two references than to a single reference sequence
-{[}ref{]}. Regardless of the polymerase, the error rate increased with
-the number of rounds of amplification. Using 30 rounds of PCR is a
-common approach across diverse types of samples. Among the data
+a chimera of two references than to a single reference sequence (18, 19,
+38). Regardless of the polymerase, the error rate increased with the
+number of rounds of amplification (Figure 1). Using 30 rounds of PCR is
+a common approach across diverse types of samples. Among the data
 generated using 30 rounds of PCR the Accuprime polymerase had the
@@
 expected from previous studies, the number of chimeras increased with
-rounds of amplification. Interestingly, the fraction of chimeric
-sequences from the mock community varied by the type of polymerase used.
-After 30 rounds of PCR, the Platinum polymerase had the highest chimera
-rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
-(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
+rounds of amplification (Figure 2A). Interestingly, the fraction of
+chimeric sequences from the mock community varied by the type of
+polymerase used. After 30 rounds of PCR, the Platinum polymerase had the
+highest chimera rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%),
+Phusion (i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
 polymerases. We used the UChime algorithm to detect chimeras in our
 pre-clustered mock community data and calculate the algorithm's
-sensitivity and specificity. For all polymerases, the specificity was
-above 95.8\% and showed a weak association with the number of cycles
-used. There was considerable inter-polymerase and inter-round of
-amplification variation in the sensitivity of UChime to detect the
-chimeras from the mock community. This suggested that the residual error
-rate after pre-clustering the sequence data did not compromise the
-sensitivity. The sensitivity of UChime varied between 50 and 84.8\%. The
-generalizability of these results is limited because we used a single
-mock community with limited genetic diversity. Although we did not know
-the true chimera rate for our four human stool samples, we were able to
-calculate the fraction of sequences that UChime identified as being
-chimeric. These results followed those from the mock communities;
-additional rounds of amplification significantly increased the rate of
-chimeras and there was variation between the polymerases that we used.
-Although it was not possible to identify the features of a polymerase
-that resulted in higher rates of chimeras, it is clear that using the
-smallest number of PCR cycles possible will minimize the impacts of
-chimeras.
+sensitivity and specificity (Figure 2A). For all polymerases, the
+specificity was above 95.7\% and showed a weak association with the
+number of cycles used. There was considerable inter-polymerase and
+inter-round of amplification variation in the sensitivity of UChime to
+detect the chimeras from the mock community. This suggested that the
+residual error rate after pre-clustering the sequence data did not
+compromise the sensitivity. The sensitivity of UChime varied between 50
+and 87.0\%. The generalizability of these results is limited because we
+used a single mock community with limited genetic diversity. Although we
+did not know the true chimera rate for our four human stool samples, we
+were able to calculate the fraction of sequences that UChime identified
+as being chimeric (Figure 2B). These results followed those from the
+mock communities; additional rounds of amplification significantly
+increased the rate of chimeras and there was variation between the
+polymerases that we used. Although it was not possible to identify the
+features of a polymerase that resulted in higher rates of chimeras, it
+is clear that using the smallest number of PCR cycles possible will
+minimize the impacts of chimeras.
 
@@
 of the 16S rRNA genes from each bacterium in the mock community to the
-data we generated across rounds of amplification and polymerase.
-Interestingly, for some bacteria, their representation became less
+data we generated across rounds of amplification and polymerase (Figure
+3). Interestingly, for some bacteria, their representation became less
 biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
@@
 data to calculate the ratio of the dominant to rare variants from
-\emph{S. enterica} (Figure XXX). The Accuprime, Phusion, Platinum, and
-Q5 polymerases converged to a ratio of 5.4; however, the ratio for the
+\emph{S. enterica} (Figure S1). The Accuprime, Phusion, Platinum, and Q5
+polymerases converged to a ratio of 5.4; however, the ratio for the
 Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
@@
 diversity of the mock community data at different phases of the sequence
-curation pipeline. First, we removed the chimeras from the mock
-community data as described above and mapped the individual reads to the
-OTUs that the 16S rRNA gene fragments would cluster into if there were
-no sequencing errors. This gave us a community distribution that
+curation pipeline (Figure 4). First, we removed the chimeras from the
+mock community data as described above and mapped the individual reads
+to the OTUs that the 16S rRNA gene fragments would cluster into if there
+were no sequencing errors. This gave us a community distribution that
 reflected the distribution following PCR without any artifacts. Alpha
@@
 polymerases except for the Kappa polymerase, where the richness remained
-constant but the diversity decreased. These data suggest that PCR had
-the effect of making the community distribution more even than it was
-originally. Next, we used the true sequence diversity, but removed
-chimeras based on mapping reads to all possible chimeras between
+constant but the diversity decreased (Figure 4A). These data suggest
+that PCR had the effect of making the community distribution more even
+than it was originally. Next, we used the true sequence diversity, but
+removed chimeras based on mapping reads to all possible chimeras between
 reference sequences, and clustered the reads to OTUs. The same trends
 were observed as with perfect sequence data except that the richness and
-diversity metrics trended higher. Finally, we used the observed sequence
-data and used the UCHIME algorithm to identify chimeras. Again, we
-observed the same trends as with the other analyses except the richness
-and diversity metrics trended higher. These comparisons demonstrated
-that although the bias at the sequence level was subtle, PCR does have a
-bias that is exacerbated by errors and chimeras. When we measured the
-Bray-Curtis distance between the communities observed after 25 rounds of
-amplification and those at 30 and 35 rounds of amplification the
-distances between 25 and 35 rounds were higher than between 25 and 30
-rounds for each of the polymerases by an average of 0.023 units. The
-Platinum polymerase varied the most across rounds of amplification (25
-vs 30 rounds: 0.13; 25 vs 35 rounds: 0.16). Although the distances
-between samples were small, the ordination of these distances showed a
-clear change in community structure with increasing rounds of PCR. This
-observation was supported by our statistical analysis, which revealed
-that the number of rounds of PCR (R\textsuperscript{2}=0.20,
-P\textless{}0.001) was a larger factor than the choice of polymerase
+diversity metrics trended higher (Figure 4A). Finally, we used the
+observed sequence data and used the UCHIME algorithm to identify
+chimeras. Again, we observed the same trends as with the other analyses
+except the richness and diversity metrics trended higher (Figure 4A).
+These comparisons demonstrated that although the bias at the sequence
+level was subtle, PCR does have a bias that is exacerbated by errors and
+chimeras. When we measured the Bray-Curtis distance between the
+communities observed after 25 rounds of amplification and those at 30
+and 35 rounds of amplification the distances between 25 and 35 rounds
+were higher than between 25 and 30 rounds for each of the polymerases by
+an average of 0.022 units (Figure 4B). The Platinum polymerase varied
+the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
+rounds: 0.15). Although the distances between samples were small, the
+ordination of these distances showed a clear change in community
+structure with increasing rounds of PCR (Figure 4C). This observation
+was supported by our statistical analysis, which revealed that the
+number of rounds of PCR (R\textsuperscript{2}=0.21, P\textless{}0.001)
+was a larger factor than the choice of polymerase
 (R\textsuperscript{2}=0.20, P\textless{}0.001). These results
@@
 Shannon diversity, inverse Simpson diversity index for each condition
-and donor. Although there were clear differences between conditions, the
-relative ordering of the stool samples did not meaningfully vary across
-conditions. When we characterized the variation between rounds of
-amplification using human stool samples, the distance between the 25 and
-30 rounds and 25 and 35 rounds varied considerably between samples and
-polymerases. In general the inter-round variation was lowest for the
-data generated using the Kappa and Accuprime polymerases. The data
-generated using the Platinum polymerase was consistent across rounds,
-but it was more biased than these polymerases. Considering the average
-distance across the four samples varied between 0.39 and 0.56,
-regardless of the polymerases and number of rounds of amplification, any
-bias due to amplification is unlikely to obscure community-wide
-differences between samples. In support of this was our principle
-coordinates analysis of the Bray-Curtis distances, which revealed
-distinct clusters by donor. Within each cluster there were no obvious
-patterns related to the polymerase or number of rounds of PCR. However,
-our statistical analysis using adonis revealed statistically significant
-differences in the community structures with the subject explaining the
-most variation (R\textsuperscript{2}=0.79, P\textless{}0.001), followed
-by the number of rounds of PCR (R\textsuperscript{2}=0.044,
-P\textless{}0.001) and the choice of polymerase
-(R\textsuperscript{2}=0.033, P\textless{}0.001). Together, these results
-indicate that for a coarse analysis of communities, the choice of number
-of rounds of amplification or polymerase are not important, but that
-they must be consistent across samples. It is difficult to develop a
-specific recommendation based on the level of bias across rounds of PCR
-or polymerases; however, the general suggestion is to use as few rounds
-of amplification as possible.
+and donor (Figure 5A). Although there were clear differences between
+conditions, the relative ordering of the stool samples did not
+meaningfully vary across conditions. When we characterized the variation
+between rounds of amplification using human stool samples, the distance
+between the 25 and 30 rounds and 25 and 35 rounds varied considerably
+between samples and polymerases (Figure 5B). In general the inter-round
+variation was lowest for the data generated using the Kappa and
+Accuprime polymerases. The data generated using the Platinum polymerase
+was consistent across rounds, but it was more biased than these
+polymerases. Considering the average distance across the four samples
+varied between 0.39 and 0.56, regardless of the polymerases and number
+of rounds of amplification, any bias due to amplification is unlikely to
+obscure community-wide differences between samples. In support of this
+was our principle coordinates analysis of the Bray-Curtis distances,
+which revealed distinct clusters by donor (Figure 5C). Within each
+cluster there were no obvious patterns related to the polymerase or
+number of rounds of PCR. However, our statistical analysis using adonis
+revealed statistically significant differences in the community
+structures with the subject explaining the most variation
+(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
+of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
+choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
+Together, these results indicate that for a coarse analysis of
+communities, the choice of number of rounds of amplification or
+polymerase are not important, but that they must be consistent across
+samples. It is difficult to develop a specific recommendation based on
+the level of bias across rounds of PCR or polymerases; however, the
+general suggestion is to use as few rounds of amplification as possible.
 
@@
 Bray-Curtis distance between replicate reactions using the same
-polymerase and number of rounds of amplification. Using the mock
-community data there were no obvious trends. The average Bray-Curtis
-distance within a set of conditions varied by 0.063 to 0.11 units.
-Although we did not obtain replicates of each of the stool samples, the
-intra-sample variation for each set of conditions was consistent and
-varied between 0.50 and 0.56 units. These data suggest that amplicon
-sequencing is robust to random variation in amplification and that any
-differences are likely to be smaller than what is considered
-biologically relevant.
+polymerase and number of rounds of amplification (Figure 6). Using the
+mock community data there were no obvious trends. The average
+Bray-Curtis distance within a set of conditions varied by 0.063 to 0.11
+units. Although we did not obtain replicates of each of the stool
+samples, the intra-sample variation for each set of conditions was
+consistent and varied between 0.50 and 0.56 units. These data suggest
+that amplicon sequencing is robust to random variation in amplification
+and that any differences are likely to be smaller than what is
+considered biologically relevant.
 
@@
 The observed sequencing error rates and alpha diversity metrics followed
-the manufacturers' measurements of their polymerases' fidelity.
-Accuprime and Platinum have fidelity that are approximately 10-times
+the manufacturers' measurements of their polymerases' fidelity (Figure
+1). Accuprime and Platinum have fidelity that are approximately 10-times
 higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
@@
 environmental and reagent contaminants can also have a significant
-impact on the analysis of low biomass samples (38). Less well understood
+impact on the analysis of low biomass samples (39). Less well understood
 is the effect of phylogenetic diversity on bias and chimera formation.
@@
 require additional rounds of PCR. A host of bioinformatics tools are
-available for removing residual sequencing errors (18, 39--41). These
+available for removing residual sequencing errors (18, 40--42). These
 tools struggle to correctly differentiate true biological diversity
@@
 little evaluation of the effects on the specificity of the algorithms
-(39, 41). Others recommend removing any read that has an abundance below
+(40, 42). Others recommend removing any read that has an abundance below
 a specified threshold (e.g.~removing all sequences that only appear
-once) (20, 39--41). This method must be approached with caution as such
+once) (20, 40--42). This method must be approached with caution as such
 approaches are likely to introduce a different bias of the community
@@
 Ultimately, researchers must test their hypotheses with multiple methods
-to validate the claims they reach with any one method (42). All methods
+to validate the claims they reach with any one method (43). All methods
 have biases and limitations and we must use complementary methods to
@@
 \textbf{\emph{Human samples.}} Fecal samples were obtained from 4
-individuals who were part of an earlier study (43). These samples were
+individuals who were part of an earlier study (44). These samples were
 collected using a protocol approved by the University of Michigan
@@
 \textbf{\emph{Sequence processing.}} The mothur software program (v
-1.41) was used for all sequence processing steps (44). The protocol has
+1.41) was used for all sequence processing steps (45). The protocol has
 been previously published (18)
@@
 abundant sequence. The VSEARCH implementation of UCHIME was used to
-screen for chimeras (35, 45). At various stages in the sequence
+screen for chimeras (35, 46). At various stages in the sequence
 processing pipeline for the mock community data, the mothur seq.error
@@
 Sequences were assigned to operational taxonomic units (OTUs) at a
-threshold of 3\% dissimilarity using the OptiClust algorithm (46). To
+threshold of 3\% dissimilarity using the OptiClust algorithm (47). To
 adjust for unequal sequencing when measuring alpha and beta diversity,
@@
 \textbf{\emph{Statistical analysis.}} All analysis was done with the R
-(v 3.5.0) software package (47). Data transformation and graphing were
-completed using the tidyverse package (v 1.2.1) (48). The distance
+(v 3.5.0) software package (48). Data transformation and graphing were
+completed using the tidyverse package (v 1.2.1) (49). The distance
 matrix data was analyzed using the adonis function within the vegan
-package (v 2.5.3) (49).
+package (v 2.5.3) (50).
 
@@
 
+\leavevmode\hypertarget{ref-Quince2011}{}%
+38. \textbf{Quince C}, \textbf{Lanzen A}, \textbf{Davenport RJ},
+\textbf{Turnbaugh PJ}. 2011. Removing noise from pyrosequenced
+amplicons. BMC Bioinformatics \textbf{12}:38.
+doi:\href{https://doi.org/10.1186/1471-2105-12-38}{10.1186/1471-2105-12-38}.
+
 \leavevmode\hypertarget{ref-Salter2014}{}%
-38. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
+39. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
 \textbf{Calus ST}, \textbf{Cookson WO}, \textbf{Moffatt MF},
@@
 \leavevmode\hypertarget{ref-Callahan2016}{}%
-39. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
+40. \textbf{Callahan BJ}, \textbf{McMurdie PJ}, \textbf{Rosen MJ},
 \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes SP}. 2016. DADA2:
@@
 \leavevmode\hypertarget{ref-Edgar2016}{}%
-40. \textbf{Edgar RC}. 2016. UNOISE2: Improved error-correction for
+41. \textbf{Edgar RC}. 2016. UNOISE2: Improved error-correction for
 illumina 16S and ITS amplicon sequencing.
@@
 \leavevmode\hypertarget{ref-Amir2017a}{}%
-41. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
+42. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
 \textbf{Kopylova E}, \textbf{Morton JT}, \textbf{Xu ZZ},
@@
 \leavevmode\hypertarget{ref-Schloss2018}{}%
-42. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
+43. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
 reproducibility, replicability, robustness, and generalizability in
@@
 \leavevmode\hypertarget{ref-Seekatz2016}{}%
-43. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+44. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
 \textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-Rognes2016}{}%
-45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \leavevmode\hypertarget{ref-Westcott2017}{}%
-46. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+47. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 improved method for assigning amplicon-based sequence data to
@@
 \leavevmode\hypertarget{ref-r_citation_2018}{}%
-47. \textbf{R Core Team}. 2018. R: A language and environment for
+48. \textbf{R Core Team}. 2018. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \leavevmode\hypertarget{ref-Tidyverse2017}{}%
-48. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
+49. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
 'tidyverse'.
@@
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-49. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
+50. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
 \textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
@@
 
-\hypertarget{figure-lengends}{%
-\subsection{Figure Lengends}\label{figure-lengends}}
-
 \textbf{Figure 1. The error rate of assembled sequence reads increases
@@
 mediated by denoising.}
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/mock_error.pdf}
+
+\newpage
 
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/chimera_plots.pdf}
+
+\newpage
+
 \textbf{Figure 3. The relative abundances of reads mapped to reference
@@
 corner of each panel.
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/species_bias.pdf}
+
+\newpage
 
@@
 
+\includegraphics[width=0.8\columnwidth]{../results/figures/mock_community.pdf}
+
+\newpage
+
 \textbf{Figure 5. Sequencing of human stool samples indicated clear
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/stool_community.pdf}
+
+\newpage
+
 \textbf{Figure 6. There is little evidence for PCR drift adversely
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/drift.pdf}
+
+\newpage
+
 \textbf{Figure S1: With the exception of the seuqence data generated
@@
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/salmonella_bias.pdf}
+
 \end{document}

### 2019-01-16 submission/manuscript.Rmd
Pat's edits

--- 2019-01-15 submission/manuscript.Rmd
@@
 
-# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase
+\vspace{25mm}
+\begin{center}
 
-# Influence of DNA Polymerase on PCR bias, chimera formation, and errors when sequencing 16S rRNA genes from complex communities
+\textbf{\huge Quantifying the impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}
 
-\begin{center}
+\vspace{50mm}
+
+\textbf{Running title:} Quantifying the effects of PCR
+
 \vspace{25mm}
@@
 ## Abstract
+
+PCR amplification of 16S rRNA genes is a critical, yet under appreciated step in the generation of sequence data to describe the taxonomic composition of microbial communities.
+Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample.
+The choice of PCR cycling conditions are known to impact the outcome of PCR and downstream inferences.
+We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community with known sequences and community structure and human stool samples.
+Although it was impossible to derive specific recommendations, we did observe general trends.
+Namely, using a polymerase with the highest possible fidelity and minimizing the number of rounds of PCR reduced the sequencing error rate, fraction of chimeric sequences, and bias.
+Evidence of bias at the sequence level was subtle and could not be ascribed to the fragments' fraction of bases that were guanines or cytosines.
+When analyzing mock community data, the amount that the community deviated from the expected composition increased with rounds of PCR.
+The effect was inconsistent for human stool samples.
+Overall the results underscore the difficulty of comparing sequence data that are generated by different PCR protocols.
+However, the results indicate that the variation in human stool samples is generally larger than that introduced by the choice of polymerase or number of rounds of PCR.
+
+### Importance
+
+A steep decline in sequencing costs drove an explosion in studies characterizing microbial communities from diverse environments.
+Although a significant amount of effort has gone into understanding the error profiles of DNA sequencers, little has been done to understand the downstream effects of the PCR amplification protocol.
+We quantified the effects of the choice of polymerase and number of PCR cycles on the quality of downstream data.
+We found that these choices can have a profound impact on the way that a microbial community is represented in the sequence data.
+The effects are relatively small compared to the variation in human stool samples, however, care should be taken to use polymerases with the highest possible fidelity and to minimize the number of rounds of PCR.
+These results underscore the difficulty of directly comparing the sequence data generated using different PCR protocols or interpreting results in an absolute sense.
 
@@
 Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017].
-The increased sampling depth that is now available relative to previous Sanger-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.
+The increased sampling depth that is now available relative to previous Sanger sequencing-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.
 
@@
 Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011].
+Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017].
 Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015]; however, these have been developed for idealized situations.
-Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017].
 To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift [@Kennedy2014].
@@
 A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@Gohl2016].
-The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@Gohl2016].
+The authors of the study reduced the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@Gohl2016].
 Yet this optimization was specific for the community they were analyzing (i.e.
 captive and semi-captive red-shanked doucs) and assumed that if the two polymerases generate the same community structure that the community structure was correct.
-In fact, the community structure generated by both methods was not free of artifacts, but has the same artifacts.
-A challenge in these types of experiments is having a priori knowledge of the true community representation.
+In fact, the community structure generated by both methods was not free of artifacts, but likely had the same artifacts.
+A challenge in these types of experiments is having *a priori* knowledge of the true community representation.
 A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias [@Schloss2011]; however, mock communities have a limited phylogenetic diversity relative to natural communities.
@@
 They can be used to validate results from mock communities and to understand the relative impacts of artifacts on the ability to differentiate biological and methodological sources of variation.
-Given the large number of DNA polymerases available to researchers it is unlikely that a specific recommendation is possible.
-Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are desired.
+Given the large number of DNA polymerases available to researchers, it is unlikely that a specific recommendation is possible.
+Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are needed.
 
@@
 We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform.
-This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol.
-To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 15, 20, 25, 30, and 35 rounds of amplification (Table 1).
+This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol [@Kozich].
+To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 20, 25, 30, and 35 rounds of amplification.
 Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.
@@
 `r filter(error, polymerase=="Q5") %>% select(contig_error)`%) polymerases (Figure 1).
-When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e.
+When we applied a pre-clustering denoising step, which merged the counts of reads within 2 nt of a more abundant sequence [@Schloss2011], the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e.
 `r filter(error, polymerase=="PL") %>% select(pc_error)`%) followed by the Accuprime (i.e.
@@
 `r filter(error, polymerase=="K") %>% select(pc_error)`%) polymerases (Figure 1).
-Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.
+Although specific recommendations are difficult to make because the phylogenetic diversity of the initial DNA template is likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.
 
@@
 ***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses.
-Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors.
-We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence.
-As expected from previous studies, the number of chimeras increased with rounds of amplification (Figure 2A).
+Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors [@Edgar2011; @Haas2011].
+Because the true sequences of the organisms in the mock community were known, we could generate all possible chimeras between pairs of V4 16S rRNA gene fragments and use these possible chimeric sequences to screen the sequences generated under the different PCR conditions for chimeras.
+The number of chimeras increased with rounds of amplification (Figure 2A).
 Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used.
@@
 `r filter(chimera, polymerase=="ACC") %>% select(pc_frac_chimera)`%) polymerases.
-We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity (Figure 2A).
-For all polymerases, the specificity was above `r min_specificity`% and showed a weak association with the number of cycles used.
-There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community.
+Because our chimera screening procedure can only be applied to mock communities, we used the UCHIME algorithm to model the chimera screening approach that is used in most sequence curation pipelines.
+By comparing the output of UCHIME to our reference-based approach, we were able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
+The specificity for all polymerases was above `r min_specificity`% and showed a weak association with the number of cycles used.
+There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UCHIME to detect the chimeras from the mock community.
 This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
-The sensitivity of UChime varied between `r sensitivity["min"]` and `r sensitivity["max"]`%.
+The sensitivity of UCHIME varied between `r sensitivity["min"]` and `r sensitivity["max"]`%.
 The generalizability of these results is limited because we used a single mock community with limited genetic diversity.
-Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric  (Figure 2B).
-These results followed those from the mock communities; additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
+Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UCHIME identified as being chimeric  (Figure 2B).
+These results followed those from the mock communities: additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
 Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
@@
 
-***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses.
-The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria.
+***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to amplify 16S rRNA gene fragments there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998].
+The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria because each bacterium had a different genome size and varied in the number of operons in its genome.
 The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community.
@@
 subtilis*).
-The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
+Contrary to prior reports [@Polz1998], The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
 Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed.
@@
 The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of `r others`; however, the ratio for the Kappa polymerase varied between `r kappa["min"]` and `r kappa["max"]` for 25 to 35 rounds of PCR.
-Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias and the previous observation regarding the guanine and cytosine relative abundance was not observed with our data.
+Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias.
 
@@
 
-***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline (Figure 4).
+***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the variation in bias between polymerases and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline (Figure 4).
 First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors.
 This gave us a community distribution that reflected the distribution following PCR without any artifacts.
-Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased (Figure 4A).
-These data suggest that PCR had the effect of making the community distribution more even than it was originally.
-Next, we used the true sequence diversity, but removed chimeras based on mapping reads to all possible chimeras between reference sequences, and clustered the reads to OTUs.
-The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher (Figure 4A).
+Although the richness did not change, the Shannon diversity increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, for which the diversity decreased (Figure 4A).
+These data suggest that PCR had the effect of making the community distribution more even than it was originally, except for the data generated using the Kappa polymerase where the evenness decreased.
+Next, we used the observed sequence errors, but removed chimeras using our reference-based approach, and clustered the reads to OTUs.
+The richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A).
 Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras.
-Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher (Figure 4A).
-These comparisons demonstrated that although the bias at the sequence level was subtle, PCR does have a bias that is exacerbated by errors and chimeras.
+Again, the richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A).
+These comparisons demonstrated that although the bias at the sequence level was subtle, PCR introduces bias that is exacerbated by errors and chimeras when sequences are clustered into OTUs.
 When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B).
@@
 To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
-We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor (Figure 5A).
+We calculated the number of observed OTUs and Shannon diversity for each condition and subject (Figure 5A).
 Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
@@
 In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases.
-The data generated using the Platinum polymerase was consistent across rounds, but it was more biased than these polymerases.
+The data generated using the Platinum polymerase was consistent across rounds, but overall, it was more biased than the other polymerases.
 Considering the average distance across the four samples varied between `r inter_diversity["min"]` and `r inter_diversity["max"]`, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
@@
 Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
-However, our statistical analysis using adonis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=`r filter(stool_vegan, effects=="subject") %>% select(r_sq)`, P<0.001), followed by the number of rounds of PCR (R^2^=`r filter(stool_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) and the choice of polymerase (R^2^=`r filter(stool_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
+However, our statistical analysis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=`r filter(stool_vegan, effects=="subject") %>% select(r_sq)`, P<0.001), followed by the number of rounds of PCR (R^2^=`r filter(stool_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) and the choice of polymerase (R^2^=`r filter(stool_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
 Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
@@
 
-***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift.
+***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift over the course of PCR.
 To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification (Figure 6).
@@
 The average Bray-Curtis distance within a set of conditions varied by `r mock_drift["min"]` to `r mock_drift["max"]` units.
-Although we did not obtain replicates of each of the stool samples, the intra-sample variation for each set of conditions was consistent and varied between `r stool_drift["min"]` and `r stool_drift["max"]` units.
+Although we did not generate technical replicates of each of the stool samples, the inter-sample variation for each set of conditions was consistent and varied between `r stool_drift["min"]` and `r stool_drift["max"]` units.
 These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.
@@
 Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between individuals.
-Based on these observations, within a study, amplicons must be generated by consistent protocols to yield meaningful comparisons.
+Based on these observations, amplicons must be generated by consistent protocols to yield meaningful comparisons.
 When comparing across studies, values like richness, diversity, and relative abundances must be made in relative and not absolute terms.
@@
 Amplicon Sequencing Variants or ASVs) and PCR or sequencing errors.
-Although these methods remove residual errors, they also risk splitting intragenomic variants into separate ASVs, merging 16S rRNA gene sequences from different taxa into the same ASV.
+Although these methods remove residual errors, they also risk splitting intragenomic variants into separate ASVs or merging 16S rRNA gene sequences from different taxa into the same ASV.
 Other tools are available for removing chimeras [@Edgar2011; @Haas2011] where there is a trade off between the sensitivity of detecting chimeras and the specificity of correctly calling a sequence a chimera.
 In recent years, parameters for these algorithms have been changed to increase their sensitivity with little evaluation of the effects on the specificity of the algorithms [@Callahan2016; @Amir2017a].
-Others recommend removing any read that has an abundance below a specified threshold (e.g.
+Others recommend removing any read that has an abundance below a specified threshold as a tool to remove PCR and sequencign artifacts (e.g.
 removing all sequences that only appear once) [@Bokulich2012; @Callahan2016; @Edgar2016; @Amir2017a].
@@
 All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2018].
-Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2017].
-The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`) [@Vegan2018].
+Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`).
+The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`).
 
@@
 
-\newpage
-
 ## Acknowledgements
 
-The authors thank the study participants in ERIN whose samples were utilized.
-We also would like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
+We appreciate the willingness of the donors to provide stool samples.
+We also thank Judy Opp and April Cockburn assistance in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
 Additional thanks to members of the Schloss lab and Dr.
@@
 
-## Figure Lengends
-
 **Figure 1.
 The error rate of assembled sequence reads increases with the number of rounds of PCR used and follows the relative error rates provided by the manufacturers; however, much of this error is mediated by denoising.**
+
+\includegraphics{../results/figures/mock_error.pdf}
+
+\newpage
 
@@
 (B) Sequencing of four human stool samples after using one of five different polymerases again demonstrated increased rate of chimera formation with increasing number of rounds of PCR and variation across polymerases.
+
+\includegraphics[width=1.0\columnwidth]{../results/figures/chimera_plots.pdf}
+
+\newpage
 
@@
 
+\includegraphics[width=1.0\columnwidth]{../results/figures/species_bias.pdf}
+
+\newpage
+
 **Figure 4.
-Despite evidence of subtle PCR bias at the genome level, there was significant evidence of bias using community-wide metrics that grew with the number of rounds of PCR when using a mock community.** (A) With the exception of data collected using the Kappa polymerase, the OTU richness and Shannon diversity values increased with number of rounds of PCR and the inclusion of residual sequencing errors and chimeras.
-The horizontal black line indicates the expected richness and diversity if there were no bias, errors, or chimeras.
-(B) Relative to the mock community sampled after 25 rounds of PCR, the Bray-Curtis distance to the communities sampled after 30 and 35 rounds of PCR increased for all polymerases.
-(C) The variation between samples collected after 20, 25, 30, and 35 rounds of PCR for the five polymerases demonstrated a significant change in the community driven by the number of rounds of PCR and the polymerase used.
+Despite evidence of subtle PCR bias at the genome level, there was significant evidence of bias using community-wide metrics that grew with the number of rounds of PCR when using a mock community.** (A) With the exception of the Kappa polymerase data, the richness and Shannon diversity values increased with number of rounds of PCR and the inclusion of residual sequencing errors and chimeras.
+The horizontal black line indicates the ideal richness and diversity.
+(B) Relative to the mock community sampled after 25 rounds of PCR, the distance to the communities sampled after 30 and 35 rounds of PCR increased for all polymerases.
+(C) The variation between samples demonstrated a significant change in the community driven by the number of rounds of PCR and the polymerase used.
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/mock_community.pdf}
+
+\newpage
 
 **Figure 5.
-Sequencing of human stool samples indicated clear increase in bias with number of rounds of PCR, however, the bias appeared to be consistent within each sample.** (A) With the exception of data collected using the Kappa polymerase, the OTU richness and Shannon diversity values increased with number of rounds of PCR.
-(B) Relative to the stool communities sampled after 25 rounds of PCR, the Bray-Curtis distance to the stool communities sampled after 30 and 35 rounds of PCR was inconsistent and there was little difference in variation for data collected using the Kappa polymerase.
+Sequencing of human stool samples indicated clear increase in bias with number of rounds of PCR, however, the bias appeared to be consistent within each sample.** (A) With the exception of data collected using the Kappa polymerase, the richness and Shannon diversity values increased with number of rounds of PCR.
+(B) Relative to the stool communities sampled after 25 rounds of PCR, the distance to the stool communities sampled after 30 and 35 rounds of PCR was inconsistent and there was little difference in variation for data collected using the Kappa polymerase.
 (C) The variation between stool samples was larger than the amount of variation introduced by varying the number of rounds of PCR or polymerase.
 
+\includegraphics[width=0.9\columnwidth]{../results/figures/stool_community.pdf}
+
+\newpage
+
 **Figure 6.
-There is little evidence for PCR drift adversely impacting the results of amplicon studies.** The average distance between replicates of sequencing the same mock community or between the the human stool samples did not vary by number of rounds of PCR or by polymerase.
+The average distance between replicates of sequencing the same mock community or between the the human stool samples (i.e.
+drift) did not vary by number of rounds of PCR or by polymerase.**
+
+\includegraphics[width=1.0\columnwidth]{../results/figures/drift.pdf}
+
+\newpage
 
@@
 enterica* V4 sequences differed by a single base.
+
+\includegraphics[width=1.0\columnwidth]{../results/figures/salmonella_bias.pdf}

### 2019-01-16 submission/manuscript.tex
Pat's edits

@@
 
-\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}
+\textbf{\huge Quantifying the impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}
 
@@
 
-\textbf{Running title:} Quantifying the choice of DNA polymerase
+\textbf{Running title:} Quantifying the effects of PCR
 
@@
 community with known sequences and community structure and human stool
-samples. Although it is impossible to derive specific recommendations,
+samples. Although it was impossible to derive specific recommendations,
 we did observe general trends. Namely, using a polymerase with the
@@
 ascribed to the fragments' fraction of bases that were guanines or
-cytosines. When analyzing the mock community data, the amount that the
+cytosines. When analyzing mock community data, the amount that the
 community deviated from the expected composition increased with rounds
@@
 results underscore the difficulty of comparing sequence data that are
-generated by different PCR protocols; however, the results indicate that
+generated by different PCR protocols. However, the results indicate that
 the variation in human stool samples is generally larger than that
@@
 
-The growth in interest in characterizing microbial communities from
-diverse communities has exploded with the steep decline in sequencing
-costs. A significant amount of effort has gone into understanding the
-error profiles of DNA sequencers; however, little work has been done to
-understand the effects of choices that researchers make to generate the
-PCR amplicons that are sequenced. Here we quantified the effects of the
-choice of polymerase and number of PCR cycles on the quality of
-downstream data. Overall, we found that these choices can have a
+A steep decline in sequencing costs drove an explosion in studies
+characterizing microbial communities from diverse environments. Although
+a significant amount of effort has gone into understanding the error
+profiles of DNA sequencers, little has been done to understand the
+downstream effects of the PCR amplification protocol. We quantified the
+effects of the choice of polymerase and number of PCR cycles on the
+quality of downstream data. We found that these choices can have a
 profound impact on the way that a microbial community is represented in
-the sequence data. Although the effects are relatively small compared to
-the variation in human stool samples, care should be taken to use
+the sequence data. The effects are relatively small compared to the
+variation in human stool samples, however, care should be taken to use
 polymerases with the highest possible fidelity and to minimize the
-number of rounds of PCR.
+number of rounds of PCR. These results underscore the difficulty of
+directly comparing the sequence data generated using different PCR
+protocols or interpreting results in an absolute sense.
 
@@
 (15--17), and analytical pipeline (15, 18--20). The increased sampling
-depth that is now available relative to previous Sanger-based methods is
-expected to compound the impacts of an investigator's choices and the
-interpretation of their results.
+depth that is now available relative to previous Sanger sequencing-based
+methods is expected to compound the impacts of an investigator's choices
+and the interpretation of their results.
 
@@
 identify chimeras, there are significant sensitivity and specificity
-tradeoffs (14, 35). Others have attempted to account for PCR bias using
-modeling approaches (29, 36); however, these have been developed for
-idealized situations. Laboratory-based solutions to minimize chimera
+tradeoffs (14, 35). Laboratory-based solutions to minimize chimera
 formation have also been proposed such as minimizing the amount of
@@
 minimizing the amount of shearing in the template DNA, and using DNA
-polymerases that have a proof-reading ability (14, 23). To minimize PCR
-drift, some investigators will pool technical replicate PCRs hoping to
-average out the drift (34). Other factors that have been shown to impact
-the formation of PCR artifacts are outside the control of an
+polymerases that have a proof-reading ability (14, 23). Others have
+attempted to account for PCR bias using modeling approaches (29, 36);
+however, these have been developed for idealized situations. To minimize
+PCR drift, some investigators will pool technical replicate PCRs hoping
+to average out the drift (34). Other factors that have been shown to
+impact the formation of PCR artifacts are outside the control of an
 investigator including the fraction of DNA bases that are guanines or
@@
 in the number of OTUs and chimeras between normal and high fidelity DNA
-polymerases (37). The authors of this study could reduce the difference
+polymerases (37). The authors of the study reduced the difference
 between two polymerases by optimizing the annealing and extension steps
@@
 the community structure generated by both methods was not free of
-artifacts, but has the same artifacts. A challenge in these types of
-experiments is having a priori knowledge of the true community
+artifacts, but likely had the same artifacts. A challenge in these types
+of experiments is having \emph{a priori} knowledge of the true community
 representation. A mock community with known composition allows
@@
 variation. Given the large number of DNA polymerases available to
-researchers it is unlikely that a specific recommendation is possible.
+researchers, it is unlikely that a specific recommendation is possible.
 Rather, the development of general best practices and understanding the
-impact of PCR artifacts on an analysis are desired.
+impact of PCR artifacts on an analysis are needed.
 
@@
 been shown to result in a relatively low sequencing error rate and is a
-widely used protocol. To better understand the impact of DNA polymerase
-choice on PCR artifacts, we selected five high fidelity DNA polymerases
-and amplified the communities using 15, 20, 25, 30, and 35 rounds of
-amplification (Table 1). Collectively, our results suggest that the
-number of rounds and to a lesser degree the choice of DNA polymerase
+widely used protocol ({\textbf{???}}). To better understand the impact
+of DNA polymerase choice on PCR artifacts, we selected five high
+fidelity DNA polymerases and amplified the communities using 20, 25, 30,
+and 35 rounds of amplification. Collectively, our results suggest that
+the number of rounds and to a lesser degree the choice of DNA polymerase
 used in PCR impact the sequence data, the effects are consistent and are
@@
 (i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
-(i.e.~0.060\%) polymerases (Figure 1). When we applied the
-pre.clustering denoising step, which merges the counts of reads within 2
-nt of a more abundant sequence, the error rates dropped considerably
-such that the Platinum polymerase had the highest error rate
-(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
-(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
-polymerases (Figure 1). Although specific recommendations are difficult
-to make because the impact of the actual community structure and
-concentration of the initial DNA template are likely to have an impact
-on the results, it is clear that using as few PCR cycles as necessary
-and a polymerase with the lowest possible error rate is a good guide to
-minimizing the impact of polymerase on the error rate.
+(i.e.~0.060\%) polymerases (Figure 1). When we applied a pre-clustering
+denoising step, which merged the counts of reads within 2 nt of a more
+abundant sequence (19), the error rates dropped considerably such that
+the Platinum polymerase had the highest error rate (i.e.~0.014\%)
+followed by the Accuprime (i.e.~0.012\%), Q5 (i.e.~0.0053\%), Phusion
+(i.e.~0.0049\%), and Kappa (i.e.~0.0049\%) polymerases (Figure 1).
+Although specific recommendations are difficult to make because the
+phylogenetic diversity of the initial DNA template is likely to have an
+impact on the results, it is clear that using as few PCR cycles as
+necessary and a polymerase with the lowest possible error rate is a good
+guide to minimizing the impact of polymerase on the error rate.
 
@@
 chimeric sequences with high specificity, their sensitivity is
-relatively low and can be reduced by the presence of sequencing errors.
-We identified those contigs that were at least three bases more similar
-to a chimera of two references than to a single reference sequence. As
-expected from previous studies, the number of chimeras increased with
-rounds of amplification (Figure 2A). Interestingly, the fraction of
-chimeric sequences from the mock community varied by the type of
-polymerase used. After 30 rounds of PCR, the Platinum polymerase had the
-highest chimera rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%),
-Phusion (i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
-polymerases. We used the UChime algorithm to detect chimeras in our
-pre-clustered mock community data and calculate the algorithm's
-sensitivity and specificity (Figure 2A). For all polymerases, the
-specificity was above 95.7\% and showed a weak association with the
-number of cycles used. There was considerable inter-polymerase and
-inter-round of amplification variation in the sensitivity of UChime to
-detect the chimeras from the mock community. This suggested that the
-residual error rate after pre-clustering the sequence data did not
-compromise the sensitivity. The sensitivity of UChime varied between 50
-and 87.0\%. The generalizability of these results is limited because we
-used a single mock community with limited genetic diversity. Although we
-did not know the true chimera rate for our four human stool samples, we
-were able to calculate the fraction of sequences that UChime identified
-as being chimeric (Figure 2B). These results followed those from the
-mock communities; additional rounds of amplification significantly
-increased the rate of chimeras and there was variation between the
-polymerases that we used. Although it was not possible to identify the
-features of a polymerase that resulted in higher rates of chimeras, it
-is clear that using the smallest number of PCR cycles possible will
-minimize the impacts of chimeras.
+relatively low and can be reduced by the presence of sequencing errors
+(14, 35). Because the true sequences of the organisms in the mock
+community were known, we could generate all possible chimeras between
+pairs of V4 16S rRNA gene fragments and use these possible chimeric
+sequences to screen the sequences generated under the different PCR
+conditions for chimeras. The number of chimeras increased with rounds of
+amplification (Figure 2A). Interestingly, the fraction of chimeric
+sequences from the mock community varied by the type of polymerase used.
+After 30 rounds of PCR, the Platinum polymerase had the highest chimera
+rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
+(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
+polymerases. Because our chimera screening procedure can only be applied
+to mock communities, we used the UCHIME algorithm to model the chimera
+screening approach that is used in most sequence curation pipelines. By
+comparing the output of UCHIME to our reference-based approach, we were
+able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
+The specificity for all polymerases was above 95.7\% and showed a weak
+association with the number of cycles used. There was considerable
+inter-polymerase and inter-round of amplification variation in the
+sensitivity of UCHIME to detect the chimeras from the mock community.
+This suggested that the residual error rate after pre-clustering the
+sequence data did not compromise the sensitivity. The sensitivity of
+UCHIME varied between 50 and 87.0\%. The generalizability of these
+results is limited because we used a single mock community with limited
+genetic diversity. Although we did not know the true chimera rate for
+our four human stool samples, we were able to calculate the fraction of
+sequences that UCHIME identified as being chimeric (Figure 2B). These
+results followed those from the mock communities: additional rounds of
+amplification significantly increased the rate of chimeras and there was
+variation between the polymerases that we used. Although it was not
+possible to identify the features of a polymerase that resulted in
+higher rates of chimeras, it is clear that using the smallest number of
+PCR cycles possible will minimize the impacts of chimeras.
 
 \textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
-Since researchers began using PCR to sequence 16S rRNA genes there has
-been concern that amplifying fragments from a mixed template pool could
-lead to a biased representation in the pool of products and would
-confound downstream analyses. The mock community was generated by mixing
-equal amounts of genomic DNA from each of the 8 bacteria resulting in
-uneven representation of the \emph{rrn} operons across the bacteria. The
-vendor of the mock community subjects each lot of genomic DNA to shotgun
-sequencing to more accurately quantify the actual abundance of each
-organism in the community. We compared the expected relative abundance
-of the 16S rRNA genes from each bacterium in the mock community to the
-data we generated across rounds of amplification and polymerase (Figure
-3). Interestingly, for some bacteria, their representation became less
-biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
-others became more biased (e.g.~\emph{E. faecalis}), and others had
-little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
-V4 region that were guanines or cytosines was not predictive of the
-amount of bias. Across the strains there was no variation in the length
-of their V4 regions and they each had the same sequence in the region
-that the primers annealed. One of the bacteria represented in the mock
-community, \emph{S. enterica}, had 6 identical copies of the V4 region
-and 1 copy that differed from those by one nucleotide. The dominant copy
-had a thymidine and the rare copy had a guanine. We used the sequence
-data to calculate the ratio of the dominant to rare variants from
-\emph{S. enterica} (Figure S1). The Accuprime, Phusion, Platinum, and Q5
-polymerases converged to a ratio of 5.4; however, the ratio for the
-Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
-Given the subtle nature of the variation in the relative abundances of
-each 16S rRNA gene fragment, it was not possible to create generalizable
-rules that would explain the bias and the previous observation regarding
-the guanine and cytosine relative abundance was not observed with our
-data.
+Since researchers began using PCR to amplify 16S rRNA gene fragments
+there has been concern that amplifying fragments from a mixed template
+pool could lead to a biased representation in the pool of products and
+would confound downstream analyses (24, 27--33). The mock community was
+generated by mixing equal amounts of genomic DNA from each of the 8
+bacteria resulting in uneven representation of the \emph{rrn} operons
+across the bacteria because each bacterium had a different genome size
+and varied in the number of operons in its genome. The vendor of the
+mock community subjects each lot of genomic DNA to shotgun sequencing to
+more accurately quantify the actual abundance of each organism in the
+community. We compared the expected relative abundance of the 16S rRNA
+genes from each bacterium in the mock community to the data we generated
+across rounds of amplification and polymerase (Figure 3). Interestingly,
+for some bacteria, their representation became less biased with
+additional rounds of PCR (e.g.~\emph{L. fermentum}), while others became
+more biased (e.g.~\emph{E. faecalis}), and others had little change
+(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), The
+percentage of bases in the V4 region that were guanines or cytosines was
+not predictive of the amount of bias. Across the strains there was no
+variation in the length of their V4 regions and they each had the same
+sequence in the region that the primers annealed. One of the bacteria
+represented in the mock community, \emph{S. enterica}, had 6 identical
+copies of the V4 region and 1 copy that differed from those by one
+nucleotide. The dominant copy had a thymidine and the rare copy had a
+guanine. We used the sequence data to calculate the ratio of the
+dominant to rare variants from \emph{S. enterica} (Figure S1). The
+Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of
+5.4; however, the ratio for the Kappa polymerase varied between 6.1 and
+6.8 for 25 to 35 rounds of PCR. Given the subtle nature of the variation
+in the relative abundances of each 16S rRNA gene fragment, it was not
+possible to create generalizable rules that would explain the bias.
 
 \textbf{\emph{At the community level, the effects of PCR amplification
-bias grow with additional rounds of PCR.}} Because the distance between
-samples and across rounds of PCR could be artificially inflated due to
-sequencing errors and chimeras, we used analyzed the alpha and beta
-diversity of the mock community data at different phases of the sequence
-curation pipeline (Figure 4). First, we removed the chimeras from the
-mock community data as described above and mapped the individual reads
-to the OTUs that the 16S rRNA gene fragments would cluster into if there
-were no sequencing errors. This gave us a community distribution that
-reflected the distribution following PCR without any artifacts. Alpha
-diversity metrics such as richness and Shannon and Inverse Simpson
-diversity metrics increased with the number of rounds of PCR for all
-polymerases except for the Kappa polymerase, where the richness remained
-constant but the diversity decreased (Figure 4A). These data suggest
-that PCR had the effect of making the community distribution more even
-than it was originally. Next, we used the true sequence diversity, but
-removed chimeras based on mapping reads to all possible chimeras between
-reference sequences, and clustered the reads to OTUs. The same trends
-were observed as with perfect sequence data except that the richness and
-diversity metrics trended higher (Figure 4A). Finally, we used the
-observed sequence data and used the UCHIME algorithm to identify
-chimeras. Again, we observed the same trends as with the other analyses
-except the richness and diversity metrics trended higher (Figure 4A).
-These comparisons demonstrated that although the bias at the sequence
-level was subtle, PCR does have a bias that is exacerbated by errors and
-chimeras. When we measured the Bray-Curtis distance between the
-communities observed after 25 rounds of amplification and those at 30
-and 35 rounds of amplification the distances between 25 and 35 rounds
-were higher than between 25 and 30 rounds for each of the polymerases by
-an average of 0.022 units (Figure 4B). The Platinum polymerase varied
-the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
-rounds: 0.15). Although the distances between samples were small, the
-ordination of these distances showed a clear change in community
-structure with increasing rounds of PCR (Figure 4C). This observation
-was supported by our statistical analysis, which revealed that the
-number of rounds of PCR (R\textsuperscript{2}=0.21, P\textless{}0.001)
-was a larger factor than the choice of polymerase
-(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
-demonstrate that subtle differences in relative abundances can have an
-impact on overall community structure. This variation underscores the
-importance of only comparing sequence data that have been generated
-using the same PCR conditions.
+bias grow with additional rounds of PCR.}} Because the variation in bias
+between polymerases and across rounds of PCR could be artificially
+inflated due to sequencing errors and chimeras, we analyzed the alpha
+and beta diversity of the mock community data at different phases of the
+sequence curation pipeline (Figure 4). First, we removed the chimeras
+from the mock community data as described above and mapped the
+individual reads to the OTUs that the 16S rRNA gene fragments would
+cluster into if there were no sequencing errors. This gave us a
+community distribution that reflected the distribution following PCR
+without any artifacts. Although the richness did not change, the Shannon
+diversity increased with the number of rounds of PCR for all polymerases
+except for the Kappa polymerase, for which the diversity decreased
+(Figure 4A). These data suggest that PCR had the effect of making the
+community distribution more even than it was originally, except for the
+data generated using the Kappa polymerase where the evenness decreased.
+Next, we used the observed sequence errors, but removed chimeras using
+our reference-based approach, and clustered the reads to OTUs. The
+richness and diversity metrics trended higher with higher error rates
+and number of rounds of PCR (Figure 4A). Finally, we used the observed
+sequence data and used the UCHIME algorithm to identify chimeras. Again,
+the richness and diversity metrics trended higher with higher error
+rates and number of rounds of PCR (Figure 4A). These comparisons
+demonstrated that although the bias at the sequence level was subtle,
+PCR introduces bias that is exacerbated by errors and chimeras when
+sequences are clustered into OTUs. When we measured the Bray-Curtis
+distance between the communities observed after 25 rounds of
+amplification and those at 30 and 35 rounds of amplification the
+distances between 25 and 35 rounds were higher than between 25 and 30
+rounds for each of the polymerases by an average of 0.022 units (Figure
+4B). The Platinum polymerase varied the most across rounds of
+amplification (25 vs 30 rounds: 0.12; 25 vs 35 rounds: 0.15). Although
+the distances between samples were small, the ordination of these
+distances showed a clear change in community structure with increasing
+rounds of PCR (Figure 4C). This observation was supported by our
+statistical analysis, which revealed that the number of rounds of PCR
+(R\textsuperscript{2}=0.21, P\textless{}0.001) was a larger factor than
+the choice of polymerase (R\textsuperscript{2}=0.20, P\textless{}0.001).
+These results demonstrate that subtle differences in relative abundances
+can have an impact on overall community structure. This variation
+underscores the importance of only comparing sequence data that have
+been generated using the same PCR conditions.
 
@@
 metrics using the human stool samples for each of the polymerases and
-rounds of amplification. We calculated the number of observed OTUs,
-Shannon diversity, inverse Simpson diversity index for each condition
-and donor (Figure 5A). Although there were clear differences between
-conditions, the relative ordering of the stool samples did not
-meaningfully vary across conditions. When we characterized the variation
-between rounds of amplification using human stool samples, the distance
-between the 25 and 30 rounds and 25 and 35 rounds varied considerably
-between samples and polymerases (Figure 5B). In general the inter-round
-variation was lowest for the data generated using the Kappa and
-Accuprime polymerases. The data generated using the Platinum polymerase
-was consistent across rounds, but it was more biased than these
-polymerases. Considering the average distance across the four samples
-varied between 0.39 and 0.56, regardless of the polymerases and number
-of rounds of amplification, any bias due to amplification is unlikely to
-obscure community-wide differences between samples. In support of this
-was our principle coordinates analysis of the Bray-Curtis distances,
-which revealed distinct clusters by donor (Figure 5C). Within each
-cluster there were no obvious patterns related to the polymerase or
-number of rounds of PCR. However, our statistical analysis using adonis
-revealed statistically significant differences in the community
-structures with the subject explaining the most variation
-(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
-of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
-choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
-Together, these results indicate that for a coarse analysis of
-communities, the choice of number of rounds of amplification or
-polymerase are not important, but that they must be consistent across
-samples. It is difficult to develop a specific recommendation based on
-the level of bias across rounds of PCR or polymerases; however, the
-general suggestion is to use as few rounds of amplification as possible.
+rounds of amplification. We calculated the number of observed OTUs and
+Shannon diversity for each condition and subject (Figure 5A). Although
+there were clear differences between conditions, the relative ordering
+of the stool samples did not meaningfully vary across conditions. When
+we characterized the variation between rounds of amplification using
+human stool samples, the distance between the 25 and 30 rounds and 25
+and 35 rounds varied considerably between samples and polymerases
+(Figure 5B). In general the inter-round variation was lowest for the
+data generated using the Kappa and Accuprime polymerases. The data
+generated using the Platinum polymerase was consistent across rounds,
+but overall, it was more biased than the other polymerases. Considering
+the average distance across the four samples varied between 0.39 and
+0.56, regardless of the polymerases and number of rounds of
+amplification, any bias due to amplification is unlikely to obscure
+community-wide differences between samples. In support of this was our
+principle coordinates analysis of the Bray-Curtis distances, which
+revealed distinct clusters by donor (Figure 5C). Within each cluster
+there were no obvious patterns related to the polymerase or number of
+rounds of PCR. However, our statistical analysis revealed statistically
+significant differences in the community structures with the subject
+explaining the most variation (R\textsuperscript{2}=0.79,
+P\textless{}0.001), followed by the number of rounds of PCR
+(R\textsuperscript{2}=0.044, P\textless{}0.001) and the choice of
+polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001). Together,
+these results indicate that for a coarse analysis of communities, the
+choice of number of rounds of amplification or polymerase are not
+important, but that they must be consistent across samples. It is
+difficult to develop a specific recommendation based on the level of
+bias across rounds of PCR or polymerases; however, the general
+suggestion is to use as few rounds of amplification as possible.
 
@@
 conditions could result in different representations of communities
-because of random drift. To test this, we determined the average
-Bray-Curtis distance between replicate reactions using the same
-polymerase and number of rounds of amplification (Figure 6). Using the
-mock community data there were no obvious trends. The average
-Bray-Curtis distance within a set of conditions varied by 0.063 to 0.11
-units. Although we did not obtain replicates of each of the stool
-samples, the intra-sample variation for each set of conditions was
-consistent and varied between 0.50 and 0.56 units. These data suggest
-that amplicon sequencing is robust to random variation in amplification
-and that any differences are likely to be smaller than what is
-considered biologically relevant.
+because of random drift over the course of PCR. To test this, we
+determined the average Bray-Curtis distance between replicate reactions
+using the same polymerase and number of rounds of amplification (Figure
+6). Using the mock community data there were no obvious trends. The
+average Bray-Curtis distance within a set of conditions varied by 0.063
+to 0.11 units. Although we did not generate technical replicates of each
+of the stool samples, the inter-sample variation for each set of
+conditions was consistent and varied between 0.50 and 0.56 units. These
+data suggest that amplicon sequencing is robust to random variation in
+amplification and that any differences are likely to be smaller than
+what is considered biologically relevant.
 
@@
 than the biological differences between individuals. Based on these
-observations, within a study, amplicons must be generated by consistent
-protocols to yield meaningful comparisons. When comparing across
-studies, values like richness, diversity, and relative abundances must
-be made in relative and not absolute terms.
+observations, amplicons must be generated by consistent protocols to
+yield meaningful comparisons. When comparing across studies, values like
+richness, diversity, and relative abundances must be made in relative
+and not absolute terms.
 
@@
 errors. Although these methods remove residual errors, they also risk
-splitting intragenomic variants into separate ASVs, merging 16S rRNA
+splitting intragenomic variants into separate ASVs or merging 16S rRNA
 gene sequences from different taxa into the same ASV. Other tools are
@@
 (40, 42). Others recommend removing any read that has an abundance below
-a specified threshold (e.g.~removing all sequences that only appear
-once) (20, 40--42). This method must be approached with caution as such
-approaches are likely to introduce a different bias of the community
-representation and ignore the fact that artifacts may be quite abundant.
-Ultimately, researchers must test their hypotheses with multiple methods
-to validate the claims they reach with any one method (43). All methods
-have biases and limitations and we must use complementary methods to
-obtain robust results.
+a specified threshold as a tool to remove PCR and sequencign artifacts
+(e.g.~removing all sequences that only appear once) (20, 40--42). This
+method must be approached with caution as such approaches are likely to
+introduce a different bias of the community representation and ignore
+the fact that artifacts may be quite abundant. Ultimately, researchers
+must test their hypotheses with multiple methods to validate the claims
+they reach with any one method (43). All methods have biases and
+limitations and we must use complementary methods to obtain robust
+results.
 
@@
 (v 3.5.0) software package (48). Data transformation and graphing were
-completed using the tidyverse package (v 1.2.1) (49). The distance
-matrix data was analyzed using the adonis function within the vegan
-package (v 2.5.3) (50).
+completed using the tidyverse package (v 1.2.1). The distance matrix
+data was analyzed using the adonis function within the vegan package (v
+2.5.3).
 
@@
 
-\newpage
-
 \hypertarget{acknowledgements}{%
@@
 
-The authors thank the study participants in ERIN whose samples were
-utilized. We also would like to thank Judy Opp and April Cockburn for
-their effort in sequencing the samples as part of the Microbiome Core
-Facility at the University of Michigan. Additional thanks to members of
-the Schloss lab and Dr.~Marcy Balunas for reading earlier drafts of the
-manuscript and providing helpful critiques. Salary support for Marc A.
-Sze came from the Canadian Institute of Health Research and NIH grant
-UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants
-P30DK034933 and 1R01CA215574.
+We appreciate the willingness of the donors to provide stool samples. We
+also thank Judy Opp and April Cockburn assistance in sequencing the
+samples as part of the Microbiome Core Facility at the University of
+Michigan. Additional thanks to members of the Schloss lab and Dr.~Marcy
+Balunas for reading earlier drafts of the manuscript and providing
+helpful critiques. Salary support for Marc A. Sze came from the Canadian
+Institute of Health Research and NIH grant UL1TR002240. Salary support
+for Patrick D. Schloss came from NIH grants P30DK034933 and
+1R01CA215574.
 
@@
 \leavevmode\hypertarget{ref-HMP2012}{}%
-2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
-the healthy human microbiome. Nature \textbf{486}:207--214.
+2. \textbf{Human Microbiome Consortium}. 2012. Structure, function and
+diversity of the healthy human microbiome. Nature \textbf{486}:207--214.
 doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
@@
 template annealing in the amplification of mixtures of 16S rRNA genes by
-pcr. Applied and environmental microbiology \textbf{62}:625--630.
+PCR. Applied and environmental microbiology \textbf{62}:625--630.
 
@@
 
-\leavevmode\hypertarget{ref-Tidyverse2017}{}%
-49. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
-'tidyverse'.
-
-\leavevmode\hypertarget{ref-Vegan2018}{}%
-50. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
-\textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
-\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Simpson GL},
-\textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
-\textbf{Wagner H}. 2018. Vegan: Community ecology package.
-
 \newpage
@@
 
-\includegraphics[width=0.9\columnwidth]{../results/figures/mock_error.pdf}
+\includegraphics{../results/figures/mock_error.pdf}
 
@@
 
-\includegraphics[width=0.9\columnwidth]{../results/figures/chimera_plots.pdf}
+\includegraphics[width=1.0\columnwidth]{../results/figures/chimera_plots.pdf}
 
@@
 
-\includegraphics[width=0.9\columnwidth]{../results/figures/species_bias.pdf}
+\includegraphics[width=1.0\columnwidth]{../results/figures/species_bias.pdf}
 
@@
 metrics that grew with the number of rounds of PCR when using a mock
-community.} (A) With the exception of data collected using the Kappa
-polymerase, the OTU richness and Shannon diversity values increased with
-number of rounds of PCR and the inclusion of residual sequencing errors
-and chimeras. The horizontal black line indicates the expected richness
-and diversity if there were no bias, errors, or chimeras. (B) Relative
-to the mock community sampled after 25 rounds of PCR, the Bray-Curtis
+community.} (A) With the exception of the Kappa polymerase data, the
+richness and Shannon diversity values increased with number of rounds of
+PCR and the inclusion of residual sequencing errors and chimeras. The
+horizontal black line indicates the ideal richness and diversity. (B)
+Relative to the mock community sampled after 25 rounds of PCR, the
 distance to the communities sampled after 30 and 35 rounds of PCR
 increased for all polymerases. (C) The variation between samples
-collected after 20, 25, 30, and 35 rounds of PCR for the five
-polymerases demonstrated a significant change in the community driven by
-the number of rounds of PCR and the polymerase used.
-
-\includegraphics[width=0.8\columnwidth]{../results/figures/mock_community.pdf}
+demonstrated a significant change in the community driven by the number
+of rounds of PCR and the polymerase used.
+
+\includegraphics[width=0.9\columnwidth]{../results/figures/mock_community.pdf}
 
@@
 appeared to be consistent within each sample.} (A) With the exception of
-data collected using the Kappa polymerase, the OTU richness and Shannon
+data collected using the Kappa polymerase, the richness and Shannon
 diversity values increased with number of rounds of PCR. (B) Relative to
-the stool communities sampled after 25 rounds of PCR, the Bray-Curtis
-distance to the stool communities sampled after 30 and 35 rounds of PCR
-was inconsistent and there was little difference in variation for data
+the stool communities sampled after 25 rounds of PCR, the distance to
+the stool communities sampled after 30 and 35 rounds of PCR was
+inconsistent and there was little difference in variation for data
 collected using the Kappa polymerase. (C) The variation between stool
@@
 
-\textbf{Figure 6. There is little evidence for PCR drift adversely
-impacting the results of amplicon studies.} The average distance between
-replicates of sequencing the same mock community or between the the
-human stool samples did not vary by number of rounds of PCR or by
-polymerase.
-
-\includegraphics[width=0.9\columnwidth]{../results/figures/drift.pdf}
+\textbf{Figure 6. The average distance between replicates of sequencing
+the same mock community or between the the human stool samples
+(i.e.~drift) did not vary by number of rounds of PCR or by polymerase.}
+
+\includegraphics[width=1.0\columnwidth]{../results/figures/drift.pdf}
 
@@
 
-\includegraphics[width=0.9\columnwidth]{../results/figures/salmonella_bias.pdf}
+\includegraphics[width=1.0\columnwidth]{../results/figures/salmonella_bias.pdf}
 

### 2019-03-01 code/vegan_analysis.R
Set random seed

@@ -1,6 +1,8 @@
 library(tidyverse)
 library(vegan)
 
+set.seed(19760620) #pds birthday
+
 ####################################################################################################
 
 read_dist <- function(dist_file_name){

### 2019-03-01 submission/manuscript.Rmd
Edit and render

@@
 
-\textbf{\huge Quantifying the impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}
+\textbf{\huge The impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}
 
@@
 
-\textbf{Running title:} Quantifying the effects of PCR
+\textbf{Running title:} Quantifying the effects of PCR conditions
 
@@
 PCR amplification of 16S rRNA genes is a critical, yet under appreciated step in the generation of sequence data to describe the taxonomic composition of microbial communities.
-Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample.
-The choice of PCR cycling conditions are known to impact the outcome of PCR and downstream inferences.
-We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community with known sequences and community structure and human stool samples.
+Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample (i.e.
+bias).
+We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community and human stool samples.
 Although it was impossible to derive specific recommendations, we did observe general trends.
@@
 When analyzing mock community data, the amount that the community deviated from the expected composition increased with rounds of PCR.
-The effect was inconsistent for human stool samples.
+This bias was inconsistent for human stool samples.
 Overall the results underscore the difficulty of comparing sequence data that are generated by different PCR protocols.
@@
 The effects are relatively small compared to the variation in human stool samples, however, care should be taken to use polymerases with the highest possible fidelity and to minimize the number of rounds of PCR.
-These results underscore the difficulty of directly comparing the sequence data generated using different PCR protocols or interpreting results in an absolute sense.
+These results also underscore that it is not possible to directly compare sequence data generated under different PCR conditions.
 
@@
 This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample.
-Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017].
+Many factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017].
 The increased sampling depth that is now available relative to previous Sanger sequencing-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.
@@
 Factors such as the choice of primers have an obvious impact on which populations will be amplified [@Parada2015; @Kozich2013].
-However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras [@Haas2011; @Wang1996; @Potapov2017; @Kebschull2015], misincorporation of nucleotides [@Potapov2017; @McInerney2014; @Cline1996], preferential amplification of populations leading to bias [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998], and accumulation of random amplification events leading to PCR drift [@Kebschull2015; @Kennedy2014; @Acinas2005; @Wagner1994].
-Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011].
+However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras [@Haas2011; @Wang1996; @Potapov2017; @Kebschull2015], misincorporation of nucleotides [@Potapov2017; @McInerney2014; @Cline1996], preferential amplification of some populations over others leading to bias [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998], and accumulation of random amplification events leading to PCR drift [@Kebschull2015; @Kennedy2014; @Acinas2005; @Wagner1994].
+Many bioinformatic tools have been developed to identify chimeras; however, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011].
 Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017].
-Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015]; however, these have been developed for idealized situations.
-To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift [@Kennedy2014].
+Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015].
+In cases where there has been success it has been with relatively small communities with consistent composition [@Brooks2015].
+To minimize PCR drift, some investigators pool technical replicate PCRs hoping to average out the drift [@Kennedy2014].
 Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community [@Polz1998; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998].
@@
 This study investigated the impact of choice of high fidelity DNA polymerase and the number of rounds of amplification on the formation of PCR artifacts using a mock community and human stool samples.
+It was hypothesized that additional rounds of PCR would exacerbate the number of artifacts.
 These PCR artifacts included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process.
@@
 We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform.
-This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol [@Kozich].
+This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol [@Kozich2013].
 To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 20, 25, 30, and 35 rounds of amplification.
-Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.
+Collectively, our results suggest that the number of rounds and to a lesser extent the choice of DNA polymerase used in PCR impact the sequence data.
+The effects are consistent and are smaller than the biological differences between individuals.
 
@@
 Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors [@Edgar2011; @Haas2011].
-Because the true sequences of the organisms in the mock community were known, we could generate all possible chimeras between pairs of V4 16S rRNA gene fragments and use these possible chimeric sequences to screen the sequences generated under the different PCR conditions for chimeras.
+Because the true sequences of the organisms in the mock community were known, we generated all possible chimeras between pairs of V4 16S rRNA gene fragments and used these possible chimeric sequences to screen the sequences generated under the different PCR conditions for chimeras.
 The number of chimeras increased with rounds of amplification (Figure 2A).
@@
 By comparing the output of UCHIME to our reference-based approach, we were able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
-The specificity for all polymerases was above `r min_specificity`% and showed a weak association with the number of cycles used.
+The specificity for all polymerases was above `r min_specificity`% and showed a weak association with the number of cycles used (Figure 2A).
 There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UCHIME to detect the chimeras from the mock community.
-This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity.
-The sensitivity of UCHIME varied between `r sensitivity["min"]` and `r sensitivity["max"]`%.
+This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity of UCHIME to detect chimeras.
+The sensitivity of UCHIME varied between `r sensitivity["min"]` and `r sensitivity["max"]`% when at least 25 cycles were used.
 The generalizability of these results is limited because we used a single mock community with limited genetic diversity.
@@
 These results followed those from the mock communities: additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used.
-Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.
+Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impact of chimeras.
 
@@
 ***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to amplify 16S rRNA gene fragments there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998].
-The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria because each bacterium had a different genome size and varied in the number of operons in its genome.
+The mock community was generated by mixing equal amounts of genomic DNA from 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria as each bacterium had a different genome size and varied in the number of operons in its genome.
 The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community.
@@
 subtilis*).
-Contrary to prior reports [@Polz1998], The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
+Contrary to prior reports [@Polz1998], the percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias.
 Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed.
@@
 We used the sequence data to calculate the ratio of the dominant to rare variants from *S.
-enterica* (Figure S1).
+enterica* expecting a ratio near 6 (Figure S1).
 The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of `r others`; however, the ratio for the Kappa polymerase varied between `r kappa["min"]` and `r kappa["max"]` for 25 to 35 rounds of PCR.
@@
 This gave us a community distribution that reflected the distribution following PCR without any artifacts.
-Although the richness did not change, the Shannon diversity increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, for which the diversity decreased (Figure 4A).
+Although the richness did not change, the Shannon diversity increased with the number of rounds of PCR for all polymerases except the Kappa polymerase, for which the diversity decreased (Figure 4A).
 These data suggest that PCR had the effect of making the community distribution more even than it was originally, except for the data generated using the Kappa polymerase where the evenness decreased.
@@
 The richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A).
-Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras.
+Finally, we used the observed sequence data and the UCHIME algorithm to identify chimeras.
 Again, the richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A).
-These comparisons demonstrated that although the bias at the sequence level was subtle, PCR introduces bias that is exacerbated by errors and chimeras when sequences are clustered into OTUs.
-When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B).
+These comparisons demonstrated that although the bias at the sequence level was subtle, PCR introduces bias at the community level that is exacerbated by errors and chimeras when sequences are clustered into OTUs.
+When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35, distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B).
 The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: `r filter(pl, col_round=="30x") %>% select(distances)`; 25 vs 35 rounds: `r filter(pl, col_round=="35x") %>% select(distances)`).
 Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR (Figure 4C).
-This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=`r filter(mock_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) was a larger factor than the choice of polymerase (R^2^=`r filter(mock_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
+This observation was supported by our statistical analysis, which revealed that the effect of the number of rounds of PCR (R^2^=`r filter(mock_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) was comparable to the choice of polymerase (R^2^=`r filter(mock_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
 These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure.
@@
 
-***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels using mock community data appeared to be small relative to the expected differences between biological samples.
+***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** We expected that the biases that we observed at the population and community levels using mock community data would be small relative to the expected differences between biological samples.
 To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification.
-We calculated the number of observed OTUs and Shannon diversity for each condition and subject (Figure 5A).
-Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
+We calculated the number of observed OTUs and Shannon diversity for each condition and stool sample (Figure 5A).
+Although there were clear differences between PCR conditions, the relative ordering of the stool samples did not meaningfully vary across conditions.
 When we characterized the variation between rounds of amplification using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases (Figure 5B).
@@
 Considering the average distance across the four samples varied between `r inter_diversity["min"]` and `r inter_diversity["max"]`, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples.
-In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by donor (Figure 5C).
+In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by stool sample (Figure 5C).
 Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR.
-However, our statistical analysis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=`r filter(stool_vegan, effects=="subject") %>% select(r_sq)`, P<0.001), followed by the number of rounds of PCR (R^2^=`r filter(stool_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) and the choice of polymerase (R^2^=`r filter(stool_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
+However, our statistical analysis revealed statistically significant differences in the community structures with the stool sample explaining the most variation (R^2^=`r filter(stool_vegan, effects=="subject") %>% select(r_sq)`, P<0.001), followed by the number of rounds of PCR (R^2^=`r filter(stool_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) and the choice of polymerase (R^2^=`r filter(stool_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001).
 Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.
@@
 Researchers should strive to minimize the number of rounds of PCR and should use a high fidelity polymerase.
-Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between individuals.
+Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between the samples we tested.
 Based on these observations, amplicons must be generated by consistent protocols to yield meaningful comparisons.
@@
 Among these polymerases, the Kappa polymerase resulted in the the lowest error rate, lowest chimera rate, and least bias across rounds of PCR.
-Considering polymerase development is an active area of commercial development and it likely that new polymerases will come to market, it is important for researchers to understand how changing the polymerase impacts downstream analyses for their type of samples.
+Considering polymerase development is an active area of commercial development with potential new polymerases becoming available, it is important for researchers to understand how changing the polymerase impacts downstream analyses for their type of samples.
 
-Over the past 20 years, a large literature has attempted to document various PCR biases and underscores the fact that data based on amplification of DNA from a mixed community are not a true representation of the actual community.
+Over the past 20 years, a large literature has attempted to document various PCR biases and underscored the fact that data based on amplification of DNA from a mixed community are not a true representation of the actual community.
 In addition to obvious biases imposed by primer selection, other factors inherent in PCR can influence the representation of communities.
@@
 Minimizing the level of DNA shearing and using the fewest number of rounds of PCR with a polymerase that has the highest possible fidelity are strategies that can be employed to minimize the formation of chimeras.
-Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are smaller than the actual biological variation in fecal communities between individuals.
-Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.
+Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that variation among polymerases is  smaller than the actual biological variation in fecal communities between individuals.
 
@@
 In recent years, parameters for these algorithms have been changed to increase their sensitivity with little evaluation of the effects on the specificity of the algorithms [@Callahan2016; @Amir2017a].
-Others recommend removing any read that has an abundance below a specified threshold as a tool to remove PCR and sequencign artifacts (e.g.
+Others recommend removing any read that has an abundance below a specified threshold as a tool to remove PCR and sequencing artifacts (e.g.
 removing all sequences that only appear once) [@Bokulich2012; @Callahan2016; @Edgar2016; @Amir2017a].
@@
 Primers targeting the V4 region of the 16S rRNA gene were used with modifications to generate MiSeq amplicon libraries [@Kozich2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/).
-The number of rounds of PCR used for each sample and polymerase 15 and increased by 5 rounds up to 35 cycles.
+The number of rounds of PCR used for each sample and polymerase started at 15 and increased by 5 rounds up to 35 cycles.
 Insufficient PCR product was generated using 15 rounds and has not been included in our analysis.
@@
 combination of polymerase and number of rounds of PCR) were replicated four times for the mock community and one time for each fecal sample.
-Libraries were generated as described previously [@Kozich2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/).
+Libraries were generated as previously described [@Kozich2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/).
 The libraries were sequenced using the Illumina MiSeq sequencing platform to generate paired 250-nt reads.
@@
 We appreciate the willingness of the donors to provide stool samples.
-We also thank Judy Opp and April Cockburn assistance in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
+We also thank Judy Opp and April Cockburn for their assistance in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan.
 Additional thanks to members of the Schloss lab and Dr.
 Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques.
-Salary support for Marc A.
-Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240.
-Salary support for Patrick D.
-Schloss came from NIH grants P30DK034933 and 1R01CA215574.
+Support for MAS came from the Canadian Institute of Health Research and NIH grant UL1TR002240 and support for PDS came from NIH grants P30DK034933 and 1R01CA215574.
 
@@
 **Figure 1.
-The error rate of assembled sequence reads increases with the number of rounds of PCR used and follows the relative error rates provided by the manufacturers; however, much of this error is mediated by denoising.**
+The error rate of assembled sequence reads increases with the number of rounds of PCR; however, much of this error was eliminated by denoising and followed the relative error rates provided by the manufacturers.**
 
@@
 Despite evidence of subtle PCR bias at the genome level, there was significant evidence of bias using community-wide metrics that grew with the number of rounds of PCR when using a mock community.** (A) With the exception of the Kappa polymerase data, the richness and Shannon diversity values increased with number of rounds of PCR and the inclusion of residual sequencing errors and chimeras.
-The horizontal black line indicates the ideal richness and diversity.
+The horizontal black line indicates the expected richness and diversity.
 (B) Relative to the mock community sampled after 25 rounds of PCR, the distance to the communities sampled after 30 and 35 rounds of PCR increased for all polymerases.
@@
 
-**Figure S1: With the exception of the seuqence data generated using the Kappa polymerase, the ratio of the two *Salmonella enterica* V4 sequences to a value was lower than the expected ratio of 6:1.** The two *S.
+**Figure S1: With the exception of the sequence data generated using the Kappa polymerase, the ratio of the two *Salmonella enterica* V4 sequences was lower than the expected ratio of 6:1.** The two *S.
 enterica* V4 sequences differed by a single base.

### 2019-03-01 submission/manuscript.tex

@@
 sequences, and the degree to which the fragments in the product
-represent their abundance in the original sample. The choice of PCR
-cycling conditions are known to impact the outcome of PCR and downstream
-inferences. We compared the performance of high fidelity polymerases and
-varying number of rounds of amplification when amplifying a mock
-community with known sequences and community structure and human stool
-samples. Although it was impossible to derive specific recommendations,
-we did observe general trends. Namely, using a polymerase with the
-highest possible fidelity and minimizing the number of rounds of PCR
-reduced the sequencing error rate, fraction of chimeric sequences, and
-bias. Evidence of bias at the sequence level was subtle and could not be
-ascribed to the fragments' fraction of bases that were guanines or
-cytosines. When analyzing mock community data, the amount that the
-community deviated from the expected composition increased with rounds
-of PCR. The effect was inconsistent for human stool samples. Overall the
-results underscore the difficulty of comparing sequence data that are
-generated by different PCR protocols. However, the results indicate that
-the variation in human stool samples is generally larger than that
-introduced by the choice of polymerase or number of rounds of PCR.
+represent their abundance in the original sample (i.e.~bias). We
+compared the performance of high fidelity polymerases and varying number
+of rounds of amplification when amplifying a mock community and human
+stool samples. Although it was impossible to derive specific
+recommendations, we did observe general trends. Namely, using a
+polymerase with the highest possible fidelity and minimizing the number
+of rounds of PCR reduced the sequencing error rate, fraction of chimeric
+sequences, and bias. Evidence of bias at the sequence level was subtle
+and could not be ascribed to the fragments' fraction of bases that were
+guanines or cytosines. When analyzing mock community data, the amount
+that the community deviated from the expected composition increased with
+rounds of PCR. This bias was inconsistent for human stool samples.
+Overall the results underscore the difficulty of comparing sequence data
+that are generated by different PCR protocols. However, the results
+indicate that the variation in human stool samples is generally larger
+than that introduced by the choice of polymerase or number of rounds of
+PCR.
 
@@
 polymerases with the highest possible fidelity and to minimize the
-number of rounds of PCR. These results underscore the difficulty of
-directly comparing the sequence data generated using different PCR
-protocols or interpreting results in an absolute sense.
+number of rounds of PCR. These results also underscore that it is not
+possible to directly compare sequence data generated under different PCR
+conditions.
 
@@
 where it is possible to characterize numerous samples with thousands of
-sequences per sample. Numerous factors can impact how a natural
-community is represented by the sequencing data including the method of
-acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
-methods (13), amplification conditions (8, 14, 15), sequencing method
-(15--17), and analytical pipeline (15, 18--20). The increased sampling
-depth that is now available relative to previous Sanger sequencing-based
-methods is expected to compound the impacts of an investigator's choices
-and the interpretation of their results.
+sequences per sample. Many factors can impact how a natural community is
+represented by the sequencing data including the method of acquiring
+samples (4--8), storage conditions (4--6, 9--12), extraction methods
+(13), amplification conditions (8, 14, 15), sequencing method (15--17),
+and analytical pipeline (15, 18--20). The increased sampling depth that
+is now available relative to previous Sanger sequencing-based methods is
+expected to compound the impacts of an investigator's choices and the
+interpretation of their results.
 
@@
 formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
-25, 26), preferential amplification of populations leading to bias (24,
-27--33), and accumulation of random amplification events leading to PCR
-drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
-identify chimeras, there are significant sensitivity and specificity
-tradeoffs (14, 35). Laboratory-based solutions to minimize chimera
-formation have also been proposed such as minimizing the amount of
-template DNA in the PCR, minimizing the number of rounds of PCR,
-minimizing the amount of shearing in the template DNA, and using DNA
-polymerases that have a proof-reading ability (14, 23). Others have
-attempted to account for PCR bias using modeling approaches (29, 36);
-however, these have been developed for idealized situations. To minimize
-PCR drift, some investigators will pool technical replicate PCRs hoping
-to average out the drift (34). Other factors that have been shown to
-impact the formation of PCR artifacts are outside the control of an
+25, 26), preferential amplification of some populations over others
+leading to bias (24, 27--33), and accumulation of random amplification
+events leading to PCR drift (24, 27, 32, 34). Many bioinformatic tools
+have been developed to identify chimeras; however, there are significant
+sensitivity and specificity tradeoffs (14, 35). Laboratory-based
+solutions to minimize chimera formation have also been proposed such as
+minimizing the amount of template DNA in the PCR, minimizing the number
+of rounds of PCR, minimizing the amount of shearing in the template DNA,
+and using DNA polymerases that have a proof-reading ability (14, 23).
+Others have attempted to account for PCR bias using modeling approaches
+(29, 36). In cases where there has been success it has been with
+relatively small communities with consistent composition (29). To
+minimize PCR drift, some investigators pool technical replicate PCRs
+hoping to average out the drift (34). Other factors that have been shown
+to impact the formation of PCR artifacts are outside the control of an
 investigator including the fraction of DNA bases that are guanines or
@@
 polymerase and the number of rounds of amplification on the formation of
-PCR artifacts using a mock community and human stool samples. These PCR
-artifacts included (i) the effect of the polymerase on the error rate of
-the bases represented in the final sequences, (ii) the fraction of
-sequences that appeared to be chimeras and the ability to detect those
-chimeras, (iii) the bias of preferentially amplifying one fragment over
-another in a mixed pool of templates, and (iv) inter-sample variation in
-community structure of samples amplified with the same polymerase across
-the amplification process. To characterize these factors we sequenced a
-mock community of 8 organisms with known sequences and community
-structure and human fecal samples with unknown sequences and community
-structures. We sequenced the V4 region of the 16S rRNA genes from a mock
-community by generating paired 250 nt reads on the Illumina MiSeq
-platform. This region and sequencing approach was used because it has
-been shown to result in a relatively low sequencing error rate and is a
-widely used protocol ({\textbf{???}}). To better understand the impact
-of DNA polymerase choice on PCR artifacts, we selected five high
-fidelity DNA polymerases and amplified the communities using 20, 25, 30,
-and 35 rounds of amplification. Collectively, our results suggest that
-the number of rounds and to a lesser degree the choice of DNA polymerase
-used in PCR impact the sequence data, the effects are consistent and are
-smaller than the biological differences between individuals.
+PCR artifacts using a mock community and human stool samples. It was
+hypothesized that additional rounds of PCR would exacerbate the number
+of artifacts. These PCR artifacts included (i) the effect of the
+polymerase on the error rate of the bases represented in the final
+sequences, (ii) the fraction of sequences that appeared to be chimeras
+and the ability to detect those chimeras, (iii) the bias of
+preferentially amplifying one fragment over another in a mixed pool of
+templates, and (iv) inter-sample variation in community structure of
+samples amplified with the same polymerase across the amplification
+process. To characterize these factors we sequenced a mock community of
+8 organisms with known sequences and community structure and human fecal
+samples with unknown sequences and community structures. We sequenced
+the V4 region of the 16S rRNA genes from a mock community by generating
+paired 250 nt reads on the Illumina MiSeq platform. This region and
+sequencing approach was used because it has been shown to result in a
+relatively low sequencing error rate and is a widely used protocol (18).
+To better understand the impact of DNA polymerase choice on PCR
+artifacts, we selected five high fidelity DNA polymerases and amplified
+the communities using 20, 25, 30, and 35 rounds of amplification.
+Collectively, our results suggest that the number of rounds and to a
+lesser extent the choice of DNA polymerase used in PCR impact the
+sequence data. The effects are consistent and are smaller than the
+biological differences between individuals.
 
@@
 (14, 35). Because the true sequences of the organisms in the mock
-community were known, we could generate all possible chimeras between
-pairs of V4 16S rRNA gene fragments and use these possible chimeric
-sequences to screen the sequences generated under the different PCR
-conditions for chimeras. The number of chimeras increased with rounds of
-amplification (Figure 2A). Interestingly, the fraction of chimeric
-sequences from the mock community varied by the type of polymerase used.
-After 30 rounds of PCR, the Platinum polymerase had the highest chimera
-rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
-(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
-polymerases. Because our chimera screening procedure can only be applied
-to mock communities, we used the UCHIME algorithm to model the chimera
-screening approach that is used in most sequence curation pipelines. By
-comparing the output of UCHIME to our reference-based approach, we were
-able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
-The specificity for all polymerases was above 95.7\% and showed a weak
-association with the number of cycles used. There was considerable
+community were known, we generated all possible chimeras between pairs
+of V4 16S rRNA gene fragments and used these possible chimeric sequences
+to screen the sequences generated under the different PCR conditions for
+chimeras. The number of chimeras increased with rounds of amplification
+(Figure 2A). Interestingly, the fraction of chimeric sequences from the
+mock community varied by the type of polymerase used. After 30 rounds of
+PCR, the Platinum polymerase had the highest chimera rate (i.e.~18.2\%)
+followed by the Q5 (i.e.~8.1\%), Phusion (i.e.~7.5\%), Kappa
+(i.e.~2.3\%), and Accuprime (i.e.~0.9\%) polymerases. Because our
+chimera screening procedure can only be applied to mock communities, we
+used the UCHIME algorithm to model the chimera screening approach that
+is used in most sequence curation pipelines. By comparing the output of
+UCHIME to our reference-based approach, we were able to calculate the
+UCHIME's sensitivity and specificity (Figure 2A). The specificity for
+all polymerases was above 95.7\% and showed a weak association with the
+number of cycles used (Figure 2A). There was considerable
 inter-polymerase and inter-round of amplification variation in the
@@
 This suggested that the residual error rate after pre-clustering the
-sequence data did not compromise the sensitivity. The sensitivity of
-UCHIME varied between 50 and 87.0\%. The generalizability of these
-results is limited because we used a single mock community with limited
-genetic diversity. Although we did not know the true chimera rate for
-our four human stool samples, we were able to calculate the fraction of
-sequences that UCHIME identified as being chimeric (Figure 2B). These
-results followed those from the mock communities: additional rounds of
+sequence data did not compromise the sensitivity of UCHIME to detect
+chimeras. The sensitivity of UCHIME varied between 50 and 87.0\% when at
+least 25 cycles were used. The generalizability of these results is
+limited because we used a single mock community with limited genetic
+diversity. Although we did not know the true chimera rate for our four
+human stool samples, we were able to calculate the fraction of sequences
+that UCHIME identified as being chimeric (Figure 2B). These results
+followed those from the mock communities: additional rounds of
 amplification significantly increased the rate of chimeras and there was
@@
 higher rates of chimeras, it is clear that using the smallest number of
-PCR cycles possible will minimize the impacts of chimeras.
+PCR cycles possible will minimize the impact of chimeras.
 
@@
 would confound downstream analyses (24, 27--33). The mock community was
-generated by mixing equal amounts of genomic DNA from each of the 8
-bacteria resulting in uneven representation of the \emph{rrn} operons
-across the bacteria because each bacterium had a different genome size
-and varied in the number of operons in its genome. The vendor of the
-mock community subjects each lot of genomic DNA to shotgun sequencing to
-more accurately quantify the actual abundance of each organism in the
+generated by mixing equal amounts of genomic DNA from 8 bacteria
+resulting in uneven representation of the \emph{rrn} operons across the
+bacteria as each bacterium had a different genome size and varied in the
+number of operons in its genome. The vendor of the mock community
+subjects each lot of genomic DNA to shotgun sequencing to more
+accurately quantify the actual abundance of each organism in the
 community. We compared the expected relative abundance of the 16S rRNA
@@
 more biased (e.g.~\emph{E. faecalis}), and others had little change
-(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), The
+(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), the
 percentage of bases in the V4 region that were guanines or cytosines was
@@
 guanine. We used the sequence data to calculate the ratio of the
-dominant to rare variants from \emph{S. enterica} (Figure S1). The
-Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of
-5.4; however, the ratio for the Kappa polymerase varied between 6.1 and
-6.8 for 25 to 35 rounds of PCR. Given the subtle nature of the variation
-in the relative abundances of each 16S rRNA gene fragment, it was not
-possible to create generalizable rules that would explain the bias.
+dominant to rare variants from \emph{S. enterica} expecting a ratio near
+6 (Figure S1). The Accuprime, Phusion, Platinum, and Q5 polymerases
+converged to a ratio of 5.4; however, the ratio for the Kappa polymerase
+varied between 6.1 and 6.8 for 25 to 35 rounds of PCR. Given the subtle
+nature of the variation in the relative abundances of each 16S rRNA gene
+fragment, it was not possible to create generalizable rules that would
+explain the bias.
 
@@
 diversity increased with the number of rounds of PCR for all polymerases
-except for the Kappa polymerase, for which the diversity decreased
-(Figure 4A). These data suggest that PCR had the effect of making the
-community distribution more even than it was originally, except for the
-data generated using the Kappa polymerase where the evenness decreased.
-Next, we used the observed sequence errors, but removed chimeras using
-our reference-based approach, and clustered the reads to OTUs. The
-richness and diversity metrics trended higher with higher error rates
-and number of rounds of PCR (Figure 4A). Finally, we used the observed
-sequence data and used the UCHIME algorithm to identify chimeras. Again,
-the richness and diversity metrics trended higher with higher error
-rates and number of rounds of PCR (Figure 4A). These comparisons
-demonstrated that although the bias at the sequence level was subtle,
-PCR introduces bias that is exacerbated by errors and chimeras when
+except the Kappa polymerase, for which the diversity decreased (Figure
+4A). These data suggest that PCR had the effect of making the community
+distribution more even than it was originally, except for the data
+generated using the Kappa polymerase where the evenness decreased. Next,
+we used the observed sequence errors, but removed chimeras using our
+reference-based approach, and clustered the reads to OTUs. The richness
+and diversity metrics trended higher with higher error rates and number
+of rounds of PCR (Figure 4A). Finally, we used the observed sequence
+data and the UCHIME algorithm to identify chimeras. Again, the richness
+and diversity metrics trended higher with higher error rates and number
+of rounds of PCR (Figure 4A). These comparisons demonstrated that
+although the bias at the sequence level was subtle, PCR introduces bias
+at the community level that is exacerbated by errors and chimeras when
 sequences are clustered into OTUs. When we measured the Bray-Curtis
 distance between the communities observed after 25 rounds of
-amplification and those at 30 and 35 rounds of amplification the
-distances between 25 and 35 rounds were higher than between 25 and 30
-rounds for each of the polymerases by an average of 0.022 units (Figure
-4B). The Platinum polymerase varied the most across rounds of
-amplification (25 vs 30 rounds: 0.12; 25 vs 35 rounds: 0.15). Although
-the distances between samples were small, the ordination of these
-distances showed a clear change in community structure with increasing
-rounds of PCR (Figure 4C). This observation was supported by our
-statistical analysis, which revealed that the number of rounds of PCR
-(R\textsuperscript{2}=0.21, P\textless{}0.001) was a larger factor than
-the choice of polymerase (R\textsuperscript{2}=0.20, P\textless{}0.001).
-These results demonstrate that subtle differences in relative abundances
-can have an impact on overall community structure. This variation
-underscores the importance of only comparing sequence data that have
-been generated using the same PCR conditions.
+amplification and those at 30 and 35, distances between 25 and 35 rounds
+were higher than between 25 and 30 rounds for each of the polymerases by
+an average of 0.022 units (Figure 4B). The Platinum polymerase varied
+the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
+rounds: 0.15). Although the distances between samples were small, the
+ordination of these distances showed a clear change in community
+structure with increasing rounds of PCR (Figure 4C). This observation
+was supported by our statistical analysis, which revealed that the
+effect of the number of rounds of PCR (R\textsuperscript{2}=0.21,
+P\textless{}0.001) was comparable to the choice of polymerase
+(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
+demonstrate that subtle differences in relative abundances can have an
+impact on overall community structure. This variation underscores the
+importance of only comparing sequence data that have been generated
+using the same PCR conditions.
 
@@
 amplification have little impact on the relative interpretation of
-community-wide metrics of diversity.}} The biases that we observed at
-the population and community levels using mock community data appeared
-to be small relative to the expected differences between biological
-samples. To study this further, we calculated alpha and beta-diversity
-metrics using the human stool samples for each of the polymerases and
-rounds of amplification. We calculated the number of observed OTUs and
-Shannon diversity for each condition and subject (Figure 5A). Although
-there were clear differences between conditions, the relative ordering
-of the stool samples did not meaningfully vary across conditions. When
-we characterized the variation between rounds of amplification using
-human stool samples, the distance between the 25 and 30 rounds and 25
-and 35 rounds varied considerably between samples and polymerases
-(Figure 5B). In general the inter-round variation was lowest for the
-data generated using the Kappa and Accuprime polymerases. The data
-generated using the Platinum polymerase was consistent across rounds,
-but overall, it was more biased than the other polymerases. Considering
-the average distance across the four samples varied between 0.39 and
-0.56, regardless of the polymerases and number of rounds of
-amplification, any bias due to amplification is unlikely to obscure
-community-wide differences between samples. In support of this was our
-principle coordinates analysis of the Bray-Curtis distances, which
-revealed distinct clusters by donor (Figure 5C). Within each cluster
-there were no obvious patterns related to the polymerase or number of
-rounds of PCR. However, our statistical analysis revealed statistically
-significant differences in the community structures with the subject
-explaining the most variation (R\textsuperscript{2}=0.79,
-P\textless{}0.001), followed by the number of rounds of PCR
-(R\textsuperscript{2}=0.044, P\textless{}0.001) and the choice of
-polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001). Together,
-these results indicate that for a coarse analysis of communities, the
-choice of number of rounds of amplification or polymerase are not
-important, but that they must be consistent across samples. It is
-difficult to develop a specific recommendation based on the level of
-bias across rounds of PCR or polymerases; however, the general
-suggestion is to use as few rounds of amplification as possible.
+community-wide metrics of diversity.}} We expected that the biases that
+we observed at the population and community levels using mock community
+data would be small relative to the expected differences between
+biological samples. To study this further, we calculated alpha and
+beta-diversity metrics using the human stool samples for each of the
+polymerases and rounds of amplification. We calculated the number of
+observed OTUs and Shannon diversity for each condition and stool sample
+(Figure 5A). Although there were clear differences between PCR
+conditions, the relative ordering of the stool samples did not
+meaningfully vary across conditions. When we characterized the variation
+between rounds of amplification using human stool samples, the distance
+between the 25 and 30 rounds and 25 and 35 rounds varied considerably
+between samples and polymerases (Figure 5B). In general the inter-round
+variation was lowest for the data generated using the Kappa and
+Accuprime polymerases. The data generated using the Platinum polymerase
+was consistent across rounds, but overall, it was more biased than the
+other polymerases. Considering the average distance across the four
+samples varied between 0.39 and 0.56, regardless of the polymerases and
+number of rounds of amplification, any bias due to amplification is
+unlikely to obscure community-wide differences between samples. In
+support of this was our principle coordinates analysis of the
+Bray-Curtis distances, which revealed distinct clusters by stool sample
+(Figure 5C). Within each cluster there were no obvious patterns related
+to the polymerase or number of rounds of PCR. However, our statistical
+analysis revealed statistically significant differences in the community
+structures with the stool sample explaining the most variation
+(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
+of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
+choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
+Together, these results indicate that for a coarse analysis of
+communities, the choice of number of rounds of amplification or
+polymerase are not important, but that they must be consistent across
+samples. It is difficult to develop a specific recommendation based on
+the level of bias across rounds of PCR or polymerases; however, the
+general suggestion is to use as few rounds of amplification as possible.
 
@@
 interpretation of the data, the effects were consistent and were smaller
-than the biological differences between individuals. Based on these
-observations, amplicons must be generated by consistent protocols to
-yield meaningful comparisons. When comparing across studies, values like
-richness, diversity, and relative abundances must be made in relative
-and not absolute terms.
+than the biological differences between the samples we tested. Based on
+these observations, amplicons must be generated by consistent protocols
+to yield meaningful comparisons. When comparing across studies, values
+like richness, diversity, and relative abundances must be made in
+relative and not absolute terms.
 
@@
 and least bias across rounds of PCR. Considering polymerase development
-is an active area of commercial development and it likely that new
-polymerases will come to market, it is important for researchers to
+is an active area of commercial development with potential new
+polymerases becoming available, it is important for researchers to
 understand how changing the polymerase impacts downstream analyses for
@@
 Over the past 20 years, a large literature has attempted to document
-various PCR biases and underscores the fact that data based on
+various PCR biases and underscored the fact that data based on
 amplification of DNA from a mixed community are not a true
@@
 should always be taken when choosing a polymerase for 16S rRNA gene
-sequencing, our observations show that the differences between a variety
-of polymerases are smaller than the actual biological variation in fecal
-communities between individuals. Therefore, if the biological signal of
-interest is similar to differences in fecal bacterial communities found
-between individuals, then the type of high fidelity DNA polymerase used
-will only minimally change the results.
+sequencing, our observations show that variation among polymerases is
+smaller than the actual biological variation in fecal communities
+between individuals.
 
@@
 (40, 42). Others recommend removing any read that has an abundance below
-a specified threshold as a tool to remove PCR and sequencign artifacts
+a specified threshold as a tool to remove PCR and sequencing artifacts
 (e.g.~removing all sequences that only appear once) (20, 40--42). This
@@
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The number of
-rounds of PCR used for each sample and polymerase 15 and increased by 5
-rounds up to 35 cycles. Insufficient PCR product was generated using 15
-rounds and has not been included in our analysis.
+rounds of PCR used for each sample and polymerase started at 15 and
+increased by 5 rounds up to 35 cycles. Insufficient PCR product was
+generated using 15 rounds and has not been included in our analysis.
 
@@
 replicated four times for the mock community and one time for each fecal
-sample. Libraries were generated as described previously (18)
+sample. Libraries were generated as previously described (18)
 (\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The libraries
@@
 \textbf{\emph{Statistical analysis.}} All analysis was done with the R
-(v 3.5.0) software package (48). Data transformation and graphing were
+(v 3.5.1) software package (48). Data transformation and graphing were
 completed using the tidyverse package (v 1.2.1). The distance matrix
 data was analyzed using the adonis function within the vegan package (v
-2.5.3).
+2.5.4).
 
@@
 We appreciate the willingness of the donors to provide stool samples. We
-also thank Judy Opp and April Cockburn assistance in sequencing the
-samples as part of the Microbiome Core Facility at the University of
-Michigan. Additional thanks to members of the Schloss lab and Dr.~Marcy
-Balunas for reading earlier drafts of the manuscript and providing
-helpful critiques. Salary support for Marc A. Sze came from the Canadian
-Institute of Health Research and NIH grant UL1TR002240. Salary support
-for Patrick D. Schloss came from NIH grants P30DK034933 and
-1R01CA215574.
+also thank Judy Opp and April Cockburn for their assistance in
+sequencing the samples as part of the Microbiome Core Facility at the
+University of Michigan. Additional thanks to members of the Schloss lab
+and Dr.~Marcy Balunas for reading earlier drafts of the manuscript and
+providing helpful critiques. Support for MAS came from the Canadian
+Institute of Health Research and NIH grant UL1TR002240 and support for
+PDS came from NIH grants P30DK034933 and 1R01CA215574.
 
@@
 \textbf{Figure 1. The error rate of assembled sequence reads increases
-with the number of rounds of PCR used and follows the relative error
-rates provided by the manufacturers; however, much of this error is
-mediated by denoising.}
+with the number of rounds of PCR; however, much of this error was
+eliminated by denoising and followed the relative error rates provided
+by the manufacturers.}
 
@@
 PCR and the inclusion of residual sequencing errors and chimeras. The
-horizontal black line indicates the ideal richness and diversity. (B)
+horizontal black line indicates the expected richness and diversity. (B)
 Relative to the mock community sampled after 25 rounds of PCR, the
@@
 
-\textbf{Figure S1: With the exception of the seuqence data generated
+\textbf{Figure S1: With the exception of the sequence data generated
 using the Kappa polymerase, the ratio of the two \emph{Salmonella
-enterica} V4 sequences to a value was lower than the expected ratio of
-6:1.} The two \emph{S. enterica} V4 sequences differed by a single base.
+enterica} V4 sequences was lower than the expected ratio of 6:1.} The
+two \emph{S. enterica} V4 sequences differed by a single base.
 

### 2019-03-01 submission/manuscript.tex
Update funding information

@@
 Institute of Health Research and NIH grant UL1TR002240 and support for
-PDS came from NIH grants P30DK034933 and 1R01CA215574.
+PDS came from NIH grants P30DK034933, R01CA215574, and U19AI09087.
 

### 2019-03-01 submission/manuscript.Rmd

@@
 Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques.
-Support for MAS came from the Canadian Institute of Health Research and NIH grant UL1TR002240 and support for PDS came from NIH grants P30DK034933 and 1R01CA215574.
+Support for MAS came from the Canadian Institute of Health Research and NIH grant UL1TR002240 and support for PDS came from NIH grants P30DK034933,  R01CA215574, and U19AI09087.
 
